### 2021-09-07 1.0-change_by_probes.rmd
first commit on nfs

---
title: "Methyl Race Patterns: Probe change frequency"
author: "Jonah Fisher"
date: "7/9/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: sandstone
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE
)

library(minfi)
library(tidyverse)
library(reshape2)
library(plotly)
```

# Get data and filter down to the samples of interest

```{r create data matrices, eval = F}
beta <- readRDS("../general_data/betaqc.rds")
pd <- readRDS("../general_data/pdqc.rds")
pd <- filter(pd, ck6ethrace %in% 1:3)

commonids <- Reduce(intersect, list(pd$idnum[pd$childteen == "C"], pd$idnum[pd$childteen == "T"]))

pdlist <- list(
  child = filter(pd, childteen == "C"),
  teen = filter(pd, childteen == "T")
)

pdlist$child <- pdlist$child[match(commonids, pdlist$child$idnum),]
pdlist$teen <- pdlist$teen[match(commonids, pdlist$teen$idnum),]

betalist <- list(
  child = beta[, match(pdlist$child$MethID, colnames(beta))],
  teen = beta[, match(pdlist$teen$MethID, colnames(beta))]
)

betadiffs <- betalist$teen - betalist$child
betadiffs <- abs(betadiffs)
```

# Different stratification

## Code for ethrace stratification --

```{r create dataframe of differences, eval = F}
changedf <- data.frame(
  probes = rownames(betadiffs),
  All = rowMeans(betadiffs),
  NHW = rowMeans(betadiffs[, pdlist$child$ck6ethrace == 1]),
  NHB = rowMeans(betadiffs[, pdlist$child$ck6ethrace == 2]),
  H = rowMeans(betadiffs[, pdlist$child$ck6ethrace == 3]),
  bins = ceiling(rowMeans(betalist$child) * 10)
)

changelong <- melt(changedf, id = c("probes", "bins"))

```

```{r create data frame of numbers of probes in various change degrees by bins, eval = F}
#We want data points of bin, ethrace group, and count
countdf <- data.frame(
  bin = rep(1:10, 4),
  ethrace = rep(c("All", "NHW", "NHB", "H"), each = 10)
)

getcounts <- function(lowerb, upperb, basemat = changedf, newmat = countdf){
  counts <- apply(newmat, 1, function(rowiter){
    basemat2 <- filter(basemat, bins == as.numeric(rowiter[1]))
    sum(basemat2[rowiter[2]] > lowerb & basemat2[rowiter[2]] < upperb)
  })
  data.frame(
    newmat,
    count = counts,
    change = sprintf("%.3f to %.3f", lowerb, upperb)
  )
}

fullcountdf <- rbind(
  getcounts(0, 0.005),
  getcounts(0.005, 0.01),
  getcounts(0.01, 0.03),
  getcounts(0.03, 0.06),
  getcounts(0.06, 1)
)

saveRDS(fullcountdf, "changecounts_bysire.rds")
```

# Overall count

## Barchart

```{r ggplot of the change stacked bar}
fullcountdf <- readRDS("changecounts_bysire.rds")

p1 <- ggplot(fullcountdf, aes(x = factor(bin, levels = 1:10), y = count, fill = change)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ethrace) +
  xlab("Cutpoint bin of methylation") +
  ylab("Overall frequency of probe count")

ggplotly(p1)

library(DT)
widecountdf <- dcast(fullcountdf, bin + change ~ ethrace, value.var = "count")
write.csv(widecountdf, "probecounts_by_sire.csv", row.names = F)

```

## Data table of overall count

```{r}
datatable(widecountdf)
```

# Proportion of probes within each range of change for every given bin and sire

```{r wide count dff}
widecountdff <- widecountdf
widecountdff[, 3:6] <- lapply(colnames(widecountdff)[3:6], function(x){
  lapply(1:10, function(y){
    vec <- filter(widecountdff, bin == y) %>% select(x) %>% unlist
    freqs <- round(vec / sum(vec), digits = 5)
    names(freqs) <- unique(widecountdff$change)
    freqs
  }) %>% unlist
}) %>% bind_cols

fullcountdff <- melt(widecountdff, id.vars = c("bin", "change"), measure.vars = c("All", "NHW", "NHB", "H"))
colnames(fullcountdff) <- c("bin", "change", "ethrace", "prop")

p2 <- ggplot(fullcountdff, aes(x = factor(bin, levels = 1:10), y = prop, fill = change)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ethrace) +
  xlab("Cutpoint bin of methylation") +
  ylab("Proportion")

ggplotly(p2)

```

## Data table of proportion of probes for given bin and SIRE that are within the change bin (0.000 - 0.005, 0.005 - 0.001, etc...)

```{r}
datatable(widecountdff)
```

### 2021-09-07 Reports/FFAnalyticPlan.Rmd
first commit on nfs

---
title: "Fragile Families Smoking Analysis"
author: "Freida Blostein"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
    highlight: tango
    number_sections: yes
    theme: sandstone
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/")
knitr::opts_knit$set(message=FALSE)
knitr::opts_knit$set(warning=FALSE)
library(tidyr)
library(dplyr)
library(compareGroups)
library(kableExtra)
library(haven)
library(sjlabelled)
library(matchmaker)
```

# Objectives & Hypotheses 
 A) Conduct an ancestry-stratified epigenome wide association study of prenatal maternal smoking and DNA methylation from child saliva samples at age nine and age fifteen. B) Examine differences in methylation trajectories from age nine to fifteen of the most significant methylation sites identified in the epigenome-wide association study between those exposed vs unexposed to prenatal smoking. C) Develop a saliva epigenetic classifier of prenatal smoke exposure using a machine learning approach and support vector machines and evaluate the agreement of the classifier at two time points.

Hypotheses: I hypothesize that different CpG loci will be associated with maternal prenatal smoking in African ancestry groups than in European and Hispanic ancestry groups. I hypothesize that the direction of associations between CpGs and prenatal maternal smoking will stay consistent between ages nine and fifteen but that the magnitude of associations will decline over time. Finally, I hypothesize that classification of prenatal maternal smoking status based upon DNA methylation information from salivary samples will be accurate and consistent at ages nine and fifteen.

# Reading in data and preprocessing:

# # Labeling phenotype data
```{r, phenotypedataReadinLabel, results='hide', warnings=FALSE}
pheno<-readRDS("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/fullpheno.rds")

#Phenotype labeling & filtering to variables of interest
#labels
#to set labels using sjlabelled, needed to fix multibyte characters 
#https://stackoverflow.com/questions/4993837/r-invalid-multibyte-string
find_offending_character <- function(x, maxStringLength=256){  
  print(x)
  for (c in 1:maxStringLength){
    offendingChar <- substr(x,c,c)
    #print(offendingChar) #uncomment if you want the indiv characters printed
    #the next character is the offending multibyte Character
  }    
}
#lapply(attributes(pheno)$var.labels, find_offending_character)
attributes(pheno)$var.labels<-gsub("\xad", "'",  attributes(pheno)$var.labels)
#lapply(attributes(pheno)$var.labels, find_offending_character)
attributes(pheno)$var.labels<-gsub("\x92", "'",  attributes(pheno)$var.labels)
#lapply(attributes(pheno)$var.labels, find_offending_character)
pheno<-set_label(pheno, c(attributes(pheno)$var.labels, 'ID'))
#value labels 
label_list<- attributes(pheno)$label.table
label_list[[12876]]<-NA
pheno<-set_labels(pheno, labels=attributes(pheno)$label.table)
names(label_list)<-attributes(pheno)$names
 pheno<-set_labels(pheno, labels=label_list)
#define vars of interest 
idvars<-c("id", "ff_id")
demovars<-c("cm1age", "cm1bsex", "cm1ethrace", "m1h3", "m1h3a", "m1h3b", "m1i4", "m1i4a", "m1i4b", "cf1age", "cf1ethrace", "f1h3", "f1h3a", "f1h3b")
smokingvars<-c("m1g4", "f1g4", "f2j5", "f2j5a", "f2j7", "f2j7a", "f3j31", "f3j32", "f4j18", "f4j19", "f5g17", "f5g18", "k5f1k", "k5f1l", "k6d40", "k6d41", "k6d41", "k6d42", "k6d43", "k6d45", "k6d46", "k6d47", "m2j5", "m2j5a", "m2j7", "m2j7a", "m4j18", "m4j19", "m5g17", "m5g18", "n5f17", "n5f18", "p3a22", "p3a23", "p3a23a", "p3a23b", "p3a24", "p4a22", "p4a23", "p5h15", "p5h15b", "p5q3cr", "p6h74", "p6h75", "p6h76", "p6h77", "p6h78")
prenataldrugusevar<-c("m1g2", "m1g3", "m1g5", "m1g6")
#filter to these variables
#FF has data inoriginal form +sjlabelled value labels
#since these dont' show in compareGroups
#also will make FF_labeled, where categorical, binary variables
#will be replaced with their labels
FF<-pheno %>% select(one_of(idvars), one_of(demovars), one_of(smokingvars), one_of(prenataldrugusevar))
#create FF_labeled
#read in csv with metadata about variables 
FFmeta<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/FFMetadata_v07.csv")
#create long data dictionary  
FFlong<-FFmeta%>% select(new_name, contains('value'), contains('label')) %>% pivot_longer(cols = contains("value"), names_to="value", values_to="old_values")
FFlong<-FFlong%>%pivot_longer(cols = contains("label"), names_to="label", values_to="new_labels")
recodes_long<-FFlong %>% filter(gsub("value", "", value)==gsub("label", "", label)& !is.na(old_values)) %>% select(new_name, old_values, new_labels) 
#store continuous, binary, categorical etc
FFmyMeta<-FFmeta %>% filter(new_name %in% c(idvars, demovars, smokingvars, prenataldrugusevar))
myContinuous<-FFmyMeta %>% filter(type=="Continuous") %>%pull(new_name); myContinuous<-colnames(FF)[colnames(FF) %in% myContinuous]
myBinary<-FFmyMeta%>%filter(type=="Binary")%>%pull(new_name); myBinary<-colnames(FF)[colnames(FF) %in% myBinary]
myCategorical_O<-FFmyMeta%>%filter(type=="Ordered Categorical")%>%pull(new_name); myCategorical_O<-colnames(FF)[colnames(FF) %in% myCategorical_O]
myCategorical_U<-FFmyMeta%>%filter(type=="Unordered Categorical")%>%pull(new_name); myCategorical_U<-colnames(FF)[colnames(FF) %in% myCategorical_U]
#filter to only vars of interest - not continuous variables  
myRecodes<-recodes_long %>% filter(new_name %in% c(myBinary, myCategorical_O, myCategorical_U))
#apply matchmaker
FF_labeled<-match_df(FF, dictionary=myRecodes, from = 'old_values', to='new_labels', by='new_name')
#reapply correct variable class types 
FF_labeled<-FF_labeled %>% mutate_at(vars(any_of(c(myBinary, myCategorical_O, myCategorical_U))), funs(as.factor(.))) %>% 
          mutate_at(vars(any_of(myContinuous)), funs(as.numeric(na_if(na_if(., "-3"), '-9'))))
#read back in the variable labels 
FF_labeled<-set_label(FF_labeled, get_label(FF))
```

# # Cleaning pdqc data 
# # # Figure out analytic subsets 1) any methylation data; 2) 2-visits of methylation data
```{r, pdqcReadin}
pdqc<-readRDS("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pdqc.rds")
#What FF ids have methylation data?
methy_ids<-pdqc$id
#ids with 9visit, 15visit or both visit
visit9ids<-pdqc%>%filter(childteen=='C')%>%pull(idnum)%>%unique()
visit15ids<-pdqc%>%filter(childteen=='T')%>%pull(idnum)%>%unique()
bothvisit_ids<-visit9ids[visit9ids %in% visit15ids]
visit9only_ids<-visit9ids[!(visit9ids %in%bothvisit_ids)]
visit15only_ids<-visit15ids[!(visit15ids %in%bothvisit_ids)]
```

```{r, createAnalyticSubsetFlags}
#create flag in FF for having methylation data
#and flag for having 2 visits of methylation data
FF_labeled<-FF_labeled %>% 
  mutate(MethylData=ifelse(id %in% methy_ids, "Analysis subset", "Not in analysis subset"),
         TwoVisitData=ifelse(id%in%bothvisit_ids,"Analysis subset - both visits", "Not in both visit analysis subset"))
table(FF_labeled$MethylData)
table(FF_labeled$TwoVisitData)
```

```{r, checkIDlengths}
if((length(visit15only_ids)+length(visit9only_ids)+length(bothvisit_ids)!=length(FF_labeled %>% filter(MethylData=="Analysis subset")%>%pull(id)))){print("WARNING ID LENGTHS DO NOT ADD UP")}
if(length(bothvisit_ids)!=length(FF_labeled %>% filter(TwoVisitData=="Analysis subset - both visits")%>%pull(id))){print("WARNING ID LENGTHS DO NOT ADD UP")}
```

# # # Filter technical replicates (only 1 technical replicate per person-visit + decide based on QC variables - pick the one with the lower probe fail percentage)

```{r}
#technical replicates will be duplicate id + childteen variables from pdqc
pdqc<-pdqc %>% mutate(newID=paste(idnum, childteen, sep='_'))
techrep_ids<- pdqc %>% select(newID) %>% group_by(newID)%>%filter(n()>1)
techreps<-pdqc %>% filter(newID %in% techrep_ids$newID)%>%arrange(idnum,childteen)
hiPctPF<-techreps %>% group_by(newID) %>%filter(probe_fail_pct==max(probe_fail_pct))%>%pull(MethID)
pdqc_clean<-pdqc %>% filter(!(MethID %in% hiPctPF) & childteen!='M')
myMethIDs<-pdqc_clean %>% pull(MethID)
```

# # Methylation summary variables

# # # Clocks

```{r, readInclocks}
clocks<-read.table(file="/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/DNAm_clocks_ffcw.txt", header = T)
colnames(clocks)[1]<-"MethID"
```

# # # Global methylation score

```{r}
betaqc<-readRDS(file="/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/betaqc.rds")
globalmethy<-colMeans(betaqc, na.rm=T)

methyldata<-left_join(pdqc_clean %>% select(MethID, childteen, idnum), left_join(as.data.frame(cbind("MethID"=names(globalmethy), "globalmethylation"=globalmethy)), clocks))
methyldata_wide <- methyldata %>% select(-MethID)%>% pivot_wider(., names_from=childteen, values_from=c(globalmethylation, horvath13, horvath18, pediatric, levine))
```

# # # Different options for methylation scores (islands; shores)

# # Creation of new variables
# # # Binary smoking variable
```{r}
FF_labeled<-FF_labeled %>%mutate(smkPreg_binary=ifelse(m1g4%in% c("1 2+pk/d","2 1<pk<2","3 <1pk/d" ), "Yes", ifelse(m1g4%in%c("-2 Don't know", "-3 Missing", "-9 Not in wave"), NA, ifelse(m1g4== "4 None", "No", "FLAG"))))
xtabs(~m1g4+smkPreg_binary, data=FF_labeled)
```

# # # Ancestry variables

Ancestry is based on csv files of ancestry-specific PCS created by erin (per Jonah meeting Oct 6 2020). If an ID was in {african/european/hispanic}, then the individual is labeled as such. 
```{r}
# read in PCs
euroPC<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/european.csv")
afrPC<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/african.csv")
hisPC<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/hispanic.csv")
allPC<-read.table("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/global_pcs.txt", header=T)
#individuals in each ancestry specific pc group can be labeled as that ancestry for categorical ancestry variable 
FF_labeled<-FF_labeled %>% mutate(ancestry=case_when(MethylData=="Analysis subset" & id %in% hisPC$idnum~"Hispanic ancestry", MethylData=="Analysis subset" & id %in%afrPC$idnum~"African ancestry", MethylData=="Analysis subset" & id %in% euroPC$idnum~"European ancestry", MethylData=="Analysis subset" & !(id %in%c(hisPC$idnum, afrPC$idnum, euroPC$idnum))~"Missing PC data", MethylData!="Analysis subset"~"Not in analysis subset"))
#check ancestry variable
table(FF_labeled$ancestry)
with(FF_labeled, xtabs(~ancestry+cm1ethrace))
MissingPCdata<-FF_labeled %>% filter(ancestry=="Missing PC data")%>%select(id)
save(MissingPCdata, file = "/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/CreatedData/MissingPC.Rdata")
```

There were `r length(MissingPCdata$id)` individuals without PC data.

# A)Conduct an ancestry-stratified epigenome wide association study of prenatal maternal smoking and DNA methylation from child saliva samples at age nine and age fifteen.

# # Comparison of analytic sample - methylation data to Fragile Families sample

```{r, warnings=F}
methyl<-createTable(compareGroups(MethylData~.-id-ff_id, data=FF_labeled), show.all=T)
export2md(methyl)
```

# # Comparison of analytic sample - two visits of methylation data to Fragile Families sample

```{r, warnings=F}
twovisit<-createTable(compareGroups(TwoVisitData~.-id-ff_id, data=FF_labeled), show.all=T)
export2md(twovisit)
```

# # Flow chart of sample exclusions

# # Comparison of variables by ancestry 
```{r, warning=F, message=F}
myFF<-FF_labeled %>% filter(MethylData=="Analysis subset")
myFF<-left_join(myFF, methyldata%>%mutate(id=idnum))
myFF<-set_label(myFF, get_label(FF_labeled))
ancestrytable<-createTable(compareGroups(ancestry~.-id-ff_id-idnum, data=myFF), show.all=T)
export2md(ancestrytable)
```

# # Distribution of main predictor and bivariate tables (stratified by ancestry)

# # # Not stratified by ancestry

```{r, warning=F, message=F}
smk_C<-createTable(compareGroups(smkPreg_binary~.-id-ff_id-idnum,data=myFF%>%filter(childteen=='C'), simplify=F), show.all=T)
smk_T<-createTable(compareGroups(smkPreg_binary~.-id-ff_id-idnum,data=myFF%>%filter(childteen=='T'), simplify=F), show.all=T)
export2md(smk_C)
export2md(smk_T)
```

# # # Bivariate tables stratified by ancestry 
```{r}
smkA_C<-createTable(compareGroups(smkPreg_binary~.-id-ff_id-idnum,data=myFF%>%filter(childteen=='C' & ancestry=='African ancestry'), simplify=F), show.all=T)
smkA_T<-createTable(compareGroups(smkPreg_binary~.-id-ff_id-idnum,data=myFF%>%filter(childteen=='T' & ancestry=='African ancestry'), simplify=F), show.all=T)
export2md(smkA_C)
export2md(smkA_T)

smkE_C<-createTable(compareGroups(smkPreg_binary~.-id-ff_id-idnum,data=myFF%>%filter(childteen=='C' & ancestry=='European ancestry'), simplify=F), show.all=T)
smkE_T<-createTable(compareGroups(smkPreg_binary~.-id-ff_id-idnum,data=myFF%>%filter(childteen=='T' & ancestry=='European ancestry'), simplify=F), show.all=T)
export2md(smkE_C)
export2md(smkE_T)

smkH_C<-createTable(compareGroups(smkPreg_binary~.-id-ff_id-idnum,data=myFF%>%filter(childteen=='C' & ancestry=='Hispanic ancestry'), simplify=F), show.all=T)
smkH_T<-createTable(compareGroups(smkPreg_binary~.-id-ff_id-idnum,data=myFF%>%filter(childteen=='T' & ancestry=='Hispanic ancestry'), simplify=F), show.all=T)
export2md(smkH_C)
export2md(smkH_T)
```

# # # Covariates by main predictor (stratified by ancestry)

# # # Methylation summary metrics (global score; clocks) by main predictor  (stratified by ancestry; age)

# # Global methylation score analysis

# # # Overall linear regressions adjusting for ancestry categorization

# # # Ancestry stratified linear regressions

# # # Different options for methylation score analysis - islands; shores etc

# # Epigenetic clock analysis

# # # Overall linear regressions adjusting for ancestry categorization

# # # Ancestry stratified linear regressions

# # Epigenome wide association analysis

# # Cell type deconvolution

# # Comparison of significant CpGs from each ancestry specific EWAS

# B) Examine differences in methylation trajectories from age nine to fifteen of the most significant methylation sites identified in the epigenome-wide association study between those exposed vs unexposed to prenatal smoking

# # Linear-mixed effect models for longitudinal trajectory of methylation 
# C) Develop a saliva epigenetic classifier of prenatal smoke exposure using a machine learning approach and ~support vector machines~ targeted maximum likelihood estimation? and evaluate the agreement of the classifier at two time points.

# # Feature selection

# # 10-fold cross validation

# # Confusion matrix

# # Cohen's kappa

### 2021-09-07 Reports/FFCleaning.Rmd
first commit on nfs

---
title: "Fragile Families Data Cleaning"
author: "Freida Blostein"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    code_folding: show
    highlight: tango
    number_sections: yes
    theme: sandstone
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/")
knitr::opts_knit$set(message=FALSE)
knitr::opts_knit$set(warning=FALSE)
library(tidyr)
library(dplyr)
library(purrr)
library(tidyselect)
library(broom)
library(compareGroups)
library(kableExtra)
library(haven)
library(sjlabelled)
library(matchmaker)
library(EpiDISH)
library("PerformanceAnalytics")
library(DescTools)
library(ggplot2)
library(ggpubr)
library(GGally)
library(ggcorrplot)
library(ggalluvial)
library(cowplot)
datadir<-"/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/"
john_datadir<-"/nfs/turbo/bakulski1/People/johndou/Fragile_Families/ProcessedFiles/jd"
outputdir<-"/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Output"
codedir<-"/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Code"
```

# Lab meeting preface & necessary background

**Purpose**: This presentation is a practice for presenting the preliminary results of the third aim of my dissertation to a committee member and important collaborator. I am seeking to get buy-in for splitting the analysis into two papers and begining to write the first paper.

**Objectives & desired feedback**: 
My objectives for this lab meeting are:

1. Practice presenting results 
2. Test drive presenting through Rmarkdown htmls

And correspondingly, I'm seeking feedback on the logical flow of the presentation and its understandability.

**Necessary background**:

#  Objectives and  hypotheses

A) Conduct an ancestry-stratified epigenome wide association study of prenatal maternal smoking and DNA methylation from child saliva samples at age nine and age fifteen.

B) Examine differences in methylation trajectories from age nine to fifteen of the most significant methylation sites identified in the epigenome-wide association study between those exposed vs unexposed to prenatal smoking.

C) Develop a saliva epigenetic classifier of prenatal smoke exposure using a machine learning approach and support vector machines and evaluate the agreement of the classifier at two time points.

Hypotheses:
- I hypothesize that different CpG loci will be associated with maternal prenatal smoking in African ancestry groups than in European and Hispanic ancestry groups. 
- I hypothesize that the direction of associations between CpGs and prenatal maternal smoking will stay consistent between ages nine and fifteen but that the magnitude of associations will decline over time. 
- Finally, I hypothesize that classification of prenatal maternal smoking status based upon DNA methylation information from salivary samples will be accurate and consistent at ages nine and fifteen.

Additional

# Reading in data and preprocessing:

## Labeling phenotype data
```{r, phenotypedataReadinLabel, include=F}
pheno<-readRDS(paste0(datadir, "OGData/fullpheno.rds"))

#Phenotype labeling & filtering to variables of interest
#labels
#to set labels using sjlabelled, needed to fix multibyte characters 
#https://stackoverflow.com/questions/4993837/r-invalid-multibyte-string
find_offending_character <- function(x, maxStringLength=256){  
  print(x)
  for (c in 1:maxStringLength){
    offendingChar <- substr(x,c,c)
    #print(offendingChar) #uncomment if you want the indiv characters printed
    #the next character is the offending multibyte Character
  }    
}
#lapply(attributes(pheno)$var.labels, find_offending_character)
attributes(pheno)$var.labels<-gsub("\xad", "'",  attributes(pheno)$var.labels)
#lapply(attributes(pheno)$var.labels, find_offending_character)
attributes(pheno)$var.labels<-gsub("\x92", "'",  attributes(pheno)$var.labels)
#lapply(attributes(pheno)$var.labels, find_offending_character)
#define vars of interest 
idvars<-c("id", "ff_id")
demovars<-c("cm1age", "cm1bsex", "cm1ethrace", "cm1edu", "cm1hhinc", "cm1inpov", "m1h3", "m1h3a", "m1h3b", "m1i4", "m1i4a", "m1i4b", "cf1age", "cf1ethrace", "f1h3", "f1h3a", "f1h3b", "ch5agem","ch6yagem", "c6yagey", "cm5b_age", "cm5b_ageyrs", "cp6yagem", "cp6yagey", "hv5_agem", "hv6_yagem", "hv6_yagey")
smokingvars<-c("m1g4", "f1g4", "f2j5", "f2j5a", "f2j7", "f2j7a", "f3j31", "f3j32", "f4j18", "f4j19", "f5g17", "f5g18", "k5f1k", "k5f1l", "k6d40", "k6d41", "k6d41", "k6d42", "k6d43", "k6d45", "k6d46", "k6d47", "m2j5", "m2j5a", "m2j7", "m2j7a", "m4j18", "m4j19", "m5g17", "m5g18", "n5f17", "n5f18", "p3a22", "p3a23", "p3a23a", "p3a23b", "p3a24", "p4a22", "p4a23", "p5h15", "p5h15b", "p5q3cr", "p6h74", "p6h75", "p6h76", "p6h77", "p6h78")
prenataldrugusevar<-c("m1g2", "m1g3", "m1g5", "m1g6")

#filter to these variables
#FF has data in original form +sjlabelled value labels
#since these dont' show in compareGroups
#also will make FF_labeled, where categorical, binary variables
#will be replaced with their labels
#FF<-pheno %>% select(one_of(idvars), one_of(demovars), one_of(smokingvars), one_of(prenataldrugusevar))
varIDX<-which(names(pheno)%in%c(idvars, demovars, smokingvars, prenataldrugusevar))
FF<-pheno[,varIDX]
varLabels<-attributes(pheno)$var.labels[varIDX]
varLabels[59]<-"ID"
FF<-set_label(FF, c(varLabels))
#create FF_labeled
#read in csv with metadata about variables 
FFmeta<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/FFMetadata_v07.csv")
#create long data dictionary  
FFlong<-FFmeta%>% select(old_name, contains('value'), contains('label')) %>% pivot_longer(cols = contains("value"), names_to="value", values_to="old_values")
FFlong<-FFlong%>%pivot_longer(cols = contains("label"), names_to="label", values_to="new_labels")
recodes_long<-FFlong %>% filter(gsub("value", "", value)==gsub("label", "", label)& !is.na(old_values)) %>% select(old_name, old_values, new_labels) 
#store continuous, binary, categorical etc
FFmyMeta<-FFmeta %>% filter(old_name %in% c(idvars, demovars, smokingvars, prenataldrugusevar))
myContinuous<-FFmyMeta %>% filter(type=="Continuous") %>%pull(old_name);
myContinuous<-colnames(FF)[colnames(FF) %in% myContinuous]
myBinary<-FFmyMeta%>%filter(type=="Binary")%>%pull(old_name); myBinary<-colnames(FF)[colnames(FF) %in% myBinary]
myCategorical_O<-FFmyMeta%>%filter(type=="Ordered Categorical")%>%pull(old_name); myCategorical_O<-colnames(FF)[colnames(FF) %in% myCategorical_O]
myCategorical_U<-FFmyMeta%>%filter(type=="Unordered Categorical")%>%pull(old_name); myCategorical_U<-colnames(FF)[colnames(FF) %in% myCategorical_U]
#filter to only vars of interest - not continuous variables  
myRecodes<-recodes_long %>% filter(old_name %in% c(myBinary, myCategorical_O, myCategorical_U))
#apply matchmaker
FF_labeled<-match_df(FF, dictionary=myRecodes, from = 'old_values', to='new_labels', by='old_name')
#reapply correct variable class types and na_codes
na_codes<-c("-1 Refuse", "-2 Don't know", "-3 Missing", "-9 Not in wave")
FF_labeled<-FF_labeled %>% mutate_at(vars(any_of(c(myBinary, myCategorical_O, myCategorical_U))), funs(as.factor(ifelse(. %in% na_codes, "Missing", .)))) %>% 
          mutate_at(vars(any_of(myContinuous)), funs(as.numeric(na_if(na_if(na_if(na_if(., "-3"), '-9'), '-6'), '-2'))))
#read back in the variable labels 
FF_labeled<-set_label(FF_labeled, get_label(FF))
FF_labels<-FF_labeled
```

## Cleaning pdqc data 
### Figure out analytic subsets 1) any methylation data; 2) 2-visits of methylation data
```{r, pdqcReadin}
pdqc_all<-readRDS(paste0(datadir, 'OGData/', "pd_qc.rds"))
pdqc<-pdqc_all %>% filter(probe_fail_10==0 & sex==predicted_sex)
#What FF ids have methylation data?
methy_ids<-pdqc$id
#create slide variable fo pdqc 
pdqc$slide <- gsub('_.*', '', pdqc$MethID)
pdqc$array<-gsub('.*_', '', pdqc$MethID)
#ids with 9visit, 15visit or both visit
visit9ids<-pdqc%>%filter(childteen=='C')%>%pull(idnum)%>%unique()
visit15ids<-pdqc%>%filter(childteen=='T')%>%pull(idnum)%>%unique()
bothvisit_ids<-visit9ids[visit9ids %in% visit15ids]
visit9only_ids<-visit9ids[!(visit9ids %in%bothvisit_ids)]
visit15only_ids<-visit15ids[!(visit15ids %in%bothvisit_ids)]
```

```{r, createAnalyticSubsetFlags}
#create flag in FF for having methylation data
#and flag for having 2 visits of methylation data
FF_labeled<-FF_labeled %>% 
  mutate(MethylData=ifelse(id %in% methy_ids, "Analysis subset", "Not in analysis subset"),
         TwoVisitData=ifelse(id%in%bothvisit_ids,"Analysis subset - both visits", "Not in both visit analysis subset"))
table(FF_labeled$MethylData)
table(FF_labeled$TwoVisitData)
```

```{r, checkIDlengths}
if((length(visit15only_ids)+length(visit9only_ids)+length(bothvisit_ids)!=length(FF_labeled %>% filter(MethylData=="Analysis subset")%>%pull(id)))){print("WARNING ID LENGTHS DO NOT ADD UP")}
if(length(bothvisit_ids)!=length(FF_labeled %>% filter(TwoVisitData=="Analysis subset - both visits")%>%pull(id))){print("WARNING ID LENGTHS DO NOT ADD UP")}
```

### Filter technical replicates (only 1 technical replicate per person-visit + decide based on QC variables - pick the one with the lower probe fail percentage)

```{r}
#technical replicates will be duplicate id + childteen variables from pdqc
pdqc<-pdqc %>% mutate(newID=paste(idnum, childteen, sep='_'))
techrep_ids<- pdqc %>% select(newID) %>% group_by(newID)%>%filter(n()>1)
techreps<-pdqc %>% filter(newID %in% techrep_ids$newID)%>%arrange(idnum,childteen)
hiPctPF<-techreps %>% group_by(newID) %>%filter(probe_fail_pct==max(probe_fail_pct))%>%pull(MethID)
pdqc_clean<-pdqc %>% filter(!(MethID %in% hiPctPF) & childteen!='M')
myMethIDs<-pdqc_clean %>% pull(MethID)
```

## Methylation summary variables

### Cell type proportions

Cell type proportions estimated using [epidDISH](https://www.bioconductor.org/packages/release/bioc/vignettes/EpiDISH/inst/doc/EpiDISH.html#how-to-estimte-cell-type-fractions-using-dnam-data) package and the EpiFibIC reference (generic for epithelial tissues).

```{r read in betaqc & celltype proportions}
betaqc<-readRDS(file=paste0(datadir, 'OGData/', "noob_filtered.rds"))
aprioriCG<-c("cg05575921", "cg04180046", "cg05549655", "cg14179389")
globalmethy<-colMeans(betaqc, na.rm=T)
globalmethdf<-as.data.frame(cbind("MethID"=names(globalmethy), "globalmethylation"=globalmethy))
aprioriCGdf<-as.data.frame(t(betaqc[aprioriCG, ]))
aprioriCGdf$MethID = colnames(betaqc)
#cell type proportions 
celltypes<-epidish(beta.m = betaqc, ref.m = centEpiFibIC.m, method = "RPC")
estF_FF<-as.data.frame(celltypes$estF)
estF_FF$MethID = row.names(estF_FF)
```

### Clocks

```{r, readInclocks}
clocks<-read.table(file=paste0(datadir, "OGData/DNAm_clocks_ffcw.txt"), header = T)
colnames(clocks)[1]<-"MethID"
```

### Global methylation score

```{r make polymethylaton score}
#read in cpgs from PMC4833289
cpgs<-read.csv(paste0(datadir, 'PMC4833289_SuppFile2CpGs.csv'), stringsAsFactors = F, header=F, skip=2)
#make headers
header1<-scan(paste0(datadir, 'PMC4833289_SuppFile2CpGs.csv'), nlines=1, what=character(), sep=',')
header1[7:26]<-c(rep("SSnewborn", 5), rep("SSnewbornCT", 5), rep("SSolder", 5), rep("anynewborn", 5))
header2<-paste(header1, scan(paste0(datadir, 'PMC4833289_SuppFile2CpGs.csv'), nlines=1, skip=1, what=character(), sep=','), sep='_')
names(cpgs)<-header2
#four meta-EWASes: SS_newborn (sustained smoking in newborns); SS_newbornCT (sustained smoking in newborns controlling for cell type) SS_older (sustained smoking in older children); any_newborn (any smoking in newborns)
#pivot_longer to get single column of coefs and a column for the analysis type
#and group into a list by analysis type
cpgs<-cpgs %>% pivot_longer(cols = c(matches("_Coef|_SE|_P|_FDR|_Direction")), names_to = c("set", ".value"), names_pattern="(.+)_(.+)")%>%rename_with(~gsub('_', '', .x))%>%group_split(set)
cpgs<-cpgs%>%setNames(lapply(cpgs, function(x){x$set %>% unique()}))
#make scores - no transform, zscore and centermean
#first, pull script 
source(file.path(codedir, 'UsefulCode', 'makePolyEpiScores.R'))
scores=makePolyEpiScore(m=betaqc, b=cpgs)
scores_z=makePolyEpiScore(m=betaqc, b=cpgs, transformopt = 'zscore')
scores_center=makePolyEpiScore(m=betaqc, b=cpgs, transformopt = 'meancenter')
polyscores=list('notransform'=scores, 'zscore'=scores_z, 'center'=scores_center)
polyscores<-lapply(polyscores, function(x) x%>% mutate(MethID=colnames(betaqc)))
save(polyscores, file = file.path(datadir, 'CreatedData', 'polymethylationscores.RDS'))
polyscores_wide<-lapply(seq_along(polyscores), function(i) polyscores[[i]] %>% setNames(c(paste0(colnames(polyscores[[i]])[1:4], '_', names(polyscores)[i]), 'MethID')))%>%reduce(left_join, by='MethID')
```

```{r make methyl dataframe}
clocks2<-clocks %>% mutate(idnum=gsub('a', '', idnum))%>%filter(MethID %in% pdqc$MethID) # currently clocks are missing 3 individuals who were in johns pipeline but not jonahs

methyldata<-left_join(left_join(left_join(left_join(left_join(pdqc_clean %>% select(MethID, slide, array,childteen, idnum), globalmethdf), clocks2), estF_FF), aprioriCGdf), polyscores_wide)
methyldata[, c("globalmethylation", aprioriCG)]<-lapply(methyldata[, c("globalmethylation", aprioriCG)],function(x) as.numeric(as.character(x)))
#lapply(methyldata, class)
methyldata_wide <- methyldata %>% select(-MethID)%>% pivot_wider(., names_from=childteen, values_from=any_of(colnames(methyldata)[4:length(colnames(methyldata))]))
```

```{r individual who is missing global methylation data but had apriori CpG and PMS data?, eval=F}
#weirdID<-methyldata %>% filter(is.na(globalmethylation))%>%pull(MethID)
#globalmethy[weirdID] #not actually missing global methylation data
#clocks %>% filter(MethID%in%weirdID) # #idnum from  not the same as in
#pdqc_clean %>% filter(MethID==weirdID) #this is just a typo in the idnum from data entry, see format of other #idnums
#pdqc_clean %>% pull(idnum) %>% head()
#solution: when creating methyldata, first trim alphabet from clocks$idnum that contain alphabetic characters --> implemented
#also misssing clocks for 3 sex outlier samples that weren't in jonahs' pipeline: 
table(is.na(methyldata$horvath13))
weirdIDs<-methyldata %>% filter(is.na(horvath13))%>%pull(MethID)
pdqc_all %>% filter(MethID %in% weirdID) %>% xtabs(~predicted_sex_outlier, data=.)
```

## Creation of new variables
### Binary smoking variable
```{r}
FF_labeled<-FF_labeled %>%mutate(smkPreg_binary=ifelse(m1g4%in% c("1 2+pk/d","2 1<pk<2","3 <1pk/d" ), "Yes", ifelse(m1g4%in%c("-2 Don't know", "-3 Missing", "-9 Not in wave"), NA, ifelse(m1g4== "4 None", "No", ifelse(m1g4=="Missing", "Missing", "FLAG")))))
#xtabs(~m1g4+smkPreg_binary, data=FF_labeled)
```

### Ancestry variables

Ancestry is based on csv files of ancestry-specific PCS created by Erin (per Jonah meeting Oct 6 2020). If an ID was in {african/european/hispanic}, then the individual is labeled as such. 
```{r}
# read in PCs
euroPC<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/european.csv")
afrPC<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/african.csv")
hisPC<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/hispanic.csv")
allPC<-read.table("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/global_pcs.txt", col.names=c("X", "idnum", paste("PC", 1:20, sep='')))
#individuals in each ancestry specific pc group can be labeled as that ancestry for categorical ancestry variable 
FF_labeled<-FF_labeled %>% mutate(ancestry=case_when(MethylData=="Analysis subset" & id %in% hisPC$idnum~"Hispanic ancestry", MethylData=="Analysis subset" & id %in%afrPC$idnum~"African ancestry", MethylData=="Analysis subset" & id %in% euroPC$idnum~"European ancestry", MethylData=="Analysis subset" & !(id %in%c(hisPC$idnum, afrPC$idnum, euroPC$idnum)) & (id %in% allPC$idnum)~"Other ancestry", MethylData=="Analysis subset" & !(id %in%c(hisPC$idnum, afrPC$idnum, euroPC$idnum)) & !(id %in% allPC$idnum)~"Missing PC data", MethylData!="Analysis subset"~"Not in analysis subset"))
#check ancestry variable
#table(FF_labeled$ancestry)
#with(FF_labeled, xtabs(~ancestry+cm1ethrace))
MissingPCdata<-FF_labeled %>% filter(ancestry=="Missing PC data")%>%select(id)
save(MissingPCdata, file = "/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/CreatedData/MissingPC.Rdata")
#add PC data to FF_labeled 
FF_labeled<-left_join(FF_labeled, allPC %>% mutate(id=as.character(idnum)) %>% select(id, PC1, PC2)%>%rename_at(vars(-id), function(x) paste0('global_', x, sep='')))
localPC<-rbind(euroPC, afrPC, hisPC)
FF_labeled<-left_join(FF_labeled, localPC  %>% mutate(id=as.character(idnum)) %>% select(id, PC1, PC2)%>% rename_at(vars(-id), function(x) paste0('local_', x, sep='')))
```

There were `r length(MissingPCdata$id)` individuals without PC data. None of these individuals had data in the global PC file, so I assume they are actually missing genetic data (for whatever reason, possibly QC).

### Prenatal exposure variables 
Recoded frequency of prenatal alcohol drinking and drug use into Yes/no binary variables for drinking and drug use during pregnancy. 
```{r PrenatalExposure}
#table(FF_labeled$m1g2) # during preg, alchohol freq
#table(FF_labeled$m1g3) # during preg, drugs freq 
with(FF_labeled, table(m1g2, m1g3))
FF_labeled<-FF_labeled %>% mutate_at(c("m1g2", "m1g3"), .funs=funs(YesNoPreg=ifelse(.=="5 Never", "No", ifelse(. =="Missing", "Missing", "Yes"))))
with(FF_labeled, table(m1g2, m1g2_YesNoPreg))
with(FF_labeled, table(m1g3, m1g3_YesNoPreg))
#table(FF_labeled$m1g5) # in last year, alc/drugs interfered with work/relationships
#table(FF_labeled$m1g6) # ever sought help for drug/alc problems
```
## Postnatal smoke exposure 
Maternal postnatal smoking at different survey points is fairly correlated, and alluvial plots of changes over time show broad consistency, so I decided to create a single smoking variable for maternal smoking at age 1 and age 5 - if mothers were smoking at either age 1 or age 5 then "Maternal smoking at age 1 or 5", if not smoking at age 1 and age 5 then "No maternal smoking at age 1 and age 5" and then if missing data at one age and no at the other, then "Missing" (and if missing at both ages, "Missing"). Also created a similar dose variable capturing maternal packs per day that could be used instead. Decided to use just age 1 and age 5 because at these two time points the mother was asked (as opposed to PCG, also the dataset of Fragile Families metadata is missing PCG questions from age 3 about smoking) and they both preceed the actual saliva collection for both saliva collection age points. Then use a smoking dose variable from the mother or PCG interview at wave of saliva collection as an additional second hand smoke exposure variable.

```{r PostnatalSmoke}
#Baseline (1) -> Year 1/Age 1 (2) ->Year 3/Age3 (3) -> Year 5 (4) ->Year 9 (5) -> year 15 (6)
#Do you smoke in past month (how many packs per day)
#Mom            ->m2j5 (m2j5a)    ->                  -> m4j18 (m4j19) /p4 -> m5g17 (m5g18) / n5f17 (n5f18) 
#PCG                              p3a23a(p3a23b)                              n5f17 (n5f18)-> p6h76 (p6h77)
#seem to be missing p3 variables (PCG survey year 3) 
#number of people in household who smoke 
#p3a23->p5h15b->p6h78

#hours child around smoke 
#p3a21 -> p4a24->p5h15c

#with(FF_labeled, table(m2j5, m4j18))
#with(FF_labeled, table(m4j18, m5g17))
#with(FF_labeled, table(n5f17, m5g17))
#with(FF_labeled, table(p6h76, m5g17))

plot_grid(FF_labeled %>% select(m2j5, m4j18, m5g17, n5f17, p6h74) %>% mutate_all(funs(as.numeric(.))) %>%ggcorr(label=T)+ggtitle("Correlation between Yes/No smoking (mother) \n at years 1, 5, & 9 & PCG at years 9 & 15")
,FF_labeled %>% select(m2j5a, m4j19, m5g18, n5f18, p6h77) %>% mutate_all(funs(as.numeric(.))) %>%ggcorr(label=T)+ggtitle("Correlation between packs per day smoking (mother) \n at years 1, 5, 9 & PCG at years 9 and 15"), nrow=2)

yesnosmkalluv<-to_lodes_form(as.data.frame(FF_labeled %>% select(m2j5, m4j18, m5g17, p6h74) %>% mutate(m5g17=stringr::str_to_title(m5g17))%>% group_by(m2j5, m4j18, m5g17, p6h74) %>% count()), axes=1:4, id='Cohort')

#ggplot(yesnosmkalluv, aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow(stat="alluvium", lode.guidance="backfront", color='darkgrey')+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9"))

smkdose<-to_lodes_form(as.data.frame(FF_labeled %>% select(m2j5a, m4j19, m5g18, p6h77)%>% mutate_at(vars(c("m2j5a", "m4j19", "m5g18")), funs(factor(substring(., 1, 1), labels=levels(FF_labeled$m2j5a)))) %>% group_by(m2j5a, m4j19, m5g18, p6h77) %>% count()), axes=1:4, id='Cohort')
#ggplot(smkdose, aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow()+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9"))

plot_grid(ggplot(yesnosmkalluv, aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow()+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9", "Age 15"))
+ggtitle("Alluvial plot of Yes/No smoking variables from age 1 to age 9 (mother) and age 15 (PCG)"), ggplot(smkdose %>% filter(!(stratum %in% c("-6 Skip", "Missing"))), aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow()+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9", "Age 15"))+ggtitle("Alluvial plot of smoking per day among those smoking from age 1 to age 9 (mother) and 15 (PCG)"), nrow=2)

#ggplot(FF_labeled, aes(x=p5h15b, y=p6h78))+geom_jitter()+stat_cor()+scale_x_continuous('# smoking in house age 9')+scale_y_continuous('# smoking in house age 15')+ggtitle("# individuals smoking in house correlation age 9 and age 15")

#Make any maternal postnatal smoking variable & dose categorization for maternal smoking at ages 1 & 5 
FF_labeled<-FF_labeled %>% mutate(PostnatalMaternalSmokingAny = case_when(
                                m2j5 =="1 Yes" ~ "Maternal smoking at age 1 or age 5",
                                m4j18=="1 Yes" ~ "Maternal smoking at age 1 or age 5",
                                m2j5 == "Missing" & m4j18=="Missing" ~ "Missing",
                                m2j5 == "Missing" & m4j18=="2 No" ~ "Missing",
                                m2j5 == "2 No" & m4j18=="Missing" ~ "Missing",
                                m2j5 == "2 No" & m4j18=="2 No" ~ "No maternal smoking at age 1 and age 5")) %>%
  mutate(PostnatalMaternalSmokingDose=case_when(
                                m2j5a %in% c("2 1pk/d", "3 1.5pk/d", "4 2pk/d", "5 >2pk/d") &
                                m4j19 %in% c("2 About pack/day", "3 Pack and half/day", "4 2 packs/day", "5   
                                          More 2 packs/day") ~ "Mom - pack/d or greater at both age 1 & 5", 
                                m2j5a %in% c("2 1pk/d", "3 1.5pk/d", "4 2pk/d", "5 >2pk/d") &
                                !(m4j19 %in% c("2 About pack/day", "3 Pack and half/day", "4 2 packs/day", "5 
                                          More 2 packs/day")) ~ "Mom - pack/d or greater at either age 1 or 5",
                                !(m2j5a %in% c("2 1pk/d", "3 1.5pk/d", "4 2pk/d", "5 >2pk/d")) &
                                m4j19 %in% c("2 About pack/day", "3 Pack and half/day", "4 2 packs/day", "5 
                                          More 2 packs/day") ~ "Mom - pack/d or greater at either age 1 or 5",
                                m2j5a %in% c("1 <1/2 pk/d") | m4j19 %in% c("1 Less half pack/day")~"Mom - less than half pack/d at either age 1 or 5", 
                                m2j5a %in% c("-6 Skip") & m4j19 %in% c("-6 Skip")~
                                  "No maternal smoking at age 1 or 5", 
                                m2j5a %in% c("Missing") & m4j19 %in% c("-6 Skip", "Missing")~"Missing", 
                                m2j5a %in% c("Missing", "-6 Skip") & m4j19 %in% c("Missing")~"Missing"))

with(FF_labeled, table(PostnatalMaternalSmokingAny, m2j5, m4j18))
#with(FF_labeled, table(m2j5a, m4j19, PostnatalMaternalSmokingDose))
#with(FF_labeled, table(PostnatalMaternalSmokingAny, PostnatalMaternalSmokingDose))
```

Maybe here best to have 1 variable capturing any maternal postnatal smoking at age 1 & 5 and then use in each crossectional model the dose of maternal postnatal smoking at current visit (and similarly for linear mixed effect model)?

```{r makeMyFF}
varLabels2<-c(paste(colnames(FF_labeled)[1:66], varLabels), "Has any methylation data?", "Has two visits of methylation data", "Any maternal smoking during pregnancy", "Ancestry from PCA of child's genetic data", "PC 1 (global)", "PC 2 (global", "PC 1 (within ancestry category)", "PC 2 (within ancestry category",
"Any alcohol during pregnancy", "Any drugs during pregnancy", "Any postnatal smoking (age 1 & 5)", "Postnatal smoking dose at age 1 & 5")
FF_labeled<-set_label(FF_labeled, varLabels2)

myFF<-FF_labeled %>% filter(MethylData=="Analysis subset")%>%mutate(idnum=id)
myFF<-left_join(myFF, methyldata)
varLabels3<-c(varLabels2, "ID", "Methylation ID", "Slide", "Array (RC)", "Visit", "Global methylation", "horvath13 methylation clock", "horvath18 methylation clock", "pediatric methylation clock", "levine methylation clock", "Epithelial cell proportion", "Fibroid cell proportion", "Immune cell proportion", "cg05575291 methylation", "cg04180046 methylation",  "cg05549655 methylation", "cg14179389 methylation", "PES: Any smoking, newborn cordblood", "PES: Sustained smoking, newborn cordblood", "PES: Sustained smoking, newborn cordblood (+CT control)", "PES: Sustained smoking, older children peripheral blood", "PES: Any smoking, newborn cordblood - zscore", "PES: Sustained smoking, newborn cordblood - zscore", "PES: Sustained smoking, newborn cordblood (+CT control) - zscore", "PES: Sustained smoking, older children peripheral blood - zscore", "PES: Any smoking, newborn cordblood - mean center", "PES: Sustained smoking, newborn cordblood - mean center", "PES: Sustained smoking, newborn cordblood (+CT control) - mean center", "PES: Sustained smoking, older children peripheral blood - mean center")
myFF<-set_label(myFF, varLabels3)
```
### Child smoking
Exclude children who reported smoking at age 9 and age 15. Interestingly, the children who report smoking at age 9 do not report smoking at age 15, and the parental perception of child smoking at age 9 does not match the children who say they have ever smoked a cigarrette. For now using the 'ever smoked' variable from age 15 for the exclusion, but there is also a 'smoked in past month' that we could use instead (lose fewer children).

```{r, ChildSmoke}
with(myFF, table(k5f1l,childteen)) # 6 kids smoking at age 9 
with(myFF, table(p5q3cr, k5f1l)) # not the same kids whose parents say somewhat or very true that the child is using tobacco
with(myFF, table(k6d40, childteen)) # 41 kids say ever smoked a cigarette at age 15
#with(myFF, summary(k6d41))
with(myFF, table(k6d42, childteen)) # of these only 13 say they've smoked in the past month 
#with(myFF, summary(k6d43))
with(myFF, table(k5f1l, k6d40, childteen))# not the same kids at age 9 and 15
```

### Maternal or PCG smoking dose at time of sample

```{r visitMaternalSmoke}
myFF<-myFF %>% mutate(SmkAtVisitPastmonth=case_when(childteen=='C' & m5g17=='1 yes' & m5g18 %in% c("2 about a pack", "3 a pack and a half", "4 about 2 packs", 
                              "5 more than two  packs") ~ "Pack or more a day", 
                              childteen=='C' & m5g17=='1 yes' & m5g18=="1 less than half a pack a day" ~ "Less than pack a day", 
                              childteen=='C' & n5f17=="1 yes" & n5f18 =='2 about a pack' ~ "Pack or more a day",
                              childteen=='C' & n5f17=="1 yes" & n5f18 =="1 less than half a pack day" ~ "Less than pack a day",
                              childteen=='C' & m5g17=="1 yes" & m5g18 %in% c("-6 Skip", "Missing")~"Missing",
                              childteen == 'C' & n5f17%in%c("-7 N/A", "2 no") & m5g17=="2 no" ~ "No smoking", 
                              childteen == 'C' & n5f17%in%c("2 no") & m5g17%in%c("2 no", "Missing") ~ "No smoking", 
                              childteen == 'C' & n5f17 %in% c("Missing", "-7 N/A") & m5g17 =="Missing" ~ "Missing", 
                              childteen=='T' & p6h76 %in% c("-6 Skip", "0 None") ~ "No smoking", 
                              childteen=='T' & p6h77%in% c("1 5 or fewer cigarettes per day", "2 About half a pack a day, 10 cigarettes")~"Less than pack a day", 
                              childteen=='T' & p6h77 %in% c("3 About a pack a day, 20 cigarettes", "4 More than 1 pack a day")~"Pack or more a day",
                              childteen=='T' & p6h77 == "Missing"~"Missing"))%>%
              mutate(SmkAtVisitYesNo=case_when(SmkAtVisitPastmonth=="No smoking"~"No", 
                                               SmkAtVisitPastmonth %in% c("Less than pack a day", "Pack or more a day") ~ "Yes", 
                                               childteen=='T'& SmkAtVisitPastmonth=="Missing" & !(p6h76 %in% c("-6 Skip", "0 None", "Missing"))~"Yes",
                                               childteen=='C' & SmkAtVisitPastmonth=="Missing" & m5g17=='1 yes'~"Yes", 
                                               childteen=='C' & SmkAtVisitPastmonth=="Missing" & m5g17!='1 yes'~"Missing", 
                                               childteen=='T'& SmkAtVisitPastmonth=="Missing" & p6h76 %in% c("-6 Skip", "0 None", "Missing")~"Missing"))

myFF %>% filter(childteen=='C') %>% xtabs(~SmkAtVisitPastmonth+SmkAtVisitYesNo, data=.)
myFF %>% filter(childteen=='T') %>% xtabs(~SmkAtVisitPastmonth+SmkAtVisitYesNo, data=.)
myFF$SmkAtVisitPastmonth<-factor(myFF$SmkAtVisitPastmonth)
```
### Child's age at time of interview/visit
```{r}
myFF %>% filter(is.na(cp6yagem) &is.na(hv6_yagem) & childteen=='T')%>%dim() #3
myFF %>% filter(is.na(cm5b_age) &is.na(hv5_agem) &childteen=='C')%>%dim() #1 
myFF %>% filter(!is.na(cm5b_age) & is.na(hv5_agem) & childteen=='C')%>%dim()#1
myFF %>% filter(!is.na(cp6yagem) &is.na(hv6_yagem))%>%dim() #990

myFF <- myFF %>% 
  mutate(ChildAgeAtIn=case_when(childteen=='C'~ cm5b_age, childteen=='T'~cp6yagem))%>%
  mutate(ChildAgeAtVisit=case_when(childteen=='C'~hv5_agem, childteen=='T'~hv6_yagem))%>% 
  mutate(ChildAgeComposite=case_when(childteen=='C' & !is.na(hv5_agem)~hv5_agem, 
         childteen=='C' & is.na(hv5_agem)~cm5b_age, 
         childteen=='T' & !is.na(hv6_yagem)~hv6_yagem, 
         childteen=='T' & is.na(hv6_yagem)~cp6yagem))

myFF %>% filter(childteen=='C')%>%group_by(ChildAgeAtVisit, ChildAgeComposite)%>%summarise(n=n())%>%filter(ChildAgeAtVisit!=ChildAgeComposite)

myFF %>% filter(childteen=='T')%>%group_by(ChildAgeAtVisit, ChildAgeComposite)%>%summarise(n=n())%>%filter(ChildAgeAtVisit!=ChildAgeComposite)

ggplot(myFF%>%filter(childteen=='C'), aes(cm5b_age, hv5_agem))+geom_point()+stat_cor()+ggtitle("Correlation between childs age at mom interview (9 yr) and childs age at home visit (9 yr)")
age5resid<-myFF %>% filter(childteen=='C')%>%mutate(age5_resid=cm5b_age-hv5_agem)%>%pull(age5_resid)
hist(age5resid)

age6resid<-myFF %>% filter(childteen=='T')%>%mutate(age6_resid=cp6yagem-hv6_yagem)%>%pull(age6_resid)
ggplot(myFF%>%filter(childteen=='T'), aes(cp6yagem,hv6_yagem))+geom_point()+stat_cor()+ggtitle("Correlation between childs age at mom interview (15 yr) and childs age at home visit (15 yr)")
hist(age6resid)

table(FF_labeled$cp6yagem-FF_labeled$hv6_yagem)
table(FF_labeled$cm5b_age-FF_labeled$hv5_agem)
table(is.na(myFF$ChildAgeComposite)) # 4

varLabels4<-c(varLabels3, "Mother/PCG (9/15 yr) smoking in month prior to interview (pk/day)", "Mother/PCG (9/15 yr) smoking in month prior to interview (yes/no)", "Child's age at maternal/PCG (9/15 yr) interview (months)", "Child's age at home vist (9/15 years) (months)", "Constructed child's age at home visit (in months) (9/15 years)")
myFF<-set_label(myFF, varLabels4)
```
I also found two variables for child's age at home visit, but the variable for child's age at home visit for 15 years has a large amount of missingness in it (missing n=`r myFF %>% filter(is.na(hv6_yagem) & childteen=='T')%>%nrow()`). The variable for child's age at home visit for 9 year visit is better (missing n=`r myFF %>% filter(is.na(hv5_agem) & childteen=='T')%>%nrow()`). The interview age variables correlate well with the home visit age variables. For the 9 year interview/home visit ages, there are 5 individuals with >12 month difference between the ages, for the 15 year visit, there are 0 (and actually, there is 0 difference for any of the individuals with nonmissing home visit ages). So one option would be to use the home visit ages for the 9 year visit and for individuals with home visit age data available for the 15 year visit, supplementing the 15 year visit age variable with  the 15 year PCG interview age visit for those missing the home visit age. Alternatively we could use baby birthdate and saliva colleciton date to create an age variable. I have the baby DOB variable, but I don't have/don't know where a saliva sample collection date would be.

# Analysis

## Randomization of prenatal maternal smoking across slide numbers

```{r slide number vs prenatal maternal smoking}
slide_count<-myFF %>% mutate(slide=as.factor(slide))%>% group_by(smkPreg_binary, slide, .drop=F)%>%count()%>%pivot_wider(id_cols = slide, names_from=smkPreg_binary, values_from=n)%>%arrange(Yes)%>%mutate(Proportion=Yes/(No+Yes))
hist_data<-slide_count %>% ggplot(aes(x=Proportion))+geom_histogram(binwidth = 0.1, fill='blue', alpha=0.5, breaks=seq(0, 1, 0.1))+ggtitle('Histogram proprotion smk (from data)')
hist_data_c<-slide_count %>% ggplot(aes(x=Yes))+geom_histogram( fill='blue', alpha=0.5, breaks=c(0:12))+scale_x_continuous(breaks=c(0:12))+ggtitle('Histogram # Yes smk (from data)')

kable(slide_count)%>%kable_styling()%>%scroll_box(width='100%', height='400px')

#simulation
set.seed(seed=20210216)
n<-nrow(myFF)
yes<-table(myFF$smkPreg_binary)[2]; no<-table(myFF$smkPreg_binary)[1]
values<-data.frame('smk'=sample(c('No', 'Yes'), n, replace=T, prob=c(no/(no+yes), yes/(no+yes))))
values_slides<-values %>% 
    group_by((row_number()-1) %/% (n()/152)) %>%
    nest()%>%unnest(cols=c(data))
colnames(values_slides)<-c('slide', 'smk')
slide_sim<-values_slides%>%mutate(slide=as.factor(slide)) %>% group_by(smk, slide, .drop=F)%>%count()%>%pivot_wider(id_cols = slide, names_from=smk, values_from=n)%>%arrange(Yes)%>%mutate(Proportion=Yes/(No+Yes))
hist_sim<-slide_sim%>% ggplot(aes(x=Proportion))+geom_histogram(binwidth = 0.1, fill='red', alpha=0.5, breaks=seq(0, 1, 0.1))+ggtitle('Histogram proprotion smk (from simulation)')
hist_sim_c<-slide_sim%>% ggplot(aes(x=Yes))+geom_histogram(fill='red', alpha=0.5, breaks=c(0:12))+scale_x_continuous(breaks=c(0:12))+ggtitle('Histogram # Yes smk (from simulation)')

hist_overlay<-slide_count %>% ggplot(aes(x=Proportion))+geom_histogram(fill='blue', binwidth = 0.1, alpha=0.5, breaks=seq(0, 1, 0.1))+geom_histogram(aes(x=slide_sim$Proportion), breaks=seq(0, 1, 0.1), fill='red', binwidth = 0.1, alpha=0.5)+xlim(0,1)
hist_overlay_c<-slide_count %>% ggplot(aes(x=Yes))+geom_histogram(fill='blue', alpha=0.5, breaks=c(0:12))+geom_histogram(aes(x=slide_sim$Yes), fill='red',  alpha=0.5, breaks=c(0:12))+scale_x_continuous(breaks=c(0:12))

plot_grid(plot_grid(hist_data, hist_sim, nrow=2), hist_overlay, nrow=1)
plot_grid(plot_grid(hist_data_c, hist_sim_c, nrow=2), hist_overlay_c, nrow=1)

```

## Comparison of analytic sample - methylation data to Fragile Families sample

```{r, warning=F, message=F}
FF_labeled<-set_label(FF_labeled, varLabels)
methyl<-createTable(compareGroups(MethylData~smkPreg_binary+cm1inpov+cm1ethrace+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+k5f1l+k6d40+f1g4, data=FF_labeled), show.all=T)
export2md(methyl, caption="Exclusion table of all Fragile Families vs those with any methylation data for main predictors and key demographic variables from first wave")
```

## Comparison of variables by ancestry 
```{r, warning=F, message=F, include=F}
myFF<-set_label(myFF, varLabels4)
ancestrytable<-strataTable(createTable(compareGroups(ancestry~globalmethylation+cg05575921+pediatric+levine+Epi+Fib+IC+smkPreg_binary+cm1inpov+cm1ethrace+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+k6d40+k6d42+f1g4, data=myFF), show.all=F), 'childteen')
export2md(ancestrytable)
```

## Flow chart of sample exclusions

```{r MakeFlowChart}
methylCohort<-pdqc_all %>% filter(childteen!='M')

basemodelvar<-c("smkPreg_binary", "ancestry", "cm1bsex", "cm1inpov", "Epi", "IC", "ChildAgeComposite")
basemodeldata<-myFF %>% filter_at(all_of(basemodelvar), all_vars(!(. %in%c("Missing", "Missing PC data"))& !is.na(.)))
child_base<-basemodeldata%>%filter(childteen=='C')
teen_base<-basemodeldata%>%filter(childteen=='T')

prenatalexposurevar<-c(basemodelvar, "m1g2_YesNoPreg", "m1g3_YesNoPreg")
prenatalexdata<-basemodeldata %>%filter_at(all_of(prenatalexposurevar), all_vars(!(. %in% c("Missing"))))
child_prenatal<-prenatalexdata%>%filter(childteen=='C')
teen_prenatal<-prenatalexdata%>%filter(childteen=='T')

secondhandsmokevars<-c(prenatalexposurevar, "PostnatalMaternalSmokingAny", "SmkAtVisitPastmonth")
secondhandsmkdata<-prenatalexdata%>%filter_at(all_of(secondhandsmokevars), all_vars(!(. %in% c("Missing"))))
child_secondhand<-secondhandsmkdata %>% filter(childteen=='C')
teen_secondhand<-secondhandsmkdata%>%filter(childteen=='T')
twovisit_secondhand<-secondhandsmkdata %>% filter(idnum %in% child_secondhand$idnum & idnum %in% teen_secondhand$idnum)

child_smoke<-secondhandsmkdata %>% filter(k5f1l!= "2 no" & childteen=='C')
child_nosmoke<-secondhandsmkdata %>% filter(k5f1l=="2 no" & childteen=='C')
teen_smoke<-secondhandsmkdata %>% filter(k6d40!="2 No" & childteen=='T')
teen_nosmoke<-secondhandsmkdata %>% filter(k6d40=="2 No" & childteen=='T')  
twovisit_nosmoke<-secondhandsmkdata %>% filter(idnum %in% child_nosmoke$idnum & idnum %in% teen_nosmoke$idnum)

fathersmkdata<-secondhandsmkdata %>% filter(f1g4!="Missing")

#Make Labels
n_fullFF<-paste0("Fragile Families cohort \n", "n=", FF_labeled %>% nrow())
n_methylQC<-paste0("Cohort with 450K methylation data \n", "n=", methylCohort$idnum%>%unique()%>%length(), "individuals with m=", methylCohort$MethID%>%length(), "samples")

n_QCloss<-paste0("Methylation data QC: 12 low-intensity samples dropped, \n", pdqc_all%>%filter(probe_fail_10==T)%>%nrow()-12, " samples with >10% of probes failing dropped, \n", pdqc_all%>%filter(sex!=predicted_sex & probe_fail_10==F & childteen!='M')%>%nrow(), " discordant sex samples dropped, \n & ", nrow(pdqc)-nrow(pdqc_clean), " duplicate samples dropped")
n_methylFF<-paste0("Cohort with quality controlled 450K data \n","n=", myFF %>% select(id) %>% unique() %>% nrow(), " individuals with m=", myFF%>%nrow(), " samples")

n_basemodel<-paste0("Base models: \n", "n=", basemodeldata %>% select(id) %>% unique() %>% nrow(), " individuals with m=", basemodeldata %>%nrow(), " samples")
n_baseexclusions<-paste0("n=",  myFF %>% filter_at(all_of(basemodelvar), any_vars((. %in%c("Missing", "Missing PC data"))))%>% select(id) %>% unique() %>% nrow(), " individuals & m=", myFF %>% filter_at(all_of(basemodelvar), any_vars((. %in%c("Missing", "Missing PC data")))) %>% nrow(), " samples missing principal components of ancestry; \n n=", myFF %>% filter_at(all_of(basemodelvar), any_vars((is.na(.))))%>%select(id)%>%unique()%>%nrow(), " individuals & m=", myFF %>% filter_at(all_of(basemodelvar), any_vars((is.na(.))))%>%nrow(), " samples missing child age at maternal/PCG interview")

n_prenatalexposures<-paste0("Prenatal exposures models: \n","n=", prenatalexdata %>% select(id) %>% unique() %>% nrow(), " individuals with m=", prenatalexdata %>%nrow(), " samples")
n_prenatalexclusions<-paste0("n=", basemodeldata %>% filter_at(all_of(prenatalexposurevar), any_vars(. %in% c("Missing"))) %>% select(id) %>% unique() %>% nrow(), " individuals & m=", basemodeldata %>% filter_at(all_of(prenatalexposurevar), any_vars(. %in% c("Missing"))) %>% nrow(), " samples missing \n data on maternal prenatal alcohol and drug use")

n_secondhandSmk<-paste0("Secondhand smoke exposure models: \n","n=", secondhandsmkdata %>% select(id) %>% unique() %>% nrow(), " individuals with m=", secondhandsmkdata %>%nrow(), " samples")
n_secondhandexclusions<-paste0("n=", prenatalexdata$id[!(prenatalexdata$id %in% secondhandsmkdata$id)] %>% unique()%>% length(), " individuals & m=", prenatalexdata %>% filter_at(all_of(secondhandsmokevars), any_vars(. %in% c("Missing"))) %>% nrow(), " samples missing data \n on maternal or primary care giver postnatal smoking")

n_Child<-paste0('n=', child_secondhand  %>% nrow(), " individuals with a 9 year visit")
n_ChildSmkexcluded<-paste0("Sensitivity analysis exclude children who smoke: \n",'n=', child_nosmoke %>% nrow(), " individuals with a 9 year visit")
n_ChildSmkexclusions<-paste0('n=', child_smoke%>%nrow(), " individual \n smoking at 9 year visit \n or missing focal child smoking data")
n_Teen<-paste0('n=', teen_secondhand  %>% nrow(), " individuals with a 15 year visit")
n_TeenSmkexcluded<-paste0("Sensitivity analysis exclude children who smoke: \n",'n=', teen_nosmoke %>% nrow(), " individuals with a 15 year visit")
n_TeenSmkexclusions<-paste0('n=', teen_smoke%>%nrow(), " individuals \n smoking at 15 year visit \n or missing focal child smoking data")

n_sensFatherSmoking<-paste0("Sensitivity analysis: compare effects of maternal and paternal prenatal smoking \n", "n=", fathersmkdata %>% select(id) %>% unique() %>% nrow(), " individuals with m=", fathersmkdata %>%nrow(), " samples", "\n with prenatal paternal smoking data")
n_sensFatherSmokingEx<- paste0("n=", secondhandsmkdata$id[!(secondhandsmkdata$id %in% fathersmkdata$id)] %>% unique()%>% length(), " individuals & m=", secondhandsmkdata %>% filter(f1g4 %in% c("Missing")) %>% nrow(), " samples missing data on paternal prenatal smoking")

n_twovisits<-paste0("n=", twovisit_secondhand%>%pull(idnum) %>%unique() %>%length(), " individuals with both a 9 and 15 year visit")
n_twovisitsSmkExcluded<-paste0("Sensitivity analysis exclude children who smoke: \n", "n=", twovisit_nosmoke%>%pull(idnum) %>%unique() %>%length(), " individuals with both a 9 and 15 year visit")
n_twovisitSmkExclusions<-paste0("n=", (twovisit_secondhand%>%pull(idnum) %>%unique() %>%length())-(twovisit_nosmoke%>%pull(idnum) %>%unique() %>%length()), " individuals smoking at \n either 9 and 15 year visit or \n missing focal child smoking data")

```

```{r drawFlowchart}
graph1<-DiagrammeR::grViz("
   digraph graph2 {
   compound=true;
   graph [layout = dot]

   node [shape = rectangle, width=4]
   a [label = '@@1']
   b [label = '@@2']
   c [label = '@@3']
   d [label = '@@4']
   e [label = '@@5']
   f [label = '@@6']
   g [label = '@@7']
   h [label = '@@8']
   i [label = '@@9', group=g1]
   j [label = '@@10']
   k [label = '@@16']
   l [label = '@@17']
   m [label = '@@19']

   #fake nodes for connecting to edges
   node [shape=point, width=0.01, height=0.01]
   a1
   b1
   c1
   d1
#   e1
  f1 [group=g1]
  g1
  k1

  #exclusion nodes
  node [shape = rectangle, width=4]
  aa [label = '@@20']
  bb [label = '@@11']
  cc [label = '@@12']
  dd [label = '@@13']
  ff [label = '@@14', width=3]
  gg [label = '@@15', width=3]
  kk [label = '@@18', width=3]

  #draw connections
  a->m
  m->a1 [arrowhead=none]
  a1->b
  b->b1 [arrowhead=none]
  b1->c
  c->c1 [arrowhead=none]
  c1->d
  d->d1 [arrowhead=none]
  d1->e
  e->f; e->g; e->k
{rank=same; h->e [dir=back, minlen=2, style=dashed]}
  f->f1 [arrowhead=none, style=dashed]
  g->g1 [arrowhead=none, style=dashed]
  k->k1 [arrowhead=none, style=dashed]
  f1->i [style=dashed]
  g1->j [style=dashed]
  k1->l [style=dashed]

{rank=same; a1->aa [minlen=2]}   
{rank=same; b1->bb [minlen=2]} 
{rank=same; c1->cc [minlen=2]}
{rank=same; d1->dd [minlen=2]}
{rank=same; f1->ff [minlen=2]}
{rank=same; g1->gg [minlen=2]}
{rank=same; k1->kk [minlen=2]}

#invisible constraints to force ordering
ff->i [style=invis]
   }

   [1]: n_fullFF
   [2]: n_methylFF
   [3]: n_basemodel
   [4]: n_prenatalexposures
   [5]: n_secondhandSmk
   [6]: n_Child
   [7]: n_Teen
   [8]: n_sensFatherSmoking
   [9]: n_ChildSmkexcluded
   [10]: n_TeenSmkexcluded
   [11]: n_baseexclusions
   [12]: n_prenatalexclusions
   [13]: n_secondhandexclusions
   [14]: n_ChildSmkexclusions
   [15]: n_TeenSmkexclusions
   [16]: n_twovisits
   [17]: n_twovisitsSmkExcluded
   [18]: n_twovisitSmkExclusions
   [19]: n_methylQC
   [20]: n_QCloss
")
```

```{r drawGraph}
graph1
```

## Cell type proportions over time

```{r, eval=F, include=F}
cellbytime<-myFF %>% pivot_longer(cols=c(Fib, Epi, IC), names_to="Celltype", values_to = "Proportion")
ggplot(cellbytime, aes(x=childteen, y=Proportion, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+facet_wrap(~Celltype)+stat_compare_means(label="..p.signif..")+ggtitle("Cell type proportions by visit")

ggplot(myFF, aes(x=childteen, y=globalmethylation, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+ggtitle("Global methylation by visit")

ggplot(myFF, aes(x=childteen, y=pediatric, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+ggtitle("Pediatric methylation clock by visit")
ggplot(myFF, aes(x=pediatric, y=IC))+geom_point()+stat_cor()+facet_wrap(~childteen)+ggtitle("Correlation between pediatric clock and IC proportion")
#ggplot(myFF, aes(x=pediatric, y=Epi))+geom_point()+stat_cor()+facet_wrap(~childteen)
#ggplot(myFF, aes(x=childteen, y=levine, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")
ggplot(myFF, aes(x=childteen, y=cg05575921, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+ggtitle("Methylation at cg05575921 by visit")

```
Cell type proportions do differ between 9 and 15 year visits and although global methylation does not look different between these visits, methylation clocks do (a good check), although, interestingly, cell type proportion does correlate with methylation age.

## Distributions & checks of methylation variables

### Correlation between all methylation data
```{r}
methylation_data<-myFF %>% select(colnames(methyldata))%>%select(-MethID, -childteen, -idnum, -slide, -array)
chart.Correlation(methylation_data, histogram=T)
```

### Correlation between polymethylation scores
```{r}
chart.Correlation(methylation_data %>%select(SSolder_notransform, SSnewbornCT_notransform, SSnewborn_notransform, anynewborn_notransform))
```

Thought this was a little weird, investigated the actual coefficients for each score:

```{r}
coefs_methyl<-cbind('SSolderCoefs'=cpgs$SSolder$Coef, 'SSnewbornCTCoefs'=cpgs$SSnewbornCT$Coef, 'SSnewbornCoefs'=cpgs$SSnewborn$Coef, 'anynewbornCoefs'=cpgs$anynewborn$Coef)
chart.Correlation(coefs_methyl)
```

```{r correlaton b/t polymethylation scores over time, eval=F}
PMS<-methyldata %>%select(idnum, childteen, colnames(polyscores_wide), -MethID)%>%pivot_longer(cols=any_of(colnames(polyscores_wide)), names_to="PMS_type", values_to="PolymethylationScore")%>%pivot_wider(names_from = childteen, values_from=PolymethylationScore)
ggplot(PMS, aes(x=C, y=T))+geom_point()+facet_grid(~PMS_type)+stat_cor()+geom_smooth(method=lm)
```
```{r Polymethylation scores over time}
gg_newborn<-ggplot(methyldata, aes(x=childteen, y=SSnewbornCT_center, group=idnum))+geom_point()+geom_line(alpha=0.2)+geom_smooth(aes(group=1), method='lm')+ggtitle("SS newborn (+CT)")
gg_older<-ggplot(methyldata, aes(x=childteen, y=SSolder_center, group=idnum))+geom_point()+geom_line(alpha=0.2)+geom_smooth(aes(group=1), method='lm')+ggtitle("SS older")
plot_grid(gg_newborn, gg_older, nrow = 2)
```
```{r Polymethylation scores over time by prenatal smoke}
gg_newborn<-ggplot(myFF, aes(x=childteen, y=SSnewbornCT_center, group=idnum, color=smkPreg_binary))+geom_point()+geom_line(alpha=0.2)+geom_smooth(aes(group=1),color='black', method='lm')+facet_wrap(~smkPreg_binary)
gg_older<-ggplot(myFF, aes(x=childteen, y=SSolder_center, group=idnum, color=smkPreg_binary))+geom_point()+geom_line(alpha=0.2)+geom_smooth(aes(group=1),color='black', method='lm')+facet_wrap(~smkPreg_binary)
plot_grid(gg_newborn, gg_older, nrow=2)
```

```{r boxplots of PMS}
PMS<-left_join(methyldata %>%select(idnum, childteen, colnames(polyscores_wide), -MethID)%>%pivot_longer(cols=any_of(colnames(polyscores_wide)), names_to="PMS_type", values_to="PolymethylationScore"), myFF[, c('idnum', 'smkPreg_binary')]%>%unique())
ggplot(PMS, aes(x=childteen, y=PolymethylationScore, color=smkPreg_binary))+geom_boxplot()+facet_wrap(~PMS_type)+stat_compare_means(label='p.signif', label.y=4)

```

## Distribution of main predictor and bivariate tables (stratified by ancestry)

### Covariates by main predictor (unstratified)

```{r, pickData}
tabledata<-rbind(child_nosmoke, teen_nosmoke)
tabledata<-set_label(tabledata, varLabels4)

```

```{r, warning=F, message=F}
smk_all<-strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+SSnewbornCT_notransform+SSolder_notransform+cg05575921+pediatric+levine+Epi+Fib+IC+cm1inpov+cm1ethrace+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+k5f1l+k6d40+f1g4, data=tabledata, simplify = F), show.all=F), 'childteen')
export2md(smk_all)
```

Note hypomethylation in exposed children at cg05575291 (AHHR) gene, which is more strongly associated at the Teen visit than the Child visit.... although not a very small p-value - check with Kelly - is this concerning?

Additionally, cell type proportion does not appear to associate with main predictor of binary prenatal smoking variable.

### Covariates by main predictor (stratified by ancestry)
These tables use the base model group from the above CONSORT diagram (n=844). 
```{r, warning=F, message=F}
smkA<-strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+SSnewbornCT_notransform+SSolder_notransform+cg05575921+pediatric+levine+Epi+Fib+IC+cm1inpov+cm1ethrace+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+k5f1l+k6d40+f1g4, data=tabledata%>%filter(ancestry=='African ancestry')%>%copy_labels(tabledata)), show.all=F), 'childteen')
export2md(smkA, caption="Smoked during pregnancy vs covariates: African ancestry only")
smkE<-strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+SSnewbornCT_notransform+SSolder_notransform+cg05575921+pediatric+levine+Epi+Fib+IC+cm1inpov+cm1ethrace+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+k5f1l+k6d40+f1g4, simplify=FALSE, data=tabledata%>%filter(ancestry=='European ancestry' & SmkAtVisitPastmonth!="Missing")%>%copy_labels(tabledata)), show.all=TRUE), 'childteen')
export2md(smkE, caption="Smoked during pregnancy vs covariates: European ancestry only")
smkH<-strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+SSnewbornCT_notransform+SSolder_notransform+cg05575921+pediatric+levine+Epi+Fib+IC+cm1inpov+cm1ethrace+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+k5f1l+k6d40+f1g4, data=tabledata%>%filter(ancestry=='Hispanic ancestry')%>%copy_labels(tabledata)), show.all=F), 'childteen')
export2md(smkH, caption="Smoked during pregnancy vs covariates: Hispanic ancestry only")

```

When stratified by ancestry, relationship between cg05575291 methylation and smoking remains in African ancestry group, pattern of hypomethylation exists in teen measures of European ancestry group but not child measures, and actually non-significant hypermethylation in Hispanic ancestry group.

## Correlation plots between all continuous variables

```{r, eval=F, include=F, eval=F}
cont_data<-myFF %>% select(globalmethylation, cg05575921, pediatric, levine, Epi, Fib, IC, myContinuous)%>%mutate_at(vars(myContinuous), funs(as.numeric(ifelse(.<0, NA, .))))
chart.Correlation(cont_data, histogram=T)
```

## Associations with global methylation 
```{r, eval=F, include=F}
#use desc tools here 
myFFbin<-myFF%>%mutate_at(vars(all_of(myBinary)), funs(recode_factor(., '-1 Refuse'='Missing', "-2 Don't know"="Missing", "-9 Not in wave"="Missing", "-3 Missing"="Missing", "-6 Skip"="Missing", "-7 N/A"="Missing", "-5 Not asked"="Missing")))
paste("globalmethylation~", paste(myBinary, collapse="+"), sep='')
z<-Desc(globalmethylation~m1g5+m1g6+m1h3a+m1i4a+f1h3a+m2j5+m2j7+f2j5+f2j7+f3j31+m4j18+f4j18+k5f1k+k5f1l+m5g17+f5g17+n5f17+p5h15+p6h74+k6d40+k6d45, data=myFFbin%>%filter(childteen=='C'), plotit = F)
x<-Desc(globalmethylation~m1g5+m1g6+m1h3a+m1i4a+f1h3a+m2j5+m2j7+f2j5+f2j7+f3j31+m4j18+f4j18+k5f1k+k5f1l+m5g17+f5g17+n5f17+p5h15+p6h74+k6d40+k6d45, data=myFFbin%>%filter(childteen=='T'), plotit = F)

plot(Desc(globalmethylation~smkPreg_binary+m2j5+n5f17, data=myFFbin %>%filter(childteen=='C'), main=c("Binary pregnancy smoke variable", "Mother smoke cig - Year 1", "PCG smoke cig - Year 9")), type='dens')

plot(Desc(globalmethylation~smkPreg_binary+m2j5+m5g17, data=myFFbin %>%filter(childteen=='T'), main=c("Binary pregnancy smoke variable (teen methyl)", "Mother smoke cig - Year 1 (teen methyl)", "Mother smoke cig - Year 9 (teen methyl)")), type='dens')

myFFcat<-myFF%>%mutate_at(vars(all_of(c(myCategorical_O, myCategorical_U))), funs(recode_factor(., '-1 Refuse'='Missing', "-2 Don't know"="Missing", "-9 Not in wave"="Missing", "-3 Missing"="Missing", "-6 Skip"="Missing", "-7 N/A"="Missing", "-5 Not asked"="Missing")))
paste("globalmethylation~", paste(c(myCategorical_O, myCategorical_U), collapse="+"), sep='')
cat_oC<-Desc(globalmethylation~m1g2+m1g3+m1g4+f1g4+m2j5a+m2j7a+f2j5a+f2j7a+f3j32+m4j19+f4j19+m5g18+f5g18+n5f18+p5q3cr+p6h76+p6h77+k6d42+k6d43+k6d46+k6d47+cm1bsex+m1h3+m1h3b+cm1ethrace+m1i4+m1i4b+f1h3+f1h3b+cf1ethrace, data=myFFcat%>%filter(childteen=='C'), plotit = F)
#n5f18 p6h76 cf1ethrace m1h3 (ethnicity) cm1bsex
cat_oT<-Desc(globalmethylation~m1g2+m1g3+m1g4+f1g4+m2j5a+m2j7a+f2j5a+f2j7a+f3j32+m4j19+f4j19+m5g18+f5g18+n5f18+p5q3cr+p6h76+p6h77+k6d42+k6d43+k6d46+k6d47+cm1bsex+m1h3+m1h3b+cm1ethrace+m1i4+m1i4b+f1h3+f1h3b+cf1ethrace, data=myFFcat%>%filter(childteen=='T'), plotit = F)
#m2j5a

plot(Desc(globalmethylation~cm1bsex, data=myFFbin %>%filter(childteen=='T'), main=c("Baby sex")), type='dens')

plot(Desc(globalmethylation~p6h74, data=myFF %>%filter(childteen=='T'), main=c("PCG ever smoked cigarretes")), type='dens')
plot(Desc(globalmethylation~p6h76, data=myFF %>%filter(childteen=='T'), main=c("How often PCG smoked cigarretes past month")))
plot(Desc(globalmethylation~p6h76, data=myFFcat %>%filter(childteen=='T'), main=c("How often PCG smoked cigarretes past month")), type='dens')

```

At child and teen visit, variables to do with Father and Mother hispanic identity and ethnicity associated with global methylation, as did baby sex.

Some maternal/primary care giver smoking variables from year 1, 9 and 15 also associated. A note on the question PCG Year 9 Smoke and related questions n5f17 and n5f18 - these variables are weird, with lots of missing values, need to check the FF question guide, but my impression is that they were only asked to children for whom the PCG was not the mom (PCG and Core mother survey were combined in wave 6 - year 15).

## Associations with cg05575921 methylation and other significant sites from Wiklund et al 2019

Base model controls for ancestry, childsex, maternal income to poverty ratio, Epi cells, IC cells, and child age in months at sample. 
```{r}
#define function for use in purr 
apriori_fun = function(response, dat, groupF, vars){
  exclude=ifelse(groupF=='C', 'cp6yagem', ifelse(groupF=='T', 'cm5b_age', NA))
  vars=vars[vars!=exclude]
  form=paste0(response, "~", paste(vars, collapse='+'))
  tidy(lm(as.formula(form), data=dat %>% filter(childteen==groupF)), conf.int=T, conf.level=0.95)
}

#read in top CpGs from Table One of Wicklund et al 2019
topCPGs<-read.csv(paste0(datadir, "TopCpGsWiklundEtAl2019.csv"))
topCPGs<-topCPGs %>% filter(CpG %in% aprioriCG)%>% mutate(estimate=as.numeric(gsub("−\\s", "-", gsub(" .*", "", β..SE.))), std.error=as.numeric(sub("\\).*", "", sub(".*\\(", "", β..SE.))), p.value=P.value, source="Wiklund et al. 2019", conf.low=(estimate-1.96*std.error), conf.high=(estimate+1.96*std.error))%>%select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source)

#purr over the aprioriCpGs
vars=aprioriCG
vars=purrr::set_names(vars)
#ages=c('C', 'T')
#ages=purrr::set_names(ages)
models_C=  vars%>% purrr::map(apriori_fun, dat=basemodeldata, groupF='C', vars=basemodelvar)
models_T=  vars%>% purrr::map(apriori_fun, dat=basemodeldata, groupF='T', vars=basemodelvar)
models=list('child'=models_C, 'teens'=models_T)
#pull out list of tidy models into a datafram 
modeldf<-left_join(bind_rows(unlist(models, recursive=F), .id="id") %>% filter(term=='smkPreg_binaryYes')%>%mutate(source=paste0("Fragile Families: ", age=gsub("\\..*", "", id)), CpG=gsub(".*\\.", "", id)), topCPGs %>%select(CpG, Chr, Position, Gene))%>%select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source)

#combine
modeldf<-rbind(modeldf, topCPGs)%>%mutate(label=paste0(Gene, ": ", CpG))

#plot
ggplot(modeldf, aes(x=estimate, xmin=conf.low, xmax=conf.high, color=source, y=label))+geom_point(position=position_dodge(width=1))+geom_errorbarh(height=0.1, position=position_dodge(width=1))+scale_x_continuous()+scale_y_discrete(name="CpGs (a priori)")+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+ggtitle("Wiklund 2019 vs Fragile Families at 4 selected a priori CpGs: \n Base model")

```

# Datasets for modeling
Variables

Base model: Binary smoking variable [X], ancestry categorization, baby sex [X], maternal income to poverty ratio [X],  child's age in month at visit of interest (note, currently using a constructed child's age at visit, where individuals having a missing child's age at home visit was supplemented with child's age at home interview) [X], cell types (Epi + IC, may need to take one out),

Prenatal exposures: Base model variables + prenatal maternal alcohol use (Yes/no) [X], prenatal maternal drug use (yes/no)

Secondhand smoke exposure models: Prenatal exposures model+ any maternal smoking at age 1 & 5 [X], maternal or PCG smoking dose at visit of interest (age 9 or 15) [X]

First hand smoking: Secondhand smoke exposure models + exclude children with missing or reported firsthand smoking at visit of interest (age 9 or 15)

Paternal smoking: Filter to individuals with data on prenatal paternal smoking and compare effect size of prenatal paternal to prenatal maternal (as per Wiklund et al. 2019, idea is that this captures additional confounding)

Ancestry specific models - by ancestry groups; replace ancestry control with control for local PCs within ancestry group

Longitduinal models - no additional filtering needed

```{r, saveData}
save(child_base, file=paste0(datadir, 'CreatedData/', 'ChildBaseModelData.Rdata'))
save(teen_base, file=paste0(datadir, 'CreatedData/', 'ChildBaseModelData.Rdata'))
save(basemodeldata, file=paste0(datadir, 'CreatedData/', 'BaseModelData.Rdata'))

save(child_prenatal, file=paste0(datadir, 'CreatedData/', 'ChildPrenatalExModelData.Rdata'))
save(teen_prenatal, file=paste0(datadir, 'CreatedData/', 'PrenatalExModelData.Rdata'))
save(prenatalexdata, file=paste0(datadir, 'CreatedData/', 'PrenatalExModelData.Rdata'))

save(child_secondhand, file=paste0(datadir, 'CreatedData/', 'ChildSecondHandSmkModelData.Rdata'))
save(teen_secondhand, file=paste0(datadir, 'CreatedData/', 'TeenSecondHandSmkModelData.Rdata'))
save(twovisit_secondhand, file=paste0(datadir, 'CreatedData/', 'TwoVisitSecondHandSmkModelData.Rdata'))
save(secondhandsmkdata, file=paste0(datadir, 'CreatedData/', 'SecondHandSmkModelData.Rdata'))

save(child_nosmoke, file=paste0(datadir, 'CreatedData/', 'ChildNoSmkModelData.Rdata'))
save(teen_nosmoke, file=paste0(datadir, 'CreatedData/', 'TeenNoSmkModelData.Rdata'))
save(twovisit_nosmoke, file=paste0(datadir, 'CreatedData/', 'TwoVisitNoSmkModelData.Rdata'))

save(fathersmkdata, file=paste0(datadir, 'CreatedData/', 'FatherSmkModelData.Rdata'))

save(list=c("basemodeldata", "prenatalexdata", "secondhandsmkdata", "child_secondhand", "teen_secondhand", "twovisit_secondhand", "twovisit_nosmoke", "child_nosmoke", "teen_nosmoke", "fathersmkdata"), file=paste0(datadir, 'CreatedData/', Sys.Date(), 'CreatedData.Rdata'))
```

### 2021-09-07 Reports/FFCleaning2.Rmd
first commit on nfs

---
title: "Fragile Families Data Cleaning"
author: "Freida Blostein"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    code_folding: show
    highlight: tango
    number_sections: yes
    theme: sandstone
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/nfs/turbo/bakulski1/People/blostein/FF_methylation")
knitr::opts_knit$set(message=FALSE)
knitr::opts_knit$set(warning=FALSE)
library(tidyr)
library(stringr)
library(dplyr)
library(purrr)
library(tidyselect)
library(broom)
library(compareGroups)
library(kableExtra)
library(haven)
library(sjlabelled)
library(matchmaker)
library(EpiDISH)
library("PerformanceAnalytics")
library(DescTools)
library(ggplot2)
library(ggpubr)
library(GGally)
library(ggcorrplot)
library(ggalluvial)
library(cowplot)
library(DiagrammeR)
library(sva)
datadir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
john_datadir<-"/nfs/turbo/bakulski1/People/johndou/Fragile_Families/ProcessedFiles/jd"
outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output/"
codedir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/"
```

# Review of important goals

## Necessary background:

* 11% of pregnant US women reported past month cigarette use in 2018.

* Prenatal maternal smoking is associated with negative health outcomes, including low birth weight, asthama and neurodevelopmental and behavioral issues in exposed children.

* Prenatal maternal smoking is hard to measure: social desirability bias, serum cotinine is short-lived in pregnant women, and in adult cohorts, possibility of recall bias.

* Finding a reliable and specific biomarker of prenatal maternal smoking is desirable.

### Previous work

Previous work has found many associations between prenatal maternal smoking and DNA methylation, a type of epigenetic change to DNA where a methyl attaches to a CpG site.

#  Objectives and  hypotheses

A) Conduct an ancestry-stratified epigenome wide association study of prenatal maternal smoking and DNA methylation from child saliva samples at age nine and age fifteen.

B) Examine differences in methylation trajectories from age nine to fifteen of the most significant methylation sites identified in the epigenome-wide association study between those exposed vs unexposed to prenatal smoking.

C) Develop a saliva epigenetic classifier of prenatal smoke exposure using a machine learning approach and support vector machines and evaluate the agreement of the classifier at two time points.

Hypotheses:
- I hypothesize that different CpG loci will be associated with maternal prenatal smoking in African ancestry groups than in European and Hispanic ancestry groups.

- I hypothesize that the direction of associations between CpGs and prenatal maternal smoking will stay consistent between ages nine and fifteen but that the magnitude of associations will decline over time.

- Finally, I hypothesize that classification of prenatal maternal smoking status based upon DNA methylation information from salivary samples will be accurate and consistent at ages nine and fifteen.

# Reading in data and preprocessing

## Labeling phenotype data
```{r, phenotypedataReadinLabel, include=F}
pheno<-readRDS(paste0(datadir, "OGData/fullpheno.rds"))

#Phenotype labeling & filtering to variables of interest
#labels
#to set labels using sjlabelled, needed to fix multibyte characters 
#https://stackoverflow.com/questions/4993837/r-invalid-multibyte-string
find_offending_character <- function(x, maxStringLength=256){  
  print(x)
  for (c in 1:maxStringLength){
    offendingChar <- substr(x,c,c)
    #print(offendingChar) #uncomment if you want the indiv characters printed
    #the next character is the offending multibyte Character
  }    
}
#lapply(attributes(pheno)$var.labels, find_offending_character)
attributes(pheno)$var.labels<-gsub("\xad", "'",  attributes(pheno)$var.labels)
#lapply(attributes(pheno)$var.labels, find_offending_character)
attributes(pheno)$var.labels<-gsub("\x92", "'",  attributes(pheno)$var.labels)
#lapply(attributes(pheno)$var.labels, find_offending_character)
#define vars of interest 
idvars<-c("id", "ff_id")
demovars<-c("cm1age", "cm1bsex", "cm1ethrace", "cm1edu", "cm1hhinc", "cm1inpov", "m1h3", "m1h3a", "m1h3b", "m1i4", "m1i4a", "m1i4b", "cf1age", "cf1ethrace", "f1h3", "f1h3a", "f1h3b", "ch5agem","ch6yagem", "c6yagey", "cm5b_age", "cm5b_ageyrs", "cp6yagem", "cp6yagey", "hv5_agem", "hv6_yagem", "hv6_yagey")
smokingvars<-c("m1g4", "f1g4", "f2j5", "f2j5a", "f2j7", "f2j7a", "f3j31", "f3j32", "f4j18", "f4j19", "f5g17", "f5g18", "k5f1k", "k5f1l", "k6d40", "k6d41", "k6d41", "k6d42", "k6d43", "k6d45", "k6d46", "k6d47", "m2j5", "m2j5a", "m2j7", "m2j7a", "m4j18", "m4j19", "m5g17", "m5g18", "n5f17", "n5f18", "p3a22", "p3a23", "p3a23a", "p3a23b", "p3a24", "p4a22", "p4a23", "p5h15", "p5h15b", "p5q3cr", "p6h74", "p6h75", "p6h76", "p6h77", "p6h78")
prenataldrugusevar<-c("m1g2", "m1g3", "m1g5", "m1g6")

#filter to these variables
#FF has data in original form +sjlabelled value labels
#since these dont' show in compareGroups
#also will make FF_labeled, where categorical, binary variables
#will be replaced with their labels
#FF<-pheno %>% select(one_of(idvars), one_of(demovars), one_of(smokingvars), one_of(prenataldrugusevar))
varIDX<-which(names(pheno)%in%c(idvars, demovars, smokingvars, prenataldrugusevar))
FF<-pheno[,varIDX]
varLabels<-attributes(pheno)$var.labels[varIDX]
varLabels[69]<-"ID"
FF<-set_label(FF, c(varLabels))
#create FF_labeled
#read in csv with metadata about variables 
FFmeta<-read.csv(paste0(datadir, "FFMetadata_v07.csv"))
#create long data dictionary  
FFlong<-FFmeta%>% select(old_name, contains('value'), contains('label')) %>% pivot_longer(cols = contains("value"), names_to="value", values_to="old_values")
FFlong<-FFlong%>%pivot_longer(cols = contains("label"), names_to="label", values_to="new_labels")
recodes_long<-FFlong %>% filter(gsub("value", "", value)==gsub("label", "", label)& !is.na(old_values)) %>% select(old_name, old_values, new_labels) 
#store continuous, binary, categorical etc
FFmyMeta<-FFmeta %>% filter(old_name %in% c(idvars, demovars, smokingvars, prenataldrugusevar))
myContinuous<-FFmyMeta %>% filter(type=="Continuous") %>%pull(old_name);
myContinuous<-colnames(FF)[colnames(FF) %in% myContinuous]
myBinary<-FFmyMeta%>%filter(type=="Binary")%>%pull(old_name); myBinary<-colnames(FF)[colnames(FF) %in% myBinary]
myCategorical_O<-FFmyMeta%>%filter(type=="Ordered Categorical")%>%pull(old_name); myCategorical_O<-colnames(FF)[colnames(FF) %in% myCategorical_O]
myCategorical_U<-FFmyMeta%>%filter(type=="Unordered Categorical")%>%pull(old_name); myCategorical_U<-colnames(FF)[colnames(FF) %in% myCategorical_U]
#filter to only vars of interest - not continuous variables  
myRecodes<-recodes_long %>% filter(old_name %in% c(myBinary, myCategorical_O, myCategorical_U))
#apply matchmaker
FF_labeled<-match_df(FF, dictionary=myRecodes, from = 'old_values', to='new_labels', by='old_name')
#reapply correct variable class types and na_codes
na_codes<-c("-1 Refuse", "-2 Don't know", "-3 Missing", "-9 Not in wave")
FF_labeled<-FF_labeled %>% mutate_at(vars(any_of(c(myBinary, myCategorical_O, myCategorical_U))), funs(as.factor(ifelse(. %in% na_codes, "Missing", .)))) %>% 
          mutate_at(vars(any_of(myContinuous)), funs(as.numeric(na_if(na_if(na_if(na_if(., "-3"), '-9'), '-6'), '-2'))))
#read back in the variable labels 
FF_labeled<-set_label(FF_labeled, get_label(FF))
FF_labels<-FF_labeled
```

## Cleaning pdqc data

```{r, pdqcReadin}
pdqc_all<-readRDS(paste0(datadir, 'OGData/', "pd_qc.rds"))
pdqc<-pdqc_all %>% filter(probe_fail_10==0 & sex==predicted_sex)
#What FF ids have methylation data?
methy_ids<-pdqc$id
#create slide variable fo pdqc 
pdqc$slide <- gsub('_.*', '', pdqc$MethID)
pdqc$array<-gsub('.*_', '', pdqc$MethID)
#ids with 9visit, 15visit or both visit
visit9ids<-pdqc%>%filter(childteen=='C')%>%pull(idnum)%>%unique()
visit15ids<-pdqc%>%filter(childteen=='T')%>%pull(idnum)%>%unique()
bothvisit_ids<-visit9ids[visit9ids %in% visit15ids]
visit9only_ids<-visit9ids[!(visit9ids %in%bothvisit_ids)]
visit15only_ids<-visit15ids[!(visit15ids %in%bothvisit_ids)]
```

```{r, createAnalyticSubsetFlags}
#create flag in FF for having methylation data
#and flag for having 2 visits of methylation data
FF_labeled<-FF_labeled %>% 
  mutate(MethylData=ifelse(id %in% methy_ids, "Analysis subset", "Not in analysis subset"),
         TwoVisitData=ifelse(id%in%bothvisit_ids,"Analysis subset - both visits", "Not in both visit analysis subset"))
#table(FF_labeled$MethylData)
#table(FF_labeled$TwoVisitData)
```

```{r, checkIDlengths}
if((length(visit15only_ids)+length(visit9only_ids)+length(bothvisit_ids)!=length(FF_labeled %>% filter(MethylData=="Analysis subset")%>%pull(id)))){print("WARNING ID LENGTHS DO NOT ADD UP")}
if(length(bothvisit_ids)!=length(FF_labeled %>% filter(TwoVisitData=="Analysis subset - both visits")%>%pull(id))){print("WARNING ID LENGTHS DO NOT ADD UP")}
```

### Filter technical replicates 
I filtered to only 1 technical replicate per person-visit, picking the replicate with the lower probe fail percentage

```{r}
#technical replicates will be duplicate id + childteen variables from pdqc
pdqc<-pdqc %>% mutate(newID=paste(idnum, childteen, sep='_'))
techrep_ids<- pdqc %>% select(newID) %>% group_by(newID)%>%filter(n()>1)
techreps<-pdqc %>% filter(newID %in% techrep_ids$newID)%>%arrange(idnum,childteen)
hiPctPF<-techreps %>% group_by(newID) %>%filter(probe_fail_pct==max(probe_fail_pct))%>%pull(MethID)
pdqc_clean<-pdqc %>% filter(!(MethID %in% hiPctPF) & childteen!='M')
myMethIDs<-pdqc_clean %>% pull(MethID)
```

### Plate, slide and array data

```{r read in  batch data}
batchData<-read.csv(file.path(datadir, 'Methylation_450K_array_batch_information.csv'))
```

# Creating new variables

## Creation of new sociodemographic and behavioral variables

### Prenatal maternal smoking exposure variable: binary smoking variable

```{r create binary smoking variable}
FF_labeled<-FF_labeled %>%mutate(smkPreg_binary=ifelse(m1g4%in% c("1 2+pk/d","2 1<pk<2","3 <1pk/d" ), "Yes", ifelse(m1g4%in%c("-2 Don't know", "-3 Missing", "-9 Not in wave"), NA, ifelse(m1g4== "4 None", "No", ifelse(m1g4=="Missing", "Missing", "FLAG")))))
createTable(compareGroups(m1g4~smkPreg_binary, data=FF_labeled))
```

### Ancestry variables

Ancestry is based on csv files of ancestry-specific PCS created by Dr. Erin Ware (per Jonah meeting Oct 6 2020). If an ID was in {african/european/hispanic}, then the individual is labeled as such.

```{r read in pc data and make ancestry variable}
# read in PCs
euroPC<-read.csv(paste0(datadir, "OGData/pcs/european.csv"))
afrPC<-read.csv(paste0(datadir, "OGData/pcs/african.csv"))
hisPC<-read.csv(paste0(datadir, "OGData/pcs/hispanic.csv"))
allPC<-read.table(paste0(datadir, "OGData/pcs/global_pcs.txt"), col.names=c("X", "idnum", paste("PC", 1:20, sep='')))
#individuals in each ancestry specific pc group can be labeled as that ancestry for categorical ancestry variable 
FF_labeled<-FF_labeled %>% mutate(ancestry=case_when(MethylData=="Analysis subset" & id %in% hisPC$idnum~"Hispanic ancestry", MethylData=="Analysis subset" & id %in%afrPC$idnum~"African ancestry", MethylData=="Analysis subset" & id %in% euroPC$idnum~"European ancestry", MethylData=="Analysis subset" & !(id %in%c(hisPC$idnum, afrPC$idnum, euroPC$idnum)) & (id %in% allPC$idnum)~"Other ancestry", MethylData=="Analysis subset" & !(id %in%c(hisPC$idnum, afrPC$idnum, euroPC$idnum)) & !(id %in% allPC$idnum)~"Missing PC data", MethylData!="Analysis subset"~"Not in analysis subset"))
#check ancestry variable
#table(FF_labeled$ancestry)
#with(FF_labeled, xtabs(~ancestry+cm1ethrace))
MissingPCdata<-FF_labeled %>% filter(ancestry=="Missing PC data")%>%select(id)
save(MissingPCdata, file = paste0(datadir, "CreatedData/MissingPC.Rdata"))
#add PC data to FF_labeled 
FF_labeled<-left_join(FF_labeled, allPC %>% mutate(id=as.character(idnum)) %>% select(id, PC1, PC2)%>%rename_at(vars(-id), function(x) paste0('global_', x, sep='')))
localPC<-rbind(euroPC, afrPC, hisPC)
FF_labeled<-left_join(FF_labeled, localPC  %>% mutate(id=as.character(idnum)) %>% select(id, PC1, PC2)%>% rename_at(vars(-id), function(x) paste0('local_', x, sep='')))
```

There were `r length(MissingPCdata$id)` individuals without PC data. None of these individuals had data in the global PC file, so I assume they are actually missing genetic data (for whatever reason, possibly QC).

### Other prenatal substance exposure variables

Recoded frequency of prenatal alcohol drinking and drug use into Yes/no binary variables for drinking and drug use during pregnancy. 
```{r PrenatalExposure}
#table(FF_labeled$m1g2) # during preg, alchohol freq
#table(FF_labeled$m1g3) # during preg, drugs freq 
#with(FF_labeled, table(m1g2, m1g3))
FF_labeled<-FF_labeled %>% mutate_at(c("m1g2", "m1g3"), .funs=funs(YesNoPreg=ifelse(.=="5 Never", "No", ifelse(. =="Missing", "Missing", "Yes"))))
with(FF_labeled, table(m1g2, m1g2_YesNoPreg))
with(FF_labeled, table(m1g3, m1g3_YesNoPreg))
#table(FF_labeled$m1g5) # in last year, alc/drugs interfered with work/relationships
#table(FF_labeled$m1g6) # ever sought help for drug/alc problems
```
### Postnatal smoke exposure

Maternal postnatal smoking at different survey points is fairly correlated, and alluvial plots of changes over time show broad consistency, so I decided to create a single smoking variable for maternal smoking at age 1 and age 5 - if mothers were smoking at either age 1 or age 5 then "Maternal smoking at age 1 or 5", if not smoking at age 1 and age 5 then "No maternal smoking at age 1 and age 5" and then if missing data at one age and no at the other, then "Missing" (and if missing at both ages, "Missing"). Also created a similar dose variable capturing maternal packs per day that could be used instead. Decided to use just age 1 and age 5 because at these two time points the mother was asked (as opposed to PCG, also the dataset of Fragile Families metadata is missing PCG questions from age 3 about smoking) and they both preceded the actual saliva collection for both saliva collection age points. Then use a smoking dose variable from the mother or PCG interview at wave of saliva collection as an additional second hand smoke exposure variable.

```{r PostnatalSmoke investigation}
#Baseline (1) -> Year 1/Age 1 (2) ->Year 3/Age3 (3) -> Year 5 (4) ->Year 9 (5) -> year 15 (6)
#Do you smoke in past month (how many packs per day)
#Mom            ->m2j5 (m2j5a)    ->                  -> m4j18 (m4j19) /p4 -> m5g17 (m5g18) / n5f17 (n5f18) 
#PCG                              p3a23a(p3a23b)                              n5f17 (n5f18)-> p6h76 (p6h77)
#seem to be missing p3 variables (PCG survey year 3) 
#number of people in household who smoke 
#p3a23->p5h15b->p6h78

#hours child around smoke 
#p3a21 -> p4a24->p5h15c

#with(FF_labeled, table(m2j5, m4j18))
#with(FF_labeled, table(m4j18, m5g17))
#with(FF_labeled, table(n5f17, m5g17))
#with(FF_labeled, table(p6h76, m5g17))

cor_plot_postnatalsmk<-plot_grid(FF_labeled %>% select(m2j5, m4j18, m5g17, n5f17, p6h74) %>% mutate_all(funs(as.numeric(.))) %>%ggcorr(label=T)+ggtitle("Correlation between Yes/No smoking (mother) \n at years 1, 5, & 9 & PCG at years 9 & 15")
,FF_labeled %>% select(m2j5a, m4j19, m5g18, n5f18, p6h77) %>% mutate_all(funs(as.numeric(.))) %>%ggcorr(label=T)+ggtitle("Correlation between packs per day smoking (mother) \n at years 1, 5, 9 & PCG at years 9 and 15"), nrow=2)

yesnosmkalluv<-to_lodes_form(as.data.frame(FF_labeled %>% select(m2j5, m4j18, m5g17, p6h74) %>% mutate(m5g17=stringr::str_to_title(m5g17))%>% group_by(m2j5, m4j18, m5g17, p6h74) %>% count()), axes=1:4, id='Cohort')

#ggplot(yesnosmkalluv, aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow(stat="alluvium", lode.guidance="backfront", color='darkgrey')+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9"))

smkdose<-to_lodes_form(as.data.frame(FF_labeled %>% select(m2j5a, m4j19, m5g18, p6h77)%>% mutate_at(vars(c("m2j5a", "m4j19", "m5g18")), funs(factor(substring(., 1, 1), labels=levels(FF_labeled$m2j5a)))) %>% group_by(m2j5a, m4j19, m5g18, p6h77) %>% count()), axes=1:4, id='Cohort')
#ggplot(smkdose, aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow()+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9"))

alluv_plot_postnatalsmk<-plot_grid(ggplot(yesnosmkalluv, aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow()+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9", "Age 15"))
+ggtitle("Alluvial plot of Yes/No smoking variables from age 1 to age 9 (mother) and age 15 (PCG)"), ggplot(smkdose %>% filter(!(stratum %in% c("-6 Skip", "Missing"))), aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow()+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9", "Age 15"))+ggtitle("Alluvial plot of smoking per day among those smoking from age 1 to age 9 (mother) and 15 (PCG)"), nrow=2)

#ggplot(FF_labeled, aes(x=p5h15b, y=p6h78))+geom_jitter()+stat_cor()+scale_x_continuous('# smoking in house age 9')+scale_y_continuous('# smoking in house age 15')+ggtitle("# individuals smoking in house correlation age 9 and age 15")
```

```{r plots for postnatal smoking}
cor_plot_postnatalsmk
alluv_plot_postnatalsmk
```

```{r make maternal postantal smoking vars for 1 and 5}
#Make any maternal postnatal smoking variable & dose categorization for maternal smoking at ages 1 & 5 
FF_labeled<-FF_labeled %>% mutate(PostnatalMaternalSmokingAny = case_when(
                                m2j5 =="1 Yes" ~ "Maternal smoking at age 1 or age 5",
                                m4j18=="1 Yes" ~ "Maternal smoking at age 1 or age 5",
                                m2j5 == "Missing" & m4j18=="Missing" ~ "Missing",
                                m2j5 == "Missing" & m4j18=="2 No" ~ "Missing",
                                m2j5 == "2 No" & m4j18=="Missing" ~ "Missing",
                                m2j5 == "2 No" & m4j18=="2 No" ~ "No maternal smoking at age 1 and age 5")) %>%
  mutate(PostnatalMaternalSmokingDose=case_when(
                                m2j5a %in% c("2 1pk/d", "3 1.5pk/d", "4 2pk/d", "5 >2pk/d") &
                                m4j19 %in% c("2 About pack/day", "3 Pack and half/day", "4 2 packs/day", "5   
                                          More 2 packs/day") ~ "Mom - pack/d or greater at both age 1 & 5", 
                                m2j5a %in% c("2 1pk/d", "3 1.5pk/d", "4 2pk/d", "5 >2pk/d") &
                                !(m4j19 %in% c("2 About pack/day", "3 Pack and half/day", "4 2 packs/day", "5 
                                          More 2 packs/day")) ~ "Mom - pack/d or greater at either age 1 or 5",
                                !(m2j5a %in% c("2 1pk/d", "3 1.5pk/d", "4 2pk/d", "5 >2pk/d")) &
                                m4j19 %in% c("2 About pack/day", "3 Pack and half/day", "4 2 packs/day", "5 
                                          More 2 packs/day") ~ "Mom - pack/d or greater at either age 1 or 5",
                                m2j5a %in% c("1 <1/2 pk/d") | m4j19 %in% c("1 Less half pack/day")~"Mom - less than half pack/d at either age 1 or 5", 
                                m2j5a %in% c("-6 Skip") & m4j19 %in% c("-6 Skip")~
                                  "No maternal smoking at age 1 or 5", 
                                m2j5a %in% c("Missing") & m4j19 %in% c("-6 Skip", "Missing")~"Missing", 
                                m2j5a %in% c("Missing", "-6 Skip") & m4j19 %in% c("Missing")~"Missing"))

with(FF_labeled, table(PostnatalMaternalSmokingAny, m2j5, m4j18))
#with(FF_labeled, table(m2j5a, m4j19, PostnatalMaternalSmokingDose))
#with(FF_labeled, table(PostnatalMaternalSmokingAny, PostnatalMaternalSmokingDose))
```

## Import and create methylation summary variables

### Cell type proportions

Cell type proportions currently estimated using [epidDISH](https://www.bioconductor.org/packages/release/bioc/vignettes/EpiDISH/inst/doc/EpiDISH.html#how-to-estimte-cell-type-fractions-using-dnam-data) package and the EpiFibIC reference (generic for epithelial tissues). This will be replaced with the saliva specific reference panel in [Middleton et al.](https://www.medrxiv.org/content/medrxiv/early/2020/09/15/2020.09.14.20191361.full.pdf).

```{r read in betaqc & celltype proportions}
betaqc<-readRDS(file=paste0(datadir, 'OGData/', "noob_filtered.rds"))

#cell type proportions 
celltypes<-epidish(beta.m = betaqc, ref.m = centEpiFibIC.m, method = "RPC")
estF_FF<-as.data.frame(celltypes$estF)
estF_FF$MethID = row.names(estF_FF)
```

In addition to the site-specific analysis proposed, Kelly and I decided to investigate certain DNA methylation summary metrics that are commonly used in the field. These include global methylation and methylation clocks (which are outcome-naive), and an ooutcome-predicated measure, polymethylation scores.

### Global methylation

We calculated global methylation as the mean of the CpG site-specific methylation values.

```{r global methylation and cpg sites of interest}
aprioriCG<-c("cg05575921", "cg04180046", "cg05549655", "cg14179389")
globalmethy<-colMeans(betaqc, na.rm=T)
globalmethdf<-as.data.frame(cbind("MethID"=names(globalmethy), "globalmethylation"=globalmethy))
aprioriCGdf<-as.data.frame(t(betaqc[aprioriCG, ]))
aprioriCGdf$MethID = colnames(betaqc)
```

### Methylation (or epigenetic) clocks

Methylation or epigenetic clocks are measures which exploit certain sets of CpG sites whose DNA methylation levels associate with subject age. These clocks are highly accurate molecular correlates of chronological age. However, variation in epigenetic clock measures among individuals of the same chronological age do exist and are assumed to reflect biological age.

Four different methylation clocks were created, based on different papers that used different tissue types and age groups for their training samples: [Horvath13](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2013-14-10-r115), [Horvath18](https://www.nature.com/articles/s41576-018-0004-3), [Pediatric](https://www.pnas.org/content/117/38/23329) and [Levine](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5940111/). Since this is a pediatric population, the pediatric clock is probably the best bet, but Jonah also suggested we look at the Levine clock.

```{r readInclocks}
clocks<-read.csv(file=paste0(datadir, "OGData/ffcw_n1776_8clocks.csv"), header = T)
colnames(clocks)[1]<-"MethID"
```

### Polymethylation score

Here, the idea is similar to a polygenetic score. Essentially, we want to create a summary metric of CpG sites that have associated with an exposure or outcome of interest in a previous independent cohort. We identify a previous study which conducted a regression analysis of our exposure of interest on CpG sites and extract those regression coefficients. The regression coefficients become weights for the CpG site values in our cohort. We sum the weighted CpG site values across sites for each sample to come up with a single per-sample measure of exposure-related methylation.

$Score_i=\sum^{m}_j Weight_j*CpG_{ij}$ where

$i$ indexes samples in our study; $j$ indexes CpG sites

$Weight_j$ is the regression coefficient from the independent external study for that $j$ CpG site

$CpG_{ij}$ is the CpG methylation value for that $j$ CpG in that $i$ sample in our study

$Score_i$ is the polymethylation score.

Because polymethylation scores are not widely used (yet), there isn't a standardized approach. Importantly, much of the time researchers use the (confusingly named) 'beta' value to measure methylation at a CpG site, which is the ratio of the methylated probe intensity to the overall intensity and ranges from 0-1. We were concerned that this could lead to sites with more methylation (closer to 1) being upweighted in the resulting score), so we implemented versions of the score where the $CpG_{ij}$ values were first mean-centered or z-score standardized before being used in the score.

For this analysis, we used a 2016 genome-wide consortium meta-analysisof prenatal maternal smoking and DNA methylation in blood samples, which used the same DNA methylation array. This study actually provided regression coefficients for all CpGs statistically significant after FDR correction ($m_{CpG}=6073$) in 4 different regressions:

1. Regression with sustained prenatal maternal smoking as the exposure, methylation in newborn cord blood samples as outcomes
2. Regression 1 adjusted for cell type confounding
3. Regression of sustained prenatal maternal smoking as the exposure, methylation in peripheral blood samples of older children as outcomes
4. Regression of any prenatal maternal smoking es the exposure, methylation in newborn cordblood as outcomes

These regression used normalized betas as the outcomes but found very little variation in results when using raw betas as the outcome (Spearman $\rho$=0.96 for regression coefficients, $\rho$=0.98 for $log_{10}(pvalues$))

```{r make polymethylaton score}
#read in cpgs from PMC4833289
cpgs<-read.csv(paste0(datadir, 'PMC4833289_SuppFile2CpGs.csv'), stringsAsFactors = F, header=F, skip=2)
#make headers
header1<-scan(paste0(datadir, 'PMC4833289_SuppFile2CpGs.csv'), nlines=1, what=character(), sep=',')
header1[7:26]<-c(rep("SSnewborn", 5), rep("SSnewbornCT", 5), rep("SSolder", 5), rep("anynewborn", 5))
header2<-paste(header1, scan(paste0(datadir, 'PMC4833289_SuppFile2CpGs.csv'), nlines=1, skip=1, what=character(), sep=','), sep='_')
names(cpgs)<-header2

#four meta-EWASes: SS_newborn (sustained smoking in newborns); SS_newbornCT (sustained smoking in newborns controlling for cell type) SS_older (sustained smoking in older children); any_newborn (any smoking in newborns)
#pivot_longer to get single column of coefs and a column for the analysis type
#and group into a list by analysis type
cpgs<-cpgs %>% pivot_longer(cols = c(matches("_Coef|_SE|_P|_FDR|_Direction")), names_to = c("set", ".value"), names_pattern="(.+)_(.+)")%>%rename_with(~gsub('_', '', .x))%>%group_split(set)
cpgs<-cpgs%>%setNames(lapply(cpgs, function(x){x$set %>% unique()}))

#make scores - no transform, zscore and centermean
#first, pull script 
source(file.path(codedir, 'UsefulCode', 'makePolyEpiScores.R'))
scores=makePolyEpiScore(m=betaqc, b=cpgs)
scores_z=makePolyEpiScore(m=betaqc, b=cpgs, transformopt = 'zscore')
scores_center=makePolyEpiScore(m=betaqc, b=cpgs, transformopt = 'meancenter')
polyscores=list('notransform'=scores, 'zscore'=scores_z, 'center'=scores_center)
polyscores<-lapply(polyscores, function(x) x%>% mutate(MethID=colnames(betaqc)))
save(polyscores, file = file.path(datadir, 'CreatedData', 'polymethylationscores.RDS'))
polyscores_wide<-lapply(seq_along(polyscores), function(i) polyscores[[i]] %>% setNames(c(paste0(colnames(polyscores[[i]])[1:4], '_', names(polyscores)[i]), 'MethID')))%>%reduce(left_join, by='MethID')
```

### Surrogate variables

```{r sva}

var.filter=5000
#make a matrix of model data 
#filter to only samples with methylation data (& kids??) 
pheno=FF_labeled%>%filter(MethylData=="Analysis subset")%>%left_join(pdqc%>%mutate(id=idnum)%>%select(MethID, id))
#check that dim of pheno and betaqc match

#want to make sure that the order of pheno and betaqc match
pheno=pheno[match(colnames(betaqc), pheno$MethID), ]
if(identical(colnames(betaqc), pheno$MethID)==FALSE){print('STOP: Sample names do not match')}

#make model matrix including variable of interest as factor variable & adjustment variables 
mod=model.matrix(~as.factor(smkPreg_binary), data=pheno)

#make null moddel containing only adjustment variables
mod0=model.matrix(~1, data=pheno)

#next, estimate the surrogate variables 
svobj=sva(betaqc, mod, mod0, vfilter=var.filter)
sv<-data.frame(svobj$sv); colnames(sv)<-paste('sv', 1:ncol(sv), sep='')
pheno=cbind(pheno, sv)

#bring the heatmap function
bring.the.heatmap.sv <- function(pd,covar,folder){
  library(gplots)

  nsv <- length(grep('sv',names(pd)))

  #generate correlation matrix between SVs and variables
  correlations <- matrix(0,nrow=length(covar),ncol=nsv)
  colnames(correlations) <- paste('sv',1:nsv,sep='')
  rownames(correlations) <- covar

  #take svs
  svs <- pd[,colnames(correlations)]

  #correlations between svs and variables:
  # cor test if continuous
  # anova is multi category
  # t-test if two category
  for(col in covar){
    var <- pd[,col]
    #check to see if it is numeric and isn't really a two category thing
    if(class(var)=='numeric' & dim(table(var))>2){
      correlations[col,] <- apply(svs,2,FUN = function(x,y) cor.test(x,y)$p.value,y=var)
    }else{
      if(length(levels(factor(var)))>2){
        correlations[col,] <- apply(svs,2,FUN = function(x,y) anova(lm(x~y))$'Pr(>F)'[1],y=var)
      }else{
        correlations[col,] <- apply(svs,2,FUN = function(x,y) t.test(x~y)$p.value,y=var)
      }
    }
  }

  pdf(paste0(folder,'heatmap-sv.pdf'))
  color.gradient <- colorRampPalette(c('firebrick','ivory3','royalblue'),bias=3)
  heatmap.2(correlations,Rowv=FALSE,Colv=FALSE,cexRow=1,cexCol=1,dendrogram='none',density.info='none',trace='none',
            col=color.gradient(20), key.xlab='p-value', main='Associations Between SVs\n and Given Covariates',
            lmat=rbind(c(0,3,3,3),c(2,0,4,0),c(0,1,1,1)), lhei=c(0.5,1,3.5), lwid=c(0.12,0.145,0.4,0.335), margin=c(4,7))
  dev.off()

  return('check your plot')
}

pheno=left_join(pheno, batchData)
bring.the.heatmap.sv(pheno, covar=c('Sample_Plate', 'Slide', 'Array'))

```

```{r make methyl dataframe}
clocks2<-clocks %>% mutate(idnum=gsub('a', '', idnum))%>%filter(MethID %in% pdqc$MethID)%>%select(MethID, idnum, childteen, horvath, pediatric, levine, grim) # currently clocks are missing 3 individuals who were in johns pipeline but not jonahs

methyldata<-left_join(left_join(left_join(left_join(left_join(left_join(pdqc_clean %>% select(MethID, slide, array,childteen, idnum), globalmethdf), clocks2), estF_FF), aprioriCGdf), polyscores_wide), batchData)
methyldata[, c("globalmethylation", aprioriCG)]<-lapply(methyldata[, c("globalmethylation", aprioriCG)],function(x) as.numeric(as.character(x)))
#lapply(methyldata, class)
methyldata_wide <- methyldata %>% select(-MethID)%>% pivot_wider(., names_from=childteen, values_from=any_of(colnames(methyldata)[4:length(colnames(methyldata))]))
```

```{r individual who is missing global methylation data but had apriori CpG and PMS data?, eval=F}
#weirdID<-methyldata %>% filter(is.na(globalmethylation))%>%pull(MethID)
#globalmethy[weirdID] #not actually missing global methylation data
#clocks %>% filter(MethID%in%weirdID) # #idnum from  not the same as in
#pdqc_clean %>% filter(MethID==weirdID) #this is just a typo in the idnum from data entry, see format of other #idnums
#pdqc_clean %>% pull(idnum) %>% head()
#solution: when creating methyldata, first trim alphabet from clocks$idnum that contain alphabetic characters --> implemented
#also misssing clocks for 3 sex outlier samples that weren't in jonahs' pipeline: 
table(is.na(methyldata$horvath13))
weirdIDs<-methyldata %>% filter(is.na(horvath13))%>%pull(MethID)
#pdqc_all %>% filter(MethID %in% weirdID) %>% xtabs(~predicted_sex_outlier, data=.)
```
## Distributions & checks of methylation variables

### Add methylation data to metadata

```{r makeMyFF}
varLabels2<-c(paste(colnames(FF_labeled)[1:66], varLabels), "Has any methylation data?", "Has two visits of methylation data", "Any maternal smoking during pregnancy", "Ancestry from PCA of child's genetic data", "PC 1 (global)", "PC 2 (global)", "PC 1 (within ancestry category)", "PC 2 (within ancestry category)",
"Any alcohol during pregnancy", "Any drugs during pregnancy", "Any postnatal maternal smoking (age 1 & 5)", "Postnatal maternal smoking dose at age 1 & 5")
FF_labeled<-set_label(FF_labeled, varLabels2)

myFF<-FF_labeled %>% filter(MethylData=="Analysis subset")%>%mutate(idnum=id)
myFF<-left_join(myFF, methyldata)
varLabels3<-c(varLabels2, "ID", "Methylation ID", "Slide", "Array (RC)", "Visit", "Global methylation", "horvath methylation clock", "pediatric methylation clock", "levine methylation clock", "GRIM methylation clock", "Epithelial cell proportion", "Fibroid cell proportion", "Immune cell proportion", "cg05575291 methylation", "cg04180046 methylation",  "cg05549655 methylation", "cg14179389 methylation", "PES: Any smoking, newborn cordblood", "PES: Sustained smoking, newborn cordblood", "PES: Sustained smoking, newborn cordblood (+CT control)", "PES: Sustained smoking, older children peripheral blood", "PES: Any smoking, newborn cordblood - zscore", "PES: Sustained smoking, newborn cordblood - zscore", "PES: Sustained smoking, newborn cordblood (+CT control) - zscore", "PES: Sustained smoking, older children peripheral blood - zscore", "PES: Any smoking, newborn cordblood - mean center", "PES: Sustained smoking, newborn cordblood - mean center", "PES: Sustained smoking, newborn cordblood (+CT control) - mean center", "PES: Sustained smoking, older children peripheral blood - mean center", 'Plate', 'Slide', 'Array')
myFF<-set_label(myFF, varLabels3)
```

## Correlation and distribution of global methylation and epigenetic clocks
```{r correlation global and clocks}
chart.Correlation(myFF %>%select(globalmethylation, levine, pediatric, grim))
```

### Correlation and distribution of polymethylation scores
```{r correlation polymethylation scores}
chart.Correlation(myFF %>%select(SSnewborn_notransform, SSnewbornCT_notransform, SSolder_notransform, anynewborn_notransform))
```

```{r save correlation chart}
postscript(file.path(outputdir, 'CorrelationPMSWeights.ps'))
chart.Correlation(myFF %>%select(SSnewborn_notransform, SSnewbornCT_notransform, SSolder_notransform, anynewborn_notransform))
dev.off()
```
### Correlation and distribution of polymethylation score coefficients

```{r correlation score coefficients}
coefs_methyl<-cbind('SSolderCoefs'=cpgs$SSolder$Coef, 'SSnewbornCTCoefs'=cpgs$SSnewbornCT$Coef, 'SSnewbornCoefs'=cpgs$SSnewborn$Coef, 'anynewbornCoefs'=cpgs$anynewborn$Coef)
chart.Correlation(coefs_methyl)
```

### Distributions of poly methylation scores with different transformations on beta methylation matrix

```{r compare PolyMethylation scores distributions, fig.height=11}
#lapply(polyscores, function(x){summary(x)})
#lapply(polyscores, function(x){apply(x, 2, sd)})
dist_plots=lapply(seq_along(polyscores), function(i){polyscores[[i]]%>%pivot_longer(cols=1:4, names_to='Score_Type', values_to='Score')%>%ggplot(aes(x=Score))+geom_histogram()+facet_wrap(~Score_Type)+ggtitle(names(polyscores)[[i]])+theme_classic()})
plot_grid(dist_plots[[1]], dist_plots[[2]], dist_plots[[3]], ncol=1)
```

```{r only newborn celltype controlled comparison of transform methods }
transform_labels<-c('Mean center', 'No transformation', 'Z-score standardization')
names(transform_labels)<-c('center', 'notransform', 'zscore')
postscript(file.path(outputdir, 'DistributionPMStransformation.ps'))
polyscores %>% bind_rows(.id='transform')%>%ggplot(aes(x=SSnewbornCT))+geom_histogram()+facet_wrap(~transform, scales='free', labeller=labeller(transform=transform_labels))+theme_classic()+theme(text=element_text(size=16))
dev.off()

```

As Kelly thought, the zscore standardized have a larger range and the biggest standard deviations, the standard deviations of the no transform and centered methods are the same, which makes sense as $V(aX+b)=a^2V(X)\implies V(\sum\beta_i*[x_i-\bar{x}])=V(\sum\beta_i*x_i-\beta_i*\bar{x})=V(\sum \beta_i*x_i-k)=V(\sum \beta_i*x_i)$.

### Correlation between methylation data
```{r}
methylation_data<-myFF %>% select(globalmethylation, pediatric, levine, Epi, Fib, IC, anynewborn_center, SSnewbornCT_center, SSolder_center)
chart.Correlation(methylation_data, histogram=T)
```

### Methylation variables over time or by visit

```{r celltype over time}
#Cell type
cellbytime<-myFF %>% pivot_longer(cols=c(Fib, Epi, IC), names_to="Celltype", values_to = "Proportion")

celltype_labels<-c('Epithelial cells', 'Fibroblasts', 'Immune cells'); names(celltype_labels)<-c('Epi', 'Fib', 'IC')

ggplot(cellbytime, aes(x=childteen, y=Proportion, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+facet_wrap(~Celltype, labeller=labeller(Celltype=celltype_labels))+stat_compare_means(label="..p.signif..")+ggtitle("Cell type proportions by visit")+theme_classic()+scale_color_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+scale_x_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+theme_classic()
```
```{r celltypes by visit spaghetti}
cairo_ps(file.path(outputdir, 'SpaghettiCellTypes.ps'))
ggplot(cellbytime, aes(x=childteen, y=Proportion, group=idnum))+geom_point()+geom_line(alpha=0.1, color='grey')+geom_smooth(aes(group=1), method='lm', fill='blue')+facet_wrap(~Celltype, labeller=labeller(Celltype=celltype_labels))+theme_classic()+scale_x_discrete('', labels=c('Age 9', 'Age 15'))+ylab('Proportion')+theme(text=element_text(size=16))
dev.off()
```

```{r clocks and global over time, fig.width=9}
global<-ggplot(myFF, aes(x=childteen, y=globalmethylation, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+ggtitle("Global methylation")+scale_color_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+scale_x_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+theme_classic()+theme(legend.position = 'bottom')
pediatric<-ggplot(myFF, aes(x=childteen, y=pediatric, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+ggtitle("Pediatric methylation clock")+scale_color_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+scale_x_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+theme_classic()+theme(legend.position = 'bottom')
levine<-ggplot(myFF, aes(x=childteen, y=levine, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+ggtitle("Levine methylation clock")+scale_color_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+scale_x_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+theme_classic()+theme(legend.position = 'bottom')
plot_grid(global, pediatric, levine, nrow=1)
```

```{r scores over time, fig.width=9}
#ggplot(myFF, aes(x=childteen, y=cg05575921, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+ggtitle("Methylation at cg05575921 by visit")
any<-ggplot(myFF, aes(x=childteen, y=anynewborn_center, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+ggtitle("Any smoking, \n newborn polymethylation score")+scale_color_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+scale_x_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+theme_classic()+theme(legend.position = 'bottom')
SSnewborn<-ggplot(myFF, aes(x=childteen, y=SSnewbornCT_center, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+ggtitle("Sustained smoking, \n newborn CT polymethylation score ")+scale_color_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+scale_x_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+theme_classic()+theme(legend.position = 'bottom')
SSolder<-ggplot(myFF, aes(x=childteen, y=SSolder_center, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+ggtitle("Sustained smoking, \n older children polymethylation score")+scale_color_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+scale_x_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+theme_classic()+theme(legend.position = 'bottom')
plot_grid(any, SSnewborn,SSolder, ncol=3)

```

Cell type proportions do differ between 9 and 15 year visits and although global methylation does not look different between these visits, methylation clocks do (a good check).

Possibly higher polymethylation scores at teen visit, although inconsistent across different score types.

### Correlation betwen polymethylation scores over time
```{r correlaton b/t polymethylation scores over time, fig.width=9}
PMS<-methyldata %>%select(idnum, childteen, colnames(polyscores_wide), -MethID)%>%pivot_longer(cols=any_of(colnames(polyscores_wide)), names_to="PMS_type", values_to="PolymethylationScore")%>%pivot_wider(names_from = childteen, values_from=PolymethylationScore)
PMSType_label<-c('Any smoking;\n newborn', 'Sustained smoking;\n newborn', 'Sustained smoking + cell type;\n newborn', 'Sustained smoking;\n older')
names(PMSType_label)<-c('anynewborn_center', 'SSnewborn_center', 'SSnewbornCT_center', 'SSolder_center')
ggplot(PMS%>% filter(str_detect(PMS_type, '_center')), aes(x=C, y=T))+geom_point()+facet_grid(~PMS_type, labeller=labeller(PMS_type=PMSType_label))+stat_cor()+geom_smooth(method=lm)+theme_classic()+scale_x_continuous('Age 9 score')+scale_y_continuous('Age 15 score')
```

```{r save correlation over time image}
postscript(file.path(outputdir, 'ScoreCorrelationVisits.ps'))
ggplot(PMS%>% filter(str_detect(PMS_type, '_center')), aes(x=C, y=T))+geom_point()+facet_grid(~PMS_type, labeller=labeller(PMS_type=PMSType_label))+stat_cor()+geom_smooth(method=lm)+theme_classic()+scale_x_continuous('Age 9 score')+scale_y_continuous('Age 15 score')+theme(text=element_text(size=16))
dev.off()
```

### Spaghetti plots of polymethylation scores over time
```{r Polymethylation scores over time, fig.width=9}
methyldata %>% pivot_longer(cols=contains('_center'), names_to='measure', values_to='Score')%>%ggplot(aes(x=childteen, y=Score, group=idnum))+geom_point()+geom_line(alpha=0.1, color='grey')+geom_smooth(aes(group=1), method='lm', fill='blue')+facet_wrap(~measure, labeller=labeller(measure=PMSType_label))+theme_classic()+scale_x_discrete('', labels=c('Age 9', 'Age 15'))+ylab('Score')
```

```{r save spaghetti poly}
cairo_ps(file.path(outputdir, 'SpaghettiScores.ps'))
methyldata %>% pivot_longer(cols=contains('_center'), names_to='measure', values_to='Score')%>%ggplot(aes(x=childteen, y=Score, group=idnum))+geom_point()+geom_line(alpha=0.1, color='grey')+geom_smooth(aes(group=1), method='lm', fill='blue')+facet_wrap(~measure, labeller=labeller(measure=PMSType_label))+theme_classic()+scale_x_discrete('', labels=c('Age 9', 'Age 15'))+ylab('Score')+theme(text=element_text(size=16))
dev.off()
```

## Make visit specific variables

### Child's age at time of interview/visit
```{r}
myFF %>% filter(is.na(cp6yagem) &is.na(hv6_yagem) & childteen=='T')%>%dim() #3
myFF %>% filter(is.na(cm5b_age) &is.na(hv5_agem) &childteen=='C')%>%dim() #1 
myFF %>% filter(!is.na(cm5b_age) & is.na(hv5_agem) & childteen=='C')%>%dim()#1
myFF %>% filter(!is.na(cp6yagem) &is.na(hv6_yagem))%>%dim() #990

myFF <- myFF %>% 
  mutate(ChildAgeAtIn=case_when(childteen=='C'~ cm5b_age, childteen=='T'~cp6yagem))%>%
  mutate(ChildAgeAtVisit=case_when(childteen=='C'~hv5_agem, childteen=='T'~hv6_yagem))%>% 
  mutate(ChildAgeComposite=case_when(childteen=='C' & !is.na(hv5_agem)~hv5_agem, 
         childteen=='C' & is.na(hv5_agem)~cm5b_age, 
         childteen=='T' & !is.na(hv6_yagem)~hv6_yagem, 
         childteen=='T' & is.na(hv6_yagem)~cp6yagem))

myFF %>% filter(childteen=='C')%>%group_by(ChildAgeAtVisit, ChildAgeComposite)%>%summarise(n=n())%>%filter(ChildAgeAtVisit!=ChildAgeComposite)

myFF %>% filter(childteen=='T')%>%group_by(ChildAgeAtVisit, ChildAgeComposite)%>%summarise(n=n())%>%filter(ChildAgeAtVisit!=ChildAgeComposite)

ggplot(myFF%>%filter(childteen=='C'), aes(cm5b_age, hv5_agem))+geom_point()+stat_cor()+ggtitle("Correlation between childs age at mom interview (9 yr) and childs age at home visit (9 yr)")
age5resid<-myFF %>% filter(childteen=='C')%>%mutate(age5_resid=cm5b_age-hv5_agem)%>%pull(age5_resid)
hist(age5resid)

age6resid<-myFF %>% filter(childteen=='T')%>%mutate(age6_resid=cp6yagem-hv6_yagem)%>%pull(age6_resid)
ggplot(myFF%>%filter(childteen=='T'), aes(cp6yagem,hv6_yagem))+geom_point()+stat_cor()+ggtitle("Correlation between childs age at mom interview (15 yr) and childs age at home visit (15 yr)")
hist(age6resid)

table(FF_labeled$cp6yagem-FF_labeled$hv6_yagem)
table(FF_labeled$cm5b_age-FF_labeled$hv5_agem)
table(is.na(myFF$ChildAgeComposite)) # 4

```

### Maternal or PCG smoking dose at time of sample

```{r visitMaternalSmoke}
myFF<-myFF %>% mutate(SmkAtVisitPastmonth=case_when(childteen=='C' & m5g17=='1 yes' & m5g18 %in% c("2 about a pack", "3 a pack and a half", "4 about 2 packs", 
                              "5 more than two  packs") ~ "Pack or more a day", 
                              childteen=='C' & m5g17=='1 yes' & m5g18=="1 less than half a pack a day" ~ "Less than pack a day", 
                              childteen=='C' & n5f17=="1 yes" & n5f18 =='2 about a pack' ~ "Pack or more a day",
                              childteen=='C' & n5f17=="1 yes" & n5f18 =="1 less than half a pack day" ~ "Less than pack a day",
                              childteen=='C' & m5g17=="1 yes" & m5g18 %in% c("-6 Skip", "Missing")~"Missing",
                              childteen == 'C' & n5f17%in%c("-7 N/A", "2 no") & m5g17=="2 no" ~ "No smoking", 
                              childteen == 'C' & n5f17%in%c("2 no") & m5g17%in%c("2 no", "Missing") ~ "No smoking", 
                              childteen == 'C' & n5f17 %in% c("Missing", "-7 N/A") & m5g17 =="Missing" ~ "Missing", 
                              childteen=='T' & p6h76 %in% c("-6 Skip", "0 None") ~ "No smoking", 
                              childteen=='T' & p6h77%in% c("1 5 or fewer cigarettes per day", "2 About half a pack a day, 10 cigarettes")~"Less than pack a day", 
                              childteen=='T' & p6h77 %in% c("3 About a pack a day, 20 cigarettes", "4 More than 1 pack a day")~"Pack or more a day",
                              childteen=='T' & p6h77 == "Missing"~"Missing"))%>%
              mutate(SmkAtVisitYesNo=case_when(SmkAtVisitPastmonth=="No smoking"~"No", 
                                               SmkAtVisitPastmonth %in% c("Less than pack a day", "Pack or more a day") ~ "Yes", 
                                               childteen=='T'& SmkAtVisitPastmonth=="Missing" & !(p6h76 %in% c("-6 Skip", "0 None", "Missing"))~"Yes",
                                               childteen=='C' & SmkAtVisitPastmonth=="Missing" & m5g17=='1 yes'~"Yes", 
                                               childteen=='C' & SmkAtVisitPastmonth=="Missing" & m5g17!='1 yes'~"Missing", 
                                               childteen=='T'& SmkAtVisitPastmonth=="Missing" & p6h76 %in% c("-6 Skip", "0 None", "Missing")~"Missing"))

myFF %>% filter(childteen=='C') %>% xtabs(~SmkAtVisitPastmonth+SmkAtVisitYesNo, data=.)
myFF %>% filter(childteen=='T') %>% xtabs(~SmkAtVisitPastmonth+SmkAtVisitYesNo, data=.)
myFF$SmkAtVisitPastmonth<-factor(myFF$SmkAtVisitPastmonth)
```
### Child smoking
Exclude children who reported smoking at age 9 and age 15. Interestingly, the children who report smoking at age 9 do not report smoking at age 15, and the parental perception of child smoking at age 9 does not match the children who say they have ever smoked a cigarette. For now using the 'ever smoked' variable from age 15 for the exclusion, but there is also a 'smoked in past month' that we could use instead (lose fewer children).

```{r, ChildSmoke}
with(myFF, table(k5f1l,childteen)) # 6 kids smoking at age 9 
with(myFF, table(p5q3cr, k5f1l)) # not the same kids whose parents say somewhat or very true that the child is using tobacco
with(myFF, table(k6d40, childteen)) # 41 kids say ever smoked a cigarette at age 15
#with(myFF, summary(k6d41))
with(myFF, table(k6d42, childteen)) # of these only 13 say they've smoked in the past month 
#with(myFF, summary(k6d43))
with(myFF, table(k5f1l, k6d40, childteen))# not the same kids at age 9 and 15
```

```{r}
varLabels4<-c(varLabels3, "Child's age at maternal/PCG (9/15 yr) interview (months)", "Child's age at home vist (9/15 years) (months)", "Constructed child's age at home visit (in months) (9/15 years)", "Mother/PCG (9/15 yr) smoking in month prior to interview (pk/day)", "Mother/PCG (9/15 yr) smoking in month prior to interview (yes/no)")
myFF<-set_label(myFF, varLabels4)
```

# Analysis

## Randomization of prenatal maternal smoking across slide numbers

```{r slide number vs prenatal maternal smoking, fig.width=8}
slide_count<-myFF %>% mutate(slide=as.factor(slide))%>% group_by(smkPreg_binary, slide, .drop=F)%>%count()%>%pivot_wider(id_cols = slide, names_from=smkPreg_binary, values_from=n)%>%arrange(Yes)%>%mutate(Proportion=Yes/(No+Yes))
hist_data<-slide_count %>% ggplot(aes(x=Proportion))+geom_histogram(binwidth = 0.1, fill='blue', alpha=0.5, breaks=seq(0, 1, 0.1))+ggtitle('Histogram proprotion smk (from data)')
hist_data_c<-slide_count %>% ggplot(aes(x=Yes))+geom_histogram( fill='blue', alpha=0.5, breaks=c(0:12))+scale_x_continuous(breaks=c(0:12))+ggtitle('Histogram # Yes smk (from data)')

kable(slide_count)%>%kable_styling()%>%scroll_box(width='100%', height='400px')

#simulation
set.seed(seed=20210216)
n<-nrow(myFF)
yes<-table(myFF$smkPreg_binary)[2]; no<-table(myFF$smkPreg_binary)[1]
values<-data.frame('smk'=sample(c('No', 'Yes'), n, replace=T, prob=c(no/(no+yes), yes/(no+yes))))
values_slides<-values %>% 
    group_by((row_number()-1) %/% (n()/152)) %>%
    nest()%>%unnest(cols=c(data))
colnames(values_slides)<-c('slide', 'smk')
slide_sim<-values_slides%>%mutate(slide=as.factor(slide)) %>% group_by(smk, slide, .drop=F)%>%count()%>%pivot_wider(id_cols = slide, names_from=smk, values_from=n)%>%arrange(Yes)%>%mutate(Proportion=Yes/(No+Yes))
hist_sim<-slide_sim%>% ggplot(aes(x=Proportion))+geom_histogram(binwidth = 0.1, fill='red', alpha=0.5, breaks=seq(0, 1, 0.1))+ggtitle('Histogram proprotion smk (from simulation)')
hist_sim_c<-slide_sim%>% ggplot(aes(x=Yes))+geom_histogram(fill='red', alpha=0.5, breaks=c(0:12))+scale_x_continuous(breaks=c(0:12))+ggtitle('Histogram # Yes smk (from simulation)')

hist_overlay<-slide_count %>% ggplot(aes(x=Proportion))+geom_histogram(fill='blue', binwidth = 0.1, alpha=0.5, breaks=seq(0, 1, 0.1))+geom_histogram(aes(x=slide_sim$Proportion), breaks=seq(0, 1, 0.1), fill='red', binwidth = 0.1, alpha=0.5)+xlim(0,1)
hist_overlay_c<-slide_count %>% ggplot(aes(x=Yes))+geom_histogram(fill='blue', alpha=0.5, breaks=c(0:12))+geom_histogram(aes(x=slide_sim$Yes), fill='red',  alpha=0.5, breaks=c(0:12))+scale_x_continuous(breaks=c(0:12))

#plot_grid(plot_grid(hist_data, hist_sim, nrow=2), hist_overlay, nrow=1)
plot_grid(plot_grid(hist_data_c, hist_sim_c, nrow=2), hist_overlay_c, nrow=1)

```

## Association of prenatal maternal smoking with plate

This was added 05/05/21 following 04/17/21 email from Jonah with batch variables from Princeton (thanks Jonah!)
```{r PreMSmk & plate (ie batch)}
Platechisq<-chisq.test(table(myFF$Sample_Plate, myFF$smkPreg_binary))
kable(table(myFF$Sample_Plate, myFF$smkPreg_binary))
```

Prenatal maternal smoking is not significantly associated with plate/batch ($\chi^2=$`r round(Platechisq$statistic, 2)`; P-value=`r round(Platechisq$p.value, 2)` )

## Flow chart of sample exclusions

```{r MakeFlowChart}
methylCohort<-pdqc_all %>% filter(childteen!='M')

basemodelvar<-c("smkPreg_binary", "ancestry", "cm1bsex", "cm1inpov", "Epi", "IC", "ChildAgeComposite")
basemodeldata<-myFF %>% filter_at(all_of(basemodelvar), all_vars(!(. %in%c("Missing", "Missing PC data"))& !is.na(.)))
child_base<-basemodeldata%>%filter(childteen=='C')
teen_base<-basemodeldata%>%filter(childteen=='T')

prenatalexposurevar<-c(basemodelvar, "m1g2_YesNoPreg", "m1g3_YesNoPreg")
prenatalexdata<-basemodeldata %>%filter_at(all_of(prenatalexposurevar), all_vars(!(. %in% c("Missing"))))
child_prenatal<-prenatalexdata%>%filter(childteen=='C')
teen_prenatal<-prenatalexdata%>%filter(childteen=='T')

secondhandsmokevars<-c(prenatalexposurevar, "PostnatalMaternalSmokingAny", "SmkAtVisitPastmonth")
secondhandsmkdata<-prenatalexdata%>%filter_at(all_of(secondhandsmokevars), all_vars(!(. %in% c("Missing"))))
child_secondhand<-secondhandsmkdata %>% filter(childteen=='C')
teen_secondhand<-secondhandsmkdata%>%filter(childteen=='T')
twovisit_secondhand<-secondhandsmkdata %>% filter(idnum %in% child_secondhand$idnum & idnum %in% teen_secondhand$idnum)

child_smoke<-secondhandsmkdata %>% filter(k5f1l!= "2 no" & childteen=='C')
child_nosmoke<-secondhandsmkdata %>% filter(k5f1l=="2 no" & childteen=='C')
teen_smoke<-secondhandsmkdata %>% filter(k6d40!="2 No" & childteen=='T')
teen_nosmoke<-secondhandsmkdata %>% filter(k6d40=="2 No" & childteen=='T')  
twovisit_nosmoke<-secondhandsmkdata %>% filter(idnum %in% child_nosmoke$idnum & idnum %in% teen_nosmoke$idnum)

fathersmkdata<-secondhandsmkdata %>% filter(f1g4!="Missing")

#Make Labels
n_fullFF<-paste0("Fragile Families cohort \n", "n=", FF_labeled %>% nrow())
n_methylQC<-paste0("Cohort with 450K methylation data \n", "n=", methylCohort$idnum%>%unique()%>%length(), "individuals with m=", methylCohort$MethID%>%length(), "samples")

n_QCloss<-paste0("Methylation data QC: 12 low-intensity samples dropped, \n", pdqc_all%>%filter(probe_fail_10==T)%>%nrow()-12, " samples with >10% of probes failing dropped, \n", pdqc_all%>%filter(sex!=predicted_sex & probe_fail_10==F & childteen!='M')%>%nrow(), " discordant sex samples dropped, \n & ", nrow(pdqc)-nrow(pdqc_clean), " duplicate samples dropped")
n_methylFF<-paste0("Cohort with quality controlled 450K data \n","n=", myFF %>% select(id) %>% unique() %>% nrow(), " individuals with m=", myFF%>%nrow(), " samples")

n_basemodel<-paste0("Base models: \n", "n=", basemodeldata %>% select(id) %>% unique() %>% nrow(), " individuals with m=", basemodeldata %>%nrow(), " samples")
n_baseexclusions<-paste0("n=",  myFF %>% filter_at(all_of(basemodelvar), any_vars((. %in%c("Missing", "Missing PC data"))))%>% select(id) %>% unique() %>% nrow(), " individuals & m=", myFF %>% filter_at(all_of(basemodelvar), any_vars((. %in%c("Missing", "Missing PC data")))) %>% nrow(), " samples missing principal components of ancestry; \n n=", myFF %>% filter_at(all_of(basemodelvar), any_vars((is.na(.))))%>%select(id)%>%unique()%>%nrow(), " individuals & m=", myFF %>% filter_at(all_of(basemodelvar), any_vars((is.na(.))))%>%nrow(), " samples missing child age at maternal/PCG interview")

n_prenatalexposures<-paste0("Prenatal exposures models: \n","n=", prenatalexdata %>% select(id) %>% unique() %>% nrow(), " individuals with m=", prenatalexdata %>%nrow(), " samples")
n_prenatalexclusions<-paste0("n=", basemodeldata %>% filter_at(all_of(prenatalexposurevar), any_vars(. %in% c("Missing"))) %>% select(id) %>% unique() %>% nrow(), " individuals & m=", basemodeldata %>% filter_at(all_of(prenatalexposurevar), any_vars(. %in% c("Missing"))) %>% nrow(), " samples missing \n data on maternal prenatal alcohol and drug use")

n_secondhandSmk<-paste0("Secondhand smoke exposure models: \n","n=", secondhandsmkdata %>% select(id) %>% unique() %>% nrow(), " individuals with m=", secondhandsmkdata %>%nrow(), " samples")
n_secondhandexclusions<-paste0("n=", prenatalexdata$id[!(prenatalexdata$id %in% secondhandsmkdata$id)] %>% unique()%>% length(), " individuals & m=", prenatalexdata %>% filter_at(all_of(secondhandsmokevars), any_vars(. %in% c("Missing"))) %>% nrow(), " samples missing data \n on maternal or primary care giver postnatal smoking")

n_Child<-paste0('n=', child_secondhand  %>% nrow(), " individuals with a 9 year visit")
n_ChildSmkexcluded<-paste0("Exclude children who smoke: \n",'n=', child_nosmoke %>% nrow(), " individuals with a 9 year visit")
n_ChildSmkexclusions<-paste0('n=', child_smoke%>%nrow(), " individual \n smoking at 9 year visit \n or missing focal child smoking data")
n_Teen<-paste0('n=', teen_secondhand  %>% nrow(), " individuals with a 15 year visit")
n_TeenSmkexcluded<-paste0("Exclude children who smoke: \n",'n=', teen_nosmoke %>% nrow(), " individuals with a 15 year visit")
n_TeenSmkexclusions<-paste0('n=', teen_smoke%>%nrow(), " individuals \n smoking at 15 year visit \n or missing focal child smoking data")

n_sensFatherSmoking<-paste0("Sensitivity analysis: compare effects of maternal and paternal prenatal smoking \n", "n=", fathersmkdata %>% select(id) %>% unique() %>% nrow(), " individuals with m=", fathersmkdata %>%nrow(), " samples", "\n with prenatal paternal smoking data")
n_sensFatherSmokingEx<- paste0("n=", secondhandsmkdata$id[!(secondhandsmkdata$id %in% fathersmkdata$id)] %>% unique()%>% length(), " individuals & m=", secondhandsmkdata %>% filter(f1g4 %in% c("Missing")) %>% nrow(), " samples missing data on paternal prenatal smoking")

n_twovisits<-paste0("n=", twovisit_secondhand%>%pull(idnum) %>%unique() %>%length(), " individuals with both a 9 and 15 year visit")
n_twovisitsSmkExcluded<-paste0("Exclude children who smoke: \n", "n=", twovisit_nosmoke%>%pull(idnum) %>%unique() %>%length(), " individuals with both a 9 and 15 year visit")
n_twovisitSmkExclusions<-paste0("n=", (twovisit_secondhand%>%pull(idnum) %>%unique() %>%length())-(twovisit_nosmoke%>%pull(idnum) %>%unique() %>%length()), " individuals smoking at \n either 9 and 15 year visit or \n missing focal child smoking data")

```

```{r drawFlowchart}
graph1<-DiagrammeR::grViz("
   digraph graph2 {
   compound=true;
   graph [layout = dot]

   node [shape = rectangle, width=4]
   a [label = '@@1']
   b [label = '@@2']
   c [label = '@@3']
   d [label = '@@4']
   e [label = '@@5']
   f [label = '@@6']
   g [label = '@@7']
   h [label = '@@8']
   i [label = '@@9', group=g1]
   j [label = '@@10']
   k [label = '@@16']
   l [label = '@@17']
   m [label = '@@19']

   #fake nodes for connecting to edges
   node [shape=point, width=0.01, height=0.01]
   a1
   b1
   c1
   d1
#   e1
  f1 [group=g1]
  g1
  k1

  #exclusion nodes
  node [shape = rectangle, width=4]
  aa [label = '@@20']
  bb [label = '@@11']
  cc [label = '@@12']
  dd [label = '@@13']
  ff [label = '@@14', width=3]
  gg [label = '@@15', width=3]
  kk [label = '@@18', width=3]

  #draw connections
  a->m
  m->a1 [arrowhead=none]
  a1->b
  b->b1 [arrowhead=none]
  b1->c
  c->c1 [arrowhead=none]
  c1->d
  d->d1 [arrowhead=none]
  d1->e
  e->f; e->g; e->k
{rank=same; h->e [dir=back, minlen=2, style=dashed]}
  f->f1 [arrowhead=none]
  g->g1 [arrowhead=none]
  k->k1 [arrowhead=none]
  f1->i 
  g1->j 
  k1->l

{rank=same; a1->aa [minlen=2]}   
{rank=same; b1->bb [minlen=2]} 
{rank=same; c1->cc [minlen=2]}
{rank=same; d1->dd [minlen=2]}
{rank=same; f1->ff [minlen=2]}
{rank=same; g1->gg [minlen=2]}
{rank=same; k1->kk [minlen=2]}

#invisible constraints to force ordering
ff->i [style=invis]
   }

   [1]: n_fullFF
   [2]: n_methylFF
   [3]: n_basemodel
   [4]: n_prenatalexposures
   [5]: n_secondhandSmk
   [6]: n_Child
   [7]: n_Teen
   [8]: n_sensFatherSmoking
   [9]: n_ChildSmkexcluded
   [10]: n_TeenSmkexcluded
   [11]: n_baseexclusions
   [12]: n_prenatalexclusions
   [13]: n_secondhandexclusions
   [14]: n_ChildSmkexclusions
   [15]: n_TeenSmkexclusions
   [16]: n_twovisits
   [17]: n_twovisitsSmkExcluded
   [18]: n_twovisitSmkExclusions
   [19]: n_methylQC
   [20]: n_QCloss
")
```

```{r drawGraph, fig.height=11, fig.1idth=8}
graph1
```

## Comparison of analytic sample: whole sample of Fragile Families vs methylation data vs analytic subset

```{r pick analysis subset}
analysis_data<-rbind(child_nosmoke, teen_nosmoke)
```

```{r, warning=F, message=F}
methyl1<-createTable(compareGroups(MethylData~smkPreg_binary+cm1inpov+cm1ethrace+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+k5f1l+k6d40+f1g4, data=FF_labeled), show.all=T)
export2md(methyl1, caption="Exclusion table of all Fragile Families vs those with quality controlled methylation data for main predictors and key demographic variables from first wave")
```

```{r final analytic subset, warning=F, message=F}
myFF$MethylData2<-ifelse(myFF$MethID %in% c(child_nosmoke$MethID, teen_nosmoke$MethID), 'Final analysis group', 'Not in final analysis group')
set_label(myFF$MethylData2)<-'Exclusions based on missing data'
methyl2<-createTable(compareGroups(MethylData2~smkPreg_binary+globalmethylation+pediatric+Epi+IC+anynewborn_center+SSnewbornCT_center+SSolder_center+cm1inpov+ancestry+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth, data=myFF), show.all=T)
export2md(strataTable(methyl2, 'childteen'), caption="Exclusion table of subset with quality controlled methylation data vs final analytic subset for main predictor, outcomes and key demographic variables")

```

# Datasets for modeling
Variables

Base model: Binary smoking variable [X], ancestry categorization, baby sex [X], maternal income to poverty ratio [X],  child's age in month at visit of interest (note, currently using a constructed child's age at visit, where individuals having a missing child's age at home visit was supplemented with child's age at home interview) [X], cell types (Epi + IC, may need to take one out),

Prenatal exposures: Base model variables + prenatal maternal alcohol use (Yes/no) [X], prenatal maternal drug use (yes/no)

Secondhand smoke exposure models: Prenatal exposures model+ any maternal smoking at age 1 & 5 [X], maternal or PCG smoking dose at visit of interest (age 9 or 15) [X]

First hand smoking: Secondhand smoke exposure models + exclude children with missing or reported firsthand smoking at visit of interest (age 9 or 15)

Paternal smoking: Filter to individuals with data on prenatal paternal smoking and compare effect size of prenatal paternal to prenatal maternal (as per Wiklund et al. 2019, idea is that this captures additional confounding)

Ancestry specific models - by ancestry groups; replace ancestry control with control for local PCs within ancestry group

Longitduinal models - no additional filtering needed

```{r, saveData}
save(child_base, file=paste0(datadir, 'CreatedData/', 'ChildBaseModelData.Rdata'))
save(teen_base, file=paste0(datadir, 'CreatedData/', 'ChildBaseModelData.Rdata'))
save(basemodeldata, file=paste0(datadir, 'CreatedData/', 'BaseModelData.Rdata'))

save(child_prenatal, file=paste0(datadir, 'CreatedData/', 'ChildPrenatalExModelData.Rdata'))
save(teen_prenatal, file=paste0(datadir, 'CreatedData/', 'PrenatalExModelData.Rdata'))
save(prenatalexdata, file=paste0(datadir, 'CreatedData/', 'PrenatalExModelData.Rdata'))

save(child_secondhand, file=paste0(datadir, 'CreatedData/', 'ChildSecondHandSmkModelData.Rdata'))
save(teen_secondhand, file=paste0(datadir, 'CreatedData/', 'TeenSecondHandSmkModelData.Rdata'))
save(twovisit_secondhand, file=paste0(datadir, 'CreatedData/', 'TwoVisitSecondHandSmkModelData.Rdata'))
save(secondhandsmkdata, file=paste0(datadir, 'CreatedData/', 'SecondHandSmkModelData.Rdata'))

save(child_nosmoke, file=paste0(datadir, 'CreatedData/', 'ChildNoSmkModelData.Rdata'))
save(teen_nosmoke, file=paste0(datadir, 'CreatedData/', 'TeenNoSmkModelData.Rdata'))
save(twovisit_nosmoke, file=paste0(datadir, 'CreatedData/', 'TwoVisitNoSmkModelData.Rdata'))

save(fathersmkdata, file=paste0(datadir, 'CreatedData/', 'FatherSmkModelData.Rdata'))

save(list=c("basemodeldata", "prenatalexdata", "secondhandsmkdata", "child_secondhand", "teen_secondhand", "twovisit_secondhand", "twovisit_nosmoke", "child_nosmoke", "teen_nosmoke", "fathersmkdata"), file=paste0(datadir, 'CreatedData/', Sys.Date(), 'CreatedData.Rdata'))
```

### 2021-09-07 Reports/FFEWAS.Rmd
first commit on nfs

---
title: "Fragile Families - Epigenome wide association study"
author: "Freida Blostein"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    code_folding: show
    highlight: tango
    number_sections: yes
    theme: sandstone
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include=FALSE, eval=F}
if (!requireNamespace("BiocManager")) install.packages("Bioconductor")

BiocManager::install(c(
  "bumphunter", # Differentially methylated region analysis
  "minfi", # Many functions for methylation analysis
  "sva", # Batch effect correction
  "limma", # Single-site association analysis
  "IlluminaHumanMethylation450kmanifest", # Manifest file
  "IlluminaHumanMethylation450kanno.ilmn12.hg19", # Annotation file
  "GO.dv", # Gene ontology queries
  # "missMethyl",
  "GEOquery"
))

install.packages(c(
  "RColorBrewer", # Color palettes for data visualization,
  "ggplot2", # Useful and aesthetic plotting functions
  "MatrixEQTL", # Fast eQTL and meQTL analyses
  "qqman", # QQ and manhattan plots
  "nlme", # Non-linear mixed effects models
  "BiasedUrn" # Hypergeometric distributions
))
```

```{r load libraries, include=F}
knitr::opts_chunk$set(echo = TRUE)
library(minfi)
library(limma)
library(matrixStats)
library(MASS)
#library(sva) --> current problem with png package, lib.png.16 https://stackoverflow.com/questions/47424249/unable-to-install-the-package-png-on-r-possible-conflict-with-anaconda/50935430
#library(Hmisc)
#library( missMethyl )
library(ggplot2)
library(qqman)
library(tidyr)
library(dplyr)
library(ggupset)
library(purrr)
library(broom.mixed)
library(future)
library(furrr)
library(tibble)
library(kableExtra)
library(cowplot)
library(matrixStats)
library(lme4)
library(lmerTest)
library(ggpubr)
data.dir="/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/"
limma.out.dir<-file.path(data.dir, 'CreatedData', 'limmaModels')
outputdir<-"/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Output"
```

```{r read in data}
#load in betaqc matrix
#betaqc<-readRDS(file=paste0(data.dir, 'OGData/', "noob_filtered.rds"))
#transform to m values in case want to use
#m_val<-function(x){log2(x/(1-x))}
#mvaluesqc<-apply(betaqc, 2, m_val)

#load in metadata - child no smoke and teen no smoke
load(file=paste0(data.dir, 'CreatedData/', 'ChildNoSmkModelData.Rdata'))
load(file=paste0(data.dir, 'CreatedData/', 'TeenNoSmkModelData.Rdata'))
#drop missing factor levels for both data sets
child_nosmoke<-child_nosmoke%>%droplevels()
teen_nosmoke<-teen_nosmoke%>%droplevels()
#set factor levels
child_nosmoke$PostnatalMaternalSmokingAny<-factor(child_nosmoke$PostnatalMaternalSmokingAny, levels=c("No maternal smoking at age 1 and age 5", "Maternal smoking at age 1 or age 5"))
child_nosmoke$SmkAtVisitPastmonth<-factor(child_nosmoke$SmkAtVisitPastmonth, levels=c("No smoking", "Less than pack a day", "Pack or more a day"))
teen_nosmoke$PostnatalMaternalSmokingAny<-factor(teen_nosmoke$PostnatalMaternalSmokingAny, levels=c("No maternal smoking at age 1 and age 5", "Maternal smoking at age 1 or age 5"))
teen_nosmoke$SmkAtVisitPastmonth<-factor(teen_nosmoke$SmkAtVisitPastmonth, levels=c("No smoking", "Less than pack a day", "Pack or more a day"))
modeldata=rbind(child_nosmoke, teen_nosmoke)
```

# Preprocessing

```{r filter invariant probes}
#probeVars<-rowVars(betaqc)
#saveRDS(probeVars, file=paste0(data.dir, 'CreatedData/', "probevariance.rds"))
probeVars<-readRDS(file=paste0(data.dir, 'CreatedData/', "probevariance.rds"))
varCut<-0.0002
a<-ggplot()+aes(probeVars)+geom_histogram(boundary=0)+theme_classic2()+ggtitle('Histogram of probe variances')+scale_x_continuous('Probe variances', breaks=seq(0, 0.15, 0.025))+annotate('rect', xmin=0, xmax=0.01, ymin=0, ymax=4e5, fill=alpha('grey', 0), color='orange')+annotate('segment', x=0.01, xend=0.04, y=1e5, yend=2e5, arrow=arrow(), color='orange')

b<-ggplot()+aes(probeVars[probeVars<0.01])+geom_histogram()+theme_classic2()+ggtitle('Histogram of probe variances<0.01')+scale_x_continuous('Probe variances', breaks=seq(0, 0.01, 0.005))+geom_vline(xintercept = varCut, color='red', linetype='dashed')+annotate('text', x=0.003, y=1e05, label=paste0('Cut ', length(probeVars[probeVars<varCut]), ' invariant probes \n', 'using variance cut of >', varCut), color='red')

#betaqc<-betaqc[rowVars(betaqc)>varCut,]
#saveRDS(betaqc, file=paste0(data.dir, 'CreatedData/', 'betaTrim.Rds'))
betaqc<-readRDS(file=paste0(data.dir, 'CreatedData/', 'betaTrim.Rds'))
```

```{r draw probe variance histogram, fig.width=8}
ggdraw(a)+draw_plot(b, 0.35, 0.35, 0.55, 0.55)
```

Based on the above histogram, I select `r varCut` as the lower limit for probe variance; all probes with variances<`r varCut` were removed from the set ($m_{CpG}$=`r length(probeVars[probeVars<varCut])` removed), leaving a total of $m_{CpG}$=`r nrow(betaqc)` probes for the EWAS.

```{r check variance filter dimensions}
if(!(nrow(betaqc)+length(probeVars[probeVars<varCut])==length(probeVars))){print('STOP: invariant probe filter cut before vs after probe count incorrect')}
```

```{r annotation}
### append annotation information
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
data(Locations)
data(Islands.UCSC)
data(Other)
pos<-Locations[rownames(betaqc),'pos']
chrnames <- as.character(Locations[rownames(betaqc),'chr'])

add_annotations<-function(data){
  #add annotations
Locations <- Locations[rownames(data), ]
Islands.UCSC <- Islands.UCSC[rownames(data),]
Other <- Other[rownames(data),]
# add columns of interest from annotation to annotated version
data.annotated <- data
data.annotated$chr <- Locations$chr
data.annotated$pos <- Locations$pos
data.annotated$island <- Islands.UCSC$Relation_to_Island
data.annotated$gene <- Other$UCSC_RefGene_Name
data.annotated$regulatory_feature <- Other$Regulatory_Feature_Group
return(data.annotated)
}

```

# Cross sectional epigenome wide association study: Children Age 9

```{r beta filtering and checks}
#filter betaqc to child samples only 
beta_c <-betaqc[, colnames(betaqc) %in% child_nosmoke$MethID]#; rm(betaqc)
#make sure that the order of the samples matches in the metadata matrix and the beta matrix are ordered the same
child_nosmoke<-child_nosmoke[match(colnames(beta_c), child_nosmoke$MethID), ]
#checks
if(dim(beta_c)[2]!=dim(child_nosmoke)[1]){print('STOP: Sample #s do not match')}
if(identical(colnames(beta_c), child_nosmoke$MethID)==FALSE){print('STOP: Sample names do not match')}
```

```{r child cross sectional ewas global model1, eval=F}
#model 1: make model matrix
mod1=model.matrix(~factor(child_nosmoke$smkPreg_binary)+child_nosmoke$global_PC1+child_nosmoke$global_PC2+factor(child_nosmoke$cm1bsex)+child_nosmoke$cm1inpov+child_nosmoke$ChildAgeComposite+child_nosmoke$Epi+child_nosmoke$IC)

#run EWAS with empirical st bayes approach to reduce standard error
out<-lmFit(beta_c, mod1)
out<-eBayes(out)
saveRDS(out, file=file.path(limma.out.dir, 'limmaModel1.Rda'))

#pull out top hits
child.ss.hits<-topTable(out, coef=2, number=nrow(beta_c), confint = T)
rm(mod1, out)

# Annotate top hits
child.m1.ss.hits.annotated <- add_annotations(child.ss.hits)
saveRDS(child.m1.ss.hits.annotated, file=file.path(limma.out.dir, 'Model1TopHits.Rda'))
```

```{r load model1}
child.m1.ss.hits.annotated <-readRDS(file=file.path(limma.out.dir, 'Model1TopHits.Rda'))
```

```{r child cross sectional ewas global model2, eval=F}
#model 2: make model matrix
mod2=model.matrix(~factor(child_nosmoke$smkPreg_binary)+child_nosmoke$global_PC1+child_nosmoke$global_PC2+factor(child_nosmoke$cm1bsex)+child_nosmoke$cm1inpov+child_nosmoke$ChildAgeComposite+child_nosmoke$Epi+child_nosmoke$IC+factor(child_nosmoke$m1g2_YesNoPreg)+factor(child_nosmoke$m1g3_YesNoPreg))

#run EWAS with empirical st bayes approach to reduce standard error
out<-lmFit(beta_c, mod2)
out<-eBayes(out)
saveRDS(out, file=file.path(limma.out.dir, 'limmaModel2.Rda'))

#pull out top hits
child.ss.hits<-topTable(out, coef=2, number=nrow(beta_c))
rm(mod2, out)

#annotate
child.m2.ss.hits.annotated <- add_annotations(child.ss.hits)
saveRDS(child.m2.ss.hits.annotated, file=file.path(limma.out.dir, 'Model2TopHits.Rda'))
```

```{r load model2}
child.m2.ss.hits.annotated <-readRDS(file=file.path(limma.out.dir, 'Model2TopHits.Rda'))
```

```{r child cross sectional ewas global model3, eval=F}
#model 3: make model matrix
mod3=model.matrix(~factor(child_nosmoke$smkPreg_binary)+child_nosmoke$global_PC1+child_nosmoke$global_PC2+factor(child_nosmoke$cm1bsex)+child_nosmoke$cm1inpov+child_nosmoke$ChildAgeComposite+child_nosmoke$Epi+child_nosmoke$IC+factor(child_nosmoke$m1g2_YesNoPreg)+factor(child_nosmoke$m1g3_YesNoPreg)+factor(child_nosmoke$PostnatalMaternalSmokingAny)+factor(child_nosmoke$SmkAtVisitPastmonth))

#run EWAS with empirical st bayes approach to reduce standard error
out<-lmFit(beta_c, mod3)
out<-eBayes(out)
saveRDS(out, file=file.path(limma.out.dir, 'limmaModel3.Rda'))
#pull out top hits
child.ss.hits<-topTable(out, coef=2, number=nrow(beta_c))
rm(mod3, out)

#annotate
child.m3.ss.hits.annotated <- add_annotations(child.ss.hits)
saveRDS(child.m3.ss.hits.annotated, file=file.path(limma.out.dir, 'Model3TopHits.Rda'))
```

```{r load model3}
child.m3.ss.hits.annotated <-readRDS(file=file.path(limma.out.dir, 'Model3TopHits.Rda'))
```

### Future analysis: compare top CpGs when run in ancestry specific models

## QQ plots

```{r calculate lambdas}
calcLambda<-function(df){
  o<- -log10(sort(df$P.Value, decreasing = F))
  e<- -log10(ppoints(length(df$P.Value)))
  lambda<-median(o)/median(e)
  return(lambda)
}

l_c1<-calcLambda(child.m1.ss.hits.annotated)
l_c2<-calcLambda(child.m2.ss.hits.annotated)
l_c3<-calcLambda(child.m3.ss.hits.annotated)

```

```{r QQPlots, fig.height=16}
par(mfrow=c(3, 1), mar=c(5.1, 4.1, 5.1, 2.1))
qq(child.m1.ss.hits.annotated$P.Value, main='Base model: prenatal smoking, genetic PCs, sex \n age, income to poverty ratio, Epi, Fib')
mtext(paste0(expression(lambda), '=', round(l_c1, 3)), side=3)
qq(child.m2.ss.hits.annotated$P.Value, main='Prenatal exposures model: \n Base model vars+ prenatal alcohol+prenatal other drug')
mtext(paste0(expression(lambda), '=', round(l_c2, 3)), side=3)
qq(child.m3.ss.hits.annotated$P.Value, main='Secondhand smoking model:  Prenatal exposure model + postnatal maternal smoking any at 1 or 5 + \n postnatal maternal smoking pks/day in month prior to visit')
mtext(paste0(expression(lambda), '=', round(l_c3, 3)), side=3)
```

## Manhattan plot

```{r Manhattan plots}
makeManhattan<-function(ss.hits, beta){
  pos.ss <- pos[match(rownames(ss.hits), rownames(beta))]
  chr.ss <- chrnames[match(rownames(ss.hits), rownames(beta))]
  chr.ss.num <- as.numeric(unlist(lapply(chr.ss, function(x) {
    strsplit(x, "chr", fixed = TRUE)[[1]][2]
  })))

  # Construct data frame for function
  formanhattan <- data.frame(pos.ss, chr.ss.num, ss.hits$P.Value)
  colnames(formanhattan) <- c("BP", "CHR", "P")

  formanhattan <- formanhattan[-which(is.na(formanhattan$CHR)), ]
  return(formanhattan)
}

m1.forman<-makeManhattan(child.m1.ss.hits.annotated, beta_c)
m2.forman<-makeManhattan(child.m2.ss.hits.annotated, beta_c)
m3.forman<-makeManhattan(child.m3.ss.hits.annotated, beta_c)

```

### Manhattan in children (Age 9) model 1

```{r manhattan model1}
manhattan(m1.forman)
```

## Beta vs Beta plots within Age 9 models

```{r beta vs beta within child}
child_models<-list('child_model1'=child.m1.ss.hits.annotated, 'child_model2'=child.m2.ss.hits.annotated, 'child_model3'=child.m3.ss.hits.annotated)%>%lapply(., function(x) rownames_to_column(x, var='CpG'))%>%Map(cbind, ., ModelType=names(.))

beta_child<-lapply(child_models, function(x) x %>%dplyr::select(CpG, logFC, ModelType)%>%pivot_wider(id_cols = CpG, values_from=logFC, names_from=ModelType))%>%purrr::reduce(left_join, by='CpG')

#bind_rows(.id='ModelType')%>%mutate('Model'=gsub('.*_', '', ModelType), 'Age'=gsub('_.*', '', ModelType))

sig_beta_m1<-child_models[[1]]%>%filter(P.Value<1e-3)%>%pull(CpG)#change this to 1

smooth_betas<-function(df, x, y){
  x<-enquo(x); y<-enquo(y)
  ggplot(df, aes(x=!!x, y=!!y))+
    geom_point()+geom_smooth()+
    stat_cor()+theme_classic()+
    scale_y_continuous(name=paste0(quo_name(y)))+
    scale_x_continuous(name=paste0(quo_name(x)))
}

a<-smooth_betas(beta_child, child_model1, child_model2)
b<-smooth_betas(beta_child, child_model1, child_model3)
c<-smooth_betas(beta_child, child_model2, child_model3)

d<-smooth_betas(beta_child%>%filter(CpG %in% sig_beta_m1), child_model1, child_model2)
e<-smooth_betas(beta_child%>%filter(CpG %in% sig_beta_m1), child_model1, child_model3)
f<-smooth_betas(beta_child%>%filter(CpG %in% sig_beta_m1), child_model2, child_model3)

all_beta<-plot_grid(a, b, c, ncol=1)
sig_beta<-plot_grid(d, e, f, ncol=1)

```

```{r child beta vs beta draw graph, figure.height=16, fig.cap='Correlation between effect estimates from the three different epigenome wide linear models in children for A) all CpGs and B) significant CpGs only'}
plot_grid(all_beta, sig_beta, ncol=2, labels=c('A', 'B'), hjust = -2)

```

## Table of top hits in children models

### Model 1

```{r top hits ewas table child1}
child.m1.ss.hits.annotated%>% dplyr::rename(Beta.Coefficient=logFC, Average.Methylation=AveExpr)%>% head(n=10)%>%kable()%>%kable_classic()%>%scroll_box(width='100%', height='400px')
```

### Model 2

```{r top hits ewas table child2}
child.m2.ss.hits.annotated%>% dplyr::rename(Beta.Coefficient=logFC, Average.Methylation=AveExpr)%>%head(n=10)%>%kable()%>%kable_classic()%>%scroll_box(width='100%', height='400px')
```

### Model 3

```{r top hits ewas table child 3}
child.m3.ss.hits.annotated%>% dplyr::rename(Beta.Coefficient=logFC, Average.Methylation=AveExpr)%>%head(n=10)%>%kable()%>%kable_classic()%>%scroll_box(width='100%', height='400px')
```

```{r remove beta_c}
rm(beta_c)
```

# Cross sectional epigenome wide association study: Teen Age 15

```{r beta filtering and checks teens, eval=F}
#load in betaqc matrix
#betaqc<-readRDS(file=paste0(data.dir, 'OGData/', "noob_filtered.rds"))
#filter betaqc to child samples only 
beta_t <-betaqc[, colnames(betaqc) %in% teen_nosmoke$MethID]#; rm(betaqc)
#make sure that the order of the samples matches in the metadata matrix and the beta matrix are ordered the same
teen_nosmoke<-teen_nosmoke[match(colnames(beta_t), teen_nosmoke$MethID), ]
#checks
if(dim(beta_t)[2]!=dim(teen_nosmoke)[1]){print('STOP: Sample #s do not match')}
if(identical(colnames(beta_t), teen_nosmoke$MethID)==FALSE){print('STOP: Sample names do not match')}
```

```{r set models2, eval=F}
mod1=model.matrix(~factor(teen_nosmoke$smkPreg_binary)+teen_nosmoke$global_PC1+teen_nosmoke$global_PC2+factor(teen_nosmoke$cm1bsex)+teen_nosmoke$cm1inpov+teen_nosmoke$ChildAgeComposite+teen_nosmoke$Epi+teen_nosmoke$IC)

mod2=model.matrix(~factor(teen_nosmoke$smkPreg_binary)+teen_nosmoke$global_PC1+teen_nosmoke$global_PC2+factor(teen_nosmoke$cm1bsex)+teen_nosmoke$cm1inpov+teen_nosmoke$ChildAgeComposite+teen_nosmoke$Epi+teen_nosmoke$IC+factor(teen_nosmoke$m1g2_YesNoPreg)+factor(teen_nosmoke$m1g3_YesNoPreg))

mod3=model.matrix(~factor(teen_nosmoke$smkPreg_binary)+teen_nosmoke$global_PC1+teen_nosmoke$global_PC2+factor(teen_nosmoke$cm1bsex)+teen_nosmoke$cm1inpov+teen_nosmoke$ChildAgeComposite+teen_nosmoke$Epi+teen_nosmoke$IC+factor(teen_nosmoke$m1g2_YesNoPreg)+factor(teen_nosmoke$m1g3_YesNoPreg)+factor(teen_nosmoke$PostnatalMaternalSmokingAny)+factor(teen_nosmoke$SmkAtVisitPastmonth))

modellist=list(mod1, mod2, mod3)
modelnames=c('limmaModel1Teen.Rda', 'limmaModel2Teen.Rda', 'limmaModel3Teen.Rda')
tophitnames=c('TeenModel1TopHits.Rda', 'TeenModel2TopHits.Rda', 'TeenModel3TopHits.Rda')
```

```{r loop through models, eval=F}
for(i in 1:length(modellist)){
  print(paste('Running Model', i))
  out=lmFit(beta_t, modellist[[i]])
  out=eBayes(out)
  saveRDS(out, file=file.path(limma.out.dir, modelnames[i]))
  teen.ss.hits<-topTable(out, coef=2, number=nrow(beta_t), confint=T)
  rm(out)
  print(paste('Annotating top hits from model', i))
  teen.ss.annotated.hits<-add_annotations(teen.ss.hits)
  saveRDS(teen.ss.annotated.hits, file=file.path(limma.out.dir, tophitnames[i]))
  rm(teen.ss.hits, teen.ss.annotated.hits)
}
```

```{r load top hits in teen models}
teen.m1.ss.hits.annotated <-readRDS(file=file.path(limma.out.dir, 'TeenModel1TopHits.Rda'))
teen.m2.ss.hits.annotated <-readRDS(file=file.path(limma.out.dir, 'TeenModel2TopHits.Rda'))
teen.m3.ss.hits.annotated <-readRDS(file=file.path(limma.out.dir, 'TeenModel3TopHits.Rda'))
```

## QQ plots

```{r calc lambda teens}
l_t1<-calcLambda(teen.m1.ss.hits.annotated)
l_t2<-calcLambda(teen.m2.ss.hits.annotated)
l_t3<-calcLambda(teen.m3.ss.hits.annotated)

```

```{r QQPlots teens, fig.height=16}
par(mfrow=c(3, 1), mar=c(5.1, 4.1, 5.1, 2.1))
qq(teen.m1.ss.hits.annotated$P.Value, main='Base model: prenatal smoking, genetic PCs, sex \n age, income to poverty ratio, Epi, Fib')
mtext(paste0(expression(lambda), '=', round(l_t1, 3)), side=3)
qq(teen.m2.ss.hits.annotated$P.Value, main='Prenatal exposures model: \n Base model vars+ prenatal alcohol+prenatal other drug')
mtext(paste0(expression(lambda), '=', round(l_t2, 3)), side=3)
qq(teen.m3.ss.hits.annotated$P.Value, main='Secondhand smoking model: \n Prenatal exposure model + postnatal maternal smoking any at 1 or 5 + \n postnatal maternal smoking pks/day in month prior to visit')
mtext(paste0(expression(lambda), '=', round(l_t3, 3)), side=3)
```

## Beta vs Beta plots within Age 15 models

```{r beta vs beta within teen}
teen_models<-list('teen_model1'=teen.m1.ss.hits.annotated, 'teen_model2'=teen.m2.ss.hits.annotated, 'teen_model3'=teen.m3.ss.hits.annotated)%>%lapply(., function(x) rownames_to_column(x, var='CpG'))%>%Map(cbind, ., ModelType=names(.))

beta_teen<-lapply(teen_models, function(x) x %>%dplyr::select(CpG, logFC, ModelType)%>%pivot_wider(id_cols = CpG, values_from=logFC, names_from=ModelType))%>%purrr::reduce(left_join, by='CpG')

sig_beta_m1_t<-teen_models[[1]]%>%filter(P.Value<10e-4)%>%pull(CpG)

a<-smooth_betas(beta_teen, teen_model1, teen_model2)
b<-smooth_betas(beta_teen, teen_model1, teen_model3)
c<-smooth_betas(beta_teen, teen_model2, teen_model3)

d<-smooth_betas(beta_teen%>%filter(CpG %in% sig_beta_m1_t), teen_model1, teen_model2)
e<-smooth_betas(beta_teen%>%filter(CpG %in% sig_beta_m1_t), teen_model1, teen_model3)
f<-smooth_betas(beta_teen%>%filter(CpG %in% sig_beta_m1_t), teen_model2, teen_model3)

all_beta<-plot_grid(a, b, c, ncol=1)
sig_beta<-plot_grid(d, e, f, ncol=1)

```

```{r teen beta vs beta draw graph, figure.height=16, fig.cap='Correlation between effect estimates from the three different epigenome wide linear models in teens for A) all CpGs and B) significant CpGs only'}
plot_grid(all_beta, sig_beta, ncol=2, labels=c('A', 'B'), vjust = 0.8)
```

## Table of top hits in teen EWAS

### Model 1

```{r top hits ewas table teen1}
teen.m1.ss.hits.annotated%>% dplyr::rename(Beta.Coefficient=logFC, Average.Methylation=AveExpr)%>%head(n=10)%>%kable()%>%kable_classic()%>%scroll_box(width='100%', height='400px')
```

### Model 2

```{r top hits ewas table teen2}
teen.m2.ss.hits.annotated%>% dplyr::rename(Beta.Coefficient=logFC, Average.Methylation=AveExpr)%>%head(n=10)%>%kable()%>%kable_classic()%>%scroll_box(width='100%', height='400px')
```

### Model 3

```{r top hits ewas table teen3}
teen.m3.ss.hits.annotated%>% dplyr::rename(Beta.Coefficient=logFC, Average.Methylation=AveExpr)%>%head(n=10)%>%kable()%>%kable_classic()%>%scroll_box(width='100%', height='400px')
```

# Volcano plots for child and teen cross-sectional EWAS

```{r volcano}
volc_child<-ggplot(child.m1.ss.hits.annotated%>%mutate(adjPval=ifelse(adj.P.Val<0.05, '<0.05', ifelse(adj.P.Val<0.2, '<0.2', '>0.2'))), aes(x=logFC, y=-log10(P.Value), color=adjPval))+geom_point()+scale_x_continuous('Beta (regression coefficient)')#+geom_hline(yintercept=-log10(0.05/dim(child.m2.ss.hits.annotated)[1]))
volc_teen<-ggplot(teen.m1.ss.hits.annotated%>%mutate(adjPval=ifelse(adj.P.Val<0.05, '<0.05', ifelse(adj.P.Val<0.2, '<0.2', '>0.2'))), aes(x=logFC, y=-log10(P.Value), color=adjPval))+geom_point()+scale_x_continuous('Beta (regression coefficient)')#+geom_hline(yintercept=-log10(0.05/dim(child.m2.ss.hits.annotated)[1]))
```

```{r save volcano}
png(file.path(outputdir, 'Volcano.png'), width=1300, height=800)
plot_grid(volc_child+theme_classic()+theme(text=element_text(size=16)), volc_teen+theme_classic()+theme(text=element_text(size=16)), ncol=2, labels=c('A Child', 'B Teen'), vjust = 0.8, label_size = 18)
dev.off()
```

# Compare top hits in child and teen EWAS, and to other studies

```{r compare}
AllModels<-c(child_models, teen_models)

AllModelsdf<-AllModels %>% bind_rows(.id='ModelType')%>%mutate('Model'=gsub('.*_', '', ModelType), 'Age'=gsub('_.*', '', ModelType))

AllModelsdf=AllModelsdf%>%group_by(Model, Age, ModelType)%>%nest()%>%mutate(TopCpGs=map(data, function(x) x %>%filter(P.Value<1e-3)%>%pull(CpG)))%>%mutate(ModelType=case_when(ModelType=='child_model1'~'Age 9: Model 1', ModelType=='child_model2'~'Age 9: Model 2', ModelType=='child_model3'~'Age 9: Model 3', ModelType=='teen_model1'~'Age 15: Model 1', ModelType=='teen_model2'~'Age 15: Model 2', ModelType=='teen_model3'~'Age 15: Model 3'))%>%mutate(ModelType=factor(ModelType, levels=c('Age 9: Model 1', 'Age 9: Model 2', 'Age 9: Model 3', 'Age 15: Model 1', 'Age 15: Model 2', 'Age 15: Model 3')))
```

## Upset plot of top ten CpGs in each model:age

```{r upset plot, fig.width=9}
a<-AllModelsdf %>% unnest(TopCpGs)%>%ungroup()%>%dplyr::select(ModelType, TopCpGs)%>%group_by(TopCpGs)%>%summarise(TypeList=list(ModelType))%>%ggplot(aes(x=TypeList))+geom_bar()+theme_classic2()+scale_x_upset(sets=c('Age 9: Model 1', 'Age 9: Model 2', 'Age 9: Model 3', 'Age 15: Model 1', 'Age 15: Model 2', 'Age 15: Model 3'), name='')+scale_y_continuous(name='')
b<-AllModelsdf %>%mutate(nCpGs=lengths(TopCpGs))%>%ggplot(aes(x=ModelType, y=-nCpGs))+geom_bar(position=position_dodge(), stat='identity')+coord_flip()+scale_y_continuous('',breaks=seq(0, -420, -50), labels=seq(0, 420, 50))+scale_x_discrete(limits=rev(levels(AllModelsdf$ModelType)))+theme_classic2()+theme(axis.text.y = element_blank(), axis.title.y = element_blank(), axis.ticks.y=element_blank(), axis.line.y = element_blank())
plot_grid(plot_grid(NULL, b + theme(plot.margin = unit(c(1, -5, -5, 1), "pt")), ncol = 1, rel_heights = c(2.15, 1)),a, nrow = 1, rel_widths = c(1, 3))
```

## Beta estimates comparison of top hits from child and teen models

```{r betas over time}
beta_all<-lapply(AllModels, function(x) x %>%dplyr::select(CpG, logFC, ModelType)%>%pivot_wider(id_cols = CpG, values_from=logFC, names_from=ModelType))%>%purrr::reduce(left_join, by='CpG')

sig_beta_m1<-child_models[[1]]%>%filter(P.Value<10e-4)%>%pull(CpG)

a<-smooth_betas(beta_all, child_model1, teen_model1)
b<-smooth_betas(beta_all, child_model2, teen_model2)
c<-smooth_betas(beta_all, child_model3, teen_model3)

d<-smooth_betas(beta_all%>%filter(CpG %in% sig_beta_m1), child_model1, teen_model1)
e<-smooth_betas(beta_all%>%filter(CpG %in% sig_beta_m1), child_model2, teen_model2)
f<-smooth_betas(beta_all%>%filter(CpG %in% sig_beta_m1), child_model2, teen_model3)

all_beta<-plot_grid(a, b, c, ncol=1)
sig_beta<-plot_grid(d, e, f, ncol=1)
all_beta_pres<-plot_grid(a, b, c, ncol=3)
sig_beta_pres<-plot_grid(d, e, f, ncol=3)

```

```{r teen beta vs child beta draw graph, figure.height=16, , fig.cap='Correlation between effect estimates from the three different epigenome wide linear models in children vs in teens for A) all CpGs and B) significant CpGs only'}
plot_grid(all_beta, sig_beta, ncol=2, labels=c('A', 'B'), vjust = 0.8)
```
```{r save beta comparison}
png(file.path(outputdir, 'BetaChildvsTeen.png'), width=1300, height=800)
plot_grid(all_beta_pres+theme(text=element_text(size=16)), sig_beta_pres+theme(text=element_text(size=16)), ncol=1, labels=c('A', 'B'), vjust = 0.8)
dev.off()
```

## Similar top hits as Wicklund et al. 2019 and Joubert et al. 2016

```{r boxplots}
sigCpGs<-AllModelsdf%>%filter(Model=='model1')%>%mutate(data=map(data, function(x) x %>%filter(adj.P.Val<0.2)), SigCpG=map(data, function(x) x%>%pull(CpG)))%>%unnest(SigCpG)%>%pull(SigCpG) %>%unique()

betaqc<-readRDS(file=paste0(data.dir, 'CreatedData/', 'betaTrim.Rds'))
mybetaqc<-betaqc[sigCpGs%>%unique(), ]
#allgaps<-gaphunter(betaqc) #terminates r studio session (memory) 
mygaps<-gaphunter(mybetaqc)

myBeta<-as.data.frame(t(mybetaqc))
myBeta$MethID =row.names(myBeta)
modeldata2<-left_join(modeldata%>%dplyr::select(-contains('cg')), myBeta)

boxplotcpg<-modeldata2 %>% pivot_longer(cols=contains('cg'), names_to='CpG', values_to='Methylation')
notgap_samplecpg<-boxplotcpg%>%filter(!(CpG %in% row.names(mygaps$proberesults)))%>%pull(CpG)%>%unique()%>%head(n=4)

gap_plots<- ggplot(boxplotcpg%>%filter(CpG %in% row.names(mygaps$proberesults)), aes(x=childteen, y=Methylation, color=smkPreg_binary))+geom_jitter(position=position_dodge2(width=0.9), alpha=0.5)+geom_boxplot(alpha=0.8)+facet_wrap(~CpG)+theme(text=element_text(size=16))+scale_color_discrete(name='Prenatal smoke exposure?')+scale_x_discrete('', labels=c('Age 9', 'Age 15'))+theme_classic()+theme(legend.position='bottom')
not_gaps_plots<-ggplot(boxplotcpg%>%filter(!(CpG %in% row.names(mygaps$proberesults))), aes(x=childteen, y=Methylation, color=smkPreg_binary))+geom_boxplot()+facet_wrap(~CpG)+stat_compare_means(label='p.signif', label.y=0.9)+theme(text=element_text(size=16))+scale_color_discrete(name='Prenatal smoke exposure?')+theme(legend.position='bottom')+scale_x_discrete('', labels=c('Age 9', 'Age 15'))
not_gaps_sample<-ggplot(boxplotcpg%>%filter(CpG %in% notgap_samplecpg), aes(x=childteen, y=Methylation, color=smkPreg_binary))+geom_jitter(position=position_dodge2(width=0.9), alpha=0.5)+geom_boxplot(alpha=0.8)+facet_wrap(~CpG)+stat_compare_means(label='p.signif', label.y=0.9)+theme(text=element_text(size=16))+scale_color_discrete(name='Prenatal smoke exposure?')+scale_x_discrete('', labels=c('Age 9', 'Age 15'))+theme_classic()+theme(legend.position='bottom')
```

```{r save gap boxplots}
gap_plots

png(file.path(outputdir, 'GapPlots.png'), width=800, height=800)
gap_plots+theme(text=element_text(size=16))
dev.off()

not_gaps_plots

png(file.path(outputdir, 'NotGapPlotsSample.png'), width=800, height=800)
not_gaps_sample+theme(text=element_text(size=16))
dev.off()
```

```{r}
#read in top CpGs from Table One of Wicklund et al 2019
topCPGs<-read.csv(paste0(data.dir, "TopCpGsWiklundEtAl2019.csv"))
topCPGs<-topCPGs %>% mutate(estimate=as.numeric(gsub("−\\s", "-", gsub(" .*", "", β..SE.))), std.error=as.numeric(sub("\\).*", "", sub(".*\\(", "", β..SE.))), p.value=P.value, source="Wiklund et al. 2019", conf.low=(estimate-1.96*std.error), conf.high=(estimate+1.96*std.error), set ='')%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)

#read in Joubert meta top cpgs 
#read in cpgs from PMC4833289
cpgs<-read.csv(paste0(data.dir, 'PMC4833289_SuppFile2CpGs.csv'), stringsAsFactors = F, header=F, skip=2)
#make headers
header1<-scan(paste0(data.dir, 'PMC4833289_SuppFile2CpGs.csv'), nlines=1, what=character(), sep=',')
header1[7:26]<-c(rep("SSnewborn", 5), rep("SSnewbornCT", 5), rep("SSolder", 5), rep("anynewborn", 5))
header2<-paste(header1, scan(paste0(data.dir, 'PMC4833289_SuppFile2CpGs.csv'), nlines=1, skip=1, what=character(), sep=','), sep='_')
names(cpgs)<-header2

#four meta-EWASes: SS_newborn (sustained smoking in newborns); SS_newbornCT (sustained smoking in newborns controlling for cell type) SS_older (sustained smoking in older children); any_newborn (any smoking in newborns)
#pivot_longer to get single column of coefs and a column for the analysis type
#and group into a list by analysis type
cpgs<-cpgs %>% pivot_longer(cols = c(matches("_Coef|_SE|_P|_FDR|_Direction")), names_to = c("set", ".value"), names_pattern="(.+)_(.+)")%>%rename_with(~gsub('_', '', .x))%>%mutate(Gene=`Mapped Gene`, estimate=Coef, conf.low=estimate-1.96*SE, conf.high=estimate+1.96*SE, source='Joubert et al. 2016', std.error=SE, p.value=P, Chr=CHR) %>%mutate(source=paste0(source, ': ', set))%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)%>%filter(set %in% c('SSolder', 'SSnewbornCT'))

myCPGs<-AllModelsdf %>%filter(Model=='model1')%>%mutate(data=map(data, function(x) x %>%filter(CpG %in% sigCpGs)))%>%unnest(cols=c(data))%>%mutate(Joubertetal=ifelse(CpG %in% cpgs$CpG, 'in Joubert et al. 2016', 'Not in Joubert et al. 2016'), Wicklundetal=ifelse(CpG %in% topCPGs$CpG, 'in Wicklund et al. 2019', 'Not in Wicklund et al. 2019'))%>%mutate(Gene=gene, estimate=logFC, conf.low=CI.L, conf.high=CI.R, source=paste0('Fragile Families:', Age), std.error='NA', p.value=P.Value, Chr=chr, set='Model1', Position=pos)

#myCPGs<-child.m1.ss.hits.annotated %>% filter(adj.P.Val<0.2)%>%mutate(CpG=row.names(.))%>%mutate(Joubertetal=ifelse(CpG %in% cpgs$CpG, 'in Joubert et al. 2016', 'Not in Joubert et al. 2016'), Wicklundetal=ifelse(CpG %in% topCPGs$CpG, 'in Wicklund et al. 2019', 'Not in Wicklund et al. 2019'))%>%mutate(Gene=gene, estimate=logFC, conf.low=CI.L, conf.high=CI.R, source='Fragile Familes Age 9', std.error='NA', p.value=P.Value, Chr=chr, set='Model1', Position=pos)

cpg_overlap=myCPGs%>%filter(Joubertetal=='in Joubert et al. 2016' | Wicklundetal=='in Wicklund et al. 2019')%>%pull(CpG)%>%unique()

myCPGs2<-myCPGs %>%ungroup()%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)%>%mutate(CpG=ifelse(CpG %in% row.names(mygaps$proberesults), paste0(CpG, '*'), CpG))

cpg_all<-rbind(myCPGs2, cpgs, topCPGs)%>% filter(CpG%in% cpg_overlap)
cpgs_new<-myCPGs2%>%filter(!(CpG %in% cpg_overlap))

#geom blank to give good axes limits
blank_data <- cpg_all %>% filter(CpG%in% cpg_overlap)%>%dplyr::select(Gene, CpG, estimate, std.error)%>%group_by(Gene)%>%mutate(min=min(estimate), max=max(as.numeric(std.error), na.rm=T))%>%dplyr::select(-estimate, -std.error)%>%unique()%>%mutate(estimate=min*-1+max, conf.high=estimate, conf.low=estimate)%>%dplyr::select(-max, -min)%>%arrange(Gene)%>%mutate(source='Wiklund et al. 2019')

blank_data2 <- cpgs_new %>%dplyr::select(Gene, CpG, estimate, std.error)%>%group_by(Gene)%>%mutate(min=min(estimate))%>%dplyr::select(-estimate, -std.error)%>%unique()%>%mutate(estimate=min*-1+0.005, conf.high=estimate, conf.low=estimate)%>%dplyr::select(-min)%>%arrange(Gene)%>%mutate(source='Fragile Families:child')

plot_sharedCpG<-ggplot(cpg_all, aes(x=estimate, xmin=conf.low, xmax=conf.high, color=source, y=CpG))+geom_point(position=position_dodge(width=1))+geom_errorbarh(height=0.1, position=position_dodge(width=1))+scale_x_continuous()+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+facet_wrap(.~Gene, scales='free', ncol=2)+theme_classic()+theme(legend.position = 'bottom')+guides(color=guide_legend(nrow=2,byrow=TRUE))+geom_blank(data = blank_data, aes(x = estimate))

plot_newCpG<-ggplot(cpgs_new, aes(x=estimate, xmin=conf.low, xmax=conf.high, color=source, y=CpG))+geom_point(position=position_dodge(width=1))+geom_errorbarh(height=0.1, position=position_dodge(width=1))+scale_x_continuous()+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+facet_wrap(.~Gene, scales='free')+theme_classic()+theme(legend.position = 'bottom')+guides(color=guide_legend(nrow=2,byrow=TRUE))+geom_blank(data = blank_data2, aes(x = estimate))

cut_cpgs<-c(paste0(row.names(mygaps$proberesults), '*'), 'cg14055351')
plot_newCpG_cut<-ggplot(cpgs_new%>%filter(!(CpG%in%cut_cpgs)), aes(x=estimate, xmin=conf.low, xmax=conf.high, color=source, y=CpG))+geom_point(position=position_dodge(width=1))+geom_errorbarh(height=0.1, position=position_dodge(width=1))+scale_x_continuous()+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+facet_wrap(.~Gene, scales='free')+theme_classic()+theme(legend.position = 'bottom')+guides(color=guide_legend(nrow=2,byrow=TRUE))+geom_blank(data = blank_data2%>%filter(!(CpG %in%cut_cpgs)), aes(x = estimate))
```

Filtering to only CpGs  that had an adjusted p-value of <0.2 in either the base child & teen models ($m_{CpG_{p_{BH}<0.2}}=$ `r length(sigCpGs)`), and then to only CpGs that overlapped with results from either Wicklund et al. 2019 or Joubert et al. 2016 ($m_{CpG_{overlap}}=$ `r length(cpg_overlap)` ), plot the estimates with error bars and compare.

```{r, fig.width=10, fig.height=10}
plot_sharedCpG
png(file.path(outputdir, 'SharedCpG.png'), width=1000, height=800)
plot_sharedCpG+theme(text=element_text(size=18))
dev.off()
```

Also several CpGs that had an adjusted p-value <0.2 but were not overlapping with results from either Wicklund et al. 2019 or Joubert et al. 2016 ($m_{CpG_{notOverlap}}=$ `r cpgs_new$CpG %>%unique()%>%length()`). Some of these CpGs annotate to interesting immune (Il1A, CFB)  and neurology (ADNP, CLN8, GDNF, KIRREL2, RGNEF, SHANK2) genes.

```{r, fig.width=10, fig.height=10}
plot_newCpG
png(file.path(outputdir, 'NewCpG.png'), width=1300, height=1000)
plot_newCpG+theme(text=element_text(size=18))
dev.off()
```
```{r, fig.width=10, fig.height=10}
plot_newCpG
png(file.path(outputdir, 'NewCpGCut.png'), width=1300, height=1000)
plot_newCpG_cut+theme(text=element_text(size=18))
dev.off()
```

# Longitudinal EWAS

```{r beta filtering and checks longitudinal, eval=F}
#load in betaqc matrix (trimmed)
#betaqc<-readRDS(file=paste0(data.dir, 'CreatedData/', 'betaTrim.Rds'))
#filter betaqc to modeldata samples only
beta_all <-betaqc[, colnames(betaqc) %in% modeldata$MethID]#; rm(betaqc)
#make sure that the order of the samples matches in the metadata matrix and the beta matrix are ordered the same
modeldata<-modeldata[match(colnames(beta_all), modeldata$MethID), ]
#checks
if(dim(beta_all)[2]!=dim(modeldata)[1]){print('STOP: Sample #s do not match')}
if(identical(colnames(beta_all), modeldata$MethID)==FALSE){print('STOP: Sample names do not match')}
```

```{r}

model_lmm<-function(df, y, x){
  xnew=paste0(x, '+', '(1|id)')
  lmer(formula(paste(y, xnew)), data=df, REML=F)
}

model_lmm<-function(y){
suppressWarnings(lmer(y~factor(smkPreg_binary)+global_PC1+global_PC2+factor(cm1bsex)+cm1inpov+ChildAgeComposite+Epi+IC+factor(smkPreg_binary)*ChildAgeComposite+(1|id), data=modeldata, REML=F))
}

data.frame(test_beta)%>%map(model_lmm)%>%map(tidy)%>%bind_rows(.id = 'cpg')

rm(betaqc)
bt=data.frame(t(beta_all)); rm(beta_all)
bt_test=bt[, 1:1000]

future::plan(multisession)

start.time=Sys.time()
mixed_models=bt_test%>%future_map(model_lmm, .progress=TRUE)%>%map(tidy)%>%bind_rows(.id = 'cpg')
end.time=Sys.time()
end.time-start.time

start.time=Sys.time()
mixed_models=bt_test%>%map(model_lmm)%>%map(tidy)%>%bind_rows(.id = 'cpg')
end.time=Sys.time()
end.time-start.time

```

```{r set models, eval=F}
mod1=model.matrix(~factor(modeldata$smkPreg_binary)+modeldata$global_PC1+modeldata$global_PC2+factor(modeldata$cm1bsex)+modeldata$cm1inpov+modeldata$ChildAgeComposite+modeldata$Epi+modeldata$IC+factor(modeldata$smkPreg_binary)*modeldata$ChildAgeComposite)

mod2=model.matrix(~factor(modeldata$smkPreg_binary)+modeldata$global_PC1+modeldata$global_PC2+factor(modeldata$cm1bsex)+modeldata$cm1inpov+modeldata$ChildAgeComposite+modeldata$Epi+modeldata$IC+factor(modeldata$m1g2_YesNoPreg)+factor(modeldata$m1g3_YesNoPreg)+factor(modeldata$smkPreg_binary)*modeldata$ChildAgeComposite)

mod3=model.matrix(~factor(modeldata$smkPreg_binary)+modeldata$global_PC1+modeldata$global_PC2+factor(modeldata$cm1bsex)+modeldata$cm1inpov+modeldata$ChildAgeComposite+modeldata$Epi+modeldata$IC+factor(modeldata$m1g2_YesNoPreg)+factor(modeldata$m1g3_YesNoPreg)+factor(modeldata$PostnatalMaternalSmokingAny)+factor(modeldata$SmkAtVisitPastmonth)+factor(modeldata$smkPreg_binary)*modeldata$ChildAgeComposite)

modellist=list(mod1, mod2, mod3)
modelnames=c('limmaModel1Long.Rda', 'limmaModel2Long.Rda', 'limmaModel3Long.Rda')
tophitnames_smk=c('LongModel1TopHits_smk.Rda', 'LongTeenModel2TopHits_smk.Rda', 'LongModel3TopHits_smk.Rda')
tophitnames_age=c('LongModel1TopHits_age.Rda', 'LongTeenModel2TopHits_age.Rda', 'LongModel3TopHits_age.Rda')
tophitnames_int=c('LongModel1TopHits_int.Rda', 'LongTeenModel2TopHits_int.Rda', 'LongModel3TopHits_int.Rda')
tophitnames=cbind(tophitnames_smk, tophitnames_age, tophitnames_int)
coef=c(2, 7, 10)
```

```{r loop through models longitudinal, eval=F}
#concern: For this function (duplicateCorrelation) to return statistically useful results, there must be at least two more arrays than the number of coefficients to be estimated, i.e., two more than the column rank of design.(from: https://rdrr.io/bioc/limma/man/dupcor.html)... because the design matrix is pretty large (many coefficients) is this a problem for us? 
#this is also taking forever, I think I should run the duplicateCorrelation on a subset of rows (see: https://support.bioconductor.org/p/117817/), see if it's normally distributed and then just use the median for everyone 
  print(paste('Fitting correlation structure for Model', i))
  corfit<-duplicateCorrelation(beta_all, design=modellist[[i]], block=modeldata$id)

for(i in 1:length(modellist)){
  print(paste('Fitting correlation structure for Model', i))
  corfit<-duplicateCorrelation(beta_all, design=modellist[[i]], block=modeldata$id)
  print(paste('Running Model', i))
  out=lmFit(beta_all, modellist[[i]], block=modeldata$id, correlation=corfit$consensus)
  out=eBayes(out)
  saveRDS(out, file=file.path(limma.out.dir, modelnames[i]))
  for(j in 1:3){
    print(paste('Annotating top hits from model', i, 'coef', coef[j]))
    long.ss.hits<-topTable(out, coef=coef[j], number=nrow(beta_all), confint=T)
    long.ss.annotated.hits<-add_annotations(long.ss.hits)
    saveRDS(long.ss.annotated.hits, file=file.path(limma.out.dir, tophitnames[i, j]))
    rm(long.ss.hits, long.ss.annotated.hits)
  }
  rm(out)
}
```

### 2021-09-07 Reports/FFModeling.Rmd
first commit on nfs

---
title: "Fragile Families - Epigenome summary metric modeling"
author: "Freida Blostein"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    code_folding: show
    highlight: tango
    number_sections: yes
    theme: sandstone
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/nfs/turbo/bakulski1/People/blostein/FF_methylation/")
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
library(tidyr)
library(stringr)
library(dplyr)
library(compareGroups)
library(kableExtra)
library(haven)
library(sjlabelled)
library(matchmaker)
library(EpiDISH)
library("PerformanceAnalytics")
library(DescTools)
library(purrr)
library(broom)
library(broom.mixed)
library(ggplot2)
library(multcomp)
library(ggpubr)
library(cowplot)
library(geepack)
library(lme4)
library(lmerTest)
library(stringr)
library(sva)
data.dir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/CreatedData"
outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output"
```

```{r Read in data, results='hide'}
file.path(data.dir, list.files(data.dir))[(str_detect(file.path(data.dir, list.files(data.dir)), 'NoSmk'))]%>%map(load, envir=.GlobalEnv)
```

```{r make model data}
modeldata<-rbind(child_nosmoke, teen_nosmoke) # dataset with the most exclusion i.e. squaring dataset off for all models (excluding kids who smoke in figure)
set_label(modeldata)<-get_label(child_nosmoke)
```

```{r factors in model data}
modeldata$PostnatalMaternalSmokingAny<-factor(modeldata$PostnatalMaternalSmokingAny, levels=c("No maternal smoking at age 1 and age 5", "Maternal smoking at age 1 or age 5"))
modeldata$SmkAtVisitPastmonth<-factor(modeldata$SmkAtVisitPastmonth, levels=c("No smoking", "Less than pack a day", "Pack or more a day", "Missing"))
modeldata$slide<-factor(modeldata$slide)
modeldata$Sample_Plate<-factor(modeldata$Sample_Plate)
```

# Bivariate analyses

## Methylation summary metrics, confounders and precision variables by prenatal maternal smoking, stratified by visit 
```{r bivariate table summary methylation data}
export2md(strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+pediatric+grim+anynewborn_center+SSnewbornCT_center+SSolder_center+Epi+Fib+IC+ChildAgeComposite+cm1inpov+ancestry+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth, data=modeldata)), 'childteen'))
```

## Methylation summary metrics, confounders and precision variables by prenatal maternal smoking, ancestry stratified at 9 year visit

```{r bivariate table stratified by visit and ancestry child visit}
export2md(strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+pediatric+levine+anynewborn_center+SSnewbornCT_center+SSolder_center+Epi+Fib+IC+ChildAgeComposite+cm1inpov+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth, data=modeldata%>%filter(childteen=='C'))), 'ancestry'))
```
## Methylation summary metrics, confounders and precision variables by prenatal maternal smoking, ancestry stratified at 15 year visit

```{r bivariate table stratified by visit and ancestry teen visit}
export2md(strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+pediatric+grim+anynewborn_center+SSnewbornCT_center+SSolder_center+Epi+Fib+IC+ChildAgeComposite+cm1inpov+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth, data=modeldata%>%filter(childteen=='T'))), 'ancestry'))
```

# Parallel cross-sectional multivariate models of summary methylation variables

```{r Define variables for models crossectional}
y_vector<-c("globalmethylation", "pediatric", "grim", "anynewborn_center", "SSnewbornCT_center", "SSolder_center")
###########################
baseModelvars_local<-"~smkPreg_binary+local_PC1+local_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+Sample_Plate"
baseModelvars_global<-"~smkPreg_binary+global_PC1+global_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+Sample_Plate"
###########################
prenatalModelvars<-paste0(baseModelvars_global, '+m1g2_YesNoPreg+m1g3_YesNoPreg')
smokeModelvars<-paste0(prenatalModelvars, '+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth')
predictors<-c('basemodel'=baseModelvars_local, 'prenatalex'=prenatalModelvars, 'secondhand'=smokeModelvars) 
##############################
data<-list('childteen'=c("T", "C"), 'outcome'=y_vector)
args<-data %>%cross_df()
data2<-list('ancestry'=c('African ancestry', 'Hispanic ancestry', 'European ancestry'), 'outcome'=y_vector)
args2<-data2 %>%cross_df()
```

```{r Define model functions}
model<-function(df, y, x){
  lm(formula(paste(y, x)), data=df)
}

model_gee<-function(df, y, x, corstr){
  geeglm(formula(paste(y, x)), data=df, family=gaussian(link='identity'), id=id, corstr = corstr)
}

model_lmm<-function(df, y, x){
  xnew=paste0(x, '+', '(1|id)')
  lmer(formula(paste(y, xnew)), data=df, REML=F)
}
```

## Global models (non-stratified by ancestry)

Three types of models:

* Base model variables: Yes/No smoking during pregnancy, 1st and 2nd principal components of ancestry from a PCA run in all the children, child sex, maternal income to poverty ratio at birth, child age at interview, epithelial cells, immune cells

* Prenatal exposures model variables: Base model variables + maternal alcohol use during pregnancy (yes/no), maternal other drug use during pregnancy (yes/no)

* Secondhand smoking exposures model variables: Prenatal exposure variables + any postnatal maternal smoking at ages 1 or 5 (yes/no)+maternal smoking packs/day in month prior to maternal interview

With six different epigenetic summary measurements as outcomes:

* Global methylation

* Levine clock

* Pediatric clock

* Polymethylation scores (mean centered), including: 
  * Constructed using 'any smoking - newborn cordblood' coefficients
  * Constructed using 'sustained smoking - newborn cordblood + cell-type controlled' coefficients
  * Constructed using 'sustained smoking - older children' coefficients

```{r loop global, results='hide'}
#listOfData<-list("Base model"=basemodeldata, "Prenatal exposures model"=prenatalexdata, "Secondhand smoking exposure model"=secondhandsmkdata, "Firsthand smoking exposure model"=rbind(child_nosmoke, teen_nosmoke))
listOfData<-list("Base model"=modeldata, "Prenatal exposures model"=modeldata, "Secondhand smoking exposure model"=modeldata)
###################
prenatalModelvars<-paste0(baseModelvars_global, '+m1g2_YesNoPreg+m1g3_YesNoPreg')
smokeModelvars<-paste0(prenatalModelvars, '+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth')
#predictors<-c('basemodel'=baseModelvars, 'prenatalex'=prenatalModelvars, 'secondhand'=smokeModelvars, 'firsthand'=smokeModelvars)
predictors<-c('basemodel'=baseModelvars_global, 'prenatalex'=prenatalModelvars, 'secondhand'=smokeModelvars)

listOfNests<-list()
for(i in 1:length(predictors)){
  listOfNests[[i]]<-listOfData[[i]]%>%group_by(childteen)%>%nest()
  listOfNests[[i]]<-left_join(args, listOfNests[[i]], by='childteen')
  listOfNests[[i]]<-listOfNests[[i]] %>% mutate(model=map2(data, outcome, model, x=predictors[i]), 
                                                tidied = map(model, tidy, conf.int=T), glanced = map(model, glance), 
                                                augmented=map(model, augment))%>%
                                        unnest(tidied)%>%unnest(glanced, names_sep='_')
}
names(listOfNests)<-names(listOfData)
global_models<-listOfNests %>%bind_rows(.id='modeltype')%>%dplyr::select(-data, -model, -augmented)%>%mutate(outcome=factor(outcome, levels=c("globalmethylation", "pediatric", "grim", "anynewborn_center", "SSnewbornCT_center", "SSolder_center"), labels=c('Global methylation', 'Pediatric clock', 'GRIM clock',  'Any smoking polymethylation score (newborns)', 'Sustained smoking polymethylation score (newborns, cell-type controlled', 'Sutained smoking polymethylation score (older children)')))
```

```{r global models with slide sensitivity}
###########################
baseModelvars_global<-"~smkPreg_binary+global_PC1+global_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+slide"
###########################
prenatalModelvars<-paste0(baseModelvars_global, '+m1g2_YesNoPreg+m1g3_YesNoPreg')
smokeModelvars<-paste0(prenatalModelvars, '+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth')
predictors<-c('basemodel'=baseModelvars_global, 'prenatalex'=prenatalModelvars, 'secondhand'=smokeModelvars) 
###########################
listOfNests<-list()
for(i in 1:length(predictors)){
  listOfNests[[i]]<-listOfData[[i]]%>%group_by(childteen)%>%nest()
  listOfNests[[i]]<-left_join(args, listOfNests[[i]], by='childteen')
  listOfNests[[i]]<-listOfNests[[i]] %>% mutate(model=map2(data, outcome, model, x=predictors[i]), 
                                                tidied = map(model, tidy, conf.int=T), glanced = map(model, glance), 
                                                augmented=map(model, augment))%>%
                                        unnest(tidied)%>%unnest(glanced, names_sep='_')
}
names(listOfNests)<-names(listOfData)
global_models_withslides<-listOfNests %>%bind_rows(.id='modeltype')%>%dplyr::select(-data, -model, -augmented)%>%mutate(outcome=factor(outcome, levels=c("globalmethylation", "pediatric", "grim", "anynewborn_center", "SSnewbornCT_center", "SSolder_center"), labels=c('Global methylation', 'Pediatric clock', 'GRIM clock',  'Any smoking polymethylation score (newborns)', 'Sustained smoking polymethylation score (newborns, cell-type controlled', 'Sutained smoking polymethylation score (older children)')))
```

### Global model fit table

```{r global table of model fit information}
global.mf<-global_models %>% filter(term=='(Intercept)') %>% dplyr::select(outcome,modeltype, childteen,glanced_nobs, glanced_adj.r.squared, glanced_AIC, glanced_logLik, glanced_p.value)%>%arrange(outcome)%>%mutate(glanced_p.value=as.character(signif(glanced_p.value, 2))) %>%pivot_wider(id_cols = c("outcome", "modeltype"), names_from = childteen, names_glue="{childteen}_{.value}", values_from = c(glanced_nobs, glanced_adj.r.squared, glanced_AIC, glanced_logLik, glanced_p.value))%>%dplyr::select(outcome, modeltype, order(colnames(.)))

global.mf<-global.mf %>% setNames(sub("^[^_]*_([^_]*)*_([^_]*).*", "\\2", colnames(global.mf)))

kable(global.mf, digits=2)%>%kable_classic()%>%add_header_above(c(" "=2, "Age 9 models"=5, "Age 15 models"=5))%>%pack_rows("Global methylation model", 1, 3)%>%pack_rows("Epigenetic clocks", 4, 9)%>%pack_rows("Polymethylation scores", 10, 18)%>%kable_styling(font_size=10)%>%scroll_box(width='100%', height='400px')
```

### Forest plot for any prenatal maternal smoking
```{r global models forest plot, fig.width=11, fig.height=10}
main_predictor_global<-global_models %>% filter(term=='smkPreg_binaryYes')%>%mutate(modeltype=factor(modeltype, levels=c("Base model", "Prenatal exposures model", "Secondhand smoking exposure model")))

visit.labs=c('Age 9', 'Age 15'); names(visit.labs)=c('C', 'T')

g1<-ggplot(main_predictor_global%>%filter(outcome %in% levels(main_predictor_global$outcome)[1:3]), aes(x=estimate, xmin=conf.low, xmax=conf.high, color=childteen, y=modeltype))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_y_discrete(name='Model', limits=rev(levels(main_predictor_global$modeltype)))+facet_grid(childteen~outcome, scales='free_x', labeller = labeller(outcome=label_wrap_gen(20), childteen=visit.labs))+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+theme(legend.position = 'bottom')+theme(text = element_text(size=16))+scale_color_discrete('', labels=c('Age 9', 'Age 15'))+theme_classic()+scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')

g2<-ggplot(main_predictor_global%>%filter(outcome %in% levels(main_predictor_global$outcome)[4:6]), aes(x=estimate, xmin=conf.low, xmax=conf.high, color=childteen, y=modeltype))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_y_discrete(name='Model', limits=rev(levels(main_predictor_global$modeltype)))+facet_grid(childteen~outcome, labeller = labeller(outcome=label_wrap_gen(20), childteen=visit.labs))+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+theme(legend.position = 'bottom')+theme(text = element_text(size=16))+scale_color_discrete('', labels=c('Age 9', 'Age 15'))+theme_classic()+scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')

plot_grid(g1, g2, nrow=2, labels = c("A", "B"))

example_coef<-main_predictor_global%>%filter(outcome %in% levels(main_predictor_global$outcome)[4] & modeltype=='Secondhand smoking exposure model' & childteen=='C')
```

```{r save forest global}

png(file.path(outputdir, 'GlobalForest.png'), width=1300, height=550)
plot_grid(g1+theme(text=element_text(size=16)), g2+theme(text=element_text(size=16)), nrow=2, labels = c("A", "B"))
dev.off()

png(file.path(outputdir, 'GlobalForestA.png'), width=1300, height=550)
g1+theme(text=element_text(size=16))
dev.off()

png(file.path(outputdir, 'GlobalForestB.png'), width=1300, height=550)
g2+theme(text=element_text(size=16))
dev.off()

```
### Check sample plate terms

```{r check sample plate coefficients}
#check how many models have at least one plate associated  
(n_samplePlateAs=global_models %>% filter(str_detect(term, 'Sample_Plate') & p.value<0.05)%>%dplyr::select(modeltype, childteen, outcome)%>%unique()%>%nrow())
# is this the same as the number of models run?
n_samplePlateAs==global_models %>% dplyr::select(modeltype, childteen, outcome)%>%unique()%>%nrow()

#Are certain plates responsible or is it widespread?
#in how many models is each plate associated 
global_models %>% filter(str_detect(term, 'Sample_Plate') & p.value<0.05)%>%dplyr::select(term)%>%group_by(term)%>%summarise(n=n())%>%arrange(-n)
# how many outcomes is each plate associated with? 
global_models %>% filter(str_detect(term, 'Sample_Plate') & p.value<0.05)%>%dplyr::select(term, outcome)%>%unique()%>%group_by(term)%>%summarise(n=n())%>%arrange(-n)
#for each outcome, how many plates are associated? 
global_models %>% filter(str_detect(term, 'Sample_Plate') & p.value<0.05)%>%dplyr::select(term, outcome, childteen, modeltype)%>%group_by(outcome, childteen, modeltype)%>%summarise(n=n(), plates=toString(term))%>%arrange(-n)%>%kable()%>%scroll_box(width='100%', height='400px')
```

Every single model has at least 1 plate that is significantly associated with the outcome methylation metric. Models with global methylation have the most plates associated with them. Sample plate 4, 12, 15 and 3 have the most associations with methylation outcomes, this does not exactly match what Jonah reports from his email (PC2 & PC6 correlate with batch, PC2 --> Plate 9 & 10, PC6--> Plate 12).

### Interpertation

Prenatal maternal smoking has a DNA methylation signature that is identifiable in saliva samples from children aged 9 and aged 15. This signature is specific, that is, there is no effect of prenatal maternal smoking on either global methylation or DNA methylation clocks estimated from CpG site methylation data from saliva samples of children aged 9 or 15. However,there is an effect of prenatal maternal smoking on polymethylation scores constructed using weights from a consortium-based meta-analysis of prenatal maternal smoking and blood methylation.

For example, among children aged 9 years old, children whose mother smoked during pregnancy had an 'Any smoking polymethylation score' that was on average `r example_coef%>%pull(estimate)%>%round(2)` (95% CI: (`r example_coef%>%pull(conf.low)%>%round(2)`, `r example_coef%>%pull(conf.high)%>%round(2)`))

Another thing to note is that the confidence intervals for the effect of prenatal maternal smoking overlap in the 9-year and 15-year visit: that is, it appears that the effect of prenatal maternal smoking is similar at both ages.

### IGES abstract

```{r code for IGES abstract}
# # in modeldata
n_tot<-modeldata$id%>%unique()%>%length()
# # smoke exposed
modeldata %>% dplyr::select(id, smkPreg_binary)%>%unique()%>%group_by(smkPreg_binary)%>%summarise(n=n())
modeldata %>% dplyr::select(id, ancestry)%>%unique()%>%group_by(ancestry)%>%summarise(n=n(), percent=n/n_tot)
#correlation between polymethylation score (newborn, ct controlled)
modeldata %>% dplyr::select(id, SSnewbornCT_center, childteen)%>%pivot_wider(id_cols=id, names_from=childteen, values_from=SSnewbornCT_center)%>%ggplot(aes(x=C, y=T))+geom_point()+stat_cor()

df<-modeldata %>% dplyr::select(id, SSnewbornCT_center, childteen)%>%pivot_wider(id_cols=id, names_from=childteen, values_from=SSnewbornCT_center)

#abstract model (global, CT-controlled newborn, w/ plate variable)
age9Coef<-main_predictor_global%>%filter(outcome %in% levels(main_predictor_global$outcome)[5] & modeltype=='Secondhand smoking exposure model' & childteen=='C')
print(paste0(round(age9Coef$estimate, 2), ', (', round(age9Coef$conf.low, 2), ', ', round(age9Coef$conf.high, 2), ')'))

age15Coef<-main_predictor_global%>%filter(outcome %in% levels(main_predictor_global$outcome)[5] & modeltype=='Secondhand smoking exposure model' & childteen=='T')
print(paste0(round(age15Coef$estimate, 2), ', (', round(age15Coef$conf.low, 2), ', ', round(age15Coef$conf.high, 2), ')'))

```

### Global models with slide as a variable 
Just checking if controlling for slide number influences models
Took out 05/05/21 because added plate variable to main models (check with kelly) 
```{r global models forest plot with slide as variable, fig.width=11, fig.height=10, eval=F}
main_predictor_global<-global_models_withslides %>% filter(term=='smkPreg_binaryYes')%>%mutate(modeltype=factor(modeltype, levels=c("Base model", "Prenatal exposures model", "Secondhand smoking exposure model")))

visit.labs=c('Age 9', 'Age 15'); names(visit.labs)=c('C', 'T')

g1<-ggplot(main_predictor_global%>%filter(outcome %in% levels(main_predictor_global$outcome)[1:3]), aes(x=estimate, xmin=conf.low, xmax=conf.high, color=childteen, y=modeltype))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_y_discrete(name='Model', limits=rev(levels(main_predictor_global$modeltype)))+facet_grid(childteen~outcome, scales='free_x', labeller = labeller(outcome=label_wrap_gen(20), childteen=visit.labs))+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+theme(legend.position = 'bottom')+theme(text = element_text(size=16))+scale_color_discrete('', labels=c('Age 9', 'Age 15'))+theme_classic()+scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')

g2<-ggplot(main_predictor_global%>%filter(outcome %in% levels(main_predictor_global$outcome)[4:6]), aes(x=estimate, xmin=conf.low, xmax=conf.high, color=childteen, y=modeltype))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_y_discrete(name='Model', limits=rev(levels(main_predictor_global$modeltype)))+facet_grid(childteen~outcome, labeller = labeller(outcome=label_wrap_gen(20), childteen=visit.labs))+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+theme(legend.position = 'bottom')+theme(text = element_text(size=16))+scale_color_discrete('', labels=c('Age 9', 'Age 15'))+theme_classic()+scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')

plot_grid(g1, g2, nrow=2, labels = c("A", "B"))
```

## Ancestry-specific models

The same three models and siix outcomes, EXCEPT that in each model type we replace the global PCs of ancestry with PCs of ancestry from PCA run within each ancestry strata, i.e. local PCs:

* Base model variables: Yes/No smoking during pregnancy, 1st and 2nd principal components of ancestry from a PCA run **only on children of the same ancestry**, child sex, maternal income to poverty ratio at birth, child age at interview, epithelial cells, immune cells

```{r loop ancestry stratified crosssectional, results='hide'}
#listOfData<-list("Base model"=basemodeldata, "Prenatal exposures model"=prenatalexdata, "Secondhand smoking exposure model"=secondhandsmkdata, "Firsthand smoking exposure model"=rbind(child_nosmoke, teen_nosmoke))
listOfData<-list("Base model"=modeldata, "Prenatal exposures model"=modeldata, "Secondhand smoking exposure model"=modeldata)
listOfNests<-list()
#######################
prenatalModelvars<-paste0(baseModelvars_local, '+m1g2_YesNoPreg+m1g3_YesNoPreg')
smokeModelvars<-paste0(prenatalModelvars, '+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth')
predictors=predictors<-c('basemodel'=baseModelvars_local, 'prenatalex'=prenatalModelvars, 'secondhand'=smokeModelvars)

####################################
for(i in 1:length(predictors)){
  listOfNests[[i]]<-listOfData[[i]]%>%group_by(childteen, ancestry)%>%nest()
  listOfNests[[i]]<-left_join(args, listOfNests[[i]], by='childteen')
  listOfNests[[i]]<-listOfNests[[i]] %>% mutate(model=map2(data, outcome, model, x=predictors[i]), 
                                                tidied = map(model, tidy, conf.int=T), glanced = map(model, glance), 
                                                augmented=map(model, augment))%>%
                                        unnest(tidied)%>%unnest(glanced, names_sep='_')
}
names(listOfNests)<-names(listOfData)
ancestry_models<-listOfNests %>%bind_rows(.id='modeltype')%>%dplyr::select(-data, -model, -augmented)%>%mutate(outcome=factor(outcome, levels=c("globalmethylation",  "pediatric", "grim", "anynewborn_center", "SSnewbornCT_center", "SSolder_center"), labels=c('Global methylation', 'Pediatric clock', 'GRIM clock','Any smoking polymethylation score (newborns)', 'Sustained smoking polymethylation score (newborns, cell-type controlled', 'Sustained smoking polymethylation score (older children)')))
```

### Ancestry-specific model fit tables
```{r model fit tables ancestry specific}
ancestry.mf<-ancestry_models %>% filter(term=='(Intercept)') %>% dplyr::select(outcome, ancestry, modeltype, childteen,glanced_nobs, glanced_adj.r.squared, glanced_AIC, glanced_logLik, glanced_p.value)%>%arrange(ancestry,outcome)%>%mutate(glanced_p.value=as.character(signif(glanced_p.value, 2))) %>%pivot_wider(id_cols = c("outcome", "ancestry", "modeltype"), names_from = childteen, names_glue="{childteen}_{.value}", values_from = c(glanced_nobs, glanced_adj.r.squared, glanced_AIC, glanced_logLik, glanced_p.value))%>%dplyr::select(ancestry, outcome, modeltype, order(colnames(.)))

african.mf<-ancestry.mf %>%filter(ancestry=="African ancestry")%>% setNames(sub("^[^_]*_([^_]*)*_([^_]*).*", "\\2", colnames(ancestry.mf)))
european.mf<-ancestry.mf %>%filter(ancestry=="European ancestry")%>% setNames(sub("^[^_]*_([^_]*)*_([^_]*).*", "\\2", colnames(ancestry.mf)))
hispanic.mf<-ancestry.mf %>%filter(ancestry=="Hispanic ancestry")%>% setNames(sub("^[^_]*_([^_]*)*_([^_]*).*", "\\2", colnames(ancestry.mf)))
```

African ancestry model fit
```{r african ancestry}
kable(african.mf, digits=2)%>%kable_classic()%>%add_header_above(c(" "=3, "Age 9 models"=5, "Age 15 models"=5))%>%kable_styling(font_size=10)%>%pack_rows("Global methylation model", 1, 3)%>%pack_rows("Epigenetic clocks", 4, 9)%>%pack_rows("Polymethylation scores", 10, 18)%>%scroll_box(width='100%', height='400px')
```

European ancestry model fit
```{r eurpoean ancestry}
kable(european.mf, digits=2)%>%kable_classic()%>%add_header_above(c(" "=3, "Age 9 models"=5, "Age 15 models"=5))%>%kable_styling(font_size=10)%>%pack_rows("Global methylation model", 1, 3)%>%pack_rows("Epigenetic clocks", 4, 9)%>%pack_rows("Polymethylation scores", 10, 18)%>%scroll_box(width='100%', height='400px')
```

Hispanic ancestry model fit
```{r hispanic ancestry}
kable(hispanic.mf, digits=2)%>%kable_classic()%>%add_header_above(c(" "=3, "Age 9 models"=5, "Age 15 models"=5))%>%kable_styling(font_size=10)%>%pack_rows("Global methylation model", 1, 3)%>%pack_rows("Epigenetic clocks", 4, 9)%>%pack_rows("Polymethylation scores", 10, 18)%>%scroll_box(width='100%', height='400px')
```

### Ancestry specific forest plots for prenatal maternal smoking 
```{r ancestry specific forestplots, fig.width=11, fig.height=10}
main_predictor<-ancestry_models%>%filter(term=="smkPreg_binaryYes")%>%mutate(modeltype=factor(modeltype, levels=c("Base model", "Prenatal exposures model", "Secondhand smoking exposure model")))%>%mutate(R2=paste0('adjR^2==', round(glanced_adj.r.squared,2)))

ggplot(main_predictor %>% filter(outcome %in% c("Pediatric clock", "Levine clock")), aes(x=estimate, xmin=conf.low, xmax=conf.high, color=childteen, y=modeltype))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_x_continuous()+scale_y_discrete(name='Model', limits=rev(levels(main_predictor$modeltype)))+facet_grid(outcome~ancestry, labeller = labeller(outcome=label_wrap_gen(20), childteen=visit.labs))+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+geom_text(aes(x=conf.high+1,label=paste0("n=", glanced_nobs)), position=position_dodge(width = 1))+theme(text = element_text(size=16))+theme(legend.position = 'bottom')+theme(text = element_text(size=16))+scale_color_discrete('', labels=c('Age 9', 'Age 15'))+theme_classic()+scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')

ggplot(main_predictor %>% filter(outcome %in% c("Global methylation")), aes(x=estimate, xmin=conf.low, xmax=conf.high, color=childteen, y=modeltype))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_x_continuous()+scale_y_discrete(name='Model', limits=rev(levels(main_predictor$modeltype)))+facet_grid(outcome~ancestry, labeller = labeller(outcome=label_wrap_gen(20), childteen=visit.labs))+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+geom_text(aes(x=conf.high+0.005,label=paste0("n=", glanced_nobs)), position=position_dodge(width = 1))+xlim(-0.015, 0.015)+theme(legend.position = 'bottom')+theme(text = element_text(size=16))+scale_color_discrete('', labels=c('Age 9', 'Age 15'))+theme_classic()+scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')

ggplot(main_predictor %>% filter(outcome %in% levels(main_predictor$outcome)[4:6]), aes(x=estimate, xmin=conf.low, xmax=conf.high, color=childteen, y=modeltype))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_x_continuous()+scale_y_discrete(name='Model', limits=rev(levels(main_predictor$modeltype)))+facet_grid(outcome~ancestry, labeller = labeller(outcome=label_wrap_gen(20), childteen=visit.labs))+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+geom_text(aes(x=conf.high+0.05,label=paste0("n=", glanced_nobs)), position=position_dodge(width = 1))+theme(legend.position = 'bottom')+theme(text = element_text(size=16))+scale_color_discrete('', labels=c('Age 9', 'Age 15'))+theme_classic()+scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')

```

```{r save forest ancestry}

png(file.path(outputdir, 'AncestryForest.png'), width=1300, height=700)
ggplot(main_predictor %>% filter(outcome %in% levels(main_predictor$outcome)[4:6]), aes(x=estimate, xmin=conf.low, xmax=conf.high, color=childteen, y=modeltype))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_x_continuous()+scale_y_discrete(name='Model', limits=rev(levels(main_predictor$modeltype)))+facet_grid(ancestry~outcome, labeller = labeller(outcome=label_wrap_gen(20), childteen=visit.labs))+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+geom_text(aes(x=conf.high+0.05,label=paste0("n=", glanced_nobs)), position=position_dodge(width = 1))+theme(legend.position = 'bottom')+theme(text = element_text(size=16))+scale_color_discrete('', labels=c('Age 9', 'Age 15'))+theme_classic()+scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+theme(text=element_text(size=16))
dev.off()

```

Very similar effect sizes across ancestry groups, although error bars cross 0 for the Hispanic and European strata models (recall, the number of women who smoked during pregnancy in these strata is very small).

Possibly something different going on with the European teens in the secondhand smoking exposure model. Interestingly, the European mom's have the heaviest distribution of postnatal smoking, with a larger percentage reporting over a pack a day.

# Longitudinal multivariate models on summary methylation variables

## Background & theory  
So far we have been ignoring the longitudinal nature of our data, and treating it as two cross-sectional samples: however, most children had methylation data measured at 2 time points, age 9 and 15.  In our data we have the following structure, where $i$ is visit such that $i \in {Age9, Age15}$ and $j$ is child

* $Y_{ij}$ is one of several epigenome summary metrics, which are all measured for each $j$ child at each $i$ visit

* $X_j$ is the binary yes/no maternal prenatal smoking variable, which was measured once for each $j$ child

* $A_j ... F_j$ are time-invariant confounders measured once for each $j$ child, such as:

  + Ancestry: First 2 local or global PC measurements from genetic data 
  + Child sex
  + Maternal income to poverty ratio at birth 
  + Maternal alcohol or other drug use during pregnancy

* $G_{ij} ... N_{ij}$ are time-variant confounders measured at each visit $i$ for each $j$ child, such as:

  + Child age at visit $i$
  + Epithelial/immune cell in sample from visit $i$
  + Yes/no postnatal maternal smoking at at maternal interview corresponding to visit $i$
  + Postnatal maternal smoking in packs/day in month prior to maternal interview corresponding to visit $i$

There are two (common) ways to model longitduinal data like this:

* LMM are conditional models, that is, estimates are subject-specific and allow for variation in subjects (which are modeled through random intercepts/slopes)-->What is the individual specific effect of the covariate?

The LMM base model for our data would be something like:

$Y_{ij} = \beta_0 + \beta_1 PrenatalSmoking_j+\beta_2 PC1_j +\beta_3 PC2_j +$
$\beta_4 ChildSex_j+\beta_5MaternalIncome_j+\beta_6ChildAge_{ij}+\beta_7EpiCells_{ij}+$
$\beta_8ImmuneCells_{ij} +\beta_9PrenatalSmoking_j*ChildAge_{ij}+b_{0i}+\epsilon_{ij}$

Where $i$ indexes visits within subject (child) $j$ and $j$ is the individual subject (child), the $\beta$'s are the fixed effects, $b_{0j}$ is the random deviation in intercept for subject $j$, and the errors $\epsilon_{ij}$ are conditional errors and represent the deviation from the fixed effects plus random effects at time $i$ for subject $j$ (conditional errors are independent, and correlation is a result of the random effects in the model). The interaction between Age and Prenatal Maternal smoking allows us to investigate if the rate of change in the epigenome summary variable (outcome) differs for those exposed vs unexposed to prenatal maternal smoking.

* GEE are marginal models, that is, estimates are population averages and you are modeling the expectation of your outcome conditional only upon the fixed design matrix --> What is the population averaged effect of the covariate?

The GEE base model for our data would be something like:

$Y_{ij} = \beta_0 + \beta_1 PrenatalSmoking_j+\beta_2 PC1_j +\beta_3 PC2_j +$
$\beta_4 ChildSex_j+\beta_5MaternalIncome_j+\beta_6ChildAge_{ij}+\beta_7EpiCells_{ij}+$
$\beta_8ImmuneCells_{ij}+\beta_9PrenatalSmoking_j*ChildAge_{ij}+\epsilon_{ij}$

Again, $i$ indexes visits within subject (child) $j$ and $j$ is the individual subject (child), and the $\beta$'s are the fixed effects. However, the marginal model does not have any random effects (intercepts or slopes), except the random errors. The errors represent the deviation from the fixed effects for each subject at each time point and these errors are not conditional on random effects and are instead are marginal errors, which we assume are correlated within a subject $j$.

Unlike LMM, GEE does not make distributional assumptions for the observations and the random effects. Additionally the robust sandwich type standard errors from a GEE will be valid asymptotic condifence intervals even if the correlation structure specified is not correct.

LMM may be preferred over GEE when: 
  + Subjects measured at many different arbitrary time points --> not true for FF
  + Interested in modeling growth trajectories for different subjects --> true for FF

GEE will work best when: 
  + Number of observation per subeject is small and number of subjects large --> true for FF
  + Measurements taken at the same time for all subjects --> true for FF

A GEE model with compound symmetry/exchangeable covariance structure will be the same as an LMM with a random intercept when using the identity link (when in a linear model framework) - since the MOST time points we have for any child is 2, every type of co-variance structure is actually the same and they are therefore all symmetric.

## Bivariate analyses: Spaghetti plots

```{r mah spahghetti global, figure.width=10}
p<-ggplot(modeldata, aes(x=ChildAgeComposite, y=globalmethylation, group=id, color=smkPreg_binary))+geom_line(alpha=0.05)+geom_smooth(aes(group=smkPreg_binary), method='lm')+scale_x_continuous('Child age (months)')+scale_y_continuous('Score')+ggtitle('Global methylation')+scale_color_discrete('Prenatal maternal smoking')+theme_classic()
plot_grid(p, p+facet_wrap(~ancestry), ncol=1)
```

```{r mah spahghetti pediatric, figure.width=10}
p<-ggplot(modeldata, aes(x=ChildAgeComposite, y=pediatric, group=id, color=smkPreg_binary))+geom_line(alpha=0.05)+geom_smooth(aes(group=smkPreg_binary), method='lm')+scale_x_continuous('Child age (months)')+scale_y_continuous('Score')+ggtitle('Pediatric clock')+scale_color_discrete('Prenatal maternal smoking')+theme_classic()
plot_grid(p, p+facet_wrap(~ancestry), ncol=1)
```

```{r mah spahghetti SSnewbornCT score, figure.width=10}

p<-ggplot(modeldata, aes(x=ChildAgeComposite, y=SSnewbornCT_center, group=id, color=smkPreg_binary))+geom_line(alpha=0.05)+geom_smooth(aes(group=smkPreg_binary), method='lm')+scale_x_continuous('Child age (months)')+scale_y_continuous('Score')+ggtitle('Sustained smoking \n newborn, cell-type controlled')+scale_color_discrete('Prenatal maternal smoking')+theme_classic()
plot_grid(p, p+facet_wrap(~ancestry), ncol=1)
```

After looking at the spaghetti plots, I think it would also be reasonable to be interested in a random slope for age/time, i.e. $b_{1j}ChildAge_{ij}$ where $b_{1j}$ is the random deviation in slope (of $Y_j$ over $i$). However, we don't have enough observations to support this, when I include a random intercept and a random slope, I get an error message that the number of observations <= number of random effects (see: https://stats.stackexchange.com/questions/193678/number-of-random-effects-is-not-correct-in-lmer-model)

## Global longitudinal models (LMM)

```{r Define variables for models longitudinal}
y_vector<-c("globalmethylation", "pediatric", "grim", "anynewborn_center", "SSnewbornCT_center", "SSolder_center")
###########################
baseModelvars_local<-"~smkPreg_binary+local_PC1+local_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+Sample_Plate"
baseModelvars_global<-"~smkPreg_binary+global_PC1+global_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+Sample_Plate"
baseModelInt<-paste0(baseModelvars_global,'+smkPreg_binary*ChildAgeComposite')
prenatalModelvars<-paste0(baseModelInt, '+m1g2_YesNoPreg+m1g3_YesNoPreg')
smokeModelvars<-paste0(prenatalModelvars, '+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth')
predictors<-c('basemodel'=baseModelvars_global, 'basemodelInt'=baseModelInt, 'prenatalex'=prenatalModelvars, 'secondhand'=smokeModelvars) 
data2<-list('ancestry'=c('African ancestry', 'Hispanic ancestry', 'European ancestry'), 'outcome'=y_vector)
args2<-data2 %>%cross_df()
```

```{r loop global lmm, warning=F}
listOfModels<-list()
listOfNests<-list()
listOfData<-list("Base model"=modeldata, "Base model with time interaction"= modeldata, "Prenatal exposures model"=modeldata, "Secondhand smoking exposure model"=modeldata)
for(i in 1:length(predictors)){
  listOfModels[[i]]<-listOfData[[i]]%>%droplevels()%>%dplyr::summarise(model=map(y_vector, model_lmm, x=predictors[i], df=.), outcome=y_vector)
  listOfNests[[i]]<-listOfModels[[i]]%>%mutate(tidied=map(model, tidy, conf.int=T), glanced=map(model, glance))%>%unnest(tidied)%>%unnest(glanced, names_sep='_')
}

names(listOfNests)<-names(listOfData)
names(listOfModels)<-names(listOfData)

global_models_lmm<-listOfNests %>%bind_rows(.id='modeltype')%>%dplyr::select(-model)%>%mutate(outcome=factor(outcome, levels=c("globalmethylation",  "pediatric","grim", "anynewborn_center", "SSnewbornCT_center", "SSolder_center"), labels=c('Global methylation',  'Pediatric clock', 'GRIM clock','Any smoking polymethylation score (newborns)', 'Sustained smoking polymethylation score (newborns, cell-type controlled', 'Sutained smoking polymethylation score (older children)')))

global_models_lmm2<-listOfModels %>%bind_rows(.id='modeltype')
global_models_lmm2 <-global_models_lmm2 %>%pivot_wider(id_cols=outcome, names_from=modeltype, values_from=model)%>% group_by(outcome)%>%mutate(aov_int=map2(`Base model`, `Base model with time interaction`, anova), aov_int_tidy=map(aov_int, tidy))
```

### Is there an effect of age?

Our expectation from the bivariate analyses from the previous markdown is that there is a significant effect of age for most methylation summary metrics, but that this effect is very small, except for the clocks, where the effect is larger.

```{r Child Age coefficient}
global_models_lmm %>% filter(term=='ChildAgeComposite')%>%dplyr::select(modeltype, outcome, effect, term, estimate, std.error, statistic,  p.value, conf.low, conf.high)%>%arrange(outcome)%>%kable()%>%kable_classic()%>%kable_styling(font_size=10)%>%pack_rows("Global methylation model", 1, 4)%>%pack_rows("Epigenetic clocks", 5, 12)%>%pack_rows("Polymethylation scores", 13, 24)%>%scroll_box(width='100%', height='400px')
```
Yes, expectation fulfilled

### Is there an effect of prenatal maternal smoking?

Our expectation from our bivariate analyses and our cross sectional models is that there will be an effect of prenatal maternal smoking for the polymethylation score variables but not for global methylation or the clocks

```{r}
global_models_lmm %>% filter(term=='smkPreg_binaryYes')%>%dplyr::select(modeltype, outcome, effect, term, estimate, std.error, statistic,  p.value, conf.low, conf.high)%>%arrange(outcome)%>%kable()%>%kable_classic()%>%kable_styling(font_size=10)%>%pack_rows("Global methylation model", 1, 4)%>%pack_rows("Epigenetic clocks", 5, 12)%>%pack_rows("Polymethylation scores", 13, 24)%>%scroll_box(width='100%', height='400px')
```

Expectation fulfilled

### Does the age-effect vary by prenatal maternal sloping: Is the slope for age different depending on prenatal maternal smoking status?

From our spaghetti plots, it didn't really look like there were different slopes for age depending on prenatal maternal smoking exposure, although the age slope is so small it is difficult to tell. For global methylation in Hispanic ancestry individuals looked the strongest.

```{r }
global_models_lmm %>% filter(term=='smkPreg_binaryYes:ChildAgeComposite')%>%dplyr::select(modeltype, outcome, effect, term, estimate, std.error, statistic,  p.value, conf.low, conf.high)%>%arrange(outcome)%>%kable()%>%kable_classic()%>%kable_styling(font_size=10)%>%pack_rows("Global methylation model", 1, 3)%>%pack_rows("Epigenetic clocks", 4, 9)%>%pack_rows("Polymethylation scores", 10, 18)%>%scroll_box(width='100%', height='400px')
```

No, maybe could make an argument for global methylation model.

<details><summary>Click here</summary>

Another way to see that would be using contrast statements:

```{r contrasts not stratified, eval=F}
K1<-matrix(c(0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0), 1)
#ageSlopeNoSmoke<-glht(model_4_g_int, linfct=K)
K2<-matrix(c(0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 1), 1)
#ageSlopeYesSmoke<-glht(model_4_g_int, linfct=K2)

contrast_ageslope<-global_models_lmm2 %>% mutate(AgeSlopeNoSmoke=map(.x=`Secondhand smoking exposure model`, .f=glht, linfct=K1), AgeSlopeSmoke=map(.x=`Secondhand smoking exposure model`, .f=glht, linfct=K2), AgeSlopeNoSmoke=map(AgeSlopeNoSmoke, confint), AgeSlopeSmoke=map(AgeSlopeSmoke, confint), AgeSlopeNoSmoke=map(AgeSlopeNoSmoke, tidy), AgeSlopeSmoke=map(AgeSlopeSmoke, tidy))%>%unnest(c(AgeSlopeNoSmoke), names_sep='_') %>% unnest(c(AgeSlopeSmoke), names_sep='_')%>%dplyr::select(outcome, AgeSlopeSmoke_estimate, AgeSlopeSmoke_conf.high, AgeSlopeSmoke_conf.low, AgeSlopeNoSmoke_estimate, AgeSlopeNoSmoke_conf.high, AgeSlopeNoSmoke_conf.low)%>%pivot_longer(2:7, names_to=c("type", ".value"), names_pattern="(.*)_(.*)")%>%mutate(outcometype=ifelse(str_detect(outcome, 'center'), 'Polymethylation score', ifelse(outcome=='globalmethylation', 'Global methylation', 'Methylation clock')))

ggplot(contrast_ageslope %>% filter(outcometype=='Global methylation'), aes(y=outcome, x=estimate, xmin=conf.low, xmax=conf.high, color=type))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_x_continuous()+scale_y_discrete()+facet_grid(outcometype~., scales='free')+scale_y_discrete(labels=c('Global methylation'))+scale_x_continuous()+scale_color_discrete(name='Age slope:', labels=c('With no prenatal maternal smoking', 'With prenatal maternal smoking'))+theme_classic()+theme(legend.position='bottom')

```

<details><summary>Click here</summary>

## Ancesty specific longitudinal models (LMM)

```{r loop ancestry stratified lmm, warning=F}
baseModelvars_local<-"~smkPreg_binary+local_PC1+local_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+Sample_Plate"
baseModelvars_global<-"~smkPreg_binary+global_PC1+global_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+Sample_Plate"
baseModelInt<-paste0(baseModelvars_local,'+smkPreg_binary*ChildAgeComposite')
prenatalModelvars<-paste0(baseModelInt, '+m1g2_YesNoPreg+m1g3_YesNoPreg')
smokeModelvars<-paste0(prenatalModelvars, '+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth')
predictors<-c('basemodel'=baseModelvars_global, 'basemodelInt'=baseModelInt, 'prenatalex'=prenatalModelvars, 'secondhand'=smokeModelvars) ##################################
listOfNests<-list()
for(i in 1:length(predictors)){
  listOfNests[[i]]<-listOfData[[i]]%>%droplevels()%>%group_by(ancestry)%>%nest()
  listOfNests[[i]]<-left_join(args2, listOfNests[[i]], by='ancestry')
  listOfNests[[i]]<-listOfNests[[i]] %>% mutate(model=map2(data, outcome, model_lmm, x=predictors[i]), tidied = map(model, tidy, conf.int=T), glanced = map(model, glance))%>%
                                        unnest(tidied)%>%unnest(glanced, names_sep='_')
}
names(listOfNests)<-names(listOfData)
ancestry_models_lmm<-listOfNests %>%bind_rows(.id='modeltype')%>%dplyr::select(-data, -model)%>%mutate(outcome=factor(outcome, levels=c("globalmethylation", "levine", "pediatric", "anynewborn_center", "SSnewbornCT_center", "SSolder_center"), labels=c('Global methylation', 'Levine clock', 'Pediatric clock', 'Any smoking polymethylation score (newborns)', 'Sustained smoking polymethylation score (newborns, cell-type controlled', 'Sustained smoking polymethylation score (older children)')))
ancestry_models_lmm2<-listOfNests %>%bind_rows(.id='modeltype')
```

```{r, eval=F}
ancestry_models_lmm %>% filter(term=='smkPreg_binaryYes:ChildAgeComposite')%>%dplyr::select(modeltype, outcome, effect, term, estimate, std.error, statistic,  p.value, conf.low, conf.high)%>%arrange(outcome)%>%kable()%>%kable_classic()%>%kable_styling(font_size=10)%>%pack_rows("Global methylation model", 1, 3)%>%pack_rows("Epigenetic clocks", 4, 9)%>%pack_rows("Polymethylation scores", 10, 18)%>%scroll_box(width='100%', height='400px')
```
Our expectation is that there will be a significant effect in the African ancestry strata, and that the direction of effect will be fairly consistent in the other ancestry strata but the confidence intervals will be wider, reflecting the small sample size post-stratification.

```{r forest plots lmm ancestry specific, figure.width=10}
main_predictor<-ancestry_models_lmm%>%filter(term=="smkPreg_binaryYes")%>%mutate(modeltype=factor(modeltype, levels=c("Base model", "Base model with time interaction", "Prenatal exposures model", "Secondhand smoking exposure model")))

ggplot(main_predictor %>% filter(outcome %in% c("Pediatric clock", "Levine clock")), aes(x=estimate, xmin=conf.low, xmax=conf.high, y=modeltype))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_x_continuous()+scale_y_discrete(name='Model', limits=rev(levels(main_predictor$modeltype)))+facet_grid(outcome~ancestry, labeller = labeller(outcome=label_wrap_gen(20), childteen=visit.labs))+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+theme(text = element_text(size=16))+theme(legend.position = 'bottom')+theme(text = element_text(size=16))+theme_classic()+scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')

ggplot(main_predictor %>% filter(outcome %in% c("Global methylation")), aes(x=estimate, xmin=conf.low, xmax=conf.high, y=modeltype))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_x_continuous()+scale_y_discrete(name='Model', limits=rev(levels(main_predictor$modeltype)))+facet_grid(outcome~ancestry, labeller = labeller(outcome=label_wrap_gen(20), childteen=visit.labs))+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+xlim(-0.015, 0.015)+theme(legend.position = 'bottom')+theme(text = element_text(size=16))+theme_classic()+scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')

ggplot(main_predictor %>% filter(outcome %in% levels(main_predictor$outcome)[4:6]), aes(x=estimate, xmin=conf.low, xmax=conf.high,  y=modeltype))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_x_continuous()+scale_y_discrete(name='Model', limits=rev(levels(main_predictor$modeltype)))+facet_grid(outcome~ancestry, labeller = labeller(outcome=label_wrap_gen(20), childteen=visit.labs))+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+theme(legend.position = 'bottom')+theme(text = element_text(size=16))+theme_classic()+scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')

```

### Contrasts in age stratified models

Let's take a closer look at the idea of different age slopes by prenatal smoking exposure in the stratified models. One way to look at this is to calculate the $\beta_{Age}$ separately in children exposed vs unexposed to prenatal maternal smoking using contrasts. Since I included an interaction in this model, we can calculate $\beta_{Age|Smoking}=\beta_{Age}+\beta_{Age*Smoke}*1$, and ditto for $\beta_{Age|NotSmoking}=\beta_{Age}+\beta_{Age*Smoke}*0=\beta_{Age}$, and then compare the two.

Age slope by prenatal maternal smoking status:

```{r contrasts age slope stratified, eval=F}
K1<-matrix(c(0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0), 1)
#ageSlopeNoSmoke<-glht(model_4_g_int, linfct=K)
K2<-matrix(c(0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 1), 1)
#ageSlopeYesSmoke<-glht(model_4_g_int, linfct=K2)

contrast_ageslope<-ancestry_models_lmm2 %>% dplyr::filter(modeltype=='Secondhand smoking exposure model' & term=='smkPreg_binaryYes')%>% mutate(AgeSlopeNoSmoke=map(.x=model, .f=glht, linfct=K1), AgeSlopeSmoke=map(.x=model, .f=glht, linfct=K2), AgeSlopeNoSmoke=map(AgeSlopeNoSmoke, confint), AgeSlopeSmoke=map(AgeSlopeSmoke, confint), AgeSlopeNoSmoke=map(AgeSlopeNoSmoke, tidy), AgeSlopeSmoke=map(AgeSlopeSmoke, tidy))%>%unnest(c(AgeSlopeNoSmoke), names_sep='_') %>% unnest(c(AgeSlopeSmoke), names_sep='_')%>%dplyr::select(outcome, ancestry, AgeSlopeSmoke_estimate, AgeSlopeSmoke_conf.high, AgeSlopeSmoke_conf.low, AgeSlopeNoSmoke_estimate, AgeSlopeNoSmoke_conf.high, AgeSlopeNoSmoke_conf.low)%>%pivot_longer(3:8, names_to=c("type", ".value"), names_pattern="(.*)_(.*)")%>%mutate(outcometype=ifelse(str_detect(outcome, 'center'), 'Polymethylation score', ifelse(outcome=='globalmethylation', 'Global methylation', 'Methylation clock')))

#ggplot(contrast_ageslope%>%filter(outcometype=='Global methylation'), aes(y=outcome, x=estimate, xmin=conf.low, xmax=conf.high, color=type))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_x_continuous()+scale_y_discrete()+facet_grid(outcometype~ancestry, scales='free')+scale_y_discrete()+scale_x_continuous()+scale_color_discrete(name='Age slope:', labels=c('With no prenatal maternal smoking', 'With prenatal maternal smoking'))+theme_classic()+theme(legend.position='bottom')+geom_vline(xintercept=0, color='red', linetype='dashed')

ggplot(contrast_ageslope%>%filter(outcometype=='Polymethylation score'), aes(y=outcome, x=estimate, xmin=conf.low, xmax=conf.high, color=type))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_x_continuous()+scale_y_discrete()+facet_grid(outcometype~ancestry, scales='free')+scale_y_discrete(labels=c('Sustained smoking score - older', 'Sustained smoking - newborn +CT', 'Sustained smoking any - newborn'))+scale_x_continuous()+scale_color_discrete(name='Age slope:', labels=c('With no prenatal maternal smoking', 'With prenatal maternal smoking'))+theme_classic()+theme(legend.position='bottom')+geom_vline(xintercept=0, color='red', linetype='dashed')

```
A good check is that our confidence intervals for the age slope are clearly wider for those with prenatal maternal smoking, which makes sense as there are less women who smoked prenatally than did not. Although these estimates are not exactly the same, we can see that the confidence limits overlap so we are insufficiently powered to detect true differences.

```{r contrasts prenatal smoking slope stratified, results=F, eval=F}
K1<-matrix(c(0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 9), 1)
#ageSlopeNoSmoke<-glht(model_4_g_int, linfct=K)
K2<-matrix(c(0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 15), 1)
#ageSlopeYesSmoke<-glht(model_4_g_int, linfct=K2)

contrast_ageslope<-ancestry_models_lmm2 %>% dplyr::filter(modeltype=='Secondhand smoking exposure model' & term=='smkPreg_binaryYes')%>% mutate(AgeSlopeNoSmoke=map(.x=model, .f=glht, linfct=K1), AgeSlopeSmoke=map(.x=model, .f=glht, linfct=K2), AgeSlopeNoSmoke=map(AgeSlopeNoSmoke, confint), AgeSlopeSmoke=map(AgeSlopeSmoke, confint), AgeSlopeNoSmoke=map(AgeSlopeNoSmoke, tidy), AgeSlopeSmoke=map(AgeSlopeSmoke, tidy))%>%unnest(c(AgeSlopeNoSmoke), names_sep='_') %>% unnest(c(AgeSlopeSmoke), names_sep='_')%>%dplyr::select(outcome, ancestry, AgeSlopeSmoke_estimate, AgeSlopeSmoke_conf.high, AgeSlopeSmoke_conf.low, AgeSlopeNoSmoke_estimate, AgeSlopeNoSmoke_conf.high, AgeSlopeNoSmoke_conf.low)%>%pivot_longer(3:8, names_to=c("type", ".value"), names_pattern="(.*)_(.*)")%>%mutate(outcometype=ifelse(str_detect(outcome, 'center'), 'Polymethylation score', ifelse(outcome=='globalmethylation', 'Global methylation', 'Methylation clock')))

ggplot(contrast_ageslope%>%filter(outcometype=='Polymethylation score'), aes(y=outcome, x=estimate, xmin=conf.low, xmax=conf.high, color=type))+geom_point(position = position_dodge(width = 1))+geom_errorbarh(height=0.1, position = position_dodge(width = 1))+scale_x_continuous()+scale_y_discrete()+facet_grid(outcometype~ancestry, scales='free')+scale_y_discrete(labels=c('Sustained smoking score - older', 'Sustained smoking - newborn +CT', 'Sustained smoking any - newborn'))+scale_x_continuous()+scale_color_discrete(name='Age slope:')+theme_classic()+theme(legend.position='bottom')

```

## Walking through LMM & GEE comparison with a single outcome: SSolder_center polymethylation score (with interpretations)

```{r LMM example, eval=F}

#random intercept only models
intercept_only1<-lmer(formula = SSolder_center~1+(1|id), data=modeldata, REML=F)#random intercept for subject
model1<-lmer(formula = SSolder_center~smkPreg_binary+ChildAgeComposite+(1|id), data=modeldata, REML=F)
model2<-lmer(formula = SSolder_center~smkPreg_binary+global_PC1+global_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+(1|id), data=modeldata, REML=F)
anova(model1, model2)
model3<-lmer(formula = SSolder_center~smkPreg_binary+global_PC1+global_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+smkPreg_binary*ChildAgeComposite+(1|id), data=modeldata, REML=F)
anova(model2, model3)
model4<-lmer(formula = SSolder_center~smkPreg_binary+global_PC1+global_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+m1g2_YesNoPreg+m1g3_YesNoPreg+(1|id), data=modeldata, REML=F)
anova(model3, model4)
model5<-lmer(formula = SSolder_center~smkPreg_binary+global_PC1+global_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+(1|id), data=modeldata, REML=F)
anova(model4, model5)
model5_int<-lmer(formula = SSolder_center~smkPreg_binary*ChildAgeComposite+global_PC1+global_PC2+cm1bsex+cm1inpov+Epi+IC+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+(1|id), data=modeldata, REML=F)
anova(model5, model5_int)

#proving to myself the interpretation of age slope is what I think it is 
intercept_only1<-lmer(formula = SSolder_center~1+(1|id), data=modeldata%>%filter(ancestry=='Hispanic ancestry'), REML=F)#random intercept for subject
model1<-lmer(formula = SSolder_center~smkPreg_binary+ChildAgeComposite+(1|id), data=modeldata%>%filter(ancestry=='Hispanic ancestry'), REML=F)
model2<-lmer(formula = SSolder_center~smkPreg_binary*ChildAgeComposite+(1|id), data=modeldata%>%filter(ancestry=='Hispanic ancestry'), REML=F)

#

```

```{r walk through GEE model}
modeldata2<-modeldata[order(modeldata$id),]%>%mutate(id=as.factor(id))%>%droplevels()

model1_g<-geeglm(SSolder_center~ChildAgeComposite+smkPreg_binary, data=modeldata2, family=gaussian(link='identity'), id=id, corstr = 'exchangeable')

model2_g<-geeglm(SSolder_center~ChildAgeComposite+smkPreg_binary, data=modeldata2, family=gaussian(link='identity'), id=id, corstr = 'independence')
#expectation that model 1 and model 2 are identical 
anova(model1_g, model2_g)#expectation fulfilled

model2_g<-geeglm(SSolder_center~ChildAgeComposite+smkPreg_binary+global_PC1+global_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC, data=modeldata2, family=gaussian(link='identity'), id=id, corstr = 'exchangeable')
anova(model1_g, model2_g)

model3_g<-geeglm(SSolder_center~ChildAgeComposite+smkPreg_binary+global_PC1+global_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+m1g2_YesNoPreg+m1g3_YesNoPreg, data=modeldata2, family=gaussian(link='identity'), id=id, corstr = 'exchangeable')
anova(model2_g, model3_g)

model4_g<-geeglm(SSolder_center~ChildAgeComposite+smkPreg_binary+global_PC1+global_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth, data=modeldata2, family=gaussian(link='identity'), id=id, corstr = 'exchangeable')

anova(model3_g, model4_g)

model_4_g_int<-geeglm(SSolder_center~ChildAgeComposite*smkPreg_binary+global_PC1+global_PC2+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth, data=modeldata2, family=gaussian(link='identity'), id=id, corstr = 'exchangeable')

```

```{r show models the same, eval=F}
model_4_g_int%>%tidy()%>%kable()
model5_int%>%tidy()%>%kable()
```

```{r contrasts, eval=F}
K<-matrix(c(0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0), 1)
ageSlopeNoSmoke<-glht(model_4_g_int, linfct=K)
K2<-matrix(c(0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1), 1)
ageSlopeYesSmoke<-glht(model_4_g_int, linfct=K2)

summary(ageSlopeNoSmoke); summary(ageSlopeYesSmoke)
```

```{r contrasts with ancestry, eval=F}
#Spaghetti plot: 
p<-ggplot(modeldata, aes(x=ChildAgeComposite, y=SSolder_center, group=id, color=smkPreg_binary))+geom_line(alpha=0.1)+geom_smooth(aes(group=smkPreg_binary), method='lm')
plot_grid(p, p+facet_wrap(~ancestry), ncol=1)

model_hispanic<-geeglm(SSolder_center~ChildAgeComposite*smkPreg_binary+local_PC1+local_PC2+cm1bsex+cm1inpov+Epi+IC, data=modeldata2%>%filter(ancestry=='Hispanic ancestry'), family=gaussian(link='identity'), id=id, corstr = 'exchangeable')

K<-matrix(c(0, 1, 0, 0, 0, 0, 0, 0, 0, 0), 1)
ageSlopeNoSmoke<-glht(model_hispanic, linfct=K)
K2<-matrix(c(0, 1, 0, 0, 0, 0, 0, 0,  0, 1), 1)
ageSlopeYesSmoke<-glht(model_hispanic, linfct=K2)

summary(ageSlopeNoSmoke); summary(ageSlopeYesSmoke)

f<-data.frame(rbind(confint(ageSlopeNoSmoke)%>%tidy(), confint(ageSlopeYesSmoke)%>%tidy()), 'coef'=c('Age Slope \n No smoking', 'Age Slope \n Yes Smoking'), 'model'=c('cell type controlled', 'cell type controlled'))

model_hispanic_nocellcounts<-geeglm(SSolder_center~ChildAgeComposite*smkPreg_binary+local_PC1+local_PC2+cm1bsex+cm1inpov, data=modeldata2%>%filter(ancestry=='Hispanic ancestry'), family=gaussian(link='identity'), id=id, corstr = 'exchangeable')

K<-matrix(c(0, 1, 0, 0, 0, 0, 0, 0), 1)
ageSlopeNoSmoke<-glht(model_hispanic_nocellcounts, linfct=K)
K2<-matrix(c(0, 1, 0, 0, 0, 0, 0, 1), 1)
ageSlopeYesSmoke<-glht(model_hispanic_nocellcounts, linfct=K2)

f2<-data.frame(rbind(confint(ageSlopeNoSmoke)%>%tidy(), confint(ageSlopeYesSmoke)%>%tidy()), 'coef'=c('Age Slope \n No smoking', 'Age Slope \n Yes Smoking'), 'model'=c('not cell type controlled', 'not cell type controlled'))

f3<-rbind(f, f2)
ageslopes<-ggplot(f3, aes(y=coef, x=estimate, xmin=conf.low, xmax=conf.high))+geom_linerange()+geom_point()+scale_x_continuous()+ggtitle('Hispanic ancestry: \n age slope by smoking & CT control')+facet_wrap(~model)+theme_classic()

ic_slopes<-ggplot(modeldata%>%filter(ancestry=='Hispanic ancestry'), aes(x=ChildAgeComposite, y=IC, group=id, color=smkPreg_binary))+geom_line(alpha=0.1)+geom_smooth(aes(group=smkPreg_binary), method='lm')+facet_wrap(~ancestry)+theme_classic()+scale_color_discrete('Prenatal maternal smoking')+scale_x_continuous('Child age (months)')+theme(legend.position='bottom')
ssolder_slopes<-ggplot(modeldata%>%filter(ancestry=='Hispanic ancestry'), aes(x=ChildAgeComposite, y=SSolder_center, group=id, color=smkPreg_binary))+geom_line(alpha=0.1)+geom_smooth(aes(group=smkPreg_binary), method='lm')+facet_wrap(~ancestry)+theme_classic()+scale_color_discrete('Prenatal maternal smoking')+scale_x_continuous('Child age (months)')+theme(legend.position='bottom')+scale_y_continuous('sustained smoking \n older children \n polymethylation score')

bottom_plot<-plot_grid(ssolder_slopes, ic_slopes,  ncol=2)

```

```{r plot age slopes cell type investigation, figure.width=10, eval=F}
plot_grid(ageslopes, bottom_plot, ncol=1, align='v', axis = 'l')
```

# At this point, move to EWAS markdown if time remains

### 2021-09-07 Reports/ReportonPolyMethylationOptions.Rmd
first commit on nfs

---
title: "Comparison of Polymethylation score options"
author: "Freida Blostein"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    code_folding: show
    highlight: tango
    number_sections: yes
    theme: sandstone
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
datadir<-"/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/"
outputdir<-"/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Output"
codedir<-"/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Code"
load(file.path(datadir, 'CreatedData', 'polymethylationscores.RDS'))
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(PerformanceAnalytics)
```

## Function code
I created a function based on Jonah's code that takes in as an option the transformation type for the methylation beta matrix, and applies it before creating the polymethylation scores. The code is available at https://github.com/blostein/FFSmoking/blob/main/UsefulCode/makePolyEpiScores.R

```{r cache=FALSE, echo=FALSE}
knitr::read_chunk(file.path(codedir, 'UsefulCode', 'makePolyEpiScores.R'))
```

```{r makePolyEpiScores.R, eval=FALSE}

```

A few notes on this code:

1) because the methylation matrix is so large, RStudio kept crashing when I attempted to zscore standardize or center-mean standardize the entire matrix. So, instead, I performed the standardization inside the loop, only upon the CpGs that are used in making each polymethylation scores. Since I think we should be standardizing CpGs, not standardizing individuals, this shouldn't matter.

2) Currently only 2 standardization options but could add more.

## Distributions of poly methylation scores with different transformations on beta methylation matrix

```{r compare PolyMethylation scores distributions}
lapply(polyscores, function(x){summary(x)})
lapply(polyscores, function(x){apply(x, 2, sd)})
lapply(seq_along(polyscores), function(i){polyscores[[i]]%>%pivot_longer(cols=1:4, names_to='Score_Type', values_to='Score')%>%ggplot(aes(x=Score))+geom_histogram()+facet_wrap(~Score_Type)+ggtitle(names(polyscores)[[i]])})
```

As Kelly thought, the zscore standardized have a larger range and the biggest standard deviations, the standard deviations of the no transform and centered methods are the same, which makes sense as $V(aX+b)=a^2V(X)\implies V(\sum\beta_i*[x_i-\bar{x}])=V(\sum\beta_i*x_i-\beta_i*\bar{x})=V(\sum \beta_i*x_i-k)=V(\sum \beta_i*x_i)$.

I think it also makes sense the means of the mean centered and zscores standardized scores are 0.

## Correlation between polymethylation scores of different transforms on the beta methylation matrix

```{r correlation b/t polymethylation scores}
polyscores_flipped=lapply(seq_along(polyscores), function(i){polyscores[[i]]%>%mutate('method'=names(polyscores)[[i]])%>%pivot_longer(cols=1:4, names_to="Score_type", values_to="Score")})%>%bind_rows()%>%pivot_wider(id_cols=c(MethID, Score_type), names_from='method', values_from='Score')%>%group_by(Score_type)%>%group_split()
names(polyscores_flipped)=lapply(polyscores_flipped, function(x){x$Score_type%>%unique()})

lapply(seq_along(polyscores_flipped), function(i){chart.Correlation(polyscores_flipped[[i]][3:5]); mtext(names(polyscores_flipped)[[i]], side=3, line=3)})
```

## Question for Kelly: 
These polymethylation scores calculated on non-COMBAT corrected methylation beta values -  should they also be calculated using COMBAT (or SVA) -corrected beta values?

## Bivariate analysis of polymethylation scores w/ different transformations & binary prenatal smoking variable

```{r bivariate analysis scores vs preg smoking}
load(file.path(datadir, 'CreatedData', 'BaseModelData.Rdata'))
polyscores_flipped<-lapply(polyscores_flipped, function(x) left_join(basemodeldata, x))
lapply(seq_along(polyscores_flipped), function(i) ggplot(polyscores_flipped[[i]]%>%pivot_longer(cols=notransform:center, names_to='transformType', values_to='polymethylationScore'), aes(x=childteen, y=polymethylationScore, color=smkPreg_binary))+geom_violin()+geom_boxplot(position=position_dodge(width=0.9))+facet_wrap(~transformType)+stat_compare_means(label='p.format')+ggtitle(names(polyscores_flipped)[i]))

```

### 2021-09-07 Reports/Sept22FF.Rmd
first commit on nfs

---
title: "September22_FFReport"
author: "F blostein"
date: "9/23/2020"
output: html_document
---
Goals: 
1. Create linked dataset with variables of interest from full pheno data 
2. Create table comparing various analytic subsets (methylation data; two-visit methylation data) to full Fragile Families dataset 
3. Address technical replicates

Questions for Kelly: 
1. Review plan for principal components of ancestry control 
2. Check plan for technical replicates 
3. 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/")
library(tidyr)
library(dplyr)
library(compareGroups)
library(kableExtra)
```

```{r}
pheno<-readRDS("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/fullpheno.rds")
#betaqc<-readRDS("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/betaqc.rds")
pdqc<-readRDS("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pdqc.rds")
```

```{r}
source("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Code/Reports/DataDictionary.R")
```

```{r, fullpheno}
#check column names of pheno for restricted health data which appear
# to begin with suffix ch (i.e. chc10e (tobacco use-ever) and
#chc10p (tobacco use-during preg)
colnames(pheno)[grep('ch', colnames(pheno))] 
#find the id varaibles
colnames(pheno)[grep('id', colnames(pheno))] 
#filter for variables of interest

FF<-pheno %>% select(one_of(idvars), one_of(demovars), one_of(smokingvars), one_of(prenataldrugusevar))
#Warning message:
#Unknown columns: `p3a22`, `p3a23`, `p3a23a`, `p3a23b`, `p3a24`, `p4a22`, `p4a23` 
```

```{r, idgroups}
#What FF ids have methylation data?
methy_ids<-pdqc$id
#technical replicates will be duplicate id + childteen variables from pdqc 
techrep_ids<-pdqc %>%select(idnum, childteen, MethID)%>%group_by(idnum, childteen)%>%filter(n()>1)%>%pull(MethID)
#ids with 9visit, 15visit or both visit
visit9ids<-pdqc%>%filter(childteen=='C')%>%pull(idnum)%>%unique()
visit15ids<-pdqc%>%filter(childteen=='T')%>%pull(idnum)%>%unique()
bothvisit_ids<-visit9ids[visit9ids %in% visit15ids]
visit9only_ids<-visit9ids[!(visit9ids %in%bothvisit_ids)]
visit15only_ids<-visit15ids[!(visit15ids %in%bothvisit_ids)]
```

```{r}
#create flag in FF for having methylation data
#and flag for having 2 visits of methylation data
FF<-FF %>% 
  mutate(MethylData=ifelse(id %in% methy_ids, "Analysis subset", "Not in analysis subset"),
         TwoVisitData=ifelse(id%in%bothvisit_ids,"Analysis subset - both visits", "Not in both visit analysis subset"))
table(FF$MethylData)
table(FF$TwoVisitData)
```

```{r, checks}
if((length(visit15only_ids)+length(visit9only_ids)+length(bothvisit_ids)!=length(FF %>% filter(MethylData=="Analysis subset")%>%pull(id)))){print("WARNING ID LENGTHS DO NOT ADD UP")}
if(length(bothvisit_ids)!=length(FF %>% filter(TwoVisitData=="Analysis subset - both visits")%>%pull(id))){print("WARNING ID LENGTHS DO NOT ADD UP")}
```

```{r, labelMetaData}
myBinary<-myBinary[myBinary %in% colnames(FF)]
myCategorical_O<-myCategorical_O[myCategorical_O %in% colnames(FF)]
myCategorical_U<-myCategorical_U[myCategorical_U%in%colnames(FF)]
FF<-FF %>% mutate_at(vars(myBinary, myCategorical_O, myCategorical_U), funs(as.factor(.)))
```

```{r}
methyl<-createTable(compareGroups(MethylData~.-id-ff_id, data=FF), show.all=T)
twovisit<-createTable(compareGroups(TwoVisitData~.-id-ff_id, data=FF), show.all=T)
my_FF<-FF %>% filter(MethylData=="Analysis subset")%>%mutate(smkPreg_binary=ifelse(m1g4%in% c(1, 2, 3), "Yes", ifelse(m1g4%in%c(-3, -2, -9), NA, ifelse(m1g4==4, "No", "FLAG"))))
xtabs(~m1g4+smkPreg_binary, data=my_FF)
smk<-createTable(compareGroups(smkPreg_binary~.-id-ff_id, data=my_FF))
```

```{r}
export2md(methyl)
export2md(twovisit)
export2md(smk)
```

```{r}
#explore technical replicates
#pdqc %>% filter(MethID %in% techrep_ids) %>% table(~qc_flag_any, data=.)
techrep_qcF<-pdqc %>% filter(MethID %in% techrep_ids & qc_flag_any==T)
techrep_qcP<-pdqc %>% filter(MethID %in% techrep_ids & qc_flag_any==F)
```

```{r}

```

### 2021-09-07 Reports/DataDictionary.R

@@ -1,0 +1,13 @@
+FFmeta<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/FFMetadata_v07.csv")
+idvars<-c("id", "ff_id")
+demovars<-c("cm1age", "cm1bsex", "cm1ethrace", "m1h3", "m1h3a", "m1h3b", "m1i4", "m1i4a", "m1i4b", "cf1age", "cf1ethrace", "f1h3", "f1h3a", "f1h3b")
+smokingvars<-c("m1g4", "f1g4", "f2j5", "f2j5a", "f2j7", "f2j7a", "f3j31", "f3j32", "f4j18", "f4j19", "f5g17", "f5g18", "k5f1k", "k5f1l", "k6d40", "k6d41", "k6d41", "k6d42", "k6d43", "k6d45", "k6d46", "k6d47", "m2j5", "m2j5a", "m2j7", "m2j7a", "m4j18", "m4j19", "m5g17", "m5g18", "n5f17", "n5f18", "p3a22", "p3a23", "p3a23a", "p3a23b", "p3a24", "p4a22", "p4a23", "p5h15", "p5h15b", "p5q3cr", "p6h74", "p6h75", "p6h76", "p6h77", "p6h78")
+prenataldrugusevar<-c("m1g2", "m1g3", "m1g5", "m1g6")
+FFmyMeta<-FFmeta %>% filter(new_name %in% c(idvars, demovars, smokingvars, prenataldrugusevar))
+
+table(FFmyMeta$type)
+myContinuousvars<-FFmyMeta %>% filter(type=="Continuous") %>%pull(new_name)
+myBinary<-FFmyMeta%>%filter(type=="Binary")%>%pull(new_name)
+myCategorical_O<-FFmyMeta%>%filter(type=="Ordered Categorical")%>%pull(new_name)
+myCategorical_U<-FFmyMeta%>%filter(type=="Unordered Categorical")%>%pull(new_name)
+

### 2021-09-07 UsefulCode/SVAAnalysis.R

@@ -1,0 +1,70 @@
+#libraries
+library(sjlabelled)
+library(tidyr)
+library(dplyr)
+library(purrr)
+library(furrr)
+library(stringr)
+library(sva)
+
+
+#directory based programming
+data.dir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
+output.dir=paste0(data.dir, 'CreatedData/')
+#variables 
+var.filter=5000
+
+#load in betaqc matrix
+betaqc<-readRDS(file=paste0(data.dir, 'OGData/', "noob_filtered.rds"))
+file.path(paste0(data.dir, 'CreatedData/'), list.files(paste0(data.dir, 'CreatedData/')))[(str_detect(file.path(paste0(data.dir, 'CreatedData/'), list.files(paste0(data.dir, 'CreatedData/'))), 'NoSmk'))]%>%map(load, envir=.GlobalEnv)
+
+pheno<-rbind(child_nosmoke, teen_nosmoke) # dataset with the most exclusion i.e. squaring dataset off for all models (excluding kids who smoke in figure)
+
+#make a matrix of model data 
+#filter beta matrix to only those in analytic subset 
+betaqc <-betaqc[, colnames(betaqc) %in% pheno$MethID]
+#check that dim of pheno and betaqc match 
+try(if(nrow(modeldata)!=ncol(betaqc)) stop("# obs in betaqc & pheno do not match"))
+
+#want to make sure that the order of pheno and betaqc match
+pheno=pheno[match(colnames(betaqc), pheno$MethID), ]
+try(if(identical(colnames(betaqc), pheno$MethID)==FALSE) stop('STOP: Sample names do not match'))
+
+#make model matrix including variable of interest as factor variable & adjustment variables 
+mod=model.matrix(~as.factor(smkPreg_binary), data=pheno)
+
+#make null model containing only adjustment variables
+mod0=model.matrix(~1, data=pheno)
+
+#run with reasonable variance filter (5k)
+svobj=sva(betaqc, mod, mod0, vfilter=var.filter)
+sv.v5k=data.frame(svobj$sv); colnames(sv.v5k)=paste('sv', 1:ncol(sv.v5k), sep='')
+sv.v5k=cbind(pheno%>%select(id, idnum, MethID), sv.v5k)
[…]

### 2021-09-07 UsefulCode/SVAbyAncestry.R

@@ -1,0 +1,47 @@
+#libraries
+library(sjlabelled)
+library(tidyr)
+library(dplyr)
+library(purrr)
+library(furrr)
+library(stringr)
+library(sva)
+
+
+#directory based programming
+data.dir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
+output.dir=paste0(data.dir, 'CreatedData/')
+#variables 
+var.filter=5000
+
+#load in betaqc matrix
+betaqc<-readRDS(file=paste0(data.dir, 'OGData/', "noob_filtered.rds"))
+file.path(paste0(data.dir, 'CreatedData/'), list.files(paste0(data.dir, 'CreatedData/')))[(str_detect(file.path(paste0(data.dir, 'CreatedData/'), list.files(paste0(data.dir, 'CreatedData/'))), 'NoSmk'))]%>%map(load, envir=.GlobalEnv)
+
+pheno<-rbind(child_nosmoke, teen_nosmoke) # dataset with the most exclusion i.e. squaring dataset off for all models (excluding kids who smoke in figure)
+
+pheno.ancestry=pheno%>% group_by(ancestry)%>%nest()
+
+#function to apply to each ancestry group
+my.sva=function(x, vf=1000,...){
+  bqc <-betaqc[, colnames(betaqc) %in% x$MethID]
+  try(if(nrow(x)!=ncol(bqc)) stop("# obs in betaqc & pheno do not match"))
+  x=x[match(colnames(bqc), x$MethID), ]
+  try(if(identical(colnames(bqc), remove_label(x$MethID))==FALSE) stop('STOP: Sample names do not match'))
+  #make model matrix including variable of interest as factor variable & adjustment variables 
+  mod=model.matrix(~as.factor(smkPreg_binary), data=x)
+  #make null model containing only adjustment variables
+  mod0=model.matrix(~1, data=x)
+  svobj=sva(bqc, mod, mod0, vfilter=vf)
+  sv=data.frame(svobj$sv)
+  colnames(sv)=paste('sv', 1:ncol(sv), sep='')
+  sv=cbind(x%>%select(id, idnum, MethID), sv)
+  return(sv)
+}
+
+#apply to each ancestry group
+sv.ancestry=pheno.ancestry%>%mutate(sv=map(data, my.sva, vf=5000))
+
+#
+save(sv.ancestry, file=paste0(output.dir, 'svAncestry5k.Rds'))
+print('Finished running SVA at Variance Filter of 5000 at different ancestries ')

### 2021-09-07 UsefulCode/makePolyEpiScores.R

@@ -1,0 +1,29 @@
+## ---- makePolyEpiScores.R
+#' Title
+#'
+#' @param m a matrix; cleaned matrix of methylation values (0-1) from your data where rows are cpg sites and columns are samples; rownames should be of format cg########
+#' @param b a list; a list of dataframes containing a column CpG with cpg sites named in format cg######## and a column Coef containing relavent coeficient, should be a named list (names will become score names)
+#' @param transformopt character, the method for transforming the m values, default none, options zscore or meancenter
+#'
+#' @return scores
+#' @export
+#'
+#' @examples
+makePolyEpiScore<-function(m, b, transformopt='none'){
+  #check inputs
+  try(if(is.matrix(m)==F) stop('m is not a matrix'))
+  try(if(is.list(b)==F) stop('b is not a list'))
+  if(is.null(names(b))){names(b)=paste0('option', 1:length(b))}
+  #generate poly methylation scores
+  scores<-lapply(b, function(x){
+    commonp<-Reduce(intersect,list(x%>%filter(!is.na(Coef))%>%pull(`CpG`), rownames(m)))%>% unique # filter to non-missing shared cpg sites
+    beta1<-m[match(commonp, rownames(m)),] #pull shared cpg sites from methylation matrix 
+    #apply transform to cpgs of the cleaned methylation matrix 
+    print(paste0('transforming the methylation matrix:', transformopt))
+    if(transformopt=='zscore'){beta1=t(scale(t(beta1)))} #center and scale the rows (cpgs) of the (shared cpg sites) methylation matrix
+    if(transformopt=='meancenter'){beta1=t(scale(t(beta1), center=T, scale=F))} #center the rows (cpgs) of (shared cpgs sites) methylation matrix only
+    x1<-x %>% filter(`CpG` %in% commonp) #pull shared cpg sites from coefficient matrix
+    (as.numeric(x1[['Coef']]) %*% beta1)%>%unlist%>%c #matrix multiplication of shared coefficients and shared methylation
+  })%>% bind_cols()
+  return(scores)
+}

### 2021-09-08 UsefulCode/SVAAnalysis.R
updated SVA analysis

@@ -52,12 +52,12 @@ my_num.sv=function(x, y='be'){
   return(my.n)
 }
 
-plan(multicore, workers=5)
-sv.n.be=possible.n %>% future_map(my_num.sv)
+
+sv.n.be=possible.n %>% map(my_num.sv)
 save(sv.n.be, file=paste0(output.dir, 'nSVbe.Rdata'))
 
-plan(multicore, workers=5)
-sv.n.leek=possible.n %>% future_map(my_num.sv, y='leek')
+
+sv.n.leek=possible.n %>% map(my_num.sv, y='leek')
 save(sv.n.leek, file=paste0(output.dir, 'nSVleek.Rdata'))
 
 #run with no variance filter 

### 2021-09-08 Reports/Poster/IGES_poster.Rmd
drafted poster for iges

---
title: Polymethylation Scores for Prenatal Maternal Smoke Exposure Persist Until Age 15 and Are Detected in Saliva
author:
  - name: Freida Blostein
    affil: 1
    twitter: blostein
    email: blostein@umich.edu
    orcid: 0000-0002-1099-3857
    main: true
  - name: Jonah Fisher
    affil: 2
  - name: John Dou
    affil: 1
  - name: Lisa Schenper
    affil: 3
  - name: Daniel A. Notterman
    affil: 3
  - name: Colter Mitchell
    affil: 2
  - name: Kelly M. Bakulski
    affil: 1
affiliation:
  - num: 1
    address: Department of Epidemiology, University of Michigan School of Public Health, Ann Arbor, Michigan, USA
  - num: 2
    address: Institute for Social Research, University of Michigan, Ann Arbor, Michigan, USA
  - num: 3
    address: Department of Molecular Biology, Princeton University, Princeton, New Jersey, USA
main_findings:
  - "Make **better posters** with RMarkdown + **posterdown**."
  - '![](https://raw.githubusercontent.com/brentthorne/posterdown/master/images/betterhexlogo.png){.main_pic}'
logoleft_name: '![](https://raw.githubusercontent.com/brentthorne/posterdown/master/images/qr-code-black.png){.main-img-left}'
logoright_name: '![](https://raw.githubusercontent.com/brentthorne/posterdown/master/images/betterhexlogo.png){.main-img-right}'
output: 
  posterdown::posterdown_betterland:
    self_contained: false
    pandoc_args: --mathjax
    highlight: haddock
    number_sections: false
link-citations: true
bibliography: packages.bib
---

```{r, include=FALSE}
knitr::opts_chunk$set(results = 'asis',
                      echo = FALSE,
                      warning = FALSE,
                      tidy = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      out.width = "100%")
options(knitr.table.format = "html") 
```

```{r myplot, include=FALSE}
svg('myplot.svg')
plot(iris$Sepal.Length, iris$Sepal.Width)
dev.off()
```

# Introduction

This is the `posterdown_betterland` template for the {posterdown} package! I was inspired by the twitter thread of [Mike Morrison](https://mobile.twitter.com/mikemorrison/status/1110191245035479041) and wanted to apply the `#betterposter` concept to the reproducible (yet simple to use) functionality of the {posterdown} package [@R-posterdown]. If you're not an R user don't sweat as you do **NOT** need to use it at all! Feel free to use only the Markdown functionality of this package :)

```{r, include=FALSE}
knitr::write_bib(c('posterdown', 'rmarkdown','pagedown'), 'packages.bib')
```

## Objectives

1. Pick a template layout.
2. Write/ create your poster content distraction free.
3. Let posterdown do its thing!

# Methods

I will show here how to include poster elements that may be useful, such as an equation using mathjax:

$$
E = mc^2
$$

To reference a citation you can add your `.bib` file to the working directory and name it in the YAML metadata or generate an automated one as done here, then you only need to reference the label value in the `.bib` file. For example this package is built on top of the wonderful {pagedown} package and I will cite it at the end of this sentance using this in the rmd `[@R-pagedown]` [@R-pagedown].

To get a better understanding of how to include features like these please refer to the {posterdown} [wiki](https://github.com/posterdown/wiki).

**_Now on to the results!_**

<br>

# Results

Here you may have some figures to show off, bellow I have made a scatterplot with the infamous Iris dataset and I can even reference to the figure automatically like this, `Figure \@ref(fig:irisfigure)`, Figure \@ref(fig:irisfigure).

```{r, irisfigure, fig.cap='Here is a caption for the figure. This can be added by using the "fig.cap" option in the r code chunk options, see this [link](https://yihui.name/knitr/options/#plots) from the legend himself, [Yihui Xie](https://twitter.com/xieyihui).', out.width="80%"}
par(mar=c(2,2,0,1))
plot(x = iris$Sepal.Length, y = iris$Sepal.Width,
     col = iris$Species, pch = 19, xlab = "Sepal Length",
     ylab = "Sepal Width")
```

Maybe you want to show off some of that fancy code you spent so much time on to make that figure, well you can do that too! Just use the `echo=TRUE` option in the r code chunk options, Figure \@ref(fig:myprettycode)!

```{r myprettycode, echo=FALSE,collapse=TRUE, fig.cap='Boxplots, so hot right now!', fig.height=3.5, out.width="80%"}
#trim whitespace
par(mar=c(2,2,0,0))
#plot boxplots
boxplot(iris$Sepal.Width~iris$Species,
        col = "#008080", 
        border = "#0b4545",
        ylab = "Sepal Width (cm)",
        xlab = "Species")
```

How about a neat table of data? See, Table \@ref(tab:iristable):

```{r, iristable}
knitr::kable(
  iris[1:8,1:5], format = "html",
  caption = "A table made with the **knitr::kable** function.",
  align = "c", col.names = c("Sepal <br> Length",
                             "Sepal <br> Width",
                             "Petal <br> Length",
                             "Petal <br> Width",
                             "Species"),
  escape = FALSE)
```

# References

### 2021-09-08 Reports/Poster/IGES_poster_vertical.Rmd
drafted poster for iges

--- 2021-09-08 Reports/Poster/IGES_poster.Rmd
@@
 ---
-title: Polymethylation Scores for Prenatal Maternal Smoke Exposure Persist Until Age 15 and Are Detected in Saliva
+main_topsize: 0.1 #percent coverage of the poster
+main_bottomsize: 0.05
+#ESSENTIALS
+title: Polymethylation Scores for Prenatal Maternal Smoke Exposure Persist Until Age
+  15 and Are Detected in Saliva
+title_textsize: 100
 author:
@@
   orcid: 0000-0002-1099-3857
-    main: true
+  main: yes
 - name: Jonah Fisher
@@
 - num: 1
-    address: Department of Epidemiology, University of Michigan School of Public Health, Ann Arbor, Michigan, USA
+  address: Department of Epidemiology, University of Michigan School of Public Health,
+    Ann Arbor, Michigan, USA
 - num: 2
-    address: Institute for Social Research, University of Michigan, Ann Arbor, Michigan, USA
+  address: Institute for Social Research, University of Michigan, Ann Arbor, Michigan,
+    USA
 - num: 3
-    address: Department of Molecular Biology, Princeton University, Princeton, New Jersey, USA
+  address: Department of Molecular Biology, Princeton University, Princeton, New Jersey,
+    USA
 main_findings:
-  - "Make **better posters** with RMarkdown + **posterdown**."
-  - '![](https://raw.githubusercontent.com/brentthorne/posterdown/master/images/betterhexlogo.png){.main_pic}'
-logoleft_name: '![](https://raw.githubusercontent.com/brentthorne/posterdown/master/images/qr-code-black.png){.main-img-left}'
-logoright_name: '![](https://raw.githubusercontent.com/brentthorne/posterdown/master/images/betterhexlogo.png){.main-img-right}'
+  - "**Polymethylation scores** for prenatal maternal smoke exposure **persist over time** and are **robust across ancestries**"
+logoleft_name: 'Images/Epid_informal_horiz_white.png'
+logoright_name: 'Images/Fragile_Families_Logo.png'
+logocenter_name: https&#58;//raw.githubusercontent.com/brentthorne/posterdown/master/images/qr-code-black.png
+primary_colour: "#0048A4"
+secondary_colour: "#5885AF"
+accent_colour: "#E35A01"
 output: 
-  posterdown::posterdown_betterland:
+  posterdown::posterdown_betterport:
     self_contained: false
     pandoc_args: --mathjax
-    highlight: haddock
     number_sections: false
+bibliography: packages.bib
 link-citations: true
-bibliography: packages.bib
 ---
 
+<style>
+#main-img-left {
+ width: 20%
+}
+#main-img-center {
+ width: 2%
+}
+#main-img-right {
+width=20%
+}
+</style>
+
 ```{r, include=FALSE}
-knitr::opts_chunk$set(results = 'asis',
-                      echo = FALSE,
+knitr::opts_chunk$set(echo = FALSE,
                       warning = FALSE,
@@
 
-```{r myplot, include=FALSE}
-svg('myplot.svg')
-plot(iris$Sepal.Length, iris$Sepal.Width)
-dev.off()
-```
-
 # Introduction
 
-This is the `posterdown_betterland` template for the {posterdown} package!
+This is the `posterdown_betterport` template for the {posterdown} package!
 I was inspired by the twitter thread of [Mike Morrison](https://mobile.twitter.com/mikemorrison/status/1110191245035479041) and wanted to apply the `#betterposter` concept to the reproducible (yet simple to use) functionality of the {posterdown} package [@R-posterdown].
@@
 
-## Objectives
+## Gaps
 
@@
 
-<br>
-
 # Results
@@
 
-```{r myprettycode, echo=FALSE,collapse=TRUE, fig.cap='Boxplots, so hot right now!', fig.height=3.5, out.width="80%"}
+```{r myprettycode, echo=TRUE, fig.cap='Boxplots, so hot right now!', fig.height=6, out.width="80%"}
 #trim whitespace
@@
 knitr::kable(
-  iris[1:8,1:5], format = "html",
+  iris[1:15,1:5], format = "html",
   caption = "A table made with the **knitr::kable** function.",

### 2021-09-08 Reports/SVAReport.Rmd
report on SVA parameters

---
title: "SVA Investigation"
author: "F blostein"
date: "9/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#libraries
library(sjlabelled)
library(tidyr)
library(dplyr)
library(purrr)
library(furrr)
library(stringr)
library(sva)
library(ggplot2)
```

# Surrogate variable analysis

```{r}
datadir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/CreatedData/"

file.path(datadir, list.files(datadir))[(str_detect(file.path(datadir, list.files(datadir)), 'NoSmk'))]%>%map(load, envir=.GlobalEnv)
modeldata=rbind(child_nosmoke, teen_nosmoke)

load(paste0(datadir, 'nSVbe.Rdata'))
load(paste0(datadir, 'nSVleek.Rdata'))

```

## Main idea

Use the beta matrix to calculate surrogate variables; control for these surrogate variables instead of for known covariates in multivariable models, i.e. regress global methylation on prenatal maternal smoking + surrogate variables etc, etc.

Surrogate variable models can be used in our EWAS and in our models that use summary methylation variables such as global methylation, methylation clocks and polymethylation scores.

I used John's code from 'Autism associated DNA methylation at birth from multiple tissues' code from the lab github as a guide.

I used the final, complete-case analytic subset from my analysis cleaning (excluding children with missing data in any of the model variables & excluding children who report ever smoking, n= `r nrow(modeldata)`). I only included the main predictor of interest in the SVA model, since that's what John did.  The tutorial says to include any variables of interest, so we may want to consider including age or ancestry as variables that we are holding out from the surrogate variable creation (or, do SVA separately by ancestry, see later)

## Variance filtering and method of SVA

When the number of features is very large ($m>100,000$), the SVA procedure can take a long time, so the function includes a variance filter term. In this approach, the features are ranked from the most to least variable and only the first $x$ features are used. This improves computational time but caution should be used, as surrogate variables are only estimated on a subsection of the matrix.

John doesn't do this, he said that running SVA on datasets he's worked on can take on the order of hours and for Fragile Families sample size he could envision it taking much longer.

I've been trying to run on the full matrix, submitting as a batch job. I've had several jobs run out of memory, so I've upped the memory available, but in the meantime I've been experimenting with different levels of variance cut-offs. I've also been varying the method of surrogate variable computation, two methods are available, 'be' which is the default (John uses this) and leek, which is an asymptomatic method intended for large sample sizes, see resource [1](https://support.bioconductor.org/p/97469/), [2](https://cebp.aacrjournals.org/content/cebp/suppl/2018/06/30/1055-9965.EPI-17-0728.DC1/187479_3_supp_4617635_p5bqpt.pdf)

### Number of surrogate variables by variance filter and method of SVA

```{r}
df.param=data.frame(SV=as.numeric(c(c(unlist(sv.n.be)[c(F, T)]),
                  c(unlist(sv.n.leek)[c(F, T)]))), 
           method=c(c(rep('be', length(sv.n.be))), 
                      c(rep('leek', length(sv.n.leek)))), 
           varianceFilter=rep(c(1000, 5000, 10000, 50000, 100000), 2))

ggplot(df.param, aes(x=varianceFilter,y=SV))+geom_point()+facet_wrap(~method)+scale_x_log10()+theme_classic()
```

I tested the number of predicted surrogate varaibles at a variance filter of 1,000 (lowest recommended), 5,000, 10,000, 50,000 and 100,000 features. At 1,000 features there were 42 surrogate variables when using the be method. John said this is high in his experience. Increasing the number of features to 500 halves the number of surrogate variables detected to 20, after that there is a minor increase in SVs.  Using the leek method, I get 0 surrogate variables (except when using a variance filter of 50k, at which point I get 2 surrogate variables)

## Surrogate variables using 5,000 features

Based on the above, I've been examining the surrogate variables resulting from sva on 5,000 most variable features and the be method of calculation.

### Heatmap of known covariates and surrogate variables

```{r load surrogate variables}
load(file=paste0(datadir, 'sv5k.Rds'))
my.pd=left_join(modeldata, sv.v5k)
```

```{r heatmap function}
bring.the.heatmap.sv <- function(pd,covar,folder){
  library(gplots)

  nsv <- length(grep('sv',names(pd)))

  #generate correlation matrix between SVs and variables
  correlations <- matrix(0,nrow=length(covar),ncol=nsv)
  colnames(correlations) <- paste('sv',1:nsv,sep='')
  rownames(correlations) <- covar

  #take svs
  svs <- pd[,colnames(correlations)]

  #correlations between svs and variables:
  # cor test if continuous
  # anova is multi category
  # t-test if two category
  for(col in covar){
    var <- pd[,col]
    #check to see if it is numeric and isn't really a two category thing
    if(class(var)=='numeric' & dim(table(var))>2){
      correlations[col,] <- apply(svs,2,FUN = function(x,y) cor.test(x,y)$p.value,y=var)
    }else{
      if(length(levels(factor(var)))>2){
        correlations[col,] <- apply(svs,2,FUN = function(x,y) anova(lm(x~y))$'Pr(>F)'[1],y=var)
      }else{
        correlations[col,] <- apply(svs,2,FUN = function(x,y) t.test(x~y)$p.value,y=var)
      }
    }
  }

  #pdf(paste0(folder,'heatmap-sv.pdf'))
  color.gradient <- colorRampPalette(c('firebrick','ivory3','royalblue'),bias=3)
  heatmap.2(correlations,Rowv=FALSE,Colv=FALSE,cexRow=1,cexCol=1,dendrogram='none',density.info='none',trace='none',
            col=color.gradient(20), key.xlab='p-value', main='Associations Between SVs\n and Given Covariates',
            #lmat=rbind(c(0,3,3,3),c(2,0,4,0),c(0,1,1,1)), lhei=c(0.5,1.5,3), lwid=c(0.12,0.145,0.4,0.335),
            margin=c(4,12)
            )
  #dev.off()

  #return('check your plot')
}
```

```{r plot heatmap, fig.width=20, fig.height=12}
bring.the.heatmap.sv(my.pd, c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'ancestry', 'cm1bsex', 'Epi', 'Fib', 'IC', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))
```

## Ancestry-stratified vs non-stratified

Another thing John and I chatted about briefly was the iquestion of whether surrogate variables should be run in the entire analytic subset or separately in each strata for stratified models, this was John's take:

> hm, reminds me of the question of whether you should re-compute genetic PCs in SNPs after stratifying where there isn't a clear answer
my only concern would be if something that strongly impacts methylation, like smoking, has huge race/ethnicity differences
I'm not sure if that would mess with surrogate variables

Again, using the variance filter to 5000 features and the 'be' method, I calculated the surrogate variables in each ancestry subset separately.

```{r}
load(file=paste0(datadir, 'svAncestry5k.Rds'))
```

Fairly similar numbers of surrogate variables were detected, with fewer in the European and Hispanic groups, supporting the notion that larger sample size may lead to the detection of more surrogate variables (number of surrogate variables = `r sv.ancestry %>% summarise(sv_num=map(sv, ncol))%>%pull(sv_num)%>%unlist()%>%paste0(., collapse=', ')` in the African, European and Hispanic ancestry groups respectively).

```{r}
pd.ancestry=sv.ancestry %>% mutate(pd=map2(data, sv, left_join))
```
```{r, echo=FALSE,results='hide',fig.keep='all'}
map(.x=pd.ancestry$pd, .f=~(bring.the.heatmap.sv(as.data.frame(.x), covar=c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'cm1bsex', 'Epi', 'Fib', 'IC', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))))
```

## Remaining questions

- What variance filter to use? or don't use any? 
- Add more variables of interest to be held out in the surrogate analysis? 
- Ancestry stratified or not for calculating SVs?

### 2021-09-13 Papers/Polymethylation_scores/index_paper.Rmd
added PMS paper folder and skeleton

---
title: "Polymethylation scores paper"
author: "F blostein"
date: "9/13/2021"
output: pdf_document
---

```{r child = here::here("Papers", "Polymethylation_scores", "01_titlepage.Rmd")}
```

```{r child = here::here("Papers", "Polymethylation_scores", "02_abstract.Rmd")}
```

### 2021-09-13 Papers/Polymethylation_scores/98_declarations.Rmd
added paper sections

# Declarations

## Ethics approval and consent to participate

## Consent for publication

Not applicable

### 2021-09-14 Reports/Poster/IGES_poster_vertical.Rmd
modify .gitignore

@@
     number_sections: false
-bibliography: packages.bib
+bibliography: "Methylation_SmokingBib.bib"
 link-citations: true
@@
 
+```{=html}
 <style>
@@
 </style>
-
+```
 ```{r, include=FALSE}
@@
 options(knitr.table.format = "html") 
+mydir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/Reports/Poster"
+imagedir=paste0(mydir, "/Images/")
+setwd(mydir)
 ```
@@
 
-This is the `posterdown_betterport` template for the {posterdown} package!
-I was inspired by the twitter thread of [Mike Morrison](https://mobile.twitter.com/mikemorrison/status/1110191245035479041) and wanted to apply the `#betterposter` concept to the reproducible (yet simple to use) functionality of the {posterdown} package [@R-posterdown].
-If you're not an R user don't sweat as you do **NOT** need to use it at all!
-Feel free to use only the Markdown functionality of this package :)
+Prenatal maternal smoking is associated with DNA methylation in children [@Joubert:2016ue].
+Thus, summary metrics of DNA methylation could serve as reliable and specific biomarkers of prenatal smoke exposure.
 
-```{r, include=FALSE}
-knitr::write_bib(c('posterdown', 'rmarkdown','pagedown'), 'packages.bib')
+```{r, joubert_fig, fig.cap='Volcano plot showing associations between CpG methylation in cord blood and prenatal maternal smoking exposure in @Joubert:2016ue'}
+knitr::include_graphics('Images/Joubert_volcano.jpeg')
 ```
 
-## Gaps
+***But some gaps remain...!***
 
-1. Pick a template layout.
-2. Write/ create your poster content distraction free.
-3. Let posterdown do its thing!
+1.
+![](Images/Icons/noun_blood%20sample_4011680.svg) **Tissue:** Previous studies have measured methylation in peripheral or cord blood.
+**Are the same methylation markers of prenatal maternal smoking also present in saliva, an easier to collect sample?**
+
+2.
+![](Images/Icons/noun_ethnicity_94199.svg) **Ancestry:** Previous study populations have been of European descent.
+**Are methylation markers of prenatal maternal smoking consistent across different ancestry groups?**
+
+3.
+![](Images/Icons/noun_Life%20Stages_398930.svg) **Time:** Previous study populations have examined methylation at a single cross-sectional time point in infancy or early childhood.
+**Are methylation markers of prenatal maternal smoking persistent as children age?**
+
+***We tested the association between prenatal maternal smoking and DNA methylation summary measures in sequential saliva samples from a diverse-ancestry, longitudinal birth cohort***
 
@@
 
-I will show here how to include poster elements that may be useful, such as an equation using mathjax:
+The **Fragile Families and Child Wellbeing study** is a *longitudinal* cohort of US children followed from birth until 22 years of age.
+
+-   *Smoking during pregnancy*: Maternal interview at birth.
+
+-   *Methylation*: Saliva samples from 9 and 15 year visit; 450K chip.
+
+-   *Covariates*: Maternal alcohol and drug use during pregnancy (birth interview), income-to-poverty ratio (birth), maternal postnatal smoking (interviews), child smoking (interviews), child's genetic ancestry categorization (principal components of genetic data), cell type proportions (estimated from methylation data)
+
+```{r, ff_fig, fig.cap='Study design and timeline of exposure, outcome and covariate assessment in the Fragile Families and Child Wellbeing cohort'}
+knitr::include_graphics('Images/FF_studydesign.png')
+```
+
+## Methylation summary measures
+
+We used several common summary measures of methylation data, including global methylation and methylation clocks.
+We also created several prenatal maternal smoking specific polymethylation scores, using regression coefficients from @Joubert:2016ue.
+
+**Polymethylation scores**
+
+For $i$ individuals with $j$ CpG sites measured:
 
 $$
-E = mc^2
+Score_i=\sum \color{darkgreen} {\beta_{regression_{j}}} *  \color{darkblue} {t(B_{CpG_{ij}})}
 $$
 
-To reference a citation you can add your `.bib` file to the working directory and name it in the YAML metadata or generate an automated one as done here, then you only need to reference the label value in the `.bib` file.
-For example this package is built on top of the wonderful {pagedown} package and I will cite it at the end of this sentance using this in the rmd `[@R-pagedown]` [@R-pagedown].
+[Regression coefficients for CpG sites from a previous epigenome wide association analysis]{style="color: darkgreen;"}<sup>a</sup> are used to weight [each individual's methylation beta value at that CpG site]{style="color: darkblue;"}<sup>b</sup>.
+These weighted methylation values are summed across all CpG sites.
 
-To get a better understanding of how to include features like these please refer to the {posterdown} [wiki](https://github.com/posterdown/wiki).
+```{=html}
+<div style="line-height: 1em; margin-bottom:5px;">
 
-**_Now on to the results!_**
+<span style="color: darkgreen; font-size:18px"><sup>a</sup>We benchmarked 4 different regression coefficients made available in @</span>
+
+<span style="color: darkblue; font-size:18px"><sup>b</sup>Methylation beta values are continuous and bounded at $[0, 1]$.
+Concerned this might upweight the importance of highly methylated probes, we benchmarked 3 different transformations (t()) of methylation beta values: no transform, mean-centered and z-score.</span>
+
+</div>
+```
 
@@
 
+```{r data read and set up}
+
+```
+
 ```{r, irisfigure, fig.cap='Here is a caption for the figure.
 This can be added by using the "fig.cap" option in the r code chunk options, see this [link](https://yihui.name/knitr/options/#plots) from the legend himself, [Yihui Xie](https://twitter.com/xieyihui).', out.width="80%"}
-par(mar=c(2,2,0,1))
-plot(x = iris$Sepal.Length, y = iris$Sepal.Width,
-     col = iris$Species, pch = 19, xlab = "Sepal Length",
-     ylab = "Sepal Width")
-```
 
-Maybe you want to show off some of that fancy code you spent so much time on to make that figure, well you can do that too!
-Just use the `echo=TRUE` option in the r code chunk options, Figure \@ref(fig:myprettycode)!
-
-```{r myprettycode, echo=TRUE, fig.cap='Boxplots, so hot right now!', fig.height=6, out.width="80%"}
-#trim whitespace
-par(mar=c(2,2,0,0))
-#plot boxplots
-boxplot(iris$Sepal.Width~iris$Species,
-        col = "#008080", 
-        border = "#0b4545",
-        ylab = "Sepal Width (cm)",
-        xlab = "Species")
 ```

### 2021-09-14 Scripts/03_summary_methyl_models.R

@@ -1,0 +1,133 @@
+###################################Libraries
+library(tidyr)
+library(stringr)
+library(dplyr)
+library(compareGroups)
+library(kableExtra)
+library(haven)
+library(sjlabelled)
+library(matchmaker)
+library(EpiDISH)
+library("PerformanceAnalytics")
+library(DescTools)
+library(purrr)
+library(broom)
+library(broom.mixed)
+library(ggplot2)
+library(multcomp)
+library(ggpubr)
+library(cowplot)
+library(geepack)
+library(lme4)
+library(lmerTest)
+library(stringr)
+library(sva)
+##################################Directories & data
+data.dir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/CreatedData"
+outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output"
+#read data 
+file.path(data.dir, list.files(data.dir))[(str_detect(file.path(data.dir, list.files(data.dir)), 'NoSmk'))]%>%
+  map(load, envir=.GlobalEnv)
+#bind data
+modeldata<-rbind(child_nosmoke, teen_nosmoke)
+#label data 
+set_label(modeldata)<-get_label(child_nosmoke)
+#set as factors 
+modeldata$PostnatalMaternalSmokingAny<-factor(modeldata$PostnatalMaternalSmokingAny, 
+                                              levels=c("No maternal smoking at age 1 and age 5", 
+                                                       "Maternal smoking at age 1 or age 5"))
+modeldata$SmkAtVisitPastmonth<-factor(modeldata$SmkAtVisitPastmonth, 
+                                      levels=c("No smoking", 
+                                               "Less than pack a day", 
+                                               "Pack or more a day", "Missing"))
+modeldata$slide<-factor(modeldata$slide)
+modeldata$Sample_Plate<-factor(modeldata$Sample_Plate)
+
+#################################Set variables for modeling
+#outcomes & outcome_labels
+y_vector<-c("globalmethylation", "pediatric", "grim", "anynewborn_center", "SSnewbornCT_center", "SSolder_center")
+outcome_labels=c('Global methylation', 'Pediatric clock', 'GRIM clock',  
+                 'Any smoking polymethylation score (newborns)', 
+                 'Sustained smoking polymethylation score (newborns, cell-type controlled)', 
+                 'Sutained smoking polymethylation score (older children)')
+
[…]

### 2021-09-14 Papers/Polymethylation_scores/08_methods.Rmd
wrote methods section: polymethylation scores

---
output: html_document
bibliography: "../Methylation_SmokingBib.bib"  
---

# Methods

## Cohort

The Fragile Families and Child Wellbeing Study (FFCW) is a nationally-representative cohort of nearly 5,000 children born in 20 U.S. cities between 1998 and 2000 [@Reichman:2001wv]. FFCW sampled births using a three-stage stratified random sample design, described in detail elsewhere [@Reichman:2001wv]. FFCW excluded the following births from the study: those with parents who planned to place the child for adoption, those where the father was deceased, those who did not speak English or Spanish well enough to be interviewed, those where the mothers or babies were too ill to complete the interview and those where the baby died before the interview could take place. At some of the hospitals, parents who were under 18 were not eligible to be interviewed due to hospital prohibition. Children were followed from birth until age fifteen. At the time of birth, medical record information was extracted from the hospital at which the mother gave birth. Interviews were conducted with the mother at birth in person and with father in person or over the telephone if he was not present. Interviews were also conducted with the mother and father when the child was one, three, five, and nine via telephone. Additionally, at ages three, five and nine, a separate primary caregiver survey was administered to the individual with whom the child was living either over the telephone or in the home. The child was interviewed at age nine in the home and at age fifteen by telephone or in the home. At ages nine and fifteen a saliva sample was taken from the child for DNA and DNA methylation measurements [@Reichman:2001wv].

## Prenatal maternal smoking measurement

## Covariates

## DNA methylation measurement

## Sample filtering

## Probe filtering

## Methylation summary measures

We created several methylation summary measures. Global methylation was calculated for each sample as the mean methylation value of each sample across the cleaned probe set. Methylation clocks were calculated for each sample according to the methods of --INSERT HERE--. To create polymethylation scores, 6,000 (CHECK) CpG sites with regression coefficients reported in @Joubert:2016ue were multiplied by methylation beta values from FFCW and summed. We calculated polymethylation scores using regression coefficients from 4 different regressions reported in @Joubert:2016ue (LIST) and 3 different methods of transforming the methylation beta values (no transformation, mean center, and z-score standardization).

## Statistical analysis

### 2021-09-14 Papers/Polymethylation_scores/index_paper.Rmd
wrote methods section: polymethylation scores

@@
 output: pdf_document
+bib: Methylation_SmokingBib.bib
 ---
@@
 ```
+
+```{r child = here::here("Papers", "Polymethylation_scores", "03_keywords.Rmd")}
+```
+
+```{r child = here::here("Papers", "Polymethylation_scores", "04_background.Rmd")}
+```
+
+```{r child = here::here("Papers", "Polymethylation_scores", "05_results.Rmd")}
+```
+
+```{r child = here::here("Papers", "Polymethylation_scores", "06_discussion.Rmd")}
+```
+
+```{r child = here::here("Papers", "Polymethylation_scores", "07_conclusions.Rmd")}
+```
+
+```{r child = here::here("Papers", "Polymethylation_scores", "98_declarations.Rmd")}
+```
+
+# References

### 2021-09-15 Reports/Poster/IGES_poster_vertical.Rmd
'added surrogate variables (complete matrix) to SVA Report and to models in script 03'

@@
 mydir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/Reports/Poster"
+codedir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/"
 imagedir=paste0(mydir, "/Images/")
@@
 
+## Table of smoking frequency in cohort and by ancestry
+
+## Polymethylation scores are a specific, persistant and consistent marker of prenatal maternal smoke exposure
+
 ```{r data read and set up}
-
+source(paste0(codedir, 'Scripts/', '03_summary_methyl_models.R'))
 ```
@@
 This can be added by using the "fig.cap" option in the r code chunk options, see this [link](https://yihui.name/knitr/options/#plots) from the legend himself, [Yihui Xie](https://twitter.com/xieyihui).', out.width="80%"}
+my_outcomes=c('Global methylation', 'Pediatric clock', 'Sustained smoking polymethylation score (older children)')
+
+local_models%>%unnest(tidied)%>%
+  filter(term=='smkPreg_binaryYes' & predictors=='secondhand smoke exposure model')%>%
+  filter(outcome%in% my_outcomes)%>%
+  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=ancestry))+
+  geom_point(position = position_dodge(width = 1))+
+  geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
+  facet_grid(.~outcome, scales='free_x', labeller = labeller(outcome=label_wrap_gen(15)))+
+  geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
+  theme_classic()+
+  theme(text = element_text(size=16))+theme(legend.position = 'bottom')+
+  scale_color_discrete('', limits=c('Age 9', 'Age 15'))+
+  scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
+  scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry'))
 
 ```
+
+## Polymethylation scores persisit from age 9 to age 15, with consistent associations
 

### 2021-09-15 Reports/SVAReport.Rmd
'added surrogate variables (complete matrix) to SVA Report and to models in script 03'

@@
 datadir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/CreatedData/"
+codedir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/Scripts/"
 
@@
 ```
+
 ```{r, echo=FALSE,results='hide',fig.keep='all'}
@@
 
+## Full matrix variable analysis
+
+```{r full surrogate variable}
+load(paste0(datadir, 'svNone.Rds'))
+#rename svs to get rid of space for heatmap function
+is_sv=function(x) str_replace_all(x, " ", "")
+sv.vNone=sv.vNone%>%rename_with(is_sv, contains('sv'))
+#left join
+my.pd.full=left_join(modeldata, sv.vNone)
+
+#heatmap
+bring.the.heatmap.sv(my.pd.full, c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'ancestry', 'cm1bsex', 'Epi', 'Fib', 'IC', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))
+
+```
+
+## Modeling with full surrogate variable as controlled variables
+
+```{r}
+source(paste0(codedir, "03_summary_methyl_models.R"))
+
+global_models%>%unnest(tidied)%>%
+  filter(term=='smkPreg_binaryYes')%>%
+  #filter(outcome%in% my_outcomes)%>%
+  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=predictors))+
+  geom_point(position = position_dodge(width = 1))+
+  geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
+  facet_grid(.~outcome, scales='free_x', labeller = labeller(outcome=label_wrap_gen(15)))+
+  geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
+  theme_classic()+
+  theme(text = element_text(size=16))+theme(legend.position = 'bottom')+
+  scale_color_discrete('', limits=c('Age 9', 'Age 15'))+
+  scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')
+
+```
 ## Remaining questions

### 2021-09-15 Scripts/03_summary_methyl_models.R

@@ -42,6 +42,11 @@ modeldata$SmkAtVisitPastmonth<-factor(modeldata$SmkAtVisitPastmonth,
                                                "Pack or more a day", "Missing"))
 modeldata$slide<-factor(modeldata$slide)
 modeldata$Sample_Plate<-factor(modeldata$Sample_Plate)
+#sv data - may change this depending on conversation w/ kelly
+load(paste0(datadir, 'svNone.Rds'))
+is_sv=function(x) str_replace_all(x, " ", "")
+sv.vNone=sv.vNone%>%rename_with(is_sv, contains('sv'))
+modeldata=left_join(modeldata, sv.vNone)
 
 #################################Set variables for modeling
 #outcomes & outcome_labels
@@ -49,7 +54,8 @@ y_vector<-c("globalmethylation", "pediatric", "grim", "anynewborn_center", "SSne
 outcome_labels=c('Global methylation', 'Pediatric clock', 'GRIM clock',  
                  'Any smoking polymethylation score (newborns)', 
                  'Sustained smoking polymethylation score (newborns, cell-type controlled)', 
-                 'Sutained smoking polymethylation score (older children)')
+                 'Sustained smoking polymethylation score (older children)')
+names(outcome_labels)=y_vector
 
 #model variables (excepting PCs, add in individually for global & ancestry specific models)
 base_model_vars<-"~smkPreg_binary+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+Sample_Plate"
@@ -59,18 +65,23 @@ secondhand_model_vars<-paste0(prenatal_model_vars, '+PostnatalMaternalSmokingAny
 global_pcs='+global_PC1+global_PC2'
 local_pcs='+local_PC1+local_PC2'
 
+#surrogate variable models
+surrogate_model_vars<-paste0('~smkPreg_binary+', 
+                             paste(colnames(modeldata)[str_detect(colnames(modeldata), 'sv')], collapse='+'))
+
 predictors=c('base model'=base_model_vars, 
              'prenatal exposures model'=prenatal_model_vars, 
              'secondhand smoke exposure model'=secondhand_model_vars)
-model_labels=c('base model', 'prenatal exposures model', 'secondhand smoke exposure model')
 
-#surrogate variable models
-surrogate_model_vars<-""
+model_labels=c('base model', 'prenatal exposures model', 'secondhand smoke exposure model', 'surrogate variable models')
+
+
 
 #stratification variables 
 ##############################
[…]

### 2021-09-27 Scripts/01_clean_phenotype_data.R
added roc models, began rewrite of processing code

@@ -1,0 +1,151 @@
+
+#####################libraries, directories, data###########################
+
+#libraries 
+library(tidyr)
+library(stringr)
+library(dplyr)
+library(purrr)
+library(tidyselect)
+library(sjlabelled)
+
+#directory based coding
+datadir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
+john_datadir<-"/nfs/turbo/bakulski1/People/johndou/Fragile_Families/ProcessedFiles/jd"
+outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output/"
+codedir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/"
+
+#read in data 
+pheno<-readRDS(paste0(datadir, "OGData/fullpheno.rds"))
+FFmeta<-read.csv(paste0(datadir, "FFMetadata_v07.csv"))
+
+############################relabel data###################################
+
+#remove offending characters from variable labels in phenotype data
+attributes(pheno)$var.labels<-gsub("\xad|\x92", "'",  attributes(pheno)$var.labels)
+#adding in a variable/value label for id
+attributes(pheno)$var.labels[which(colnames(pheno)=='id')]='ID'
+attributes(pheno)$label.table[which(colnames(pheno)=='id')]=NA
+#label data using sj labelled - variable names
+pheno=set_label(pheno, attributes(pheno)$var.labels)
+#label data using sj labelled - value labels 
+pheno=set_labels(pheno, labels=attributes(pheno)$label.table)
+#identify categorical variables
+FF_factors=FFmeta%>%
+  filter(type%in% c('Binary', 'Ordered Categorical', 'Unordered Categorical'))%>%
+  pull(old_name)
+#recode binary, categorical variables so they have label names as values (nicer for table and plotting)
+pheno=pheno %>% dplyr::mutate(across(any_of(FF_factors), ~sjlabelled::as_label(., drop.levels=TRUE)))
+
+#select variables
+idvars<-c("id", "ff_id")
[…]

### 2021-09-27 Scripts/02_make_methyl_summary.R

@@ -1,0 +1,33 @@
+#libraries 
+library(tidyr)
+library(stringr)
+library(dplyr)
+library(purrr)
+library(tidyselect)
+library(broom)
+library(compareGroups)
+library(kableExtra)
+library(haven)
+library(sjlabelled)
+library(matchmaker)
+library(EpiDISH)
+library("PerformanceAnalytics")
+library(DescTools)
+library(ggplot2)
+library(ggpubr)
+library(GGally)
+library(ggcorrplot)
+library(ggalluvial)
+library(cowplot)
+library(DiagrammeR)
+library(sva)
+
+#directory based coding
+datadir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
+john_datadir<-"/nfs/turbo/bakulski1/People/johndou/Fragile_Families/ProcessedFiles/jd"
+outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output/"
+codedir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/"
+
+#
+#polymethylation scores, global methylation data, clocks, 
+#cell-type proportions, surrogate variable analysis

### 2021-09-27 Scripts/03_summary_methyl_models.R

@@ -19,6 +19,7 @@ library(ggpubr)
 library(cowplot)
 library(geepack)
 library(lme4)
+library(pROC)
 library(lmerTest)
 library(stringr)
 library(sva)
@@ -42,25 +43,28 @@ modeldata$SmkAtVisitPastmonth<-factor(modeldata$SmkAtVisitPastmonth,
                                                "Pack or more a day", "Missing"))
 modeldata$slide<-factor(modeldata$slide)
 modeldata$Sample_Plate<-factor(modeldata$Sample_Plate)
-#sv data - may change this depending on conversation w/ kelly
-load(paste0(datadir, 'svNone.Rds'))
+
+#sv data - no variance filter, whole subset
+load(paste0(data.dir, '/svNone.Rds'))
 is_sv=function(x) str_replace_all(x, " ", "")
 sv.vNone=sv.vNone%>%rename_with(is_sv, contains('sv'))
 modeldata=left_join(modeldata, sv.vNone)
 
 #################################Set variables for modeling
 #outcomes & outcome_labels
-y_vector<-c("globalmethylation", "pediatric", "grim", "anynewborn_center", "SSnewbornCT_center", "SSolder_center")
+y_vector<-c("globalmethylation", "pediatric", "grim", "anynewborn_center", "SSnewbornCT_center", "SSolder_center", "cg05575921")
 outcome_labels=c('Global methylation', 'Pediatric clock', 'GRIM clock',  
                  'Any smoking polymethylation score (newborns)', 
                  'Sustained smoking polymethylation score (newborns, cell-type controlled)', 
-                 'Sustained smoking polymethylation score (older children)')
+                 'Sustained smoking polymethylation score (older children)', 
+                 'AHHR: cg05575921')
 names(outcome_labels)=y_vector
 
 #model variables (excepting PCs, add in individually for global & ancestry specific models)
-base_model_vars<-"~smkPreg_binary+cm1bsex+cm1inpov+ChildAgeComposite+Epi+IC+Sample_Plate"
+base_model_vars<-"~smkPreg_binary+cm1bsex+cm1inpov+Epi+IC+Sample_Plate"
 prenatal_model_vars<-paste0(base_model_vars, "+m1g2_YesNoPreg+m1g3_YesNoPreg")
 secondhand_model_vars<-paste0(prenatal_model_vars, '+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth')
[…]

### 2021-09-27 Reports/FFModeling.Rmd
Reports/*.Rmdreport on ROC code

@@
 ```{r bivariate table stratified by visit and ancestry child visit}
-export2md(strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+pediatric+levine+anynewborn_center+SSnewbornCT_center+SSolder_center+Epi+Fib+IC+ChildAgeComposite+cm1inpov+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth, data=modeldata%>%filter(childteen=='C'))), 'ancestry'))
+export2md(strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+pediatric+levine+anynewborn_center+SSnewbornCTw_center+SSolder_center+Epi+Fib+IC+ChildAgeComposite+cm1inpov+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth, data=modeldata%>%filter(childteen=='C'))), 'ancestry'))
 ```

### 2021-09-27 Reports/ROC_report.Rmd
Reports/*.Rmdreport on ROC code

---
title: "ROC curve analysis"
author: "F blostein"
date: "9/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source()
```

```{r}
#investigate adding in covariates and effect on ROC curve 
#question for kelly --> only use methylation or use methylation + base model
#don't think it's reasonable to use prenatal exposures + postantal smoking 
# could we predict peoples prenatal smoke exposure when we don't have info on maternal smoking/behaviors

global_roc%>%dplyr::select(covariates, methylation, childteen, test)%>%unnest(test) %>% 
  filter(str_detect(methylation, 'SSnewbornCT|cg055|global|pediatric'))%>%
  ggplot(aes(x=spec, y=sens, color=covariates))+geom_line()+
  facet_grid(childteen~methylation, labeller=labeller(childteen=age_labels, methylation=outcome_labels))+
  theme_classic()+theme(legend.position = 'bottom')+guides(color=guide_legend(nrow=6,byrow=TRUE))+
  geom_abline(intercept=1, slope=-1, color='black')

#is prediction better when using PMS? 
global_roc%>%dplyr::select(predictors, covariates, methylation, childteen, test)%>%
  mutate(modeltype=ifelse(str_detect(predictors, '\\+'), covariates, 'No covariates'))%>%
  unnest(test) %>% 
  filter(str_detect(methylation, 'SSnewbornCT|cg055|global|pediatric') &
           !str_detect(covariates, 'm1g2|sv'))%>%
  ggplot(aes(x=spec, y=sens, color=methylation, linetype=modeltype))+geom_line()+
  facet_grid(childteen~., labeller=labeller(childteen=age_labels))+
  scale_color_discrete(labels=outcome_labels)+
  theme_classic()+theme(legend.position = 'right')+guides(color=guide_legend(nrow=4,byrow=TRUE))+
  geom_abline(intercept=1, slope=-1, color='black')

global_roc%>%dplyr::select(predictors, covariates, methylation, childteen, test)%>%
  mutate(modeltype=ifelse(str_detect(predictors, '\\+'), covariates, 'No covariates'))%>%
  unnest(test) %>% 
  filter(str_detect(methylation, 'SSnewbornCT|cg055|global|pediatric') &
           covariates=='')%>%
  ggplot(aes(x=spec, y=sens, color=methylation, linetype=modeltype))+geom_line()+
  facet_grid(childteen~., labeller=labeller(childteen=age_labels))+
  scale_color_discrete(labels=outcome_labels)+
  theme_classic()+theme(legend.position = 'right')+guides(color=guide_legend(nrow=4,byrow=TRUE))+
  geom_abline(intercept=1, slope=-1, color='black')

#is best methylation predictor consistent across ancestries
local_roc %>% dplyr::select(covariates, predictors, childteen, ancestry, test)%>%unnest(test)%>% 
  filter(str_detect(predictors, 'SSnewbornCT|cg055|global|pediatric') & covariates=='')%>%
  ggplot(aes(x=spec, y=sens, color=predictors))+
  facet_grid(childteen~ancestry)+geom_line()

#things get weird when using the base model vars
local_roc %>% dplyr::select(covariates, predictors, childteen, ancestry, test)%>%unnest(test)%>% 
  filter(str_detect(predictors, 'SSnewbornCT|cg055|global|pediatric') & covariates!='')%>%
  ggplot(aes(x=spec, y=sens, color=predictors))+
  facet_grid(childteen~ancestry)+geom_line()

#investigated change in ROC when adding in each base model var individually
local_roc_test%>% dplyr::select(covariates, predictors, childteen, ancestry, test)%>%unnest(test)%>% 
  filter(str_detect(predictors, 'global'))%>%
  ggplot(aes(x=spec, y=sens, color=predictors))+
  facet_grid(childteen~ancestry)+geom_line()
#big jump in ROC curve for European & hispanic - batch variable 
#big jump in ROC curve for European - poverty and hispanic - local pcs
modeldata %>%filter(ancestry=='Hispanic ancestry' & childteen=='T')%>%xtabs(~Sample_Plate+smkPreg_binary, data=.)

#is prediction improved when using two time points?

global_roc%>%dplyr::select(covariates, methylation, childteen, test)%>%unnest(test) %>% 
  filter(str_detect(methylation, 'SSnewbornCT|cg055|global|pediatric') & covariates=='')%>%
  rbind(long_global_roc%>% mutate(childteen='C+T', methylation=gsub('_C.*', '', methylation))%>%
          dplyr::select(covariates, methylation, childteen, test)%>%
          unnest(test) %>% 
          filter(str_detect(methylation, 'SSnewbornCT|cg055|global|pediatric') & covariates==''))%>%
  ggplot(aes(x=spec, y=sens, color=childteen))+geom_line()+
  facet_wrap(~methylation)+
  scale_color_discrete(labels=outcome_labels)+
  theme_classic()+theme(legend.position = 'right')+guides(color=guide_legend(nrow=4,byrow=TRUE))+
  geom_abline(intercept=1, slope=-1, color='black')

global_roc%>%dplyr::select(covariates, methylation, childteen, test)%>%unnest(test) %>% 
  filter(str_detect(methylation, 'SSnewbornCT|cg055') & covariates=='')%>%
  rbind(long_global_roc%>% mutate(childteen='C+T', methylation=gsub('_C.*', '', methylation))%>%
          dplyr::select(covariates, methylation, childteen, test)%>%
          unnest(test) %>% 
          filter(str_detect(methylation, 'SSnewbornCT|cg055') & covariates==''))%>%
  ggplot(aes(x=spec, y=sens, color=childteen))+geom_line()+
  facet_wrap(~methylation)+
  scale_color_discrete(labels=outcome_labels)+
  theme_classic()+theme(legend.position = 'right')+guides(color=guide_legend(nrow=4,byrow=TRUE))+
  geom_abline(intercept=1, slope=-1, color='black')

```

### 2021-09-27 Reports/SVAReport.Rmd
Reports/*.Rmdreport on ROC code

@@
 date: "9/8/2021"
-output: html_document
+output:
+    html_document:
+      code_folding: show
+      highlight: tango
+      number_sections: yes
+      theme: sandstone
+      toc: yes
+      toc_depth: 2
+      toc_float: yes
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(warning = FALSE)
+knitr::opts_chunk$set(message = FALSE)
+
 #libraries
@@
 John said this is high in his experience.
-Increasing the number of features to 500 halves the number of surrogate variables detected to 20, after that there is a minor increase in SVs.
+Increasing the number of features to 5000 halves the number of surrogate variables detected to 20, after that there is a minor increase in SVs.
+In the full matrix SV analysis (using all the CpG sites), there are 36 syrrigate variables.
 Using the leek method, I get 0 surrogate variables (except when using a variance filter of 50k, at which point I get 2 surrogate variables)
@@
 ```{r heatmap function}
-bring.the.heatmap.sv <- function(pd,covar,folder){
+bring.the.heatmap.sv <- function(pd,covar,folder, name='Associations Between SVs\n and Given Covariates'){
   library(gplots)
@@
   heatmap.2(correlations,Rowv=FALSE,Colv=FALSE,cexRow=1,cexCol=1,dendrogram='none',density.info='none',trace='none',
-            col=color.gradient(20), key.xlab='p-value', main='Associations Between SVs\n and Given Covariates',
+            col=color.gradient(20), key.xlab='p-value', main=name,
             #lmat=rbind(c(0,3,3,3),c(2,0,4,0),c(0,1,1,1)), lhei=c(0.5,1.5,3), lwid=c(0.12,0.145,0.4,0.335),
@@
 
-```{r plot heatmap, fig.width=20, fig.height=12}
+```{r plot heatmap}
 bring.the.heatmap.sv(my.pd, c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'ancestry', 'cm1bsex', 'Epi', 'Fib', 'IC', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))
@@
 ```{r, echo=FALSE,results='hide',fig.keep='all'}
-map(.x=pd.ancestry$pd, .f=~(bring.the.heatmap.sv(as.data.frame(.x), covar=c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'cm1bsex', 'Epi', 'Fib', 'IC', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))))
+map2(.x=pd.ancestry$pd, .y=paste('SV and covariate \n associations in', pd.ancestry$ancestry),  .f=~(bring.the.heatmap.sv(as.data.frame(.x), covar=c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'cm1bsex', 'Epi', 'Fib', 'IC', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'), name=.y)))
 ```
@@
 
+### Global w/ full matrix SVs
+
 ```{r}
@@
   geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
-  facet_grid(.~outcome, scales='free_x', labeller = labeller(outcome=label_wrap_gen(15)))+
+  facet_wrap(.~outcome, scales='free_x', labeller = labeller(outcome=label_wrap_gen(15)), ncol=3)+
   geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
   theme_classic()+
-  theme(text = element_text(size=16))+theme(legend.position = 'bottom')+
+  theme(text = element_text(size=12))+theme(legend.position = 'bottom')+
+  scale_color_discrete('', limits=c('Age 9', 'Age 15'))+
+  scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')
+```
+
+### Local (ancestry-specific)
+
+```{r}
+local_models%>%unnest(tidied)%>%
+  filter(term=='smkPreg_binaryYes')%>%
+  filter(outcome%in% c('Global methylation', 'Pediatric clock', 'GRIM clock'))%>%
+  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=predictors))+
+  geom_point(position = position_dodge(width = 1))+
+  geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
+  facet_grid(ancestry~outcome, scales='free_x', labeller = labeller(outcome=label_wrap_gen(15)))+
+  geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
+  theme_classic()+
+  theme(text = element_text(size=12))+theme(legend.position = 'bottom')+
   scale_color_discrete('', limits=c('Age 9', 'Age 15'))+
@@
 
+local_models%>%unnest(tidied)%>%
+  filter(term=='smkPreg_binaryYes')%>%
+  filter(str_detect(outcome, 'score'))%>%
+  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=predictors))+
+  geom_point(position = position_dodge(width = 1))+
+  geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
+  facet_grid(ancestry~outcome, scales='free_x', labeller = labeller(outcome=label_wrap_gen(15)))+
+  geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
+  theme_classic()+
+  theme(text = element_text(size=12))+theme(legend.position = 'bottom')+
+  scale_color_discrete('', limits=c('Age 9', 'Age 15'))+
+  scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')
 ```
+
 ## Remaining questions

### 2021-09-27 Reports/Poster/IGES_poster_vertical.Rmd
updates to rmd poster

--- 2021-09-14 Reports/Poster/IGES_poster_vertical.Rmd
@@
 }
+.mybreak {
+   break-before: column;
+}
 </style>
 ```
+
 ```{r, include=FALSE}
@@
 
-```{r, joubert_fig, fig.cap='Volcano plot showing associations between CpG methylation in cord blood and prenatal maternal smoking exposure in @Joubert:2016ue'}
+```{r, joubertfigure, fig.cap='Volcano plot showing associations between CpG methylation in cord blood and prenatal maternal smoking exposure in @Joubert:2016ue'}
 knitr::include_graphics('Images/Joubert_volcano.jpeg')
@@
 
-# Methods
+# Methods {.mybreak}
 
@@
 
-```{r, ff_fig, fig.cap='Study design and timeline of exposure, outcome and covariate assessment in the Fragile Families and Child Wellbeing cohort'}
+```{r, FFfigure, fig.cap='Study design and timeline of exposure, outcome and covariate assessment in the Fragile Families and Child Wellbeing cohort'}
 knitr::include_graphics('Images/FF_studydesign.png')
@@
 
-Here you may have some figures to show off, bellow I have made a scatterplot with the infamous Iris dataset and I can even reference to the figure automatically like this, `Figure \@ref(fig:irisfigure)`, Figure \@ref(fig:irisfigure).
+```{r data read and set up}
+source(paste0(codedir, 'Scripts/', '03_summary_methyl_models.R'))
+```
 
-```{r data read and set up}
+**Methylation summary metrics by prenatal maternal smoking and child age**
+
+Table \@table1 shows the distributions of methylation summary measures ancestry (inferred from ellipses of genetic principal components) by prenatal maternal smoke exposure (yes=exposed, no=unexposed) and visit (C=age 9 visit, T=age 15 visit).
+
+```{r, table1}
+export2md(strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+pediatric+SSnewbornCT_center+ancestry, data=modeldata)), 'childteen'))
+```
+
+## Polymethylation scores are correlated and stable between age 9 and age 15
+
+```{r}
+outcome_labels=c('Global methylation', 'Pediatric clock', 'Polymethylation score \n (newborn, cell-type controlled, mean centered)')
+names(outcome_labels)=c('globalmethylation', 'pediatric', 'SSnewbornCT_center')
+modeldata%>%dplyr::select(idnum, childteen, SSnewbornCT_center, globalmethylation, pediatric)%>%pivot_longer(cols=c(globalmethylation, pediatric, SSnewbornCT_center), names_to='Methylation summary measure type', values_to='Methylation summary measure')%>%pivot_wider(names_from=childteen, values_from=`Methylation summary measure`)%>%ggplot(aes(x=C, y=T))+geom_point()+stat_cor()+geom_smooth(method=lm)+theme_classic()+scale_x_continuous('Age 9 score')+scale_y_continuous('Age 15 score')+facet_wrap(~`Methylation summary measure type`, scales='free', labeller = labeller(`Methylation summary measure type`=outcome_labels))
+```
+
+```{r}
+modeldata %>% pivot_longer(cols=c('globalmethylation', 'pediatric', 'SSnewbornCT_center'), names_to='measuretype', values_to='Methylation summary measure')%>%ggplot(aes(x=childteen, y=`Methylation summary measure`, group=idnum))+geom_point()+geom_line(alpha=0.1, color='grey')+geom_smooth(aes(group=1), method='lm', fill='blue')+facet_wrap(~measuretype, scales='free_y', labeller = labeller(measuretype=outcome_labels))+theme_classic()+scale_x_discrete('', labels=c('Age 9', 'Age 15'))
+```
+
+## Polymethylation scores are a specific, persistant and consistent marker of prenatal maternal smoke exposure
+
+```{r, forestplot, fig.cap='Forest plot from ancestry-specific regressions of prenatal maternal smoking against various methylation summary metrics.
+All models control for child sex, mother income to poverty ratio at birth, epithelial and immune cell proportions (estimated from methylation data), maternal other drug and other alcohol use during pregnancy, batch, genetic ancestry (1st two principal components of genetic data), postnatal smoke exposure (any postnatal maternal smoking and maternal packs per day in month prior to visit) ', out.width="80%"}
+my_outcomes=c('Global methylation', 'Pediatric clock', 'Sustained smoking polymethylation score (newborns, cell-type controlled)')
+
+data_text=local_models%>%
+  mutate(n=map_dbl(data, nrow), n=paste0('n=', n), x=-0.01)%>%
+  unnest(tidied)%>%
+  filter(outcome=='Global methylation' & term=='smkPreg_binaryYes'& predictors=='secondhand smoke exposure model')
+
+local_models%>%
+  mutate(n=map_dbl(data, nrow), n=paste0('n=', n))%>%
+  unnest(tidied)%>%
+  filter(term=='smkPreg_binaryYes' & predictors=='secondhand smoke exposure model')%>%
+  filter(outcome%in% my_outcomes)%>%
+  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=ancestry))+
+  geom_point(position = position_dodge(width = 1))+
+  geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
+  facet_grid(.~outcome, scales='free_x', labeller = labeller(outcome=label_wrap_gen(15)))+
+  geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
+  geom_text(data=data_text, aes(x=x,label=n), position=position_dodge(width = 1), hjust=0.5)+
+  theme_classic()+
+  theme(text = element_text(size=16))+theme(legend.position = 'bottom')+
+  scale_color_discrete('', limits=c('Age 9', 'Age 15'))+
+  scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
+  scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry'))
 
@@
 
-```{r, irisfigure, fig.cap='Here is a caption for the figure.
-This can be added by using the "fig.cap" option in the r code chunk options, see this [link](https://yihui.name/knitr/options/#plots) from the legend himself, [Yihui Xie](https://twitter.com/xieyihui).', out.width="80%"}
-
-```
-
-How about a neat table of data? See, Table \@ref(tab:iristable):
-
-```{r, iristable}
-knitr::kable(
-  iris[1:15,1:5], format = "html",
-  caption = "A table made with the **knitr::kable** function.",
-  align = "c", col.names = c("Sepal <br> Length",
-                             "Sepal <br> Width",
-                             "Petal <br> Length",
-                             "Petal <br> Width",
-                             "Species"),
-  escape = FALSE)
-```
+# Conclusions
 

### 2021-09-30 Reports/FFCleaning2.Rmd
September 30 updated pipeline and results

@@
 ```
+
 ```{r celltypes by visit spaghetti}

### 2021-09-30 Reports/ROC_report.Rmd
September 30 updated pipeline and results

@@
   geom_abline(intercept=1, slope=-1, color='black')
+```
+
+```{r}
 
@@
 
+```
+
+```{r}
 #is best methylation predictor consistent across ancestries
@@
   facet_grid(childteen~ancestry)+geom_line()
+```
 
+```{r}
 #investigated change in ROC when adding in each base model var individually
@@
 
+```
+
+```{r}
 #is prediction improved when using two time points?

### 2021-09-30 Reports/Results.Rmd
September 30 updated pipeline and results

---
title: "Polymethylation scores results"
author: "F blostein"
date: "9/30/2021"
output:
  html_document:
    code_folding: show
    highlight: tango
    number_sections: no
    theme: sandstone
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message=FALSE)

library(ggplot2)
library(kableExtra)
library(ggpubr)
library(tidyr)
library(stringr)
library(dplyr)
library(compareGroups)
library(kableExtra)
library(haven)
library(sjlabelled)
library(matchmaker)
library(EpiDISH)
library("PerformanceAnalytics")
library(DescTools)
library(purrr)
library(broom)
library(broom.mixed)
library(ggplot2)
library(multcomp)
library(ggpubr)
library(cowplot)
library(geepack)
library(lme4)
library(pROC)
library(lmerTest)
library(stringr)
library(sva)
```

```{r temp read n data}
codedir=''
```

```{r future run scripts, cache=T}

source('/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/Scripts/03_summary_methyl_models.R')
```

```{r plotting variables}
theme_set(theme_classic())
```

```{r}
n_total=modeldata %>% dplyr::select(id)%>%unique()%>%nrow()
n_obs=modeldata%>%nrow()

my_round=function(x, d){
  y=c()
  d_orig=d
  for(i in 1:length(x)){
    d=d_orig
    while(round(x[i], d)==0){
    d=d+1
    }
    y[i]=round(x[i], d)
  }
  return(y)
}
```

# Coding tasks

- X replace epidish cell estimation with ewastools Middleton

- check with jonah about final sample size and exclusions
- determine what to do w/ 1 sample missing clock data (related to above)
- rerun clocks i plan on using on all samples

- make this an entire story line, combined supplemental and non supplemental

# Target journal and deadline

[Clinical epigenetics](https://www.biomedcentral.com/collections/epibio): Special issue on biomarkers; due date December 1st
 >Epigenetic biomarkers have significantly contributed to improved understanding of the origins and progression of disease.  Moreover, increasing evidence shows that epigenetic biomarkers have potential for personalized medicine. However, routine use of epigenetic biomarkers in in-vitro diagnostics (IVD) is lagging behind the discoveries of new biomarkers. In this thematic series of Clinical Epigenetics, we discuss current progress in development of epigenetic biomarkers for clinical use in common complex diseases of ageing and welcome both original research and reviews.
The sections of the collection include:
  - Types of epigenetic biomarkers and readiness for IVD use
    - DNA methylation as biomarker
    - Histone modification as biomarker
    - Upcoming field of circulating free nucleosomes as biomarker
    - Progress towards clinical use of biomarker
  - *Biomarkers of environmental exposures and causality of the biomarker*
  - Disease predisposition biomarkers
  - Biomarkers used for disease detection
  - Clinical disease prevention and management: predictive and prognostic
  - Monitoring of chronic disease
  - *Single versus multiple biomarkers and risk scores*
  - Major stages of clinical validation of the biomarker
  - Biomarker detection technologies suitable for IVD use and significance of bioinformatics
  - Ethics and regulatory aspects of IVD epigenetic biomarkers

# Figure 1 - CONSORT Diagram

# Supplemental Table 1 - Comparison of 450K subsample to Fragile Families cohort

# Supplemental Figure 1 - Cell type proportion estimation

# Supplemental Figure 2 - Choice of beta methylation transformation

# Supplemental Figure 3 - Distribution and correlation of methylation summary measures

#Supplemental Figure 4 - Change in methylation summary measures between visits

# Figure 2: Polymethylation scores are correlated and stable between age 9 and age 15 in a diverse sample of `r n_total`US children

```{r change in scores time, fig.width=12}

outcome_labels=c('Global methylation', 'Pediatric clock', 'AHHR: cg05575921', 'Polymethylation score \n (newborn, mean centered)', 'Polymethylation score  \n (older children, mean centered')
names(outcome_labels)=c('globalmethylation', 'pediatric', 'cg05575921', 'SSnewbornCT_center','SSolder_center')

fig2a=modeldata%>%
  dplyr::select(idnum, childteen, SSolder_center, SSnewbornCT_center, 
                cg05575921, globalmethylation, pediatric)%>%
  pivot_longer(cols=c(globalmethylation, pediatric, SSnewbornCT_center, SSolder_center, cg05575921),
               names_to='Methylation summary measure type',
               values_to='Methylation summary measure')%>%
  pivot_wider(names_from=childteen, 
              values_from=`Methylation summary measure`)%>%
  mutate(meth_f=factor(`Methylation summary measure type`, 
                       levels=c('SSnewbornCT_center', 'SSolder_center', 
                                'cg05575921',  'globalmethylation', 'pediatric')))%>%
  ggplot(aes(x=C, y=T))+geom_point()+stat_cor()+geom_smooth(method=lm)+theme_classic()+
  scale_x_continuous('Age 9 score')+scale_y_continuous('Age 15 score')+
  facet_wrap(~meth_f, scales='free', labeller = labeller(meth_f=outcome_labels), nrow=1)

fig2b=modeldata%>%
  dplyr::select(idnum, childteen, SSolder_center, SSnewbornCT_center, 
                cg05575921, globalmethylation, pediatric)%>%
  pivot_longer(cols=c(globalmethylation, pediatric, SSnewbornCT_center, SSolder_center, cg05575921),
               names_to='Methylation summary measure type', 
               values_to='Methylation summary measure')%>%
  pivot_wider(names_from=childteen,
              values_from=`Methylation summary measure`)%>%
  mutate(meth_f=factor(`Methylation summary measure type`, 
                       levels=c('SSnewbornCT_center', 'SSolder_center', '
                                cg05575921',  'globalmethylation', 'pediatric')),
         diff=T-C)%>%
  ggplot(aes(x=diff))+geom_histogram()+
  facet_wrap(~meth_f, scales='free_x', labeller = labeller(meth_f=outcome_labels), nrow=1)+
  theme_classic()+xlab('Difference in measure (Age 15- Age 9)')+
  geom_vline(aes(xintercept=0), color='red', linetype=2)+geom_rug()

plot_grid(fig2a, fig2b, nrow=2, align = 'v', axis = 'lr')

```

# Table 1: Bivariate associations between prenatal maternal smoking, DNA methylation summary measures and important covariates among a diverse sample of

```{r temp relabel varables}
set_label(modeldata$cg05575921)<-'AHHR: cg05575921'
set_label(modeldata$SmkAtVisitPastmonth)<-'Past month maternal/primary caregiver smokin (pack/day)'
set_label(modeldata$PostnatalMaternalSmokingAny)<-'Any maternal postnatal smoking when child 1 and 5 years of age'
modeldata$childteen=factor(case_when(modeldata$childteen=='C'~'Age 9', 
                              modeldata$childteen=='T'~'Age 15'), levels=c('Age 9', 'Age 15'))
```

```{r bivariate table}
export2md(strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+pediatric+SSnewbornCT_center+cg05575921+Epi+Fib+IC+ChildAgeComposite+cm1inpov+ancestry+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth, data=modeldata)), 'childteen'))
```

#Supplemental Table XX - Bivariate associations between prenatal maternal smoking and covariates in ancestry strata

# Supplemental Table XX: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable models among `r n_total` US children

```{r}
my_main_outcomes=c('AHHR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)', 'Sustained smoking polymethylation score (older children)', 'Pediatric clock', 'Global methylation')

my_names=c('Methylation measure', paste0(c(paste(c('Age 9 model \n n=', 'Age 15 model \n n='), modeldata%>%group_by(childteen)%>%summarise(n=n())%>%pull(n)), paste0('Longitudinal models <br> n=', n_total)), '<br> Beta (95% CI); p-value'))

tb2=rbind(global_models, longitudinal_global_models %>% mutate(childteen='Longitudinal'))%>%filter(predictors=='secondhand smoke exposure model')%>%filter(outcome %in% my_main_outcomes)%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%mutate(estimate=paste0(my_round(estimate, 2), ' (', my_round(conf.low, 2), ', ', my_round(conf.high, 2), '); ', signif(p.value, 2)))%>%dplyr::select(childteen, outcome, estimate)%>%pivot_wider(names_from=childteen, values_from=c(estimate))
names(tb2)=my_names

tb2%>%kable(., escape=F)%>%kable_minimal()%>%
  kableExtra::scroll_box(width = "100%")%>% footnote(general = "All models adjusted for:child sex, mother income to poverty ratio at birth, epithelial and immune cell proportions (estimated from methylation data), maternal other drug and other alcohol use during pregnancy, batch of array, genetic ancestry (1st two principal components of genetic data), postnatal smoke exposure (any postnatal maternal smoking and maternal packs per day in month prior to visit).")%>%kable_styling(font_size = 10)
```

```{r, width=11}
#alternatively 
my_names=c('Methylation measure', paste0(c(paste0(paste0('Unstratified <br> n=', n_total), footnote_marker_alphabet(1, 'html')), paste0(modeldata%>%group_by(ancestry)%>%dplyr::select(ancestry, idnum)%>%unique()%>%summarise(n=n())%>%transmute(label=paste0(ancestry, ' <br> n=', n))%>%pull(label), footnote_marker_alphabet(2, 'html'))), '<br> Beta (95% CI); p-value'))

tb2=rbind(longitudinal_global_models %>% mutate(ancestry='Unstratified'), longitudinal_local_models%>%dplyr::select(-data))%>%filter(predictors=='secondhand smoke exposure model')%>%filter(outcome %in% my_main_outcomes)%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%mutate(estimate=paste0(my_round(estimate, 2), ' (', my_round(conf.low, 2), ', ', my_round(conf.high, 2), ') p=', signif(p.value, 2)))%>%dplyr::select(ancestry, outcome, estimate)%>%pivot_wider(names_from=ancestry, values_from=c(estimate))

names(tb2)=my_names

tb2%>%kable(., escape=F)%>%kable_minimal()%>%
  kableExtra::scroll_box(width = "100%")%>% footnote(alphabet =  c("Model adjusted for:child sex, mother income to poverty ratio at birth, epithelial and immune cell proportions (estimated from methylation data), maternal other drug and other alcohol use during pregnancy, batch of array, genetic ancestry (1st two principal components of genetic data), postnatal smoke exposure (any postnatal maternal smoking and maternal packs per day in month prior to visit).", "Models adjusted as in (a) but principal components of genetic ancestry rerun within ancestry strata and first two components from these stratified principal components analyses used"))%>%kable_styling(font_size = 10)

```

# Figure 3 - Polymethylation scores are associated with prenatal maternal smoke exposure consistently across age and ancestry group in a sample of `r n_total` US children

```{r, fig.width=12}
my_outcomes=c('Global methylation', 'Pediatric clock', 'AHHR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)')

outcome_labels=c('Global methylation', 'Pediatric clock', 'AHHR: cg05575921',  'Polymethylation score \n (newborn, mean centered)')

all_models=rbind(local_models, global_models%>%mutate(ancestry='Unstratified'))

data_text=all_models%>%
  mutate(n=map_dbl(data, nrow), n=paste0('n=', n), x=-0.25)%>%
  unnest(tidied)%>%
  filter(outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & term=='smkPreg_binaryYes'& predictors=='secondhand smoke exposure model')%>%mutate(outcome_f=outcome)

names(outcome_labels)=my_outcomes

all_models%>%
  mutate(n=map_dbl(data, nrow), n=paste0('n=', n))%>%
  unnest(tidied)%>%
  filter(term=='smkPreg_binaryYes' & predictors=='secondhand smoke exposure model')%>%
  filter(outcome%in% my_outcomes)%>%
    mutate(outcome_f=factor(outcome, levels=rev(my_outcomes)))%>%
  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=ancestry))+
  geom_point(position = position_dodge(width = 1))+
  geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
  facet_grid(.~outcome_f, scales='free_x', labeller = labeller(outcome_f=outcome_labels))+
  geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
  geom_text(data=data_text, aes(x=x,label=n), position=position_dodge(width = 1), hjust=0.05)+
  theme_classic()+
  theme(text = element_text(size=16))+theme(legend.position = 'bottom')+
  scale_color_discrete('', limits=c('Age 9', 'Age 15'))+
  scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
  scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))
```

# Supplemental Figure XX: Surrogate variable associations with known covariates

# Supplemental Table XX: Surrogate variable models

# Figure 4 - Polymethylation scores more accurately delineate prenatal smoke exposure status than non-targeted DNA methylation summary measures and *a priori* CpGs

```{r, fig.width=10}

names(my_outcomes)=c( 'globalmethylation', 'pediatric','cg05575921', 'SSnewbornCT_center')
my_outcomes=c('Global methylation', 'Pediatric clock', "AHHR: cg05575921", 'Polymethylation score \n (newborn, mean centered)')

fig4a= global_roc%>%
  dplyr::select(predictors, covariates, methylation, childteen, test)%>%
  mutate(modeltype=ifelse(str_detect(predictors, '\\+'), covariates, 'No covariates'))%>%
  unnest(test) %>% 
  filter(str_detect(methylation, 'SSnewbornCT|cg055|global|pediatric') &
           covariates=='')%>%
  ggplot(aes(x=spec, y=sens, color=methylation))+geom_line()+
  facet_wrap(childteen~., labeller=labeller(childteen=age_labels), nrow=1)+
  scale_color_discrete('', labels=my_outcomes)+
  theme_classic()+theme(legend.position = 'right')+guides(color=guide_legend(nrow=2, byrow=TRUE))+
  geom_abline(intercept=1, slope=-1, color='black')+
  xlab('Specificity')+ylab('Sensitivity')+
  theme(legend.position='bottom')+ coord_fixed()

fig4b=global_roc%>%
  dplyr::select(covariates, methylation, childteen, test)%>%unnest(test) %>% 
  filter(str_detect(methylation, 'SSnewbornCT') 
         & covariates=='')%>%
  rbind(long_global_roc%>% 
          mutate(childteen='C+T', methylation=gsub('_C.*', '', methylation))%>%
          dplyr::select(covariates, methylation, childteen, test)%>%
          unnest(test) %>% 
          filter(str_detect(methylation, 'SSnewbornCT') 
                 & covariates==''))%>%
  ggplot(aes(x=spec, y=sens, color=childteen, linetype=childteen))+geom_line(size=2)+
  facet_wrap(~methylation, labeller = labeller(methylation=c('SSnewbornCT_center'='Polymethylation score, \n newborn mean centered')))+
  scale_color_discrete(labels=my_outcomes)+
  theme_classic()+theme(legend.position = 'right')+guides(color=guide_legend(nrow=2,byrow=TRUE))+
  geom_abline(intercept=1, slope=-1, color='black')+
  scale_color_grey('', labels=c('Age 9 only', 'Age 9 & 15', 'Age 15 only'))+
  scale_linetype('', labels=c('Age 9 only', 'Age 9 & 15', 'Age 15 only'))+
  xlab('Specificity')+ylab('Sensitivity')+
  theme(legend.position='bottom')+ coord_fixed()

cowplot::plot_grid(fig4a, fig4b, labels=c('A', 'B'), nrow=1, rel_widths=c(0.65, 0.35), align='h', axis='t')

```

add p value comparing base model to adding in curves

# Question for group: add EWAS results to this paper

- Arguments in favor: recent paper in Clinical Epigenetics included both 
- Arguments against: enough here for a compellling story; deadline of December 1st; submit and if asked for we have it available.

# Cross sectional EWAS results:

# Longitudinal EWAS results

# METAL meta-analysis of 450K + EPIC chip data

### 2021-09-30 Scripts/01_clean_phenotype_data.R

@@ -44,7 +44,8 @@ smokingvars<-c("m1g4", "f1g4", "f2j5", "f2j5a", "f2j7", "f2j7a", "f3j31", "f3j32
 prenataldrugusevar<-c("m1g2", "m1g3", "m1g5", "m1g6")
 FF_labeled=pheno %>%dplyr::select(any_of(c(idvars, demovars, smokingvars, prenataldrugusevar)))
 #apply NA codes
-my_na_codes<-c("-1 Refuse", "-2 Don't know", "-3 Missing", "-9 Not in wave")
+my_na_codes<-c(-10:-1, "-1 Refuse", "-2 Don't know", "-3 Missing", "-9 Not in wave", "-7 N/A")
+my_na_codes_numeric=
 na_codes <- function(x, ...) {
   x[x %in% c(...)] <- NA
   if(is.factor(x)==TRUE){x<-droplevels(x)}
@@ -138,14 +139,12 @@ myFF<-myFF %>%
       (childteen=='T' & as.numeric(substr(p6h77, 1, 1))>2) ~ "Pack or more a day", 
     (childteen=='C' & ( as.numeric(substr(m5g18, 1, 1))==1 | as.numeric(substr(n5f18, 1, 1))==1) ) | 
       (childteen=='T' & (as.numeric(substr(p6h77, 1, 1))%in% c(1, 2)))  ~ "Less than pack a day", 
-    (childteen=='C' & ( is.na(m5g18) & (is.na(n5f17) | n5f17=='-7 N/A') )) |
+    (childteen=='C' & ( is.na(m5g18) & (is.na(n5f18)) )) |
       (childteen=='T' & is.na(p6h77))~ 'Missing', 
-    childteen=='C' & (m5g18=='-6 Skip' & (n5f18 %in% c('-7 N/A', "-6 Skip")))|
+    childteen=='C' & ((m5g17=='2 no' & n5f17 %in% c(NA_character_, '2 no')) | (n5f17=='2 no' & m5g17%in%c('2 no', NA_character_)))|
     childteen=='T' & p6h77=='-6 Skip'~ 'No smoking'))
-
+#throws NA warning because of as.numeric in these statements - 
+if(myFF%>%filter(is.na(SmkAtVisitPastmonth))%>%nrow()!=0){stop('NA in SmkAtVisitPastmonth')}
 #######################complete case dataset###################################
 
 
-#######################multiple imputation dataset###################################
-#this should be its ownscript after joining in methylation data 
- 

### 2021-09-30 Scripts/02_make_methyl_summary.R

@@ -1,33 +1,73 @@
-#libraries 
-library(tidyr)
-library(stringr)
-library(dplyr)
-library(purrr)
-library(tidyselect)
-library(broom)
-library(compareGroups)
-library(kableExtra)
-library(haven)
-library(sjlabelled)
-library(matchmaker)
-library(EpiDISH)
-library("PerformanceAnalytics")
-library(DescTools)
-library(ggplot2)
-library(ggpubr)
-library(GGally)
-library(ggcorrplot)
-library(ggalluvial)
-library(cowplot)
-library(DiagrammeR)
-library(sva)
-
 #directory based coding
 datadir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
 john_datadir<-"/nfs/turbo/bakulski1/People/johndou/Fragile_Families/ProcessedFiles/jd"
 outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output/"
 codedir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/"
 
-#
-#polymethylation scores, global methylation data, clocks, 
-#cell-type proportions, surrogate variable analysis
+#######################################chose apriori cpgs, if any
+aprioriCG<-c("cg05575921", "cg04180046", "cg05549655", "cg14179389")
+
+################################read in methylation data
+betaqc<-readRDS(file=paste0(datadir, 'OGData/', "noob_filtered.rds"))
+
+###########################################cell type proportions 
+celltypes<-epidish(beta.m = betaqc, ref.m = centEpiFibIC.m, method = "RPC")
+#RPC method w/ centEpiFibIC
+estF_FF<-as.data.frame(celltypes$estF)%>%
+  tibble::rownames_to_column(var='MethID')%>%
+  rename_with(~paste0(., '_RPC'), .cols=c('Epi', 'Fib', 'IC'))
+#Middleton method
+estL_FF<-estimateLC(betaqc, 'saliva')%>%
+  mutate(MethID=colnames(betaqc))%>%
+  rename_with(~paste0(., '_saliva'), .cols=c('Leukocytes', 'Epithelial.cells'))
+
+cell_types=left_join(estF_FF, estL_FF)
+
+################################################global methylation
+globalmethy<-colMeans(betaqc, na.rm=T)
+globalmethdf<-as.data.frame(cbind("MethID"=names(globalmethy), "globalmethylation"=globalmethy))
+
+#################################################apriori cpgs
[…]

### 2021-09-30 Scripts/03_sva_variables.R

@@ -1,0 +1,43 @@
+#libraries
+library(sjlabelled)
+library(tidyr)
+library(dplyr)
+library(purrr)
+library(furrr)
+library(stringr)
+library(sva)
+
+#directory based programming
+data.dir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
+output.dir=paste0(data.dir, 'CreatedData/')
+
+#load in betaqc matrix
+betaqc<-readRDS(file=paste0(data.dir, 'OGData/', "noob_filtered.rds"))
+file.path(paste0(data.dir, 'CreatedData/'), list.files(paste0(data.dir, 'CreatedData/')))[(str_detect(file.path(paste0(data.dir, 'CreatedData/'), list.files(paste0(data.dir, 'CreatedData/'))), 'NoSmk'))]%>%map(load, envir=.GlobalEnv)
+
+pheno<-rbind(child_nosmoke, teen_nosmoke) # dataset with the most exclusion i.e. squaring dataset off for all models (excluding kids who smoke in figure)
+
+#make a matrix of model data 
+#filter beta matrix to only those in analytic subset 
+betaqc <-betaqc[, colnames(betaqc) %in% pheno$MethID]
+#check that dim of pheno and betaqc match 
+try(if(nrow(pheno)!=ncol(betaqc)) stop("# obs in betaqc & pheno do not match"))
+
+#want to make sure that the order of pheno and betaqc match
+pheno=pheno[match(colnames(betaqc), pheno$MethID), ]
+try(if(identical(colnames(betaqc), pheno$MethID)==FALSE) stop('STOP: Sample names do not match'))
+
+#make model matrix including variable of interest as factor variable & adjustment variables 
+mod=model.matrix(~as.factor(smkPreg_binary), data=pheno)
+
+#make null model containing only adjustment variables
+mod0=model.matrix(~1, data=pheno)
+
+#run with no variance filter 
+svobj=sva(betaqc, mod, mod0)
+sv.vNone=data.frame(svobj$sv); colnames(sv.vNone)=paste('sv', 1:ncol(sv.vNone))
+sv.vNone=cbind(pheno%>%select(id, idnum, MethID), sv.vNone)
+saveRDS(sv.vNone, file=paste0(output.dir, 'svNone.Rds'))
+print('Finished running SVA with no variance filter')
+
+

### 2021-09-30 Scripts/04_summary_methyl_models.R

@@ -1,0 +1,245 @@
+###################################Libraries
+library(tidyr)
+library(stringr)
+library(dplyr)
+library(compareGroups)
+library(kableExtra)
+library(haven)
+library(sjlabelled)
+library(matchmaker)
+library(EpiDISH)
+library("PerformanceAnalytics")
+library(DescTools)
+library(purrr)
+library(broom)
+library(broom.mixed)
+library(ggplot2)
+library(multcomp)
+library(ggpubr)
+library(cowplot)
+library(geepack)
+library(lme4)
+library(pROC)
+library(lmerTest)
+library(stringr)
+library(sva)
+#############
+#investigate why clocks have 1 fewer observation in child models
+##################################Directories & data
+data.dir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/CreatedData"
+outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output"
+#read data 
+file.path(data.dir, list.files(data.dir))[(str_detect(file.path(data.dir, list.files(data.dir)), 'NoSmk'))]%>%
+  map(load, envir=.GlobalEnv)
+#bind data
+modeldata<-rbind(child_nosmoke, teen_nosmoke)
+#label data 
+set_label(modeldata)<-get_label(child_nosmoke)
+#set as factors 
+modeldata$PostnatalMaternalSmokingAny<-factor(modeldata$PostnatalMaternalSmokingAny, 
+                                              levels=c("No maternal smoking at age 1 and age 5", 
+                                                       "Maternal smoking at age 1 or age 5"))
+modeldata$SmkAtVisitPastmonth<-factor(modeldata$SmkAtVisitPastmonth, 
+                                      levels=c("No smoking", 
+                                               "Less than pack a day", 
+                                               "Pack or more a day", "Missing"))
+modeldata$slide<-factor(modeldata$slide)
+modeldata$Sample_Plate<-factor(modeldata$Sample_Plate)
+
+#sv data - no variance filter, whole subset
+load(paste0(data.dir, '/svNone.Rds'))
+is_sv=function(x) str_replace_all(x, " ", "")
+sv.vNone=sv.vNone%>%rename_with(is_sv, contains('sv'))
+modeldata=left_join(modeldata, sv.vNone)
+
+#################################Set variables for modeling
+#outcomes & outcome_labels
[…]

### 2021-09-30 Reports/Poster/IGES_poster_vertical.Rmd
poster changes

@@
 ```{r}
-outcome_labels=c('Global methylation', 'Pediatric clock', 'Polymethylation score \n (newborn, cell-type controlled, mean centered)')
+outcome_labels=c('Global methylation', 'Pediatric clock', 'Polymethylation score \n (newborn, mean centered)')
 names(outcome_labels)=c('globalmethylation', 'pediatric', 'SSnewbornCT_center')
-modeldata%>%dplyr::select(idnum, childteen, SSnewbornCT_center, globalmethylation, pediatric)%>%pivot_longer(cols=c(globalmethylation, pediatric, SSnewbornCT_center), names_to='Methylation summary measure type', values_to='Methylation summary measure')%>%pivot_wider(names_from=childteen, values_from=`Methylation summary measure`)%>%ggplot(aes(x=C, y=T))+geom_point()+stat_cor()+geom_smooth(method=lm)+theme_classic()+scale_x_continuous('Age 9 score')+scale_y_continuous('Age 15 score')+facet_wrap(~`Methylation summary measure type`, scales='free', labeller = labeller(`Methylation summary measure type`=outcome_labels))
+
+modeldata%>%dplyr::select(idnum, childteen, SSnewbornCT_center, globalmethylation, pediatric)%>%pivot_longer(cols=c(globalmethylation, pediatric, SSnewbornCT_center), names_to='Methylation summary measure type', values_to='Methylation summary measure')%>%pivot_wider(names_from=childteen, values_from=`Methylation summary measure`)%>%mutate(meth_f=factor(`Methylation summary measure type`, levels=c('SSnewbornCT_center', 'globalmethylation', 'pediatric')))%>%ggplot(aes(x=C, y=T))+geom_point()+stat_cor()+geom_smooth(method=lm)+theme_classic()+scale_x_continuous('Age 9 score')+scale_y_continuous('Age 15 score')+facet_wrap(~meth_f, scales='free', labeller = labeller(meth_f=outcome_labels))+theme(text=element_text(size=16))
+
 ```
@@
 ```{r}
-modeldata %>% pivot_longer(cols=c('globalmethylation', 'pediatric', 'SSnewbornCT_center'), names_to='measuretype', values_to='Methylation summary measure')%>%ggplot(aes(x=childteen, y=`Methylation summary measure`, group=idnum))+geom_point()+geom_line(alpha=0.1, color='grey')+geom_smooth(aes(group=1), method='lm', fill='blue')+facet_wrap(~measuretype, scales='free_y', labeller = labeller(measuretype=outcome_labels))+theme_classic()+scale_x_discrete('', labels=c('Age 9', 'Age 15'))
+modeldata %>%group_by(idnum)%>%filter(n()>1)%>%pivot_longer(cols=c('globalmethylation', 'pediatric', 'SSnewbornCT_center'), names_to='measuretype', values_to='Methylation summary measure')%>%group_by(measuretype, idnum) %>%
+    mutate(slope = (`Methylation summary measure`[childteen=='T'] - `Methylation summary measure`[childteen=='C'])/(2-1))%>%filter(!is.na(`Methylation summary measure`))%>%
+  ggplot(aes(x=childteen, y=`Methylation summary measure`, group=idnum))+geom_point()+geom_line(alpha=0.1, aes(color=slope>0))+geom_smooth(aes(group=1), method='lm', fill='blue')+facet_wrap(~measuretype, scales='free_y', labeller = labeller(measuretype=outcome_labels))+theme_classic()+scale_x_discrete('', labels=c('Age 9', 'Age 15'))+scale_color_manual(name='', labels=c('slope<0', 'slope>0'), values=c('blue', 'red'))
+
+modeldata %>%group_by(idnum)%>%filter(n()>1)%>%pivot_longer(cols=c('globalmethylation', 'pediatric', 'SSnewbornCT_center'), names_to='measuretype', values_to='Methylation summary measure')%>%group_by(measuretype, idnum) %>%
+    mutate(slope = (`Methylation summary measure`[childteen=='T'] - `Methylation summary measure`[childteen=='C'])/(2-1), slope_factor=case_when(slope>0~'Positive', slope<=0~'Negative'))%>%filter(!is.na(`Methylation summary measure`))%>%ggplot(aes(x=slope))+geom_histogram()+geom_blank(aes(x=-slope)) + facet_wrap(~measuretype, scales='free', ncol=1)
 ```
@@
 
-data_text=local_models%>%
-  mutate(n=map_dbl(data, nrow), n=paste0('n=', n), x=-0.01)%>%
+all_models=rbind(local_models, global_models%>%mutate(ancestry='Unstratified'))
+
+data_text=all_models%>%
+  mutate(n=map_dbl(data, nrow), n=paste0('n=', n), x=-0.25)%>%
   unnest(tidied)%>%
-  filter(outcome=='Global methylation' & term=='smkPreg_binaryYes'& predictors=='secondhand smoke exposure model')
+  filter(outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & term=='smkPreg_binaryYes'& predictors=='secondhand smoke exposure model')%>%mutate(outcome_f=outcome)
 
-local_models%>%
+names(outcome_labels)=my_outcomes
+all_models%>%
   mutate(n=map_dbl(data, nrow), n=paste0('n=', n))%>%
@@
   filter(outcome%in% my_outcomes)%>%
+    mutate(outcome_f=factor(outcome, levels=rev(my_outcomes)))%>%
   ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=ancestry))+
@@
   geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
-  facet_grid(.~outcome, scales='free_x', labeller = labeller(outcome=label_wrap_gen(15)))+
+  facet_grid(.~outcome_f, scales='free_x', labeller = labeller(outcome_f=outcome_labels))+
   geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
-  geom_text(data=data_text, aes(x=x,label=n), position=position_dodge(width = 1), hjust=0.5)+
+  geom_text(data=data_text, aes(x=x,label=n), position=position_dodge(width = 1), hjust=0.05)+
   theme_classic()+
@@
   scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
-  scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry'))
+  scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))
 

### 2021-10-01 Papers/Polymethylation_scores/04_background.Rmd
worked on outlining introduction

# Background

P1: Maternal prenatal smoking is a public health concern but measuring prenatal maternal smoking behavior is challenging. Developing an in vitro biomarker for prenatal smoke exposure would be beneficial for personalized medicine.

P2: DNA methylation could provide a reliable and specific biomarker of prenatal maternal smoke exposure. Cordblood and peripheral blood DNA methylation has been significantly associated with prenatal maternal smoking. **Gap**: Most previous studies have used blood samples, but saliva is an easier to collect tissue. To our knowledge, no study has measured associations between prenatal maternal smoking and DNA methylation in saliva samples.

P3: DNA methylation has been associated with prenatal maternal smoking at various cross-sectional time points across the life course. Ideally, a biomarker of a prenatal exposure would persist across the life course. **Gap**: Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times.

P4: DNA methylation has been well replicated in primarily European ancestry cohorts. The performance of other biomarkers and risk scores varies across ancestry groups. To be applied clinically without exacerbating health disparities, it is important to understand how the perfomance of biomarkers vary across ancestry groups. **Gap**: Yet few (no?) studies have compared associations of prenatal maternal smoking across diffferent ancestry groups.

P5: Choice of how to construct a DNA methylation based biomarker.

P6: To translate DNA methylation to a *in vitro* biomarker for prenatal exposures, we must understand how performance [check with kelly if ok to use this word here now that have ROC plots] varies across tissue, age and ancestry. **To address these gaps we used XXXX longitudinal saliva samples from 9 and 14 year old children enrolled in a  diverse US birth cohort to test associations between DNA methylation summary measures and prenatal maternal smoke exposure.**  We found that maternal prenatal smoking was significantly associated with polymethylation scores. This association was persistent across time-point and consistent across ancestry strata. Polymethylation scores outperformed other non-target methylation summary measures, including global methylation and methylation clocks, as well as individual, *a priori* selected CpG sites. Our results suggest that polymethylation scores can serve as a persistent and consistent biomarker of prenatal smoke exposure.

### 2021-10-01 Scripts/01_clean_phenotype_data.R
repaired filtering of postnatal smoke exposure

@@ -90,9 +90,9 @@ FF_labeled=FF_labeled %>%
   mutate(across(c('m1g2', 'm1g3'), 
                 ~case_when(.x=='5 Never'~'No', is.na(.x)~ NA_character_, TRUE ~ 'Yes'), 
                 .names="{.col}_YesNoPreg"))%>%
-  mutate(PostnatalMaternalSmokingAny=case_when(m2j5=='1 Yes' | m4j18=='Yes'~ 'Maternal smoking at age 1 or age 5', 
-                                               is.na(m2j5) | is.na(m4j18) ~ 'Missing', 
-                                               m2j5 == '2 No' & m4j18=='2 No' ~ 'No maternal smoking at age 1 and 5'), 
+  mutate(PostnatalMaternalSmokingAny=case_when(m2j5 == '2 No' & m4j18=='2 No' ~ 'No maternal smoking at age 1 and 5',
+                                               m2j5=='1 Yes' | m4j18=='1 Yes'~ 'Maternal smoking at age 1 or age 5',
+                                               (is.na(m2j5) | m2j5=='2 No') & (is.na(m4j18) | m4j18=='2 No') ~ 'Missing'), 
          PostnatalMaternalSmokingDose=case_when(m2j5a %in% c("2 1pk/d", "3 1.5pk/d", "4 2pk/d", "5 >2pk/d") |
                                                 m4j19 %in% c("2 About pack/day", "3 Pack and half/day", "4 2 packs/day", "5 More 2 packs/day")~
                                                   'Pack/day or greater when child aged 1 or 5', 
@@ -143,8 +143,31 @@ myFF<-myFF %>%
       (childteen=='T' & is.na(p6h77))~ 'Missing', 
     childteen=='C' & ((m5g17=='2 no' & n5f17 %in% c(NA_character_, '2 no')) | (n5f17=='2 no' & m5g17%in%c('2 no', NA_character_)))|
     childteen=='T' & p6h77=='-6 Skip'~ 'No smoking'))
-#throws NA warning because of as.numeric in these statements - 
+#throws NA warning because of as.numeric in these statements - sow rite a check to make sure there aren't NAs here 
 if(myFF%>%filter(is.na(SmkAtVisitPastmonth))%>%nrow()!=0){stop('NA in SmkAtVisitPastmonth')}
+
 #######################complete case dataset###################################
 
[…]

### 2021-10-01 Scripts/02_make_methyl_summary.R

@@ -1,8 +1,9 @@
 #directory based coding
 datadir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
-john_datadir<-"/nfs/turbo/bakulski1/People/johndou/Fragile_Families/ProcessedFiles/jd"
-outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output/"
 codedir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/"
+
+####################################################################source functions
+source(file.path(codedir, 'UsefulCode', 'makePolyEpiScores.R'))
 
 #######################################chose apriori cpgs, if any
 aprioriCG<-c("cg05575921", "cg04180046", "cg05549655", "cg14179389")

### 2021-10-01 Papers/Polymethylation_scores/04_background.Rmd
started discussion outline

@@
 P2: DNA methylation could provide a reliable and specific biomarker of prenatal maternal smoke exposure.
-Cordblood and peripheral blood DNA methylation has been significantly associated with prenatal maternal smoking.
+Explanation of different approaches: polymethylation scores, individual CpG sites, summary methylation measures like clocks, etc.
+
+P3: Cordblood and peripheral blood DNA methylation has been significantly associated with prenatal maternal smoking.
+
 **Gap**: Most previous studies have used blood samples, but saliva is an easier to collect tissue.
@@
 
-P3: DNA methylation has been associated with prenatal maternal smoking at various cross-sectional time points across the life course.
+P4: DNA methylation has been associated with prenatal maternal smoking at various cross-sectional time points across the life course.
 Ideally, a biomarker of a prenatal exposure would persist across the life course.
+
 **Gap**: Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times.
 
-P4: DNA methylation has been well replicated in primarily European ancestry cohorts.
+P5: DNA methylation has been well replicated in primarily European ancestry cohorts.
 The performance of other biomarkers and risk scores varies across ancestry groups.
 To be applied clinically without exacerbating health disparities, it is important to understand how the perfomance of biomarkers vary across ancestry groups.
+
 **Gap**: Yet few (no?) studies have compared associations of prenatal maternal smoking across diffferent ancestry groups.
-
-P5: Choice of how to construct a DNA methylation based biomarker.
 

### 2021-10-01 Papers/Polymethylation_scores/06_discussion.Rmd
started discussion outline

# Discussion

Restate findings

How are results similar or different to other results

Tissue

Age

Ancestry

### 2021-10-03 Scripts/01_clean_phenotype_data.R
added clock code, added Middletone cell estimation, revised labeling

@@ -146,6 +146,16 @@ myFF<-myFF %>%
 #throws NA warning because of as.numeric in these statements - sow rite a check to make sure there aren't NAs here 
 if(myFF%>%filter(is.na(SmkAtVisitPastmonth))%>%nrow()!=0){stop('NA in SmkAtVisitPastmonth')}
 
+################################################################################add labels
+set_label(myFF)=c(get_label(myFF)[1:69], 'Has 450K Illumina chip data', 'Maternal prenatal smoking', 
+                  'Maternal prenatal alcohol use', 'Maternal prenatal any drug use', 
+                  'Any postnatal maternal smoking when child age 1 or 5', 'Postnatal maternal smoking dose when child age 1 or 5', 
+                  'Ancestry categorization from child principal components of genetic data', 
+                  paste0('Within ancestry strata principal component', 1:20), 
+                  paste0('Within all samples principal component', 1:20), 
+                  'Methylation data ID', 'Visit', 'ID', 'Batch', 'Slide', 'Array', 
+                  'Child age at visit', 'Maternal/primary care giver smoking in month prior to visit (pks/day)')
+
 #######################complete case dataset###################################
 
 methylCohort<-pdqc_all %>% filter(childteen!='M')
@@ -163,11 +173,14 @@ child_smoke<-secondhandsmkdata %>% filter(k5f1l!= "2 no" & childteen=='C')
 child_nosmoke<-secondhandsmkdata %>% filter(k5f1l=="2 no" & childteen=='C')
 teen_smoke<-secondhandsmkdata %>% filter(k6d40!="2 No" & childteen=='T')
 teen_nosmoke<-secondhandsmkdata %>% filter(k6d40=="2 No" & childteen=='T')
-completecase<-rbind(child_nosmoke, teen_nosmoke)
+completecase<-rbind(child_nosmoke, teen_nosmoke)%>%copy_labels(basemodeldata)%>%copy_labels(myFF)
 
 fathersmkdata<-completecase %>% filter(f1g4!="Missing" & !is.na(f1g4))
 
+
+
+
 #####################################################################################Save all data 
-save(file=paste0(datadir, '/CreatedData/allPhenoData.Rdata'))
+save.image(file=paste0(datadir, '/CreatedData/allPhenoData.Rdata'))
 #####################################################################################Save my specific data for modeling
 save(completecase, file=paste0(datadir, '/CreatedData/completeCasepheno.Rdata'))

### 2021-10-03 Scripts/02_make_methyl_summary.R

@@ -1,7 +1,19 @@
 #directory based coding
 datadir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
 codedir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/"
+clock_coefs="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/OGData/clock_coefs/"
 
+####################################################################source functionslibrary
+library(dplyr)
+library(EpiDISH)
+library(ewastools)
+library(tidyr)
+library(purrr)
+library(sjlabelled)
+library(wateRmelon)
+library(data.table)
+library(DunedinPoAm)
+library(knitr)
 ####################################################################source functions
 source(file.path(codedir, 'UsefulCode', 'makePolyEpiScores.R'))
 
@@ -58,17 +70,66 @@ save(polyscores, file = file.path(datadir, 'CreatedData', 'polymethylationscores
 polyscores_wide<-lapply(seq_along(polyscores), function(i) polyscores[[i]] %>% setNames(c(paste0(colnames(polyscores[[i]])[1:4], '_', names(polyscores)[i]), 'MethID')))%>%reduce(left_join, by='MethID')
 
 #################################################epigenetic clocks
-clocks<-read.csv(file=paste0(datadir, "OGData/ffcw_n1776_8clocks.csv"), header = T)
-colnames(clocks)[1]<-"MethID"
-
+###################THIS NEEDS TO BE CHANGED -- clocks weren't calculated on johns betaqc but on jonahs, 
+#have different probe filter/sample filter sets
+#I've been working with Johns betaqc but using jonahs clocks. 
+#clocks<-read.csv(file=paste0(datadir, "OGData/ffcw_n1776_8clocks.csv"), header = T)
+#colnames(clocks)[1]<-"MethID"
+#Note that there are 3 IDS missing here from Jonah's clocks because of the sex filtration step 
+#i.e. John dropped sex discordant pairs, jonah dropped sex outliers. so there wer 3 sex outliers here in 
+#johns betaqc that weren't in jonahs. But I think all these clocks need to be rerun 
+#since they were created using jonahs betaqc but I have calculated the pms, sva & the global methylation 
+#using johns betaqc 
+#options: either recalculate the clocks using johns betaqc or switch back to jonahs betaqc, fitler out discordant pairs and use 
+#that only 
+#pdqc 
+pdqc_all<-readRDS(paste0(datadir, 'OGData/', "pd_qc.rds"))
[…]

### 2021-10-03 Scripts/04_summary_methyl_models.R

@@ -29,12 +29,8 @@ library(sva)
 data.dir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/CreatedData"
 outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output"
 #read data 
-file.path(data.dir, list.files(data.dir))[(str_detect(file.path(data.dir, list.files(data.dir)), 'NoSmk'))]%>%
-  map(load, envir=.GlobalEnv)
-#bind data
-modeldata<-rbind(child_nosmoke, teen_nosmoke)
-#label data 
-set_label(modeldata)<-get_label(child_nosmoke)
+load(paste0(datadir, '/CreatedData/completeCasemethyl.Rdata'))
+
 #set as factors 
 modeldata$PostnatalMaternalSmokingAny<-factor(modeldata$PostnatalMaternalSmokingAny, 
                                               levels=c("No maternal smoking at age 1 and age 5", 

### 2021-10-03 Scripts/04_summary_methyl_models.R
fixed global methylation to be numeric

@@ -26,14 +26,15 @@ library(sva)
 #############
 #investigate why clocks have 1 fewer observation in child models
 ##################################Directories & data
-data.dir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/CreatedData"
+datadir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/CreatedData"
 outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output"
 #read data 
-load(paste0(datadir, '/CreatedData/completeCasemethyl.Rdata'))
+load(paste0(datadir, '/completeCasemethyl.Rdata'))
+modeldata=completecase%>%droplevels()
 
 #set as factors 
 modeldata$PostnatalMaternalSmokingAny<-factor(modeldata$PostnatalMaternalSmokingAny, 
-                                              levels=c("No maternal smoking at age 1 and age 5", 
+                                              levels=c("No maternal smoking at age 1 and 5", 
                                                        "Maternal smoking at age 1 or age 5"))
 modeldata$SmkAtVisitPastmonth<-factor(modeldata$SmkAtVisitPastmonth, 
                                       levels=c("No smoking", 
@@ -43,15 +44,15 @@ modeldata$slide<-factor(modeldata$slide)
 modeldata$Sample_Plate<-factor(modeldata$Sample_Plate)
 
 #sv data - no variance filter, whole subset
-load(paste0(data.dir, '/svNone.Rds'))
+load(paste0(datadir, '/svNone.Rds'))
 is_sv=function(x) str_replace_all(x, " ", "")
 sv.vNone=sv.vNone%>%rename_with(is_sv, contains('sv'))
 modeldata=left_join(modeldata, sv.vNone)
 
 #################################Set variables for modeling
 #outcomes & outcome_labels
-y_vector<-c("globalmethylation", "pediatric", "grim", "anynewborn_center", "SSnewbornCT_center", "SSolder_center", "cg05575921")
-outcome_labels=c('Global methylation', 'Pediatric clock', 'GRIM clock',  
+y_vector<-c("globalmethylation", "pediatric", "anynewborn_center", "SSnewbornCT_center", "SSolder_center", "cg05575921")
+outcome_labels=c('Global methylation', 'Pediatric clock',  
                  'Any smoking polymethylation score (newborns)', 
[…]

### 2021-10-05 Papers/Polymethylation_scores/04_background.Rmd
outline intro and discussion

--- 2021-10-01 Papers/Polymethylation_scores/04_background.Rmd
@@
+---
+output:
+  word_document: default
+  html_document: default
+---
 # Background
@@
 P1: Maternal prenatal smoking is a public health concern but measuring prenatal maternal smoking behavior is challenging.
-Developing an in vitro biomarker for prenatal smoke exposure would be beneficial for personalized medicine.
+
+  - In 2018, 11% of pregnant US women aged 15-44 reported 
+   any past-month cigarette use.
+
+  - Health effects in exposed children include low-birth weight, 
+  neurodevelopmental outcomes, and asthma.
+
+  - Maternal prenatal smoking is underreported due to stigma, so 
+  self-reports are unideal proxies for fetal smoke exposure.
+
+  - Serum cotinine is a biomarker of smoking but has a short 
+  half-life of nine hours in pregnant women, thus 
+  misclassification can still occur.
+
+  - Further, serum cotinine measures and maternal self-report 
+  data are rarely available outside of birth cohorts.
+
+  - Developing a persistent, reliable and feasible biomarker of prenatal maternal smoke exposure is critical for advancing research and precision health.
 
 P2: DNA methylation could provide a reliable and specific biomarker of prenatal maternal smoke exposure.
-Cordblood and peripheral blood DNA methylation has been significantly associated with prenatal maternal smoking.
-**Gap**: Most previous studies have used blood samples, but saliva is an easier to collect tissue.
+
+  - Cordblood, peripheral blood and placental DNA methylation has been significantly associated with prenatal maternal smoking. [Joubert 2016; Bustamante 2021; Davey Smith 2018]
+
+  - Changes in DNA methylation are specific to smoke exposure *in utero*
+
+  - Previous work has found reasonable performance of cord blood based [London 2017] (accuracy 91%, sens 58%, spec 97%) [Fallin 2016] (81% accuracy) and peripheral blood based biomarkers (AUC: 0.72) [Davey Smith 2018] of prenatal maternal smoking
+
+**Gap**: Previous studies have used blood or placental samples.
+Cord blood or placenta samples are unlikely to be available in a clinical setting.
+Peripheral blood is a more realistic clinical sample, but saliva is even easier to collect.
 To our knowledge, no study has measured associations between prenatal maternal smoking and DNA methylation in saliva samples.
@@
 P3: DNA methylation has been associated with prenatal maternal smoking at various cross-sectional time points across the life course.
-Ideally, a biomarker of a prenatal exposure would persist across the life course.
-**Gap**: Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times.
+
+- Ideally, a biomarker of a prenatal exposure would persist across the life course.
+
+- Blood DNA methylation signatures also were significantly associated with prenatal maternal smoking exposure in cohorts of older children and adults [Terry 2019, Davey Smith 2018]
+
+- A 2018 study found that DNA methylation based scores could predict predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [Davey Smith 2018]
+
+**Gap**: Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times or evaluated if the inclusion of multiple time points improves performance.
 
 P4: DNA methylation has been well replicated in primarily European ancestry cohorts.
-The performance of other biomarkers and risk scores varies across ancestry groups.
-To be applied clinically without exacerbating health disparities, it is important to understand how the perfomance of biomarkers vary across ancestry groups.
-**Gap**: Yet few (no?) studies have compared associations of prenatal maternal smoking across diffferent ancestry groups.
 
-P5: Choice of how to construct a DNA methylation based biomarker.
+- The performance of other biomarkers and risk scores varies across ancestry groups.
+To be applied clinically without exacerbating health disparities, it is important to understand how the performance of biomarkers vary across ancestry groups.
 
-P6: To translate DNA methylation to a *in vitro* biomarker for prenatal exposures, we must understand how performance [check with kelly if ok to use this word here now that have ROC plots] varies across tissue, age and ancestry.
-**To address these gaps we used XXXX longitudinal saliva samples from 9 and 14 year old children enrolled in a  diverse US birth cohort to test associations between DNA methylation summary measures and prenatal maternal smoke exposure.**  We found that maternal prenatal smoking was significantly associated with polymethylation scores.
-This association was persistent across time-point and consistent across ancestry strata.
-Polymethylation scores outperformed other non-target methylation summary measures, including global methylation and methylation clocks, as well as individual, *a priori* selected CpG sites.
-Our results suggest that polymethylation scores can serve as a persistent and consistent biomarker of prenatal smoke exposure.
+- DNA methylation can vary across genetic haplotypes and therefore ancestry groups.
+
+- Few studies have used non-European ancestry populations and those that have had small sample sizes: Fallin 2016 186 individuals of African or admixed ancestry, Terry 2018 60% African or hispanic but only 89 participants,  Burchard 2019 572 Latino children
+
+**Gap**: No studies have compared associations of prenatal maternal smoking DNA methylation biomarkers across different ancestry groups.
+
+P5: The goal of this study was to assess the *in vitro* potential of salivary DNA methylation biomarkers for prenatal maternal smoking.
+We tested associations between prenatal maternal smoke exposure and saliva DNA methylation biomarkers from the Illumina 450K chip, including epigenetic clocks, polymethylation scores, and individual *a priori* CpG sites using longitudinal samples from the Fragile Families and Child Wellbeing study.
+We performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a consistent biomarker of prenatal maternal smoking across populations and time.

### 2021-10-05 Papers/Polymethylation_scores/06_discussion.Rmd
outline intro and discussion

@@
 
-How are results similar or different to other results
+How do findings fit with previous literature: 
+- Performance was worse than the performance of biomarkers using cord blood (AUC: 0.82-0.97), but comparable to performance of biomarkers using peripheral blood (AUC 0.72...
+NOTE: I don't think this was adjusted for anything else, i.e.
+if any other predictors were included in the model--if not, then ours is worse).
 
-Tissue
+What does our study add:
 
-Age
+Strengths:
 
-Ancestry
+- Large sample size and longitudinal measures
+
+-
+
+Limitations:
+
+- Available coefficients for construction of scores were from blood, primarily European ancestry meta-analysis
+
+- Used maternal self report as opposed to serum cotinine levels as exposure measurement - exposure misclassification/information bias
+
+- Possibility for selection bias into cohort due to live birth bias
+
+- Residual confounding is always possible.

### 2021-10-05 Reports/Results.Rmd
editing results.rmd

@@
 codedir=''
+datadir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data"
+load(paste0(datadir, '/CreatedData/allPhenoData.Rdata'))
+load(paste0(datadir, '/CreatedData/allmethyl.Rdata'))
+load(paste0(datadir, '/CreatedData/completeCasemethyl.Rdata'))
 ```
@@
 
-# Coding tasks
-
-- X replace epidish cell estimation with ewastools Middleton
-
-- check with jonah about final sample size and exclusions
-- determine what to do w/ 1 sample missing clock data (related to above)
-- rerun clocks i plan on using on all samples
-
-- make this an entire story line, combined supplemental and non supplemental
+# Current issues
 
@@
 
-# Supplemental Table 1 - Comparison of 450K subsample to Fragile Families cohort
+```{r, eval=T}
+#Make Labels
+n_fullFF<-paste0("Fragile Families cohort \n", "n=", FF_labeled %>% nrow())
+n_methylQC<-paste0("Cohort with 450K methylation data \n", "n=", methylCohort$idnum%>%unique()%>%length(), " individuals with m=", methylCohort$MethID%>%length(), " samples")
 
-# Supplemental Figure 1 - Cell type proportion estimation
+n_QCloss<-paste0("Methylation data QC: ", 
+                 pdqc_all%>%filter(probe_fail_10==T & childteen!='M')%>%nrow(), " samples / ", 
+                 pdqc_all%>%filter(probe_fail_10==T & childteen!='M')%>%dplyr::select(idnum)%>%unique()%>%nrow(), 
+                 " individuals with >10% of probes failing dropped, \n",
+                 pdqc_all%>%filter(sex_new!=predicted_sex & probe_fail_10==F & childteen!='M')%>%nrow(), " samples / ", 
+                 pdqc_all%>%filter(sex_new!=predicted_sex & probe_fail_10==F & childteen!='M')%>%dplyr::select(idnum)%>%unique()%>%nrow(),
+                 " individuals with discordant sex samples dropped, \n",
+                 pdqc_all%>%filter(predicted_sex_outlier==T & sex_new==predicted_sex & probe_fail_10==F & childteen!='M')%>%nrow(), ' samples / ',
+                  pdqc_all%>%filter(predicted_sex_outlier==T & sex_new==predicted_sex & probe_fail_10==F & childteen!='M')%>%dplyr::select(idnum)%>%unique()%>%nrow(), 
+                  ' individuals with abnormal sex chromosome intensity dropped & \n',
+                 length(reps_remove), " duplicate samples dropped")
 
-# Supplemental Figure 2 - Choice of beta methylation transformation
+n_methylFF<-paste0("Cohort with quality controlled 450K data \n","n=", myFF %>% dplyr::select(id) %>% unique() %>% nrow(), " individuals with m=", myFF%>%nrow(), " samples")
 
-# Supplemental Figure 3 - Distribution and correlation of methylation summary measures
+n_baseexclusions<-paste0("n individuals=",  
+                         myFF %>% filter_at(all_of(basemodelvar), any_vars((. %in%c("Missing", "Missing PC data"))))%>% dplyr::select(id) %>% unique() %>% nrow(), " \ m samples=", 
+                         myFF %>% filter_at(all_of(basemodelvar), any_vars((. %in%c("Missing", "Missing PC data")))) %>% nrow(), " missing principal components of ancestry; \n n individuals=",
+                         myFF %>% filter_at(all_of(basemodelvar), any_vars((is.na(.))))%>%dplyr::select(id)%>%unique()%>%nrow(), " individuals \ m=", myFF %>% filter_at(all_of(basemodelvar), any_vars((is.na(.))))%>%nrow(), " samples missing child age at maternal/PCG interview")
 
-#Supplemental Figure 4 - Change in methylation summary measures between visits
+n_prenatalexposures<-paste0("Prenatal exposures models: \n","n=", prenatalexdata %>% dplyr::select(id) %>% unique() %>% nrow(), " individuals with m=", prenatalexdata %>%nrow(), " samples")
+n_prenatalexclusions<-paste0("n=", basemodeldata %>% filter_at(all_of(prenatalexposurevar), any_vars(.
+%in% c("Missing"))) %>% dplyr::select(id) %>% unique() %>% nrow(), " individuals & m=", basemodeldata %>% filter_at(all_of(prenatalexposurevar), any_vars(.
+%in% c("Missing"))) %>% nrow(), " samples missing \n data on maternal prenatal alcohol and drug use")
+
+n_secondhandSmk<-paste0("Secondhand smoke exposure models: \n","n=", secondhandsmkdata %>% dplyr::select(id) %>% unique() %>% nrow(), " individuals with m=", secondhandsmkdata %>%nrow(), " samples")
+n_secondhandexclusions<-paste0("n=", prenatalexdata$id[!(prenatalexdata$id %in% secondhandsmkdata$id)] %>% unique()%>% length(), " individuals & m=", prenatalexdata %>% filter_at(all_of(secondhandsmokevars), any_vars(.
+%in% c("Missing"))) %>% nrow(), " samples missing data \n on maternal or primary care giver postnatal smoking")
+
+n_Child<-paste0('n=', secondhandsmkdata %>%filter(childteen=='C')  %>% nrow(), " individuals with a 9 year visit")
+n_ChildSmkexcluded<-paste0("Exclude children who smoke: \n",'n=', child_nosmoke %>% nrow(), " individuals with a 9 year visit")
+n_ChildSmkexclusions<-paste0('n=', child_smoke%>%nrow(), " individual \n smoking at 9 year visit \n or missing focal child smoking data")
+n_Teen<-paste0('n=', secondhandsmkdata %>%filter(childteen=='T')   %>% nrow(), " individuals with a 15 year visit")
+n_TeenSmkexcluded<-paste0("Exclude children who smoke: \n",'n=', teen_nosmoke %>% nrow(), " individuals with a 15 year visit")
+n_TeenSmkexclusions<-paste0('n=', teen_smoke%>%nrow(), " individuals \n smoking at 15 year visit \n or missing focal child smoking data")
+
+graph1<-DiagrammeR::grViz("
+   digraph graph2 {
+   compound=true;
+   graph [layout = dot]
+
+   node [shape = rectangle, width=4]
+   a [label = '@@1']
+   b [label = '@@2']
+   c [label = '@@3']
+   d [label = '@@4']
+   e [label = '@@5']
+   f [label = '@@6']
+   g [label = '@@7']
+   h [label = '@@8']
+   i [label = '@@9', group=g1]
+   j [label = '@@10']
+   k [label = '@@16']
+   l [label = '@@17']
+   m [label = '@@19']
+
+   #fake nodes for connecting to edges
+   node [shape=point, width=0.01, height=0.01]
+   a1
+   b1
+   c1
+   d1
+#   e1
+  f1 [group=g1]
+  g1
+  k1
+
+  #exclusion nodes
+  node [shape = rectangle, width=4]
+  aa [label = '@@20']
+  bb [label = '@@11']
+  cc [label = '@@12']
+  dd [label = '@@13']
+  ff [label = '@@14', width=3]
+  gg [label = '@@15', width=3]
+  kk [label = '@@18', width=3]
+
+  #draw connections
+  a->m
+  m->a1 [arrowhead=none]
+  a1->b
+  b->b1 [arrowhead=none]
+  b1->c
+  c->c1 [arrowhead=none]
+  c1->d
+  d->d1 [arrowhead=none]
+  d1->e
+  e->f; e->g; e->k
+{rank=same; h->e [dir=back, minlen=2, style=dashed]}
+  f->f1 [arrowhead=none]
+  g->g1 [arrowhead=none]
+  k->k1 [arrowhead=none]
+  f1->i 
+  g1->j 
+  k1->l
+
+{rank=same; a1->aa [minlen=2]}   
+{rank=same; b1->bb [minlen=2]} 
+{rank=same; c1->cc [minlen=2]}
+{rank=same; d1->dd [minlen=2]}
+{rank=same; f1->ff [minlen=2]}
+{rank=same; g1->gg [minlen=2]}
+{rank=same; k1->kk [minlen=2]}
+
+#invisible constraints to force ordering
+ff->i [style=invis]
+   }
+
+   [1]: n_fullFF
+   [2]: n_methylFF
+   [3]: n_basemodel
+   [4]: n_prenatalexposures
+   [5]: n_secondhandSmk
+   [6]: n_Child
+   [7]: n_Teen
+   [8]: n_sensFatherSmoking
+   [9]: n_ChildSmkexcluded
+   [10]: n_TeenSmkexcluded
+   [11]: n_baseexclusions
+   [12]: n_prenatalexclusions
+   [13]: n_secondhandexclusions
+   [14]: n_ChildSmkexclusions
+   [15]: n_TeenSmkexclusions
+   [16]: n_twovisits
+   [17]: n_twovisitsSmkExcluded
+   [18]: n_twovisitSmkExclusions
+   [19]: n_methylQC
+   [20]: n_QCloss
+")
+```
+
+# Supplemental Table 1 - Comparison of excluded sample to included sample
+
+```{r}
+myFF$MethylData2<-ifelse(myFF$MethID %in% completecase$MethID, 'Final analysis group', 'Not in final analysis group')
+set_label(myFF$MethylData2)<-'Exclusions based on missing data'
+myFF=left_join(myFF, methyldata%>%mutate(idnum=as.character(idnum)))
+methyl2<-createTable(compareGroups(MethylData2~smkPreg_binary+globalmethylation+pediatric+Epithelial.cells_saliva+Leukocytes_saliva+anynewborn_center+SSnewbornCT_center+SSolder_center+cm1inpov+ancestry+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth, data=myFF), show.all=T)
+export2md(strataTable(methyl2, 'childteen'), caption="Exclusion table of subset with quality controlled methylation data vs final analytic subset for main predictor, outcomes and key demographic variables")
+```
+
+# Supplemental Figure 2 - Correlation of methylation summary measures
+
+```{r}
+library(Hmisc)
+library(corrplot)
+res2<-completecase%>%dplyr::select(globalmethylation, pediatric, grim, Epithelial.cells_saliva, Leukocytes_saliva, anynewborn_center, SSnewbornCT_center, SSolder_center, cg05575921)%>%as.matrix(.)%>%rcorr()
+corrplot.mixed(res2$r, upper="ellipse", lower='number', order='hclust')
+```
+
+#Supplemental Figure 3 - Change in methylation summary measures between visits
+
+```{r}
+
+```
 
@@
   dplyr::select(predictors, covariates, methylation, childteen, test)%>%
-  mutate(modeltype=ifelse(str_detect(predictors, '\\+'), covariates, 'No covariates'))%>%
+  mutate(modeltype=ifelse(methylation=='', 'Base model', paste0('Base model+', methylation)))%>%
   unnest(test) %>% 
-  filter(str_detect(methylation, 'SSnewbornCT|cg055|global|pediatric') &
-           covariates=='')%>%
-  ggplot(aes(x=spec, y=sens, color=methylation))+geom_line()+
+  filter(str_detect(methylation, 'SSnewbornCT|SSolder|cg055|global|pediatric'))%>%
+  ggplot(aes(x=spec, y=sens, color=modeltype))+geom_line()+
   facet_wrap(childteen~., labeller=labeller(childteen=age_labels), nrow=1)+

### 2021-10-05 Scripts/01_clean_phenotype_data.R

@@ -6,12 +6,11 @@ library(tidyr)
 library(stringr)
 library(dplyr)
 library(purrr)
-library(tidyselect)
+library(tidydplyr::select)
 library(sjlabelled)
 
 #directory based coding
 datadir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
-john_datadir<-"/nfs/turbo/bakulski1/People/johndou/Fragile_Families/ProcessedFiles/jd"
 outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output/"
 codedir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/"
 
@@ -37,7 +36,7 @@ FF_factors=FFmeta%>%
 #recode binary, categorical variables so they have label names as values (nicer for table and plotting)
 pheno=pheno %>% dplyr::mutate(across(any_of(FF_factors), ~sjlabelled::as_label(., drop.levels=TRUE)))
 
-#select variables
+#dplyr::select variables
 idvars<-c("id", "ff_id")
 demovars<-c("cm1age", "cm1bsex", "cm1ethrace", "cm1edu", "cm1hhinc", "cm1inpov", "m1h3", "m1h3a", "m1h3b", "m1i4", "m1i4a", "m1i4b", "cf1age", "cf1ethrace", "f1h3", "f1h3a", "f1h3b", "ch5agem","ch6yagem", "c6yagey", "cm5b_age", "cm5b_ageyrs", "cp6yagem", "cp6yagey", "hv5_agem", "hv6_yagem", "hv6_yagey")
 smokingvars<-c("m1g4", "f1g4", "f2j5", "f2j5a", "f2j7", "f2j7a", "f3j31", "f3j32", "f4j18", "f4j19", "f5g17", "f5g18", "k5f1k", "k5f1l", "k6d40", "k6d41", "k6d41", "k6d42", "k6d43", "k6d45", "k6d46", "k6d47", "m2j5", "m2j5a", "m2j7", "m2j7a", "m4j18", "m4j19", "m5g17", "m5g18", "n5f17", "n5f18", "p3a22", "p3a23", "p3a23a", "p3a23b", "p3a24", "p4a22", "p4a23", "p5h15", "p5h15b", "p5q3cr", "p6h74", "p6h75", "p6h76", "p6h77", "p6h78")
[…]

### 2021-10-05 Scripts/02_make_methyl_summary.R

@@ -3,7 +3,7 @@ datadir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
 codedir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/"
 clock_coefs="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/OGData/clock_coefs/"
 
-####################################################################source functionslibrary
+####################################################################library
 library(dplyr)
 library(EpiDISH)
 library(ewastools)
@@ -21,8 +21,9 @@ source(file.path(codedir, 'UsefulCode', 'makePolyEpiScores.R'))
 aprioriCG<-c("cg05575921", "cg04180046", "cg05549655", "cg14179389")
 
 ################################read in methylation data
-betaqc<-readRDS(file=paste0(datadir, 'OGData/', "noob_filtered.rds"))
-
+#as of october 4 switched to beta file with  423668 probes (see Jonah email search Check in on sex filtration/clocks Fragile Families)
+betaqc<-readRDS(file=paste0(datadir, 'OGData/', "betaqc.rds"))
+if(nrow(betaqc)!=423668){stop("Probe number not what thought, stop and check betaqc")}
 ###########################################cell type proportions 
 celltypes<-epidish(beta.m = betaqc, ref.m = centEpiFibIC.m, method = "RPC")
 #RPC method w/ centEpiFibIC
@@ -38,7 +39,7 @@ cell_types=left_join(estF_FF, estL_FF)
 
 ################################################global methylation
 globalmethy<-colMeans(betaqc, na.rm=T)
-globalmethdf<-as.data.frame(cbind("MethID"=names(globalmethy), "globalmethylation"=globalmethy))
+globalmethdf<-data.frame("MethID"=names(globalmethy), "globalmethylation"=globalmethy)
 
 #################################################apriori cpgs
 aprioriCGdf<-as.data.frame(t(betaqc[aprioriCG, ]))
@@ -67,22 +68,20 @@ scores_center=makePolyEpiScore(m=betaqc, b=cpgs, transformopt = 'meancenter')
 polyscores=list('notransform'=scores, 'zscore'=scores_z, 'center'=scores_center)
 polyscores<-lapply(polyscores, function(x) x%>% mutate(MethID=colnames(betaqc)))
 save(polyscores, file = file.path(datadir, 'CreatedData', 'polymethylationscores.RDS'))
[…]

### 2021-10-05 Scripts/03_sva_variables.R

@@ -12,10 +12,11 @@ data.dir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
 output.dir=paste0(data.dir, 'CreatedData/')
 
 #load in betaqc matrix
-betaqc<-readRDS(file=paste0(data.dir, 'OGData/', "noob_filtered.rds"))
-file.path(paste0(data.dir, 'CreatedData/'), list.files(paste0(data.dir, 'CreatedData/')))[(str_detect(file.path(paste0(data.dir, 'CreatedData/'), list.files(paste0(data.dir, 'CreatedData/'))), 'NoSmk'))]%>%map(load, envir=.GlobalEnv)
+#as of october 4 switched to beta file with  423668 probes (see Jonah email search Check in on sex filtration/clocks Fragile Families)
+betaqc<-readRDS(file=paste0(data.dir, 'OGData/', "betaqc.rds"))
 
-pheno<-rbind(child_nosmoke, teen_nosmoke) # dataset with the most exclusion i.e. squaring dataset off for all models (excluding kids who smoke in figure)
+load(paste0(datadir, '/CreatedData/completeCasepheno.Rdata'))
+pheno<-completecase # dataset with the most exclusion i.e. squaring dataset off for all models (excluding kids who smoke in figure)
 
 #make a matrix of model data 
 #filter beta matrix to only those in analytic subset 

### 2021-10-05 Scripts/04_summary_methyl_models.R

@@ -184,8 +184,7 @@ modeldata=modeldata%>%mutate(smkPreg_binaryN=case_when(smkPreg_binary=='Yes'~1, 
 
 #global predictors, cross sectional
 roc_predictors=gsub('~smkPreg_binary', '', paste0(base_model_vars, global_pcs))
-args=list('childteen'=age_vector, 'methylation'=c('', y_vector), 'covariates'=roc_predictors)%>%cross_df()%>%mutate(predictors=paste0(methylation, covariates))
-
+args=list('childteen'=age_vector, 'methylation'=c('', y_vector, 'cg05549655'), 'covariates'=roc_predictors)%>%cross_df()%>%mutate(predictors=paste0(methylation, covariates))
 
 global_roc=modeldata%>%
   group_by(childteen)%>%nest()%>%left_join(args)%>%
@@ -196,23 +195,11 @@ global_roc=modeldata%>%
          test=map(roc, ~data.frame('sens'=.x$sensitivities, spec=.x$specificities)))
 
 #local predictors, crosssectional 
-local_predictors=c('', gsub('~smkPreg_binary', '', c(paste0(predictors, local_pcs), surrogate_model_vars)))
-args=list('childteen'=age_vector, 'methylation'=y_vector, 'covariates'=local_predictors[1:2])%>%cross_df()%>%mutate(predictors=paste0(methylation, covariates))
+roc_predictors=gsub('~smkPreg_binary', '', paste0(base_model_vars, local_pcs))
+args=list('childteen'=age_vector, 'methylation'=c('', y_vector), 'covariates'=roc_predictors)%>%cross_df()%>%mutate(predictors=paste0(methylation, covariates))
 
 
 local_roc=modeldata%>%group_by(childteen, ancestry)%>%nest()%>%left_join(args)%>%
-  mutate(model=pmap(list(data, 'smkPreg_binaryN~', predictors), model_reversed), 
-         tidied=map(model, tidy, conf.int=T), 
-         augmented=map(model, augment, type.predict = "response"), 
-         roc=map(augmented, ~roc(.x$smkPreg_binaryN, .x$.fitted)), 
-         test=map(roc, ~data.frame('sens'=.x$sensitivities, spec=.x$specificities)))
-
-
-#investigation 
-predictors_test=c(paste0('+', str_split(base_model_vars, '\\+')[[1]][-1]), local_pcs, '')
[…]

### 2021-10-11 Papers/Polymethylation_scores/04_background.Rmd
work on intro and background outline

--- 2021-10-01 Papers/Polymethylation_scores/04_background.Rmd
@@
+---
+output:
+  word_document: default
+  html_document: default
+---
 # Background
 
-P1: Maternal prenatal smoking is a public health concern but measuring prenatal maternal smoking behavior is challenging.
-Developing an in vitro biomarker for prenatal smoke exposure would be beneficial for personalized medicine.
+Maternal prenatal smoking is a public health concern: in 2018, 11% of pregnant US women aged 15-44 reported any past-month cigarette use, and exposed children are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma.
+Measuring prenatal maternal smoking behavior is challenging.
+Maternal prenatal smoking is underreported due to stigma.
+The gold-standard for smoking measurements, serum cotinine, has a short half-life of nine hours in pregnant women.
+Further, serum cotinine measures during pregnancy are rarely available outside of birth cohorts, limiting clinical application as a marker of prenatal exposures.
+Developing a persistent, reliable and feasible biomarker of prenatal maternal smoke exposure is critical for advancing research and precision health.
 
-P2: DNA methylation could provide a reliable and specific biomarker of prenatal maternal smoke exposure.
-Cordblood and peripheral blood DNA methylation has been significantly associated with prenatal maternal smoking.
-**Gap**: Most previous studies have used blood samples, but saliva is an easier to collect tissue.
-To our knowledge, no study has measured associations between prenatal maternal smoking and DNA methylation in saliva samples.
+DNA methylation has been associated with prenatal maternal smoking at various cross-sectional time points across the life course in primarily European ancestry groups.
+Cordblood, peripheral blood and placental DNA methylation has been significantly associated with prenatal maternal smoking.
+[Joubert 2016; Bustamante 2021; Davey Smith 2018].
+Blood DNA methylation signatures in cohorts of older children and adults also were significantly associated with prenatal maternal smoking exposure [Terry 2019, Davey Smith 2018].
+Yet few studies have examined associations at different ages in the same participants.
+Additionally, while associations are well-replicated in cohorts of European ancestry, few studies have used non-European ancestry populations and those that have had small sample sizes: Fallin 2016 186 individuals of African or admixed ancestry, Terry 2018 60% African or hispanic but only 89 participants, Burchard 2019 572 Latino children].
+DNA methylation is associated with prenatal maternal smoking, but the consistency of this association across age and ancestry is understudied.
 
-P3: DNA methylation has been associated with prenatal maternal smoking at various cross-sectional time points across the life course.
-Ideally, a biomarker of a prenatal exposure would persist across the life course.
-**Gap**: Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times.
+The association between prenatal maternal smoking and DNA methylation suggests DNA methylation could provide a reliable and specific biomarker of prenatal maternal smoke exposure.
+Importantly, changes in DNA methylation are specific to smoke exposure *in utero*.
+Previous work has found reasonable performance of cord blood based [London 2017] (accuracy 91%, sens 58%, spec 97%) [Fallin 2016] (81% accuracy) and peripheral blood based biomarkers (AUC: 0.72) [Davey Smith 2018] of prenatal maternal smoking.A 2018 study found that DNA methylation based scores could even predict predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [Davey Smith 2018].
+However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across age and ancestry has not been evalutated.
+Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times or evaluated if the inclusion of multiple time points improves performance.
+Additionally, previous work has relied on blood and placental samples.
+Cord blood or placenta samples are unlikely to be available in a clinical setting.
+While peripheral blood is a more realistic clinical sample,  saliva is even easier to collect.
+To our knowledge, no study has evaluated salivary DNA methylation as a biomarker for prenatal maternal smoking.
 
-P4: DNA methylation has been well replicated in primarily European ancestry cohorts.
-The performance of other biomarkers and risk scores varies across ancestry groups.
-To be applied clinically without exacerbating health disparities, it is important to understand how the perfomance of biomarkers vary across ancestry groups.
-**Gap**: Yet few (no?) studies have compared associations of prenatal maternal smoking across diffferent ancestry groups.
-
-P5: Choice of how to construct a DNA methylation based biomarker.
-
-P6: To translate DNA methylation to a *in vitro* biomarker for prenatal exposures, we must understand how performance [check with kelly if ok to use this word here now that have ROC plots] varies across tissue, age and ancestry.
-**To address these gaps we used XXXX longitudinal saliva samples from 9 and 14 year old children enrolled in a  diverse US birth cohort to test associations between DNA methylation summary measures and prenatal maternal smoke exposure.**  We found that maternal prenatal smoking was significantly associated with polymethylation scores.
-This association was persistent across time-point and consistent across ancestry strata.
-Polymethylation scores outperformed other non-target methylation summary measures, including global methylation and methylation clocks, as well as individual, *a priori* selected CpG sites.
-Our results suggest that polymethylation scores can serve as a persistent and consistent biomarker of prenatal smoke exposure.
+DNA methylation changes hold promise as a biomarker for prenatal maternal smoking.
+However, before translation to clinical applications, we must understand how and if the signal varies across age, ancestry and tissue sample.
+The goal of this study was to assess the *in vitro* potential of salivary DNA methylation biomarkers for prenatal maternal smoking.
+We tested associations between prenatal maternal smoke exposure and saliva DNA methylation biomarkers from the Illumina 450K chip, including epigenetic clocks, polymethylation scores, and individual *a priori* CpG sites using longitudinal samples from the Fragile Families and Child Wellbeing study.
+We performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a consistent biomarker of prenatal maternal smoking across populations and time.

### 2021-10-11 Papers/Polymethylation_scores/06_discussion.Rmd
work on intro and background outline

--- 2021-10-01 Papers/Polymethylation_scores/06_discussion.Rmd
@@
 
-Restate findings
+P1 (restate findings): In this longitudinal birth cohort, we found an association between prenatal maternal smoking and salivary DNA methylation.
 
-How are results similar or different to other results
+- Polymethylation scores were associated with prenatal maternal smoking across ages and ancestry while global methylation and epigenetic clocks were not.
 
-Tissue
+- Polymethylation scores had reasonable accuracy and accuracy was more consistent across time when using polymethylation scores than when using individual *a priori* CpG sites.
 
-Age
+P2 (compares to previous findings): Our findings fit within the previous literature on DNA methylation as a possible biomarker of prenatal maternal smoking.
 
-Ancestry
+- {if including EWAS results: many top hits from previous EWAS analyses using blood samples were replicated} (joubert 2016).
+
+- Fits well with previous findings of an association in cord and peripheral blood (Joubert 2016).
+Although no studies of prenatal maternal smoking and saliva methylation, own-smoking and DNA methylation association in saliva, including at cg05575921 (Philibert 2019, Long 2019).
+
+- Performance was worse than the performance of biomarkers using cord blood (AUC: 0.82-0.97 from Fallin 2016 London 2017), but comparable to performance of biomarkers using peripheral blood (AUC 0.72, Davey Smith 2018...
+NOTE: I don't think this was adjusted for anything else, i.e.
+other predictors were included in the model--so technically then ours is worse).
+
+- For own smoking saliva DNA methylation at cg05575921 has been found to have a higher accuracy (in models incorporating age gender and methylations status) was 0.995 in blood, 0.971 in saliva, improved to 0.981 when incorporating another methylation marker of tissue heterogeneity [Philibert, Long 2019]
+
+P3 (what we add): We contribute to the translation of DNA methylation to a clinical biomarker by evaluating consistency across ages and ancestries.
+
+- Even when polymethylation score coefficients were based on coefficients from a metaanalysis of primarily European ancestry, we found an association with prenatal maternal smoking.
+Effect estimates relatively consistent across ancestries
+
+- AUC for polymethylation scores calculated using coefficients from models adjusting for cell-type better than those not adjusting for cell-type, cell-type control should be instituted both when creating weights and when applying scores to new population.
+
+- AUC better for polymethylationn scores than cg055, and more consistent across ages. Better to incorporate multiple methylation sites than single methylation site?
+
+P4 (limitations & strengths):
+
+- Available coefficients for construction of scores were from blood, primarily European ancestry meta-analysis
+
+- Exposure measurement not ideal: Used maternal self report as opposed to serum cotinine levels as exposure measurement - exposure misclassification/information bias; any vs sustained smoking
+
+- Possibility for selection bias into cohort due to live birth bias
+
+- Residual confounding is always possible.
+
+- Large sample size and longitudinal measures

### 2021-10-11 Papers/Polymethylation_scores/index_paper.Rmd
worked on outlining introduction
work on intro and background outline

@@
 date: "9/13/2021"
-output: pdf_document
-bib: Methylation_SmokingBib.bib
+output:
+  word_document: default
+  html_document:
+    df_print: paged
+bib: ../Methylation_SmokingBib.bib
 ---

### 2021-10-14 Reports/FFCleaning2.Rmd
05 ewas script and results section improvements

@@
 
-```{r sva}
+```{r sva, eval=F}
 
@@
 
-```{r, saveData}
+```{r saveData, eval=F}
 save(child_base, file=paste0(datadir, 'CreatedData/', 'ChildBaseModelData.Rdata'))

### 2021-10-14 Reports/Results.Rmd
05 ewas script and results section improvements

--- 2021-09-30 Reports/Results.Rmd
@@
 library(kableExtra)
+library(gtsummary)
+library(gtools)
 library(haven)
@@
 codedir=''
-```
-
-```{r future run scripts, cache=T}
-
-source('/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/Scripts/03_summary_methyl_models.R')
+datadir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data"
+load(paste0(datadir, '/CreatedData/allPhenoData.Rdata'))
+load(paste0(datadir, '/CreatedData/allmethyl.Rdata'))
+load(paste0(datadir, '/CreatedData/completeCaseSVA.Rdata'))
+load(paste0(datadir, '/CreatedData/methylation_summarymodels.Rdata'))
 ```
@@
 ```{r plotting variables}
-theme_set(theme_classic())
+theme_set(theme_bw())
 ```
@@
 n_obs=modeldata%>%nrow()
+
+all_methyl=c("globalmethylation", "pediatric", "grim", "Epithelial.cells_saliva", "Leukocytes_saliva", "SSnewborn_center", "SSnewbornCT_center", "SSolder_center", "cg05575921")
+main_methyl=c("globalmethylation", "pediatric", "SSnewbornCT_center", "cg05575921")
+
+all_methyl_labels=modeldata%>%dplyr::select(any_of(all_methyl))%>%get_label()
+all_methyl_labels[str_detect(all_methyl_labels, 'Polymethylation', )]=paste0(rep('Score:', 3), c('Sustained smoking \n newborns', 'Sustained smoking \n newborns (w/ CT control)', 'Sustained smoking \n older children'))
+names(all_methyl_labels)=all_methyl
+
+all_methyl_levels=c("SSnewbornCT_center", "SSnewborn_center",  "SSolder_center",
+                                       "pediatric" , "grim", "cg05575921", 
+                                      "globalmethylation","Epithelial.cells_saliva", "Leukocytes_saliva")
+
+main_methyl_labels=modeldata%>%dplyr::select(any_of(main_methyl))%>%get_label()
 
@@
 
-# Figure 1 - CONSORT Diagram
+# Figure 1 - Selection of samples from the Fragile Families and Child Wellbeing (FFCW) study into this analysis.
+N represents the number of individuals at each step in the selection procedure, M represents the number of samples, as individuals with repeated measures can have more than one sample.
 
-# Supplemental Table 1 - Comparison of 450K subsample to Fragile Families cohort
+```{r, eval=T}
+#function
+N_M_count=function(filter_statement, statement, df){
+  paste0('m=', df %>% filter(eval(parse(text = filter_statement)))%>%nrow(), 
+         ' samples & ', 
+         'n=', df %>% filter(eval(parse(text = filter_statement))) %>% dplyr::select(idnum) %>%unique()%>%nrow(), 
+         ' individuals with', statement)
+}
+#Make Labels
+n_fullFF<-paste0("Fragile Families cohort \n", "n=", FF_labeled %>% nrow())
+n_methylQC<-paste0("Cohort with 450K methylation data \n", "n=", methylCohort$idnum%>%unique()%>%length(), " individuals with m=", methylCohort$MethID%>%length(), " samples")
 
-# Supplemental Figure 1 - Cell type proportion estimation
+n_QCloss<-paste0("Methylation data QC: ", 
+                 N_M_count("probe_fail_10==T & childteen!='M'", " with >10% of probes failing dropped, \n ", pdqc_all),
+                 N_M_count("sex_new!=predicted_sex & probe_fail_10==F & childteen!='M'", 
+                           "  with discordant sex samples dropped, \n ", pdqc_all),
+                 N_M_count("predicted_sex_outlier==T & sex_new==predicted_sex & probe_fail_10==F & childteen!='M'", 
+                           ' with abnormal sex chromosome intensity dropped & \n ', pdqc_all),
+                 length(reps_remove), " duplicate samples dropped")
 
-# Supplemental Figure 2 - Choice of beta methylation transformation
+n_methylFF<-paste0("Cohort with quality controlled 450K data \n","n=", myFF %>% dplyr::select(id) %>% unique() %>% nrow(), " individuals with m=", myFF%>%nrow(), " samples")
 
-# Supplemental Figure 3 - Distribution and correlation of methylation summary measures
+n_missingmaternalprenataldata=paste0('Missing maternal data: \n', 
+                                     N_M_count("is.na(smkPreg_binary)", " with missing prenatal maternal smoking \n ", myFF), 
+                                     N_M_count("!is.na(smkPreg_binary) & is.na(m1g2_YesNoPreg)", " with missing prenatal alcohol \n ", myFF), 
+                                     N_M_count("!is.na(smkPreg_binary) & !is.na(m1g2_YesNoPreg) & is.na(m1g3_YesNoPreg)", 
+                                               " with missing prenatal other drug use \n ", myFF),
+                                     N_M_count("!is.na(smkPreg_binary) & !is.na(m1g2_YesNoPreg) & !is.na(m1g3_YesNoPreg) & is.na(cm1inpov)", 
+                                               " with missing maternal income to poverty ratio data \n ", myFF), 
+                                     N_M_count("!is.na(smkPreg_binary) & !is.na(m1g2_YesNoPreg) & !is.na(m1g3_YesNoPreg) & !is.na(cm1inpov) &
+                                               PostnatalMaternalSmokingAny=='Missing'", " with missing postnatal maternal smoking data \n ", myFF), 
+                                     N_M_count("!is.na(smkPreg_binary) & !is.na(m1g2_YesNoPreg) & !is.na(m1g3_YesNoPreg) & !is.na(cm1inpov) &
+                                               PostnatalMaternalSmokingAny!='Missing' & SmkAtVisitPastmonth=='Missing'", 
+                                               " with missing past month maternal/primary care giver smoking packs/day", myFF))
 
-#Supplemental Figure 4 - Change in methylation summary measures between visits
+maternaldata=myFF %>% filter(!is.na(smkPreg_binary) & !is.na(m1g2_YesNoPreg) & !is.na(m1g3_YesNoPreg) & !is.na(cm1inpov) & PostnatalMaternalSmokingAny!='Missing' & SmkAtVisitPastmonth!='Missing')  
+n_withmaternaldata=paste0('m=', maternaldata %>% nrow(), ' samples & n=', maternaldata %>% dplyr::select(idnum)%>%unique()%>%nrow(), ' individuals with maternal covariate data')
 
-# Figure 2: Polymethylation scores are correlated and stable between age 9 and age 15 in a diverse sample of `r n_total`US children
+n_missingchilddata=paste0('Missing child data: \n', 
+                          N_M_count("is.na(ancestry) | ancestry=='Missing PC data'", 
+                                    " with missing ancestry/genetic data  \n ", maternaldata), 
+                          N_M_count("!is.na(ancestry) & ancestry!='Missing PC data' & is.na(cm1bsex)", 
+                                    " with missing child sex data \n ", maternaldata), 
+                          N_M_count("is.na(ancestry) | ancestry=='Missing PC data' & !is.na(cm1bsex) & is.na(ChildAgeComposite)", 
+                                    " with missing child age at sample \n ", maternaldata))
+
+childdata= maternaldata %>% filter(ancestry!='Missing PC data' & !is.na(cm1bsex) & !is.na(ChildAgeComposite))
+n_withchilddata=paste0('m=', childdata %>% nrow(), ' samples & n=', childdata %>% dplyr::select(idnum)%>%unique()%>%nrow(), ' individuals with child covariate data')
+
+n_Child<-paste0('n=', childdata %>%filter(childteen=='C')  %>% nrow(), " individuals with a 9 year visit")
+n_ChildSmkexcluded<-paste0("Exclude children who smoke: \n",'n=', childdata%>%filter(childteen=='C' & k5f1l=='2 no')%>% nrow(), 
+                           " individuals with a 9 year visit")
+n_ChildSmkexclusions<-paste0('n=', childdata%>%filter(k5f1l!= "2 no" & childteen=='C')%>%nrow(), 
+                             " individuals smoking \n or missing child smoking data \n at 9 year visit")
+n_Teen<-paste0('n=', childdata %>%filter(childteen=='T')   %>% nrow(), " individuals with a 15 year visit")
+n_TeenSmkexcluded<-paste0("Exclude children who smoke: \n",'n=', childdata %>%filter(childteen=='T' & k6d40=="2 No") %>% nrow(), 
+                          " individuals with a 15 year visit")
+n_TeenSmkexclusions<-paste0('n=', childdata %>% filter(k6d40!="2 No" & childteen=='T') %>% nrow,
+                            " individuals smoking \n or missing focal child smoking data \n at 15 year visit") 
+if(childdata %>% filter((childteen=='C' & k5f1l=='2 no') | (childteen=='T' & k6d40=='2 No'))%>%nrow()!=nrow(modeldata)){stop('filter N and completecase data set N do not match')}
+
+```
+
+```{r}
+graph1<-DiagrammeR::grViz("
+   digraph graph2 {
+   compound=true;
+   graph [layout = dot]
+
+   node [shape = rectangle, width=4]
+   a [label = '@@1']
+   b [label = '@@2']
+   c [label = '@@3']
+   d [label = '@@4']
+   e [label = '@@5']
+   f [label = '@@6']
+   g [label = '@@7']
+   h [label = '@@11']
+   i [label = '@@12']
+
+   #fake nodes for connecting to edges
+   node [shape=point, width=0.01, height=0.01]
+   a1
+   b1
+   c1
+   d1
+   f1 [group=g1]
+   g1
+   h1
+
+  #exclusion nodes
+  node [shape = rectangle, width=4]
+  aa [label = '@@8']
+  bb [label = '@@9']
+  cc [label = '@@10']
+  hh [label = '@@13']
+  ii [label = '@@14']
+
+  #draw connections
+#  a->m
+#  m->a1 [arrowhead=none]
+  a->a1 [arrowhead=none]
+  a1->b
+  b->b1 [arrowhead=none]
+  b1->c
+  c->c1 [arrowhead=none]
+  c1->d
+  d->d1 [arrowhead=none]
+  d1->e
+  e->f; e->g
+  f->g1 [arrowhead=none]; g->h1 [arrowhead=none]
+  g1->h; h1->i
+
+{rank=same; b1->aa [minlen=2]}   
+{rank=same; c1->bb [minlen=2]} 
+{rank=same; d1->cc [minlen=2]}
+{rank=same; g1->hh [minlen=2]}
+{rank=same; h1->ii [minlen=2]}
+
+#invisible constraints to force ordering
+#ff->i [style=invis]
+   }
+
+   [1]: n_fullFF
+   [2]: n_methylQC
+   [3]: n_methylFF
+   [4]: n_withmaternaldata
+   [5]: n_withchilddata
+   [6]: n_Child
+   [7]: n_Teen
+   [8]: n_QCloss
+   [9]: n_missingmaternalprenataldata
+   [10]: n_missingchilddata
+   [11]: n_ChildSmkexcluded
+   [12]: n_TeenSmkexcluded
+   [13]: n_ChildSmkexclusions
+   [14]: n_TeenSmkexclusions
+")
+
+graph1
+```
+
+# Supplemental Table 1 - Descriptive statistics of excluded vs included samples with methylation data available from the Fragile and Families and Child Wellbeing study
+
+```{r}
+myFF$MethylData2<-ifelse(myFF$MethID %in% modeldata$MethID, 'Final analysis group', 'Not in final analysis group')
+set_label(myFF$MethylData2)<-'Exclusions based on missing data'
+myFF=left_join(myFF, methyldata%>%mutate(idnum=as.character(idnum)))
+methyl2<-createTable(compareGroups(MethylData2~smkPreg_binary+m1g2_YesNoPreg+m1g3_YesNoPreg+cm1inpov+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+ancestry+cm1bsex+globalmethylation+pediatric+Epithelial.cells_saliva+Leukocytes_saliva+SSnewbornCT_center+SSolder_center+cg05575921, data=myFF, include.miss = T), show.all=T)
+export2md(strataTable(methyl2, 'childteen'), caption="Exclusion table of subset with quality controlled methylation data vs final analytic subset for main predictor, outcomes and key demographic variables")
+```
+# Supplemental Figure - Comparison of choice of A) regression coefficient and B) transform of methylation beta value on distributions of polymethylation scores
+
+```{r}
+scores=gsub('_center', '', colnames(modeldata)[str_detect(colnames(modeldata), 'center')])
+age_long_score=modeldata%>%
+  dplyr::select(any_of(c('idnum', 'childteen', 'ancestry', 'smkPreg_binary')),
+                contains(scores))%>%
+  pivot_longer(cols=contains(scores),
+               names_to='Scores',
+               values_to='Polymethylation score')
+
+plot_grid(
+  age_long_score%>%filter(str_detect(Scores, 'no')) %>% 
+    ggplot(aes(x=`Polymethylation score`))+geom_histogram()+
+    facet_wrap(~Scores, labeller=labeller(Scores=get_label(modeldata%>%dplyr::select(contains('notransform'))))), 
+  age_long_score%>%filter(str_detect(Scores, 'CT')) %>% 
+    ggplot(aes(x=`Polymethylation score`))+geom_histogram()+
+    facet_wrap(~Scores, scales='free', ncol=3,
+               labeller=labeller(Scores=get_label(modeldata%>%dplyr::select(contains('CT'))))),
+  labels=c('A', 'B'), nrow=2
+)
+
+```
+
+# Supplemental Figure 2 - Correlation of methylation summary measures
+
+```{r}
+library(Hmisc)
+library(corrplot)
+res2<-modeldata%>%dplyr::select(any_of(all_methyl))%>%cor()
+#labels
+colnames(res2)=all_methyl_labels
+rownames(res2)=all_methyl_labels
+corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8)
+```
+
+#Supplemental Figure 3 - Correlation between methylation summary measures between visits
+
+```{r}
+age_long_methyl=modeldata%>%
+  dplyr::select(any_of(c('idnum', 'childteen', 'ancestry', 'smkPreg_binary', all_methyl)))%>%
+  pivot_longer(cols=any_of(all_methyl),
+               names_to='Methylation summary measure type',
+               values_to='Methylation summary measure')%>%
+  mutate(Methyl_factor=factor(`Methylation summary measure type`, 
+                              levels=all_methyl_levels))
+
+age_wide_methyl=age_long_methyl%>%
+  pivot_wider(names_from=childteen, 
+              values_from=`Methylation summary measure`)
+
+age_wide_methyl%>%
+  ggplot(aes(x=C, y=T))+geom_point()+stat_cor()+geom_smooth(method=lm)+theme_classic()+
+  scale_x_continuous('Age 9 score')+scale_y_continuous('Age 15 score')+
+  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels))
+```
+
+# Supplemental Figure XX: Change in methylation summary measures between age 9 and age 15
+
+```{r}
+age_long_methyl%>%
+  ggplot(aes(x=childteen, y=`Methylation summary measure`, color=childteen))+
+  geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+
+  scale_color_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+
+  scale_x_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+
+  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels))+
+  theme(legend.position = 'bottom')
+
+age_long_methyl%>%
+  ggplot(aes(x=childteen, y=`Methylation summary measure`, group=idnum))+
+  geom_point()+geom_line(alpha=0.1, color='grey')+geom_smooth(aes(group=1), method='lm', fill='blue')+
+  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels))+
+  theme_classic()+scale_x_discrete('', labels=c('Age 9', 'Age 15'))
+```
+
+# Figure 2: Polymethylation scores are correlated and stable between age 9 and age 15 in a diverse sample of `r n_total` US children
 
@@
 
+#add line for mean ? 
 outcome_labels=c('Global methylation', 'Pediatric clock', 'AHHR: cg05575921', 'Polymethylation score \n (newborn, mean centered)', 'Polymethylation score  \n (older children, mean centered')
@@
 
-fig2a=modeldata%>%
-  dplyr::select(idnum, childteen, SSolder_center, SSnewbornCT_center, 
-                cg05575921, globalmethylation, pediatric)%>%
-  pivot_longer(cols=c(globalmethylation, pediatric, SSnewbornCT_center, SSolder_center, cg05575921),
-               names_to='Methylation summary measure type',
-               values_to='Methylation summary measure')%>%
-  pivot_wider(names_from=childteen, 
-              values_from=`Methylation summary measure`)%>%
-  mutate(meth_f=factor(`Methylation summary measure type`, 
-                       levels=c('SSnewbornCT_center', 'SSolder_center', 
-                                'cg05575921',  'globalmethylation', 'pediatric')))%>%
+fig2a=age_wide_methyl%>%filter(Methyl_factor%in% main_methyl)%>%
   ggplot(aes(x=C, y=T))+geom_point()+stat_cor()+geom_smooth(method=lm)+theme_classic()+
   scale_x_continuous('Age 9 score')+scale_y_continuous('Age 15 score')+
-  facet_wrap(~meth_f, scales='free', labeller = labeller(meth_f=outcome_labels), nrow=1)
+  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels), nrow=1)
 
-fig2b=modeldata%>%
-  dplyr::select(idnum, childteen, SSolder_center, SSnewbornCT_center, 
-                cg05575921, globalmethylation, pediatric)%>%
-  pivot_longer(cols=c(globalmethylation, pediatric, SSnewbornCT_center, SSolder_center, cg05575921),
-               names_to='Methylation summary measure type', 
-               values_to='Methylation summary measure')%>%
-  pivot_wider(names_from=childteen,
-              values_from=`Methylation summary measure`)%>%
-  mutate(meth_f=factor(`Methylation summary measure type`, 
-                       levels=c('SSnewbornCT_center', 'SSolder_center', '
-                                cg05575921',  'globalmethylation', 'pediatric')),
-         diff=T-C)%>%
+mean_summary=age_wide_methyl%>%filter(Methyl_factor%in% main_methyl)%>%mutate(diff=T-C)%>%group_by(Methyl_factor)%>%summarise(mean_diff=mean(diff, na.rm=T))
+
+fig2b=age_wide_methyl%>%filter(Methyl_factor%in% main_methyl)%>%mutate(diff=T-C)%>%
   ggplot(aes(x=diff))+geom_histogram()+
-  facet_wrap(~meth_f, scales='free_x', labeller = labeller(meth_f=outcome_labels), nrow=1)+
-  theme_classic()+xlab('Difference in measure (Age 15- Age 9)')+
-  geom_vline(aes(xintercept=0), color='red', linetype=2)+geom_rug()
+  facet_wrap(~Methyl_factor, scales='free_x', labeller = labeller(Methyl_factor=all_methyl_labels), nrow=1)+
+  theme_classic()+xlab('Difference in measure (Age 15 - Age 9)')+
+  geom_vline(aes(xintercept=0), color='red', linetype=2)+geom_rug()+
+  geom_vline(data=mean_summary, aes(xintercept=mean_diff), color='blue')
 
@@
 
-```{r temp relabel varables}
-set_label(modeldata$cg05575921)<-'AHHR: cg05575921'
-set_label(modeldata$SmkAtVisitPastmonth)<-'Past month maternal/primary caregiver smokin (pack/day)'
-set_label(modeldata$PostnatalMaternalSmokingAny)<-'Any maternal postnatal smoking when child 1 and 5 years of age'
-modeldata$childteen=factor(case_when(modeldata$childteen=='C'~'Age 9', 
-                              modeldata$childteen=='T'~'Age 15'), levels=c('Age 9', 'Age 15'))
+```{r bivariate table}
+export2md(strataTable(createTable(compareGroups(smkPreg_binary~m1g2_YesNoPreg+m1g3_YesNoPreg+cm1inpov+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+ancestry+cm1bsex+globalmethylation+pediatric+Epithelial.cells_saliva+Leukocytes_saliva+SSnewbornCT_center+SSolder_center+cg05575921, data=modeldata%>%mutate(childteen=factor(case_when(childteen=='C'~'Age 9', childteen=='T'~ 'Age 15'), levels=c('Age 9', 'Age 15'))), include.miss = T)), 'childteen'))
 ```
 
-```{r bivariate table}
-export2md(strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+pediatric+SSnewbornCT_center+cg05575921+Epi+Fib+IC+ChildAgeComposite+cm1inpov+ancestry+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth, data=modeldata)), 'childteen'))
+# Supplemental Figure: *A priori* CpG sites methylation and prenatal maternal smoking by age and ancestry
+
+# Supplemental Figure - A) Ancestry and B) age differences in methylation summary measures by prenatal maternal smoking exposure
+
+```{r}
+plot_grid(
+  age_long_methyl%>%filter(Methyl_factor %in% main_methyl)%>%
+    ggplot(aes(x=smkPreg_binary, y=`Methylation summary measure`, color=ancestry))+
+    geom_violin()+geom_boxplot(width=0.2, position=position_dodge(0.9))+stat_compare_means(label="..p.signif..")+
+    scale_color_discrete(name='Ancestry')+
+    scale_x_discrete(name='Prenatal smoke exposure', labels=c('No', 'Yes'))+
+    facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels))+
+    theme(legend.position = 'bottom'),
+
+age_long_methyl%>%filter(Methyl_factor %in% main_methyl)%>%
+  ggplot(aes(x=smkPreg_binary, y=`Methylation summary measure`, color=childteen))+
+  geom_violin()+geom_boxplot(width=0.2, position=position_dodge(0.9))+stat_compare_means(label="..p.signif..")+
+  scale_color_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+
+  scale_x_discrete(name='Prenatal smoke exposure', labels=c('No', 'Yes'))+
+  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels))+
+  theme(legend.position = 'bottom'),
+
+labels=c('A', 'B'), nrow=1)
 ```
-
-#Supplemental Table XX - Bivariate associations between prenatal maternal smoking and covariates in ancestry strata
 
@@
 
-my_names=c('Methylation measure', paste0(c(paste(c('Age 9 model \n n=', 'Age 15 model \n n='), modeldata%>%group_by(childteen)%>%summarise(n=n())%>%pull(n)), paste0('Longitudinal models <br> n=', n_total)), '<br> Beta (95% CI); p-value'))
+test=rbind(global_models, longitudinal_global_models %>% mutate(childteen='Longitudinal'))%>%
+  filter(predictors=='secondhand smoke exposure model')%>%
+  filter(outcome %in% my_main_outcomes)%>%
+  mutate(childteen=factor(case_when(childteen=='C'~'Age 9 models', 
+                             childteen=='T'~'Age 15 models', 
+                             childteen=='Longitudinal'~'Longitudinal models'),levels=c('Age 9 models', 'Age 15 models', 'Longitudinal models')), 
+         tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
+                                        show_single_row='smkPreg_binary', 
+                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
 
-tb2=rbind(global_models, longitudinal_global_models %>% mutate(childteen='Longitudinal'))%>%filter(predictors=='secondhand smoke exposure model')%>%filter(outcome %in% my_main_outcomes)%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%mutate(estimate=paste0(my_round(estimate, 2), ' (', my_round(conf.low, 2), ', ', my_round(conf.high, 2), '); ', signif(p.value, 2)))%>%dplyr::select(childteen, outcome, estimate)%>%pivot_wider(names_from=childteen, values_from=c(estimate))
-names(tb2)=my_names
+table_list=test %>%split(f=.$childteen)%>%map(~tbl_merge(.x$tbl, 
+                                                         tab_spanner = as.character(.x$outcome)))
 
-tb2%>%kable(., escape=F)%>%kable_minimal()%>%
-  kableExtra::scroll_box(width = "100%")%>% footnote(general = "All models adjusted for:child sex, mother income to poverty ratio at birth, epithelial and immune cell proportions (estimated from methylation data), maternal other drug and other alcohol use during pregnancy, batch of array, genetic ancestry (1st two principal components of genetic data), postnatal smoke exposure (any postnatal maternal smoking and maternal packs per day in month prior to visit).")%>%kable_styling(font_size = 10)
+regression_table_age=tbl_stack(table_list, group_header=names(table_list))
+
 ```
 
-```{r, width=11}
+```{r}
+test=rbind(longitudinal_global_models %>% mutate(ancestry='Unstratified'),
+           longitudinal_local_models%>%dplyr::select(-data))%>%
+  filter(predictors=='secondhand smoke exposure model')%>%
+  filter(outcome %in% my_main_outcomes)%>%
+  mutate(tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
+                                        show_single_row='smkPreg_binary', 
+                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
+
+table_list=test %>%split(f=.$ancestry)%>%map(~tbl_merge(.x$tbl, 
+                                                         tab_spanner = as.character(.x$outcome)))
+
+regression_table_ancestry=tbl_stack(table_list, group_header=names(table_list))
+
+```
+
+```{r, width=11, eval=F}
 #alternatively 
@@
   theme(text = element_text(size=16))+theme(legend.position = 'bottom')+
-  scale_color_discrete('', limits=c('Age 9', 'Age 15'))+
   scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
-  scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))
+  scale_y_discrete('', limits=c('european', 'hispanic', 'african', 'Unstratified'), labels=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))
 ```
@@
 
+```{r}
+#bring the heatmap function
+covar=c('Sample_Plate', 'Slide', 'Array', 'cm1bsex', 'smkPreg_binary', 'childteen', 'ancestry')
+
+bring.the.heatmap.sv <- function(pd,covar,folder){
+
+  nsv <- length(grep('sv',names(pd)))
+
+  #generate correlation matrix between SVs and variables
+  correlations <- matrix(0,nrow=length(covar),ncol=nsv)
+  colnames(correlations) <- paste('sv',1:nsv,sep='')
+  rownames(correlations) <- covar
+
+  #take svs
+  svs <- pd[,colnames(correlations)]
+
+  #correlations between svs and variables:
+  # cor test if continuous
+  # anova is multi category
+  # t-test if two category
+  for(col in covar){
+    var <- pd[,col]
+    #check to see if it is numeric and isn't really a two category thing
+    if(class(var)=='numeric' & dim(table(var))>2){
+      correlations[col,] <- apply(svs,2,FUN = function(x,y) cor.test(x,y)$p.value,y=var)
+    }else{
+      if(length(levels(factor(var)))>2){
+        correlations[col,] <- apply(svs,2,FUN = function(x,y) anova(lm(x~y))$'Pr(>F)'[1],y=var)
+      }else{
+        correlations[col,] <- apply(svs,2,FUN = function(x,y) t.test(x~y)$p.value,y=var)
+      }
+    }
+  }
+  color.gradient <- colorRampPalette(c('firebrick','ivory3','royalblue'),bias=3)
+  correlations=correlations%>%as.data.frame()%>%tibble::rownames_to_column('var1')%>%
+    pivot_longer(-var1, names_to='var2')%>%
+    mutate(var2=factor(var2, levels=unique(mixedsort(var2))),
+           var1=factor(var1, levels=pd[, covar]%>%colnames(), labels=get_label(pd[, covar])))
+  correlations%>%
+    ggplot(aes(x=var1, y=var2, fill=value))+geom_tile()+
+    scale_fill_gradientn(colours = color.gradient(20), values = seq(0, 1, 0.5))+
+    scale_x_discrete(labels=str_wrap(get_label(pd[, covar]), 10))
+}
+
+bring.the.heatmap.sv(modeldata, covar=c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'ancestry', 'cm1bsex', 'Leukocytes_saliva', 'Epithelial.cells_saliva', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))
+```
+
 # Supplemental Table XX: Surrogate variable models
+
+```{r}
+test=rbind(global_models, longitudinal_global_models %>% mutate(childteen='Longitudinal'))%>%
+  filter(predictors=='surrogate variable models')%>%
+  filter(outcome %in% my_main_outcomes)%>%
+  mutate(childteen=factor(case_when(childteen=='C'~'Age 9 models', 
+                             childteen=='T'~'Age 15 models', 
+                             childteen=='Longitudinal'~'Longitudinal models'),levels=c('Age 9 models', 'Age 15 models', 'Longitudinal models')), 
+         tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
+                                        show_single_row='smkPreg_binary', 
+                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
+
+table_list=test %>%split(f=.$childteen)%>%map(~tbl_merge(.x$tbl, 
+                                                         tab_spanner = as.character(.x$outcome)))
+
+regression_table_age_svs=tbl_stack(table_list, group_header=names(table_list))
+
+test=rbind(longitudinal_global_models %>% mutate(ancestry='Unstratified'),
+           longitudinal_local_models%>%dplyr::select(-data))%>%
+  filter(predictors=='surrogate variable models')%>%
+  filter(outcome %in% my_main_outcomes)%>%
+  mutate(tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
+                                        show_single_row='smkPreg_binary', 
+                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
+
+table_list=test %>%split(f=.$ancestry)%>%map(~tbl_merge(.x$tbl, 
+                                                         tab_spanner = as.character(.x$outcome)))
+
+regression_table_ancestry_svs=tbl_stack(table_list, group_header=names(table_list))
+```
 
@@
 
-names(my_outcomes)=c( 'globalmethylation', 'pediatric','cg05575921', 'SSnewbornCT_center')
-my_outcomes=c('Global methylation', 'Pediatric clock', "AHHR: cg05575921", 'Polymethylation score \n (newborn, mean centered)')
+age_labels=c('Age 9', 'Age 15'); names(age_labels)=c('C', 'T')
+methyl_labels_roc=paste0('Base model+', all_methyl_labels)
+names(methyl_labels_roc)=paste0('Base model+', names(all_methyl_labels))
 
 fig4a= global_roc%>%
-  dplyr::select(predictors, covariates, methylation, childteen, test)%>%
-  mutate(modeltype=ifelse(str_detect(predictors, '\\+'), covariates, 'No covariates'))%>%
+    dplyr::select(predictors, covariates, methylation, childteen, test, modeltype)%>%
     unnest(test) %>% 
-  filter(str_detect(methylation, 'SSnewbornCT|cg055|global|pediatric') &
-           covariates=='')%>%
-  ggplot(aes(x=spec, y=sens, color=methylation))+geom_line()+
-  facet_wrap(childteen~., labeller=labeller(childteen=age_labels), nrow=1)+
-  scale_color_discrete('', labels=my_outcomes)+
-  theme_classic()+theme(legend.position = 'right')+guides(color=guide_legend(nrow=2, byrow=TRUE))+
+    filter(str_detect(methylation, 'SSnewbornCT|cg0557|global') | modeltype=='Base model')%>%
+    ggplot(aes(x=spec, y=sens, color=modeltype))+geom_line()+
+    facet_wrap(childteen~., nrow=1, labeller=labeller(childteen=age_labels))+
+    theme_classic()+
+    scale_color_manual('', values=c('red', 'darkred', 'darkblue', 'darkgreen', 'blue', 'green'))+
+    theme(legend.position = 'right')+guides(color=guide_legend(nrow=2, byrow=TRUE))+
     geom_abline(intercept=1, slope=-1, color='black')+
@@
   dplyr::select(covariates, methylation, childteen, test)%>%unnest(test) %>% 
-  filter(str_detect(methylation, 'SSnewbornCT') 
-         & covariates=='')%>%
+  filter(str_detect(methylation, 'SSnewbornCT'))%>%
   rbind(long_global_roc%>% 
@@
           filter(str_detect(methylation, 'SSnewbornCT') 
-                 & covariates==''))%>%
+                 & covariates!=''))%>%
   ggplot(aes(x=spec, y=sens, color=childteen, linetype=childteen))+geom_line(size=2)+
   facet_wrap(~methylation, labeller = labeller(methylation=c('SSnewbornCT_center'='Polymethylation score, \n newborn mean centered')))+
-  scale_color_discrete(labels=my_outcomes)+
   theme_classic()+theme(legend.position = 'right')+guides(color=guide_legend(nrow=2,byrow=TRUE))+
@@
 
-add p value comparing base model to adding in curves
+```{r}
+roc_table=global_roc %>% dplyr::select(childteen, methylation, modeltype, roc, auc)
+
+roc_base=roc_table %>% filter(modeltype=='Base model+No methylation')%>%mutate(roc_basemodel=roc)%>%dplyr::select(childteen, roc_basemodel)
+roc_ss=roc_table %>% filter(modeltype=='Sustained smoking w/ cell type correction (newborn)')%>%mutate(roc_pms=roc)%>%dplyr::select(childteen, roc_pms)
+
+roc_table1=roc_table%>%filter(!str_detect(modeltype, 'Base model'))%>%left_join(roc_ss)%>%mutate(roctest=map2(roc, roc_pms, ~roc.test(.x, .y)))
+roc_table1%>%mutate(`pval comparison to sustained smoking score (newborns)`=map_dbl(roctest, ~.x$p.value))%>%dplyr::select(childteen, modeltype, auc, `pval comparison to sustained smoking score (newborns)`)
+
+roc_table2=roc_table%>%filter(str_detect(modeltype, 'Base model'))%>%left_join(roc_base)%>%mutate(roctest=map2(roc, roc_basemodel, ~roc.test(.x, .y)))
+roc_table2%>%mutate(`pval comparison to base model`=map_dbl(roctest, ~.x$p.value))%>%dplyr::select(childteen, modeltype, auc, `pval comparison to base model`)
+```
 

### 2021-10-14 Scripts/03_sva_variables.R

@@ -15,7 +15,7 @@ output.dir=paste0(data.dir, 'CreatedData/')
 #as of october 4 switched to beta file with  423668 probes (see Jonah email search Check in on sex filtration/clocks Fragile Families)
 betaqc<-readRDS(file=paste0(data.dir, 'OGData/', "betaqc.rds"))
 
-load(paste0(datadir, '/CreatedData/completeCasepheno.Rdata'))
+load(paste0(data.dir, '/CreatedData/completeCasepheno.Rdata'))
 pheno<-completecase # dataset with the most exclusion i.e. squaring dataset off for all models (excluding kids who smoke in figure)
 
 #make a matrix of model data 

### 2021-10-14 Scripts/04_summary_methyl_models.R

@@ -30,7 +30,7 @@ datadir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/CreatedData"
 outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output"
 #read data 
 load(paste0(datadir, '/completeCasemethyl.Rdata'))
-modeldata=completecase%>%droplevels()
+modeldata=completecase%>%droplevels()%>%copy_labels(completecase)
 
 #set as factors 
 modeldata$PostnatalMaternalSmokingAny<-factor(modeldata$PostnatalMaternalSmokingAny, 
@@ -40,14 +40,18 @@ modeldata$SmkAtVisitPastmonth<-factor(modeldata$SmkAtVisitPastmonth,
                                       levels=c("No smoking", 
                                                "Less than pack a day", 
                                                "Pack or more a day", "Missing"))
-modeldata$slide<-factor(modeldata$slide)
+#modeldata$slide<-factor(modeldata$slide)
 modeldata$Sample_Plate<-factor(modeldata$Sample_Plate)
 
 #sv data - no variance filter, whole subset
-load(paste0(datadir, '/svNone.Rds'))
+sv.vNone=readRDS(paste0(datadir, '/svNone.Rds'))
 is_sv=function(x) str_replace_all(x, " ", "")
 sv.vNone=sv.vNone%>%rename_with(is_sv, contains('sv'))
-modeldata=left_join(modeldata, sv.vNone)
+modeldata=left_join(modeldata, sv.vNone)%>%copy_labels(modeldata)
+set_label(modeldata$Sample_Plate)='Batch'
+set_label(modeldata$PostnatalMaternalSmokingAny)='Any postnatal maternal smoking (Age 1 & 5)'
+set_label(modeldata$SmkAtVisitPastmonth)='Primary care giver smoke past month dose'
+save(modeldata, file=file.path(datadir, file='completeCaseSVA.Rdata'))
 
 #################################Set variables for modeling
 #outcomes & outcome_labels
@@ -183,8 +187,15 @@ longitudinal_local_models=my_clean_argnames(longitudinal_local_models, local_pre
 modeldata=modeldata%>%mutate(smkPreg_binaryN=case_when(smkPreg_binary=='Yes'~1, smkPreg_binary=='No'~0))
 
 #global predictors, cross sectional
-roc_predictors=gsub('~smkPreg_binary', '', paste0(base_model_vars, global_pcs))
[…]

### 2021-10-14 Scripts/05_ewas.R

@@ -1,0 +1,85 @@
+#ewas
+library(dplyr)
+library(furrr) 
+library(purrr)
+library(lme4) 
+library(lmerTest)
+library(matrixStats)
+library('variancePartition')
+library('edgeR')
+library('BiocParallel')
+
+datadir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data"
+outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output"
+
+#variance trim of beta matrix
+betaqc<-readRDS(file=file.path(datadir, 'OGData', "betaqc.rds"))
+varCut<-0.0002
+betaqc<-betaqc[rowVars(betaqc)>varCut,]
+dim(betaqc)
+
+#filter to only samples in final analytic subsample
+load(file=file.path(datadir, 'CreatedData', 'completeCaseSVA.Rdata'))
+betaqc<-betaqc[, colnames(betaqc) %in% modeldata$MethID]
+
+#match rows to columns
+modeldata=modeldata[match(colnames(betaqc), modeldata$MethID),]
+#rownames of modeldata 
+rownames(modeldata)=modeldata$MethID
+
+#functions for  EWAS 
+model_lmm<-function(y, local=F, model='1'){
+  #create new variable for PCS
+  if(local==F){modeldata=modeldata %>% mutate(PC1=global_PC1, PC2=global_PC2)}
+  if(local==T){modeldata=modeldata %>% mutate(PC1=local_PC1, PC2=local_PC2)}
+  if(model=='1'){my_model=suppressMessages(suppressWarnings(lmer(y~smkPreg_binary+PC1+PC2+
+                                                    cm1bsex+cm1inpov+ChildAgeComposite+Leukocytes_saliva+(1|id), data=modeldata, REML=F)))}
+  if(model=='2'){my_model=suppressMessages(suppressWarnings(lmer(y~smkPreg_binary+PC1+PC2+
+                                                                   cm1bsex+cm1inpov+ChildAgeComposite+Leukocytes_saliva+
+                                                                   m1g2_YesNoPreg+m1g3_YesNoPreg+(1|id), data=modeldata, REML=F)))}
+  if(model=='3'){my_model=suppressMessages(suppressWarnings(lmer(y~smkPreg_binary+PC1+PC2+
+                                                          cm1bsex+cm1inpov+ChildAgeComposite+Leukocytes_saliva+
[…]

### 2021-10-26 Papers/Polymethylation_scores/04_background.Rmd
wrote discussion and methods

@@
   html_document: default
+bibliography: Methylation_SmokingBib.bib
 ---
+
 # Background
 
-Maternal prenatal smoking is a public health concern: in 2018, 11% of pregnant US women aged 15-44 reported any past-month cigarette use, and exposed children are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma.
+Maternal prenatal smoking is a public health concern: in 2018, 11% of pregnant US women aged 15-44 reported any past-month cigarette use [@SAMHSA2018], and exposed children are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu].
 Measuring prenatal maternal smoking behavior is challenging.
-Maternal prenatal smoking is underreported due to stigma.
-The gold-standard for smoking measurements, serum cotinine, has a short half-life of nine hours in pregnant women.
+Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua].
+The gold-standard for smoking measurements, serum cotinine, has a short half-life of nine hours in pregnant women [@Dempsey:2002wr].
 Further, serum cotinine measures during pregnancy are rarely available outside of birth cohorts, limiting clinical application as a marker of prenatal exposures.
-Developing a persistent, reliable and feasible biomarker of prenatal maternal smoke exposure is critical for advancing research and precision health.
+Developing a portable and reliable biomarker of prenatal maternal smoke exposure would therefore have important implications for research and clinical practice.
 
 DNA methylation has been associated with prenatal maternal smoking at various cross-sectional time points across the life course in primarily European ancestry groups.
-Cordblood, peripheral blood and placental DNA methylation has been significantly associated with prenatal maternal smoking.
-[Joubert 2016; Bustamante 2021; Davey Smith 2018].
-Blood DNA methylation signatures in cohorts of older children and adults also were significantly associated with prenatal maternal smoking exposure [Terry 2019, Davey Smith 2018].
+Cordblood, peripheral blood and placental DNA methylation has been significantly associated with prenatal maternal smoking [@Joubert:2016ue; @Everson:2021td; @Richmond:2018tu; @Cardenas2019].
+Blood DNA methylation signatures in cohorts of older children and adults also were significantly associated with prenatal maternal smoking exposure [@Tehranifar:2018tb ; @Richmond:2018tu].
 Yet few studies have examined associations at different ages in the same participants.
-Additionally, while associations are well-replicated in cohorts of European ancestry, few studies have used non-European ancestry populations and those that have had small sample sizes: Fallin 2016 186 individuals of African or admixed ancestry, Terry 2018 60% African or hispanic but only 89 participants, Burchard 2019 572 Latino children].
-DNA methylation is associated with prenatal maternal smoking, but the consistency of this association across age and ancestry is understudied.
+Additionally, while associations are well-replicated in cohorts of European ancestry, only a few, small sample size studies are in non-European ancestry populations.
+A 2016 study with 186 participants of African or admixed ancestry [@Ladd-Acosta:2016aa], while a 2018 study of 89 participants had 60% of African or admized ancestry [@Tehranifar:2018tb].
+A larger 2019 study of 572 Latino children replicated associations with prenatal maternal smoking at 7 out of 148 tested *a priori* CpG sites from Joubert et al.
+at a FDR-adjusted *P*-value\<0.05 [@Neophytou:2019ty; @Joubert:2016ue].
+Thus, while DNA methylation is associated with prenatal maternal smoking, the consistency of this association across age and ancestry is understudied.
 
@@
 Importantly, changes in DNA methylation are specific to smoke exposure *in utero*.
-Previous work has found reasonable performance of cord blood based [London 2017] (accuracy 91%, sens 58%, spec 97%) [Fallin 2016] (81% accuracy) and peripheral blood based biomarkers (AUC: 0.72) [Davey Smith 2018] of prenatal maternal smoking.A 2018 study found that DNA methylation based scores could even predict predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [Davey Smith 2018].
-However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across age and ancestry has not been evalutated.
+Previous work has found reasonable performance of cord-blood accuracy ranging from 81% to 91%) [@Reese:2017ts; @Ladd-Acosta:2016aa].
+A 2018 study found that DNA methylation based scores from adult peripheral blood could even predict predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu].
+However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across age and ancestry has not been evaluated.
 Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times or evaluated if the inclusion of multiple time points improves performance.
 Additionally, previous work has relied on blood and placental samples.
+DNA methylation varies across tissues, an epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples[@Everson:2021td].
 Cord blood or placenta samples are unlikely to be available in a clinical setting.

### 2021-10-26 Papers/Polymethylation_scores/index_paper.Rmd
wrote discussion and methods

--- 2021-09-14 Papers/Polymethylation_scores/index_paper.Rmd
@@
 date: "9/13/2021"
-output: pdf_document
-bib: Methylation_SmokingBib.bib
+output:
+  word_document: default
+  html_document:
+    df_print: paged
+bibliography: Methylation_SmokingBib.bib
+csl: springer-vancouver.csl
 ---
@@
 
+```{r child = here::here("Papers", "Polymethylation_scores", "08_methods.Rmd")}
+```
+
 ```{r child = here::here("Papers", "Polymethylation_scores", "98_declarations.Rmd")}

### 2021-10-26 Reports/FFEWAS.Rmd
results for oct 25 FF meeting

@@
 data.dir="/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/"
-limma.out.dir<-file.path(data.dir, 'CreatedData', 'limmaModels')
+limma.out.dir<-file.path(datadir, 'CreatedData', 'limmaModels')
 outputdir<-"/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Output"
@@
 
-AllModelsdf<-AllModels %>% bind_rows(.id='ModelType')%>%mutate('Model'=gsub('.*_', '', ModelType), 'Age'=gsub('_.*', '', ModelType))
-
 AllModelsdf=AllModelsdf%>%group_by(Model, Age, ModelType)%>%nest()%>%mutate(TopCpGs=map(data, function(x) x %>%filter(P.Value<1e-3)%>%pull(CpG)))%>%mutate(ModelType=case_when(ModelType=='child_model1'~'Age 9: Model 1', ModelType=='child_model2'~'Age 9: Model 2', ModelType=='child_model3'~'Age 9: Model 3', ModelType=='teen_model1'~'Age 15: Model 1', ModelType=='teen_model2'~'Age 15: Model 2', ModelType=='teen_model3'~'Age 15: Model 3'))%>%mutate(ModelType=factor(ModelType, levels=c('Age 9: Model 1', 'Age 9: Model 2', 'Age 9: Model 3', 'Age 15: Model 1', 'Age 15: Model 2', 'Age 15: Model 3')))
@@
 ```
+
 ```{r save beta comparison}

### 2021-10-26 Reports/quickResults.Rmd
results for oct 25 FF meeting

---
title: "Polymethylation scores results"
author: "F blostein"
date: "9/30/2021"
output:
  html_document:
    code_folding: show
    highlight: tango
    number_sections: no
    theme: sandstone
  pdf_document: default
  word_document: default
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message=FALSE)

library(ggplot2)
library(kableExtra)
library(ggpubr)
library(tidyr)
library(stringr)
library(dplyr)
library(compareGroups)
library(kableExtra)
library(gtsummary)
library(gtools)
library(haven)
library(sjlabelled)
library(matchmaker)
library(EpiDISH)
library("PerformanceAnalytics")
library(DescTools)
library(purrr)
library(broom)
library(broom.mixed)
library(ggplot2)
library(multcomp)
library(ggpubr)
library(cowplot)
library(geepack)
library(lme4)
library(pROC)
library(lmerTest)
library(stringr)
library(sva)
library(gtsummary)
library(gt)
library(here)
library(DiagrammeRsvg)
library(xml2)
library(ggrepel)
```

```{r temp read n data}
basedir=here::here()
outdir=file.path(gsub('Code', 'Output', basedir), 'Figures')
datadir=gsub('Code', 'Data', basedir)
load(paste0(datadir, '/CreatedData/allPhenoData.Rdata'))
load(paste0(datadir, '/CreatedData/allmethyl.Rdata'))
load(paste0(datadir, '/CreatedData/completeCaseSVA.Rdata'))
load(paste0(datadir, '/CreatedData/methylation_summarymodels.Rdata'))
```

```{r plotting variables}
theme_set(theme_bw())
```

```{r}
modeldata=modeldata %>% mutate(childteen=factor(childteen, levels=c('Age 9', 'Age 15')))
n_total=modeldata %>% dplyr::select(id)%>%unique()%>%nrow()
n_obs=modeldata%>%nrow()

all_methyl=c("globalmethylation", "pediatric", "grim", "Epithelial.cells_saliva", "Leukocytes_saliva", "SSnewborn_center", "SSnewbornCT_center", "SSolder_center", "cg05575921")
main_methyl=c("globalmethylation", "pediatric", "SSnewbornCT_center", "cg05575921")

all_methyl_labels=modeldata%>%dplyr::select(any_of(all_methyl))%>%get_label()
all_methyl_labels[str_detect(all_methyl_labels, 'Polymethylation', )]=paste0(rep('Score:', 3), c('Sustained smoking \n newborns', 'Sustained smoking \n newborns (w/ CT control)', 'Sustained smoking \n older children'))
names(all_methyl_labels)=all_methyl

all_methyl_levels=c("SSnewbornCT_center", "SSnewborn_center",  "SSolder_center",
                                       "pediatric" , "grim", "cg05575921", 
                                      "globalmethylation","Epithelial.cells_saliva", "Leukocytes_saliva")

main_methyl_labels=modeldata%>%dplyr::select(any_of(main_methyl))%>%get_label()

my_round=function(x, d){
  y=c()
  d_orig=d
  for(i in 1:length(x)){
    d=d_orig
    while(round(x[i], d)==0){
    d=d+1
    }
    y[i]=round(x[i], d)
  }
  return(y)
}
```

```{r figure counter}
main_fig=1
sup_fig=1
```

# Target journal and deadline

[Clinical epigenetics](https://www.biomedcentral.com/collections/epibio): Special issue on biomarkers; due date December 1st

 "Epigenetic biomarkers have significantly contributed to improved understanding of the origins and progression of disease.  Moreover, increasing evidence shows that epigenetic biomarkers have potential for personalized medicine. However, routine use of epigenetic biomarkers in in-vitro diagnostics (IVD) is lagging behind the discoveries of new biomarkers. In this thematic series of Clinical Epigenetics, we discuss current progress in development of epigenetic biomarkers for clinical use in common complex diseases of ageing and welcome both original research and reviews.
The sections of the collection include:

  - Types of epigenetic biomarkers and readiness for IVD use

    - DNA methylation as biomarker

    - Histone modification as biomarker

    - Upcoming field of circulating free nucleosomes as biomarker

    - Progress towards clinical use of biomarker

  - *Biomarkers of environmental exposures and causality of the biomarker*

  - Disease predisposition biomarkers

  - Biomarkers used for disease detection

  - Clinical disease prevention and management: predictive and prognostic

  - Monitoring of chronic disease

  - *Single versus multiple biomarkers and risk scores*

  - Major stages of clinical validation of the biomarker

  - Biomarker detection technologies suitable for IVD use and significance of bioinformatics

  - Ethics and regulatory aspects of IVD epigenetic biomarkers

# Figure 1 - Selection of samples from the Fragile Families and Child Wellbeing (FFCW) study into this analysis. N represents the number of individuals at each step in the selection procedure, M represents the number of samples, as individuals with repeated measures can have more than one sample.

```{r, eval=T}
#function
N_M_count=function(filter_statement, statement, df){
  paste0('m=', df %>% filter(eval(parse(text = filter_statement)))%>%nrow(), 
         ' samples & ', 
         'n=', df %>% filter(eval(parse(text = filter_statement))) %>% dplyr::select(idnum) %>%unique()%>%nrow(), 
         ' individuals with', statement)
}
#Make Labels
n_fullFF<-paste0("Fragile Families cohort \n", "n=", FF_labeled %>% nrow())
n_methylQC<-paste0("Cohort with 450K methylation data \n", "n=", methylCohort$idnum%>%unique()%>%length(), " individuals with m=", methylCohort$MethID%>%length(), " samples")

n_QCloss<-paste0("Methylation data QC: ", 
                 N_M_count("probe_fail_10==T & childteen!='M'", " with >10% of probes failing dropped, \n ", pdqc_all),
                 N_M_count("sex_new!=predicted_sex & probe_fail_10==F & childteen!='M'", 
                           "  with discordant sex samples dropped, \n ", pdqc_all),
                 N_M_count("predicted_sex_outlier==T & sex_new==predicted_sex & probe_fail_10==F & childteen!='M'", 
                           ' with abnormal sex chromosome intensity dropped & \n ', pdqc_all),
                 length(reps_remove), " duplicate samples dropped")

n_methylFF<-paste0("Cohort with quality controlled 450K data \n","n=", myFF %>% dplyr::select(id) %>% unique() %>% nrow(), " individuals with m=", myFF%>%nrow(), " samples")

n_missingmaternalprenataldata=paste0('Missing maternal data: \n', 
                                     N_M_count("is.na(smkPreg_binary)", " with missing prenatal maternal smoking \n ", myFF), 
                                     N_M_count("!is.na(smkPreg_binary) & is.na(m1g2_YesNoPreg)", " with missing prenatal alcohol \n ", myFF), 
                                     N_M_count("!is.na(smkPreg_binary) & !is.na(m1g2_YesNoPreg) & is.na(m1g3_YesNoPreg)", 
                                               " with missing prenatal other drug use \n ", myFF),
                                     N_M_count("!is.na(smkPreg_binary) & !is.na(m1g2_YesNoPreg) & !is.na(m1g3_YesNoPreg) & is.na(cm1inpov)", 
                                               " with missing maternal income to poverty ratio data \n ", myFF), 
                                     N_M_count("!is.na(smkPreg_binary) & !is.na(m1g2_YesNoPreg) & !is.na(m1g3_YesNoPreg) & !is.na(cm1inpov) &
                                               PostnatalMaternalSmokingAny=='Missing'", " with missing postnatal maternal smoking data \n ", myFF), 
                                     N_M_count("!is.na(smkPreg_binary) & !is.na(m1g2_YesNoPreg) & !is.na(m1g3_YesNoPreg) & !is.na(cm1inpov) &
                                               PostnatalMaternalSmokingAny!='Missing' & SmkAtVisitPastmonth=='Missing'", 
                                               " with missing past month maternal/primary care giver smoking packs/day", myFF))

maternaldata=myFF %>% filter(!is.na(smkPreg_binary) & !is.na(m1g2_YesNoPreg) & !is.na(m1g3_YesNoPreg) & !is.na(cm1inpov) & PostnatalMaternalSmokingAny!='Missing' & SmkAtVisitPastmonth!='Missing')  
n_withmaternaldata=paste0('m=', maternaldata %>% nrow(), ' samples & n=', maternaldata %>% dplyr::select(idnum)%>%unique()%>%nrow(), ' individuals with maternal covariate data')

n_missingchilddata=paste0('Missing child data: \n', 
                          N_M_count("is.na(ancestry) | ancestry=='Missing PC data'", 
                                    " with missing ancestry/genetic data  \n ", maternaldata), 
                          N_M_count("!is.na(ancestry) & ancestry!='Missing PC data' & is.na(cm1bsex)", 
                                    " with missing child sex data \n ", maternaldata), 
                          N_M_count("is.na(ancestry) | ancestry=='Missing PC data' & !is.na(cm1bsex) & is.na(ChildAgeComposite)", 
                                    " with missing child age at sample \n ", maternaldata))

childdata= maternaldata %>% filter(ancestry!='Missing PC data' & !is.na(cm1bsex) & !is.na(ChildAgeComposite))
n_withchilddata=paste0('m=', childdata %>% nrow(), ' samples & n=', childdata %>% dplyr::select(idnum)%>%unique()%>%nrow(), ' individuals with child covariate data')

n_Child<-paste0('n=', childdata %>%filter(childteen=='C')  %>% nrow(), " individuals with a 9 year visit")
n_ChildSmkexcluded<-paste0("Exclude children who smoke: \n",'n=', childdata%>%filter(childteen=='C' & k5f1l=='2 no')%>% nrow(), 
                           " individuals with a 9 year visit")
n_ChildSmkexclusions<-paste0('n=', childdata%>%filter(k5f1l!= "2 no" & childteen=='C')%>%nrow(), 
                             " individuals smoking \n or missing child smoking data \n at 9 year visit")
n_Teen<-paste0('n=', childdata %>%filter(childteen=='T')   %>% nrow(), " individuals with a 15 year visit")
n_TeenSmkexcluded<-paste0("Exclude children who smoke: \n",'n=', childdata %>%filter(childteen=='T' & k6d40=="2 No") %>% nrow(), 
                          " individuals with a 15 year visit")
n_TeenSmkexclusions<-paste0('n=', childdata %>% filter(k6d40!="2 No" & childteen=='T') %>% nrow,
                            " individuals smoking \n or missing focal child smoking data \n at 15 year visit") 
if(childdata %>% filter((childteen=='C' & k5f1l=='2 no') | (childteen=='T' & k6d40=='2 No'))%>%nrow()!=nrow(modeldata)){stop('filter N and completecase data set N do not match')}

```

```{r, fig.height=12}

graph1<-DiagrammeR::grViz("
   digraph graph2 {
   compound=true;
   graph [layout = dot]

   node [shape = rectangle, width=4]
   a [label = '@@1']
   b [label = '@@2']
   c [label = '@@3']
   d [label = '@@4']
   e [label = '@@5']
   f [label = '@@6']
   g [label = '@@7']
   h [label = '@@11']
   i [label = '@@12']

   #fake nodes for connecting to edges
   node [shape=point, width=0.01, height=0.01]
   a1
   b1
   c1
   d1
   f1 [group=g1]
   g1
   h1

  #exclusion nodes
  node [shape = rectangle, width=4]
  aa [label = '@@8']
  bb [label = '@@9']
  cc [label = '@@10']
  hh [label = '@@13']
  ii [label = '@@14']

  #draw connections
#  a->m
#  m->a1 [arrowhead=none]
  a->a1 [arrowhead=none]
  a1->b
  b->b1 [arrowhead=none]
  b1->c
  c->c1 [arrowhead=none]
  c1->d
  d->d1 [arrowhead=none]
  d1->e
  e->f; e->g
  f->g1 [arrowhead=none]; g->h1 [arrowhead=none]
  g1->h; h1->i

{rank=same; b1->aa [minlen=2]}   
{rank=same; c1->bb [minlen=2]} 
{rank=same; d1->cc [minlen=2]}
{rank=same; g1->hh [minlen=2]}
{rank=same; h1->ii [minlen=2]}

#invisible constraints to force ordering
#ff->i [style=invis]
   }

   [1]: n_fullFF
   [2]: n_methylQC
   [3]: n_methylFF
   [4]: n_withmaternaldata
   [5]: n_withchilddata
   [6]: n_Child
   [7]: n_Teen
   [8]: n_QCloss
   [9]: n_missingmaternalprenataldata
   [10]: n_missingchilddata
   [11]: n_ChildSmkexcluded
   [12]: n_TeenSmkexcluded
   [13]: n_ChildSmkexclusions
   [14]: n_TeenSmkexclusions
")

graph1

graph1 %>%
  export_svg() %>%
  read_xml() %>%
  write_xml(file.path(outdir, paste0("Figure", main_fig, ".svg")))

main_fig=main_fig+1            
```

# Supplemental Table 1 - Descriptive statistics of excluded vs included samples with methylation data available from the Fragile and Families and Child Wellbeing study

```{r}
myFF$MethylData2<-ifelse(myFF$MethID %in% modeldata$MethID, 'Final analysis group', 'Not in final analysis group')
set_label(myFF$MethylData2)<-'Exclusions based on missing data'
myFF=left_join(myFF, methyldata%>%mutate(idnum=as.character(idnum)))
myFF=myFF%>%mutate(across(c(PostnatalMaternalSmokingAny, SmkAtVisitPastmonth), ~ifelse(.x=='Missing', 
                                                       NA, .x)))%>%
  mutate(ancestry=ifelse(ancestry=='Missing PC data', NA, ancestry))%>%copy_labels(myFF)

exclusion_tbl=myFF %>% mutate(childteen=case_when(childteen=='C'~'Age 9', childteen=='T'~'Age 15'))%>%
  dplyr::select(smkPreg_binary, m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,
         SmkAtVisitPastmonth, ancestry, cm1bsex, childteen, globalmethylation, pediatric, grim,
         Epithelial.cells_saliva, Leukocytes_saliva, SSnewbornCT_center, SSolder_center,
         cg05575921, MethylData2)%>%
  tbl_summary(., by=MethylData2, statistic=list(all_continuous()~"{mean} ({sd})"))%>%
  add_n()%>%add_p(test=list(all_continuous()~"t.test"))

exclusion_tbl
```

# Supplemental Figure 2 - Correlation of methylation summary measures

```{r, fig.width=12, fig.height=12}
library(Hmisc)
library(corrplot)
res2<-modeldata%>%dplyr::select(any_of(all_methyl))%>%cor()
#labels
colnames(res2)=all_methyl_labels
rownames(res2)=all_methyl_labels
corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8)

pdf(file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')))
corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8)
dev.off()
sup_fig=sup_fig+1

```

# Supplemental Figure 3 - Correlation between visits among DNA methylation summary measures

```{r}
age_long_methyl=modeldata%>%
  dplyr::select(any_of(c('idnum', 'childteen', 'ancestry', 'smkPreg_binary', all_methyl)))%>%
  pivot_longer(cols=any_of(all_methyl),
               names_to='Methylation summary measure type',
               values_to='Methylation summary measure')%>%
  mutate(Methyl_factor=factor(`Methylation summary measure type`, 
                              levels=all_methyl_levels))

age_wide_methyl=age_long_methyl%>%
  pivot_wider(names_from=childteen, 
              values_from=`Methylation summary measure`)

visit_cor=age_wide_methyl%>%
  ggplot(aes(x=`Age 9`, y=`Age 15`))+geom_point()+stat_cor()+geom_smooth(method=lm)+theme_classic()+
  scale_x_continuous('Age 9 score')+scale_y_continuous('Age 15 score')+
  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels))

visit_cor

ggexport(visit_cor, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')))
sup_fig=sup_fig+1
```

# Supplemental Figure XX: Distributions of methylation summary measures at age 9 and age 15

```{r, fig.height=11}
dist_methyl=age_long_methyl%>%
  ggplot(aes(x=childteen, y=`Methylation summary measure`, color=childteen))+
  geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..", label.y.npc=0.87)+
  scale_color_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+
  scale_x_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+
  geom_point(alpha=0)+geom_line(alpha=0, aes(group=idnum))+geom_smooth(aes(group=1), method='lm', color='black')+
  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels))+
  theme(legend.position = 'bottom')

dist_methyl

ggexport(dist_methyl, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')))
sup_fig=sup_fig+1

```

# Figure 2: Polymethylation scores are correlated and stable between age 9 and age 15 in a diverse sample of `r n_total` US children

```{r change in scores time, fig.width=12}

#add line for mean ? 
outcome_labels=c('Global methylation', 'Pediatric clock', 'AHHR: cg05575921', 'Polymethylation score \n (newborn, mean centered)', 'Polymethylation score  \n (older children, mean centered')
names(outcome_labels)=c('globalmethylation', 'pediatric', 'cg05575921', 'SSnewbornCT_center','SSolder_center')

fig2=age_long_methyl%>%
    filter(Methyl_factor%in% main_methyl)%>%
    ggplot(aes(x=childteen, y=`Methylation summary measure`, color=smkPreg_binary))+
    geom_violin()+geom_boxplot(width=0.2, position = position_dodge(0.9))+
  stat_compare_means(label='..p.signif..', label.y.npc = 0.88)+
    facet_wrap(~Methyl_factor, scales='free',labeller = labeller(Methyl_factor=all_methyl_labels))+
    scale_x_discrete('Visit', labels=c('Age 9', 'Age 15'))+
    scale_color_grey('', labels=c('Not prenatal maternal smoke exposed', 'Prenatal smoke exposed'))+
    theme(legend.position = 'bottom')

fig2

ggexport(fig2, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')))
main_fig=main_fig+1

```

# Table 1: Bivariate associations between prenatal maternal smoking, DNA methylation summary measures and important covariates among a diverse sample of

```{r bivariate table}
smoking_tbl=modeldata %>% 
  dplyr::select(smkPreg_binary, m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,
         SmkAtVisitPastmonth, ancestry, cm1bsex, globalmethylation, pediatric, Epithelial.cells_saliva,
         Leukocytes_saliva, SSnewbornCT_center, SSolder_center, cg05575921, childteen)%>%
  tbl_strata(strata=childteen, .tbl_fun=~.x %>%
               tbl_summary(by=smkPreg_binary, statistic=
                             list(all_continuous()~"{mean} ({sd})"))%>%
               add_n()%>%add_p(test=list(all_continuous()~"t.test")))

smoking_tbl=smoking_tbl%>%as_gt()%>%
  tab_row_group(label='Maternal', rows = 1:11)%>%
  tab_row_group(label='Child', rows=12:25)

smoking_tbl
```

# Figure 3 - Polymethylation scores are associated with prenatal maternal smoke exposure consistently across age and ancestry group in a sample of `r n_total` US children

```{r, fig.width=12}
my_outcomes=c('Global methylation', 'Pediatric clock', 'AHHR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)')

outcome_labels=c('Global methylation', 'Pediatric clock', 'AHHR: cg05575921',  'Polymethylation score \n (newborn, mean centered)')

all_models=rbind(local_models, global_models%>%mutate(ancestry='Unstratified'))

data_text=all_models%>%
  mutate(n=map_dbl(data, nrow), n=paste0('n=', n), x=-0.25)%>%
  unnest(tidied)%>%
  filter(outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & term=='smkPreg_binaryYes'& predictors=='secondhand smoke exposure model')%>%mutate(outcome_f=outcome)

names(outcome_labels)=my_outcomes

forest_plot=all_models%>%
  mutate(n=map_dbl(data, nrow), n=paste0('n=', n))%>%
  unnest(tidied)%>%
  filter(term=='smkPreg_binaryYes' & predictors=='secondhand smoke exposure model')%>%
  filter(outcome%in% my_outcomes)%>%
    mutate(outcome_f=factor(outcome, levels=rev(my_outcomes)))%>%
  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=ancestry))+
  geom_point(position = position_dodge(width = 1))+
  geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
  facet_grid(.~outcome_f, scales='free_x', labeller = labeller(outcome_f=outcome_labels))+
  geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
  geom_text(data=data_text, aes(x=x,label=n), position=position_dodge(width = 1), hjust=0.05)+
  theme_classic()+
  theme(text = element_text(size=16))+theme(legend.position = 'bottom')+
  scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
  scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'), labels=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))

forest_plot

ggexport(forest_plot, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')), width=15)
main_fig=main_fig+1
```

# Internal Table and Figure: Question: Remove postnatal maternal smoking variable from models and/or use base model as main paper figures?

Somewhat concerned about extrapolated inference: very few exposed children (post-ancestry stratification) who were not also exposed to postnatal smoking.

```{r}
with(modeldata, table(ancestry, PostnatalMaternalSmokingAny, smkPreg_binary))%>%kbl()

forest_plot_compareModels=all_models%>%
    mutate(n=map_dbl(data, nrow), n=paste0('n=', n))%>%
    unnest(tidied)%>%
    filter(term=='smkPreg_binaryYes')%>%
    filter(outcome%in% my_outcomes)%>%
    mutate(outcome_f=factor(outcome, levels=rev(my_outcomes)))%>%
    ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=ancestry))+
    geom_point(position = position_dodge(width = 1))+
    geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
    facet_grid(predictors~outcome_f, scales='free_x', labeller = labeller(outcome_f=outcome_labels))+
    geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
    theme_classic()+
    theme(text = element_text(size=16))+theme(legend.position = 'bottom')+
    scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
    scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'), labels=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))

forest_plot_compareModels

ggexport(forest_plot_compareModels, filename=file.path(outdir, paste0('InternalFigure', '1', '.pdf')), width=15, height=15)
```

# Supplemental Table XX: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable models among `r n_total` US children by age

```{r}
#my_main_outcomes=c('AHHR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)', 'Sustained smoking polymethylation score (older children)', 'Pediatric clock', 'Global methylation')

test=rbind(global_models, longitudinal_global_models %>% mutate(childteen='Longitudinal'))%>%
  filter(predictors=='secondhand smoke exposure model')%>%
  filter(outcome %in% my_outcomes)%>%
  mutate(childteen=factor(case_when(childteen=='Age 9'~'Age 9 models', 
                             childteen=='Age 15'~'Age 15 models', 
                             childteen=='Longitudinal'~'Longitudinal models'),levels=c('Age 9 models', 'Age 15 models', 'Longitudinal models')), 
         tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
                                        show_single_row='smkPreg_binary', 
                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))

table_list=test %>%split(f=.$childteen)%>%map(~tbl_merge(.x$tbl, 
                                                         tab_spanner = as.character(.x$outcome)))

regression_table_age=tbl_stack(table_list, group_header=names(table_list))

regression_table_age

```

# Supplemental Table XX: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable models among `r n_total` US children by ancestry

```{r}
test=rbind(longitudinal_global_models %>% mutate(ancestry='Unstratified'),
           longitudinal_local_models%>%dplyr::select(-data))%>%
  filter(predictors=='secondhand smoke exposure model')%>%
  filter(outcome %in% my_outcomes)%>%
  mutate(tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
                                        show_single_row='smkPreg_binary', 
                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))

table_list=test %>%split(f=.$ancestry)%>%map(~tbl_merge(.x$tbl, 
                                                         tab_spanner = as.character(.x$outcome)))

regression_table_ancestry=tbl_stack(table_list, group_header=names(table_list))

regression_table_ancestry

```

# Supplemental Figure XX: Surrogate variable associations with known covariates

```{r, fig.width=12, fig.height=8}
#bring the heatmap function
covar=c('Sample_Plate', 'Slide', 'Array', 'cm1bsex', 'smkPreg_binary', 'childteen', 'ancestry')

bring.the.heatmap.sv <- function(pd,covar,folder){

  nsv <- length(grep('sv',names(pd)))

  #generate correlation matrix between SVs and variables
  correlations <- matrix(0,nrow=length(covar),ncol=nsv)
  colnames(correlations) <- paste('sv',1:nsv,sep='')
  rownames(correlations) <- covar

  #take svs
  svs <- pd[,colnames(correlations)]

  #correlations between svs and variables:
  # cor test if continuous
  # anova is multi category
  # t-test if two category
  for(col in covar){
    var <- pd[,col]
    #check to see if it is numeric and isn't really a two category thing
    if(class(var)=='numeric' & dim(table(var))>2){
      correlations[col,] <- apply(svs,2,FUN = function(x,y) cor.test(x,y)$p.value,y=var)
    }else{
      if(length(levels(factor(var)))>2){
        correlations[col,] <- apply(svs,2,FUN = function(x,y) anova(lm(x~y))$'Pr(>F)'[1],y=var)
      }else{
        correlations[col,] <- apply(svs,2,FUN = function(x,y) t.test(x~y)$p.value,y=var)
      }
    }
  }
  color.gradient <- colorRampPalette(c('firebrick','ivory3','royalblue'),bias=3)
  correlations=correlations%>%as.data.frame()%>%tibble::rownames_to_column('var1')%>%
    pivot_longer(-var1, names_to='var2')%>%
    mutate(var2=factor(var2, levels=unique(mixedsort(var2))),
           var1=factor(var1, levels=pd[, covar]%>%colnames(), labels=get_label(pd[, covar])))
  correlations%>%
    ggplot(aes(x=var1, y=var2, fill=value))+geom_tile()+
    scale_fill_gradientn('P value', colours = color.gradient(20), values = seq(0, 1, 0.5))+
    scale_x_discrete('', labels=str_wrap(get_label(pd[, covar]), 10))+
    scale_y_discrete('')
}

bring.the.heatmap.sv(modeldata, covar=c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'ancestry', 'cm1bsex', 'Leukocytes_saliva', 'Epithelial.cells_saliva', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))

pdf(file=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=11, height=13)
bring.the.heatmap.sv(modeldata, covar=c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'ancestry', 'cm1bsex', 'Leukocytes_saliva', 'Epithelial.cells_saliva', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))
dev.off()
sup_fig=sup_fig+1
```

# Supplemental Table XX: Surrogate variable models

```{r}
test=rbind(global_models, longitudinal_global_models %>% mutate(childteen='Longitudinal'))%>%
  filter(predictors=='surrogate variable models')%>%
  filter(outcome %in% my_outcomes)%>%
  mutate(childteen=factor(case_when(childteen=='Age 9'~'Age 9 models', 
                             childteen=='Age 15'~'Age 15 models', 
                             childteen=='Longitudinal'~'Longitudinal models'),levels=c('Age 9 models', 'Age 15 models', 'Longitudinal models')), 
         tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
                                        show_single_row='smkPreg_binary', 
                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))

table_list=test %>%split(f=.$childteen)%>%map(~tbl_merge(.x$tbl, 
                                                         tab_spanner = as.character(.x$outcome)))

regression_table_age_svs=tbl_stack(table_list, group_header=names(table_list))

test=rbind(longitudinal_global_models %>% mutate(ancestry='Unstratified'),
           longitudinal_local_models%>%dplyr::select(-data))%>%
  filter(predictors=='surrogate variable models')%>%
  filter(outcome %in% my_outcomes)%>%
  mutate(tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
                                        show_single_row='smkPreg_binary', 
                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))

table_list=test %>%split(f=.$ancestry)%>%map(~tbl_merge(.x$tbl, 
                                                         tab_spanner = as.character(.x$outcome)))

regression_table_ancestry_svs=tbl_stack(table_list, group_header=names(table_list))

regression_table_ancestry_svs
```

# Figure 4 - Polymethylation scores more accurately delineate prenatal smoke exposure status than non-targeted DNA methylation summary measures and *a priori* CpGs

```{r, fig.width=10}

age_labels=c('Age 9', 'Age 15'); names(age_labels)=c('Age 9', 'Age 15')
methyl_labels_roc=paste0('Base model+', all_methyl_labels)
names(methyl_labels_roc)=paste0('Base model+', names(all_methyl_labels))

fig4a= global_roc%>%
    dplyr::select(predictors, covariates, methylation, childteen, test, modeltype)%>%
    unnest(test) %>% 
    filter(str_detect(methylation, 'SSnewbornCT|cg0557|global') | modeltype=='Base model')%>%
    ggplot(aes(x=spec, y=sens, color=modeltype))+geom_line()+
    facet_wrap(childteen~., nrow=1, labeller=labeller(childteen=age_labels))+
    theme_classic()+
    scale_color_manual('', values=c('red', 'darkred', 'darkblue', 'darkgreen', 'blue', 'green'), 
                       function(x) str_wrap(x, width = 30))+
    theme(legend.position = 'right')+guides(color=guide_legend(nrow=4, byrow=TRUE))+
    geom_abline(intercept=1, slope=-1, color='black')+
    xlab('Specificity')+ylab('Sensitivity')+
    theme(legend.position='bottom')+ coord_fixed()

fig4b=global_roc%>%
  dplyr::select(covariates, methylation, childteen, test)%>% 
  filter(str_detect(methylation, 'SSnewbornCT') & covariates!='')%>%unnest(test)%>%
  rbind(long_global_roc%>% 
          mutate(childteen='Age 9 & Age 15', methylation=gsub('_Age_9.*', '', methylation))%>%
          dplyr::select(covariates, methylation, childteen, test)%>% 
          filter(str_detect(methylation, 'SSnewbornCT') 
                 & covariates!='')%>%
          unnest(test))%>%
  ggplot(aes(x=spec, y=sens, color=childteen, linetype=childteen))+geom_line(size=2)+
  facet_wrap(~methylation, labeller = labeller(methylation=c('SSnewbornCT_center'='Polymethylation score, \n newborn mean centered')))+
  theme_classic()+theme(legend.position = 'right')+guides(color=guide_legend(nrow=4,byrow=TRUE))+
  geom_abline(intercept=1, slope=-1, color='black')+
  scale_color_grey('', labels=c('Age 9 only', 'Age 9 & 15', 'Age 15 only'))+
  scale_linetype('', labels=c('Age 9 only', 'Age 9 & 15', 'Age 15 only'))+
  xlab('Specificity')+ylab('Sensitivity')+
  theme(legend.position='bottom')+ coord_fixed()

auc_fig=cowplot::plot_grid(fig4a, fig4b, labels=c('A', 'B'), nrow=1, rel_widths=c(0.65, 0.35), align='h', axis='t')

auc_fig

ggexport(auc_fig, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')), width=9)
main_fig=main_fig+1
```

# Supplemental Table with AUC and p-value comparison to sustained smoking, newborn cell type controls polymethylation score

```{r}
roc_table=global_roc %>% dplyr::select(childteen, methylation, modeltype, roc, auc)
roc_base=roc_table %>% filter(modeltype=='Base model+No methylation')%>%mutate(roc_basemodel=roc)%>%dplyr::select(childteen, roc_basemodel)
roc_ss=roc_table %>% filter(modeltype=='Sustained smoking w/ cell type correction (newborn)')%>%mutate(roc_pms=roc)%>%dplyr::select(childteen, roc_pms)

roc_table1=roc_table%>%filter(!str_detect(modeltype, 'Base model'))%>%left_join(roc_ss)%>%mutate(roctest=map2(roc, roc_pms, ~roc.test(.x, .y)))%>%mutate(auc=round(auc, 2))

roc_table1%>%mutate(`pval comparison to sustained smoking score (newborns)`=map_dbl(roctest, ~round(.x$p.value, 4)))%>%dplyr::select(childteen, modeltype, auc, `pval comparison to sustained smoking score (newborns)`)%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value (vs polymethylation score, newborns, cell type controlled)'))
```

# Supplemental Table with AUC and p-value comparison to base model with no methylation variables (i.e., using only child sex, income to poverty ratio, saliva/epithelial cell proportions and sample plate)

```{r}

roc_table2=roc_table%>%filter(str_detect(modeltype, 'Base model'))%>%left_join(roc_base)%>%mutate(roctest=map2(roc, roc_basemodel, ~roc.test(.x, .y)))%>%mutate(auc=round(auc, 2))

roc_table2%>%mutate(`pval comparison to base model`=map_dbl(roctest, ~round(.x$p.value, 4)))%>%dplyr::select(childteen, modeltype, auc, `pval comparison to base model`)%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value (vs base w/ no methyl)'))

```

### 2021-10-26 Scripts/01_clean_phenotype_data.R

@@ -6,7 +6,8 @@ library(tidyr)
 library(stringr)
 library(dplyr)
 library(purrr)
-library(tidydplyr::select)
+library(tidyverse)
+library(dplyr)
 library(sjlabelled)
 
 #directory based coding
@@ -120,6 +121,7 @@ FF_labeled=FF_labeled %>%
 #read in local and global pc data
 pc_files=list.files(path=paste0(datadir, "OGData/pcs"), pattern='*.csv', full.names=TRUE)
 names(pc_files)=gsub('.csv', '', list.files(path=paste0(datadir, "OGData/pcs"), pattern='*.csv'))
+names(pc_files)=paste0(str_to_title(names(pc_files)), ' ancestry')
 localpc=pc_files%>%map_dfr(read.csv, .id='ancestry')%>%rename_with(~str_c('local_', .), contains('PC'))
 allPC<-read.table(paste0(datadir, "OGData/pcs/global_pcs.txt"), col.names=c("X", "idnum", paste("global_PC", 1:20, sep='')))
 

### 2021-10-26 Scripts/02_make_methyl_summary.R

@@ -18,7 +18,7 @@ library(knitr)
 source(file.path(codedir, 'UsefulCode', 'makePolyEpiScores.R'))
 
 #######################################chose apriori cpgs, if any
-aprioriCG<-c("cg05575921", "cg04180046", "cg05549655", "cg14179389")
+aprioriCG<-c("cg05575921", "cg04180046", "cg05549655", "cg14179389", "cg22132788")
 
 ################################read in methylation data
 #as of october 4 switched to beta file with  423668 probes (see Jonah email search Check in on sex filtration/clocks Fragile Families)
@@ -31,7 +31,7 @@ estF_FF<-as.data.frame(celltypes$estF)%>%
   tibble::rownames_to_column(var='MethID')%>%
   rename_with(~paste0(., '_RPC'), .cols=c('Epi', 'Fib', 'IC'))
 #Middleton method
-estL_FF<-estimateLC(betaqc, 'saliva')%>%
+estL_FF<-estimateLC(betaqc, 'saliva', constrain=TRUE)%>%
   mutate(MethID=colnames(betaqc))%>%
   rename_with(~paste0(., '_saliva'), .cols=c('Leukocytes', 'Epithelial.cells'))
 
@@ -115,15 +115,15 @@ pms_labels=c('Polymethylation score: coefficients for any smoking from newborn c
               'Polymethylation score: coeffcients for sustained smoking from older children peripheral blood meta-analysis (no transform)')
 
 methyl_labels_myclocks=c('Methylation data ID', 'Epithelial cell proportion', 'Fibroblast cel proportion', 'Immune cell proportion',
-                         'Leukocytes proportion (saliva)','Epithelial cell proportion (saliva)', 
-                         'Global methylation', 'AHHR: cg05575921', 'MYO1G: cg04180046', 'CYP1A1: cg05549655', 'GFI1: cg14179389', 
+                         'Immune cell proportion (saliva)','Epithelial cell proportion (saliva)', 
+                         'Global methylation', 'AHHR: cg05575921', 'MYO1G: cg04180046', 'CYP1A1: cg05549655', 'GFI1: cg14179389', "MYO1G: cg22132788", 
                          pms_labels, gsub('no transform', 'z-score standardized', pms_labels), gsub('no transform', 'mean-centered', pms_labels), 
                          'ID', 'Visit',
                          'Horvath clock', 'SkinBlood clock', 'Pediatric clock', 'PoAm clock 38', 'PoAm clock 45')
 
[…]

### 2021-10-26 Scripts/04_summary_methyl_models.R

@@ -33,6 +33,7 @@ load(paste0(datadir, '/completeCasemethyl.Rdata'))
 modeldata=completecase%>%droplevels()%>%copy_labels(completecase)
 
 #set as factors 
+modeldata=modeldata %>% mutate(childteen=factor(childteen, levels=c('Age 9', 'Age 15')))
 modeldata$PostnatalMaternalSmokingAny<-factor(modeldata$PostnatalMaternalSmokingAny, 
                                               levels=c("No maternal smoking at age 1 and 5", 
                                                        "Maternal smoking at age 1 or age 5"))
@@ -55,6 +56,7 @@ save(modeldata, file=file.path(datadir, file='completeCaseSVA.Rdata'))
 
 #################################Set variables for modeling
 #outcomes & outcome_labels
+
 y_vector<-c("globalmethylation", "pediatric", "anynewborn_center", "SSnewbornCT_center", "SSolder_center", "cg05575921")
 outcome_labels=c('Global methylation', 'Pediatric clock',  
                  'Any smoking polymethylation score (newborns)', 
@@ -63,8 +65,9 @@ outcome_labels=c('Global methylation', 'Pediatric clock',
                  'AHHR: cg05575921')
 names(outcome_labels)=y_vector
 
+
 #model variables (excepting PCs, add in individually for global & ancestry specific models)
-base_model_vars<-"~smkPreg_binary+cm1bsex+cm1inpov+Leukocytes_saliva+Epithelial.cells_saliva+Sample_Plate"
+base_model_vars<-"~smkPreg_binary+cm1bsex+cm1inpov+Leukocytes_saliva+Sample_Plate+ChildAgeComposite"
 prenatal_model_vars<-paste0(base_model_vars, "+m1g2_YesNoPreg+m1g3_YesNoPreg")
 secondhand_model_vars<-paste0(prenatal_model_vars, '+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth')
 interaction_model_vars<-paste0(secondhand_model_vars, '+ChildAgeComposite:smkPreg_binary')
@@ -140,7 +143,7 @@ global_models=my_clean_argnames(global_models, global_predictors)
 #########################Local
 #arguments
 local_predictors=c(paste0(predictors, local_pcs), surrogate_model_vars)
-args=list('childteen'=c('C', 'T'), 'outcome'=y_vector, 'predictors'=local_predictors, 'ancestry'=ancestry_vector)%>%cross_df()
[…]

### 2021-10-26 Scripts/05_ewas.R

@@ -5,12 +5,13 @@ library(purrr)
 library(lme4) 
 library(lmerTest)
 library(matrixStats)
+library(limma)
 library('variancePartition')
 library('edgeR')
 library('BiocParallel')
 
 datadir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data"
-outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output"
+outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/CreatedData/EWAS"
 
 #variance trim of beta matrix
 betaqc<-readRDS(file=file.path(datadir, 'OGData', "betaqc.rds"))
@@ -27,6 +28,59 @@ modeldata=modeldata[match(colnames(betaqc), modeldata$MethID),]
 #rownames of modeldata 
 rownames(modeldata)=modeldata$MethID
 
+#function to add annotation
+add_annotations<-function(data, type='lm'){
+  #add annotations
+  if(type=='lme'){rownames(data)=data$CPG}
+  Locations <- Locations[rownames(data), ]
+  Islands.UCSC <- Islands.UCSC[rownames(data),]
+  Other <- Other[rownames(data),]
+  # add columns of interest from annotation to annotated version
+  data.annotated <- data
+  data.annotated$chr <- Locations$chr
+  data.annotated$pos <- Locations$pos
+  data.annotated$island <- Islands.UCSC$Relation_to_Island
+  data.annotated$gene <- Other$UCSC_RefGene_Name
+  data.annotated$regulatory_feature <- Other$Regulatory_Feature_Group
+  return(data.annotated)
+}
+library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
+data(Locations)
+data(Islands.UCSC)
+data(Other)
+
+##################cross-sectional EWAS###############
+#########child & teen#################
+#filter betaqc to child samples only 
+child_nosmoke=modeldata%>%filter(childteen=='Age 9')
+teen_nosmoke=modeldata%>%filter(childteen=='Age 15')
+#models
+mod1C=model.matrix(~factor(child_nosmoke$smkPreg_binary)+child_nosmoke$global_PC1+child_nosmoke$global_PC2+
+                    factor(child_nosmoke$cm1bsex)+child_nosmoke$cm1inpov+child_nosmoke$ChildAgeComposite+
[…]

### 2021-10-26 Papers/Polymethylation_scores/06_discussion.Rmd

--- 2021-10-01 Papers/Polymethylation_scores/06_discussion.Rmd
@@
+---
+output:
+  word_document: default
+  html_document: default
+bibliography: Methylation_SmokingBib.bib
+---
+
 # Discussion
 
-Restate findings
+In the longitudinal Fragile Families and Child Wellbeing birth cohort, we found an association between prenatal maternal smoking and DNA methylation in sequential saliva samples from ages 9 and 15.
+Polymethylation scores were associated with prenatal maternal smoking across ages and ancestry while global methylation and epigenetic clocks were not, demonstrating a specific and persistent signal of prenatal smoke exposure in DNA methylation.
+Polymethylation scores had reasonable accuracy for classifying prenatal maternal smoking.
+This accuracy was more consistent across time than when using some individual *a priori* CpG sites.
 
-How are results similar or different to other results
+Our findings fit within the previous literature on associations between prenatal maternal smoking and DNA methylation.
+We replicate top hits from previous epigenome wide association studies, including a large metaanalysis of DNA methylation in blood samples [@Joubert:2016ue].
+Although there are no previous studies of prenatal maternal smoking and saliva methylation, our findings fit with findings on the consistency of DNA methylation signatures of own-smoking between saliva and blood samples [@Dawes2019; @Philibert2020].
 
-Tissue
+We contribute to the literature on DNA methylation and prenatal maternal smoking by evaluating the consistency of this association in saliva and across ages and ancestries.
+Polymethylation scores built using coefficients from metaanalysis of blood DNA methylation still associated with prenatal maternal smoking in our independent population of saliva samples.
+Similarly, even though the source population for construction of the polymethylation scores was primarily of European ancestry, an association with prenatal maternal smoking was apparent in our primarily African ancestry population.
+Effect estimates were consistent across ancestries at age 9, although Hispanic and European ancestry age 15 samples had a lower effect estimate.
+This could reflect higher unreported smoking initiation in European and Hispanic ancestry US teens than African ancestry teens.
+Genetic ancestry correlates with race, and White and Latino teens have much higher rates of teen smoking and earlier ages at initiation than Black children [@El-Toukhy2016].
+Own-smoking could influence DNA methylation at ages 9 and 15, creating outcome misclassification.
+This could also explain the observed attenuation of hypomethylation at AHRR:CpG05575921 by age 15 in our sample, as methylation at this probe is known to vary by own-smoking exposure as well[@Dawes2019; @Philibert2020].
 
-Age
+Our findings also fit with previous work on DNA methylation as a biomarker for prenatal maternal smoke exposure.
+Accuracy of the polymethylation scores in our sample (Age 9: 0.67 and Age 15: 0.66) was worse than the performance of biomarkers using cord blood (AUCs ranging from 0.82 to 0.97) [@Reese:2017ts; @Ladd-Acosta:2016aa].
+Differences in methylation patterns across tissues and over time could be responsible for this diminished accuracy.
+However, our accuracy was comparable to accuracy when using peripheral blood from older adults (AUC 0.72) [@Richmond:2018tu].
+The diminished accuracy from cord blood may be an acceptable trade-off considering saliva is a more readily available tissue for epidemiological and clinical applications.
 
-Ancestry
+Our analysis answers pertinent questions for the development of a clinically applicable methylation biomarker of prenatal maternal smoking.
+We show that DNA methylation scores calculated from children's saliva can accurately classify prenatal maternal smoke exposure in non-European ancestry children.
+Other bomarkers and risk scores are known to differ across ancestry groups, potentially contributing to and exacerbating health disparities.
+As new biomarkers are developed, understanding their consistency across ancestry groups is critical.
+We also demonstrated that accuracy is not substantially affected when using coefficients from newborn cordblood (no cell type control, age 9: 0.62, age 15: 0.61) vs older children peripheral blood (no cell type control age 9: 0.63, age 15: 0.60).
+However, using coefficients from models that incorporated cell-type control at the time of calculation did result in better accuracy, suggesting cell-type control should be instituted both when creating weights and when applying scores to new population.
+Including samples from both ages did not improve accuracy, thus repeated samplings may be unnecessary for clinical applications.
+Additionally, accuracy was more consistent when using polymethylation scores than using the single top-hit from Joubert et al, AHHR:CpG05575921 [@Joubert:2016ue].
+Salivary AHHR:CpG05575921 methylation was less accurate in our sample at predicting prenatal maternal smoke exposure than when it has been used to classify own smoking in blood or saliva [@Philibert2020].
+While using a single CpG site as a biomarker may be clinically pragamatic, single sites may be less robust to time-since-exposure and new environmental exposures.
+Alternatively, incorporating information from across several CpGs in a polymethylation score may be more robust to these .
+
+Our analysis is not without its limitations.
+We used maternal self report of prenatal maternal smoker as serum cotinine levels were not available.
+Due to social desirability bias, this could result in exposure misclassification.
+We would expect this to bias our results towards the null.
+In any analysis of a prenatal exposure and postnatal outcome, there is the possibility of selection bias into the cohort due to live birth bias.
+Selection bias is also possible due to loss-to-follow-up between birth and age 9.
+Additionally, while we controlled for postnatal secondhand smoke exposure and excluded children who reported any own-smoking, we cannot exclude the possibility of residual confounding in this observational cohort.
+However, the longitudinal design with repeated measures, exposure-outcome temporality and large sample size are strengths of our analysis.
+
+In total, our analysis contributes to the clinical implementation of a DNA methylation-based biomarker of prenatal maternal smoking.
+We showed that DNA methylation in saliva, a clinically practical tissue, had strong associations with and reasonable accuracy for prenatal maternal smoke exposure.
+Further, we demonstrated that polymethylation scores could be applied as a biomarker of prenatal maternal smoke exposure across ancestry-groups, an important consideration for health equity.
+The development and application of biomarker for prenatal maternal smoke exposure has important implications for epidemiological research and clinical practice.
+Given the difficulty of measuring prenatal maternal smoke exposure, such a biomarker could allow for confounder control in research areas where such control is currently impossible.
+Given the prevalence and negative health consequences of prenatal maternal smoke exposure, such a biomarker could be used to identify exposed individuals who might especially benefit from certain interventions.

### 2021-10-26 Papers/Polymethylation_scores/08_methods.Rmd
work on intro and background outline

@@
 ---
-output: html_document
-bibliography: "../Methylation_SmokingBib.bib"  
+output:
+  word_document: default
+  html_document: default
+bibliography: Methylation_SmokingBib.bib
 ---
 
-# Methods
+## Methods
 
-## Cohort
+### Cohort
 
-The Fragile Families and Child Wellbeing Study (FFCW) is a nationally-representative cohort of nearly 5,000 children born in 20 U.S.
+The Fragile Families and Child Wellbeing Study (FFCW) is a birth cohort of nearly 5,000 children born in 20 U.S.
 cities between 1998 and 2000 [@Reichman:2001wv].
-FFCW sampled births using a three-stage stratified random sample design, described in detail elsewhere [@Reichman:2001wv].
+FFCW sampled births using a three-stage stratified random sample design, described in detail elsewhere, which oversampled unmarried mothers by a ratio of 3:1 [@Reichman:2001wv].
 FFCW excluded the following births from the study: those with parents who planned to place the child for adoption, those where the father was deceased, those who did not speak English or Spanish well enough to be interviewed, those where the mothers or babies were too ill to complete the interview and those where the baby died before the interview could take place.
-At some of the hospitals, parents who were under 18 were not eligible to be interviewed due to hospital prohibition.
-Children were followed from birth until age fifteen.
-At the time of birth, medical record information was extracted from the hospital at which the mother gave birth.
-Interviews were conducted with the mother at birth in person and with father in person or over the telephone if he was not present.
-Interviews were also conducted with the mother and father when the child was one, three, five, and nine via telephone.
-Additionally, at ages three, five and nine, a separate primary caregiver survey was administered to the individual with whom the child was living either over the telephone or in the home.
-The child was interviewed at age nine in the home and at age fifteen by telephone or in the home.
+Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; additional follow up is ongoing.
+Assessments include medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher and child.
 At ages nine and fifteen a saliva sample was taken from the child for DNA and DNA methylation measurements [@Reichman:2001wv].
 
-## Prenatal maternal smoking measurement
+### Covariates and exposure measurement
 
-## Covariates
+Demographic and prenatal maternal substance use variables were derived from maternal self-report questionnaire data at baseline (child's birth).
+Demographic variables included maternal income to poverty ratio and child sex.
+Prenatal maternal alcohol and other drug use were dichotomized to any vs no prenatal maternal alcohol and drug use respectively.
+Prenatal maternal smoking was ascertained from the categorical interview question
 
-## DNA methylation measurement
+> "During your pregnancy, how many cigarettes did you smoke? Did you smoke...:
+>
+> 2 or more packs a day
+>
+> 1 or more but less than 2 packs per day
+>
+> 	Less than 1 pack a day
+>
+> 	None
 
-## Sample filtering
+Because few participants reported smoking a pack or more a day, this question was also dichotomized to any vs no prenatal maternal smoking.
 
-## Probe filtering
+Postnatal second-hand smoke exposure was encoded in two variables.
+To encapsulate general early childhood smoke exposure, we used a binary any postnatal exposure to maternal or primary caregiver smoking at ages 1 or 5 , constructed from questions about maternal or primary caregiver smoking in the age 1 and 5 interviews.
+To encapsulate recent postnatal smoke exposure, we used a categorical packs per day maternal or primary care giver smoking in the month prior to age 9 and 15 maternal interview.
 
-## Methylation summary measures
+Principal components of child genetic ancestry were calculated from genetic data (CHIP?) from child saliva samples.
+In non-stratified models, the first two components of genetic ancestry from principal component analysis run in the entire analytic subset were used, in ancestry-stratified models the first two principal components from principal component analysis run within each strata were used.
 
-We created several methylation summary measures.
+Confounder variables included principal components of child genetic ancestry, prenatal maternal alcohol and other drug use.
+
+Precision variables (variables strongly associated with DNA methylation controlled for to decrease standard errors of coefficient of interest) included child sex, cell proportions estimated from DNA methylation data
+
+### DNA methylation measurement
+
+Salivary samples from the children in FFCW were taken at ages nine and fifteen using the Oragene•DNA sample collection kit (DNA Genotek Inc., Ontario).
+Saliva DNA was extracted manually using DNA Genotek's purification protocol using prepIT L2P.
+DNA was bisulfite treated and cleaned using the Zymo Research EZ DNA Methylation kit.
+Samples were randomized and plated across slides by demographic characteristics.
+Saliva DNA methylation was measured using the Illumina HumanMethylation 450k BeadArray [@Sandoval2011] and imaged using the Illumina iScan system.
+All samples were run in a single batch to minimize technical variability.
+
+DNA methylation image data (IDAT) were processed in R statistical software (3.5) using the minfi package [@Aryee2014].
+The IDAT pairs (n=1811 ????) were read into R and the minfi *preprocessNoob* function was used to normalize dye bias and apply background correction.
+Further quality control was applied using the ewastools packages [@Heiss2018].
+We dropped samples with \>10% of sites have detection p-value \>0.01 (n=43), sex discordance between DNA methylation predicted sex and recorded sex (n=20), and abnormal sex chromosome intensity (n=3).
+CpG sites were removed if they had detection p-value \>0.01 in 5% of samples (n=26,830) or were identified as cross-reactive (n=27,782) [@Chen2013].
+Relative proportions of immune and epithelial cell types were estimated from DNA methylation measures using a childhood saliva reference panel [@Middleton2021].
+
 Global methylation was calculated for each sample as the mean methylation value of each sample across the cleaned probe set.
-Methylation clocks were calculated for each sample according to the methods of --INSERT HERE--.
-To create polymethylation scores, 6,000 (CHECK) CpG sites with regression coefficients reported in @Joubert:2016ue were multiplied by methylation beta values from FFCW and summed.
-We calculated polymethylation scores using regression coefficients from 4 different regressions reported in @Joubert:2016ue (LIST) and 3 different methods of transforming the methylation beta values (no transformation, mean center, and z-score standardization).
+Pediatric epigenetic age was calculated for each sample.
+To create polymethylation scores for prenatal maternal smoke exposure, 6,074 CpG sites with regression coefficients reported in Joubert et al.
+2016 [@Joubert:2016ue] were multiplied by mean-centered methylation beta values from FFCW and summed.
+We calculated polymethylation scores using regression coefficients from 4 different regressions reported in Joubert et al.
+2016 [@Joubert:2016ue] (sustained smoking exposure and DNA methylation in newborn cordblood with cell-type control, sustained smoking exposure and DNA methylation in newborn cordblood without cell-type control, sustained smoking exposure and DNA methylation in older children peripheral blood, any smoking exposure and DNA methylation in newborn cordblood ).
 
-## Statistical analysis
+Epigenome-wide association analyses were run separately at age 9 and age 15 using the linear models and the R package limma [@Ritchie2015].
+We also ran epigenome wide association analyses using DNA methylation measures from both time points using linear mixed effect models with a random intercept for child ID.
+Gap probes were identified using the function *gaphunter* from the R package minfi [@Aryee2014].

### 2021-10-26 Papers/Polymethylation_scores/06_discussion.Rmd
edits to results and scripts for 8/25 FF meeting

@@
 Including samples from both ages did not improve accuracy, thus repeated samplings may be unnecessary for clinical applications.
-Additionally, accuracy was more consistent when using polymethylation scores than using the single top-hit from Joubert et al, AHHR:CpG05575921 [@Joubert:2016ue].
-Salivary AHHR:CpG05575921 methylation was less accurate in our sample at predicting prenatal maternal smoke exposure than when it has been used to classify own smoking in blood or saliva [@Philibert2020].
+Additionally, accuracy was more consistent when using polymethylation scores than using the single top-hit from Joubert et al, AHRR:CpG05575921 [@Joubert:2016ue].
+Salivary AHRR:CpG05575921 methylation was less accurate in our sample at predicting prenatal maternal smoke exposure than when it has been used to classify own smoking in blood or saliva [@Philibert2020].
 While using a single CpG site as a biomarker may be clinically pragamatic, single sites may be less robust to time-since-exposure and new environmental exposures.

### 2021-10-26 Papers/Polymethylation_scores/index_paper.Rmd
edits to results and scripts for 8/25 FF meeting

@@
 ---
+
+```{r, echo=F, include=F, warning=F, message=F}
+library(stringr)
+results_run=str_detect(here::here(), 'nfs/turbo/bakulski1')
+```
 
@@
 
-```{r child = here::here("Papers", "Polymethylation_scores", "05_results.Rmd")}
+```{r child = here::here("Papers", "Polymethylation_scores", "05_results.Rmd"), eval=results_run}
 ```

### 2021-10-26 Reports/Results.Rmd
edits to results and scripts for 8/25 FF meeting

@@
   pdf_document: default
+  word_document: default
+always_allow_html: true
 ---
@@
 library(sva)
+library(gtsummary)
+library(gt)
+library(here)
+library(DiagrammeRsvg)
+library(xml2)
+library(ggrepel)
 ```
@@
 ```{r temp read n data}
-codedir=''
-datadir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data"
+basedir=here::here()
+outdir=file.path(gsub('Code', 'Output', basedir), 'Figures')
+datadir=gsub('Code', 'Data', basedir)
 load(paste0(datadir, '/CreatedData/allPhenoData.Rdata'))
@@
 ```{r}
+modeldata=modeldata %>% mutate(childteen=factor(childteen, levels=c('Age 9', 'Age 15')))
+set_label(modeldata$childteen)='Visit'
+
 n_total=modeldata %>% dplyr::select(id)%>%unique()%>%nrow()
-n_obs=modeldata%>%nrow()
+m_obs=modeldata%>%nrow()
+n_twovisit=modeldata %>% group_by(id)%>%summarise(n=n())%>%filter(n==2)%>%nrow()
 
@@
 
-# Current issues
+```{r figure counter}
+main_fig=1
+sup_fig=1
+main_tab=1
+sup_tab=1
+```
 
-# Target journal and deadline
-
-[Clinical epigenetics](https://www.biomedcentral.com/collections/epibio): Special issue on biomarkers; due date December 1st
- >Epigenetic biomarkers have significantly contributed to improved understanding of the origins and progression of disease.
-Moreover, increasing evidence shows that epigenetic biomarkers have potential for personalized medicine.
-However, routine use of epigenetic biomarkers in in-vitro diagnostics (IVD) is lagging behind the discoveries of new biomarkers.
-In this thematic series of Clinical Epigenetics, we discuss current progress in development of epigenetic biomarkers for clinical use in common complex diseases of ageing and welcome both original research and reviews.
-The sections of the collection include:
-  - Types of epigenetic biomarkers and readiness for IVD use
-    - DNA methylation as biomarker
-    - Histone modification as biomarker
-    - Upcoming field of circulating free nucleosomes as biomarker
-    - Progress towards clinical use of biomarker
-  - *Biomarkers of environmental exposures and causality of the biomarker*
-  - Disease predisposition biomarkers
-  - Biomarkers used for disease detection
-  - Clinical disease prevention and management: predictive and prognostic
-  - Monitoring of chronic disease
-  - *Single versus multiple biomarkers and risk scores*
-  - Major stages of clinical validation of the biomarker
-  - Biomarker detection technologies suitable for IVD use and significance of bioinformatics
-  - Ethics and regulatory aspects of IVD epigenetic biomarkers
-
-# Figure 1 - Selection of samples from the Fragile Families and Child Wellbeing (FFCW) study into this analysis.
+# Figure `r main_fig` - Selection of samples from the Fragile Families and Child Wellbeing (FFCW) study into this analysis.
 N represents the number of individuals at each step in the selection procedure, M represents the number of samples, as individuals with repeated measures can have more than one sample.
@@
 
-```{r}
+```{r, fig.height=12}
+
 graph1<-DiagrammeR::grViz("
@@
 graph1
+
+graph1 %>%
+  export_svg() %>%
+  read_xml() %>%
+  write_xml(file.path(outdir, paste0("Figure", main_fig, ".svg")))
+
+main_fig=main_fig+1            
 ```
 
-# Supplemental Table 1 - Descriptive statistics of excluded vs included samples with methylation data available from the Fragile and Families and Child Wellbeing study
+# Supplemental Table `r sup_tab` - Descriptive statistics of excluded vs included samples with methylation data available from the Fragile and Families and Child Wellbeing study
 
@@
 myFF=left_join(myFF, methyldata%>%mutate(idnum=as.character(idnum)))
-methyl2<-createTable(compareGroups(MethylData2~smkPreg_binary+m1g2_YesNoPreg+m1g3_YesNoPreg+cm1inpov+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+ancestry+cm1bsex+globalmethylation+pediatric+Epithelial.cells_saliva+Leukocytes_saliva+SSnewbornCT_center+SSolder_center+cg05575921, data=myFF, include.miss = T), show.all=T)
-export2md(strataTable(methyl2, 'childteen'), caption="Exclusion table of subset with quality controlled methylation data vs final analytic subset for main predictor, outcomes and key demographic variables")
-```
-# Supplemental Figure - Comparison of choice of A) regression coefficient and B) transform of methylation beta value on distributions of polymethylation scores
+myFF=myFF%>%mutate(across(c(PostnatalMaternalSmokingAny, SmkAtVisitPastmonth), ~ifelse(.x=='Missing', 
+                                                       NA, .x)))%>%
+  mutate(ancestry=ifelse(ancestry=='Missing PC data', NA, ancestry))%>%copy_labels(myFF)
 
-```{r}
-scores=gsub('_center', '', colnames(modeldata)[str_detect(colnames(modeldata), 'center')])
-age_long_score=modeldata%>%
-  dplyr::select(any_of(c('idnum', 'childteen', 'ancestry', 'smkPreg_binary')),
-                contains(scores))%>%
-  pivot_longer(cols=contains(scores),
-               names_to='Scores',
-               values_to='Polymethylation score')
+exclusion_tbl=myFF %>% mutate(childteen=case_when(childteen=='C'~'Age 9', childteen=='T'~'Age 15'))%>%
+  dplyr::select(smkPreg_binary, m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,
+         SmkAtVisitPastmonth, ancestry, cm1bsex, childteen, globalmethylation, pediatric, grim,
+         Epithelial.cells_saliva, Leukocytes_saliva, SSnewbornCT_center, SSolder_center,
+         cg05575921, MethylData2)%>%
+  tbl_summary(., by=MethylData2, statistic=list(all_continuous()~"{mean} ({sd})"))%>%
+  modify_header(stat_by = "**{level}**, M samples = {n}")%>%
+  add_n(col_label='M samples')#%>%
+  #add_p(test=list(all_continuous()~"t.test"))
 
-plot_grid(
-  age_long_score%>%filter(str_detect(Scores, 'no')) %>% 
-    ggplot(aes(x=`Polymethylation score`))+geom_histogram()+
-    facet_wrap(~Scores, labeller=labeller(Scores=get_label(modeldata%>%dplyr::select(contains('notransform'))))), 
-  age_long_score%>%filter(str_detect(Scores, 'CT')) %>% 
-    ggplot(aes(x=`Polymethylation score`))+geom_histogram()+
-    facet_wrap(~Scores, scales='free', ncol=3,
-               labeller=labeller(Scores=get_label(modeldata%>%dplyr::select(contains('CT'))))),
-  labels=c('A', 'B'), nrow=2
-)
+exclusion_tbl
 
+sup_tab=sup_tab+1
 ```
 
-# Supplemental Figure 2 - Correlation of methylation summary measures
+# Supplemental Figure `r sup_fig` - Pearson correlation of methylation summary measures from an anlysis of M=`r m_obs` samples and N=`r n_total`children in the Fragile Families and Child Wellbeing cohort
 
-```{r}
+```{r, fig.width=12, fig.height=12}
 library(Hmisc)
@@
 corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8)
+
+pdf(file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')))
+corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8)
+dev.off()
+
+sup_fig=sup_fig+1
 ```
 
-#Supplemental Figure 3 - Correlation between methylation summary measures between visits
+# Supplemental Figure `r sup_fig` - Correlation between age 9 and age 15 visits for selected DNA methylation summary measures among `r n_twovisit` children with age 9 and age 15 visit data in the Fragile Families and Child Wellbeing study
 
@@
 
-age_wide_methyl%>%
-  ggplot(aes(x=C, y=T))+geom_point()+stat_cor()+geom_smooth(method=lm)+theme_classic()+
+#checks
+if(age_wide_methyl%>%filter(Methyl_factor=='globalmethylation')%>%filter(!is.na(`Age 9`) & !is.na(`Age 15`))%>%nrow()!=n_twovisit){stop('check sample size two visit vs age_wide methyl')}
+
+visit_cor=age_wide_methyl%>%
+  ggplot(aes(x=`Age 9`, y=`Age 15`))+geom_point()+stat_cor()+geom_smooth(method=lm)+theme_classic()+
   scale_x_continuous('Age 9 score')+scale_y_continuous('Age 15 score')+
   facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels))
+
+visit_cor
+
+ggexport(visit_cor, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')))
+sup_fig=sup_fig+1
 ```
 
-# Supplemental Figure XX: Change in methylation summary measures between age 9 and age 15
+# Supplemental Figure `r sup_fig`: Distributions of selected methylation summary measures at age 9 and age 15 among `r n_total` children in the Fragile Families and Child Wellbeing study
 
-```{r}
-age_long_methyl%>%
+```{r, fig.height=11}
+dist_methyl=age_long_methyl%>%
   ggplot(aes(x=childteen, y=`Methylation summary measure`, color=childteen))+
-  geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+
+  geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..", label.y.npc=0.87)+
   scale_color_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+
   scale_x_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+
+  geom_point(alpha=0)+geom_line(alpha=0, aes(group=idnum))+geom_smooth(aes(group=1), method='lm', color='black')+
   facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels))+
@@
 
-age_long_methyl%>%
-  ggplot(aes(x=childteen, y=`Methylation summary measure`, group=idnum))+
-  geom_point()+geom_line(alpha=0.1, color='grey')+geom_smooth(aes(group=1), method='lm', fill='blue')+
-  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels))+
-  theme_classic()+scale_x_discrete('', labels=c('Age 9', 'Age 15'))
+dist_methyl
+
+ggexport(dist_methyl, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')))
+sup_fig=sup_fig+1
+
 ```
 
-# Figure 2: Polymethylation scores are correlated and stable between age 9 and age 15 in a diverse sample of `r n_total` US children
+# Figure `r main_fig`: Differences in selected DNA methylation summary measures by prenatal maternal smoke exposure among `r n_total` US children in the Fragile Families and Child Wellbeing study
 
@@
 #add line for mean ? 
-outcome_labels=c('Global methylation', 'Pediatric clock', 'AHHR: cg05575921', 'Polymethylation score \n (newborn, mean centered)', 'Polymethylation score  \n (older children, mean centered')
+outcome_labels=c('Global methylation', 'Pediatric clock', 'AHRR: cg05575921', 'Polymethylation score \n (newborn, mean centered)', 'Polymethylation score  \n (older children, mean centered')
 names(outcome_labels)=c('globalmethylation', 'pediatric', 'cg05575921', 'SSnewbornCT_center','SSolder_center')
 
-fig2a=age_wide_methyl%>%filter(Methyl_factor%in% main_methyl)%>%
-  ggplot(aes(x=C, y=T))+geom_point()+stat_cor()+geom_smooth(method=lm)+theme_classic()+
-  scale_x_continuous('Age 9 score')+scale_y_continuous('Age 15 score')+
-  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels), nrow=1)
+fig2=age_long_methyl%>%
+    filter(Methyl_factor%in% main_methyl)%>%
+    ggplot(aes(x=childteen, y=`Methylation summary measure`, color=smkPreg_binary))+
+    geom_violin()+geom_boxplot(width=0.2, position = position_dodge(0.9))+
+  stat_compare_means(label='..p.signif..', label.y.npc = 0.88)+
+    facet_wrap(~Methyl_factor, scales='free',labeller = labeller(Methyl_factor=all_methyl_labels))+
+    scale_x_discrete('Visit', labels=c('Age 9', 'Age 15'))+
+    scale_color_grey('', labels=c('Not prenatal maternal smoke exposed', 'Prenatal smoke exposed'))+
+    theme(legend.position = 'bottom')
 
-mean_summary=age_wide_methyl%>%filter(Methyl_factor%in% main_methyl)%>%mutate(diff=T-C)%>%group_by(Methyl_factor)%>%summarise(mean_diff=mean(diff, na.rm=T))
+fig2
 
-fig2b=age_wide_methyl%>%filter(Methyl_factor%in% main_methyl)%>%mutate(diff=T-C)%>%
-  ggplot(aes(x=diff))+geom_histogram()+
-  facet_wrap(~Methyl_factor, scales='free_x', labeller = labeller(Methyl_factor=all_methyl_labels), nrow=1)+
-  theme_classic()+xlab('Difference in measure (Age 15 - Age 9)')+
-  geom_vline(aes(xintercept=0), color='red', linetype=2)+geom_rug()+
-  geom_vline(data=mean_summary, aes(xintercept=mean_diff), color='blue')
-
-plot_grid(fig2a, fig2b, nrow=2, align = 'v', axis = 'lr')
+ggexport(fig2, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')))
+main_fig=main_fig+1
 
@@
 
-# Table 1: Bivariate associations between prenatal maternal smoking, DNA methylation summary measures and important covariates among a diverse sample of
+# Table `r main_tab`: Bivariate associations between prenatal maternal smoking, selected DNA methylation summary measures and important covariates among a diverse sample of `r n_total` children in the Fragile Families and Child Wellbeing study
 
 ```{r bivariate table}
-export2md(strataTable(createTable(compareGroups(smkPreg_binary~m1g2_YesNoPreg+m1g3_YesNoPreg+cm1inpov+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+ancestry+cm1bsex+globalmethylation+pediatric+Epithelial.cells_saliva+Leukocytes_saliva+SSnewbornCT_center+SSolder_center+cg05575921, data=modeldata%>%mutate(childteen=factor(case_when(childteen=='C'~'Age 9', childteen=='T'~ 'Age 15'), levels=c('Age 9', 'Age 15'))), include.miss = T)), 'childteen'))
+smoking_tbl=modeldata %>% 
+  dplyr::select(smkPreg_binary, m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,
+         SmkAtVisitPastmonth, ancestry, cm1bsex, globalmethylation, pediatric, Epithelial.cells_saliva,
+         Leukocytes_saliva, SSnewbornCT_center, SSolder_center, cg05575921, childteen)%>%
+  tbl_strata(strata=childteen, .tbl_fun=~.x %>%
+               tbl_summary(by=smkPreg_binary, statistic=
+                             list(all_continuous()~"{mean} ({sd})"))%>%
+               add_n()%>%add_p(test=list(all_continuous()~"t.test")))
+
+smoking_tbl=smoking_tbl%>%as_gt()%>%
+  tab_row_group(label='Maternal', rows = 1:11)%>%
+  tab_row_group(label='Child', rows=12:25)
+
+smoking_tbl
+
+main_tab=main_tab+1
 ```
 
-# Supplemental Figure: *A priori* CpG sites methylation and prenatal maternal smoking by age and ancestry
-
-# Supplemental Figure - A) Ancestry and B) age differences in methylation summary measures by prenatal maternal smoking exposure
-
-```{r}
-plot_grid(
-  age_long_methyl%>%filter(Methyl_factor %in% main_methyl)%>%
-    ggplot(aes(x=smkPreg_binary, y=`Methylation summary measure`, color=ancestry))+
-    geom_violin()+geom_boxplot(width=0.2, position=position_dodge(0.9))+stat_compare_means(label="..p.signif..")+
-    scale_color_discrete(name='Ancestry')+
-    scale_x_discrete(name='Prenatal smoke exposure', labels=c('No', 'Yes'))+
-    facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels))+
-    theme(legend.position = 'bottom'),
-
-age_long_methyl%>%filter(Methyl_factor %in% main_methyl)%>%
-  ggplot(aes(x=smkPreg_binary, y=`Methylation summary measure`, color=childteen))+
-  geom_violin()+geom_boxplot(width=0.2, position=position_dodge(0.9))+stat_compare_means(label="..p.signif..")+
-  scale_color_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+
-  scale_x_discrete(name='Prenatal smoke exposure', labels=c('No', 'Yes'))+
-  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels))+
-  theme(legend.position = 'bottom'),
-
-labels=c('A', 'B'), nrow=1)
-```
-
-# Supplemental Table XX: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable models among `r n_total` US children
-
-```{r}
-my_main_outcomes=c('AHHR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)', 'Sustained smoking polymethylation score (older children)', 'Pediatric clock', 'Global methylation')
-
-test=rbind(global_models, longitudinal_global_models %>% mutate(childteen='Longitudinal'))%>%
-  filter(predictors=='secondhand smoke exposure model')%>%
-  filter(outcome %in% my_main_outcomes)%>%
-  mutate(childteen=factor(case_when(childteen=='C'~'Age 9 models', 
-                             childteen=='T'~'Age 15 models', 
-                             childteen=='Longitudinal'~'Longitudinal models'),levels=c('Age 9 models', 'Age 15 models', 'Longitudinal models')), 
-         tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
-                                        show_single_row='smkPreg_binary', 
-                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
-
-table_list=test %>%split(f=.$childteen)%>%map(~tbl_merge(.x$tbl, 
-                                                         tab_spanner = as.character(.x$outcome)))
-
-regression_table_age=tbl_stack(table_list, group_header=names(table_list))
-
-```
-
-```{r}
-test=rbind(longitudinal_global_models %>% mutate(ancestry='Unstratified'),
-           longitudinal_local_models%>%dplyr::select(-data))%>%
-  filter(predictors=='secondhand smoke exposure model')%>%
-  filter(outcome %in% my_main_outcomes)%>%
-  mutate(tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
-                                        show_single_row='smkPreg_binary', 
-                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
-
-table_list=test %>%split(f=.$ancestry)%>%map(~tbl_merge(.x$tbl, 
-                                                         tab_spanner = as.character(.x$outcome)))
-
-regression_table_ancestry=tbl_stack(table_list, group_header=names(table_list))
-
-```
-
-```{r, width=11, eval=F}
-#alternatively 
-my_names=c('Methylation measure', paste0(c(paste0(paste0('Unstratified <br> n=', n_total), footnote_marker_alphabet(1, 'html')), paste0(modeldata%>%group_by(ancestry)%>%dplyr::select(ancestry, idnum)%>%unique()%>%summarise(n=n())%>%transmute(label=paste0(ancestry, ' <br> n=', n))%>%pull(label), footnote_marker_alphabet(2, 'html'))), '<br> Beta (95% CI); p-value'))
-
-tb2=rbind(longitudinal_global_models %>% mutate(ancestry='Unstratified'), longitudinal_local_models%>%dplyr::select(-data))%>%filter(predictors=='secondhand smoke exposure model')%>%filter(outcome %in% my_main_outcomes)%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%mutate(estimate=paste0(my_round(estimate, 2), ' (', my_round(conf.low, 2), ', ', my_round(conf.high, 2), ') p=', signif(p.value, 2)))%>%dplyr::select(ancestry, outcome, estimate)%>%pivot_wider(names_from=ancestry, values_from=c(estimate))
-
-names(tb2)=my_names
-
-tb2%>%kable(., escape=F)%>%kable_minimal()%>%
-  kableExtra::scroll_box(width = "100%")%>% footnote(alphabet =  c("Model adjusted for:child sex, mother income to poverty ratio at birth, epithelial and immune cell proportions (estimated from methylation data), maternal other drug and other alcohol use during pregnancy, batch of array, genetic ancestry (1st two principal components of genetic data), postnatal smoke exposure (any postnatal maternal smoking and maternal packs per day in month prior to visit).", "Models adjusted as in (a) but principal components of genetic ancestry rerun within ancestry strata and first two components from these stratified principal components analyses used"))%>%kable_styling(font_size = 10)
-
-```
-
-# Figure 3 - Polymethylation scores are associated with prenatal maternal smoke exposure consistently across age and ancestry group in a sample of `r n_total` US children
+# Figure `r main_fig` - Polymethylation scores are associated with prenatal maternal smoke exposure consistently across age and ancestry group in a sample of `r n_total` US children
 
 ```{r, fig.width=12}
-my_outcomes=c('Global methylation', 'Pediatric clock', 'AHHR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)')
+my_outcomes=c('Global methylation', 'Pediatric clock', 'AHRR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)')
 
-outcome_labels=c('Global methylation', 'Pediatric clock', 'AHHR: cg05575921',  'Polymethylation score \n (newborn, mean centered)')
+outcome_labels=c('Global methylation', 'Pediatric clock', 'AHRR: cg05575921',  'Polymethylation score \n (newborn, mean centered)')
 
@@
 
-all_models%>%
+forest_plot=all_models%>%
   mutate(n=map_dbl(data, nrow), n=paste0('n=', n))%>%
@@
   scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
-  scale_y_discrete('', limits=c('european', 'hispanic', 'african', 'Unstratified'), labels=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))
+  scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'), labels=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))
+
+forest_plot
+
+ggexport(forest_plot, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')), width=15)
+main_fig=main_fig+1
 ```
 
-# Supplemental Figure XX: Surrogate variable associations with known covariates
+# Supplemental Figure `r sup_fig`: Sensitivity analyses for controlling for different variable sets in multivariable analyses of prenatal maternal smoking and DNA methylation summary measures
+
+```{r, fig.width=14}
+with(modeldata, table(ancestry, PostnatalMaternalSmokingAny, smkPreg_binary))%>%kbl()
+
+forest_plot_compareModels=all_models%>%
+    mutate(n=map_dbl(data, nrow), n=paste0('n=', n))%>%
+    unnest(tidied)%>%
+    filter(term=='smkPreg_binaryYes')%>%
+    filter(outcome%in% my_outcomes)%>%
+    mutate(outcome_f=factor(outcome, levels=rev(my_outcomes)))%>%
+    ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=ancestry))+
+    geom_point(position = position_dodge(width = 1))+
+    geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
+    facet_grid(predictors~outcome_f, scales='free_x', labeller = labeller(outcome_f=outcome_labels))+
+    geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
+    theme_classic()+
+    theme(text = element_text(size=16))+theme(legend.position = 'bottom')+
+    scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
+    scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'), labels=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))
+
+forest_plot_compareModels
+
+ggexport(forest_plot_compareModels, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=15, height=15)
+sup_fig=sup_fig+1
+```
+
+# Supplemental Figure `r sup_fig` DNA methylation summary measures across prenatal and postnatal exposures
+
+```{r, fig.width=14, fig.height=10}
+
+my_comparisons <- list(c("No prenatal smoke exposed & No maternal smoking at age 1 and 5", 
+                        "No prenatal smoke exposed & Maternal smoking at age 1 or age 5"), 
+                        c("No prenatal smoke exposed & Maternal smoking at age 1 or age 5", 
+                          "Yes prenatal smoke exposed & Maternal smoking at age 1 or age 5"), 
+                      c("Yes prenatal smoke exposed & No maternal smoking at age 1 and 5", 
+                        "Yes prenatal smoke exposed & Maternal smoking at age 1 or age 5"))
+
+n_count=modeldata%>%mutate(exposed=paste0(smkPreg_binary, ' prenatal smoke exposed & ', PostnatalMaternalSmokingAny))%>% group_by(childteen, exposed)%>%summarise(n=n())
+
+give.n <- function(x){
+  return(c(y = min(x)-0.1, label = length(x))) 
+  # experiment with the multiplier to find the perfect position
+}
+
+modeldata%>%
+  dplyr::select(any_of(c('idnum', 'childteen', 'ancestry', 'smkPreg_binary', 'PostnatalMaternalSmokingAny', all_methyl)))%>%
+  pivot_longer(cols=any_of(all_methyl),
+               names_to='Methylation summary measure type',
+               values_to='Methylation summary measure')%>%
+  mutate(Methyl_factor=factor(`Methylation summary measure type`, 
+                              levels=all_methyl_levels))%>%
+
+  mutate(exposed=paste0(smkPreg_binary, ' prenatal smoke exposed & ', PostnatalMaternalSmokingAny))%>% 
+  filter(Methyl_factor%in% main_methyl[3:4])%>%
+  ggplot(aes(x=exposed, y=`Methylation summary measure`, color=exposed))+
+    geom_violin()+geom_boxplot(width=0.2, position = position_dodge(0.9))+
+  stat_summary(fun.data = give.n, geom = "text", fun.y = max,
+                  position = position_dodge(width = 0.75))+
+  stat_compare_means(comparisons=my_comparisons)+
+    facet_grid(Methyl_factor~childteen, scales='free',labeller = labeller(Methyl_factor=all_methyl_labels))+
+    scale_x_discrete(labels=function(x) str_wrap(x, width = 10))+
+    theme(legend.position = 'right')+
+  guides(color=guide_legend(nrow=4,byrow=TRUE))
+sup_fig=sup_fig+1
+```
+
+# Supplemental Table `r sup_tab`: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable models among `r n_total` US children by age
 
 ```{r}
+#my_main_outcomes=c('AHRR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)', 'Sustained smoking polymethylation score (older children)', 'Pediatric clock', 'Global methylation')
+
+test=rbind(global_models, longitudinal_global_models %>% mutate(childteen='Longitudinal'))%>%
+  filter(predictors=='secondhand smoke exposure model')%>%
+  filter(outcome %in% my_outcomes)%>%
+  mutate(childteen=factor(case_when(childteen=='Age 9'~'Age 9 models', 
+                             childteen=='Age 15'~'Age 15 models', 
+                             childteen=='Longitudinal'~'Longitudinal models'),levels=c('Age 9 models', 'Age 15 models', 'Longitudinal models')), 
+         tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
+                                        show_single_row='smkPreg_binary', 
+                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
+
+table_list=test %>%split(f=.$childteen)%>%map(~tbl_merge(.x$tbl, 
+                                                         tab_spanner = as.character(.x$outcome)))
+
+regression_table_age=tbl_stack(table_list, group_header=names(table_list))
+
+regression_table_age
+
+sup_tab=sup_tab+1
+```
+
+# Supplemental Table `r sup_tab`: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable models among `r n_total` US children by ancestry
+
+```{r}
+test=rbind(longitudinal_global_models %>% mutate(ancestry='Unstratified'),
+           longitudinal_local_models%>%dplyr::select(-data))%>%
+  filter(predictors=='secondhand smoke exposure model')%>%
+  filter(outcome %in% my_outcomes)%>%
+  mutate(tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
+                                        show_single_row='smkPreg_binary', 
+                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
+
+table_list=test %>%split(f=.$ancestry)%>%map(~tbl_merge(.x$tbl, 
+                                                         tab_spanner = as.character(.x$outcome)))
+
+regression_table_ancestry=tbl_stack(table_list, group_header=names(table_list))
+
+regression_table_ancestry
+
+sup_tab=sup_tab+1
+
+```
+
+# Supplemental Figure `r sup_fig`: Surrogate variable associations with known covariates
+
+```{r, fig.width=12, fig.height=8}
 #bring the heatmap function
@@
     ggplot(aes(x=var1, y=var2, fill=value))+geom_tile()+
-    scale_fill_gradientn(colours = color.gradient(20), values = seq(0, 1, 0.5))+
-    scale_x_discrete(labels=str_wrap(get_label(pd[, covar]), 10))
+    scale_fill_gradientn('P value', colours = color.gradient(20), values = seq(0, 1, 0.5))+
+    scale_x_discrete('', labels=str_wrap(get_label(pd[, covar]), 10))+
+    scale_y_discrete('')
 }
@@
 bring.the.heatmap.sv(modeldata, covar=c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'ancestry', 'cm1bsex', 'Leukocytes_saliva', 'Epithelial.cells_saliva', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))
+
+pdf(file=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=11, height=13)
+bring.the.heatmap.sv(modeldata, covar=c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'ancestry', 'cm1bsex', 'Leukocytes_saliva', 'Epithelial.cells_saliva', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))
+dev.off()
+sup_fig=sup_fig+1
 ```
 
-# Supplemental Table XX: Surrogate variable models
+# Supplemental Table `r sup_tab`: Coefficients for multivariable models of prenatal maternal smoke exposure regressed on selected DNA methylation summary variables (outcomes) among children in the Fragile Families and Child Wellbeing study
 
@@
   filter(predictors=='surrogate variable models')%>%
-  filter(outcome %in% my_main_outcomes)%>%
-  mutate(childteen=factor(case_when(childteen=='C'~'Age 9 models', 
-                             childteen=='T'~'Age 15 models', 
+  filter(outcome %in% my_outcomes)%>%
+  mutate(childteen=factor(case_when(childteen=='Age 9'~'Age 9 models', 
+                             childteen=='Age 15'~'Age 15 models', 
                              childteen=='Longitudinal'~'Longitudinal models'),levels=c('Age 9 models', 'Age 15 models', 'Longitudinal models')), 
@@
   filter(predictors=='surrogate variable models')%>%
-  filter(outcome %in% my_main_outcomes)%>%
+  filter(outcome %in% my_outcomes)%>%
   mutate(tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
@@
 regression_table_ancestry_svs=tbl_stack(table_list, group_header=names(table_list))
+
+regression_table_ancestry_svs
+
+sup_tab=sup_tab+1
 ```
 
-# Figure 4 - Polymethylation scores more accurately delineate prenatal smoke exposure status than non-targeted DNA methylation summary measures and *a priori* CpGs
+# Figure `r main_fig` - Polymethylation scores more accurately delineate prenatal smoke exposure status than non-targeted DNA methylation summary measures and *a priori* CpGs
 
@@
 
-age_labels=c('Age 9', 'Age 15'); names(age_labels)=c('C', 'T')
+age_labels=c('Age 9', 'Age 15'); names(age_labels)=c('Age 9', 'Age 15')
 methyl_labels_roc=paste0('Base model+', all_methyl_labels)
@@
     theme_classic()+
-    scale_color_manual('', values=c('red', 'darkred', 'darkblue', 'darkgreen', 'blue', 'green'))+
-    theme(legend.position = 'right')+guides(color=guide_legend(nrow=2, byrow=TRUE))+
+    scale_color_manual('', values=c('red', 'darkred', 'darkblue', 'darkgreen', 'blue', 'green'), 
+                       function(x) str_wrap(x, width = 30))+
+    theme(legend.position = 'right')+guides(color=guide_legend(nrow=4, byrow=TRUE))+
     geom_abline(intercept=1, slope=-1, color='black')+
@@
 fig4b=global_roc%>%
-  dplyr::select(covariates, methylation, childteen, test)%>%unnest(test) %>% 
-  filter(str_detect(methylation, 'SSnewbornCT'))%>%
+  dplyr::select(covariates, methylation, childteen, test)%>% 
+  filter(str_detect(methylation, 'SSnewbornCT') & covariates!='')%>%unnest(test)%>%
   rbind(long_global_roc%>% 
-          mutate(childteen='C+T', methylation=gsub('_C.*', '', methylation))%>%
+          mutate(childteen='Age 9 & Age 15', methylation=gsub('_Age_9.*', '', methylation))%>%
           dplyr::select(covariates, methylation, childteen, test)%>% 
-          unnest(test) %>% 
           filter(str_detect(methylation, 'SSnewbornCT') 
-                 & covariates!=''))%>%
+                 & covariates!='')%>%
+          unnest(test))%>%
   ggplot(aes(x=spec, y=sens, color=childteen, linetype=childteen))+geom_line(size=2)+
   facet_wrap(~methylation, labeller = labeller(methylation=c('SSnewbornCT_center'='Polymethylation score, \n newborn mean centered')))+
-  theme_classic()+theme(legend.position = 'right')+guides(color=guide_legend(nrow=2,byrow=TRUE))+
+  theme_classic()+theme(legend.position = 'right')+guides(color=guide_legend(nrow=4,byrow=TRUE))+
   geom_abline(intercept=1, slope=-1, color='black')+
@@
 
-cowplot::plot_grid(fig4a, fig4b, labels=c('A', 'B'), nrow=1, rel_widths=c(0.65, 0.35), align='h', axis='t')
+auc_fig=cowplot::plot_grid(fig4a, fig4b, labels=c('A', 'B'), nrow=1, rel_widths=c(0.65, 0.35), align='h', axis='t')
 
+auc_fig
+
+ggexport(auc_fig, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')), width=9)
+main_fig=main_fig+1
 ```
+
+# Supplemental Table with AUC and p-value comparison to sustained smoking, newborn cell type controls polymethylation score
 
@@
 roc_table=global_roc %>% dplyr::select(childteen, methylation, modeltype, roc, auc)
-
 roc_base=roc_table %>% filter(modeltype=='Base model+No methylation')%>%mutate(roc_basemodel=roc)%>%dplyr::select(childteen, roc_basemodel)
@@
 
-roc_table1=roc_table%>%filter(!str_detect(modeltype, 'Base model'))%>%left_join(roc_ss)%>%mutate(roctest=map2(roc, roc_pms, ~roc.test(.x, .y)))
-roc_table1%>%mutate(`pval comparison to sustained smoking score (newborns)`=map_dbl(roctest, ~.x$p.value))%>%dplyr::select(childteen, modeltype, auc, `pval comparison to sustained smoking score (newborns)`)
+roc_table1=roc_table%>%filter(!str_detect(modeltype, 'Base model'))%>%left_join(roc_ss)%>%mutate(roctest=map2(roc, roc_pms, ~roc.test(.x, .y)))%>%mutate(auc=round(auc, 2))
 
-roc_table2=roc_table%>%filter(str_detect(modeltype, 'Base model'))%>%left_join(roc_base)%>%mutate(roctest=map2(roc, roc_basemodel, ~roc.test(.x, .y)))
-roc_table2%>%mutate(`pval comparison to base model`=map_dbl(roctest, ~.x$p.value))%>%dplyr::select(childteen, modeltype, auc, `pval comparison to base model`)
+roc_table1%>%mutate(`pval comparison to sustained smoking score (newborns)`=map_dbl(roctest, ~round(.x$p.value, 4)))%>%dplyr::select(childteen, modeltype, auc, `pval comparison to sustained smoking score (newborns)`)%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value (vs polymethylation score, newborns, cell type controlled)'))
+```
+
+# Supplemental Table with AUC and p-value comparison to base model with no methylation variables (i.e., using only child sex, income to poverty ratio, saliva/epithelial cell proportions and sample plate)
+
+```{r}
+
+roc_table2=roc_table%>%filter(str_detect(modeltype, 'Base model'))%>%left_join(roc_base)%>%mutate(roctest=map2(roc, roc_basemodel, ~roc.test(.x, .y)))%>%mutate(auc=round(auc, 2))
+
+roc_table2%>%mutate(`pval comparison to base model`=map_dbl(roctest, ~round(.x$p.value, 4)))%>%dplyr::select(childteen, modeltype, auc, `pval comparison to base model`)%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value (vs base w/ no methyl)'))
+
 ```
@@
 - Arguments in favor: recent paper in Clinical Epigenetics included both 
-- Arguments against: enough here for a compellling story; deadline of December 1st; submit and if asked for we have it available.
+- Arguments against: enough here for a compelling story; deadline of December 1st; submit and if asked for we have it available.
 
-# Cross sectional EWAS results:
+```{r}
 
-# Longitudinal EWAS results
+calcLambda<-function(df, type='lm'){
+  if(type=='lme'){df=df%>%mutate(P.Value=p.value)}
+  o<- -log10(sort(df$P.Value, decreasing = F))
+  e<- -log10(ppoints(length(df$P.Value)))
+  lambda<-median(o)/median(e)
+  return(lambda)
+}
+```
 
-# METAL meta-analysis of 450K + EPIC chip data
+```{r}
+child_m1=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'child1_TopHits.RDS'))
+teen_m1=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'teen1_TopHits.RDS'))
+```
+
+```{r}
+#lme_base=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'lme_ewas_base.RDS'))
+#lme_smk = lme_base %>% filter(term=='smkPreg_binaryYes')%>%mutate(p.bh=p.adjust(p.value, 'hochberg'))
+#lme_smk=add_annotations(lme_smk, type='lme')
+lme_smk=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'lme_smoking.RDS'))
+#mutate columns to get matching columns in lme_smk as in child & teen _m1
+lme_smk = lme_smk %>% mutate(logFC=estimate, CI.L=conf.low, CI.R=conf.high, P.Value=p.value, adj.P.Val=p.bh)
+```
+
+```{r, eval=F}
+library(variancePartition)
+dream=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'dream_ewas.RDS'))
+dream_smk=topTable(dream, coef='smkPreg_binaryYes', number=nrow(dream$p.value))
+#dream_smk=add_annotations(dream_smk)
+```
+
+## QQ plots
+
+```{r}
+
+png(file=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.png')))
+par(mfrow=c(3, 1), mar=c(5.1, 4.1, 5.1, 2.1))
+  qqman::qq(child_m1$P.Value, main='Age 9: Base model EWAS')
+  mtext(paste0(expression(lambda), '=', round(calcLambda(child_m1), 3)), side=3)
+
+  qqman::qq(teen_m1$P.Value, main='Age 15: Base model EWAS')
+  mtext(paste0(expression(lambda), '=', round(calcLambda(teen_m1), 3)), side=3)
+
+  qqman::qq(lme_smk$P.Value, main='Mixed effect models: Base model EWAS')
+  mtext(paste0(expression(lambda), '=', round(calcLambda(lme_smk), 3)), side=3)
+dev.off()
+
+par(mfrow=c(3, 1), mar=c(5.1, 4.1, 5.1, 2.1))
+  qqman::qq(child_m1$P.Value, main='Age 9: Base model EWAS')
+  mtext(paste0(expression(lambda), '=', round(calcLambda(child_m1), 3)), side=3)
+
+  qqman::qq(teen_m1$P.Value, main='Age 15: Base model EWAS')
+  mtext(paste0(expression(lambda), '=', round(calcLambda(teen_m1), 3)), side=3)
+
+  qqman::qq(lme_smk$P.Value, main='Mixed effect models: Base model EWAS')
+  mtext(paste0(expression(lambda), '=', round(calcLambda(lme_smk), 3)), side=3)
+
+sup_fig=sup_fig+1
+
+```
+
+## Beta vs beta plot: Age 9 vs age 15
+
+```{r, cache=T}
+library(tibble)
+base_models=list('Age_9'=child_m1%>%rownames_to_column(var='CPG'), 'Age_15'=teen_m1%>%rownames_to_column(var='CPG'), "Mixed_effect_model"=lme_smk)%>%Map(cbind, ., ModelType=names(.))
+
+beta_base<-lapply(base_models, function(x) x %>%dplyr::select(CPG, logFC, ModelType)%>%pivot_wider(id_cols = CPG, values_from=logFC, names_from=ModelType))%>%purrr::reduce(left_join, by='CPG')
+
+smooth_betas<-function(df, x, y){
+  ggplot(df, aes_string(x=x, y=y))+
+    geom_point()+geom_smooth(method='lm')+
+    stat_cor()+theme_classic()+
+    scale_y_continuous(name=y)+
+   scale_x_continuous(name=x)+
+    coord_fixed()
+}
+
+sig_beta_m1<-child_m1%>%rownames_to_column(var='CPG')%>%filter(P.Value<10e-4)%>%pull(CPG)
+
+beta_compare=ggarrange( 
+  smooth_betas(beta_base, "Age_9", "Age_15"),
+  smooth_betas(beta_base %>%filter(CPG %in% sig_beta_m1), 'Age_9', 'Age_15'), 
+  labels=c('A', 'B'), align='h')
+
+beta_compare
+ggexport(beta_compare, filename = file.path(outdir, paste0('SupplementalFigure', sup_fig, '.png'))) 
+```
+
+## Upset plots
+
+```{r}
+library(ggupset)
+sig_cpg_list<-map(base_models, ~.x %>%filter(P.Value<10e-4)%>%dplyr::select(CPG, ModelType))%>%
+  bind_rows()%>%
+  group_by(CPG)%>%summarise(Models=list(ModelType))
+
+upset_cpgs=ggplot(sig_cpg_list, aes(x=Models))+
+    geom_bar() +
+    scale_x_upset()
+ggexport(upset_cpgs, filename = file.path(outdir, paste0('SupplementaalFigure', sup_fig, '.pdf')))
+sup_fig=sup_fig+1
+```
+
+## Volcano plot
+
+```{r, fig.width=14}
+volcs=ggarrange(
+ggplot(child_m1%>%mutate(adjPval=ifelse(adj.P.Val<0.05, '<0.05', ifelse(adj.P.Val<0.2, '<0.2', '>0.2'))), aes(x=logFC, y=-log10(P.Value), color=adjPval))+geom_point()+scale_x_continuous('Beta (regression coefficient)')+
+  scale_color_manual('Benjamini-Hochberg adjusted P value', values=c('red', 'orange', 'black'))+
+  scale_y_continuous(limits=c(0, 9), breaks=c(0, 2, 4, 6, 8))+
+  scale_x_continuous(limits=c(-0.06, 0.06), breaks=c(-0.05, 0, 0.05)),
+ggplot(teen_m1%>%mutate(adjPval=ifelse(adj.P.Val<0.05, '<0.05', ifelse(adj.P.Val<0.2, '<0.2', '>0.2'))), aes(x=logFC, y=-log10(P.Value), color=adjPval))+geom_point()+scale_x_continuous('Beta (regression coefficient)')+
+  scale_color_manual('Benjamini-Hochberg adjusted P value', values=c('red', 'orange', 'black'))+
+  scale_y_continuous(limits=c(0, 9), breaks=c(0, 2, 4, 6, 8))+
+  scale_x_continuous(limits=c(-0.06, 0.06), breaks=c(-0.05, 0, 0.05)),
+ggplot(lme_smk%>%mutate(adjPval=ifelse(adj.P.Val<0.05, '<0.05', ifelse(adj.P.Val<0.2, '<0.2', '>0.2'))), aes(x=logFC, y=-log10(P.Value), color=adjPval))+geom_point()+scale_x_continuous('Beta (regression coefficient)')+
+  scale_color_manual('Benjamini-Hochberg adjusted P value', values=c('red', 'orange', 'black'))+
+  scale_y_continuous(limits=c(0, 9), breaks=c(0, 2, 4, 6, 8))+
+  scale_x_continuous(limits=c(-0.06, 0.06), breaks=c(-0.05, 0, 0.05)),
+labels=c('A', 'B', 'C'), 
+common.legend = TRUE, 
+nrow=1
+)
+
+ggexport(volcs, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')))
+ggexport(volcs, filename=file.path(outdir, paste0('Figure', main_fig, '.png')))
+
+volcs
+```
+
+## Gap hunter
+
+```{r}
+all_gaps=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'gaphunter_gaps.RDS'))
+```
+
+## Compare across models and studies
+
+```{r, fig.width=14}
+topCPGs<-read.csv(paste0(datadir, "/TopCpGsWiklundEtAl2019.csv"))
+topCPGs<-topCPGs %>% mutate(estimate=as.numeric(gsub("−\\s", "-", gsub(" .*", "", β..SE.))), std.error=as.numeric(sub("\\).*", "", sub(".*\\(", "", β..SE.))), p.value=P.value, source="Wiklund et al.
+2019", conf.low=(estimate-1.96*std.error), conf.high=(estimate+1.96*std.error), set ='')%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)
+
+#read in Joubert meta top cpgs 
+#read in cpgs from PMC4833289
+cpgs<-read.csv(paste0(datadir, '/PMC4833289_SuppFile2CpGs.csv'), stringsAsFactors = F, header=F, skip=2)
+#make headers
+header1<-scan(paste0(datadir, '/PMC4833289_SuppFile2CpGs.csv'), nlines=1, what=character(), sep=',')
+header1[7:26]<-c(rep("SSnewborn", 5), rep("SSnewbornCT", 5), rep("SSolder", 5), rep("anynewborn", 5))
+header2<-paste(header1, scan(paste0(datadir, '/PMC4833289_SuppFile2CpGs.csv'), nlines=1, skip=1, what=character(), sep=','), sep='_')
+names(cpgs)<-header2
+
+#four meta-EWASes: SS_newborn (sustained smoking in newborns); SS_newbornCT (sustained smoking in newborns controlling for cell type) SS_older (sustained smoking in older children); any_newborn (any smoking in newborns)
+#pivot_longer to get single column of coefs and a column for the analysis type
+#and group into a list by analysis type
+cpgs<-cpgs %>% pivot_longer(cols = c(matches("_Coef|_SE|_P|_FDR|_Direction")), names_to = c("set", ".value"), names_pattern="(.+)_(.+)")%>%rename_with(~gsub('_', '', .x))%>%mutate(Gene=`Mapped Gene`, estimate=Coef, conf.low=estimate-1.96*SE, conf.high=estimate+1.96*SE, source='Joubert et al.
+2016', std.error=SE, p.value=P, Chr=CHR) %>%mutate(source=paste0(source, ': ', set))%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)%>%filter(set %in% c('SSolder', 'SSnewbornCT'))
+
+#myCPGs<-AllModelsdf %>%filter(Model=='model1')%>%mutate(data=map(data, function(x) x %>%filter(CpG %in% sigCpGs)))%>%unnest(cols=c(data))%>%mutate(Joubertetal=ifelse(CpG %in% cpgs$CpG, 'in Joubert et al.
+2016', 'Not in Joubert et al.
+2016'), Wicklundetal=ifelse(CpG %in% topCPGs$CpG, 'in Wicklund et al.
+2019', 'Not in Wicklund et al.
+2019'))%>%mutate(Gene=gene, estimate=logFC, conf.low=CI.L, conf.high=CI.R, source=paste0('Fragile Families:', Age), std.error='NA', p.value=P.Value, Chr=chr, set='Model1', Position=pos)
+
+```
+
+```{r, fig.width=14}
+comparison=left_join(
+  lme_smk%>%dplyr::select(CPG, gene, estimate, p.value)%>%dplyr::rename_with(~paste0('FF_', .x), .cols=c(estimate, p.value)), 
+  cpgs%>%filter(set=='SSnewbornCT')%>%mutate(CPG=CpG)%>%dplyr::select(CPG, Gene, estimate, p.value)%>%dplyr::rename_with( ~paste0('Joubert_', .x), .cols=c(estimate, p.value))
+  )%>%filter(!is.na(Joubert_estimate))
+
+comp_nofilter=ggplot(comparison, aes(x=Joubert_estimate, y=FF_estimate, color=FF_p.value))+
+  geom_point()+stat_cor()+
+  scale_y_continuous('Estimate in Fragile Families')+
+  scale_x_continuous('Estimate in Joubert et al. metaanalysis')+
+  scale_color_continuous('P value in Fragile Families')
+
+comparison %>% mutate(s1=sign(FF_estimate), s2=sign(Joubert_estimate), agreement=case_when(s1==s2~'Direction consistent', s1<s2~'FF -, Joubert +', s1>s2~'FF+, Joubert-'))%>%group_by(agreement)%>%summarise(n=n())
+
+comp_filter=ggplot(comparison%>%filter(FF_p.value<10e-4), aes(x=Joubert_estimate, y=FF_estimate, color=FF_p.value, label=ifelse(abs(Joubert_estimate)>0.04 , gene, NA)))+
+  geom_point()+stat_cor()+
+  geom_text_repel()+
+  scale_y_continuous('Estimate in Fragile Families')+
+  scale_x_continuous('Estimate in Joubert et al. metaanalysis')+
+  scale_color_continuous('P value in Fragile Families')
+
+comp_joubert=ggarrange(comp_nofilter, comp_filter, common.legend = F, labels=c('A', 'B'))
+ggexport(comp_joubert, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.png')))
+
+comp_joubert
+```
+
+## Additional covariate control and ancestry stratified models at top hits
+
+```{r}
+top_cpgs = map(base_models, ~.x %>%filter(adj.P.Val<0.05)%>%dplyr::select(CPG, ModelType, gene))%>%bind_rows()%>%mutate(label=paste0(gene, ': ', CPG))
+top_cpg_gaps=top_cpgs[top_cpgs$CPG %in% rownames(all_gaps$proberesults),]
+top_cpgs=top_cpgs[!(top_cpgs$CPG %in% rownames(all_gaps$proberesults)),]
+top_cpg_groups=top_cpgs %>%group_by(CPG)%>%dplyr::summarise(models=list(ModelType))
+top_cpg_names=c(top_cpgs%>%pull(label)%>%unique(), 'AHRR: cg05575921', 'MYO1G: cg22132788')
+top_cpgs=c(top_cpgs %>% pull(CPG)%>%unique(), 'cg05575921', 'cg22132788')
+
+betaqc<-readRDS(file=file.path(datadir, 'OGData', "betaqc.rds"))
+my_beta=as.data.frame(t(betaqc[row.names(betaqc)%in% top_cpgs, ])) %>% mutate(MethID=row.names(.))
+modeldata=left_join(modeldata, my_beta)
+```
+
+```{r, eval=F}
+y_vector<-top_cpgs
+outcome_labels=top_cpg_names
+names(outcome_labels)=y_vector
+
+#model variables (excepting PCs, add in individually for global & ancestry specific models)
+base_model_vars<-"~smkPreg_binary+cm1bsex+cm1inpov+Leukocytes_saliva+Epithelial.cells_saliva+Sample_Plate"
+prenatal_model_vars<-paste0(base_model_vars, "+m1g2_YesNoPreg+m1g3_YesNoPreg")
+secondhand_model_vars<-paste0(prenatal_model_vars, '+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth')
+interaction_model_vars<-paste0(secondhand_model_vars, '+ChildAgeComposite:smkPreg_binary')
+#pc variables
+global_pcs='+global_PC1+global_PC2'
+local_pcs='+local_PC1+local_PC2'
+
+#surrogate variable models
+surrogate_model_vars<-paste0('~smkPreg_binary+', 
+                             paste(colnames(modeldata)[str_detect(colnames(modeldata), 'sv')], collapse='+'))
+
+predictors=c('base model'=base_model_vars, 
+             'prenatal exposures model'=prenatal_model_vars, 
+             'secondhand smoke exposure model'=secondhand_model_vars)
+
+predictors_interaction=c(predictors, 'interaction model'=interaction_model_vars)
+
+model_labels=c('base model', 'prenatal exposures model', 'secondhand smoke exposure model', 'surrogate variable models')
+
+model_labels_interaction=c('base model', 'prenatal exposures model', 'secondhand smoke exposure model', 'interaction model (age*smoke)', 'surrogate variable models')
+
+#stratification variables 
+##############################
+age_labels=c('Age 9', 'Age 15')
+age_vector=modeldata$childteen%>%unique()
+names(age_labels)=age_vector
+ancestry_vector=modeldata$ancestry%>%unique()
+
+#################################Source functions#######################################
+model<-function(df, y, x){
+  lm(formula(paste(y, x)), data=df)
+}
+
+model_lmm<-function(df, y, x){
+  xnew=paste0(x, '+', '(1|id)')
+  lmer(formula(paste(y, xnew)), data=df, REML=F)
+}
+
+my_clean_argnames=function(df, my_preds, childteen=T, model_labs=model_labels){
+  names(model_labs)=my_preds
+  df=df %>% mutate(predictors=recode_factor(predictors, !!!model_labs),
+                outcome=recode_factor(outcome, !!!outcome_labels))
+  if(childteen==T){df=df%>%mutate(age=recode_factor(childteen, !!!age_labels))} 
+  return(df)
+}
+
+#################################Cross sectional models###########################
+#########################Global
+#set predictor variables & arguments
+global_predictors=c(paste0(predictors, global_pcs), surrogate_model_vars)
+args=list('childteen'=age_vector, 'outcome'=top_cpgs, 'predictors'=global_predictors)%>%cross_df()
+
+#run models
+global_models=modeldata%>%group_by(childteen)%>%nest()%>%left_join(args)%>%
+  mutate(model=pmap(list(data, outcome, predictors), model), 
+         tidied=map(model, tidy, conf.int=T), 
+         glanced=map(model, glance))
+
+#clean argument names for plotting
+global_models=my_clean_argnames(global_models, global_predictors)
+
+#########################Local
+#arguments
+local_predictors=c(paste0(predictors, local_pcs), surrogate_model_vars)
+args=list('childteen'=c('Age 9', 'Age 15'), 'outcome'=top_cpgs, 'predictors'=local_predictors, 'ancestry'=ancestry_vector)%>%cross_df()
+
+#run models
+local_models=modeldata%>%group_by(childteen, ancestry)%>%nest()%>%left_join(args)%>%
+  mutate(model=pmap(list(data, outcome, predictors), model), 
+         tidied=map(model, tidy, conf.int=T), 
+         glanced=map(model, glance))
+
+#clean argument names for plotting
+local_models=my_clean_argnames(local_models, local_predictors)
+
+#################################Longitudinal models##################
+#########################Global
+#set predictor variables & arguments
+#include a model with an interaction term b/t age and prenatal smoke exposure
+#to determine if there is a age-slope difference b/t exposed & not exposed
+global_predictors=c(paste0(predictors_interaction, global_pcs, '+ChildAgeComposite'), surrogate_model_vars)
+args=list('outcome'=top_cpgs, 'predictors'=global_predictors)%>%cross_df()
+
+#run models
+longitudinal_global_models=args %>% 
+    mutate(model=pmap(list(y=outcome, x=predictors), model_lmm, df=modeldata), 
+           tidied=map(model, tidy, conf.int=T), 
+           glanced=map(model, glance))
+
+#clean argument names
+longitudinal_global_models=my_clean_argnames(longitudinal_global_models, global_predictors, childteen = F, model_labs = model_labels_interaction)
+#########################Local
+#set predictor variables and arguments
+local_predictors=c(paste0(predictors_interaction, local_pcs, '+ChildAgeComposite'), surrogate_model_vars)
+args=list('outcome'=top_cpgs, 'predictors'=local_predictors, 'ancestry'=ancestry_vector)%>%cross_df()
+
+#run models
+longitudinal_local_models=modeldata%>%group_by(ancestry)%>%nest()%>%left_join(args)%>%
+  mutate(model=pmap(list(data, outcome, predictors), model_lmm), 
+         tidied=map(model, tidy, conf.int=T), 
+         glanced=map(model, glance))
+
+#clean argument names
+longitudinal_local_models=my_clean_argnames(longitudinal_local_models, local_predictors, childteen = F, model_labs = model_labels_interaction)
+```
+
+```{r compare over time, eval=F}
+
+rbind(global_models%>%dplyr::select(childteen, predictors, outcome, tidied)%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes'), 
+      longitudinal_global_models%>%dplyr::select(predictors, outcome, tidied)%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%mutate(childteen='Linear mixed model'))%>%
+  filter(predictors %in% c('base model', 'secondhand smoke exposure model'))%>%
+  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=childteen, y=predictors))+
+    geom_point(position = position_dodge(width = 1))+
+    geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
+    facet_grid(.~outcome, scales='free_x', labeller = labeller(outcome_f=top_cpg_names))+
+    geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
+    theme_classic()
+
+cpg_timevariant=top_cpg_groups%>%mutate(count=map_dbl(models, length))%>%filter(count<3)%>%pull(CPG)
+
+age_long_methyl%>%
+  filter(`Methylation summary measure type` %in% top_cpgs)%>%
+  ggplot(aes(x=childteen, y=`Methylation summary measure`, color=childteen))+
+  geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+
+  scale_color_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+
+  scale_x_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+
+  geom_point(alpha=0)+geom_line(alpha=0, aes(group=idnum))+geom_smooth(aes(group=1), method='lm', color='black')+
+  facet_wrap(~Methyl_factor, scales='free')+
+  theme(legend.position = 'bottom')
+
+```
+
+```{r compare over ancestries, eval=F}
+longitudinal_local_models%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%
+  filter(predictors %in% c('base model', 'secondhand smoke exposure model'))%>%
+  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=ancestry, y=predictors))+
+    geom_point(position = position_dodge(width = 1))+
+    geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
+    facet_grid(.~outcome, scales='free_x', labeller = labeller(outcome_f=top_cpg_names))+
+    geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
+    theme_classic()
+
+longitudinal_local_models%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%
+    filter(predictors %in% c('base model'))%>%group_by(outcome)%>%summarise(est=list(estimate))%>%mutate(sign=map_dbl(est, ~sum(sign(.x))))%>%filter(abs(sign)!=3)
+
+```
+
+```{r, fig.width=14}
+
+age_long_methyl=modeldata%>%
+  dplyr::select(any_of(c('idnum', 'childteen', 'ancestry', 'smkPreg_binary', top_cpgs)))%>%
+  pivot_longer(cols=any_of(top_cpgs),
+               names_to='Methylation summary measure type',
+               values_to='Methylation summary measure')%>%
+  mutate(Methyl_factor=factor(`Methylation summary measure type`, 
+                              levels=top_cpgs))
+
+ancestry_differences=age_long_methyl%>%
+    filter(`Methylation summary measure type` %in% top_cpgs)%>%
+    ggplot(aes(x=ancestry, y=`Methylation summary measure`, color=smkPreg_binary))+
+    geom_violin()+geom_boxplot(width=0.2, position=position_dodge(0.9))+stat_compare_means(label="..p.signif..", label.y.npc = 0.95)+
+    scale_color_discrete(name='')+
+    scale_x_discrete()+
+    geom_point(alpha=0)+
+    facet_grid(childteen~Methyl_factor, scales='free')+theme_bw()+
+    theme(legend.position = 'bottom')+
+  theme(axis.text.x = element_text(angle=45, hjust=1))
+
+ancestry_differences
+
+ggexport(ancestry_differences, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=11)
+
+ggexport(ancestry_differences, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.png')))
+```
+
+```{r}
+
+myCPGs<-base_models%>%bind_rows(.id='modeltype')%>%mutate(CpG=CPG, Gene=gene, estimate=logFC, conf.low=CI.L, conf.high=CI.R, source=paste0('Fragile Families:', modeltype), std.error='NA', p.value=P.Value, Chr=chr, set='Model1', Position=pos)%>%dplyr::select(colnames(cpgs))
+
+#cpg_overlap=myCPGs%>%filter(Joubertetal=='in Joubert et al. 2016' | Wicklundetal=='in Wicklund et al. 2019')%>%pull(CpG)%>%unique()
+
+#myCPGs2<-myCPGs %>%ungroup()%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)%>%mutate(CpG=ifelse(CpG %in% row.names(mygaps$proberesults), paste0(CpG, '*'), CpG))
+
+cpg_all<-rbind(myCPGs%>%filter(CpG %in% top_cpgs), cpgs%>%filter(CpG %in% top_cpgs), topCPGs%>%filter(CpG%in%top_cpgs))
+
+#geom blank to give good axes limits
+blank_data <- cpg_all %>%dplyr::select(Gene, CpG, estimate, std.error)%>%group_by(Gene)%>%mutate(min=min(estimate), max=max(as.numeric(std.error), na.rm=T))%>%dplyr::select(-estimate, -std.error)%>%unique()%>%mutate(estimate=min*-1+max, conf.high=estimate, conf.low=estimate)%>%dplyr::select(-max, -min)%>%arrange(Gene)%>%mutate(source='Wiklund et al.
+2019')
+
+blank_data2 <- cpg_all %>%dplyr::select(Gene, CpG, estimate, std.error)%>%group_by(Gene)%>%mutate(min=min(estimate))%>%dplyr::select(-estimate, -std.error)%>%unique()%>%mutate(estimate=min*-1+0.005, conf.high=estimate, conf.low=estimate)%>%dplyr::select(-min)%>%arrange(Gene)%>%mutate(source='Fragile Families: Age_9')
+
+plot_sharedCpG<-ggplot(cpg_all, aes(x=estimate, xmin=conf.low, xmax=conf.high, color=source, y=CpG))+geom_point(position=position_dodge(width=1))+geom_errorbarh(height=0.1, position=position_dodge(width=1))+scale_x_continuous()+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+facet_wrap(.~Gene, scales='free', ncol=2)+theme_classic()+theme(legend.position = 'bottom')+guides(color=guide_legend(nrow=4,byrow=TRUE))+geom_blank(data = blank_data, aes(x = estimate))+scale_color_discrete('', labels=c('Fragile families saliva: age 15', 'Fragile families saliva: age 9', 'Fragile families: mixed effect model', 'Joubert et al.
+2016 newborn cord blood', 'Joubert et al.
+2016 peripheral blood older children', 'Wicklund et al.
+peripheral blood >16 years'))
+
+ggexport(plot_sharedCpG, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')))
+
+plot_sharedCpG
+
+```

### 2021-10-26 Scripts/01_clean_phenotype_data.R

@@ -119,6 +119,10 @@ FF_labeled=FF_labeled %>%
 
 #ancestry
 #read in local and global pc data
+id_list=list.files(path=paste0(datadir, "OGData/pcs"), pattern='*idnums*', full.names=TRUE)
+id_list%>%map_dfr(read.table, .id='ancestry')
+
+
 pc_files=list.files(path=paste0(datadir, "OGData/pcs"), pattern='*.csv', full.names=TRUE)
 names(pc_files)=gsub('.csv', '', list.files(path=paste0(datadir, "OGData/pcs"), pattern='*.csv'))
 names(pc_files)=paste0(str_to_title(names(pc_files)), ' ancestry')

### 2021-10-26 Scripts/02_make_methyl_summary.R

@@ -77,7 +77,10 @@ polyscores_wide<-lapply(seq_along(polyscores), function(i) polyscores[[i]] %>% s
 if(nrow(betaqc)==423668){
   clocks<-read.csv(file=paste0(datadir, "OGData/ffcw_n1776_8clocks.csv"), header = T)
   colnames(clocks)[1]<-"MethID"
-}
+  grim_pk=read.csv(file=file.path(datadir, 'OGData', 'dnampackyrs_fromgrimoutput.csv'), header=T)
+  colnames(grim_pk)[1]<-'MethID'
+  clocks<-left_join(clocks, grim_pk)
+  }
 #the below code will run if the # CpGs in your chosen CpG matrix isn't 423668 (see note at top re: Jonah email) as the clocks
 #from ffcw_n1776_8clocks.csv were run using the 423668 matrix. if you switch to some other beta matrix, this code will recreate 
 # the clocks for you, except levine and GRIM clock which requires different code. 
@@ -126,7 +129,7 @@ methyl_labels_jonahclocks=c('Methylation data ID', 'Epithelial cell proportion',
                             'Global methylation', 'AHHR: cg05575921', 'MYO1G: cg04180046', 'CYP1A1: cg05549655', 'GFI1: cg14179389', "MYO1G: cg22132788", 
                             pms_labels, gsub('no transform', 'z-score standardized', pms_labels), gsub('no transform', 'mean-centered', pms_labels), 
                             'ID', 'Visit',
-                            'Horvath clock', 'SkinBlood clock', 'Hannum clock', 'Pediatric clock', 'Levine clock', 'PoAm clock 38', 'PoAm clock 45', 'GRIM clock')
+                            'Horvath clock', 'SkinBlood clock', 'Hannum clock', 'Pediatric clock', 'Levine clock', 'PoAm clock 38', 'PoAm clock 45', 'GRIM clock', 'GRIM pack/yrs component')
 
 if(nrow(betaqc)==423668){
   set_label(methyldata)=methyl_labels_jonahclocks

### 2021-10-28 Papers/Polymethylation_scores/02_abstract.Rmd
1st rough draft

# Abstract

**Background**: Prenatal maternal smoking is a difficult to measure prenatal exposure with negative implications for child health.  DNA methylation signatures vary by prenatal maternal smoke exposure in blood and placental samples and can function as a biomarker. However little work has assessed how DNA methylation signatures of prenatal maternal smoking vary across ages, ancestry groups or tissues. Using children in the Fragile Families and Child Wellbeing study, we tested if prenatal maternal smoking associates with salivary DNA methylation summary measures, including global methylation, epigenetic age, and polymethylation scores. We assessed the consistency of associations at ages 9 and 15 and their portability across  African, European and Hispanic ancestries.

**Results**: We created a polymethylation score using coefficients from a metanalysis of prenatal maternal smoke exposure and DNA methylation in newborn cordblood. Prenatal maternal smoking was associated with this polymethylation score () but not epigenetic age or global methylation (). The association with the polymethylation score was consistent at age 9 () and age 15 ().  The direction and magnitude of the association was portable across ancestries at age 9 (African: European: Hispanic), although the association was attenuated among European children at age 15 (African: European: Hispanic). The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (Area under the curve from receiver operating curve (AUC-ROC age 9: age 15: ), and more accurate than using the top CpG site identified in the metaanalysis, AHRR:CpG05575921 (AUC-ROC age 9:, age 15:, P value).

**Conclusions**: Prenatal maternal smoking associates with DNA methylation signatures in saliva samples, a clinically practical tissue. Polymethylation scores for prenatal maternal smoking are portable across ancestries and more accurate than some individual CpG sites.  Thus, DNA polymethylation scores from saliva samples could serve as robust and practical clinical biomarkers of prenatal maternal smoke exposure.

### 2021-10-28 Papers/Polymethylation_scores/04_background.Rmd
1st rough draft

@@
 
-Maternal prenatal smoking is a public health concern: in 2018, 11% of pregnant US women aged 15-44 reported any past-month cigarette use [@SAMHSA2018], and exposed children are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu].
+Maternal prenatal smoking is a public health concern: in 2018, 11% of pregnant US women aged 15-44 reported any past-month cigarette use [@SAMHSA2018].
+Exposed children are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu].
 Measuring prenatal maternal smoking behavior is challenging.
@@
 Further, serum cotinine measures during pregnancy are rarely available outside of birth cohorts, limiting clinical application as a marker of prenatal exposures.
-Developing a portable and reliable biomarker of prenatal maternal smoke exposure would therefore have important implications for research and clinical practice.
+Developing a portable and reliable biomarker of prenatal maternal smoke exposure would have important implications for research and clinical practice.
 
-DNA methylation has been associated with prenatal maternal smoking at various cross-sectional time points across the life course in primarily European ancestry groups.
-Cordblood, peripheral blood and placental DNA methylation has been significantly associated with prenatal maternal smoking [@Joubert:2016ue; @Everson:2021td; @Richmond:2018tu; @Cardenas2019].
-Blood DNA methylation signatures in cohorts of older children and adults also were significantly associated with prenatal maternal smoking exposure [@Tehranifar:2018tb ; @Richmond:2018tu].
+Differences in DNA methylation signatures of cordblood and placental tissues by prenatal maternal smoke exposure exist in primarily European cohorts of newborns [@Joubert:2016ue; @Everson:2021td; @Cardenas2019].
+Blood DNA methylation signatures in cohorts of older children and adults also differ by prenatal maternal smoking exposure [@Tehranifar:2018tb ; @Richmond:2018tu].
 Yet few studies have examined associations at different ages in the same participants.
 Additionally, while associations are well-replicated in cohorts of European ancestry, only a few, small sample size studies are in non-European ancestry populations.
-A 2016 study with 186 participants of African or admixed ancestry [@Ladd-Acosta:2016aa], while a 2018 study of 89 participants had 60% of African or admized ancestry [@Tehranifar:2018tb].
+One 2016 study included 186 participants of African or admixed ancestry [@Ladd-Acosta:2016aa], while a 2018 study of 89 participants had 60% of African or admized ancestry [@Tehranifar:2018tb].
 A larger 2019 study of 572 Latino children replicated associations with prenatal maternal smoking at 7 out of 148 tested *a priori* CpG sites from Joubert et al.
@@
 Additionally, previous work has relied on blood and placental samples.
-DNA methylation varies across tissues, an epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples[@Everson:2021td].
+DNA methylation varies across cell-types and tissues, and epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples[@Everson:2021td].
 Cord blood or placenta samples are unlikely to be available in a clinical setting.

### 2021-10-28 Papers/Polymethylation_scores/05_results.Rmd
1st rough draft

---
output:
  word_document: default
  html_document: default
bibliography: Methylation_SmokingBib.bib
always_allow_html: true
---

```{r, echo=F, include=FALSE}
library(knitr)
knit_child(here::here("Reports", "Results.Rmd"))
```

```{r, echo=F, include=FALSE}
coefCI=function(df, filter_statement, my_term, si){
  my_df=df%>%
    filter(eval(parse(text = filter_statement)))%>%
    unnest(tidied)%>%
    mutate(val=paste0(round(estimate, si), ' (95%CI: ', 
                      round(conf.low, si), ', ',
                      round(conf.high, si), ')'))%>%
    filter(term==my_term)%>%
    pull(val)
  return(my_df)
}
```

# Results

## Study sample descriptive statistics

Of `r methylCohort$MethID%>%length()` samples from `r methylCohort$idnum%>%unique()%>%length()` unique individuals with DNA methylation data measured on the 450K Illumina array, `r m_obs` samples from `r n_total` individuals had data on additional key covariates (Figure 1). `r n_twovisit` individuals had both age 9 and age 15 samples. Excluded samples were similar to included samples, except that included samples were slightly more likely to be from children of African ancestry and less likely to be from children of Hispanic ancestry (Supplemental Table 1). Twenty percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alchol use and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use. The mean income to poverty ratio of mothers at birth was `r round(mean(modeldata_wide$cm1inpov), 2)`. Of the children in the analytic sample, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Hispanic ancestry']`% were of Hispanic ancestry.

We calculated several common summary measures of DNA methylation, including global methylation, epigenetic clocks (Pediatric and GRIM), and polymethylation scores. We calculated polymethylation scores using regression coefficients for 6074 CpGs made from three different models in Joubert et al.: a regression using newborn cordblood samples controlling for cell-type proportions, a regression using newborn cordblood samples not controlling for cell-type proportions and a regression using peripheral blood samples from older children. We also examined the top probe from Joubert et al., AHRR:cg05575921. Many of the methylation summary measures were correlated with each other (Supplemental Figure 1). Notably, polymethylation scores calculated using coefficients from a cell-type corrected regression as weights did not correlate with estimated cell-type proportions, while scores calculated using coefficients from a regression not corrected for cell-type did. Among the `r n_twovisit` individuals with data from both visits, methylation summary measures from the same individual were correlated between the age 9 and age 15 (Supplemental Figure 2). The correlation was strongest for the polymethylation score calculated using weights from a regression for sustained smoking in newborn cordblood with cell-type control (Pearson $\rho$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value<0.0001). The distribution of epigenetic ages shifted between the age 9 and age 15 visit, as expected (Supplemental Figure 3; mean pediatric clock age 9: `r round(mean(modeldata_wide$pediatric_Age_9, na.rm=T))` age 15: `r round(mean(modeldata_wide$pediatric_Age_15, na.rm=T))`, mean GRIM clock age 9 `r round(mean(modeldata_wide$grim_Age_9, na.rm=T))`, age 15: `r round(mean(modeldata_wide$grim_Age_15, na.rm=T))`. The distribution of estimated cell-type proportions also shifted between visits while the distribution of global methylation and polymethylation scores were more consistent (Supplemental Figure 3). Based on these results, we focused on the polymethylation score from newborn cordblood with cell-type control, the pediatric clock, AHRR:cg05575921 and global methylation for bivariate and multivariable analyses.

## Bivariate associations between prenatal maternal smoking and DNA methylation summary measures

Mothers who reported smoking during pregnancy had lower income to poverty ratios and were more likely to report prenatal alcohol use, prenatal drug use and postnatal smoking than mothers who did not report prenatal maternal smoking (Table 1).  Children of mothers who reported smoking during pregnancy were more likely to be of European ancestry and had higher prenatal maternal smoking polymethylation scores at both the age 9 and age 15 visit (Table 1). Children of mothers who reported prenatal maternal smoking did not differ from those whose mothers did not in terms of global methylation, epigenetic age estimaated from pediatric clock or estimate cell-type proportions. Compared to children of mothers not reporting prenatal materna smoking, exposed children were significantly more hypomethylated at AHRR:cg05575921 at age 9 (mean methylation `r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`) but not age 15 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)` *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).

## Multivariable models

The association between prenatal maternal smoking and the cell-type controlled polymethylation score was also observed in multivariable models controlling for precision and confounder variables (Figure 3; Supplemental Table 2). In the entire sample, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` higher polymethylation score at age 9 and `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` higher polymethylation score at age 15. A consistent association was observed when stratifying to only African ancestry children (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`). The direction and magnitude of the association was similar in European and Hispanic ancestry children at age 9, though attenuated at age 15 (Figure 3; Supplemental Tables 2). Prenatal maternal smoking was associated with hypomethylation at AHRR:cg055975921 at age 9 but not age 15 and not associated with global methylation or epigenetic age from the pediatric clock at either age. Results were similar in linear mixed effect models and in sensitivity models controlling for surrogate variables instead of known covariates (Supplemental Tables 3-4; Supplemental Figure 4-5).

## Accuracy of polymethylation scores as a biomarker of prenatal maternal smoking

Next, we compared the accuracy of different DNA methylation summary measures for classifying prenatal maternal smoking using receiver operating curves (Figure 4A). The polymethylation score with cell-type control had the largest area under the curve (AUC) both when used as a single predictor (age 9 AUC; `r rct1%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 9')%>%pull(auc)`, age 15 AUC: `r rct1%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 15')%>%pull(auc)` ) and when used in models including child sex, maternal income-poverty ratio, immune cell proportion and batch (age 9 AUC: `r rct2%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 9')%>%pull(auc)`, age 15 AUC: `r rct2%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 15')%>%pull(auc)`). The AUC for the polymethylation score was significantly larger than using only the top CpG from Joubert et al., AHRR:cg05575921 (age 9 AUC: `r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 9')%>%pull(auc)` *P* value=`r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 9')%>%pull(4)%>%round(3)` , age 15 AUC: `r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 15')%>%pull(auc)` *P* value=`r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 15')%>%pull(4)%>%round(3)`). Accuracy was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B).

### 2021-10-28 Papers/Polymethylation_scores/06_discussion.Rmd
1st rough draft

@@
 However, the longitudinal design with repeated measures, exposure-outcome temporality and large sample size are strengths of our analysis.
-
-In total, our analysis contributes to the clinical implementation of a DNA methylation-based biomarker of prenatal maternal smoking.
-We showed that DNA methylation in saliva, a clinically practical tissue, had strong associations with and reasonable accuracy for prenatal maternal smoke exposure.
-Further, we demonstrated that polymethylation scores could be applied as a biomarker of prenatal maternal smoke exposure across ancestry-groups, an important consideration for health equity.
-The development and application of biomarker for prenatal maternal smoke exposure has important implications for epidemiological research and clinical practice.
-Given the difficulty of measuring prenatal maternal smoke exposure, such a biomarker could allow for confounder control in research areas where such control is currently impossible.
-Given the prevalence and negative health consequences of prenatal maternal smoke exposure, such a biomarker could be used to identify exposed individuals who might especially benefit from certain interventions.

### 2021-10-28 Papers/Polymethylation_scores/07_conclusions.Rmd
1st rough draft

# Conclusions

Our analysis contributes to the clinical implementation of a DNA methylation-based biomarker of prenatal maternal smoking. We showed that DNA methylation in saliva, a clinically practical tissue, had strong associations with and reasonable accuracy for prenatal maternal smoke exposure. Further, we demonstrated that polymethylation scores could be applied as a biomarker of prenatal maternal smoke exposure across ancestry-groups, an important consideration for health equity. The development and application of biomarker for prenatal maternal smoke exposure has important implications for epidemiological research and clinical practice. Given the difficulty of measuring prenatal maternal smoke exposure, such a biomarker could allow for confounder control in research areas where such control is currently impossible. Given the prevalence and negative health consequences of prenatal maternal smoke exposure, such a biomarker could be used to identify exposed individuals who might especially benefit from certain interventions.

### 2021-10-28 Papers/Polymethylation_scores/index_paper.Rmd
1st rough draft

@@
 csl: springer-vancouver.csl
+always_allow_html: true
 ---

### 2021-10-28 Scripts/01_clean_phenotype_data.R
added new people from most recent genetic data pull

@@ -123,11 +123,17 @@ id_list=list.files(path=paste0(datadir, "OGData/pcs"), pattern='*idnums*', full.
 id_list%>%map_dfr(read.table, .id='ancestry')
 
 
-pc_files=list.files(path=paste0(datadir, "OGData/pcs"), pattern='*.csv', full.names=TRUE)
-names(pc_files)=gsub('.csv', '', list.files(path=paste0(datadir, "OGData/pcs"), pattern='*.csv'))
-names(pc_files)=paste0(str_to_title(names(pc_files)), ' ancestry')
+pc_files=list.files(path=paste0(datadir, "OGData/pcs"), pattern='FFCWDNA_.*.csv', full.names=TRUE)
+names(pc_files)=gsub('.csv', '', list.files(path=paste0(datadir, "OGData/pcs"), pattern='FFCWDNA_.*.csv'))
+names(pc_files)=paste0(str_to_title(str_match(names(pc_files), '.*_(.*)_n.*')[,2]), ' ancestry')
 localpc=pc_files%>%map_dfr(read.csv, .id='ancestry')%>%rename_with(~str_c('local_', .), contains('PC'))
-allPC<-read.table(paste0(datadir, "OGData/pcs/global_pcs.txt"), col.names=c("X", "idnum", paste("global_PC", 1:20, sep='')))
+allPC<-read.table(paste0(datadir, "OGData/pcs/pcs_n50.eigenvec"), col.names=c("idnum", 'idnum2', paste("global_PC", 1:50, sep='')))
+
+old_pc=list.files(path=paste0(datadir, "OGData/pcs"), pattern='.*[n|c].csv', full.names=TRUE)
+names(old_pc)=gsub('.csv', '', list.files(path=paste0(datadir, "OGData/pcs"), pattern='FFCWDNA_.*.csv'))
+names(old_pc)=paste0(str_to_title(str_match(names(old_pc), '.*_(.*)_n.*')[,2]), ' ancestry')
+local_old=old_pc%>%map_dfr(read.csv, .id='ancestry')%>%rename_with(~str_c('local_', .), contains('PC'))
+checkPC=left_join(localpc, local_old%>%mutate(old_ancestry=ancestry)%>%dplyr::select(idnum, old_ancestry))
 
 #individuals in each ancestry specific pc group can be labeled as that ancestry for categorical ancestry variable 
[…]

### 2021-10-28 Scripts/04_summary_methyl_models.R

@@ -62,7 +62,7 @@ outcome_labels=c('Global methylation', 'Pediatric clock',
                  'Any smoking polymethylation score (newborns)', 
                  'Sustained smoking polymethylation score (newborns, cell-type controlled)', 
                  'Sustained smoking polymethylation score (older children)', 
-                 'AHHR: cg05575921')
+                 'AHRR: cg05575921')
 names(outcome_labels)=y_vector
 
 
@@ -239,7 +239,7 @@ modeldata_wide = modeldata%>%
   dplyr::select(any_of(c('idnum', 'ancestry', 'childteen', 'global_PC1', 
                          'global_PC2', 'local_PC1', 'local_PC2', 
                          'smkPreg_binaryN',  y_vector, 
-                         str_split(secondhand_model_vars,  '\\+')[[1]][-1])))%>%
+                         str_split(str_remove(secondhand_model_vars, 'ChildAgeComposite'),  '\\+')[[1]][-1])))%>%
   pivot_wider(names_from = childteen, values_from=any_of(c(y_vector, 'Leukocytes_saliva', 'Epithelial.cells_saliva', 'Sample_Plate', 'SmkAtVisitPastmonth')))
 
 #arguments

### 2021-10-28 Reports/Results.Rmd
rewriting results to include/exclude ewas results

@@
 library(ggrepel)
+
+include_ewas=F
+stand_alone=T
 ```
@@
 
-all_methyl=c("globalmethylation", "pediatric", "grim", "Epithelial.cells_saliva", "Leukocytes_saliva", "SSnewborn_center", "SSnewbornCT_center", "SSolder_center", "cg05575921")
+all_methyl=c("globalmethylation", "pediatric", "grim", 'DNAmPACKYRS', "Epithelial.cells_saliva", "Leukocytes_saliva", "SSnewborn_center", "SSnewbornCT_center", "SSolder_center", "cg05575921")
 main_methyl=c("globalmethylation", "pediatric", "SSnewbornCT_center", "cg05575921")
@@
 all_methyl_levels=c("SSnewbornCT_center", "SSnewborn_center",  "SSolder_center",
-                                       "pediatric" , "grim", "cg05575921", 
+                                       "pediatric" , "grim", 'DNAmPACKYRS',  "cg05575921", 
                                       "globalmethylation","Epithelial.cells_saliva", "Leukocytes_saliva")
@@
 maternaldata=myFF %>% filter(!is.na(smkPreg_binary) & !is.na(m1g2_YesNoPreg) & !is.na(m1g3_YesNoPreg) & !is.na(cm1inpov) & PostnatalMaternalSmokingAny!='Missing' & SmkAtVisitPastmonth!='Missing')
+
 n_withmaternaldata=paste0('m=', maternaldata %>% nrow(), ' samples & n=', maternaldata %>% dplyr::select(idnum)%>%unique()%>%nrow(), ' individuals with maternal covariate data')
@@
                            " individuals with a 9 year visit")
-n_ChildSmkexclusions<-paste0('n=', childdata%>%filter(k5f1l!= "2 no" & childteen=='C')%>%nrow(), 
+n_ChildSmkexclusions<-paste0('n=', childdata %>% filter((k5f1l=="1 yes" | is.na(k5f1l)) & childteen=='C') %>% nrow(), 
                              " individuals smoking \n or missing child smoking data \n at 9 year visit")
@@
                           " individuals with a 15 year visit")
-n_TeenSmkexclusions<-paste0('n=', childdata %>% filter(k6d40!="2 No" & childteen=='T') %>% nrow,
+n_TeenSmkexclusions<-paste0('n=', childdata %>% filter((k6d40=="1 Yes" | is.na(k6d40) | k5f1l=='1 yes') & childteen=='T') %>% nrow,
                             " individuals smoking \n or missing focal child smoking data \n at 15 year visit")
-if(childdata %>% filter((childteen=='C' & k5f1l=='2 no') | (childteen=='T' & k6d40=='2 No'))%>%nrow()!=nrow(modeldata)){stop('filter N and completecase data set N do not match')}
+
+if(childdata %>% filter((childteen=='C' & k5f1l=='2 no') | (childteen=='T' & k6d40=='2 No' & k5f1l!='1 yes'))%>%nrow()!=nrow(modeldata)){stop('filter N and completecase data set N do not match')}
 
@@
 
-graph1
-
 graph1 %>%
@@
 main_fig=main_fig+1            
+```
+```{r, eval=stand_alone}
+graph1
 ```
@@
 
+sup_tab=sup_tab+1
+```
+
+```{r, eval=stand_alone}
 exclusion_tbl
-
-sup_tab=sup_tab+1
+```
+
+```{r wide data}
+modeldata_wide = modeldata%>%
+  mutate(childteen=gsub(' ', '_', childteen))%>%
+  dplyr::select(any_of(c('idnum', 'ancestry', 'childteen', 'cm1bsex', 'ancestry', 
+                         'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'cm1inpov',
+                         'smkPreg_binary',  all_methyl)))%>%
+  pivot_wider(names_from = childteen, values_from=any_of(all_methyl))
+
+model_twovisit=modeldata_wide[complete.cases(modeldata_wide), ]
 ```
@@
 
-```{r, fig.width=12, fig.height=12}
+```{r}
 library(Hmisc)
@@
 rownames(res2)=all_methyl_labels
-corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8)
 
@@
 sup_fig=sup_fig+1
+```
+
+```{r, eval=stand_alone, fig.width=12, fig.height=12}
+corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8)
+
 ```
@@
   scale_x_continuous('Age 9 score')+scale_y_continuous('Age 15 score')+
-  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels))
-
-visit_cor
+  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels), ncol=3)
 
@@
 
+```{r, eval=stand_alone, fig.width=12, fig.height=12}
+visit_cor
+```
+
 # Supplemental Figure `r sup_fig`: Distributions of selected methylation summary measures at age 9 and age 15 among `r n_total` children in the Fragile Families and Child Wellbeing study
 
-```{r, fig.height=11}
+```{r}
 dist_methyl=age_long_methyl%>%
@@
   geom_point(alpha=0)+geom_line(alpha=0, aes(group=idnum))+geom_smooth(aes(group=1), method='lm', color='black')+
-  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels))+
+  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels), ncol=3)+
   theme(legend.position = 'bottom')
-
-dist_methyl
 
@@
 
+```
+
+```{r, eval=stand_alone, fig.height=11}
+dist_methyl
 ```
@@
 
-fig2
-
 ggexport(fig2, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')))
@@
 
+```
+
+```{r, eval=stand_alone}
+fig2
 ```
@@
   dplyr::select(smkPreg_binary, m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,
-         SmkAtVisitPastmonth, ancestry, cm1bsex, globalmethylation, pediatric, Epithelial.cells_saliva,
+         SmkAtVisitPastmonth, ancestry, cm1bsex, globalmethylation, pediatric, grim, DNAmPACKYRS, Epithelial.cells_saliva,
          Leukocytes_saliva, SSnewbornCT_center, SSolder_center, cg05575921, childteen)%>%
@@
 
+main_tab=main_tab+1
+```
+
+```{r, eval=stand_alone}
 smoking_tbl
-
-main_tab=main_tab+1
 ```
@@
 
-```{r, fig.width=12}
+```{r}
 my_outcomes=c('Global methylation', 'Pediatric clock', 'AHRR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)')
@@
 
-forest_plot
-
 ggexport(forest_plot, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')), width=15)
@@
 
+```{r, eval=stand_alone, fig.width=12}
+forest_plot
+```
+
 # Supplemental Figure `r sup_fig`: Sensitivity analyses for controlling for different variable sets in multivariable analyses of prenatal maternal smoking and DNA methylation summary measures
 
-```{r, fig.width=14}
+```{r}
 with(modeldata, table(ancestry, PostnatalMaternalSmokingAny, smkPreg_binary))%>%kbl()
@@
 
-forest_plot_compareModels
-
 ggexport(forest_plot_compareModels, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=15, height=15)
@@
 
+```{r, eval=stand_alone, fig.width=14}
+forest_plot_compareModels
+```
+
 # Supplemental Figure `r sup_fig` DNA methylation summary measures across prenatal and postnatal exposures
 
-```{r, fig.width=14, fig.height=10}
+```{r}
 
@@
 
-modeldata%>%
+dna_prepost=modeldata%>%
   dplyr::select(any_of(c('idnum', 'childteen', 'ancestry', 'smkPreg_binary', 'PostnatalMaternalSmokingAny', all_methyl)))%>%
@@
   guides(color=guide_legend(nrow=4,byrow=TRUE))
+
+ggexport(dna_prepost, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=10, height=10)
 sup_fig=sup_fig+1
+```
+
+```{r, eval=stand_alone, fig.width=14, fig.height=10}
+dna_prepost
 ```
@@
 
+sup_tab=sup_tab+1
+```
+
+```{r, eval=stand_alone}
 regression_table_age
-
-sup_tab=sup_tab+1
 ```
@@
 
+sup_tab=sup_tab+1
+
+```
+
+```{r, eval=stand_alone}
 regression_table_ancestry
-
-sup_tab=sup_tab+1
-
 ```
@@
 
-```{r, fig.width=12, fig.height=8}
+```{r}
 #bring the heatmap function
@@
 
-bring.the.heatmap.sv(modeldata, covar=c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'ancestry', 'cm1bsex', 'Leukocytes_saliva', 'Epithelial.cells_saliva', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))
-
 pdf(file=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=11, height=13)
@@
 sup_fig=sup_fig+1
+```
+
+```{r, eval=stand_alone, fig.width=12, fig.height=8}
+bring.the.heatmap.sv(modeldata, covar=c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'ancestry', 'cm1bsex', 'Leukocytes_saliva', 'Epithelial.cells_saliva', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))
 ```
@@
 
+sup_tab=sup_tab+1
+```
+
+```{r, eval=stand_alone}
 regression_table_ancestry_svs
-
-sup_tab=sup_tab+1
 ```
@@
 
-```{r, fig.width=10}
+```{r}
 
@@
 
-auc_fig
-
 ggexport(auc_fig, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')), width=9)
 main_fig=main_fig+1
+```
+
+```{r, eval=stand_alone, fig.width=10}
+auc_fig
 ```
@@
 
-roc_table1%>%mutate(`pval comparison to sustained smoking score (newborns)`=map_dbl(roctest, ~round(.x$p.value, 4)))%>%dplyr::select(childteen, modeltype, auc, `pval comparison to sustained smoking score (newborns)`)%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value (vs polymethylation score, newborns, cell type controlled)'))
+rct1=roc_table1%>%mutate(`pval comparison to sustained smoking score (newborns)`=map_dbl(roctest, ~round(.x$p.value, 4)))%>%dplyr::select(childteen, modeltype, auc, `pval comparison to sustained smoking score (newborns)`)
+
+```
+
+```{r, eval=stand_alone}
+rct1%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value (vs polymethylation score, newborns, cell type controlled)'))
 ```
@@
 
-roc_table2%>%mutate(`pval comparison to base model`=map_dbl(roctest, ~round(.x$p.value, 4)))%>%dplyr::select(childteen, modeltype, auc, `pval comparison to base model`)%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value (vs base w/ no methyl)'))
-
+rct2=roc_table2%>%mutate(`pval comparison to base model`=map_dbl(roctest, ~round(.x$p.value, 4)))%>%dplyr::select(childteen, modeltype, auc, `pval comparison to base model`)
+
+rct2%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value (vs base w/ no methyl)'))
+
+```
+
+```{r, eval=stand_alone}
+rct2%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value (vs base w/ no methyl)'))
 ```
@@
 
-```{r}
+```{r, eval=include_ewas}
 
@@
 
-```{r}
+```{r, eval=include_ewas}
 child_m1=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'child1_TopHits.RDS'))
@@
 
-```{r}
+```{r, include=include_ewas}
 #lme_base=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'lme_ewas_base.RDS'))
@@
 
+```
+
+```{r, eval=stand_alone, eval=include_ewas}
 par(mfrow=c(3, 1), mar=c(5.1, 4.1, 5.1, 2.1))
@@
 sup_fig=sup_fig+1
-
 ```
@@
 
-```{r, cache=T}
+```{r, cache=T, eval=include_ewas}
 library(tibble)
@@
 
+```{r, eval=stand_alone, eval=include_ewas}
+beta_compare
+```
+
 ## Upset plots
 
-```{r}
+```{r, eval=include_ewas}
 library(ggupset)
@@
 
-```{r, fig.width=14}
+```{r, eval=include_ewas}
 volcs=ggarrange(
@@
 ggexport(volcs, filename=file.path(outdir, paste0('Figure', main_fig, '.png')))
-
+```
+
+```{r, eval=stand_alone, fig.width=14, }
 volcs
@@
 
-```{r}
+```{r, eval=include_ewas}
 all_gaps=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'gaphunter_gaps.RDS'))
@@
 
-```{r, fig.width=14}
+```{r, eval=include_ewas}
 topCPGs<-read.csv(paste0(datadir, "/TopCpGsWiklundEtAl2019.csv"))
@@
 
-```{r, fig.width=14}
+```{r, fig.width=14, eval=include_ewas}
 comparison=left_join(
@@
 
-comparison %>% mutate(s1=sign(FF_estimate), s2=sign(Joubert_estimate), agreement=case_when(s1==s2~'Direction consistent', s1<s2~'FF -, Joubert +', s1>s2~'FF+, Joubert-'))%>%group_by(agreement)%>%summarise(n=n())
+#comparison %>% mutate(s1=sign(FF_estimate), s2=sign(Joubert_estimate), agreement=case_when(s1==s2~'Direction consistent', s1<s2~'FF -, Joubert +', s1>s2~'FF+, Joubert-'))%>%group_by(agreement)%>%summarise(n=n())
 
@@
 ggexport(comp_joubert, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.png')))
-
+```
+
+```{r, eval=stand_alone, fig.width=14}
 comp_joubert
@@
 
-```{r}
+```{r, eval=include_ewas}
 top_cpgs = map(base_models, ~.x %>%filter(adj.P.Val<0.05)%>%dplyr::select(CPG, ModelType, gene))%>%bind_rows()%>%mutate(label=paste0(gene, ': ', CPG))
@@
 
-```{r, fig.width=14}
+```{r, fig.width=14, eval=include_ewas}
 
@@
 
-```{r}
+```{r, eval=stand_alone, fig.width=14}
+ancestry_differences
+```
+
+```{r, eval=include_ewas}
 
@@
 ggexport(plot_sharedCpG, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')))
-
+```
+
+```{r, eval=stand_alone, fig.width=14}
 plot_sharedCpG
-
-```
+```

### 2021-11-01 Papers/Polymethylation_scores/index_paper.Rmd
updates to index paper

@@
 
-```{r child = here::here("Papers", "Polymethylation_scores", "01_titlepage.Rmd")}
+```{r, child = here::here("Papers", "Polymethylation_scores", "01_titlepage.Rmd")}
 ```
 
-```{r child = here::here("Papers", "Polymethylation_scores", "02_abstract.Rmd")}
+```{r, child = here::here("Papers", "Polymethylation_scores", "02_abstract.Rmd")}
 ```
 
-```{r child = here::here("Papers", "Polymethylation_scores", "03_keywords.Rmd")}
+```{r, child = here::here("Papers", "Polymethylation_scores", "03_keywords.Rmd")}
 ```

### 2021-11-01 Reports/Results.Rmd
updates to index paper

@@
 
-```{r, eval=stand_alone, fig.width=14, }
+```{r, eval=stand_alone, eval=include_ewas, fig.width=14, }
 volcs
@@
 
-```{r, eval=stand_alone, fig.width=14}
+```{r, eval=stand_alone, eval=include_ewas, fig.width=14}
 comp_joubert
@@
 
-```{r, eval=stand_alone, fig.width=14}
+```{r, eval=stand_alone, eval=include_ewas, fig.width=14}
 ancestry_differences
@@
 
-```{r, eval=stand_alone, fig.width=14}
+```{r, eval=stand_alone, eval=include_ewas, fig.width=14}
 plot_sharedCpG

### 2021-11-02 Papers/Polymethylation_scores/01_titlepage.Rmd
responses to kelly edits november2

# Title Page

**Title:** Polymethylation Scores for Prenatal Maternal Smoke Exposure Persist Until Age 15 and Are Detected in Saliva

**Authors:** Freida A. Blostein^1, Jonah Fisher^2, John Dou^1, Lisa Schenper^3, Erin Ware^2 Daniel A. Notterman^3, Colter Mitchell^2, Kelly M. Bakulski^1

**Affiliations:**

^1 Department of Epidemiology, University of Michigan School of Public Health, Ann Arbor, Michigan, United States of America

^2 Institute for Social Research, University of Michigan, Ann Arbor, Michigan, United States of America

^3 Department of Molecular Biology, Princeton University, Princeton, New Jersey, United States of America

### 2021-11-02 Papers/Polymethylation_scores/04_background.Rmd
responses to kelly edits november2

@@
 
-Maternal prenatal smoking is a public health concern: in 2018, 11% of pregnant US women aged 15-44 reported any past-month cigarette use [@SAMHSA2018].
-Exposed children are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu].
+Maternal prenatal smoking is a public health concern.
+In 2018, 11% of pregnant women in the United States aged 15-44 reported any past-month cigarette use [@SAMHSA2018].
+Children exposed to cigarette smoking in utero are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu].
 Measuring prenatal maternal smoking behavior is challenging.
 Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua].
-The gold-standard for smoking measurements, serum cotinine, has a short half-life of nine hours in pregnant women [@Dempsey:2002wr].
+The gold-standard for smoking measurements, serum cotinine levels, have a short half-life of nine hours in pregnant women [@Dempsey:2002wr].
 Further, serum cotinine measures during pregnancy are rarely available outside of birth cohorts, limiting clinical application as a marker of prenatal exposures.
@@
 
-Differences in DNA methylation signatures of cordblood and placental tissues by prenatal maternal smoke exposure exist in primarily European cohorts of newborns [@Joubert:2016ue; @Everson:2021td; @Cardenas2019].
-Blood DNA methylation signatures in cohorts of older children and adults also differ by prenatal maternal smoking exposure [@Tehranifar:2018tb ; @Richmond:2018tu].
-Yet few studies have examined associations at different ages in the same participants.
-Additionally, while associations are well-replicated in cohorts of European ancestry, only a few, small sample size studies are in non-European ancestry populations.
-One 2016 study included 186 participants of African or admixed ancestry [@Ladd-Acosta:2016aa], while a 2018 study of 89 participants had 60% of African or admized ancestry [@Tehranifar:2018tb].
-A larger 2019 study of 572 Latino children replicated associations with prenatal maternal smoking at 7 out of 148 tested *a priori* CpG sites from Joubert et al.
-at a FDR-adjusted *P*-value\<0.05 [@Neophytou:2019ty; @Joubert:2016ue].
-Thus, while DNA methylation is associated with prenatal maternal smoking, the consistency of this association across age and ancestry is understudied.
+Maternal cigarette smoking during pregnancy is associated with reproducible DNA methylation signatures, measured in infant cord blood and placental tissues, and primarily tested in European cohorts [@Joubert:2016ue; @Everson:2021td; @Cardenas2019].
+Prenatal maternal cigarette smoking exposure is associated with postnatal DNA methylation signatures in blood from cohorts of older children and adults[@Tehranifar:2018tb ; @Richmond:2018tu].
+Few studies have considered the persistence of the association between maternal prenatal cigarette smoking and DNA methylation at different ages in the same individuals.
+Additionally, little work has considered the portability of the association between maternal prenatal cigarette smoking and DNA methylation across tissues and ancestry groups.
+DNA methylation drives cell differentiation and cell type proportions differ across tissue types.
+Epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples [@Everson:2021td].
+Cord blood and placental samples are unlikely to be available in clinical settings.
+While peripheral blood is a more realistic clinical samples, saliva is even easier to collect.
+To our knowledge, no study has evaluated associations between maternal prenatal cigarette smoking and salivary DNA methylation.
+Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017], and DNA methylation signatures of own-smoking differ across ancestry groups [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012].
+Only a few studies have examined associations between prenatal maternal cigarette smoking and DNA methylation in non-European ancestry populations.
+Among 954 infants, 70% of whom had mothers identifying as Black, prenatal maternal smoking was associated with cord blood DNA methylation at 38 CpG sites [@Xu2021].
+The direction of these association between prenatal maternal cigarette smoking and DNA methylation at these sites were generally similar in children of Black and non-Black participants [@Xu2021].
+Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking was associated with DNA methylation in peripheral blood samples but ancestry-stratified analyses were not performed [@Ladd-Acosta:2016aa].
+Similarly, among 89 middle-aged women, of whom reported African American or Hispanic ethnicity, prenatal maternal cigarette smoking was associated with DNA methylation at 17 of 190 tested CpG sites [@Tehranifar:2018tb].
+Adjusting for ancestry did not substantially affect the association between prenatal maternal smoking and DNA methylation at any of the sites, although the authors note that African American women had on average 2--5% higher mean DNA methylation on the absolute scale as compared with White and Hispanic women at 5 *CYP1A1* CpG sites [@Tehranifar:2018tb].
+Of 148 CpG sites selected based on their association prenatal maternal smoking among primarily European children, 7 CpG sites were associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue].
+Thus, while prenatal maternal cigarette smoking is associated with DNA methylation, the consistency of this association across age, tissue and ancestry is understudied.
 
-The association between prenatal maternal smoking and DNA methylation suggests DNA methylation could provide a reliable and specific biomarker of prenatal maternal smoke exposure.
-Importantly, changes in DNA methylation are specific to smoke exposure *in utero*.
-Previous work has found reasonable performance of cord-blood accuracy ranging from 81% to 91%) [@Reese:2017ts; @Ladd-Acosta:2016aa].
-A 2018 study found that DNA methylation based scores from adult peripheral blood could even predict predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu].
-However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across age and ancestry has not been evaluated.
+An important next step after association testing is the evaluation of DNA methylation as a biomarker of prenatal maternal smoke exposure, which is often done by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior.
+<!--#DEFINE POLYMETHYLATION SCORES HERE?-->DNA methylation signatures from cord blood classified maternal smoking behavior with accuracy of 91% in a primarily European cohort [@Reese:2017ts].
+Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking behavior could also be classified with an accuracy of 81% when using a support vector machine approach and DNA methylation at 26 sites measured in peripheral blood at a single time point [@Ladd-Acosta:2016aa].
+Among middle-aged adults, peripheral blood from a single time point could even predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu].
+However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across tissue, age and ancestry has not been evaluated.
 Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times or evaluated if the inclusion of multiple time points improves performance.
-Additionally, previous work has relied on blood and placental samples.
-DNA methylation varies across cell-types and tissues, and epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples[@Everson:2021td].
-Cord blood or placenta samples are unlikely to be available in a clinical setting.
-While peripheral blood is a more realistic clinical sample, saliva is even easier to collect.
-To our knowledge, no study has evaluated salivary DNA methylation as a biomarker for prenatal maternal smoking.
+No study has evaluated the accuracy of salivary DNA methylation as a biomarker for prenatal maternal smoking.
 
@@
 However, before translation to clinical applications, we must understand how and if the signal varies across age, ancestry and tissue sample.
-The goal of this study was to assess the *in vitro* potential of salivary DNA methylation biomarkers for prenatal maternal smoking.
-We tested associations between prenatal maternal smoke exposure and saliva DNA methylation biomarkers from the Illumina 450K chip, including epigenetic clocks, polymethylation scores, and individual *a priori* CpG sites using longitudinal samples from the Fragile Families and Child Wellbeing study.
-We performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a consistent biomarker of prenatal maternal smoking across populations and time.
+In the Fragile Families and Child Wellbeing study, a diverse longitudinal cohort of children, we aimed to assess the potential of salivary DNA methylation biomarkers for prenatal maternal cigarette smoking.
+We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including epigenetic clocks, polymethylation scores, and individual *a priori* CpG sites.
+We then performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a persistent and portable biomarker of prenatal maternal smoking.

### 2021-11-02 Papers/Polymethylation_scores/06_discussion.Rmd
responses to kelly edits november2

@@
 
-In the longitudinal Fragile Families and Child Wellbeing birth cohort, we found an association between prenatal maternal smoking and DNA methylation in sequential saliva samples from ages 9 and 15.
-Polymethylation scores were associated with prenatal maternal smoking across ages and ancestry while global methylation and epigenetic clocks were not, demonstrating a specific and persistent signal of prenatal smoke exposure in DNA methylation.
+In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed prenatal maternal smoking was associated with DNA methylation in children's saliva samples from ages 9 and 15.
+Prenatal maternal smoking was associated with polymethylation scores across strata of child age and ancestry.
+Global methylation and epigenetic clocks were not associated with maternal smoking exposure demonstrating a specific and persistent signal of prenatal smoke exposure in DNA methylation.
 Polymethylation scores had reasonable accuracy for classifying prenatal maternal smoking.
-This accuracy was more consistent across time than when using some individual *a priori* CpG sites.
+This accuracy was more consistent across time than when using individual *a priori* CpG sites.
 
-Our findings fit within the previous literature on associations between prenatal maternal smoking and DNA methylation.
-We replicate top hits from previous epigenome wide association studies, including a large metaanalysis of DNA methylation in blood samples [@Joubert:2016ue].
-Although there are no previous studies of prenatal maternal smoking and saliva methylation, our findings fit with findings on the consistency of DNA methylation signatures of own-smoking between saliva and blood samples [@Dawes2019; @Philibert2020].
+Our findings are consistent with the previous literature on associations between prenatal maternal smoking and DNA methylation.
+We replicated top hits from previous epigenome wide association studies, including a large meta-analysis of DNA methylation in newborn blood samples [@Joubert:2016ue].
+There are no previous studies of prenatal maternal smoking and saliva DNA methylation.
+However, previous work has that the majority of CpG sites were similarly methylated in blood and saliva (reviewed in @Langie2017).
+DNA methylation at AHRR:cg05575291 can accurately classify own--smoking behavior in both blood (area under the curve 0.995) and saliva (area under the curve 0.971) [@Dawes2019; @Philibert2020].
 
-We contribute to the literature on DNA methylation and prenatal maternal smoking by evaluating the consistency of this association in saliva and across ages and ancestries.
-Polymethylation scores built using coefficients from metaanalysis of blood DNA methylation still associated with prenatal maternal smoking in our independent population of saliva samples.
-Similarly, even though the source population for construction of the polymethylation scores was primarily of European ancestry, an association with prenatal maternal smoking was apparent in our primarily African ancestry population.
-Effect estimates were consistent across ancestries at age 9, although Hispanic and European ancestry age 15 samples had a lower effect estimate.
-This could reflect higher unreported smoking initiation in European and Hispanic ancestry US teens than African ancestry teens.
+We advance the literature on prenatal maternal smoking and DNA methylation by evaluating the persistence of the association as children age and its portability across tissue and ancestry.
+Polymethylation scores built using coefficients from meta-analysis of cord blood DNA methylation from primarily European-ancestry newborns were still associated with prenatal maternal smoking in our independent saliva samples from a diverse cohort at ages 9 and 15 [@Joubert:2016ue].
+The portability of other risk scores, such as polygenic risk scores, across ancestries is a complex research area [@Martin2019; ] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Hüls2020].
+In this case, the polymethylation score for prenatal maternal smoking appears to be portable across ancestry groups.
+Effect estimates were consistent across ancestries at age 9, although Hispanic and European genetic ancestry samples had lower effect estimates at age 15.
+This could reflect higher unreported smoking initiation in European and Hispanic ancestry teens in the United States than African ancestry teens.
 Genetic ancestry correlates with race, and White and Latino teens have much higher rates of teen smoking and earlier ages at initiation than Black children [@El-Toukhy2016].
-Own-smoking could influence DNA methylation at ages 9 and 15, creating outcome misclassification.
-This could also explain the observed attenuation of hypomethylation at AHRR:CpG05575921 by age 15 in our sample, as methylation at this probe is known to vary by own-smoking exposure as well[@Dawes2019; @Philibert2020].
+Though we excluded children who reported own-smoking from our analytic sample, under-reporting of own-smoking could influence DNA methylation at ages 9 and 15, creating outcome misclassification.
+This could also explain the observed attenuation the effect on AHRR:cg05575921 methylation by age 15 in our sample, as DNA methylation at this probe is known to vary by personal cigarette smoking[@Dawes2019; @Philibert2020].
 
-Our findings also fit with previous work on DNA methylation as a biomarker for prenatal maternal smoke exposure.
-Accuracy of the polymethylation scores in our sample (Age 9: 0.67 and Age 15: 0.66) was worse than the performance of biomarkers using cord blood (AUCs ranging from 0.82 to 0.97) [@Reese:2017ts; @Ladd-Acosta:2016aa].
-Differences in methylation patterns across tissues and over time could be responsible for this diminished accuracy.
-However, our accuracy was comparable to accuracy when using peripheral blood from older adults (AUC 0.72) [@Richmond:2018tu].
-The diminished accuracy from cord blood may be an acceptable trade-off considering saliva is a more readily available tissue for epidemiological and clinical applications.
+Our findings are also consistent with previous work on DNA methylation as a biomarker for prenatal maternal smoke exposure.
+Prenatal maternal smoking classification using saliva DNA methylation in our sample of 9 and 15 year olds () performed comparably to classification using peripheral blood from older adults () [@Richmond:2018tu].
+However, previous biomarkers of prenatal maternal smoking using DNA methylation of cord blood from newborns () and peripheral blood from 3-5 year old children () more accurately classified prenatal maternal smoke exposure [@Reese:2017ts; @Ladd-Acosta:2016aa].
+Differences in DNA methylation patterns across tissues and over time may influence the performance of DNA methylation biomarkers.
+The similarity in performance between saliva DNA methylation in our study and peripheral blood DNA methylation in older adults is encouraging for the use of saliva DNA methylation as a readily accessible clinical sample.
 
-Our analysis answers pertinent questions for the development of a clinically applicable methylation biomarker of prenatal maternal smoking.
-We show that DNA methylation scores calculated from children's saliva can accurately classify prenatal maternal smoke exposure in non-European ancestry children.
-Other bomarkers and risk scores are known to differ across ancestry groups, potentially contributing to and exacerbating health disparities.
-As new biomarkers are developed, understanding their consistency across ancestry groups is critical.
-We also demonstrated that accuracy is not substantially affected when using coefficients from newborn cordblood (no cell type control, age 9: 0.62, age 15: 0.61) vs older children peripheral blood (no cell type control age 9: 0.63, age 15: 0.60).
-However, using coefficients from models that incorporated cell-type control at the time of calculation did result in better accuracy, suggesting cell-type control should be instituted both when creating weights and when applying scores to new population.
-Including samples from both ages did not improve accuracy, thus repeated samplings may be unnecessary for clinical applications.
-Additionally, accuracy was more consistent when using polymethylation scores than using the single top-hit from Joubert et al, AHRR:CpG05575921 [@Joubert:2016ue].
-Salivary AHRR:CpG05575921 methylation was less accurate in our sample at predicting prenatal maternal smoke exposure than when it has been used to classify own smoking in blood or saliva [@Philibert2020].
-While using a single CpG site as a biomarker may be clinically pragamatic, single sites may be less robust to time-since-exposure and new environmental exposures.
-Alternatively, incorporating information from across several CpGs in a polymethylation score may be more robust to these .
+Our analysis contributes to the development of an accurate methylation biomarker for prenatal maternal smoking by evaluating the impact of specific methodological choices on accuracy of classification.
+DNA methylation biomarkers may be especially susceptible to confounding by subject age at methylation measurement, tissue of measurement, and cell-type proportion [@Hüls2020].
+Accuracy of polymethylation score classification was similar when using coefficients from cord blood vs peripheral blood samples from older children.
+Accuracy was not improved by using two time-points of methylation measurement.
+Accuracy was improved when polymethylation scores used coefficients which incorporated cell-type control.
+Our analysis therefore suggests that cell-type control when both generating and applying coefficients for polymethylation scores may positively influence accuracy, although more work comparing coefficients from different populations is still needed.
+
+Our analysis suggests that polymethylation scores may be more accurate than using single CpG sites as biomarkers.
+AHRR:cg05575921 is a consistent marker of prenatal maternal smoke exposure in meta-analyses of newborn cord blood[@Joubert:2016ue] and a very accurate DNA methylation marker of own smoking [@Philibert2020].
+However, salivary AHRR:cg05575921 was not a persistent marker of prenatal maternal smoking in an analysis of middle-aged adult women [@Richmond:2018tu].
+In our sample, salivary AHRR:CpG05575921 methylation was less accurate at predicting prenatal maternal smoke exposure than polymethylation scores.
+The accuracy of AHRR:CpG05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures.
+Incorporating information across multiple sites of DNA methylation may yield a biomarker more robust to these influences.
 
@@
 Additionally, while we controlled for postnatal secondhand smoke exposure and excluded children who reported any own-smoking, we cannot exclude the possibility of residual confounding in this observational cohort.
-However, the longitudinal design with repeated measures, exposure-outcome temporality and large sample size are strengths of our analysis.
+
+However, our analysis also has several strengths.
+We analyzed samples from a large sample of diverse participants underrepresented in genetic and epigenetic research [@Sirugo2019; @Bentley2017].
+While our exposure measurement of prenatal maternal smoking was self reported, it was assessed prospectively and preceded outcome measurements.
+We analyzed repeated measures of DNA methylation with reproducible array measures conducted in a single batch.
+We tested associations between prenatal maternal smoking and multiple DNA summary measures to evaluate the specificity of the polymethylation scores.
+In sensitivity analyses, we adjusted for other prenatal exposures and postnatal smoke exposure to examine the specificity of the biomarker to nature and timing of exposure.

### 2021-11-02 Papers/Polymethylation_scores/07_conclusions.Rmd
responses to kelly edits november2

@@
 
-Our analysis contributes to the clinical implementation of a DNA methylation-based biomarker of prenatal maternal smoking.
-We showed that DNA methylation in saliva, a clinically practical tissue, had strong associations with and reasonable accuracy for prenatal maternal smoke exposure.
-Further, we demonstrated that polymethylation scores could be applied as a biomarker of prenatal maternal smoke exposure across ancestry-groups, an important consideration for health equity.
+In a large, prospective study of diverse participants, we showed that DNA methylation in children's saliva had strong associations with and reasonable accuracy for prenatal maternal smoke exposure.
+Further, we demonstrated that polymethylation scores could be applied as a biomarker of prenatal maternal smoke exposure across ancestry-groups, an important consideration for the equitable biomarker development.
 The development and application of biomarker for prenatal maternal smoke exposure has important implications for epidemiological research and clinical practice.
 Given the difficulty of measuring prenatal maternal smoke exposure, such a biomarker could allow for confounder control in research areas where such control is currently impossible.
-Given the prevalence and negative health consequences of prenatal maternal smoke exposure, such a biomarker could be used to identify exposed individuals who might especially benefit from certain interventions.
+Prenatal maternal smoke exposure is prevalent and has negative health consequences, thus an exposure biomarker could be used to provide support and health interventions for children.

### 2021-11-02 Papers/Polymethylation_scores/08_methods.Rmd
1st rough draft
responses to kelly edits november2

--- 2021-09-14 Papers/Polymethylation_scores/08_methods.Rmd
@@
 ---
-output: html_document
-bibliography: "../Methylation_SmokingBib.bib"  
+output:
+  word_document: default
+  html_document: default
+bibliography: Methylation_SmokingBib.bib
 ---
 
-# Methods
+## Methods
 
-## Cohort
+### Cohort
 
-The Fragile Families and Child Wellbeing Study (FFCW) is a nationally-representative cohort of nearly 5,000 children born in 20 U.S.
-cities between 1998 and 2000 [@Reichman:2001wv].
-FFCW sampled births using a three-stage stratified random sample design, described in detail elsewhere [@Reichman:2001wv].
-FFCW excluded the following births from the study: those with parents who planned to place the child for adoption, those where the father was deceased, those who did not speak English or Spanish well enough to be interviewed, those where the mothers or babies were too ill to complete the interview and those where the baby died before the interview could take place.
-At some of the hospitals, parents who were under 18 were not eligible to be interviewed due to hospital prohibition.
-Children were followed from birth until age fifteen.
-At the time of birth, medical record information was extracted from the hospital at which the mother gave birth.
-Interviews were conducted with the mother at birth in person and with father in person or over the telephone if he was not present.
-Interviews were also conducted with the mother and father when the child was one, three, five, and nine via telephone.
-Additionally, at ages three, five and nine, a separate primary caregiver survey was administered to the individual with whom the child was living either over the telephone or in the home.
-The child was interviewed at age nine in the home and at age fifteen by telephone or in the home.
-At ages nine and fifteen a saliva sample was taken from the child for DNA and DNA methylation measurements [@Reichman:2001wv].
+The Fragile Families and Child Wellbeing Study is a birth cohort of nearly 5,000 children born in 20 cities in the United States between 1998 and 2000 [@Reichman:2001wv].
+Participants were selected at delivery using a three-stage stratified random sample design which oversampled unmarried mothers by a ratio of 3:1 [@Reichman:2001wv].
+Participants were excluded on the following criteria: those with parents who planned to place the child for adoption, those where the father was deceased, those who did not speak English or Spanish well enough to be interviewed, births where the mothers or babies were too ill to complete the interview, and those where the baby died before the interview could take place.
+Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; additional follow up is ongoing.
+Assessments included medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher and child.
+At ages nine and fifteen a saliva sample was taken from the child [@Reichman:2001wv].
+A subsample of the Fragile Families cohort was selected for saliva DNA and DNA methylation processing.
+<!--#This subsampling was ....
+PUT SUBSAMPLE technique here -->
 
-## Prenatal maternal smoking measurement
+### Covariates and exposure measurement
 
-## Covariates
+Demographic and prenatal maternal substance use variables were derived from maternal self-report questionnaire data at baseline (child's birth).
 
-## DNA methylation measurement
+Maternal covariates included maternal income to poverty ratio at baseline, prenatal smoking, prenatal maternal drug and alcohol use, and postnatal maternal or primary caregiver smoking.
+Maternal income to poverty ratio is a constructed variable of the ratio of total household income (as self-reported by the mother) to the official poverty thresholds designated by the United States Census Bureau for the year preceding the interview.
+At baseline, mother's answered categorical questions about their prenatal smoking, drug and alcohol use.
+For maternal prenatal smoking, mother's were asked
 
-## Sample filtering
+> During your pregnancy, how many cigarettes did you smoke? Did you smoke...:
+>
+> 2 or more packs a day
+>
+> 1 or more but less than 2 packs per day
+>
+> Less than 1 pack a day
+>
+> None
 
-## Probe filtering
+Few participants reported smoking a pack or more a day, thus we dichotomized to any vs no prenatal maternal smoking. For maternal prenatal drug and alcohol use the mothers were asked
 
-## Methylation summary measures
+> During your pregnancy how often did you use drugs/drink alcohol (respectively):
+>
+> Never
+>
+> Less than 1 time per month
+>
+> Several times per month
+>
+> Several times per week
+>
+> Every day
 
-We created several methylation summary measures.
-Global methylation was calculated for each sample as the mean methylation value of each sample across the cleaned probe set.
-Methylation clocks were calculated for each sample according to the methods of --INSERT HERE--.
-To create polymethylation scores, 6,000 (CHECK) CpG sites with regression coefficients reported in @Joubert:2016ue were multiplied by methylation beta values from FFCW and summed.
-We calculated polymethylation scores using regression coefficients from 4 different regressions reported in @Joubert:2016ue (LIST) and 3 different methods of transforming the methylation beta values (no transformation, mean center, and z-score standardization).
+When the child was 1, 5, 9 and 15 years of age, primary caregivers responded to questions about maternal and in-home smoking.
+To encapsulate general early childhood smoke exposure, we created a binary variable for postnatal exposure at ages 1 or 5.
+To encapsulate recent postnatal smoke exposure, we used a categorical variable for packs per day (no smoking, less than one pack/day, one or more packs/day) in the month prior to the age 9 and 15 interview.
 
-## Statistical analysis
+Child covariates included child sex, child report of personal cigarette smoking, and child genetic ancestry.
+Mothers reported sex (male/female) of their child at baseline.
+At ages 9 and 15, the child responded to interview questions about ever smoking a cigarette.
+
+Child genetic ancestry was calculated. Principal components of child genetic ancestry were calculated from genetic data (CHIP?) from child saliva samples.
+
+### DNA methylation measurement
+
+Salivary samples from the children were collected at ages nine and fifteen using the Oragene•DNA sample collection kit (DNA Genotek Inc., Ontario).
+Saliva DNA was extracted manually using DNA Genotek's purification protocol using prepIT L2P.
+DNA was bisulfite treated and cleaned using the EZ DNA Methylation kit (Zymo Research, California).
+Samples were randomized and plated across slides by demographic characteristics.
+Saliva DNA methylation was measured using the Illumina HumanMethylation 450k BeadArray [@Sandoval2011] and imaged using the Illumina iScan system.
+All samples were run in a single batch to minimize technical variability.
+
+DNA methylation image data were processed in R statistical software (3.5) using the minfi package [@Aryee2014].
+The red and green image pairs (n=1811) were read into R and the minfi *preprocessNoob* function was used to normalize dye bias and apply background correction.
+Further quality control was applied using the ewastools packages [@Heiss2018].
+We dropped samples with \>10% of sites have detection p-value \>0.01 (n=43), sex discordance between DNA methylation predicted sex and recorded sex (n=20), and abnormal sex chromosome intensity (n=3).
+CpG sites were removed if they had detection p-value \>0.01 in 5% of samples (n=26,830) or were identified as cross-reactive (n=27,782) (Figure 1) [@Chen2013].
+Relative proportions of immune and epithelial cell types were estimated from DNA methylation measures using a childhood saliva reference panel [@Middleton2021].
+
+We created polymethylation scores for prenatal maternal smoke exposure.
+From an independent meta-analysis of prenatal smoke exposure and newborn DNA methylation, we extracted the regression coefficients of 6,074 CpG sites associated with prenatal maternal smoking at a false discovery ratio corrected $P$ value \<0.05 [@Joubert:2016ue].
+We mean-centered the DNA methylation beta values in our study, weighted them by the independent refression coefficience and took the sum.
+We calculated polymethylation scores using regression coefficients from 4 different regressions (sustained smoking exposure and DNA methylation in newborn cordblood with cell-type control, sustained smoking exposure and DNA methylation in newborn cordblood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children, any smoking exposure and DNA methylation in newborn cordblood) [@Joubert:2016ue].
+Global DNA methylation was calculated for each sample as the mean methylation value of each sample across the cleaned probe set.
+Pediatric epigenetic age was calculated for each sample using the coefficients and methods provided by the creators (see [<https://github.com/kobor-lab/Public-Scripts>](https://github.com/kobor-lab/Public-Scripts){.uri}) [@mcewen2019].
+The GRIM age clock, including the smoking pack-years subscale, was calculated as previously described [@lu2019].
+
+Single *a priori* CpG sites were selected from the previous large meta-analysis of prenatal maternal smoking and DNA methylation in children's cord and peripheral blood samples [@Joubert:2016ue].
+
+Surrogate variables were calculated from DNA methylation data using the function sva from the R package sva version 3.38.
+
+### Statistical analyses
+
+#### Inclusion exclusion criteria
+
+From the 1692 samples from 884 unique individuals with quality controlled DNA methylation data, we further excluded any samples missing data on: maternal prenatal smoking (0 samples), alcohol (4 samples) or other drug use (6 samples), maternal income to poverty ratio (0 samples), maternal postnatal smoking data ( samples).
+We also excluded samples with missing child sex (0 samples), child age (0 samples) or genetic data ( samples).
+Finally, if a child reported ever smoking a cigarette at age 9, we excluded their samples.
+Age 15 samples from children who reported ever smoking a cigarette at age 15 were also excluded.
+
+#### Primary model
+
+In our primary model we adjusted for maternal income to poverty ratio at baseline, child sex, child age, plate from DNA methylation processing, and estimated immune cell proportion estimated from DNA methylation.
+In nonstratified models, we adjusted for the first two components of genetic ancestry from principal component analysis.
+In ancestry-stratified models, the first two principal components from principal component analysis run within each ancestry strata were used.
+While child sex, child age, plate from DNA methylation processing and immune cell proportions are not confounders (as they cannot casually effect prenatal maternal smoke exposure), these variables can strongly effect DNA methylation and so were adjusted for as precision variables.
+
+#### Sensitivity analyses
+
+In sensitivity analyses we performed additional adjustments: 1) models adjusting for prenatal drug and alcohol use, 2) models adjusting for prenatal drug and alcohol use and postnatal maternal/primary caregiver smoking, 3) models adjusting for surrogate variables calculated from DNA methylation data.
+
+Code to perform all analyses is available (www.github.com/bakulskilab)

### 2021-11-04 Papers/Polymethylation_scores/02_abstract.Rmd
more revisions post Kelly 1st round comments

@@
 **Background**: Prenatal maternal smoking is a difficult to measure prenatal exposure with negative implications for child health.
-DNA methylation signatures vary by prenatal maternal smoke exposure in blood and placental samples and can function as a biomarker.
+DNA methylation signatures in blood and placental samples can function as a biomarker of prenatal maternal smoking.
 However little work has assessed how DNA methylation signatures of prenatal maternal smoking vary across ages, ancestry groups or tissues.
-Using children in the Fragile Families and Child Wellbeing study, we tested if prenatal maternal smoking associates with salivary DNA methylation summary measures, including global methylation, epigenetic age, and polymethylation scores.
+Using children in the Fragile Families and Child Wellbeing study, we tested if prenatal maternal smoking is associated with salivary polymethylation scores.
 We assessed the consistency of associations at ages 9 and 15 and their portability across  African, European and Hispanic ancestries.
 
-**Results**: We created a polymethylation score using coefficients from a metanalysis of prenatal maternal smoke exposure and DNA methylation in newborn cordblood.
-Prenatal maternal smoking was associated with this polymethylation score () but not epigenetic age or global methylation ().
-The association with the polymethylation score was consistent at age 9 () and age 15 ().
-The direction and magnitude of the association was portable across ancestries at age 9 (African: European: Hispanic), although the association was attenuated among European children at age 15 (African: European: Hispanic).
-The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (Area under the curve from receiver operating curve (AUC-ROC age 9: age 15: ), and more accurate than using the top CpG site identified in the metaanalysis, AHRR:CpG05575921 (AUC-ROC age 9:, age 15:, P value).
+**Results**: We created a polymethylation score using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cordblood.
+At age 9, prenatal maternal smoke exposure was associated with a 0.52 (95%CI: 0.36, 0.67) standard deviation higher polymethylation score.
+At age 15, At age 9, prenatal maternal smoke exposure was associated with a 0.46 (95%CI: 0.3, 0.62) higher polymetylation score.
+The direction and magnitude of the association was portable across genetic ancestries at age 9 although the association was attenuated among European and Hispanic genetic ancestry samples at age 15.
+The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (Area under the curve from receiver operating curve (AUC-ROC age 9: 0.77 age 15: 0.77), and more accurate than using a single *a priori* CpG site, AHRR:CpG05575921 (AUC-ROC age 9:0.74, *P* value=0.02; age 15:0.73, *P* value=0.02).
 

### 2021-11-04 Papers/Polymethylation_scores/04_background.Rmd
more revisions post Kelly 1st round comments

@@
 Adjusting for ancestry did not substantially affect the association between prenatal maternal smoking and DNA methylation at any of the sites, although the authors note that African American women had on average 2--5% higher mean DNA methylation on the absolute scale as compared with White and Hispanic women at 5 *CYP1A1* CpG sites [@Tehranifar:2018tb].
-Of 148 CpG sites selected based on their association prenatal maternal smoking among primarily European children, 7 CpG sites were associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue].
+Of 148 CpG sites selected based on their association prenatal maternal smoking among primarily European children, 7 CpG sites were also associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue].
 Thus, while prenatal maternal cigarette smoking is associated with DNA methylation, the consistency of this association across age, tissue and ancestry is understudied.
@@
 An important next step after association testing is the evaluation of DNA methylation as a biomarker of prenatal maternal smoke exposure, which is often done by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior.
-<!--#DEFINE POLYMETHYLATION SCORES HERE?-->DNA methylation signatures from cord blood classified maternal smoking behavior with accuracy of 91% in a primarily European cohort [@Reese:2017ts].
+Many methods for constructing a DNA methylation biomarker exist, ranging in complexity from single CpG sites to machine learning algorithms such as support vector machines [@Philibert2020; @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa].
+One attractive option is a polymethylation score, analogous to a polygenic score.
+In a polymethylation score, CpG sites are weighted by the strength of their association with the variable of interest in a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a].
+Using a subset of CpG sites selected with LASSO regression, one such DNA methylation score calculated from cord blood classified maternal smoking behavior with accuracy of 91% in a primarily European cohort [@Reese:2017ts].
 Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking behavior could also be classified with an accuracy of 81% when using a support vector machine approach and DNA methylation at 26 sites measured in peripheral blood at a single time point [@Ladd-Acosta:2016aa].
-Among middle-aged adults, peripheral blood from a single time point could even predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu].
+Among middle-aged adults, a score calculated using DNA methylation in peripheral blood samples from a single time point could even predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu].
+Ideally, a biomarker for prenatal maternal smoking would be consistent over participant age at methylation sampling and portable across genetic ancestry groups.
 However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across tissue, age and ancestry has not been evaluated.

### 2021-11-04 Papers/Polymethylation_scores/05_results.Rmd
more revisions post Kelly 1st round comments

@@
 output:
+  html_document: default
   word_document: default
-  html_document: default
 bibliography: Methylation_SmokingBib.bib
@@
 library(knitr)
-knit_child(here::here("Reports", "Results.Rmd"))
+knitr::knit_child(here::here("Reports", "Paper1.Rmd"))
 ```
@@
 
-Of `r methylCohort$MethID%>%length()` samples from `r methylCohort$idnum%>%unique()%>%length()` unique individuals with DNA methylation data measured on the 450K Illumina array, `r m_obs` samples from `r n_total` individuals had data on additional key covariates (Figure 1).
-`r n_twovisit` individuals had both age 9 and age 15 samples.
-Excluded samples were similar to included samples, except that included samples were slightly more likely to be from children of African ancestry and less likely to be from children of Hispanic ancestry (Supplemental Table 1).
-Twenty percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alchol use and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use.
+Of `r methylCohort$MethID%>%length()` samples from `r methylCohort$idnum%>%unique()%>%length()` unique participants with DNA methylation data measured on the Illumina 450K  array, `r m_obs` samples from `r n_total` participants had complete data on additional key covariates (Figure 1).
+There were `r n_twovisit` participants with both age 9 and age 15 DNA methylation samples.
+Excluded samples were similar to included samples, except that included samples were slightly more likely to be from children of African genetic ancestry and less likely to be from children of Hispanic genetic ancestry (Supplemental Table 1).
+In the included sample, 20% percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alcohol use and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use.
 The mean income to poverty ratio of mothers at birth was `r round(mean(modeldata_wide$cm1inpov), 2)`.
-Of the children in the analytic sample, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Hispanic ancestry']`% were of Hispanic ancestry.
+Of the children in the analytic sample, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African genetic ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Hispanic ancestry']`% were of Hispanic genetic ancestry.
 
-We calculated several common summary measures of DNA methylation, including global methylation, epigenetic clocks (Pediatric and GRIM), and polymethylation scores.
-We calculated polymethylation scores using regression coefficients for 6074 CpGs made from three different models in Joubert et al.: a regression using newborn cordblood samples controlling for cell-type proportions, a regression using newborn cordblood samples not controlling for cell-type proportions and a regression using peripheral blood samples from older children.
-We also examined the top probe from Joubert et al., AHRR:cg05575921.
+We calculated several summary measures of DNA methylation, including polymethylation scores, global methylation, and epigenetic clocks.
 Many of the methylation summary measures were correlated with each other (Supplemental Figure 1).
-Notably, polymethylation scores calculated using coefficients from a cell-type corrected regression as weights did not correlate with estimated cell-type proportions, while scores calculated using coefficients from a regression not corrected for cell-type did.
-Among the `r n_twovisit` individuals with data from both visits, methylation summary measures from the same individual were correlated between the age 9 and age 15 (Supplemental Figure 2).
-The correlation was strongest for the polymethylation score calculated using weights from a regression for sustained smoking in newborn cordblood with cell-type control (Pearson $\rho$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value<0.0001).
-The distribution of epigenetic ages shifted between the age 9 and age 15 visit, as expected (Supplemental Figure 3; mean pediatric clock age 9: `r round(mean(modeldata_wide$pediatric_Age_9, na.rm=T))` age 15: `r round(mean(modeldata_wide$pediatric_Age_15, na.rm=T))`, mean GRIM clock age 9 `r round(mean(modeldata_wide$grim_Age_9, na.rm=T))`, age 15: `r round(mean(modeldata_wide$grim_Age_15, na.rm=T))`.
+For example, as expected, the percent of estimated immune cells was perfectly inversely correlated with the percent of estimate epithelial cells (Pearson $rho$=`r cor(modeldata$Epithelial.cells_saliva, modeldata$Leukocytes_saliva)`; *P* value<0.0001).
+Our primary polymethylation score  measure, which was calculated using coefficients from a cell-type corrected regression, was very slightly correlated estimated cell-type proportion of immune cells (Pearson $rho$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value, 3)`).
+
+We compared epigenetic measures between the age 9 and age 15 visit among the `r n_twovisit` individuals with data from both visits (Supplemental Figure 2).
+The correlation across ages was strongest for the primary polymethylation score (Pearson $\rho$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value<0.0001).
+The distribution of epigenetic ages shifted between the age 9 and age 15 visit, as expected (Supplemental Figure 3).
+Among age 9 samples, the mean estimated age from the pediatric clock was `r round(mean(modeldata_wide$pediatric_Age_9, na.rm=T))`  and the mean estimated age from the GRIM clock was `r round(mean(modeldata_wide$grim_Age_9, na.rm=T))`.
+Among age 15 samples, the mean estimated age from the pediatric clock was `r round(mean(modeldata_wide$pediatric_Age_15, na.rm=T))` and the mean estimated age from the GRIM clock was `r round(mean(modeldata_wide$grim_Age_15, na.rm=T))`.
 The distribution of estimated cell-type proportions also shifted between visits while the distribution of global methylation and polymethylation scores were more consistent (Supplemental Figure 3).
-Based on these results, we focused on the polymethylation score from newborn cordblood with cell-type control, the pediatric clock, AHRR:cg05575921 and global methylation for bivariate and multivariable analyses.
+
+In bivariate analyses and multivariable models, we focused on the polymethylation scores and the single top CpG site from Joubert et al, AHRR:cg05575921, as hypothesized biomarkers, and used global methylation and the pediatric clock as negative controls.
 
@@
 
-Mothers who reported smoking during pregnancy had lower income to poverty ratios and were more likely to report prenatal alcohol use, prenatal drug use and postnatal smoking than mothers who did not report prenatal maternal smoking (Table 1).
-Children of mothers who reported smoking during pregnancy were more likely to be of European ancestry and had higher prenatal maternal smoking polymethylation scores at both the age 9 and age 15 visit (Table 1).
-Children of mothers who reported prenatal maternal smoking did not differ from those whose mothers did not in terms of global methylation, epigenetic age estimaated from pediatric clock or estimate cell-type proportions.
-Compared to children of mothers not reporting prenatal materna smoking, exposed children were significantly more hypomethylated at AHRR:cg05575921 at age 9 (mean methylation `r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`) but not age 15 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)` *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
+Mothers who reported smoking during pregnancy had lower income to poverty ratios than those who did not ( `r modeldata_people%>%filter(smkPreg_binary=='Yes')%>%summarise(mean(cm1inpov))%>%round(2)` vs `r modeldata_people%>%filter(smkPreg_binary=='No')%>%summarise(mean(cm1inpov))%>%round(2)`).
+Mothers who smoked were more likely to report prenatal alcohol use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%), prenatal drug use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%) and postnatal smoking (`r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[1, 2]%>%round(2)*100`%) than mothers who did not report prenatal maternal smoking (Table 1).
+Children of mothers who reported smoking during pregnancy were more likely to be of European genetic ancestry (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[2, 2], 2)*100`%) than children of mothers who did not (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[1, 2], 2)*100`%, *P* value=`r round(with(modeldata_people, chisq.test(table(smkPreg_binary, ancestry)))$p.value, 4)`).
+Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both the age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) visits (Table 1).
+At age 9, children exposed to prenatal maternal smoking had lower DNA methylation at AHRR:cg05575921 (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)*100`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)*100`%, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
+At age 15, children exposed to prenatal smoking had lower DNA methylation at AHRR:cg05575921 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)*100`%) than children of non-smoking mothers, although this difference was not significant (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)*100`% *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
 
-## Multivariable models
+## Multivariable associations between prenatal maternal smoking and DNA methylation summary measures
 
-The association between prenatal maternal smoking and the cell-type controlled polymethylation score was also observed in multivariable models controlling for precision and confounder variables (Figure 3; Supplemental Table 2).
-In the entire sample, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` higher polymethylation score at age 9 and `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` higher polymethylation score at age 15.
-A consistent association was observed when stratifying to only African ancestry children (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`).
-The direction and magnitude of the association was similar in European and Hispanic ancestry children at age 9, though attenuated at age 15 (Figure 3; Supplemental Tables 2).
-Prenatal maternal smoking was associated with hypomethylation at AHRR:cg055975921 at age 9 but not age 15 and not associated with global methylation or epigenetic age from the pediatric clock at either age.
-Results were similar in linear mixed effect models and in sensitivity models controlling for surrogate variables instead of known covariates (Supplemental Tables 3-4; Supplemental Figure 4-5).
+The association between prenatal maternal smoking and the polymethylation score was also observed in multivariable models  (Figure 3; Supplemental Table 2).
+At age 9, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.
+At age 15, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.
+A consistent association was observed when stratifying to the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`).
+In the African genetic ancestry sample at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` higher polymethylation score.
+The direction and magnitude of the association was similar in the European and Hispanic genetic ancestry samples at age 9, though attenuated at age 15 (Figure 3; Supplemental Tables 2).
+
+Prenatal maternal smoking was associated with lower methylation at AHRR:cg055975921 at age 9 (`r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921'", 'smkPreg_binaryYes', 2)`) and age 15 (`r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921'", 'smkPreg_binaryYes', 2)`).
+Associations between prenatal maternal smoking and AHRR:cg05575921 were less robust across ancestries (Figure 3).
+
+## Sensitivity analyses
+
+When also adjusting for other prenatal exposures and postnatal smoke exposure, the association between prenatal maternal smoking and the polymethylation score was very consistent (Supplemental Figure 4).
+The direction of the association between prenatal maternal smoking and AHRR:cg055975921 was also consistent, although no longer significant at age 15.
+Children exposed to postnatal smoke but not prenatal smoke did not have higher polymethylation scores than children unexposed to both postnatal and prenatal smoke.
+(Supplemental Figure 5).
+Results were also similar when controlling for surrogate variables instead of known covariates (Supplemental Figure 6; Supplemental Table 2).
+
+Results were similar when using linear mixed effect models instead of age-stratified models (Supplemental Table 3 & 4).
+
+Results were also similar when using polymethylation scores constructed using regression coefficients (see Methods, Supplemental Figure 7).
+The association between prenatal maternal smoking and the polymethylation score using regression coefficients from newborn cord blood with cell-type controls was the strongest.
+Prenatal maternal smoking also associated with other selected *a priori* CpG sites consistently with associations reported in previous meta-analyses (Supplemental Figure 8).
 
@@
 
-Next, we compared the accuracy of different DNA methylation summary measures for classifying prenatal maternal smoking using receiver operating curves (Figure 4A).
-The polymethylation score with cell-type control had the largest area under the curve (AUC) both when used as a single predictor (age 9 AUC; `r rct1%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 9')%>%pull(auc)`, age 15 AUC: `r rct1%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 15')%>%pull(auc)` ) and when used in models including child sex, maternal income-poverty ratio, immune cell proportion and batch (age 9 AUC: `r rct2%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 9')%>%pull(auc)`, age 15 AUC: `r rct2%>%filter(str_detect(modeltype, 'cell type')&childteen=='Age 15')%>%pull(auc)`).
-The AUC for the polymethylation score was significantly larger than using only the top CpG from Joubert et al., AHRR:cg05575921 (age 9 AUC: `r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 9')%>%pull(auc)` *P* value=`r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 9')%>%pull(4)%>%round(3)` , age 15 AUC: `r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 15')%>%pull(auc)` *P* value=`r rct1%>%filter(str_detect(modeltype, 'AH')&childteen=='Age 15')%>%pull(4)%>%round(3)`).
+Next, we compared the accuracy of different DNA methylation summary measures for classifying prenatal maternal smoking using receiver operating curves (Figure 4A; Supplemental Table 5).
+We considered classification when using DNA methylation summary measures alone.
+We also considered classification when using DNA methylation summary measures in addition to a base model of child sex, maternal income to poverty ratio at baseline, child age at methylation measurement, estimated immune cell proportion, plate from methylation processing and the first two components of genetic ancestry.
+When used without DNA methylation measures, these base model variables had an area under the curve (AUC) of `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 9')%>%pull(auc)` at age 9 and `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 15')%>%pull(auc)` at age 15.
+
+At age 9, including the polymethylation score  significantly improved the base model accuracy (AUC: `r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(auc)`, *P* value comparing to base model=`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(4)`).
+Similarly, including the polymethylation significantly improved the base model accuracy at age 15 (AUC:`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(auc)` , *P* value=`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(4)`).
+
+At age 9,the base model with the polymethylation score had a larger AUC than the base model with AHRR:cg05575921 (AUC base model+AHRR:cg05575921: `r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(auc)`; *P* value =`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(5)`).
+At age 15, this was also true (AUC for base model+AHRR:cg05575921=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(auc)`,  *P* value=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(5)`).
 Accuracy was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B).
+
+## Sensitivity analyses
+
+Results were similar when using other coefficients as the weights in construction of the polymethylation scores.
+However, the AUC was always significantly higher when using the coefficients from a model which incorporated cell-type control as weights (Supplemental Table 6).

### 2021-11-04 Papers/Polymethylation_scores/06_discussion.Rmd
more revisions post Kelly 1st round comments

@@
 
-In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed prenatal maternal smoking was associated with DNA methylation in children's saliva samples from ages 9 and 15.
+In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed that prenatal maternal smoking was associated with DNA methylation in children's saliva samples from ages 9 and 15.
 Prenatal maternal smoking was associated with polymethylation scores across strata of child age and ancestry.
@@
 Our findings are consistent with the previous literature on associations between prenatal maternal smoking and DNA methylation.
-We replicated top hits from previous epigenome wide association studies, including a large meta-analysis of DNA methylation in newborn blood samples [@Joubert:2016ue].
+We replicated top hits from previous epigenome wide association studies of DNA methylation in cord and peripheral blood[@Joubert:2016ue; @wiklund2019].
 There are no previous studies of prenatal maternal smoking and saliva DNA methylation.
-However, previous work has that the majority of CpG sites were similarly methylated in blood and saliva (reviewed in @Langie2017).
-DNA methylation at AHRR:cg05575291 can accurately classify own--smoking behavior in both blood (area under the curve 0.995) and saliva (area under the curve 0.971) [@Dawes2019; @Philibert2020].
+However, previous work has shown that the majority of CpG sites are similarly methylated in blood and saliva (reviewed in @Langie2017).
 
@@
 Polymethylation scores built using coefficients from meta-analysis of cord blood DNA methylation from primarily European-ancestry newborns were still associated with prenatal maternal smoking in our independent saliva samples from a diverse cohort at ages 9 and 15 [@Joubert:2016ue].
-The portability of other risk scores, such as polygenic risk scores, across ancestries is a complex research area [@Martin2019; ] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Hüls2020].
+The portability of other risk scores, such as polygenic risk scores, across ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Hüls2020].
 In this case, the polymethylation score for prenatal maternal smoking appears to be portable across ancestry groups.
@@
 Our findings are also consistent with previous work on DNA methylation as a biomarker for prenatal maternal smoke exposure.
-Prenatal maternal smoking classification using saliva DNA methylation in our sample of 9 and 15 year olds () performed comparably to classification using peripheral blood from older adults () [@Richmond:2018tu].
-However, previous biomarkers of prenatal maternal smoking using DNA methylation of cord blood from newborns () and peripheral blood from 3-5 year old children () more accurately classified prenatal maternal smoke exposure [@Reese:2017ts; @Ladd-Acosta:2016aa].
+Prenatal maternal smoking classification using saliva DNA methylation in our sample of 9 and 15 year olds performed comparably to classification using peripheral blood from older adults (AUC 0.72) [@Richmond:2018tu].
+However, previous biomarkers of prenatal maternal smoking using DNA methylation of cord blood from newborns (AUCs ranging from 0.82-0.97) and peripheral blood from 3-5 year old children (AUC=0.87) more accurately classified prenatal maternal smoke exposure [@Reese:2017ts; @Ladd-Acosta:2016aa].
 Differences in DNA methylation patterns across tissues and over time may influence the performance of DNA methylation biomarkers.
@@
 Our analysis contributes to the development of an accurate methylation biomarker for prenatal maternal smoking by evaluating the impact of specific methodological choices on accuracy of classification.
-DNA methylation biomarkers may be especially susceptible to confounding by subject age at methylation measurement, tissue of measurement, and cell-type proportion [@Hüls2020].
+DNA methylation biomarkers may be especially susceptible to confounding by subject age at methylation measurement and cell-type proportion [@Hüls2020].
 Accuracy of polymethylation score classification was similar when using coefficients from cord blood vs peripheral blood samples from older children.
@@
 Our analysis suggests that polymethylation scores may be more accurate than using single CpG sites as biomarkers.
-AHRR:cg05575921 is a consistent marker of prenatal maternal smoke exposure in meta-analyses of newborn cord blood[@Joubert:2016ue] and a very accurate DNA methylation marker of own smoking [@Philibert2020].
+AHRR:cg05575921 is a consistent marker of prenatal maternal smoke exposure in meta-analyses of newborn cord blood[@Joubert:2016ue].
+DNA methylation at AHRR:cg05575291 can accurately classify own--smoking behavior in both blood (area under the curve 0.995) and saliva (area under the curve 0.971) [@Dawes2019; @Philibert2020].
 However, salivary AHRR:cg05575921 was not a persistent marker of prenatal maternal smoking in an analysis of middle-aged adult women [@Richmond:2018tu].

### 2021-11-04 Papers/Polymethylation_scores/08_methods.Rmd
more revisions post Kelly 1st round comments

@@
 A subsample of the Fragile Families cohort was selected for saliva DNA and DNA methylation processing.
+To be eligible for DNA methylation children had to 1) have participated and given saliva at age 9 and 2) participated in the SAND {ACRONYM HELP HERE} study, which oversampled from Detroit, Toledo and Chicago.
+A random sample of these eligible participants was then selected.
 <!--#This subsampling was ....
@@
 Mothers reported sex (male/female) of their child at baseline.
-At ages 9 and 15, the child responded to interview questions about ever smoking a cigarette.
+At ages 9 children were asked if they had ever smoked a cigarette or used tobacco (yes/no) and at age 15 they were asked if they had ever smoked an entire cigarette (yes/no).
 
@@
 From an independent meta-analysis of prenatal smoke exposure and newborn DNA methylation, we extracted the regression coefficients of 6,074 CpG sites associated with prenatal maternal smoking at a false discovery ratio corrected $P$ value \<0.05 [@Joubert:2016ue].
-We mean-centered the DNA methylation beta values in our study, weighted them by the independent refression coefficience and took the sum.
-We calculated polymethylation scores using regression coefficients from 4 different regressions (sustained smoking exposure and DNA methylation in newborn cordblood with cell-type control, sustained smoking exposure and DNA methylation in newborn cordblood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children, any smoking exposure and DNA methylation in newborn cordblood) [@Joubert:2016ue].
+We mean-centered the DNA methylation beta values in our study, weighted them by the independent regression coefficients and took the sum.
+We calculated polymethylation scores using regression coefficients from 4 different regressions.
+For our main analysis we used coefficients from a regression of sustained smoking exposure and DNA methylation in newborn cordblood with cell-type control.
+As sensitivity analyses, we used re sustained smoking exposure and DNA methylation in newborn cordblood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children, any smoking exposure and DNA methylation in newborn cordblood) [@Joubert:2016ue].
 Global DNA methylation was calculated for each sample as the mean methylation value of each sample across the cleaned probe set.
-Pediatric epigenetic age was calculated for each sample using the coefficients and methods provided by the creators (see [<https://github.com/kobor-lab/Public-Scripts>](https://github.com/kobor-lab/Public-Scripts){.uri}) [@mcewen2019].
+Pediatric epigenetic age was calculated for each sample using the coefficients and methods provided by the creators (see [\<https://github.com/kobor-lab/Public-Scripts\>](https://github.com/kobor-lab/Public-Scripts){.uri}) [@mcewen2019].
 The GRIM age clock, including the smoking pack-years subscale, was calculated as previously described [@lu2019].
 
-Single *a priori* CpG sites were selected from the previous large meta-analysis of prenatal maternal smoking and DNA methylation in children's cord and peripheral blood samples [@Joubert:2016ue].
+Single *a priori* CpG sites were selected from previous large meta-analyses of prenatal maternal smoking and DNA methylation in children's cord and peripheral blood samples [@Joubert:2016ue; @wiklund2019].
 
@@
 We also excluded samples with missing child sex (0 samples), child age (0 samples) or genetic data ( samples).
-Finally, if a child reported ever smoking a cigarette at age 9, we excluded their samples.
+Finally, if a child reported ever smoking a cigarette or using tobacco age 9, we excluded all of their available samples.
 Age 15 samples from children who reported ever smoking a cigarette at age 15 were also excluded.
+Children who were missing a response to the question at age 9 but answered that they had never smoked a whole cigarette at age 15 were kept in the sample.
 
@@
 
+#### Receiver operator curve analysis
+
+To evaluate the accuracy of the DNA methylation summary measures as biomarkers of prenatal maternal smoking, we used a receiver operating curve.
+First, we regressed exposure to prenatal maternal smoking (outcome) againts the DNA methylation summary measures in individual logistic regressions, while adjusting for the base model variables listed above.
+Next, we calculated receiver operating curves (ROC) and area under the curves (AUCs) using the function roc from the R library pROC version 1.18.0.
+We compared ROC curves and AUC using the delong method and the function roc.test from the R library pROC version 1.18.0.
+
 #### Sensitivity analyses
@@
 
+In addition to the cross-sectional models within visit-strata, we performed a linear mixed effect model with a random intercept for child.
+
 Code to perform all analyses is available (www.github.com/bakulskilab)

### 2021-11-04 Papers/Polymethylation_scores/98_declarations.Rmd
added declarations
more revisions post Kelly 1st round comments

@@
 
-## Ethics approval and consent to participate
+### Ethics approval and consent to participate
 
-## Consent for publication
+Participants provided written informed consent for the study.
+The data used in this manuscript were prepared by the Fragile Families and Childhood Wellbeing Study administrators following approval of the manuscript proposal.
+These secondary data analyses were approved by the University of Michigan Institutional Review Board (IRB, HUM00129826)
+
+### Consent for publication
 
 Not applicable
+
+### Availability of data and materials
+
+Many of the variables used in this analysis are publicly available in the Fragile Families dataset.
+Some of the data used in this analysis, including genetic and epigenetic data, is restricted use but available to researchers upon reasonable application.
+
+### Competing interests
+
+The authors have no competing interests to declare. check here
+
+### Funding
+
+This research was made possible through several grants (add). FB was supported by the National Institutes of Health National Institute of Dental and Craniofacial Research (F31DE029992).
+
+### Authors' contributions
+
+FB performed the analysis with contributions from JD, JF, and EW.
+FB and KB wrote the paper with all authors contributing to revisions.
+DN, CM, KM, EW and FB all contributed to the design and conception of the study.
+
+### Acknowledgements
+
+Not applicable

### 2021-11-04 Papers/Polymethylation_scores/99_FigureCaptions.Rmd
more revisions post Kelly 1st round comments

---
output:
  html_document: default
  word_document: default
bibliography: Methylation_SmokingBib.bib
always_allow_html: true
---

# Figure 1 - Selection of samples from the Fragile Families and Child Wellbeing study into analytic subset. N represents the number of individuals at each step in the selection procedure, M represents the number of samples. Individuals with repeated measures can have more than one sample.

# Figure 2 - Differences in selected DNA methylation summary measures by self-report of prenatal maternal smoking among `r n_total` children in the Fragile Families and Child Wellbeing study (CT=cell-type control)

# Figure 3 - Prental maternal smoke exposure is consistently associated with polymethylation scores at ages 9 and 15 and is portable across genetic ancestry groups in a sample of `r n_total` children in the Fragile Families and Child Wellbeing study

# Figure 4 - A) Polymethylation scores accurately classify maternal report of prenatal smoke exposure status at ages 9 and 15 among a `r n_total` children in the Fragile Families and Child Wellbeing Study B) Inclusion of two-visits of DNA methylation data does not improve classification of maternal report of prenatal smoke exposure status among `r n_total` children in the Fragile Families and Child Wellbeing study

### 2021-11-04 Papers/Polymethylation_scores/index_paper.Rmd
more revisions post Kelly 1st round comments

@@
 ---
-title: "Polymethylation scores paper"
-author: "F blostein"
-date: "9/13/2021"
+title: "PAPER DRAFT"
+author: "------------"
+date: "`r Sys.Date()`"
 output:
@@
 
+```{r child = here::here("Papers", "Polymethylation_scores", "99_FigureCaptions.Rmd"), eval=results_run}
+```
+
 # References

### 2021-11-04 Reports/Paper1.Rmd
more revisions post Kelly 1st round comments

--- 2021-10-28 Reports/Results.Rmd
@@
 
-# Figure `r main_fig` - Selection of samples from the Fragile Families and Child Wellbeing (FFCW) study into this analysis.
+# Figure `r main_fig` - Selection of samples from the Fragile Families and Child Wellbeing (FFCW) study into analytic subset.
 N represents the number of individuals at each step in the selection procedure, M represents the number of samples, as individuals with repeated measures can have more than one sample.
@@
 
-if(childdata %>% filter((childteen=='C' & k5f1l=='2 no') | (childteen=='T' & k6d40=='2 No' & k5f1l!='1 yes'))%>%nrow()!=nrow(modeldata)){stop('filter N and completecase data set N do not match')}
+if(childdata %>% filter(((k5f1l=="2 no" | (is.na(k5f1l) & k6d40=='2 No')) & childteen=='C') | (childteen=='T' & k6d40=="2 No" & (k5f1l=='2 no' |  is.na(k5f1l))))%>%nrow()!=nrow(modeldata)){stop('filter N and completecase data set N do not match')}
 
@@
 model_twovisit=modeldata_wide[complete.cases(modeldata_wide), ]
+
+modeldata_people=modeldata%>%
+  dplyr::select(any_of(c('idnum', 'ancestry', 'cm1bsex', 'ancestry', 
+                         'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'cm1inpov',
+                         'smkPreg_binary', 'PostnatalMaternalSmokingAny')))%>%unique()
 ```
@@
   unnest(tidied)%>%
-  filter(outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & term=='smkPreg_binaryYes'& predictors=='secondhand smoke exposure model')%>%mutate(outcome_f=outcome)
+  filter(outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & term=='smkPreg_binaryYes'& predictors=='base model')%>%mutate(outcome_f=outcome)
 
@@
   unnest(tidied)%>%
-  filter(term=='smkPreg_binaryYes' & predictors=='secondhand smoke exposure model')%>%
+  filter(term=='smkPreg_binaryYes' & predictors=='base model')%>%
   filter(outcome%in% my_outcomes)%>%
@@
 
-# Supplemental Table `r sup_tab`: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable models among `r n_total` US children by age
-
-```{r}
-#my_main_outcomes=c('AHRR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)', 'Sustained smoking polymethylation score (older children)', 'Pediatric clock', 'Global methylation')
-
-test=rbind(global_models, longitudinal_global_models %>% mutate(childteen='Longitudinal'))%>%
-  filter(predictors=='secondhand smoke exposure model')%>%
-  filter(outcome %in% my_outcomes)%>%
-  mutate(childteen=factor(case_when(childteen=='Age 9'~'Age 9 models', 
-                             childteen=='Age 15'~'Age 15 models', 
-                             childteen=='Longitudinal'~'Longitudinal models'),levels=c('Age 9 models', 'Age 15 models', 'Longitudinal models')), 
-         tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
-                                        show_single_row='smkPreg_binary', 
-                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
-
-table_list=test %>%split(f=.$childteen)%>%map(~tbl_merge(.x$tbl, 
-                                                         tab_spanner = as.character(.x$outcome)))
-
-regression_table_age=tbl_stack(table_list, group_header=names(table_list))
-
-sup_tab=sup_tab+1
-```
-
-```{r, eval=stand_alone}
-regression_table_age
-```
-
-# Supplemental Table `r sup_tab`: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable models among `r n_total` US children by ancestry
-
-```{r}
-test=rbind(longitudinal_global_models %>% mutate(ancestry='Unstratified'),
-           longitudinal_local_models%>%dplyr::select(-data))%>%
-  filter(predictors=='secondhand smoke exposure model')%>%
-  filter(outcome %in% my_outcomes)%>%
-  mutate(tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
-                                        show_single_row='smkPreg_binary', 
-                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
-
-table_list=test %>%split(f=.$ancestry)%>%map(~tbl_merge(.x$tbl, 
-                                                         tab_spanner = as.character(.x$outcome)))
-
-regression_table_ancestry=tbl_stack(table_list, group_header=names(table_list))
-
-sup_tab=sup_tab+1
-
-```
-
-```{r, eval=stand_alone}
-regression_table_ancestry
-```
-
 # Supplemental Figure `r sup_fig`: Surrogate variable associations with known covariates
@@
 
-# Supplemental Table `r sup_tab`: Coefficients for multivariable models of prenatal maternal smoke exposure regressed on selected DNA methylation summary variables (outcomes) among children in the Fragile Families and Child Wellbeing study
+# Supplemental Table `r sup_tab`: Coefficients for multivariable models of prenatal maternal smoke exposure regressed on selected DNA methylation summary variables (outcomes) among children in the Fragile Families and Child Wellbeing study using surrogate variables
 
@@
 regression_table_ancestry_svs
+```
+
+# Supplemental Table `r sup_tab`: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable linear models stratified by age or in multivariable linear mixed effect models among `r n_total` US children
+
+```{r}
+#my_main_outcomes=c('AHRR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)', 'Sustained smoking polymethylation score (older children)', 'Pediatric clock', 'Global methylation')
+
+test=rbind(global_models, longitudinal_global_models %>% mutate(childteen='Longitudinal'))%>%
+  filter(predictors=='base model')%>%
+  filter(outcome %in% my_outcomes)%>%
+  mutate(childteen=factor(case_when(childteen=='Age 9'~'Age 9 models', 
+                             childteen=='Age 15'~'Age 15 models', 
+                             childteen=='Longitudinal'~'Longitudinal models'),levels=c('Age 9 models', 'Age 15 models', 'Longitudinal models')), 
+         tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
+                                        show_single_row='smkPreg_binary', 
+                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
+
+table_list=test %>%split(f=.$childteen)%>%map(~tbl_merge(.x$tbl, 
+                                                         tab_spanner = as.character(.x$outcome)))
+
+regression_table_age=tbl_stack(table_list, group_header=names(table_list))
+
+sup_tab=sup_tab+1
+```
+
+```{r, eval=stand_alone}
+regression_table_age
+```
+
+# Supplemental Table `r sup_tab`: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable models among `r n_total` US children by ancestry
+
+```{r}
+test=rbind(longitudinal_global_models %>% mutate(ancestry='Unstratified'),
+           longitudinal_local_models%>%dplyr::select(-data))%>%
+  filter(predictors=='base model')%>%
+  filter(outcome %in% my_outcomes)%>%
+  mutate(tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
+                                        show_single_row='smkPreg_binary', 
+                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
+
+table_list=test %>%split(f=.$ancestry)%>%map(~tbl_merge(.x$tbl, 
+                                                         tab_spanner = as.character(.x$outcome)))
+
+regression_table_ancestry=tbl_stack(table_list, group_header=names(table_list))
+
+sup_tab=sup_tab+1
+
+```
+
+```{r, eval=stand_alone}
+regression_table_ancestry
+```
+
+# Supplemental Figure `r sup_fig`: Associations between prenatal maternal smoking and polymethylation scores calculated using alternative coefficient sets among `n n_total` children in the Fragile Families and Child Wellbeing study
+
+```{r} 
+ compare_pms=all_models%>%
+    mutate(n=map_dbl(data, nrow), n=paste0('n=', n))%>%
+    unnest(tidied)%>%
+    filter(term=='smkPreg_binaryYes')%>%
+    filter(str_detect(outcome, 'smoking'))%>%
+  filter(ancestry=='Unstratified')%>%
+ ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=outcome))+
+    geom_point(position = position_dodge(width = 1))+
+    geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
+    facet_grid(predictors~ancestry, scales='free_x')+
+    geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
+    theme_classic()
+
+ggexport(compare_pms, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=9)
+sup_fig=sup_fig+1
+
+```
+
+```{r, eval=stand_alone, fig.width=10}
+compare_pms
+```
+
+# Supplemental Figure `r sup_fig` - Associations between prenatal maternal smoking and other *a priori* CpG sites selected from previous meta-analyses among `n n_total` children in the Fragile Families and Child Wellbeing study
+
+```{r}
+topCPGs<-read.csv(paste0(datadir, "/TopCpGsWiklundEtAl2019.csv"))
+topCPGs<-topCPGs %>% mutate(estimate=as.numeric(gsub("−\\s", "-", gsub(" .*", "", β..SE.))), std.error=as.numeric(sub("\\).*", "", sub(".*\\(", "", β..SE.))), p.value=P.value, source="Wiklund et al.
+2019", conf.low=(estimate-1.96*std.error), conf.high=(estimate+1.96*std.error), set ='')%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)
+
+#read in Joubert meta top cpgs 
+#read in cpgs from PMC4833289
+cpgs<-read.csv(paste0(datadir, '/PMC4833289_SuppFile2CpGs.csv'), stringsAsFactors = F, header=F, skip=2)
+#make headers
+header1<-scan(paste0(datadir, '/PMC4833289_SuppFile2CpGs.csv'), nlines=1, what=character(), sep=',')
+header1[7:26]<-c(rep("SSnewborn", 5), rep("SSnewbornCT", 5), rep("SSolder", 5), rep("anynewborn", 5))
+header2<-paste(header1, scan(paste0(datadir, '/PMC4833289_SuppFile2CpGs.csv'), nlines=1, skip=1, what=character(), sep=','), sep='_')
+names(cpgs)<-header2
+
+#four meta-EWASes: SS_newborn (sustained smoking in newborns); SS_newbornCT (sustained smoking in newborns controlling for cell type) SS_older (sustained smoking in older children); any_newborn (any smoking in newborns)
+#pivot_longer to get single column of coefs and a column for the analysis type
+#and group into a list by analysis type
+cpgs<-cpgs %>% pivot_longer(cols = c(matches("_Coef|_SE|_P|_FDR|_Direction")), names_to = c("set", ".value"), names_pattern="(.+)_(.+)")%>%rename_with(~gsub('_', '', .x))%>%mutate(Gene=`Mapped Gene`, estimate=Coef, conf.low=estimate-1.96*SE, conf.high=estimate+1.96*SE, source='Joubert et al.
+2016', std.error=SE, p.value=P, Chr=CHR) %>%mutate(source=paste0(source, ': ', set))%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)%>%filter(set %in% c('SSolder', 'SSnewbornCT'))
+
+myCpGs=all_models%>%filter(str_detect(outcome, 'cg') & predictors=='base model')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%
+    filter(ancestry=='Unstratified')%>%mutate(CpG=gsub( '.*: ', '', outcome), Gene=gsub(':.*', '', outcome), source=paste0('Fragile Families:', age))
+
+top_cpgs=myCpGs$CpG
+
+cpg_all<-rbind(myCpGs%>%filter(CpG %in% top_cpgs), cpgs%>%filter(CpG %in% top_cpgs)%>%dplyr::select(-Chr), topCPGs%>%filter(CpG%in%top_cpgs)%>%mutate_at(c('p.value'), as.numeric))
+
+#geom blank to give good axes limits
+blank_data <- cpg_all %>%dplyr::select(Gene, CpG, estimate, std.error)%>%group_by(Gene)%>%mutate(min=min(estimate), max=max(as.numeric(std.error), na.rm=T))%>%dplyr::select(-estimate, -std.error)%>%unique()%>%mutate(estimate=min*-1+max, conf.high=estimate, conf.low=estimate)%>%dplyr::select(-max, -min)%>%arrange(Gene)%>%mutate(source='Wiklund et al.
+2019')
+
+blank_data2 <- cpg_all %>%dplyr::select(Gene, CpG, estimate, std.error)%>%group_by(Gene)%>%mutate(min=min(estimate))%>%dplyr::select(-estimate, -std.error)%>%unique()%>%mutate(estimate=min*-1+0.005, conf.high=estimate, conf.low=estimate)%>%dplyr::select(-min)%>%arrange(Gene)%>%mutate(source='Fragile Families: Age_9')
+
+plot_aprioriCpG<-ggplot(cpg_all, aes(x=estimate, xmin=conf.low, xmax=conf.high, color=source, y=CpG))+geom_point(position=position_dodge(width=1))+geom_errorbarh(height=0.1, position=position_dodge(width=1))+scale_x_continuous()+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+facet_wrap(.~Gene, scales='free', ncol=2)+theme_classic()+theme(legend.position = 'bottom')+guides(color=guide_legend(nrow=4,byrow=TRUE))+geom_blank(data = blank_data, aes(x = estimate))+scale_color_discrete('', labels=c('Fragile families saliva: age 15', 'Fragile families saliva: age 9', 'Fragile families: mixed effect model', 'Joubert et al.
+2016 newborn cord blood', 'Joubert et al.
+2016 peripheral blood older children', 'Wicklund et al.
+peripheral blood >16 years'))
+
+ggexport(plot_aprioriCpG, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=9)
+sup_fig=sup_fig+1
+```
+```{r, eval=stand_alone, fig.width=10}
+plot_aprioriCpG
 ```
@@
   ggplot(aes(x=spec, y=sens, color=childteen, linetype=childteen))+geom_line(size=2)+
-  facet_wrap(~methylation, labeller = labeller(methylation=c('SSnewbornCT_center'='Polymethylation score, \n newborn mean centered')))+
+  facet_wrap(~methylation, labeller = labeller(methylation=c('SSnewbornCT_center_scale'='Polymethylation score, \n newborn mean centered')))+
   theme_classic()+theme(legend.position = 'right')+guides(color=guide_legend(nrow=4,byrow=TRUE))+
@@
 
-auc_fig=cowplot::plot_grid(fig4a, fig4b, labels=c('A', 'B'), nrow=1, rel_widths=c(0.65, 0.35), align='h', axis='t')
+auc_fig=cowplot::plot_grid(fig4a, fig4b, labels=c('A', 'B'), nrow=1, rel_widths=c(0.65, 0.35), align='h', axis='tb')
 
@@
 
-# Supplemental Table with AUC and p-value comparison to sustained smoking, newborn cell type controls polymethylation score
+# Supplemental Table: Area under the curve for models shown in main figure 4 for classifying prenatal maternal smoke exposure, with *P* value when comparing to base model only or polymethylation score
 
@@
 roc_base=roc_table %>% filter(modeltype=='Base model+No methylation')%>%mutate(roc_basemodel=roc)%>%dplyr::select(childteen, roc_basemodel)
-roc_ss=roc_table %>% filter(modeltype=='Sustained smoking w/ cell type correction (newborn)')%>%mutate(roc_pms=roc)%>%dplyr::select(childteen, roc_pms)
-
-roc_table1=roc_table%>%filter(!str_detect(modeltype, 'Base model'))%>%left_join(roc_ss)%>%mutate(roctest=map2(roc, roc_pms, ~roc.test(.x, .y)))%>%mutate(auc=round(auc, 2))
-
-rct1=roc_table1%>%mutate(`pval comparison to sustained smoking score (newborns)`=map_dbl(roctest, ~round(.x$p.value, 4)))%>%dplyr::select(childteen, modeltype, auc, `pval comparison to sustained smoking score (newborns)`)
+roc_ss=roc_table %>% filter(modeltype=='Sustained smoking polymethylation score (newborns, cell-type controlled)')%>%mutate(roc_pms=roc)%>%dplyr::select(childteen, roc_pms)
+
+rct1=roc_table%>% 
+  filter(methylation %in% c('globalmethylation', 'cg05575921', 'SSnewbornCT_center_scale', ''))%>%
+  left_join(roc_base)%>%
+  mutate(roctest_base=map2(roc, roc_basemodel, ~roc.test(.x, .y)))%>%
+  left_join(roc_ss)%>%
+  mutate(roctest_ss=map2(roc, roc_pms, ~roc.test(.x, .y)))%>%
+  mutate(auc=round(auc, 2))%>%
+  mutate(`P value vs polymethylation score`=map_dbl(roctest_ss, ~round(.x$p.value, 4)))%>%
+  mutate(`P value vs base model`=map_dbl(roctest_base, ~round(.x$p.value, 4)))%>%
+  dplyr::select(childteen, modeltype, auc, `P value vs base model`, `P value vs polymethylation score`)
 
@@
 ```{r, eval=stand_alone}
-rct1%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value (vs polymethylation score, newborns, cell type controlled)'))
-```
-
-# Supplemental Table with AUC and p-value comparison to base model with no methylation variables (i.e., using only child sex, income to poverty ratio, saliva/epithelial cell proportions and sample plate)
-
-```{r}
-
-roc_table2=roc_table%>%filter(str_detect(modeltype, 'Base model'))%>%left_join(roc_base)%>%mutate(roctest=map2(roc, roc_basemodel, ~roc.test(.x, .y)))%>%mutate(auc=round(auc, 2))
-
-rct2=roc_table2%>%mutate(`pval comparison to base model`=map_dbl(roctest, ~round(.x$p.value, 4)))%>%dplyr::select(childteen, modeltype, auc, `pval comparison to base model`)
-
-rct2%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value (vs base w/ no methyl)'))
-
+rct1%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value vs base model', 'P value vs polymethylation score'))
+```
+
+# Supplemental Table: Area under the curve when using different coefficients to construct polymethylation scores for predicting prenatal maternal smoke exposure
+
+```{r} 
+rct2=roc_table%>% 
+    filter(str_detect(modeltype, 'smoking|No methylation'))%>%
+    left_join(roc_base)%>%
+    mutate(roctest_base=map2(roc, roc_basemodel, ~roc.test(.x, .y)))%>%
+    left_join(roc_ss)%>%
+    mutate(roctest_ss=map2(roc, roc_pms, ~roc.test(.x, .y)))%>%
+    mutate(auc=round(auc, 4))%>%
+    mutate(`P value vs polymethylation score with cell-type control`=map_dbl(roctest_ss, ~round(.x$p.value, 4)))%>%
+    mutate(`P value vs base model`=map_dbl(roctest_base, ~round(.x$p.value, 4)))%>%
+    dplyr::select(childteen, modeltype, auc, `P value vs base model`, `P value vs polymethylation score with cell-type control`)
 ```
@@
 ```{r, eval=stand_alone}
-rct2%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value (vs base w/ no methyl)'))
-```
-
-# Question for group: add EWAS results to this paper
-
--   Arguments in favor: recent paper in Clinical Epigenetics included both
--   Arguments against: enough here for a compelling story; deadline of December 1st; submit and if asked for we have it available.
-
-```{r, eval=include_ewas}
-
-calcLambda<-function(df, type='lm'){
-  if(type=='lme'){df=df%>%mutate(P.Value=p.value)}
-  o<- -log10(sort(df$P.Value, decreasing = F))
-  e<- -log10(ppoints(length(df$P.Value)))
-  lambda<-median(o)/median(e)
-  return(lambda)
-}
-```
-
-```{r, eval=include_ewas}
-child_m1=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'child1_TopHits.RDS'))
-teen_m1=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'teen1_TopHits.RDS'))
-```
-
-```{r, include=include_ewas}
-#lme_base=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'lme_ewas_base.RDS'))
-#lme_smk = lme_base %>% filter(term=='smkPreg_binaryYes')%>%mutate(p.bh=p.adjust(p.value, 'hochberg'))
-#lme_smk=add_annotations(lme_smk, type='lme')
-lme_smk=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'lme_smoking.RDS'))
-#mutate columns to get matching columns in lme_smk as in child & teen _m1
-lme_smk = lme_smk %>% mutate(logFC=estimate, CI.L=conf.low, CI.R=conf.high, P.Value=p.value, adj.P.Val=p.bh)
-```
-
-```{r, eval=F}
-library(variancePartition)
-dream=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'dream_ewas.RDS'))
-dream_smk=topTable(dream, coef='smkPreg_binaryYes', number=nrow(dream$p.value))
-#dream_smk=add_annotations(dream_smk)
-```
-
-## QQ plots
-
-```{r, eval=include_ewas}
-
-png(file=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.png')))
-par(mfrow=c(3, 1), mar=c(5.1, 4.1, 5.1, 2.1))
-  qqman::qq(child_m1$P.Value, main='Age 9: Base model EWAS')
-  mtext(paste0(expression(lambda), '=', round(calcLambda(child_m1), 3)), side=3)
-
-  qqman::qq(teen_m1$P.Value, main='Age 15: Base model EWAS')
-  mtext(paste0(expression(lambda), '=', round(calcLambda(teen_m1), 3)), side=3)
-
-  qqman::qq(lme_smk$P.Value, main='Mixed effect models: Base model EWAS')
-  mtext(paste0(expression(lambda), '=', round(calcLambda(lme_smk), 3)), side=3)
-dev.off()
-
-```
-
-```{r, eval=stand_alone, eval=include_ewas}
-par(mfrow=c(3, 1), mar=c(5.1, 4.1, 5.1, 2.1))
-  qqman::qq(child_m1$P.Value, main='Age 9: Base model EWAS')
-  mtext(paste0(expression(lambda), '=', round(calcLambda(child_m1), 3)), side=3)
-
-  qqman::qq(teen_m1$P.Value, main='Age 15: Base model EWAS')
-  mtext(paste0(expression(lambda), '=', round(calcLambda(teen_m1), 3)), side=3)
-
-  qqman::qq(lme_smk$P.Value, main='Mixed effect models: Base model EWAS')
-  mtext(paste0(expression(lambda), '=', round(calcLambda(lme_smk), 3)), side=3)
-
-sup_fig=sup_fig+1
-```
-
-## Beta vs beta plot: Age 9 vs age 15
-
-```{r, cache=T, eval=include_ewas}
-library(tibble)
-base_models=list('Age_9'=child_m1%>%rownames_to_column(var='CPG'), 'Age_15'=teen_m1%>%rownames_to_column(var='CPG'), "Mixed_effect_model"=lme_smk)%>%Map(cbind, ., ModelType=names(.))
-
-beta_base<-lapply(base_models, function(x) x %>%dplyr::select(CPG, logFC, ModelType)%>%pivot_wider(id_cols = CPG, values_from=logFC, names_from=ModelType))%>%purrr::reduce(left_join, by='CPG')
-
-smooth_betas<-function(df, x, y){
-  ggplot(df, aes_string(x=x, y=y))+
-    geom_point()+geom_smooth(method='lm')+
-    stat_cor()+theme_classic()+
-    scale_y_continuous(name=y)+
-   scale_x_continuous(name=x)+
-    coord_fixed()
-}
-
-sig_beta_m1<-child_m1%>%rownames_to_column(var='CPG')%>%filter(P.Value<10e-4)%>%pull(CPG)
-
-beta_compare=ggarrange( 
-  smooth_betas(beta_base, "Age_9", "Age_15"),
-  smooth_betas(beta_base %>%filter(CPG %in% sig_beta_m1), 'Age_9', 'Age_15'), 
-  labels=c('A', 'B'), align='h')
-
-beta_compare
-ggexport(beta_compare, filename = file.path(outdir, paste0('SupplementalFigure', sup_fig, '.png'))) 
-```
-
-```{r, eval=stand_alone, eval=include_ewas}
-beta_compare
-```
-
-## Upset plots
-
-```{r, eval=include_ewas}
-library(ggupset)
-sig_cpg_list<-map(base_models, ~.x %>%filter(P.Value<10e-4)%>%dplyr::select(CPG, ModelType))%>%
-  bind_rows()%>%
-  group_by(CPG)%>%summarise(Models=list(ModelType))
-
-upset_cpgs=ggplot(sig_cpg_list, aes(x=Models))+
-    geom_bar() +
-    scale_x_upset()
-ggexport(upset_cpgs, filename = file.path(outdir, paste0('SupplementaalFigure', sup_fig, '.pdf')))
-sup_fig=sup_fig+1
-```
-
-## Volcano plot
-
-```{r, eval=include_ewas}
-volcs=ggarrange(
-ggplot(child_m1%>%mutate(adjPval=ifelse(adj.P.Val<0.05, '<0.05', ifelse(adj.P.Val<0.2, '<0.2', '>0.2'))), aes(x=logFC, y=-log10(P.Value), color=adjPval))+geom_point()+scale_x_continuous('Beta (regression coefficient)')+
-  scale_color_manual('Benjamini-Hochberg adjusted P value', values=c('red', 'orange', 'black'))+
-  scale_y_continuous(limits=c(0, 9), breaks=c(0, 2, 4, 6, 8))+
-  scale_x_continuous(limits=c(-0.06, 0.06), breaks=c(-0.05, 0, 0.05)),
-ggplot(teen_m1%>%mutate(adjPval=ifelse(adj.P.Val<0.05, '<0.05', ifelse(adj.P.Val<0.2, '<0.2', '>0.2'))), aes(x=logFC, y=-log10(P.Value), color=adjPval))+geom_point()+scale_x_continuous('Beta (regression coefficient)')+
-  scale_color_manual('Benjamini-Hochberg adjusted P value', values=c('red', 'orange', 'black'))+
-  scale_y_continuous(limits=c(0, 9), breaks=c(0, 2, 4, 6, 8))+
-  scale_x_continuous(limits=c(-0.06, 0.06), breaks=c(-0.05, 0, 0.05)),
-ggplot(lme_smk%>%mutate(adjPval=ifelse(adj.P.Val<0.05, '<0.05', ifelse(adj.P.Val<0.2, '<0.2', '>0.2'))), aes(x=logFC, y=-log10(P.Value), color=adjPval))+geom_point()+scale_x_continuous('Beta (regression coefficient)')+
-  scale_color_manual('Benjamini-Hochberg adjusted P value', values=c('red', 'orange', 'black'))+
-  scale_y_continuous(limits=c(0, 9), breaks=c(0, 2, 4, 6, 8))+
-  scale_x_continuous(limits=c(-0.06, 0.06), breaks=c(-0.05, 0, 0.05)),
-labels=c('A', 'B', 'C'), 
-common.legend = TRUE, 
-nrow=1
-)
-
-ggexport(volcs, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')))
-ggexport(volcs, filename=file.path(outdir, paste0('Figure', main_fig, '.png')))
-```
-
-```{r, eval=stand_alone, fig.width=14, }
-volcs
-```
-
-## Gap hunter
-
-```{r, eval=include_ewas}
-all_gaps=readRDS(file.path(datadir, 'CreatedData', 'EWAS', 'gaphunter_gaps.RDS'))
-```
-
-## Compare across models and studies
-
-```{r, eval=include_ewas}
-topCPGs<-read.csv(paste0(datadir, "/TopCpGsWiklundEtAl2019.csv"))
-topCPGs<-topCPGs %>% mutate(estimate=as.numeric(gsub("−\\s", "-", gsub(" .*", "", β..SE.))), std.error=as.numeric(sub("\\).*", "", sub(".*\\(", "", β..SE.))), p.value=P.value, source="Wiklund et al.
-2019", conf.low=(estimate-1.96*std.error), conf.high=(estimate+1.96*std.error), set ='')%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)
-
-#read in Joubert meta top cpgs 
-#read in cpgs from PMC4833289
-cpgs<-read.csv(paste0(datadir, '/PMC4833289_SuppFile2CpGs.csv'), stringsAsFactors = F, header=F, skip=2)
-#make headers
-header1<-scan(paste0(datadir, '/PMC4833289_SuppFile2CpGs.csv'), nlines=1, what=character(), sep=',')
-header1[7:26]<-c(rep("SSnewborn", 5), rep("SSnewbornCT", 5), rep("SSolder", 5), rep("anynewborn", 5))
-header2<-paste(header1, scan(paste0(datadir, '/PMC4833289_SuppFile2CpGs.csv'), nlines=1, skip=1, what=character(), sep=','), sep='_')
-names(cpgs)<-header2
-
-#four meta-EWASes: SS_newborn (sustained smoking in newborns); SS_newbornCT (sustained smoking in newborns controlling for cell type) SS_older (sustained smoking in older children); any_newborn (any smoking in newborns)
-#pivot_longer to get single column of coefs and a column for the analysis type
-#and group into a list by analysis type
-cpgs<-cpgs %>% pivot_longer(cols = c(matches("_Coef|_SE|_P|_FDR|_Direction")), names_to = c("set", ".value"), names_pattern="(.+)_(.+)")%>%rename_with(~gsub('_', '', .x))%>%mutate(Gene=`Mapped Gene`, estimate=Coef, conf.low=estimate-1.96*SE, conf.high=estimate+1.96*SE, source='Joubert et al.
-2016', std.error=SE, p.value=P, Chr=CHR) %>%mutate(source=paste0(source, ': ', set))%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)%>%filter(set %in% c('SSolder', 'SSnewbornCT'))
-
-#myCPGs<-AllModelsdf %>%filter(Model=='model1')%>%mutate(data=map(data, function(x) x %>%filter(CpG %in% sigCpGs)))%>%unnest(cols=c(data))%>%mutate(Joubertetal=ifelse(CpG %in% cpgs$CpG, 'in Joubert et al.
-2016', 'Not in Joubert et al.
-2016'), Wicklundetal=ifelse(CpG %in% topCPGs$CpG, 'in Wicklund et al.
-2019', 'Not in Wicklund et al.
-2019'))%>%mutate(Gene=gene, estimate=logFC, conf.low=CI.L, conf.high=CI.R, source=paste0('Fragile Families:', Age), std.error='NA', p.value=P.Value, Chr=chr, set='Model1', Position=pos)
-
-```
-
-```{r, fig.width=14, eval=include_ewas}
-comparison=left_join(
-  lme_smk%>%dplyr::select(CPG, gene, estimate, p.value)%>%dplyr::rename_with(~paste0('FF_', .x), .cols=c(estimate, p.value)), 
-  cpgs%>%filter(set=='SSnewbornCT')%>%mutate(CPG=CpG)%>%dplyr::select(CPG, Gene, estimate, p.value)%>%dplyr::rename_with( ~paste0('Joubert_', .x), .cols=c(estimate, p.value))
-  )%>%filter(!is.na(Joubert_estimate))
-
-comp_nofilter=ggplot(comparison, aes(x=Joubert_estimate, y=FF_estimate, color=FF_p.value))+
-  geom_point()+stat_cor()+
-  scale_y_continuous('Estimate in Fragile Families')+
-  scale_x_continuous('Estimate in Joubert et al. metaanalysis')+
-  scale_color_continuous('P value in Fragile Families')
-
-#comparison %>% mutate(s1=sign(FF_estimate), s2=sign(Joubert_estimate), agreement=case_when(s1==s2~'Direction consistent', s1<s2~'FF -, Joubert +', s1>s2~'FF+, Joubert-'))%>%group_by(agreement)%>%summarise(n=n())
-
-comp_filter=ggplot(comparison%>%filter(FF_p.value<10e-4), aes(x=Joubert_estimate, y=FF_estimate, color=FF_p.value, label=ifelse(abs(Joubert_estimate)>0.04 , gene, NA)))+
-  geom_point()+stat_cor()+
-  geom_text_repel()+
-  scale_y_continuous('Estimate in Fragile Families')+
-  scale_x_continuous('Estimate in Joubert et al. metaanalysis')+
-  scale_color_continuous('P value in Fragile Families')
-
-comp_joubert=ggarrange(comp_nofilter, comp_filter, common.legend = F, labels=c('A', 'B'))
-ggexport(comp_joubert, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.png')))
-```
-
-```{r, eval=stand_alone, fig.width=14}
-comp_joubert
-```
-
-## Additional covariate control and ancestry stratified models at top hits
-
-```{r, eval=include_ewas}
-top_cpgs = map(base_models, ~.x %>%filter(adj.P.Val<0.05)%>%dplyr::select(CPG, ModelType, gene))%>%bind_rows()%>%mutate(label=paste0(gene, ': ', CPG))
-top_cpg_gaps=top_cpgs[top_cpgs$CPG %in% rownames(all_gaps$proberesults),]
-top_cpgs=top_cpgs[!(top_cpgs$CPG %in% rownames(all_gaps$proberesults)),]
-top_cpg_groups=top_cpgs %>%group_by(CPG)%>%dplyr::summarise(models=list(ModelType))
-top_cpg_names=c(top_cpgs%>%pull(label)%>%unique(), 'AHRR: cg05575921', 'MYO1G: cg22132788')
-top_cpgs=c(top_cpgs %>% pull(CPG)%>%unique(), 'cg05575921', 'cg22132788')
-
-betaqc<-readRDS(file=file.path(datadir, 'OGData', "betaqc.rds"))
-my_beta=as.data.frame(t(betaqc[row.names(betaqc)%in% top_cpgs, ])) %>% mutate(MethID=row.names(.))
-modeldata=left_join(modeldata, my_beta)
-```
-
-```{r, eval=F}
-y_vector<-top_cpgs
-outcome_labels=top_cpg_names
-names(outcome_labels)=y_vector
-
-#model variables (excepting PCs, add in individually for global & ancestry specific models)
-base_model_vars<-"~smkPreg_binary+cm1bsex+cm1inpov+Leukocytes_saliva+Epithelial.cells_saliva+Sample_Plate"
-prenatal_model_vars<-paste0(base_model_vars, "+m1g2_YesNoPreg+m1g3_YesNoPreg")
-secondhand_model_vars<-paste0(prenatal_model_vars, '+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth')
-interaction_model_vars<-paste0(secondhand_model_vars, '+ChildAgeComposite:smkPreg_binary')
-#pc variables
-global_pcs='+global_PC1+global_PC2'
-local_pcs='+local_PC1+local_PC2'
-
-#surrogate variable models
-surrogate_model_vars<-paste0('~smkPreg_binary+', 
-                             paste(colnames(modeldata)[str_detect(colnames(modeldata), 'sv')], collapse='+'))
-
-predictors=c('base model'=base_model_vars, 
-             'prenatal exposures model'=prenatal_model_vars, 
-             'secondhand smoke exposure model'=secondhand_model_vars)
-
-predictors_interaction=c(predictors, 'interaction model'=interaction_model_vars)
-
-model_labels=c('base model', 'prenatal exposures model', 'secondhand smoke exposure model', 'surrogate variable models')
-
-model_labels_interaction=c('base model', 'prenatal exposures model', 'secondhand smoke exposure model', 'interaction model (age*smoke)', 'surrogate variable models')
-
-#stratification variables 
-##############################
-age_labels=c('Age 9', 'Age 15')
-age_vector=modeldata$childteen%>%unique()
-names(age_labels)=age_vector
-ancestry_vector=modeldata$ancestry%>%unique()
-
-#################################Source functions#######################################
-model<-function(df, y, x){
-  lm(formula(paste(y, x)), data=df)
-}
-
-model_lmm<-function(df, y, x){
-  xnew=paste0(x, '+', '(1|id)')
-  lmer(formula(paste(y, xnew)), data=df, REML=F)
-}
-
-my_clean_argnames=function(df, my_preds, childteen=T, model_labs=model_labels){
-  names(model_labs)=my_preds
-  df=df %>% mutate(predictors=recode_factor(predictors, !!!model_labs),
-                outcome=recode_factor(outcome, !!!outcome_labels))
-  if(childteen==T){df=df%>%mutate(age=recode_factor(childteen, !!!age_labels))} 
-  return(df)
-}
-
-#################################Cross sectional models###########################
-#########################Global
-#set predictor variables & arguments
-global_predictors=c(paste0(predictors, global_pcs), surrogate_model_vars)
-args=list('childteen'=age_vector, 'outcome'=top_cpgs, 'predictors'=global_predictors)%>%cross_df()
-
-#run models
-global_models=modeldata%>%group_by(childteen)%>%nest()%>%left_join(args)%>%
-  mutate(model=pmap(list(data, outcome, predictors), model), 
-         tidied=map(model, tidy, conf.int=T), 
-         glanced=map(model, glance))
-
-#clean argument names for plotting
-global_models=my_clean_argnames(global_models, global_predictors)
-
-#########################Local
-#arguments
-local_predictors=c(paste0(predictors, local_pcs), surrogate_model_vars)
-args=list('childteen'=c('Age 9', 'Age 15'), 'outcome'=top_cpgs, 'predictors'=local_predictors, 'ancestry'=ancestry_vector)%>%cross_df()
-
-#run models
-local_models=modeldata%>%group_by(childteen, ancestry)%>%nest()%>%left_join(args)%>%
-  mutate(model=pmap(list(data, outcome, predictors), model), 
-         tidied=map(model, tidy, conf.int=T), 
-         glanced=map(model, glance))
-
-#clean argument names for plotting
-local_models=my_clean_argnames(local_models, local_predictors)
-
-#################################Longitudinal models##################
-#########################Global
-#set predictor variables & arguments
-#include a model with an interaction term b/t age and prenatal smoke exposure
-#to determine if there is a age-slope difference b/t exposed & not exposed
-global_predictors=c(paste0(predictors_interaction, global_pcs, '+ChildAgeComposite'), surrogate_model_vars)
-args=list('outcome'=top_cpgs, 'predictors'=global_predictors)%>%cross_df()
-
-#run models
-longitudinal_global_models=args %>% 
-    mutate(model=pmap(list(y=outcome, x=predictors), model_lmm, df=modeldata), 
-           tidied=map(model, tidy, conf.int=T), 
-           glanced=map(model, glance))
-
-#clean argument names
-longitudinal_global_models=my_clean_argnames(longitudinal_global_models, global_predictors, childteen = F, model_labs = model_labels_interaction)
-#########################Local
-#set predictor variables and arguments
-local_predictors=c(paste0(predictors_interaction, local_pcs, '+ChildAgeComposite'), surrogate_model_vars)
-args=list('outcome'=top_cpgs, 'predictors'=local_predictors, 'ancestry'=ancestry_vector)%>%cross_df()
-
-#run models
-longitudinal_local_models=modeldata%>%group_by(ancestry)%>%nest()%>%left_join(args)%>%
-  mutate(model=pmap(list(data, outcome, predictors), model_lmm), 
-         tidied=map(model, tidy, conf.int=T), 
-         glanced=map(model, glance))
-
-#clean argument names
-longitudinal_local_models=my_clean_argnames(longitudinal_local_models, local_predictors, childteen = F, model_labs = model_labels_interaction)
-```
-
-```{r compare over time, eval=F}
-
-rbind(global_models%>%dplyr::select(childteen, predictors, outcome, tidied)%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes'), 
-      longitudinal_global_models%>%dplyr::select(predictors, outcome, tidied)%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%mutate(childteen='Linear mixed model'))%>%
-  filter(predictors %in% c('base model', 'secondhand smoke exposure model'))%>%
-  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=childteen, y=predictors))+
-    geom_point(position = position_dodge(width = 1))+
-    geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
-    facet_grid(.~outcome, scales='free_x', labeller = labeller(outcome_f=top_cpg_names))+
-    geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
-    theme_classic()
-
-cpg_timevariant=top_cpg_groups%>%mutate(count=map_dbl(models, length))%>%filter(count<3)%>%pull(CPG)
-
-age_long_methyl%>%
-  filter(`Methylation summary measure type` %in% top_cpgs)%>%
-  ggplot(aes(x=childteen, y=`Methylation summary measure`, color=childteen))+
-  geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+
-  scale_color_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+
-  scale_x_discrete(name='Visit', labels=c('Age 9', 'Age 15'))+
-  geom_point(alpha=0)+geom_line(alpha=0, aes(group=idnum))+geom_smooth(aes(group=1), method='lm', color='black')+
-  facet_wrap(~Methyl_factor, scales='free')+
-  theme(legend.position = 'bottom')
-
-```
-
-```{r compare over ancestries, eval=F}
-longitudinal_local_models%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%
-  filter(predictors %in% c('base model', 'secondhand smoke exposure model'))%>%
-  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=ancestry, y=predictors))+
-    geom_point(position = position_dodge(width = 1))+
-    geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
-    facet_grid(.~outcome, scales='free_x', labeller = labeller(outcome_f=top_cpg_names))+
-    geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
-    theme_classic()
-
-longitudinal_local_models%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%
-    filter(predictors %in% c('base model'))%>%group_by(outcome)%>%summarise(est=list(estimate))%>%mutate(sign=map_dbl(est, ~sum(sign(.x))))%>%filter(abs(sign)!=3)
-
-```
-
-```{r, fig.width=14, eval=include_ewas}
-
-age_long_methyl=modeldata%>%
-  dplyr::select(any_of(c('idnum', 'childteen', 'ancestry', 'smkPreg_binary', top_cpgs)))%>%
-  pivot_longer(cols=any_of(top_cpgs),
-               names_to='Methylation summary measure type',
-               values_to='Methylation summary measure')%>%
-  mutate(Methyl_factor=factor(`Methylation summary measure type`, 
-                              levels=top_cpgs))
-
-ancestry_differences=age_long_methyl%>%
-    filter(`Methylation summary measure type` %in% top_cpgs)%>%
-    ggplot(aes(x=ancestry, y=`Methylation summary measure`, color=smkPreg_binary))+
-    geom_violin()+geom_boxplot(width=0.2, position=position_dodge(0.9))+stat_compare_means(label="..p.signif..", label.y.npc = 0.95)+
-    scale_color_discrete(name='')+
-    scale_x_discrete()+
-    geom_point(alpha=0)+
-    facet_grid(childteen~Methyl_factor, scales='free')+theme_bw()+
-    theme(legend.position = 'bottom')+
-  theme(axis.text.x = element_text(angle=45, hjust=1))
-
-ancestry_differences
-
-ggexport(ancestry_differences, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=11)
-
-ggexport(ancestry_differences, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.png')))
-```
-
-```{r, eval=stand_alone, fig.width=14}
-ancestry_differences
-```
-
-```{r, eval=include_ewas}
-
-myCPGs<-base_models%>%bind_rows(.id='modeltype')%>%mutate(CpG=CPG, Gene=gene, estimate=logFC, conf.low=CI.L, conf.high=CI.R, source=paste0('Fragile Families:', modeltype), std.error='NA', p.value=P.Value, Chr=chr, set='Model1', Position=pos)%>%dplyr::select(colnames(cpgs))
-
-#cpg_overlap=myCPGs%>%filter(Joubertetal=='in Joubert et al. 2016' | Wicklundetal=='in Wicklund et al. 2019')%>%pull(CpG)%>%unique()
-
-#myCPGs2<-myCPGs %>%ungroup()%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)%>%mutate(CpG=ifelse(CpG %in% row.names(mygaps$proberesults), paste0(CpG, '*'), CpG))
-
-cpg_all<-rbind(myCPGs%>%filter(CpG %in% top_cpgs), cpgs%>%filter(CpG %in% top_cpgs), topCPGs%>%filter(CpG%in%top_cpgs))
-
-#geom blank to give good axes limits
-blank_data <- cpg_all %>%dplyr::select(Gene, CpG, estimate, std.error)%>%group_by(Gene)%>%mutate(min=min(estimate), max=max(as.numeric(std.error), na.rm=T))%>%dplyr::select(-estimate, -std.error)%>%unique()%>%mutate(estimate=min*-1+max, conf.high=estimate, conf.low=estimate)%>%dplyr::select(-max, -min)%>%arrange(Gene)%>%mutate(source='Wiklund et al.
-2019')
-
-blank_data2 <- cpg_all %>%dplyr::select(Gene, CpG, estimate, std.error)%>%group_by(Gene)%>%mutate(min=min(estimate))%>%dplyr::select(-estimate, -std.error)%>%unique()%>%mutate(estimate=min*-1+0.005, conf.high=estimate, conf.low=estimate)%>%dplyr::select(-min)%>%arrange(Gene)%>%mutate(source='Fragile Families: Age_9')
-
-plot_sharedCpG<-ggplot(cpg_all, aes(x=estimate, xmin=conf.low, xmax=conf.high, color=source, y=CpG))+geom_point(position=position_dodge(width=1))+geom_errorbarh(height=0.1, position=position_dodge(width=1))+scale_x_continuous()+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+facet_wrap(.~Gene, scales='free', ncol=2)+theme_classic()+theme(legend.position = 'bottom')+guides(color=guide_legend(nrow=4,byrow=TRUE))+geom_blank(data = blank_data, aes(x = estimate))+scale_color_discrete('', labels=c('Fragile families saliva: age 15', 'Fragile families saliva: age 9', 'Fragile families: mixed effect model', 'Joubert et al.
-2016 newborn cord blood', 'Joubert et al.
-2016 peripheral blood older children', 'Wicklund et al.
-peripheral blood >16 years'))
-
-ggexport(plot_sharedCpG, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')))
-```
-
-```{r, eval=stand_alone, fig.width=14}
-plot_sharedCpG
-```
+rct2%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value vs base model', 'P value vs polymethylation score with cell-type control'))
+```

### 2021-11-04 Scripts/01_clean_phenotype_data.R

@@ -199,9 +199,11 @@ secondhandsmokevars<-c(prenatalexposurevar, "PostnatalMaternalSmokingAny", "SmkA
 secondhandsmkdata<-prenatalexdata%>%filter_at(all_of(secondhandsmokevars), all_vars(!(. %in% c("Missing")) & !is.na(.)))
 
 child_smoke<-secondhandsmkdata %>% filter(k5f1l!= "2 no" & childteen=='C')
-child_nosmoke<-secondhandsmkdata %>% filter(k5f1l=="2 no" & childteen=='C')
+#only children who report never cigarette use at age 9 OR were missing never cigarette at age 9 but reported no smoking at age 15
+child_nosmoke<-secondhandsmkdata %>% filter((k5f1l=="2 no" | (is.na(k5f1l) & k6d40=='2 No')) & childteen=='C')
 teen_smoke<-secondhandsmkdata %>% filter(k6d40!="2 No" & childteen=='T')
-teen_nosmoke<-secondhandsmkdata %>% filter(k6d40=="2 No" & k5f1l!='1 yes' & childteen=='T')
+#only children who report never smoking at age 15 AND were either missing or reported not smoking at age 9
+teen_nosmoke<-secondhandsmkdata %>% filter(k6d40=="2 No" & (k5f1l=='2 no' |  is.na(k5f1l)) & childteen=='T')
 completecase<-rbind(child_nosmoke, teen_nosmoke)%>%copy_labels(basemodeldata)%>%copy_labels(myFF)
 
 fathersmkdata<-completecase %>% filter(f1g4!="Missing" & !is.na(f1g4))

### 2021-11-04 Scripts/02_make_methyl_summary.R

@@ -121,6 +121,7 @@ methyl_labels_myclocks=c('Methylation data ID', 'Epithelial cell proportion', 'F
                          'Immune cell proportion (saliva)','Epithelial cell proportion (saliva)', 
                          'Global methylation', 'AHHR: cg05575921', 'MYO1G: cg04180046', 'CYP1A1: cg05549655', 'GFI1: cg14179389', "MYO1G: cg22132788", 
                          pms_labels, gsub('no transform', 'z-score standardized', pms_labels), gsub('no transform', 'mean-centered', pms_labels), 
+                         #paste0(pms_labels, '/n (coefficients) & z-score standardized (score)'), gsub('no transform', '/n z-score standardized (coefficients) & z-score standardized (score)', pms_labels), gsub('no transform', '/n mean-centered (coefficients) & z-score standardized (score)', pms_labels),
                          'ID', 'Visit',
                          'Horvath clock', 'SkinBlood clock', 'Pediatric clock', 'PoAm clock 38', 'PoAm clock 45')
 
@@ -128,6 +129,7 @@ methyl_labels_jonahclocks=c('Methylation data ID', 'Epithelial cell proportion',
                             'Immune cell proportion (saliva)','Epithelial cell proportion (saliva)', 
                             'Global methylation', 'AHHR: cg05575921', 'MYO1G: cg04180046', 'CYP1A1: cg05549655', 'GFI1: cg14179389', "MYO1G: cg22132788", 
                             pms_labels, gsub('no transform', 'z-score standardized', pms_labels), gsub('no transform', 'mean-centered', pms_labels), 
+                            #paste0(pms_labels, '/n (coefficients) & z-score standardized (score)'), gsub('no transform', '/n z-score standardized (coefficients) & z-score standardized (score)', pms_labels), gsub('no transform', '/n mean-centered (coefficients) & z-score standardized (score)', pms_labels),
                             'ID', 'Visit',
                             'Horvath clock', 'SkinBlood clock', 'Hannum clock', 'Pediatric clock', 'Levine clock', 'PoAm clock 38', 'PoAm clock 45', 'GRIM clock', 'GRIM pack/yrs component')
 
@@ -148,6 +150,12 @@ summary(methyldata)
 load(paste0(datadir, '/CreatedData/completeCasepheno.Rdata'))
 completecase=left_join(completecase, methyldata%>%mutate(idnum=as.character(idnum)))%>%
[…]

### 2021-11-04 Scripts/04_summary_methyl_models.R

@@ -24,7 +24,6 @@ library(lmerTest)
 library(stringr)
 library(sva)
 #############
-#investigate why clocks have 1 fewer observation in child models
 ##################################Directories & data
 datadir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/CreatedData"
 outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output"
@@ -57,12 +56,16 @@ save(modeldata, file=file.path(datadir, file='completeCaseSVA.Rdata'))
 #################################Set variables for modeling
 #outcomes & outcome_labels
 
-y_vector<-c("globalmethylation", "pediatric", "anynewborn_center", "SSnewbornCT_center", "SSolder_center", "cg05575921")
+y_vector<-c("globalmethylation", "pediatric", "anynewborn_center_scale", 
+            "SSnewbornCT_center_scale", "SSnewborn_center_scale",  "SSolder_center_scale", 
+            "cg05575921", "cg04180046", "cg05549655", "cg14179389","cg22132788")
 outcome_labels=c('Global methylation', 'Pediatric clock',  
                  'Any smoking polymethylation score (newborns)', 
                  'Sustained smoking polymethylation score (newborns, cell-type controlled)', 
+                 'Sustained smoking polymethylation score (newborns)',
                  'Sustained smoking polymethylation score (older children)', 
-                 'AHRR: cg05575921')
+                 'AHRR: cg05575921', "MYO1G: cg04180046", "CYP1A1: cg05549655" ,  "GFI1: cg14179389" , "MYO1G: cg22132788")
+
 names(outcome_labels)=y_vector
 
 
@@ -191,16 +194,12 @@ modeldata=modeldata%>%mutate(smkPreg_binaryN=case_when(smkPreg_binary=='Yes'~1, 
 
 #global predictors, cross sectional
 roc_predictors=c('', gsub('~smkPreg_binary', '', paste0(base_model_vars, global_pcs)))
-roc_outcomes=c('', y_vector, 'cg05549655', 'cg22132788', 'SSnewborn_center')
+roc_outcomes=c('', y_vector)
 args=list('childteen'=age_vector, 'methylation'=roc_outcomes, 'covariates'=roc_predictors)%>%cross_df()%>%mutate(predictors=paste0(methylation, covariates))%>%
   filter(predictors!='')
 
 
-methyl_names=data.frame(name_methyl=c('No methylation', 'Global methylation', 'Pediatric clock', 
[…]

### 2021-11-05 Scripts/02_make_methyl_summary.R
added mean methylation at different regions (islands shores etc)

@@ -9,6 +9,7 @@ library(EpiDISH)
 library(ewastools)
 library(tidyr)
 library(purrr)
+library(tibble)
 library(sjlabelled)
 library(wateRmelon)
 library(data.table)
@@ -40,6 +41,21 @@ cell_types=left_join(estF_FF, estL_FF)
 ################################################global methylation
 globalmethy<-colMeans(betaqc, na.rm=T)
 globalmethdf<-data.frame("MethID"=names(globalmethy), "globalmethylation"=globalmethy)
+#add annotations for stratified methylation mean calculations
+library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
+data(Locations)
+data(Islands.UCSC)
+data(Other)
+Locations <- Locations[rownames(betaqc), ]
+Islands.UCSC <- Islands.UCSC[rownames(betaqc),]
+Other <- Other[rownames(betaqc),]
+#list of regions
+isles=Islands.UCSC%>%as.data.frame()%>%mutate(Relation_to_Island=gsub('._', '', Relation_to_Island))%>%rownames_to_column('cpgs')%>%group_by(Relation_to_Island)%>%summarise(cpgs=list(cpgs))
+#mean methylation per region
+methyMean_regions=lapply(isles$cpgs, function(x) as.data.frame(colMeans(betaqc[x, ], na.rm=T)))
+methyMean_regions=methyMean_regions%>% bind_cols()%>%rownames_to_column('MethID')
+colnames(methyMean_regions)=c('MethID', paste0(isles$Relation_to_Island, '_mean'))
+globalmethdf=left_join(globalmethdf, methyMean_regions)
 
 #################################################apriori cpgs
 aprioriCGdf<-as.data.frame(t(betaqc[aprioriCG, ]))
@@ -119,7 +135,8 @@ pms_labels=c('Polymethylation score: coefficients for any smoking from newborn c
 
 methyl_labels_myclocks=c('Methylation data ID', 'Epithelial cell proportion', 'Fibroblast cel proportion', 'Immune cell proportion',
                          'Immune cell proportion (saliva)','Epithelial cell proportion (saliva)', 
-                         'Global methylation', 'AHHR: cg05575921', 'MYO1G: cg04180046', 'CYP1A1: cg05549655', 'GFI1: cg14179389', "MYO1G: cg22132788", 
[…]

### 2021-11-09 Papers/Polymethylation_scores/05_results.Rmd
added figure captions

@@
 Children of mothers who reported smoking during pregnancy were more likely to be of European genetic ancestry (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[2, 2], 2)*100`%) than children of mothers who did not (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[1, 2], 2)*100`%, *P* value=`r round(with(modeldata_people, chisq.test(table(smkPreg_binary, ancestry)))$p.value, 4)`).
-Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both the age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) visits (Table 1).
+Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both the age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) visits (Table 1, Supplemental Table2).
 At age 9, children exposed to prenatal maternal smoking had lower DNA methylation at AHRR:cg05575921 (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)*100`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)*100`%, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
@@
 A consistent association was observed when stratifying to the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`).
-In the African genetic ancestry sample at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` higher polymethylation score.
+In the African genetic ancestry sample at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.
 The direction and magnitude of the association was similar in the European and Hispanic genetic ancestry samples at age 9, though attenuated at age 15 (Figure 3; Supplemental Tables 2).
@@
 
-Results were also similar when using polymethylation scores constructed using regression coefficients (see Methods, Supplemental Figure 7).
+Results were also similar when using polymethylation scores constructed using alternative regression coefficients (see Methods, Supplemental Figure 7).
 The association between prenatal maternal smoking and the polymethylation score using regression coefficients from newborn cord blood with cell-type controls was the strongest.
-Prenatal maternal smoking also associated with other selected *a priori* CpG sites consistently with associations reported in previous meta-analyses (Supplemental Figure 8).
+Prenatal maternal smoking also associated with other selected *a priori* CpG sites consistently with associations reported in previous meta-analyses (Supplemental Table 2, Supplemental Figure 8).
 

### 2021-11-09 Papers/Polymethylation_scores/08_methods.Rmd
added figure captions

@@
 We calculated polymethylation scores using regression coefficients from 4 different regressions.
-For our main analysis we used coefficients from a regression of sustained smoking exposure and DNA methylation in newborn cordblood with cell-type control.
-As sensitivity analyses, we used re sustained smoking exposure and DNA methylation in newborn cordblood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children, any smoking exposure and DNA methylation in newborn cordblood) [@Joubert:2016ue].
+For our main analysis we used coefficients from a regression of sustained smoking exposure and DNA methylation in newborn cord blood with cell-type control.
+As sensitivity analyses, we used re sustained smoking exposure and DNA methylation in newborn cord blood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children, any smoking exposure and DNA methylation in newborn cord blood) [@Joubert:2016ue].
+
 Global DNA methylation was calculated for each sample as the mean methylation value of each sample across the cleaned probe set.
+Mean DNA methylation restricted to probes in genomic regions (CpG island, shore, shelf or open sea, as identified in the R package IlluminaHumanMethylation450kanno.ilmn12.hg19 v 0.6.0) was also calculated.
+
 Pediatric epigenetic age was calculated for each sample using the coefficients and methods provided by the creators (see [\<https://github.com/kobor-lab/Public-Scripts\>](https://github.com/kobor-lab/Public-Scripts){.uri}) [@mcewen2019].
-The GRIM age clock, including the smoking pack-years subscale, was calculated as previously described [@lu2019].
+The GRIM age clock, including the smoking pack-years sub-scale, was calculated as previously described [@lu2019].
 
@@
 To evaluate the accuracy of the DNA methylation summary measures as biomarkers of prenatal maternal smoking, we used a receiver operating curve.
-First, we regressed exposure to prenatal maternal smoking (outcome) againts the DNA methylation summary measures in individual logistic regressions, while adjusting for the base model variables listed above.
+First, we regressed exposure to prenatal maternal smoking (outcome) against the DNA methylation summary measures in individual logistic regressions, while adjusting for the base model variables listed above.
 Next, we calculated receiver operating curves (ROC) and area under the curves (AUCs) using the function roc from the R library pROC version 1.18.0.
-We compared ROC curves and AUC using the delong method and the function roc.test from the R library pROC version 1.18.0.
+We compared ROC curves and AUC using the Delong method and the function roc.test from the R library pROC version 1.18.0.
 

### 2021-11-09 Papers/Polymethylation_scores/99_FigureCaptions.Rmd
added figure captions

@@
 
-# Figure 1 - Selection of samples from the Fragile Families and Child Wellbeing study into analytic subset.
+```{r, echo=F, include=FALSE}
+print_figs=TRUE
+```
+
+#### Table 1: Bivariate associations between prenatal maternal smoking (no/yes) and selected DNA methylation summary measures and important covariates among a diverse sample of `r n_total` children in the Fragile Families and Child Wellbeing study
+
+```{r,  eval=print_figs, include=T}
+smoking_tbl
+```
+
+#### Figure 1 - Selection of samples from the Fragile Families and Child Wellbeing study into analytic subset.
 N represents the number of individuals at each step in the selection procedure, M represents the number of samples.
@@
 
-# Figure 2 - Differences in selected DNA methylation summary measures by self-report of prenatal maternal smoking among `r n_total` children in the Fragile Families and Child Wellbeing study (CT=cell-type control)
+```{r, echo=F, eval=print_figs, include=T, fig.height=11, fig.width=7}
+graph1
+```
 
-# Figure 3 - Prental maternal smoke exposure is consistently associated with polymethylation scores at ages 9 and 15 and is portable across genetic ancestry groups in a sample of `r n_total` children in the Fragile Families and Child Wellbeing study
+#### Figure 2 - Differences in selected DNA methylation summary measures by self-report of prenatal maternal smoking among `r n_total` children in the Fragile Families and Child Wellbeing study (CT=cell-type control)
 
-# Figure 4 - A) Polymethylation scores accurately classify maternal report of prenatal smoke exposure status at ages 9 and 15 among a `r n_total` children in the Fragile Families and Child Wellbeing Study B) Inclusion of two-visits of DNA methylation data does not improve classification of maternal report of prenatal smoke exposure status among `r n_total` children in the Fragile Families and Child Wellbeing study
+```{r, echo=F, eval=print_figs, include=T, fig.height=5, fig.width=7}
+fig2
+```
+
+#### Figure 3 - Prental maternal smoke exposure is consistently associated with polymethylation scores at ages 9 and 15 and is portable across genetic ancestry groups in a sample of `r n_total` children in the Fragile Families and Child Wellbeing study
+
+```{r, echo=F, eval=print_figs, include=T, fig.width=12}
+forest_plot
+```
+
+#### Figure 4 - A) Polymethylation scores accurately classify maternal report of prenatal smoke exposure status at ages 9 and 15 among a `r n_total` children in the Fragile Families and Child Wellbeing Study B) Inclusion of two-visits of DNA methylation data does not improve classification of maternal report of prenatal smoke exposure status among `r n_total` children in the Fragile Families and Child Wellbeing study
+
+```{r, echo=F, eval=print_figs, include=T,  fig.width=12}
+auc_fig
+```

### 2021-11-09 Reports/Paper1.Rmd
added figure captions

@@
 ---
-title: "Polymethylation scores results"
-author: "F blostein"
-date: "9/30/2021"
+title: "Supplemental Materials: Polymethylation Scores for Prenatal Maternal Smoke Exposure Persist Until Age 15 and Are Detected in Saliva"
+author: "Freida A. Blostein^1, Jonah Fisher^2, John Dou^1, Lisa Schenper^3, Erin Ware^2 Daniel A. Notterman^3, Colter Mitchell^2, Kelly M. Bakulski^1"
+date: "`9/30/2021`r Sys.Date()`"
 output:
+  pdf_document: default
+  word_document: default
   html_document:
@@
     theme: sandstone
-  pdf_document: default
-  word_document: default
 always_allow_html: true
@@
 
-include_ewas=F
-stand_alone=T
+as_supp=T
+stand_alone=F
 ```
@@
 ```{r}
-modeldata=modeldata %>% mutate(childteen=factor(childteen, levels=c('Age 9', 'Age 15')))
+modeldata=modeldata %>% mutate(childteen=factor(childteen, levels=c('Age 9', 'Age 15')))%>%droplevels()
 set_label(modeldata$childteen)='Visit'
@@
 
-# Figure `r main_fig` - Selection of samples from the Fragile Families and Child Wellbeing (FFCW) study into analytic subset.
-N represents the number of individuals at each step in the selection procedure, M represents the number of samples, as individuals with repeated measures can have more than one sample.
-
-```{r, eval=T}
+```{r figure 1, eval=T}
 #function
@@
 
-```{r, fig.height=12}
+```{r figure 1 make figure, fig.height=12}
 
@@
 ```
-```{r, eval=stand_alone}
+
+```{r figure 1 plot figure, eval=stand_alone}
 graph1
@@
 
-```{r}
+```{r supplemental table exclusions}
 myFF$MethylData2<-ifelse(myFF$MethID %in% modeldata$MethID, 'Final analysis group', 'Not in final analysis group')
@@
 
-```{r, eval=stand_alone}
+```{r exclusion table include, eval=as_supp}
 exclusion_tbl
@@
 
-```{r}
+```{r make supplemental figure correlation plot}
 library(Hmisc)
@@
 
-```{r, eval=stand_alone, fig.width=12, fig.height=12}
+```{r supplemental correlation plot plot it , eval=as_supp, fig.width=12, fig.height=12}
 corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8)
@@
 
-```{r}
+```{r supplemental figure time correlation plot}
 age_long_methyl=modeldata%>%
@@
 
-```{r, eval=stand_alone, fig.width=12, fig.height=12}
+```{r supplemenetal figure time correlation plot, eval=as_supp, fig.width=12, fig.height=12}
 visit_cor
@@
 
-```{r}
+```{r supplemental figure methylation time distributions}
 dist_methyl=age_long_methyl%>%
@@
 
-```{r, eval=stand_alone, fig.height=11}
+```{r supplemental figure methylation time distribution, eval=as_supp, fig.height=11}
 dist_methyl
@@
 
-# Figure `r main_fig`: Differences in selected DNA methylation summary measures by prenatal maternal smoke exposure among `r n_total` US children in the Fragile Families and Child Wellbeing study
-
-```{r change in scores time, fig.width=12}
+```{r main figure 2 change in scores by smoking status, fig.width=12}
 
@@
 
-```{r, eval=stand_alone}
+```{r main figure 2 change in scores smoking status, eval=stand_alone}
 fig2
 ```
-
-# Table `r main_tab`: Bivariate associations between prenatal maternal smoking, selected DNA methylation summary measures and important covariates among a diverse sample of `r n_total` children in the Fragile Families and Child Wellbeing study
 
@@
   dplyr::select(smkPreg_binary, m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,
-         SmkAtVisitPastmonth, ancestry, cm1bsex, globalmethylation, pediatric, grim, DNAmPACKYRS, Epithelial.cells_saliva,
-         Leukocytes_saliva, SSnewbornCT_center, SSolder_center, cg05575921, childteen)%>%
+         SmkAtVisitPastmonth, ancestry, cm1bsex, globalmethylation, pediatric, Epithelial.cells_saliva,
+         Leukocytes_saliva, SSnewbornCT_center,  cg05575921, childteen)%>%
   tbl_strata(strata=childteen, .tbl_fun=~.x %>%
@@
 smoking_tbl=smoking_tbl%>%as_gt()%>%
-  tab_row_group(label='Maternal', rows = 1:11)%>%
-  tab_row_group(label='Child', rows=12:25)
+  tab_row_group(label='Maternal', rows = 1:10)%>%
+  tab_row_group(label='Child', rows=11:23)
 
@@
 
-# Figure `r main_fig` - Polymethylation scores are associated with prenatal maternal smoke exposure consistently across age and ancestry group in a sample of `r n_total` US children
+# Supplemental Table `r sup_tab`:  Bivariate associations between prenatal maternal smoking (no/yes) and additional DNA methylation summary measures  among a diverse sample of `r n_total` children in the Fragile Families and Child Wellbeing study
 
-```{r}
+```{r supplemental table additional methylation measures}
+smoking_tbl_extendedMeth=modeldata %>% 
+  dplyr::select(smkPreg_binary, globalmethylation, Island_mean, OpenSea_mean, Shelf_mean, Shore_mean, 
+                pediatric, grim, DNAmPACKYRS, Epithelial.cells_saliva, Leukocytes_saliva,
+                SSnewbornCT_center, SSnewborn_center, SSolder_center, anynewborn_center,
+                matches('cg'),  childteen)%>%
+  tbl_strata(strata=childteen, .tbl_fun=~.x %>%
+               tbl_summary(by=smkPreg_binary, statistic=
+                             list(all_continuous()~"{mean} ({sd})"))%>%
+               add_n()%>%add_p(test=list(all_continuous()~"t.test")))
+
+sup_tab=sup_tab+1
+
+```
+
+```{r supplemental table additional methylation measures print, eval=as_supp}
+smoking_tbl_extendedMeth
+```
+
+```{r main figure 3 forest plot}
 my_outcomes=c('Global methylation', 'Pediatric clock', 'AHRR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)')
@@
 data_text=all_models%>%
-  mutate(n=map_dbl(data, nrow), n=paste0('n=', n), x=-0.25)%>%
+  mutate(n=map_dbl(data, nrow), n=paste0('n=', n), x=-0.5)%>%
   unnest(tidied)%>%
@@
 
-```{r, eval=stand_alone, fig.width=12}
+```{r main figure 3 print forest plot, eval=stand_alone, fig.width=12}
 forest_plot
@@
 
-```{r}
-with(modeldata, table(ancestry, PostnatalMaternalSmokingAny, smkPreg_binary))%>%kbl()
+```{r supplemental figure all models forest}
+#with(modeldata, table(ancestry, PostnatalMaternalSmokingAny, smkPreg_binary))%>%kbl()
 
@@
 
-```{r, eval=stand_alone, fig.width=14}
+```{r, eval=as_supp, fig.width=14, fig.height=10}
 forest_plot_compareModels
@@
 
-```{r}
+```{r supplemental figure pre post smoke exposure}
 
@@
 
-```{r, eval=stand_alone, fig.width=14, fig.height=10}
+```{r supplemental figure pre post smoke exposure print, eval=as_supp, fig.width=14, fig.height=10}
 dna_prepost
@@
 
-```{r}
+```{r surrogate variables heatmap}
 #bring the heatmap function
@@
 
-```{r, eval=stand_alone, fig.width=12, fig.height=8}
+```{r surrogate variable heatmap plot, eval=as_supp, fig.width=12, fig.height=8}
 bring.the.heatmap.sv(modeldata, covar=c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'ancestry', 'cm1bsex', 'Leukocytes_saliva', 'Epithelial.cells_saliva', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))
@@
 
-```{r}
+```{r supplemental table surrogate variable table}
 test=rbind(global_models, longitudinal_global_models %>% mutate(childteen='Longitudinal'))%>%
@@
 
-```{r, eval=stand_alone}
+```{r surrogate variable table print, eval=as_supp}
 regression_table_ancestry_svs
@@
 
-```{r}
+```{r supplemental table age and linear mixed effect}
 #my_main_outcomes=c('AHRR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)', 'Sustained smoking polymethylation score (older children)', 'Pediatric clock', 'Global methylation')
@@
 
-```{r, eval=stand_alone}
+```{r supplemental table age and mixed effect print, eval=as_supp}
 regression_table_age
@@
 
-# Supplemental Table `r sup_tab`: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable models among `r n_total` US children by ancestry
+# Supplemental Table `r sup_tab`: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable models among `r n_total` US children by ancestry using linear mixed effect models
 
-```{r}
+```{r supplemental table by ancestry using linear mixed effect models}
 test=rbind(longitudinal_global_models %>% mutate(ancestry='Unstratified'),
@@
 
-```{r, eval=stand_alone}
+```{r supplemental table by ancestry using lme print, eval=as_supp}
 regression_table_ancestry
@@
 
-```{r} 
+```{r supplemental figure compare across pms coefficients} 
  compare_pms=all_models%>%
@@
     geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
-    theme_classic()
+    theme_classic()+
+  scale_color_discrete(name='')+scale_x_continuous("Estimate (95% CI)")+scale_y_discrete('')
 
@@
 
-```{r, eval=stand_alone, fig.width=10}
+```{r compare across pms print, eval=as_supp, fig.width=10}
 compare_pms
@@
 
-# Supplemental Figure `r sup_fig` - Associations between prenatal maternal smoking and other *a priori* CpG sites selected from previous meta-analyses among `n n_total` children in the Fragile Families and Child Wellbeing study
+# Supplemental Figure `r sup_fig` - Associations between prenatal maternal smoking and other *a priori* CpG sites selected from previous meta-analyses among `r n_total` children in the Fragile Families and Child Wellbeing study
 
-```{r}
+```{r supplemental figure compare to top cpgs}
 topCPGs<-read.csv(paste0(datadir, "/TopCpGsWiklundEtAl2019.csv"))
@@
 2016 peripheral blood older children', 'Wicklund et al.
-peripheral blood >16 years'))
+peripheral blood >16 years'))+scale_x_continuous('Estimate (95% CI)')
 
@@
 ```
-```{r, eval=stand_alone, fig.width=10}
+
+```{r compare to top cpgs print, eval=as_supp, fig.width=10}
 plot_aprioriCpG
@@
 
-# Figure `r main_fig` - Polymethylation scores more accurately delineate prenatal smoke exposure status than non-targeted DNA methylation summary measures and *a priori* CpGs
-
-```{r}
+```{r main figure 4 auc}
 
@@
 
-```{r, eval=stand_alone, fig.width=10}
+```{r main figure 4 auc print, eval=stand_alone, fig.width=10}
 auc_fig
@@
 
-# Supplemental Table: Area under the curve for models shown in main figure 4 for classifying prenatal maternal smoke exposure, with *P* value when comparing to base model only or polymethylation score
+# Supplemental Table `r sup_tab`: Area under the curve for models shown in Main Figure 4 for classifying prenatal maternal smoke exposure, with *P* value when comparing to base model only or to model using polymethylation score only
 
-```{r}
+```{r supplemental table auc}
 roc_table=global_roc %>% dplyr::select(childteen, methylation, modeltype, roc, auc)
@@
 
+sup_tab=sup_tab+1  
 ```
 
-```{r, eval=stand_alone}
+```{r supplemental table auc print, eval=as_supp}
 rct1%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value vs base model', 'P value vs polymethylation score'))
@@
 
-# Supplemental Table: Area under the curve when using different coefficients to construct polymethylation scores for predicting prenatal maternal smoke exposure
+# Supplemental Table `r sup_tab`: Area under the curve when using different coefficient sets to construct polymethylation scores for predicting prenatal maternal smoke exposure
 
-```{r} 
+```{r supplemental table auc other coefficients} 
 rct2=roc_table%>% 
@@
 
-```{r, eval=stand_alone}
+```{r supplemental table auc other coefficients print, eval=as_supp}
 rct2%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value vs base model', 'P value vs polymethylation score with cell-type control'))

### 2021-11-10 Papers/Polymethylation_scores/01_titlepage.Rmd
kelly revisions round 2

@@
+---
+output:
+  html_document: default
+  word_document: default
+bibliography: Methylation_SmokingBib.bib
+always_allow_html: true
+---
+
 # Title Page
@@
 
-**Authors:** Freida A. Blostein^1, Jonah Fisher^2, John Dou^1, Lisa Schenper^3, Erin Ware^2 Daniel A. Notterman^3, Colter Mitchell^2, Kelly M. Bakulski^1
+**Authors:** Freida A. Blostein$^1$, Jonah Fisher$^2$,  John Dou$^1$, Lisa Schenper$^3$, Erin Ware$^2$ Daniel A. Notterman$^3$, Colter Mitchell$^2$, Kelly M. Bakulski$^1$`
 
@@
 
-^1 Department of Epidemiology, University of Michigan School of Public Health, Ann Arbor, Michigan, United States of America
+$^1$Department of Epidemiology, University of Michigan School of Public Health, Ann Arbor, Michigan, United States of America
 
-^2 Institute for Social Research, University of Michigan, Ann Arbor, Michigan, United States of America
+$^2$Institute for Social Research, University of Michigan, Ann Arbor, Michigan, United States of America
 
-^3 Department of Molecular Biology, Princeton University, Princeton, New Jersey, United States of America
+$^3$Department of Molecular Biology, Princeton University, Princeton, New Jersey, United States of America

### 2021-11-10 Papers/Polymethylation_scores/08_methods.Rmd
kelly revisions round 2

@@
 
-#### Primary model
+#### Base model
 
-In our primary model we adjusted for maternal income to poverty ratio at baseline, child sex, child age, plate from DNA methylation processing, and estimated immune cell proportion estimated from DNA methylation.
+In our base model we adjusted for maternal income to poverty ratio at baseline, child sex, child age, plate from DNA methylation processing, and estimated immune cell proportion estimated from DNA methylation.
 In nonstratified models, we adjusted for the first two components of genetic ancestry from principal component analysis.

### 2021-11-10 Papers/Polymethylation_scores/02_abstract.Rmd

@@
 
-**Background**: Prenatal maternal smoking is a difficult to measure prenatal exposure with negative implications for child health.
-DNA methylation signatures in blood and placental samples can function as a biomarker of prenatal maternal smoking.
-However little work has assessed how DNA methylation signatures of prenatal maternal smoking vary across ages, ancestry groups or tissues.
-Using children in the Fragile Families and Child Wellbeing study, we tested if prenatal maternal smoking is associated with salivary polymethylation scores.
-We assessed the consistency of associations at ages 9 and 15 and their portability across  African, European and Hispanic ancestries.
+**Background**: Prenatal maternal smoking has negative implications for child health.
+DNA methylation signatures  can function as a biomarker of prenatal maternal smoking.
+However little work has assessed how DNA methylation signatures of prenatal maternal smoking vary across ages, ancestry groups, or tissues.
+In the Fragile Families and Child Wellbeing study, we tested whether prenatal maternal smoking was associated with salivary polymethylation scores.
+We assessed the consistency of associations at ages 9 and 15, their portability across  participants from African, European, and Hispanic genetic ancestries and the accuracy of exposure classification using Area Under the Curve (AUC) from receiver operating curve analyses.
 
-**Results**: We created a polymethylation score using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cordblood.
-At age 9, prenatal maternal smoke exposure was associated with a 0.52 (95%CI: 0.36, 0.67) standard deviation higher polymethylation score.
-At age 15, At age 9, prenatal maternal smoke exposure was associated with a 0.46 (95%CI: 0.3, 0.62) higher polymetylation score.
-The direction and magnitude of the association was portable across genetic ancestries at age 9 although the association was attenuated among European and Hispanic genetic ancestry samples at age 15.
-The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (Area under the curve from receiver operating curve (AUC-ROC age 9: 0.77 age 15: 0.77), and more accurate than using a single *a priori* CpG site, AHRR:CpG05575921 (AUC-ROC age 9:0.74, *P* value=0.02; age 15:0.73, *P* value=0.02).
+**Results**: We created saliva polymethylation scores using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cordblood.
+In the full sample at age 9 (n=756), prenatal maternal smoke exposure was associated with a 0.52 (95%CI: 0.36, 0.67) standard deviation higher polymethylation score.
+The direction and magnitude of the association was consistent across when stratified genetic ancestries.
+In the full sample at age 15 (n=749), prenatal maternal smoke exposure was associated with a 0.46 (95%CI: 0.3, 0.62) standard deviation higher polymetylation score, and the association was attenuated among the European and Hispanic genetic ancestry samples.
+The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (AUC age 9=0.77, *P* value<0.001, age 15=0.77, *P* value<0.001), and more accurate than using a single *a priori* site in the AHRR gene (cg05575921 AUC=0.74, *P* value=0.03; age 15=0.73, *P* value=0.01).
 
-**Conclusions**: Prenatal maternal smoking associates with DNA methylation signatures in saliva samples, a clinically practical tissue.
-Polymethylation scores for prenatal maternal smoking are portable across ancestries and more accurate than some individual CpG sites.
-Thus, DNA polymethylation scores from saliva samples could serve as robust and practical clinical biomarkers of prenatal maternal smoke exposure.
+**Conclusions**: Prenatal maternal smoking was associated with DNA methylation signatures in saliva samples, a clinically practical tissue.
+Polymethylation scores for prenatal maternal smoking were portable across genetic ancestries and more accurate than individual DNA methylation sites.
+DNA polymethylation scores from saliva samples could serve as robust and practical clinical biomarkers of prenatal maternal smoke exposure.

### 2021-11-10 Papers/Polymethylation_scores/04_background.Rmd

--- 2021-11-02 Papers/Polymethylation_scores/04_background.Rmd
@@
 Maternal cigarette smoking during pregnancy is associated with reproducible DNA methylation signatures, measured in infant cord blood and placental tissues, and primarily tested in European cohorts [@Joubert:2016ue; @Everson:2021td; @Cardenas2019].
-Prenatal maternal cigarette smoking exposure is associated with postnatal DNA methylation signatures in blood from cohorts of older children and adults[@Tehranifar:2018tb ; @Richmond:2018tu].
+Prenatal maternal cigarette smoking exposure is associated with postnatal DNA methylation signatures in blood from cohorts of older children and adults [@Tehranifar:2018tb ; @Richmond:2018tu].
 Few studies have considered the persistence of the association between maternal prenatal cigarette smoking and DNA methylation at different ages in the same individuals.
-Additionally, little work has considered the portability of the association between maternal prenatal cigarette smoking and DNA methylation across tissues and ancestry groups.
+Additionally, little work has considered the portability of the association between maternal prenatal cigarette smoking and DNA methylation across tissues.
 DNA methylation drives cell differentiation and cell type proportions differ across tissue types.
@@
 Cord blood and placental samples are unlikely to be available in clinical settings.
-While peripheral blood is a more realistic clinical samples, saliva is even easier to collect.
+While peripheral blood is a more realistic clinical sample, saliva is even easier to collect.
 To our knowledge, no study has evaluated associations between maternal prenatal cigarette smoking and salivary DNA methylation.
-Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017], and DNA methylation signatures of own-smoking differ across ancestry groups [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012].
+
+Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017], and DNA methylation signatures of own-smoking differ across genetic ancestry groups [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012].
 Only a few studies have examined associations between prenatal maternal cigarette smoking and DNA methylation in non-European ancestry populations.
 Among 954 infants, 70% of whom had mothers identifying as Black, prenatal maternal smoking was associated with cord blood DNA methylation at 38 CpG sites [@Xu2021].
-The direction of these association between prenatal maternal cigarette smoking and DNA methylation at these sites were generally similar in children of Black and non-Black participants [@Xu2021].
-Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking was associated with DNA methylation in peripheral blood samples but ancestry-stratified analyses were not performed [@Ladd-Acosta:2016aa].
+The direction of the associations between prenatal maternal cigarette smoking and DNA methylation at these sites were generally similar in children of Black and non-Black participants [@Xu2021].
+Among 572 3-5 year old children, 186 of whom were of African or admixed genetic ancestry, prenatal maternal smoking was associated with DNA methylation in peripheral blood samples but ancestry-stratified analyses were not performed [@Ladd-Acosta:2016aa].
 Similarly, among 89 middle-aged women, of whom reported African American or Hispanic ethnicity, prenatal maternal cigarette smoking was associated with DNA methylation at 17 of 190 tested CpG sites [@Tehranifar:2018tb].
 Adjusting for ancestry did not substantially affect the association between prenatal maternal smoking and DNA methylation at any of the sites, although the authors note that African American women had on average 2--5% higher mean DNA methylation on the absolute scale as compared with White and Hispanic women at 5 *CYP1A1* CpG sites [@Tehranifar:2018tb].
-Of 148 CpG sites selected based on their association prenatal maternal smoking among primarily European children, 7 CpG sites were associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue].
-Thus, while prenatal maternal cigarette smoking is associated with DNA methylation, the consistency of this association across age, tissue and ancestry is understudied.
+Of 148 CpG sites selected based on their association with prenatal maternal smoking among primarily European children, 7 CpG sites were also associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue].
+While prenatal maternal cigarette smoking is associated with DNA methylation, the consistency of this association across genetic ancestry is understudied.
 
-An important next step after association testing is the evaluation of DNA methylation as a biomarker of prenatal maternal smoke exposure, which is often done by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior.
-<!--#DEFINE POLYMETHYLATION SCORES HERE?-->DNA methylation signatures from cord blood classified maternal smoking behavior with accuracy of 91% in a primarily European cohort [@Reese:2017ts].
-Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking behavior could also be classified with an accuracy of 81% when using a support vector machine approach and DNA methylation at 26 sites measured in peripheral blood at a single time point [@Ladd-Acosta:2016aa].
-Among middle-aged adults, peripheral blood from a single time point could even predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu].
-However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across tissue, age and ancestry has not been evaluated.
-Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times or evaluated if the inclusion of multiple time points improves performance.
-No study has evaluated the accuracy of salivary DNA methylation as a biomarker for prenatal maternal smoking.
+After association testing, an important next step is to evaluate DNA methylation as a biomarker of prenatal maternal smoke exposure by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior.
+Biomarkers can consist of single DNA methylation sites or summary meausres across multiple sites [@Philibert2020; @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa].
+One attractive option is a polymethylation score, analogous to a polygenic score.
+In the polymethylation score, CpG sites are weighted by the strength of their association with maternal smoking from a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a].
+One such DNA methylation score calculated from cord blood classified maternal smoking behavior with an area under the curve (AUC) of 0.82 in a primarily European cohort [@Reese:2017ts].
+Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking behavior was classified with an AUC of 0.87 [@Ladd-Acosta:2016aa].
+Among middle-aged adults, a score calculated using DNA methylation in peripheral blood samples from a single time point could even predict prenatal smoke exposure (30 years previously) with an AUC of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu].
+Ideally, a biomarker for prenatal maternal smoking would be consistent over participant age at DNA methylation sampling and portable across genetic ancestry groups.
+However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across tissue, age, and ancestry has not been evaluated in the same study.
 
-DNA methylation changes hold promise as a biomarker for prenatal maternal smoking.
-However, before translation to clinical applications, we must understand how and if the signal varies across age, ancestry and tissue sample.
+DNA methylation differences hold promise as a biomarker for prenatal maternal smoking.
+Before translation to clinical applications, we must understand how and if the signal varies across age, ancestry, and tissue sample.
 In the Fragile Families and Child Wellbeing study, a diverse longitudinal cohort of children, we aimed to assess the potential of salivary DNA methylation biomarkers for prenatal maternal cigarette smoking.
 We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including epigenetic clocks, polymethylation scores, and individual *a priori* CpG sites.
-We then performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a persistent and portable biomarker of prenatal maternal smoking.
+We then performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a persistent and consistent biomarker of prenatal maternal smoking.

### 2021-11-10 Papers/Polymethylation_scores/05_results.Rmd

@@
 
-Of `r methylCohort$MethID%>%length()` samples from `r methylCohort$idnum%>%unique()%>%length()` unique participants with DNA methylation data measured on the Illumina 450K  array, `r m_obs` samples from `r n_total` participants had complete data on additional key covariates (Figure 1).
+In a sample of the Fragile Families and Child Wellbeing Study, saliva DNA methylation was measured on `r methylCohort$MethID%>%length()` samples from `r methylCohort$idnum%>%unique()%>%length()` unique participants with the Illumina 450K  array.
+Complete data on covariates of interest was available on `r m_obs` samples from `r n_total` participants who were included in the analysis (Figure 1).
 There were `r n_twovisit` participants with both age 9 and age 15 DNA methylation samples.
@@
 
-In bivariate analyses and multivariable models, we focused on the polymethylation scores and the single top CpG site from Joubert et al, AHRR:cg05575921, as hypothesized biomarkers, and used global methylation and the pediatric clock as negative controls.
+In bivariate analyses and multivariable models, we focused on the polymethylation scores and the single top CpG site from prior research, cg05575921 in the *AHRR* gene,, as hypothesized biomarkers, and used global methylation and the pediatric clock as negative controls.
 
@@
 
-Mothers who reported smoking during pregnancy had lower income to poverty ratios than those who did not ( `r modeldata_people%>%filter(smkPreg_binary=='Yes')%>%summarise(mean(cm1inpov))%>%round(2)` vs `r modeldata_people%>%filter(smkPreg_binary=='No')%>%summarise(mean(cm1inpov))%>%round(2)`).
+Mothers who reported smoking during pregnancy had lower income to poverty ratios (`r modeldata_people%>%filter(smkPreg_binary=='Yes')%>%summarise(mean(cm1inpov))%>%round(2)`) than those who did not ( `r modeldata_people%>%filter(smkPreg_binary=='No')%>%summarise(mean(cm1inpov))%>%round(2)`).
 Mothers who smoked were more likely to report prenatal alcohol use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%), prenatal drug use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%) and postnatal smoking (`r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[1, 2]%>%round(2)*100`%) than mothers who did not report prenatal maternal smoking (Table 1).
-Children of mothers who reported smoking during pregnancy were more likely to be of European genetic ancestry (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[2, 2], 2)*100`%) than children of mothers who did not (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[1, 2], 2)*100`%, *P* value=`r round(with(modeldata_people, chisq.test(table(smkPreg_binary, ancestry)))$p.value, 4)`).
-Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both the age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) visits (Table 1, Supplemental Table2).
-At age 9, children exposed to prenatal maternal smoking had lower DNA methylation at AHRR:cg05575921 (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)*100`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)*100`%, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
-At age 15, children exposed to prenatal smoking had lower DNA methylation at AHRR:cg05575921 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)*100`%) than children of non-smoking mothers, although this difference was not significant (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)*100`% *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
+Children of mothers who reported smoking during pregnancy were more likely to be of European genetic ancestry (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[2, 2], 2)*100`%) than children of mothers who did not (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[1, 2], 2)*100`%, *P* value=<0.001).
+Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both the age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) visits (Table 1, Supplemental Table 2).
+At age 9, children exposed to prenatal maternal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`%, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
+At age 15, children exposed to prenatal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers, although this difference was not significant (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`% *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
 
@@
 
-The association between prenatal maternal smoking and the polymethylation score was also observed in multivariable models  (Figure 3; Supplemental Table 2).
+The association between prenatal maternal smoking and the polymethylation score was also observed in multivariable models, adjusting for base model covariates of child sex, maternal income to poverty ratio at baseline, proportion of salivary immune cells, sample plate from DNA methlation analysis, and child age  (Figure 3; Supplemental Table 2).
 At age 9, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.
@@
 
-Prenatal maternal smoking was associated with lower methylation at AHRR:cg055975921 at age 9 (`r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921'", 'smkPreg_binaryYes', 2)`) and age 15 (`r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921'", 'smkPreg_binaryYes', 2)`).
-Associations between prenatal maternal smoking and AHRR:cg05575921 were less robust across ancestries (Figure 3).
+At age 9, prenatal maternal smoking was associated with `r round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95% CI: `r paste(round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`) after adjusting forbase model covariates.
+Similarly, at age 15 prenatal maternal smoking was associated with `r round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95%CI: (`r paste(round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`)).
+The magnitude of the association between prenatal maternal smoking and cg05575921 was reduced in European genetic ancestry samples at age 15 (Figure 3).
+The direction of the association between prenatal maternal smoking and cg05575921 was reversed in Hispanic genetic ancestry sample (Figure 3).
 
-## Sensitivity analyses
+Prenatal maternal smoking was not associated with global DNA methylation or epigenetic age (Figure 3).
 
-When also adjusting for other prenatal exposures and postnatal smoke exposure, the association between prenatal maternal smoking and the polymethylation score was very consistent (Supplemental Figure 4).
-The direction of the association between prenatal maternal smoking and AHRR:cg055975921 was also consistent, although no longer significant at age 15.
+## Sensitivity analyses for association testing
+
+After additionally adjusting for other prenatal exposures and postnatal smoke exposure, the association between prenatal maternal smoking and the polymethylation score was very consistent (Supplemental Figure 4).
+Similarly, after additionally adjusting for other prenatal exposures and postnatal smoke exposure, the direction of the association between prenatal maternal smoking and cg055975921 was consistent, although no longer significant at age 15.
 Children exposed to postnatal smoke but not prenatal smoke did not have higher polymethylation scores than children unexposed to both postnatal and prenatal smoke.
@@
 
-Results were similar when using linear mixed effect models instead of age-stratified models (Supplemental Table 3 & 4).
+Results were similar when using linear mixed effect models instead of age-stratified models and adjusting for base model covariates (Supplemental Table 3 & 4).
 
@@
 The association between prenatal maternal smoking and the polymethylation score using regression coefficients from newborn cord blood with cell-type controls was the strongest.
-Prenatal maternal smoking also associated with other selected *a priori* CpG sites consistently with associations reported in previous meta-analyses (Supplemental Table 2, Supplemental Figure 8).
+Prenatal maternal smoking was also associated with other selected *a priori* CpG sites consistently with associations reported in previous meta-analyses (Supplemental Table 2, Supplemental Figure 8).
 
@@
 We considered classification when using DNA methylation summary measures alone.
-We also considered classification when using DNA methylation summary measures in addition to a base model of child sex, maternal income to poverty ratio at baseline, child age at methylation measurement, estimated immune cell proportion, plate from methylation processing and the first two components of genetic ancestry.
+We also considered classification when using DNA methylation summary measures in addition to a base model of child sex, maternal income to poverty ratio at baseline, child age at DNA methylation measurement, estimated immune cell proportion, plate from methylation processing and the first two components of genetic ancestry.
 When used without DNA methylation measures, these base model variables had an area under the curve (AUC) of `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 9')%>%pull(auc)` at age 9 and `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 15')%>%pull(auc)` at age 15.
 
-At age 9, including the polymethylation score  significantly improved the base model accuracy (AUC: `r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(auc)`, *P* value comparing to base model=`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(4)`).
-Similarly, including the polymethylation significantly improved the base model accuracy at age 15 (AUC:`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(auc)` , *P* value=`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(4)`).
+At age 9, including the polymethylation score  significantly improved the base model accuracy (AUC: `r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(auc)`, *P* value comparing to base model<0.001).
+At age 15, similarly, including the polymethylation score significantly improved the base model accuracy (AUC:`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(auc)` , *P* value<0.001).
 
-At age 9,the base model with the polymethylation score had a larger AUC than the base model with AHRR:cg05575921 (AUC base model+AHRR:cg05575921: `r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(auc)`; *P* value =`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(5)`).
-At age 15, this was also true (AUC for base model+AHRR:cg05575921=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(auc)`,  *P* value=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(5)`).
+At age 9, the base model with the polymethylation score had a larger AUC than the base model with cg05575921 (AUC base model + cg05575921: `r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(auc)`; *P* value =`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(5)`).
+At age 15, this was also true (AUC for base model + cg05575921=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(auc)`,  *P* value=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(5)`).
 Accuracy was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B).
 
-## Sensitivity analyses
+## Sensitivity analyses for biomarker accuracy assessment
 

### 2021-11-10 Papers/Polymethylation_scores/06_discussion.Rmd

@@
 Prenatal maternal smoking classification using saliva DNA methylation in our sample of 9 and 15 year olds performed comparably to classification using peripheral blood from older adults (AUC 0.72) [@Richmond:2018tu].
-However, previous biomarkers of prenatal maternal smoking using DNA methylation of cord blood from newborns (AUCs ranging from 0.82-0.97) and peripheral blood from 3-5 year old children (AUC=0.87) more accurately classified prenatal maternal smoke exposure [@Reese:2017ts; @Ladd-Acosta:2016aa].
+However, previous prenatal maternal smoke exposure biomarkers created using LASSO regression and cord blood from newborns performed better (AUCs ranging from 0.82-0.97) [@Ladd-Acosta:2016aa] A support vector machine approach performed on and peripheral blood from 3-5 year old children also more accurately classified prenatal maternal smoke exposure (AUC=0.87) [@Reese:2017ts].
 Differences in DNA methylation patterns across tissues and over time may influence the performance of DNA methylation biomarkers.
+Different methods for summarizing across multiple DNA methylation sites may also influence biomarker performance.
 The similarity in performance between saliva DNA methylation in our study and peripheral blood DNA methylation in older adults is encouraging for the use of saliva DNA methylation as a readily accessible clinical sample.

### 2021-11-10 Papers/Polymethylation_scores/99_FigureCaptions.Rmd

@@
 
-#### Figure 1 - Selection of samples from the Fragile Families and Child Wellbeing study into analytic subset.
-N represents the number of individuals at each step in the selection procedure, M represents the number of samples.
+#### **Figure 1 - Selection of samples from the Fragile Families and Child Wellbeing study into analytic subset.** N represents the number of individuals at each step in the selection procedure, M represents the number of samples.
 Individuals with repeated measures can have more than one sample.
@@
 
-#### Figure 2 - Differences in selected DNA methylation summary measures by self-report of prenatal maternal smoking among `r n_total` children in the Fragile Families and Child Wellbeing study (CT=cell-type control)
+#### **Figure 2 - Differences in selected DNA methylation summary measures by self-report of prenatal maternal smoking among `r n_total` children in the Fragile Families and Child Wellbeing study at ages 9 and 15.** Samples from children exposed to prenatal maternal smoke in grey, samples from children unexposed to prenatal maternal smoke in black.
+From top-left, clockwise: Polymethylation scores for prenatal maternal smoke exposure, constructed using regression coefficients for prenatal smoke exposure predicting DNA methylation in newborn cordblood samples, accounting for cell-type control.
+DNA methylation values from samples in the Fragile Families and ChildWellbeing study were mean-centered, then multiplied by these regression coefficients and summed.
+Pediatric epigentic clock (years).
+AHRR gene: percent cg05575921 methylation.
+Percent global DNA methylation.
 
@@
 
-#### Figure 3 - Prental maternal smoke exposure is consistently associated with polymethylation scores at ages 9 and 15 and is portable across genetic ancestry groups in a sample of `r n_total` children in the Fragile Families and Child Wellbeing study
+#### **Figure 3 - Prenatal maternal smoke exposure is consistently associated with polymethylation scores at ages 9 and 15 and is portable across genetic ancestry groups in a sample of `r n_total` children in the Fragile Families and Child Wellbeing study** All models shown controlled for: first two principal components of child genetic ancestry (from ancestry-stratified principal components for ancestry stratified models), child sex, maternal income-to-poverty ratio at birth, immune cell proportion estimated from methylation data, yes/no other maternal prenatal drug use, yes/no maternal prenatal alcohol use, postnatal maternal smoking when child age 1 or age 5, postnatal maternal/primary care give smoking packs/day in month prior to saliva sample.
 
@@
 
-#### Figure 4 - A) Polymethylation scores accurately classify maternal report of prenatal smoke exposure status at ages 9 and 15 among a `r n_total` children in the Fragile Families and Child Wellbeing Study B) Inclusion of two-visits of DNA methylation data does not improve classification of maternal report of prenatal smoke exposure status among `r n_total` children in the Fragile Families and Child Wellbeing study
+####  **Figure 4 - Polymethylation scores accurately classify prenatal maternal smok exposure at ages 9 and 15 among `r n_total` children in the Fragile Families and Child Wellbeing study** A) Receiver operator curve for select DNA methylation measures for predicting prenatal smoke exposure using no other variables (light colors) or using base model variables (dark colors, other variables included: child sex, maternal income-poverty ratio at birth, immune cell proportion and batch of methylation data processing).
+B) Receiver operator curve for include polymethylation scores individually at each visit (black & light grey) or jointly (dark grey).
 

### 2021-11-10 Reports/Paper1.Rmd

@@
 output:
-  pdf_document: default
-  word_document: default
   html_document:
@@
     theme: sandstone
+  word_document: default
+  pdf_document: default
 always_allow_html: true
@@
 theme_set(theme_bw())
+set_label(modeldata$cm1bsex)='Child gender'
+set_label(modeldata$cm1inpov)='Maternal income to poverty threshold (at baseline)'
+set_label(modeldata$PostnatalMaternalSmokingDose)='Primary caregiver smoking past month by dose'
+set_label(modeldata$SSnewbornCT_center)='Polymethylation score for smoke exposure'
+set_label(modeldata$pediatric)='Pediatric epigenetic clock (years)'
+set_label(modeldata$grim)='GRIM epigenetic clock (years)'
+set_label(modeldata$globalmethylation)='Percent global DNA methylation'
+set_label(modeldata$cg05575921)='AHRR gene: percent cg05575921 methylation'
 ```
@@
 
-# Supplemental Table `r sup_tab` - Descriptive statistics of excluded vs included samples with methylation data available from the Fragile and Families and Child Wellbeing study
+# Supplemental Table `r sup_tab` - Descriptive statistics of excluded vs included samples with methylation data available from the Fragile and Families and Child Wellbeing study.
+Samples excluded according to the flow diagram in Main Figure 1
 
@@
 exclusion_tbl=myFF %>% mutate(childteen=case_when(childteen=='C'~'Age 9', childteen=='T'~'Age 15'))%>%
-  dplyr::select(smkPreg_binary, m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,
-         SmkAtVisitPastmonth, ancestry, cm1bsex, childteen, globalmethylation, pediatric, grim,
-         Epithelial.cells_saliva, Leukocytes_saliva, SSnewbornCT_center, SSolder_center,
-         cg05575921, MethylData2)%>%
+    dplyr::select(smkPreg_binary, childteen, MethylData2, SSnewbornCT_center, cg05575921, pediatric, globalmethylation, Leukocytes_saliva, Epithelial.cells_saliva, ancestry, cm1bsex,  m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,SmkAtVisitPastmonth )%>%
+  mutate(smkPreg_binary=factor(smkPreg_binary, levels=c('No', 'Yes'), labels=c('Not exposed', 'Smoke exposed')), 
+         cm1bsex=factor(cm1bsex, levels=c('1 Boy', '2 Girl'), labels=c('Boy', 'Girl')))%>%
+  mutate(cm1bsex=set_label(cm1bsex, 'Child gender'), smkPreg_binary=set_label(smkPreg_binary, 'Prenatal maternal smoking'),
+         childteen=set_label(childteen, 'Age at DNA methylation sample'))%>%
   tbl_summary(., by=MethylData2, statistic=list(all_continuous()~"{mean} ({sd})"))%>%
@@
 
-# Supplemental Figure `r sup_fig` - Pearson correlation of methylation summary measures from an anlysis of M=`r m_obs` samples and N=`r n_total`children in the Fragile Families and Child Wellbeing cohort
+# Supplemental Figure `r sup_fig` - Pearson correlation of selected DNA methylation summary measures from an anlysis of M=`r m_obs` samples and N=`r n_total`children in the Fragile Families and Child Wellbeing cohort.
+Lower diaganol shows the Pearson rho correlation coefficient, upper triangle shows ellipses illustrating relationship between variables colored by the magnitude and direction of the correlation.
 
@@
 
-# Supplemental Figure `r sup_fig` - Correlation between age 9 and age 15 visits for selected DNA methylation summary measures among `r n_twovisit` children with age 9 and age 15 visit data in the Fragile Families and Child Wellbeing study
+# Supplemental Figure `r sup_fig` - Correlation between age 9 and age 15 visit DNA methylation summary measures among `r n_twovisit` children with age 9 and age 15 visit data in the Fragile Families and Child Wellbeing study.
+Each point represents one individual, x-axis is the DNA methylation summary measure at age 9 and y-axis is the DNA methylation measure at age 15.
+Blue lines are simple linear regression lines of best fit, with Pearson rho and associated $P$ value shown.
+From top-left: Polymethylation score for prenatal maternal smoking, score created using coefficients from a regression of prenatal maternal smoking and DNA methylation in newborn cord blood, incorporating cell-type control.
+Polymethylation score for prenatal maternal smoking using coefficients from a regression of prenatal maternal smoking and DNA methylation in newborn cord blood, no cell-type control.
+Polymethylation score calculated using regression coefficients from a regression of prenatal maternal smoking and DNA methylation in older children's peripheral blood samples.
+Pediatric epigenetic clock (years).
+GRIM epigenetic clock (years).
+Percent DNA methylation at cg05575921 in AHRR.
+Percent global DNA methylation (mean across all DNA methylation sites in cleaned probe set).
+Proportion of epithelial cells calculated from DNA methylation using reference panel of children's saliva.
+Proportion of immune cells calculated from DNA methylation using a reference panel of children's saliva.
 
@@
 
-# Supplemental Figure `r sup_fig`: Distributions of selected methylation summary measures at age 9 and age 15 among `r n_total` children in the Fragile Families and Child Wellbeing study
+# Supplemental Figure `r sup_fig`: Distributions of selected methylation summary measures at age 9 and age 15 among `r n_total` children in the Fragile Families and Child Wellbeing study.
+Captive box plots within violin plots show the distribution of selected DNA methylation measures in samples from age 9 (red) and age 15 (blue).
+Black lines show lines of best fit.
+From top-left: Polymethylation score for prenatal maternal smoking, score created using coefficients from a regression of prenatal maternal smoking and DNA methylation in newborn cord blood, incorporating cell-type control.
+Polymethylation score for prenatal maternal smoking using coefficients from a regression of prenatal maternal smoking and DNA methylation in newborn cord blood, no cell-type control.
+Polymethylation score calculated using regression coefficients from a regression of prenatal maternal smoking and DNA methylation in older children's peripheral blood samples.
+Pediatric epigenetic clock (years).
+GRIM epigenetic clock (years).
+Percent DNA methylation at cg05575921 in AHRR.
+Percent global DNA methylation (mean across all DNA methylation sites in cleaned probe set).
+Proportion of epithelial cells calculated from DNA methylation using reference panel of children's saliva.
+Proportion of immune cells calculated from DNA methylation using a reference panel of children's saliva.
 
@@
 ```{r bivariate table}
+
 smoking_tbl=modeldata %>% 
-  dplyr::select(smkPreg_binary, m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,
-         SmkAtVisitPastmonth, ancestry, cm1bsex, globalmethylation, pediatric, Epithelial.cells_saliva,
-         Leukocytes_saliva, SSnewbornCT_center,  cg05575921, childteen)%>%
+  dplyr::select(smkPreg_binary, childteen, SSnewbornCT_center, cg05575921, pediatric, globalmethylation, Leukocytes_saliva, Epithelial.cells_saliva, ancestry, cm1bsex,  m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,SmkAtVisitPastmonth )%>%
+  mutate(smkPreg_binary=factor(smkPreg_binary, levels=c('No', 'Yes'), labels=c('Not exposed', 'Smoke exposed')), 
+         cm1bsex=factor(cm1bsex, levels=c('1 Boy', '2 Girl', labels=c('Boy', 'Girl'))))%>%
   tbl_strata(strata=childteen, .tbl_fun=~.x %>%
@@
 smoking_tbl=smoking_tbl%>%as_gt()%>%
-  tab_row_group(label='Maternal', rows = 1:10)%>%
-  tab_row_group(label='Child', rows=11:23)
+  tab_row_group(label='Child characteristics', rows = 1:13)%>%
+  tab_row_group(label='Maternal characteristics', rows=14:23)%>%
+     row_group_order(
+      groups = c("Child characteristics", "Maternal characteristics")
+    )
 
@@
 
-# Supplemental Table `r sup_tab`:  Bivariate associations between prenatal maternal smoking (no/yes) and additional DNA methylation summary measures  among a diverse sample of `r n_total` children in the Fragile Families and Child Wellbeing study
+# Supplemental Table `r sup_tab`:  Bivariate associations between prenatal maternal smoking and additional DNA methylation summary measures among a diverse sample of `r n_total` children in the Fragile Families and Child Wellbeing study.
+Percent DNA methylation shown calculated across all DNA methylation sites available (global DNA methylation) and stratified by genomic region (Island, Open Sea, Shelf, Shore).
+Polymethylation score for smoke exposure from main analysis and from sensitivity analyses where scores were calculated using alternative available regression coefficients.
 
@@
                 matches('cg'),  childteen)%>%
+    mutate(smkPreg_binary=factor(smkPreg_binary, levels=c('No', 'Yes'), labels=c('Not exposed', 'Smoke exposed')))%>%
   tbl_strata(strata=childteen, .tbl_fun=~.x %>%
@@
 
-# Supplemental Figure `r sup_fig`: Sensitivity analyses for controlling for different variable sets in multivariable analyses of prenatal maternal smoking and DNA methylation summary measures
+# Supplemental Figure `r sup_fig`: Sensitivity analyses for controlling for different variable sets in multivariable analyses of prenatal maternal smoking and DNA methylation summary measures.
+Base model: Child sex, maternal income to poverty ratio, immune cell proportion, plate, 1st two principal components from PCA on child genetic data, child age.
+Prenatal exposures model: base model + yes/no any prenatal maternal alcohol use, yes/no any prenatal other drug use.
+Secondhand smoking models: prenatal exposures model + yes/no any postnatal maternal/primary care giver smoking at age 1 or 5, pk/day in month prior to primary caregiver/maternal interview at age 9 or 15.
+Surrogate variable: including all surrogate variables
 
@@
 
-```{r, eval=as_supp, fig.width=14, fig.height=10}
+```{r, eval=as_supp, fig.width=14, fig.height=12}
 forest_plot_compareModels
@@
 
-# Supplemental Figure `r sup_fig` DNA methylation summary measures across prenatal and postnatal exposures
+# Supplemental Figure `r sup_fig` DNA methylation summary measures across combinations of exposure to prenatal and postnatal  maternal smoking.
+Sample size is shown at the bottom of violin plots with captured boxplots.
+POlymethylation score shown in top facets, percent DNA methylation at cg05575921 in AHRR gene shown in bottom facets, age 9 samples shown in left facets, age 15 shown in right facets
 
@@
 
-# Supplemental Figure `r sup_fig`: Surrogate variable associations with known covariates
+# Supplemental Figure `r sup_fig`: Heatmap of *P* values from Pearson correlations between known covariates and surrogate variables from surrogate variable models
 
@@
 covar=c('Sample_Plate', 'Slide', 'Array', 'cm1bsex', 'smkPreg_binary', 'childteen', 'ancestry')
+modeldata$Sample_Plate=set_label(modeldata$Sample_Plate, 'Batch')
+modeldata$cm1bsex=set_label(modeldata$cm1bsex, 'Child gender')
+modeldata$PostnatalMaternalSmokingAny=set_label(modeldata$PostnatalMaternalSmokingAny, 'Postnatal maternal smoking (age 1 & 5)')
+modeldata$SmkAtVisitPastmonth=set_label(modeldata$SmkAtVisitPastmonth, 'Postnatal maternal smoking (month prior to DNA methylation sample')
 
@@
 
-# Supplemental Table `r sup_tab`: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable linear models stratified by age or in multivariable linear mixed effect models among `r n_total` US children
+# Supplemental Table `r sup_tab`: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable linear models stratified by age or in multivariable linear mixed effect models among `r n_total` US children.
+All models controlled for:  first two principal components of child genetic ancestry (recalculated within ancestry strata for stratified models), child sex, maternal income-to-poverty ratio at birth, immune cell proportion estimated from methylation data, child age
 
@@
 
-# Supplemental Table `r sup_tab`: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable models among `r n_total` US children by ancestry using linear mixed effect models
+# Supplemental Table `r sup_tab`: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable models among `r n_total` US children by ancestry using linear mixed effect models.
+All models controlled for:  first two principal components of child genetic ancestry (recalculated within ancestry strata for stratified models), child sex, maternal income-to-poverty ratio at birth, immune cell proportion estimated from methylation data, child age
 
@@
 
-# Supplemental Figure `r sup_fig`: Associations between prenatal maternal smoking and polymethylation scores calculated using alternative coefficient sets among `n n_total` children in the Fragile Families and Child Wellbeing study
+# Supplemental Figure `r sup_fig`: Associations between prenatal maternal smoking and polymethylation scores calculated using alternative coefficient sets among `n n_total` children in the Fragile Families and Child Wellbeing study.
+Base model: Child sex, maternal income to poverty ratio, immune cell proportion, plate, 1st two principal components from PCA on child genetic data, child age.
+Prenatal exposures model: base model + yes/no any prenatal maternal alcohol use, yes/no any prenatal other drug use.
+Secondhand smoking models: prenatal exposures model + yes/no any postnatal maternal/primary care giver smoking at age 1 or 5, pk/day in month prior to primary caregiver/maternal interview at age 9 or 15.
+Surrogate variable: including all surrogate variables
 
@@
     geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
-    facet_grid(predictors~ancestry, scales='free_x')+
+    facet_wrap(predictors~., scales='free_x')+
     geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
@@
 
-```{r compare across pms print, eval=as_supp, fig.width=10}
+```{r compare across pms print, eval=as_supp, fig.width=10, fig.height=12}
 compare_pms
@@
 topCPGs<-topCPGs %>% mutate(estimate=as.numeric(gsub("−\\s", "-", gsub(" .*", "", β..SE.))), std.error=as.numeric(sub("\\).*", "", sub(".*\\(", "", β..SE.))), p.value=P.value, source="Wiklund et al.
-2019", conf.low=(estimate-1.96*std.error), conf.high=(estimate+1.96*std.error), set ='')%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)
+2019", conf.low=(estimate-1.96*std.error), conf.high=(estimate+1.96*std.error), set ='')%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)%>%mutate(estimate=estimate*100, conf.low=100*conf.low, conf.high=100*conf.high)
 
@@
 #and group into a list by analysis type
-cpgs<-cpgs %>% pivot_longer(cols = c(matches("_Coef|_SE|_P|_FDR|_Direction")), names_to = c("set", ".value"), names_pattern="(.+)_(.+)")%>%rename_with(~gsub('_', '', .x))%>%mutate(Gene=`Mapped Gene`, estimate=Coef, conf.low=estimate-1.96*SE, conf.high=estimate+1.96*SE, source='Joubert et al.
-2016', std.error=SE, p.value=P, Chr=CHR) %>%mutate(source=paste0(source, ': ', set))%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)%>%filter(set %in% c('SSolder', 'SSnewbornCT'))
+cpgs<-cpgs %>% pivot_longer(cols = c(matches("_Coef|_SE|_P|_FDR|_Direction")), names_to = c("set", ".value"), names_pattern="(.+)_(.+)")%>%rename_with(~gsub('_', '', .x))%>%mutate(Gene=`Mapped Gene`, estimate=Coef, conf.low=(estimate-1.96*SE), conf.high=(estimate+1.96*SE), source='Joubert et al.
+2016', std.error=SE, p.value=P, Chr=CHR) %>%mutate(source=paste0(source, ': ', set))%>%dplyr::select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source, set)%>%filter(set %in% c('SSolder', 'SSnewbornCT'))%>%mutate(estimate=estimate*100, conf.low=100*conf.low, conf.high=100*conf.high)
 
@@
 
-blank_data2 <- cpg_all %>%dplyr::select(Gene, CpG, estimate, std.error)%>%group_by(Gene)%>%mutate(min=min(estimate))%>%dplyr::select(-estimate, -std.error)%>%unique()%>%mutate(estimate=min*-1+0.005, conf.high=estimate, conf.low=estimate)%>%dplyr::select(-min)%>%arrange(Gene)%>%mutate(source='Fragile Families: Age_9')
+blank_data2 <- cpg_all %>%dplyr::select(Gene, CpG, estimate, std.error)%>%group_by(Gene)%>%mutate(min=min(estimate))%>%dplyr::select(-estimate, -std.error)%>%unique()%>%mutate(estimate=min*-1, conf.high=estimate, conf.low=estimate)%>%dplyr::select(-min)%>%arrange(Gene)%>%mutate(source='Fragile Families: Age_9')
 
@@
   ggplot(aes(x=spec, y=sens, color=childteen, linetype=childteen))+geom_line(size=2)+
-  facet_wrap(~methylation, labeller = labeller(methylation=c('SSnewbornCT_center_scale'='Polymethylation score, \n newborn mean centered')))+
+  facet_wrap(~methylation, labeller = labeller(methylation=c('SSnewbornCT_center_scale'='Polymethylation score')))+
   theme_classic()+theme(legend.position = 'right')+guides(color=guide_legend(nrow=4,byrow=TRUE))+
@@
 
-auc_fig=cowplot::plot_grid(fig4a, fig4b, labels=c('A', 'B'), nrow=1, rel_widths=c(0.65, 0.35), align='h', axis='tb')
+auc_fig=cowplot::plot_grid(fig4a, fig4b, labels=c('A', 'B'), nrow=1, rel_widths=c(0.63, 0.37), align='h', axis='tb')
 
@@
 
-```{r main figure 4 auc print, eval=stand_alone, fig.width=10}
+```{r main figure 4 auc print, eval=stand_alone, fig.width=12}
 auc_fig
@@
 
-# Supplemental Table `r sup_tab`: Area under the curve for models shown in Main Figure 4 for classifying prenatal maternal smoke exposure, with *P* value when comparing to base model only or to model using polymethylation score only
+# Supplemental Table `r sup_tab`: Area under the curve for models shown in Main Figure 4 for classifying prenatal maternal smoke exposure, with *P* value when comparing to base model only or to model using polymethylation score + base model.
+Base model: Child sex, maternal income to poverty ratio, immune cell proportion, plate, 1st two principal components from PCA on child genetic data, child age.
 
@@
 roc_base=roc_table %>% filter(modeltype=='Base model+No methylation')%>%mutate(roc_basemodel=roc)%>%dplyr::select(childteen, roc_basemodel)
-roc_ss=roc_table %>% filter(modeltype=='Sustained smoking polymethylation score (newborns, cell-type controlled)')%>%mutate(roc_pms=roc)%>%dplyr::select(childteen, roc_pms)
+roc_ss=roc_table %>% filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)')%>%mutate(roc_pms=roc)%>%dplyr::select(childteen, roc_pms)
 
@@
 ```{r supplemental table auc print, eval=as_supp}
-rct1%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value vs base model', 'P value vs polymethylation score'))
+rct1%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value vs base model', 'P value vs base model + polymethylation score'))
 ```
 
-# Supplemental Table `r sup_tab`: Area under the curve when using different coefficient sets to construct polymethylation scores for predicting prenatal maternal smoke exposure
+# Supplemental Table `r sup_tab`: Area under the curve when using different coefficient sets to construct polymethylation scores for predicting prenatal maternal smoke exposure.
+With *P* value when comparing to base model only or to model using polymethylation score + base model.
+Base model: Child sex, maternal income to poverty ratio, immune cell proportion, plate, 1st two principal components from PCA on child genetic data, child age.
 
@@
 ```{r supplemental table auc other coefficients print, eval=as_supp}
-rct2%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value vs base model', 'P value vs polymethylation score with cell-type control'))
+rct2%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value vs base model', 'P value vs base model + polymethylation score with cell-type control'))
 ```

### 2021-11-10 Reports/Results.Rmd

@@
 include_ewas=F
-stand_alone=T
+stand_alone=F
 ```

### 2021-11-10 Reports/Paper1.Rmd

@@
 as_supp=T
-stand_alone=F
+stand_alone=T
 ```
@@
 load(paste0(datadir, '/CreatedData/methylation_summarymodels.Rdata'))
-```
-
-```{r plotting variables}
-theme_set(theme_bw())
-set_label(modeldata$cm1bsex)='Child gender'
-set_label(modeldata$cm1inpov)='Maternal income to poverty threshold (at baseline)'
-set_label(modeldata$PostnatalMaternalSmokingDose)='Primary caregiver smoking past month by dose'
-set_label(modeldata$SSnewbornCT_center)='Polymethylation score for smoke exposure'
-set_label(modeldata$pediatric)='Pediatric epigenetic clock (years)'
-set_label(modeldata$grim)='GRIM epigenetic clock (years)'
-set_label(modeldata$globalmethylation)='Percent global DNA methylation'
-set_label(modeldata$cg05575921)='AHRR gene: percent cg05575921 methylation'
 ```
@@
 sup_tab=1
+```
+
+```{r plotting variables}
+theme_set(theme_bw())
+set_label(modeldata$cm1bsex)='Child gender'
+set_label(modeldata$cm1inpov)='Maternal income to poverty threshold (at baseline)'
+set_label(modeldata$PostnatalMaternalSmokingDose)='Primary caregiver smoking past month by dose'
+set_label(modeldata$SSnewbornCT_center)='Polymethylation score for smoke exposure'
+set_label(modeldata$pediatric)='Pediatric epigenetic clock (years)'
+set_label(modeldata$grim)='GRIM epigenetic clock (years)'
+set_label(modeldata$globalmethylation)='Percent global DNA methylation'
+set_label(modeldata$cg05575921)='AHRR gene: percent cg05575921 methylation'
+set_label(modeldata$PostnatalMaternalSmokingAny)='Postnatal maternal smoking at ages 1 or 5'
+set_label(modeldata$SmkAtVisitPastmonth)='Postnatal primary caregiver smoking in past month prior to visit'
 ```
@@
   mutate(smkPreg_binary=factor(smkPreg_binary, levels=c('No', 'Yes'), labels=c('Not exposed', 'Smoke exposed')), 
-         cm1bsex=factor(cm1bsex, levels=c('1 Boy', '2 Girl', labels=c('Boy', 'Girl'))))%>%
+         cm1bsex=factor(cm1bsex, levels=c('1 Boy', '2 Girl'), labels=c('Boy', 'Girl')))%>%
+    mutate(cm1bsex=set_label(cm1bsex, 'Child gender'), smkPreg_binary=set_label(smkPreg_binary, 'Prenatal maternal smoking'),
+         childteen=set_label(childteen, 'Age at DNA methylation sample'))%>%
   tbl_strata(strata=childteen, .tbl_fun=~.x %>%

### 2021-11-11 Reports/Paper1.Rmd
modifications to labeling in supplemental figures

@@
 load(paste0(datadir, '/CreatedData/methylation_summarymodels.Rdata'))
+modeldata=modeldata %>% mutate(childteen=factor(childteen, levels=c('Age 9', 'Age 15')))%>%droplevels()%>%copy_labels(modeldata)
+```
+
+```{r plotting variables}
+theme_set(theme_bw())
+set_label(modeldata$cm1bsex)='Child gender'
+set_label(modeldata$cm1inpov)='Maternal income to poverty threshold (at baseline)'
+set_label(modeldata$PostnatalMaternalSmokingDose)='Primary caregiver smoking past month by dose'
+set_label(modeldata$SSnewbornCT_center)='Polymethylation score for smoke exposure'
+set_label(modeldata$pediatric)='Pediatric epigenetic clock (years)'
+set_label(modeldata$grim)='GRIM epigenetic clock (years)'
+set_label(modeldata$globalmethylation)='Percent global DNA methylation'
+set_label(modeldata$cg05575921)='AHRR gene: \n percent cg05575921 methylation'
+set_label(modeldata$PostnatalMaternalSmokingAny)='Postnatal maternal smoking at ages 1 or 5'
+set_label(modeldata$SmkAtVisitPastmonth)='Postnatal primary caregiver smoking in past month prior to visit'
+set_label(modeldata$childteen)='Visit of DNA methylation measure'
 ```
@@
 ```{r}
-modeldata=modeldata %>% mutate(childteen=factor(childteen, levels=c('Age 9', 'Age 15')))%>%droplevels()
-set_label(modeldata$childteen)='Visit'
-
 n_total=modeldata %>% dplyr::select(id)%>%unique()%>%nrow()
@@
 sup_tab=1
-```
-
-```{r plotting variables}
-theme_set(theme_bw())
-set_label(modeldata$cm1bsex)='Child gender'
-set_label(modeldata$cm1inpov)='Maternal income to poverty threshold (at baseline)'
-set_label(modeldata$PostnatalMaternalSmokingDose)='Primary caregiver smoking past month by dose'
-set_label(modeldata$SSnewbornCT_center)='Polymethylation score for smoke exposure'
-set_label(modeldata$pediatric)='Pediatric epigenetic clock (years)'
-set_label(modeldata$grim)='GRIM epigenetic clock (years)'
-set_label(modeldata$globalmethylation)='Percent global DNA methylation'
-set_label(modeldata$cg05575921)='AHRR gene: percent cg05575921 methylation'
-set_label(modeldata$PostnatalMaternalSmokingAny)='Postnatal maternal smoking at ages 1 or 5'
-set_label(modeldata$SmkAtVisitPastmonth)='Postnatal primary caregiver smoking in past month prior to visit'
 ```
@@
 
-pdf(file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')))
-corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8)
+pdf(file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=12)
+corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8, number.cex= 7/ncol(res2))
 dev.off()
@@
 ```{r supplemental correlation plot plot it , eval=as_supp, fig.width=12, fig.height=12}
-corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8)
+corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8, number.cex= 7/ncol(res2))
 
@@
 
-outcome_labels=c('Global methylation', 'Pediatric clock', 'AHRR: cg05575921',  'Polymethylation score \n (newborn, mean centered)')
+outcome_labels=c('Percent global  \n DNA methylation  (percent)', 'Pediatric epigenetic \n clock (years)', 'Percent DNA methylation of \n cg05575921 in AHRR (percent)',  'Polymethylation score for prenatal maternal \n smoke exposure (standard deviations)')
 
@@
   scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
-  scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'), labels=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))
+  scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'), labels=c('European genetic ancestry', 'Hispanic genetic ancestry', 'African genetic ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))
 
-ggexport(forest_plot, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')), width=15)
+ggexport(forest_plot, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')), width=19)
 main_fig=main_fig+1
@@
     scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
-    scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'), labels=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))
+    scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'), labels=c('European genetic ancestry', 'Hispanic genetic ancestry', 'African genetic ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))
 
@@
 covar=c('Sample_Plate', 'Slide', 'Array', 'cm1bsex', 'smkPreg_binary', 'childteen', 'ancestry')
-modeldata$Sample_Plate=set_label(modeldata$Sample_Plate, 'Batch')
+modeldata$Sample_Plate=set_label(modeldata$Sample_Plate, 'Sample Plate')
 modeldata$cm1bsex=set_label(modeldata$cm1bsex, 'Child gender')
@@
 
-plot_aprioriCpG<-ggplot(cpg_all, aes(x=estimate, xmin=conf.low, xmax=conf.high, color=source, y=CpG))+geom_point(position=position_dodge(width=1))+geom_errorbarh(height=0.1, position=position_dodge(width=1))+scale_x_continuous()+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+facet_wrap(.~Gene, scales='free', ncol=2)+theme_classic()+theme(legend.position = 'bottom')+guides(color=guide_legend(nrow=4,byrow=TRUE))+geom_blank(data = blank_data, aes(x = estimate))+scale_color_discrete('', labels=c('Fragile families saliva: age 15', 'Fragile families saliva: age 9', 'Fragile families: mixed effect model', 'Joubert et al.
+plot_aprioriCpG<-ggplot(cpg_all, aes(x=estimate, xmin=conf.low, xmax=conf.high, color=source, y=CpG))+geom_point(position=position_dodge(width=1))+geom_errorbarh(height=0.1, position=position_dodge(width=1))+scale_x_continuous()+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+facet_wrap(.~Gene, scales='free', ncol=2)+theme_classic()+theme(legend.position = 'bottom')+guides(color=guide_legend(nrow=4,byrow=TRUE))+geom_blank(data = blank_data, aes(x = estimate))+scale_color_discrete('', labels=c('Fragile families saliva: age 15', 'Fragile families saliva: age 9', 'Joubert et al.
 2016 newborn cord blood', 'Joubert et al.
@@
     filter(str_detect(methylation, 'SSnewbornCT|cg0557|global') | modeltype=='Base model')%>%
+  mutate(modeltype=gsub("Sustained smoking polymethylation score \\(newborns, cell-type controlled\\)", "Polymethylation score for prenatal maternal smoke exposure", modeltype))%>%
     ggplot(aes(x=spec, y=sens, color=modeltype))+geom_line()+

### 2021-11-11 Scripts/01_clean_phenotype_data.R
update directory based programming with here::here

@@ -9,11 +9,12 @@ library(purrr)
 library(tidyverse)
 library(dplyr)
 library(sjlabelled)
+library(here)
 
 #directory based coding
-datadir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
-outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output/"
-codedir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/"
+datadir<-gsub('Code', 'Data/', here::here())
+outputdir<-gsub('Code', 'Output/', here::here())
+codedir<-paste0(here::here(), '/')
 
 #read in data 
 pheno<-readRDS(paste0(datadir, "OGData/fullpheno.rds"))

### 2021-11-11 Scripts/02_make_methyl_summary.R

@@ -1,7 +1,9 @@
+library(here)
 #directory based coding
-datadir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
-codedir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Code/"
-clock_coefs="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/OGData/clock_coefs/"
+datadir<-gsub('Code', 'Data/', here::here())
+codedir<-paste0(here::here(), '/')
+clock_coefs=file.path(paste0(datadir, 'OGData'), 'clock_coefs')
+
 
 ####################################################################library
 library(dplyr)
@@ -89,9 +91,6 @@ save(polyscores, file = file.path(datadir, 'CreatedData', 'polymethylationscores
 polyscores_wide<-lapply(seq_along(polyscores), function(i) polyscores[[i]] %>% setNames(c(paste0(colnames(polyscores[[i]])[1:4], '_', names(polyscores)[i]), 'MethID')))%>%purrr::reduce(left_join, by='MethID')
 
 #################################################epigenetic clocks
-###################THIS NEEDS TO BE CHANGED -- clocks weren't calculated on johns betaqc but on jonahs, 
-#have different probe filter/sample filter sets
-#I've been working with Johns betaqc but using jonahs clocks. 
 if(nrow(betaqc)==423668){
   clocks<-read.csv(file=paste0(datadir, "OGData/ffcw_n1776_8clocks.csv"), header = T)
   colnames(clocks)[1]<-"MethID"
@@ -164,9 +163,8 @@ if(nrow(betaqc)!=423668){
 
 #################################################checks
 summary(methyldata)
-#note the three missing clock ids when left joining to clocks. Also see above. needs to make a decision here about what to do 
 
-#load in complete case analysis
+#load in complete case data set
 load(paste0(datadir, '/CreatedData/completeCasepheno.Rdata'))
 completecase=left_join(completecase, methyldata%>%mutate(idnum=as.character(idnum)))%>%
   mutate(childteen=case_when(childteen=='C'~'Age 9',childteen=='T'~'Age 15'))

### 2021-11-11 Scripts/03_sva_variables.R

@@ -8,7 +8,7 @@ library(stringr)
 library(sva)
 
 #directory based programming
-data.dir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/"
+data.dir=gsub('Code', 'Data/', here::here())
 output.dir=paste0(data.dir, 'CreatedData/')
 
 #load in betaqc matrix

### 2021-11-11 Scripts/04_summary_methyl_models.R

@@ -23,10 +23,12 @@ library(pROC)
 library(lmerTest)
 library(stringr)
 library(sva)
+library(here)
 #############
 ##################################Directories & data
-datadir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/CreatedData"
-outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Output"
+datadir=gsub('Code', 'Data/CreatedData', here::here())
+outputdir<-gsub('Code', 'Output', here::here())
+
 #read data 
 load(paste0(datadir, '/completeCasemethyl.Rdata'))
 modeldata=completecase%>%droplevels()%>%copy_labels(completecase)

### 2021-11-11 Reports/Paper1.Rmd
added a chunk option to run data instead of loading

@@
 stand_alone=T
+load_data=T
+run_scripts=F
 ```
 
-```{r temp read n data}
+```{r temp read n data, eval=load_data}
 basedir=here::here()
@@
 datadir=gsub('Code', 'Data', basedir)
+scriptdir=file.path(basedir, 'Scripts')
 load(paste0(datadir, '/CreatedData/allPhenoData.Rdata'))
@@
 modeldata=modeldata %>% mutate(childteen=factor(childteen, levels=c('Age 9', 'Age 15')))%>%droplevels()%>%copy_labels(modeldata)
+```
+
+```{r sources scripts instead of loading, eval=run_scripts}
+source(file.path(scriptdir, '01_clean_phenotype_data.R'))
+source(file.path(scriptdir, '02_make_methyl_summary.R'))
+#assuming 03_sva_variables.R has been run and exists separately as this would take a day to run. 
+source(file.path(scriptdir, '04_summary_methyl_models.R'))
 ```

### 2021-11-11 Reports/Archive/try1.Rmd
moved some reports to archive folder

--- 2021-09-07 Reports/FFCleaning.Rmd
@@
 
-# Lab meeting preface & necessary background
-
-**Purpose**: This presentation is a practice for presenting the preliminary results of the third aim of my dissertation to a committee member and important collaborator.
-I am seeking to get buy-in for splitting the analysis into two papers and begining to write the first paper.
-
-**Objectives & desired feedback**: 
-My objectives for this lab meeting are:
-
-1. Practice presenting results 
-2. Test drive presenting through Rmarkdown htmls
-
-And correspondingly, I'm seeking feedback on the logical flow of the presentation and its understandability.
-
-**Necessary background**:
-
-#  Objectives and  hypotheses
-
+# Objectives & Hypotheses 
  A) Conduct an ancestry-stratified epigenome wide association study of prenatal maternal smoking and DNA methylation from child saliva samples at age nine and age fifteen.
-
 B) Examine differences in methylation trajectories from age nine to fifteen of the most significant methylation sites identified in the epigenome-wide association study between those exposed vs unexposed to prenatal smoking.
-
 C) Develop a saliva epigenetic classifier of prenatal smoke exposure using a machine learning approach and support vector machines and evaluate the agreement of the classifier at two time points.
 
-Hypotheses:
-- I hypothesize that different CpG loci will be associated with maternal prenatal smoking in African ancestry groups than in European and Hispanic ancestry groups. 
-- I hypothesize that the direction of associations between CpGs and prenatal maternal smoking will stay consistent between ages nine and fifteen but that the magnitude of associations will decline over time.
-- Finally, I hypothesize that classification of prenatal maternal smoking status based upon DNA methylation information from salivary samples will be accurate and consistent at ages nine and fifteen.
-
-Additional
+Hypotheses: I hypothesize that different CpG loci will be associated with maternal prenatal smoking in African ancestry groups than in European and Hispanic ancestry groups.
+I hypothesize that the direction of associations between CpGs and prenatal maternal smoking will stay consistent between ages nine and fifteen but that the magnitude of associations will decline over time.
+Finally, I hypothesize that classification of prenatal maternal smoking status based upon DNA methylation information from salivary samples will be accurate and consistent at ages nine and fifteen.
 
@@
 ```{r, phenotypedataReadinLabel, include=F}
-pheno<-readRDS(paste0(datadir, "OGData/fullpheno.rds"))
-
-#Phenotype labeling & filtering to variables of interest
-#labels
-#to set labels using sjlabelled, needed to fix multibyte characters 
-#https://stackoverflow.com/questions/4993837/r-invalid-multibyte-string
-find_offending_character <- function(x, maxStringLength=256){  
-  print(x)
-  for (c in 1:maxStringLength){
-    offendingChar <- substr(x,c,c)
-    #print(offendingChar) #uncomment if you want the indiv characters printed
-    #the next character is the offending multibyte Character
-  }    
-}
-#lapply(attributes(pheno)$var.labels, find_offending_character)
-attributes(pheno)$var.labels<-gsub("\xad", "'",  attributes(pheno)$var.labels)
-#lapply(attributes(pheno)$var.labels, find_offending_character)
-attributes(pheno)$var.labels<-gsub("\x92", "'",  attributes(pheno)$var.labels)
-#lapply(attributes(pheno)$var.labels, find_offending_character)
-#define vars of interest 
-idvars<-c("id", "ff_id")
-demovars<-c("cm1age", "cm1bsex", "cm1ethrace", "cm1edu", "cm1hhinc", "cm1inpov", "m1h3", "m1h3a", "m1h3b", "m1i4", "m1i4a", "m1i4b", "cf1age", "cf1ethrace", "f1h3", "f1h3a", "f1h3b", "ch5agem","ch6yagem", "c6yagey", "cm5b_age", "cm5b_ageyrs", "cp6yagem", "cp6yagey", "hv5_agem", "hv6_yagem", "hv6_yagey")
-smokingvars<-c("m1g4", "f1g4", "f2j5", "f2j5a", "f2j7", "f2j7a", "f3j31", "f3j32", "f4j18", "f4j19", "f5g17", "f5g18", "k5f1k", "k5f1l", "k6d40", "k6d41", "k6d41", "k6d42", "k6d43", "k6d45", "k6d46", "k6d47", "m2j5", "m2j5a", "m2j7", "m2j7a", "m4j18", "m4j19", "m5g17", "m5g18", "n5f17", "n5f18", "p3a22", "p3a23", "p3a23a", "p3a23b", "p3a24", "p4a22", "p4a23", "p5h15", "p5h15b", "p5q3cr", "p6h74", "p6h75", "p6h76", "p6h77", "p6h78")
-prenataldrugusevar<-c("m1g2", "m1g3", "m1g5", "m1g6")
-
-#filter to these variables
-#FF has data in original form +sjlabelled value labels
-#since these dont' show in compareGroups
-#also will make FF_labeled, where categorical, binary variables
-#will be replaced with their labels
-#FF<-pheno %>% select(one_of(idvars), one_of(demovars), one_of(smokingvars), one_of(prenataldrugusevar))
-varIDX<-which(names(pheno)%in%c(idvars, demovars, smokingvars, prenataldrugusevar))
-FF<-pheno[,varIDX]
-varLabels<-attributes(pheno)$var.labels[varIDX]
-varLabels[59]<-"ID"
-FF<-set_label(FF, c(varLabels))
-#create FF_labeled
-#read in csv with metadata about variables 
-FFmeta<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/FFMetadata_v07.csv")
-#create long data dictionary  
-FFlong<-FFmeta%>% select(old_name, contains('value'), contains('label')) %>% pivot_longer(cols = contains("value"), names_to="value", values_to="old_values")
-FFlong<-FFlong%>%pivot_longer(cols = contains("label"), names_to="label", values_to="new_labels")
-recodes_long<-FFlong %>% filter(gsub("value", "", value)==gsub("label", "", label)& !is.na(old_values)) %>% select(old_name, old_values, new_labels) 
-#store continuous, binary, categorical etc
-FFmyMeta<-FFmeta %>% filter(old_name %in% c(idvars, demovars, smokingvars, prenataldrugusevar))
-myContinuous<-FFmyMeta %>% filter(type=="Continuous") %>%pull(old_name);
-myContinuous<-colnames(FF)[colnames(FF) %in% myContinuous]
-myBinary<-FFmyMeta%>%filter(type=="Binary")%>%pull(old_name); myBinary<-colnames(FF)[colnames(FF) %in% myBinary]
-myCategorical_O<-FFmyMeta%>%filter(type=="Ordered Categorical")%>%pull(old_name); myCategorical_O<-colnames(FF)[colnames(FF) %in% myCategorical_O]
-myCategorical_U<-FFmyMeta%>%filter(type=="Unordered Categorical")%>%pull(old_name); myCategorical_U<-colnames(FF)[colnames(FF) %in% myCategorical_U]
-#filter to only vars of interest - not continuous variables  
-myRecodes<-recodes_long %>% filter(old_name %in% c(myBinary, myCategorical_O, myCategorical_U))
-#apply matchmaker
-FF_labeled<-match_df(FF, dictionary=myRecodes, from = 'old_values', to='new_labels', by='old_name')
-#reapply correct variable class types and na_codes
-na_codes<-c("-1 Refuse", "-2 Don't know", "-3 Missing", "-9 Not in wave")
-FF_labeled<-FF_labeled %>% mutate_at(vars(any_of(c(myBinary, myCategorical_O, myCategorical_U))), funs(as.factor(ifelse(. %in% na_codes, "Missing", .)))) %>% 
-          mutate_at(vars(any_of(myContinuous)), funs(as.numeric(na_if(na_if(na_if(na_if(., "-3"), '-9'), '-6'), '-2'))))
-#read back in the variable labels 
-FF_labeled<-set_label(FF_labeled, get_label(FF))
-FF_labels<-FF_labeled
-```
-
-## Cleaning pdqc data 
-### Figure out analytic subsets 1) any methylation data; 2) 2-visits of methylation data
-```{r, pdqcReadin}
-pdqc_all<-readRDS(paste0(datadir, 'OGData/', "pd_qc.rds"))
-pdqc<-pdqc_all %>% filter(probe_fail_10==0 & sex==predicted_sex)
-#What FF ids have methylation data?
-methy_ids<-pdqc$id
-#create slide variable fo pdqc 
-pdqc$slide <- gsub('_.*', '', pdqc$MethID)
-pdqc$array<-gsub('.*_', '', pdqc$MethID)
-#ids with 9visit, 15visit or both visit
-visit9ids<-pdqc%>%filter(childteen=='C')%>%pull(idnum)%>%unique()
-visit15ids<-pdqc%>%filter(childteen=='T')%>%pull(idnum)%>%unique()
-bothvisit_ids<-visit9ids[visit9ids %in% visit15ids]
-visit9only_ids<-visit9ids[!(visit9ids %in%bothvisit_ids)]
-visit15only_ids<-visit15ids[!(visit15ids %in%bothvisit_ids)]
-```
-
-```{r, createAnalyticSubsetFlags}
-#create flag in FF for having methylation data
-#and flag for having 2 visits of methylation data
-FF_labeled<-FF_labeled %>% 
-  mutate(MethylData=ifelse(id %in% methy_ids, "Analysis subset", "Not in analysis subset"),
-         TwoVisitData=ifelse(id%in%bothvisit_ids,"Analysis subset - both visits", "Not in both visit analysis subset"))
-table(FF_labeled$MethylData)
-table(FF_labeled$TwoVisitData)
-```
-
-```{r, checkIDlengths}
-if((length(visit15only_ids)+length(visit9only_ids)+length(bothvisit_ids)!=length(FF_labeled %>% filter(MethylData=="Analysis subset")%>%pull(id)))){print("WARNING ID LENGTHS DO NOT ADD UP")}
-if(length(bothvisit_ids)!=length(FF_labeled %>% filter(TwoVisitData=="Analysis subset - both visits")%>%pull(id))){print("WARNING ID LENGTHS DO NOT ADD UP")}
-```
-
-### Filter technical replicates (only 1 technical replicate per person-visit + decide based on QC variables - pick the one with the lower probe fail percentage)
-
-```{r}
-#technical replicates will be duplicate id + childteen variables from pdqc
-pdqc<-pdqc %>% mutate(newID=paste(idnum, childteen, sep='_'))
-techrep_ids<- pdqc %>% select(newID) %>% group_by(newID)%>%filter(n()>1)
-techreps<-pdqc %>% filter(newID %in% techrep_ids$newID)%>%arrange(idnum,childteen)
-hiPctPF<-techreps %>% group_by(newID) %>%filter(probe_fail_pct==max(probe_fail_pct))%>%pull(MethID)
-pdqc_clean<-pdqc %>% filter(!(MethID %in% hiPctPF) & childteen!='M')
-myMethIDs<-pdqc_clean %>% pull(MethID)
-```
-
-## Methylation summary variables
-
-### Cell type proportions
-
-Cell type proportions estimated using [epidDISH](https://www.bioconductor.org/packages/release/bioc/vignettes/EpiDISH/inst/doc/EpiDISH.html#how-to-estimte-cell-type-fractions-using-dnam-data) package and the EpiFibIC reference (generic for epithelial tissues).
-
-```{r read in betaqc & celltype proportions}
-betaqc<-readRDS(file=paste0(datadir, 'OGData/', "noob_filtered.rds"))
-aprioriCG<-c("cg05575921", "cg04180046", "cg05549655", "cg14179389")
-globalmethy<-colMeans(betaqc, na.rm=T)
-globalmethdf<-as.data.frame(cbind("MethID"=names(globalmethy), "globalmethylation"=globalmethy))
-aprioriCGdf<-as.data.frame(t(betaqc[aprioriCG, ]))
-aprioriCGdf$MethID = colnames(betaqc)
-#cell type proportions 
-celltypes<-epidish(beta.m = betaqc, ref.m = centEpiFibIC.m, method = "RPC")
-estF_FF<-as.data.frame(celltypes$estF)
-estF_FF$MethID = row.names(estF_FF)
-```
-
-### Clocks
-
-```{r, readInclocks}
-clocks<-read.table(file=paste0(datadir, "OGData/DNAm_clocks_ffcw.txt"), header = T)
-colnames(clocks)[1]<-"MethID"
-```
-
-### Global methylation score
-
-```{r make polymethylaton score}
-#read in cpgs from PMC4833289
-cpgs<-read.csv(paste0(datadir, 'PMC4833289_SuppFile2CpGs.csv'), stringsAsFactors = F, header=F, skip=2)
-#make headers
-header1<-scan(paste0(datadir, 'PMC4833289_SuppFile2CpGs.csv'), nlines=1, what=character(), sep=',')
-header1[7:26]<-c(rep("SSnewborn", 5), rep("SSnewbornCT", 5), rep("SSolder", 5), rep("anynewborn", 5))
-header2<-paste(header1, scan(paste0(datadir, 'PMC4833289_SuppFile2CpGs.csv'), nlines=1, skip=1, what=character(), sep=','), sep='_')
-names(cpgs)<-header2
-#four meta-EWASes: SS_newborn (sustained smoking in newborns); SS_newbornCT (sustained smoking in newborns controlling for cell type) SS_older (sustained smoking in older children); any_newborn (any smoking in newborns)
-#pivot_longer to get single column of coefs and a column for the analysis type
-#and group into a list by analysis type
-cpgs<-cpgs %>% pivot_longer(cols = c(matches("_Coef|_SE|_P|_FDR|_Direction")), names_to = c("set", ".value"), names_pattern="(.+)_(.+)")%>%rename_with(~gsub('_', '', .x))%>%group_split(set)
-cpgs<-cpgs%>%setNames(lapply(cpgs, function(x){x$set %>% unique()}))
-#make scores - no transform, zscore and centermean
-#first, pull script 
-source(file.path(codedir, 'UsefulCode', 'makePolyEpiScores.R'))
-scores=makePolyEpiScore(m=betaqc, b=cpgs)
-scores_z=makePolyEpiScore(m=betaqc, b=cpgs, transformopt = 'zscore')
-scores_center=makePolyEpiScore(m=betaqc, b=cpgs, transformopt = 'meancenter')
-polyscores=list('notransform'=scores, 'zscore'=scores_z, 'center'=scores_center)
-polyscores<-lapply(polyscores, function(x) x%>% mutate(MethID=colnames(betaqc)))
-save(polyscores, file = file.path(datadir, 'CreatedData', 'polymethylationscores.RDS'))
-polyscores_wide<-lapply(seq_along(polyscores), function(i) polyscores[[i]] %>% setNames(c(paste0(colnames(polyscores[[i]])[1:4], '_', names(polyscores)[i]), 'MethID')))%>%reduce(left_join, by='MethID')
-```
-
-```{r make methyl dataframe}
-clocks2<-clocks %>% mutate(idnum=gsub('a', '', idnum))%>%filter(MethID %in% pdqc$MethID) # currently clocks are missing 3 individuals who were in johns pipeline but not jonahs
-
-methyldata<-left_join(left_join(left_join(left_join(left_join(pdqc_clean %>% select(MethID, slide, array,childteen, idnum), globalmethdf), clocks2), estF_FF), aprioriCGdf), polyscores_wide)
-methyldata[, c("globalmethylation", aprioriCG)]<-lapply(methyldata[, c("globalmethylation", aprioriCG)],function(x) as.numeric(as.character(x)))
-#lapply(methyldata, class)
-methyldata_wide <- methyldata %>% select(-MethID)%>% pivot_wider(., names_from=childteen, values_from=any_of(colnames(methyldata)[4:length(colnames(methyldata))]))
-```
-
-```{r individual who is missing global methylation data but had apriori CpG and PMS data?, eval=F}
-#weirdID<-methyldata %>% filter(is.na(globalmethylation))%>%pull(MethID)
-#globalmethy[weirdID] #not actually missing global methylation data
-#clocks %>% filter(MethID%in%weirdID) # #idnum from  not the same as in
-#pdqc_clean %>% filter(MethID==weirdID) #this is just a typo in the idnum from data entry, see format of other #idnums
-#pdqc_clean %>% pull(idnum) %>% head()
-#solution: when creating methyldata, first trim alphabet from clocks$idnum that contain alphabetic characters --> implemented
-#also misssing clocks for 3 sex outlier samples that weren't in jonahs' pipeline: 
-table(is.na(methyldata$horvath13))
-weirdIDs<-methyldata %>% filter(is.na(horvath13))%>%pull(MethID)
-pdqc_all %>% filter(MethID %in% weirdID) %>% xtabs(~predicted_sex_outlier, data=.)
-```
-
-## Creation of new variables
-### Binary smoking variable
-```{r}
-FF_labeled<-FF_labeled %>%mutate(smkPreg_binary=ifelse(m1g4%in% c("1 2+pk/d","2 1<pk<2","3 <1pk/d" ), "Yes", ifelse(m1g4%in%c("-2 Don't know", "-3 Missing", "-9 Not in wave"), NA, ifelse(m1g4== "4 None", "No", ifelse(m1g4=="Missing", "Missing", "FLAG")))))
-#xtabs(~m1g4+smkPreg_binary, data=FF_labeled)
-```
-
-### Ancestry variables
-
-Ancestry is based on csv files of ancestry-specific PCS created by Erin (per Jonah meeting Oct 6 2020). If an ID was in {african/european/hispanic}, then the individual is labeled as such. 
-```{r}
-# read in PCs
-euroPC<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/european.csv")
-afrPC<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/african.csv")
-hisPC<-read.csv("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/hispanic.csv")
-allPC<-read.table("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/pcs/global_pcs.txt", col.names=c("X", "idnum", paste("PC", 1:20, sep='')))
-#individuals in each ancestry specific pc group can be labeled as that ancestry for categorical ancestry variable 
-FF_labeled<-FF_labeled %>% mutate(ancestry=case_when(MethylData=="Analysis subset" & id %in% hisPC$idnum~"Hispanic ancestry", MethylData=="Analysis subset" & id %in%afrPC$idnum~"African ancestry", MethylData=="Analysis subset" & id %in% euroPC$idnum~"European ancestry", MethylData=="Analysis subset" & !(id %in%c(hisPC$idnum, afrPC$idnum, euroPC$idnum)) & (id %in% allPC$idnum)~"Other ancestry", MethylData=="Analysis subset" & !(id %in%c(hisPC$idnum, afrPC$idnum, euroPC$idnum)) & !(id %in% allPC$idnum)~"Missing PC data", MethylData!="Analysis subset"~"Not in analysis subset"))
-#check ancestry variable
-#table(FF_labeled$ancestry)
-#with(FF_labeled, xtabs(~ancestry+cm1ethrace))
-MissingPCdata<-FF_labeled %>% filter(ancestry=="Missing PC data")%>%select(id)
-save(MissingPCdata, file = "/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/CreatedData/MissingPC.Rdata")
-#add PC data to FF_labeled 
-FF_labeled<-left_join(FF_labeled, allPC %>% mutate(id=as.character(idnum)) %>% select(id, PC1, PC2)%>%rename_at(vars(-id), function(x) paste0('global_', x, sep='')))
-localPC<-rbind(euroPC, afrPC, hisPC)
-FF_labeled<-left_join(FF_labeled, localPC  %>% mutate(id=as.character(idnum)) %>% select(id, PC1, PC2)%>% rename_at(vars(-id), function(x) paste0('local_', x, sep='')))
-```
-
-There were `r length(MissingPCdata$id)` individuals without PC data.
-None of these individuals had data in the global PC file, so I assume they are actually missing genetic data (for whatever reason, possibly QC).
-
-### Prenatal exposure variables 
-Recoded frequency of prenatal alcohol drinking and drug use into Yes/no binary variables for drinking and drug use during pregnancy. 
-```{r PrenatalExposure}
-#table(FF_labeled$m1g2) # during preg, alchohol freq
-#table(FF_labeled$m1g3) # during preg, drugs freq 
-with(FF_labeled, table(m1g2, m1g3))
-FF_labeled<-FF_labeled %>% mutate_at(c("m1g2", "m1g3"), .funs=funs(YesNoPreg=ifelse(.=="5 Never", "No", ifelse(. =="Missing", "Missing", "Yes"))))
-with(FF_labeled, table(m1g2, m1g2_YesNoPreg))
-with(FF_labeled, table(m1g3, m1g3_YesNoPreg))
-#table(FF_labeled$m1g5) # in last year, alc/drugs interfered with work/relationships
-#table(FF_labeled$m1g6) # ever sought help for drug/alc problems
-```
-## Postnatal smoke exposure 
-Maternal postnatal smoking at different survey points is fairly correlated, and alluvial plots of changes over time show broad consistency, so I decided to create a single smoking variable for maternal smoking at age 1 and age 5 - if mothers were smoking at either age 1 or age 5 then "Maternal smoking at age 1 or 5", if not smoking at age 1 and age 5 then "No maternal smoking at age 1 and age 5" and then if missing data at one age and no at the other, then "Missing" (and if missing at both ages, "Missing").
-Also created a similar dose variable capturing maternal packs per day that could be used instead.
-Decided to use just age 1 and age 5 because at these two time points the mother was asked (as opposed to PCG, also the dataset of Fragile Families metadata is missing PCG questions from age 3 about smoking) and they both preceed the actual saliva collection for both saliva collection age points.
-Then use a smoking dose variable from the mother or PCG interview at wave of saliva collection as an additional second hand smoke exposure variable.
-
-```{r PostnatalSmoke}
-#Baseline (1) -> Year 1/Age 1 (2) ->Year 3/Age3 (3) -> Year 5 (4) ->Year 9 (5) -> year 15 (6)
-#Do you smoke in past month (how many packs per day)
-#Mom            ->m2j5 (m2j5a)    ->                  -> m4j18 (m4j19) /p4 -> m5g17 (m5g18) / n5f17 (n5f18) 
-#PCG                              p3a23a(p3a23b)                              n5f17 (n5f18)-> p6h76 (p6h77)
-#seem to be missing p3 variables (PCG survey year 3) 
-#number of people in household who smoke 
-#p3a23->p5h15b->p6h78
-
-#hours child around smoke 
-#p3a21 -> p4a24->p5h15c
-
-#with(FF_labeled, table(m2j5, m4j18))
-#with(FF_labeled, table(m4j18, m5g17))
-#with(FF_labeled, table(n5f17, m5g17))
-#with(FF_labeled, table(p6h76, m5g17))
-
-plot_grid(FF_labeled %>% select(m2j5, m4j18, m5g17, n5f17, p6h74) %>% mutate_all(funs(as.numeric(.))) %>%ggcorr(label=T)+ggtitle("Correlation between Yes/No smoking (mother) \n at years 1, 5, & 9 & PCG at years 9 & 15")
-,FF_labeled %>% select(m2j5a, m4j19, m5g18, n5f18, p6h77) %>% mutate_all(funs(as.numeric(.))) %>%ggcorr(label=T)+ggtitle("Correlation between packs per day smoking (mother) \n at years 1, 5, 9 & PCG at years 9 and 15"), nrow=2)
-
-yesnosmkalluv<-to_lodes_form(as.data.frame(FF_labeled %>% select(m2j5, m4j18, m5g17, p6h74) %>% mutate(m5g17=stringr::str_to_title(m5g17))%>% group_by(m2j5, m4j18, m5g17, p6h74) %>% count()), axes=1:4, id='Cohort')
-
-#ggplot(yesnosmkalluv, aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow(stat="alluvium", lode.guidance="backfront", color='darkgrey')+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9"))
-
-smkdose<-to_lodes_form(as.data.frame(FF_labeled %>% select(m2j5a, m4j19, m5g18, p6h77)%>% mutate_at(vars(c("m2j5a", "m4j19", "m5g18")), funs(factor(substring(., 1, 1), labels=levels(FF_labeled$m2j5a)))) %>% group_by(m2j5a, m4j19, m5g18, p6h77) %>% count()), axes=1:4, id='Cohort')
-#ggplot(smkdose, aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow()+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9"))
-
-plot_grid(ggplot(yesnosmkalluv, aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow()+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9", "Age 15"))
-+ggtitle("Alluvial plot of Yes/No smoking variables from age 1 to age 9 (mother) and age 15 (PCG)"), ggplot(smkdose %>% filter(!(stratum %in% c("-6 Skip", "Missing"))), aes(x=x, stratum = stratum, alluvium=Cohort, fill=stratum, label=stratum, y=n))+geom_flow()+geom_stratum()+scale_x_discrete(labels=c("Age 1", "Age 5", "Age 9", "Age 15"))+ggtitle("Alluvial plot of smoking per day among those smoking from age 1 to age 9 (mother) and 15 (PCG)"), nrow=2)
-
-#ggplot(FF_labeled, aes(x=p5h15b, y=p6h78))+geom_jitter()+stat_cor()+scale_x_continuous('# smoking in house age 9')+scale_y_continuous('# smoking in house age 15')+ggtitle("# individuals smoking in house correlation age 9 and age 15")
-
-#Make any maternal postnatal smoking variable & dose categorization for maternal smoking at ages 1 & 5 
-FF_labeled<-FF_labeled %>% mutate(PostnatalMaternalSmokingAny = case_when(
-                                m2j5 =="1 Yes" ~ "Maternal smoking at age 1 or age 5",
-                                m4j18=="1 Yes" ~ "Maternal smoking at age 1 or age 5",
-                                m2j5 == "Missing" & m4j18=="Missing" ~ "Missing",
-                                m2j5 == "Missing" & m4j18=="2 No" ~ "Missing",
-                                m2j5 == "2 No" & m4j18=="Missing" ~ "Missing",
-                                m2j5 == "2 No" & m4j18=="2 No" ~ "No maternal smoking at age 1 and age 5")) %>%
-  mutate(PostnatalMaternalSmokingDose=case_when(
-                                m2j5a %in% c("2 1pk/d", "3 1.5pk/d", "4 2pk/d", "5 >2pk/d") &
-                                m4j19 %in% c("2 About pack/day", "3 Pack and half/day", "4 2 packs/day", "5   
-                                          More 2 packs/day") ~ "Mom - pack/d or greater at both age 1 & 5", 
-                                m2j5a %in% c("2 1pk/d", "3 1.5pk/d", "4 2pk/d", "5 >2pk/d") &
-                                !(m4j19 %in% c("2 About pack/day", "3 Pack and half/day", "4 2 packs/day", "5 
-                                          More 2 packs/day")) ~ "Mom - pack/d or greater at either age 1 or 5",
-                                !(m2j5a %in% c("2 1pk/d", "3 1.5pk/d", "4 2pk/d", "5 >2pk/d")) &
-                                m4j19 %in% c("2 About pack/day", "3 Pack and half/day", "4 2 packs/day", "5 
-                                          More 2 packs/day") ~ "Mom - pack/d or greater at either age 1 or 5",
-                                m2j5a %in% c("1 <1/2 pk/d") | m4j19 %in% c("1 Less half pack/day")~"Mom - less than half pack/d at either age 1 or 5", 
-                                m2j5a %in% c("-6 Skip") & m4j19 %in% c("-6 Skip")~
-                                  "No maternal smoking at age 1 or 5", 
-                                m2j5a %in% c("Missing") & m4j19 %in% c("-6 Skip", "Missing")~"Missing", 
-                                m2j5a %in% c("Missing", "-6 Skip") & m4j19 %in% c("Missing")~"Missing"))
-
-with(FF_labeled, table(PostnatalMaternalSmokingAny, m2j5, m4j18))
-#with(FF_labeled, table(m2j5a, m4j19, PostnatalMaternalSmokingDose))
-#with(FF_labeled, table(PostnatalMaternalSmokingAny, PostnatalMaternalSmokingDose))
-```
-
-Maybe here best to have 1 variable capturing any maternal postnatal smoking at age 1 & 5 and then use in each crossectional model the dose of maternal postnatal smoking at current visit (and similarly for linear mixed effect model)?
-
-```{r makeMyFF}
-varLabels2<-c(paste(colnames(FF_labeled)[1:66], varLabels), "Has any methylation data?", "Has two visits of methylation data", "Any maternal smoking during pregnancy", "Ancestry from PCA of child's genetic data", "PC 1 (global)", "PC 2 (global", "PC 1 (within ancestry category)", "PC 2 (within ancestry category",
-"Any alcohol during pregnancy", "Any drugs during pregnancy", "Any postnatal smoking (age 1 & 5)", "Postnatal smoking dose at age 1 & 5")
-FF_labeled<-set_label(FF_labeled, varLabels2)
-
-myFF<-FF_labeled %>% filter(MethylData=="Analysis subset")%>%mutate(idnum=id)
-myFF<-left_join(myFF, methyldata)
-varLabels3<-c(varLabels2, "ID", "Methylation ID", "Slide", "Array (RC)", "Visit", "Global methylation", "horvath13 methylation clock", "horvath18 methylation clock", "pediatric methylation clock", "levine methylation clock", "Epithelial cell proportion", "Fibroid cell proportion", "Immune cell proportion", "cg05575291 methylation", "cg04180046 methylation",  "cg05549655 methylation", "cg14179389 methylation", "PES: Any smoking, newborn cordblood", "PES: Sustained smoking, newborn cordblood", "PES: Sustained smoking, newborn cordblood (+CT control)", "PES: Sustained smoking, older children peripheral blood", "PES: Any smoking, newborn cordblood - zscore", "PES: Sustained smoking, newborn cordblood - zscore", "PES: Sustained smoking, newborn cordblood (+CT control) - zscore", "PES: Sustained smoking, older children peripheral blood - zscore", "PES: Any smoking, newborn cordblood - mean center", "PES: Sustained smoking, newborn cordblood - mean center", "PES: Sustained smoking, newborn cordblood (+CT control) - mean center", "PES: Sustained smoking, older children peripheral blood - mean center")
-myFF<-set_label(myFF, varLabels3)
-```
-### Child smoking
-Exclude children who reported smoking at age 9 and age 15.
-Interestingly, the children who report smoking at age 9 do not report smoking at age 15, and the parental perception of child smoking at age 9 does not match the children who say they have ever smoked a cigarrette.
-For now using the 'ever smoked' variable from age 15 for the exclusion, but there is also a 'smoked in past month' that we could use instead (lose fewer children).
-
-```{r, ChildSmoke}
-with(myFF, table(k5f1l,childteen)) # 6 kids smoking at age 9 
-with(myFF, table(p5q3cr, k5f1l)) # not the same kids whose parents say somewhat or very true that the child is using tobacco
-with(myFF, table(k6d40, childteen)) # 41 kids say ever smoked a cigarette at age 15
-#with(myFF, summary(k6d41))
-with(myFF, table(k6d42, childteen)) # of these only 13 say they've smoked in the past month 
-#with(myFF, summary(k6d43))
-with(myFF, table(k5f1l, k6d40, childteen))# not the same kids at age 9 and 15
-```
-
-### Maternal or PCG smoking dose at time of sample
-
-```{r visitMaternalSmoke}
-myFF<-myFF %>% mutate(SmkAtVisitPastmonth=case_when(childteen=='C' & m5g17=='1 yes' & m5g18 %in% c("2 about a pack", "3 a pack and a half", "4 about 2 packs", 
-                              "5 more than two  packs") ~ "Pack or more a day", 
-                              childteen=='C' & m5g17=='1 yes' & m5g18=="1 less than half a pack a day" ~ "Less than pack a day", 
-                              childteen=='C' & n5f17=="1 yes" & n5f18 =='2 about a pack' ~ "Pack or more a day",
-                              childteen=='C' & n5f17=="1 yes" & n5f18 =="1 less than half a pack day" ~ "Less than pack a day",
-                              childteen=='C' & m5g17=="1 yes" & m5g18 %in% c("-6 Skip", "Missing")~"Missing",
-                              childteen == 'C' & n5f17%in%c("-7 N/A", "2 no") & m5g17=="2 no" ~ "No smoking", 
-                              childteen == 'C' & n5f17%in%c("2 no") & m5g17%in%c("2 no", "Missing") ~ "No smoking", 
-                              childteen == 'C' & n5f17 %in% c("Missing", "-7 N/A") & m5g17 =="Missing" ~ "Missing", 
-                              childteen=='T' & p6h76 %in% c("-6 Skip", "0 None") ~ "No smoking", 
-                              childteen=='T' & p6h77%in% c("1 5 or fewer cigarettes per day", "2 About half a pack a day, 10 cigarettes")~"Less than pack a day", 
-                              childteen=='T' & p6h77 %in% c("3 About a pack a day, 20 cigarettes", "4 More than 1 pack a day")~"Pack or more a day",
-                              childteen=='T' & p6h77 == "Missing"~"Missing"))%>%
-              mutate(SmkAtVisitYesNo=case_when(SmkAtVisitPastmonth=="No smoking"~"No", 
-                                               SmkAtVisitPastmonth %in% c("Less than pack a day", "Pack or more a day") ~ "Yes", 
-                                               childteen=='T'& SmkAtVisitPastmonth=="Missing" & !(p6h76 %in% c("-6 Skip", "0 None", "Missing"))~"Yes",
-                                               childteen=='C' & SmkAtVisitPastmonth=="Missing" & m5g17=='1 yes'~"Yes", 
-                                               childteen=='C' & SmkAtVisitPastmonth=="Missing" & m5g17!='1 yes'~"Missing", 
-                                               childteen=='T'& SmkAtVisitPastmonth=="Missing" & p6h76 %in% c("-6 Skip", "0 None", "Missing")~"Missing"))
-
-myFF %>% filter(childteen=='C') %>% xtabs(~SmkAtVisitPastmonth+SmkAtVisitYesNo, data=.)
-myFF %>% filter(childteen=='T') %>% xtabs(~SmkAtVisitPastmonth+SmkAtVisitYesNo, data=.)
-myFF$SmkAtVisitPastmonth<-factor(myFF$SmkAtVisitPastmonth)
-```
-### Child's age at time of interview/visit
-```{r}
-myFF %>% filter(is.na(cp6yagem) &is.na(hv6_yagem) & childteen=='T')%>%dim() #3
-myFF %>% filter(is.na(cm5b_age) &is.na(hv5_agem) &childteen=='C')%>%dim() #1 
-myFF %>% filter(!is.na(cm5b_age) & is.na(hv5_agem) & childteen=='C')%>%dim()#1
-myFF %>% filter(!is.na(cp6yagem) &is.na(hv6_yagem))%>%dim() #990
-
-myFF <- myFF %>% 
-  mutate(ChildAgeAtIn=case_when(childteen=='C'~ cm5b_age, childteen=='T'~cp6yagem))%>%
-  mutate(ChildAgeAtVisit=case_when(childteen=='C'~hv5_agem, childteen=='T'~hv6_yagem))%>% 
-  mutate(ChildAgeComposite=case_when(childteen=='C' & !is.na(hv5_agem)~hv5_agem, 
-         childteen=='C' & is.na(hv5_agem)~cm5b_age, 
-         childteen=='T' & !is.na(hv6_yagem)~hv6_yagem, 
-         childteen=='T' & is.na(hv6_yagem)~cp6yagem))
-
-myFF %>% filter(childteen=='C')%>%group_by(ChildAgeAtVisit, ChildAgeComposite)%>%summarise(n=n())%>%filter(ChildAgeAtVisit!=ChildAgeComposite)
-
-myFF %>% filter(childteen=='T')%>%group_by(ChildAgeAtVisit, ChildAgeComposite)%>%summarise(n=n())%>%filter(ChildAgeAtVisit!=ChildAgeComposite)
-
-ggplot(myFF%>%filter(childteen=='C'), aes(cm5b_age, hv5_agem))+geom_point()+stat_cor()+ggtitle("Correlation between childs age at mom interview (9 yr) and childs age at home visit (9 yr)")
-age5resid<-myFF %>% filter(childteen=='C')%>%mutate(age5_resid=cm5b_age-hv5_agem)%>%pull(age5_resid)
-hist(age5resid)
-
-age6resid<-myFF %>% filter(childteen=='T')%>%mutate(age6_resid=cp6yagem-hv6_yagem)%>%pull(age6_resid)
-ggplot(myFF%>%filter(childteen=='T'), aes(cp6yagem,hv6_yagem))+geom_point()+stat_cor()+ggtitle("Correlation between childs age at mom interview (15 yr) and childs age at home visit (15 yr)")
-hist(age6resid)
-
-table(FF_labeled$cp6yagem-FF_labeled$hv6_yagem)
-table(FF_labeled$cm5b_age-FF_labeled$hv5_agem)
-table(is.na(myFF$ChildAgeComposite)) # 4
-
-varLabels4<-c(varLabels3, "Mother/PCG (9/15 yr) smoking in month prior to interview (pk/day)", "Mother/PCG (9/15 yr) smoking in month prior to interview (yes/no)", "Child's age at maternal/PCG (9/15 yr) interview (months)", "Child's age at home vist (9/15 years) (months)", "Constructed child's age at home visit (in months) (9/15 years)")
-myFF<-set_label(myFF, varLabels4)
-```
-I also found two variables for child's age at home visit, but the variable for child's age at home visit for 15 years has a large amount of missingness in it (missing n=`r myFF %>% filter(is.na(hv6_yagem) & childteen=='T')%>%nrow()`).
-The variable for child's age at home visit for 9 year visit is better (missing n=`r myFF %>% filter(is.na(hv5_agem) & childteen=='T')%>%nrow()`).
-The interview age variables correlate well with the home visit age variables.
-For the 9 year interview/home visit ages, there are 5 individuals with >12 month difference between the ages, for the 15 year visit, there are 0 (and actually, there is 0 difference for any of the individuals with nonmissing home visit ages).
-So one option would be to use the home visit ages for the 9 year visit and for individuals with home visit age data available for the 15 year visit, supplementing the 15 year visit age variable with  the 15 year PCG interview age visit for those missing the home visit age.
-Alternatively we could use baby birthdate and saliva colleciton date to create an age variable.
-I have the baby DOB variable, but I don't have/don't know where a saliva sample collection date would be.
-
-# Analysis
-
-## Randomization of prenatal maternal smoking across slide numbers
-
-```{r slide number vs prenatal maternal smoking}
-slide_count<-myFF %>% mutate(slide=as.factor(slide))%>% group_by(smkPreg_binary, slide, .drop=F)%>%count()%>%pivot_wider(id_cols = slide, names_from=smkPreg_binary, values_from=n)%>%arrange(Yes)%>%mutate(Proportion=Yes/(No+Yes))
-hist_data<-slide_count %>% ggplot(aes(x=Proportion))+geom_histogram(binwidth = 0.1, fill='blue', alpha=0.5, breaks=seq(0, 1, 0.1))+ggtitle('Histogram proprotion smk (from data)')
-hist_data_c<-slide_count %>% ggplot(aes(x=Yes))+geom_histogram( fill='blue', alpha=0.5, breaks=c(0:12))+scale_x_continuous(breaks=c(0:12))+ggtitle('Histogram # Yes smk (from data)')
-
-kable(slide_count)%>%kable_styling()%>%scroll_box(width='100%', height='400px')
-
-#simulation
-set.seed(seed=20210216)
-n<-nrow(myFF)
-yes<-table(myFF$smkPreg_binary)[2]; no<-table(myFF$smkPreg_binary)[1]
-values<-data.frame('smk'=sample(c('No', 'Yes'), n, replace=T, prob=c(no/(no+yes), yes/(no+yes))))
-values_slides<-values %>% 
-    group_by((row_number()-1) %/% (n()/152)) %>%
-    nest()%>%unnest(cols=c(data))
-colnames(values_slides)<-c('slide', 'smk')
-slide_sim<-values_slides%>%mutate(slide=as.factor(slide)) %>% group_by(smk, slide, .drop=F)%>%count()%>%pivot_wider(id_cols = slide, names_from=smk, values_from=n)%>%arrange(Yes)%>%mutate(Proportion=Yes/(No+Yes))
-hist_sim<-slide_sim%>% ggplot(aes(x=Proportion))+geom_histogram(binwidth = 0.1, fill='red', alpha=0.5, breaks=seq(0, 1, 0.1))+ggtitle('Histogram proprotion smk (from simulation)')
-hist_sim_c<-slide_sim%>% ggplot(aes(x=Yes))+geom_histogram(fill='red', alpha=0.5, breaks=c(0:12))+scale_x_continuous(breaks=c(0:12))+ggtitle('Histogram # Yes smk (from simulation)')
-
-hist_overlay<-slide_count %>% ggplot(aes(x=Proportion))+geom_histogram(fill='blue', binwidth = 0.1, alpha=0.5, breaks=seq(0, 1, 0.1))+geom_histogram(aes(x=slide_sim$Proportion), breaks=seq(0, 1, 0.1), fill='red', binwidth = 0.1, alpha=0.5)+xlim(0,1)
-hist_overlay_c<-slide_count %>% ggplot(aes(x=Yes))+geom_histogram(fill='blue', alpha=0.5, breaks=c(0:12))+geom_histogram(aes(x=slide_sim$Yes), fill='red',  alpha=0.5, breaks=c(0:12))+scale_x_continuous(breaks=c(0:12))
-
-plot_grid(plot_grid(hist_data, hist_sim, nrow=2), hist_overlay, nrow=1)
-plot_grid(plot_grid(hist_data_c, hist_sim_c, nrow=2), hist_overlay_c, nrow=1)
+pheno<-readRDS("/scratch/bfoxman_root/bfoxman/blostein/FragileFamilies/Files/OGData/fullpheno.rds")
 
 ```
-
-## Comparison of analytic sample - methylation data to Fragile Families sample
-
-```{r, warning=F, message=F}
-FF_labeled<-set_label(FF_labeled, varLabels)
-methyl<-createTable(compareGroups(MethylData~smkPreg_binary+cm1inpov+cm1ethrace+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+k5f1l+k6d40+f1g4, data=FF_labeled), show.all=T)
-export2md(methyl, caption="Exclusion table of all Fragile Families vs those with any methylation data for main predictors and key demographic variables from first wave")
-```
-
-## Comparison of variables by ancestry 
-```{r, warning=F, message=F, include=F}
-myFF<-set_label(myFF, varLabels4)
-ancestrytable<-strataTable(createTable(compareGroups(ancestry~globalmethylation+cg05575921+pediatric+levine+Epi+Fib+IC+smkPreg_binary+cm1inpov+cm1ethrace+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+k6d40+k6d42+f1g4, data=myFF), show.all=F), 'childteen')
-export2md(ancestrytable)
-```
-
-## Flow chart of sample exclusions
-
-```{r MakeFlowChart}
-methylCohort<-pdqc_all %>% filter(childteen!='M')
-
-basemodelvar<-c("smkPreg_binary", "ancestry", "cm1bsex", "cm1inpov", "Epi", "IC", "ChildAgeComposite")
-basemodeldata<-myFF %>% filter_at(all_of(basemodelvar), all_vars(!(. %in%c("Missing", "Missing PC data"))& !is.na(.)))
-child_base<-basemodeldata%>%filter(childteen=='C')
-teen_base<-basemodeldata%>%filter(childteen=='T')
-
-prenatalexposurevar<-c(basemodelvar, "m1g2_YesNoPreg", "m1g3_YesNoPreg")
-prenatalexdata<-basemodeldata %>%filter_at(all_of(prenatalexposurevar), all_vars(!(. %in% c("Missing"))))
-child_prenatal<-prenatalexdata%>%filter(childteen=='C')
-teen_prenatal<-prenatalexdata%>%filter(childteen=='T')
-
-secondhandsmokevars<-c(prenatalexposurevar, "PostnatalMaternalSmokingAny", "SmkAtVisitPastmonth")
-secondhandsmkdata<-prenatalexdata%>%filter_at(all_of(secondhandsmokevars), all_vars(!(. %in% c("Missing"))))
-child_secondhand<-secondhandsmkdata %>% filter(childteen=='C')
-teen_secondhand<-secondhandsmkdata%>%filter(childteen=='T')
-twovisit_secondhand<-secondhandsmkdata %>% filter(idnum %in% child_secondhand$idnum & idnum %in% teen_secondhand$idnum)
-
-child_smoke<-secondhandsmkdata %>% filter(k5f1l!= "2 no" & childteen=='C')
-child_nosmoke<-secondhandsmkdata %>% filter(k5f1l=="2 no" & childteen=='C')
-teen_smoke<-secondhandsmkdata %>% filter(k6d40!="2 No" & childteen=='T')
-teen_nosmoke<-secondhandsmkdata %>% filter(k6d40=="2 No" & childteen=='T')  
-twovisit_nosmoke<-secondhandsmkdata %>% filter(idnum %in% child_nosmoke$idnum & idnum %in% teen_nosmoke$idnum)
-
-fathersmkdata<-secondhandsmkdata %>% filter(f1g4!="Missing")
-
-#Make Labels
-n_fullFF<-paste0("Fragile Families cohort \n", "n=", FF_labeled %>% nrow())
-n_methylQC<-paste0("Cohort with 450K methylation data \n", "n=", methylCohort$idnum%>%unique()%>%length(), "individuals with m=", methylCohort$MethID%>%length(), "samples")
-
-n_QCloss<-paste0("Methylation data QC: 12 low-intensity samples dropped, \n", pdqc_all%>%filter(probe_fail_10==T)%>%nrow()-12, " samples with >10% of probes failing dropped, \n", pdqc_all%>%filter(sex!=predicted_sex & probe_fail_10==F & childteen!='M')%>%nrow(), " discordant sex samples dropped, \n & ", nrow(pdqc)-nrow(pdqc_clean), " duplicate samples dropped")
-n_methylFF<-paste0("Cohort with quality controlled 450K data \n","n=", myFF %>% select(id) %>% unique() %>% nrow(), " individuals with m=", myFF%>%nrow(), " samples")
-
-n_basemodel<-paste0("Base models: \n", "n=", basemodeldata %>% select(id) %>% unique() %>% nrow(), " individuals with m=", basemodeldata %>%nrow(), " samples")
-n_baseexclusions<-paste0("n=",  myFF %>% filter_at(all_of(basemodelvar), any_vars((.
-%in%c("Missing", "Missing PC data"))))%>% select(id) %>% unique() %>% nrow(), " individuals & m=", myFF %>% filter_at(all_of(basemodelvar), any_vars((.
-%in%c("Missing", "Missing PC data")))) %>% nrow(), " samples missing principal components of ancestry; \n n=", myFF %>% filter_at(all_of(basemodelvar), any_vars((is.na(.))))%>%select(id)%>%unique()%>%nrow(), " individuals & m=", myFF %>% filter_at(all_of(basemodelvar), any_vars((is.na(.))))%>%nrow(), " samples missing child age at maternal/PCG interview")
-
-n_prenatalexposures<-paste0("Prenatal exposures models: \n","n=", prenatalexdata %>% select(id) %>% unique() %>% nrow(), " individuals with m=", prenatalexdata %>%nrow(), " samples")
-n_prenatalexclusions<-paste0("n=", basemodeldata %>% filter_at(all_of(prenatalexposurevar), any_vars(.
-%in% c("Missing"))) %>% select(id) %>% unique() %>% nrow(), " individuals & m=", basemodeldata %>% filter_at(all_of(prenatalexposurevar), any_vars(.
-%in% c("Missing"))) %>% nrow(), " samples missing \n data on maternal prenatal alcohol and drug use")
-
-n_secondhandSmk<-paste0("Secondhand smoke exposure models: \n","n=", secondhandsmkdata %>% select(id) %>% unique() %>% nrow(), " individuals with m=", secondhandsmkdata %>%nrow(), " samples")
-n_secondhandexclusions<-paste0("n=", prenatalexdata$id[!(prenatalexdata$id %in% secondhandsmkdata$id)] %>% unique()%>% length(), " individuals & m=", prenatalexdata %>% filter_at(all_of(secondhandsmokevars), any_vars(.
-%in% c("Missing"))) %>% nrow(), " samples missing data \n on maternal or primary care giver postnatal smoking")
-
-n_Child<-paste0('n=', child_secondhand  %>% nrow(), " individuals with a 9 year visit")
-n_ChildSmkexcluded<-paste0("Sensitivity analysis exclude children who smoke: \n",'n=', child_nosmoke %>% nrow(), " individuals with a 9 year visit")
-n_ChildSmkexclusions<-paste0('n=', child_smoke%>%nrow(), " individual \n smoking at 9 year visit \n or missing focal child smoking data")
-n_Teen<-paste0('n=', teen_secondhand  %>% nrow(), " individuals with a 15 year visit")
-n_TeenSmkexcluded<-paste0("Sensitivity analysis exclude children who smoke: \n",'n=', teen_nosmoke %>% nrow(), " individuals with a 15 year visit")
-n_TeenSmkexclusions<-paste0('n=', teen_smoke%>%nrow(), " individuals \n smoking at 15 year visit \n or missing focal child smoking data")
-
-n_sensFatherSmoking<-paste0("Sensitivity analysis: compare effects of maternal and paternal prenatal smoking \n", "n=", fathersmkdata %>% select(id) %>% unique() %>% nrow(), " individuals with m=", fathersmkdata %>%nrow(), " samples", "\n with prenatal paternal smoking data")
-n_sensFatherSmokingEx<- paste0("n=", secondhandsmkdata$id[!(secondhandsmkdata$id %in% fathersmkdata$id)] %>% unique()%>% length(), " individuals & m=", secondhandsmkdata %>% filter(f1g4 %in% c("Missing")) %>% nrow(), " samples missing data on paternal prenatal smoking")
-
-n_twovisits<-paste0("n=", twovisit_secondhand%>%pull(idnum) %>%unique() %>%length(), " individuals with both a 9 and 15 year visit")
-n_twovisitsSmkExcluded<-paste0("Sensitivity analysis exclude children who smoke: \n", "n=", twovisit_nosmoke%>%pull(idnum) %>%unique() %>%length(), " individuals with both a 9 and 15 year visit")
-n_twovisitSmkExclusions<-paste0("n=", (twovisit_secondhand%>%pull(idnum) %>%unique() %>%length())-(twovisit_nosmoke%>%pull(idnum) %>%unique() %>%length()), " individuals smoking at \n either 9 and 15 year visit or \n missing focal child smoking data")
-
-```
-
-```{r drawFlowchart}
-graph1<-DiagrammeR::grViz("
-   digraph graph2 {
-   compound=true;
-   graph [layout = dot]
-
-   node [shape = rectangle, width=4]
-   a [label = '@@1']
-   b [label = '@@2']
-   c [label = '@@3']
-   d [label = '@@4']
-   e [label = '@@5']
-   f [label = '@@6']
-   g [label = '@@7']
-   h [label = '@@8']
-   i [label = '@@9', group=g1]
-   j [label = '@@10']
-   k [label = '@@16']
-   l [label = '@@17']
-   m [label = '@@19']
-
-   #fake nodes for connecting to edges
-   node [shape=point, width=0.01, height=0.01]
-   a1
-   b1
-   c1
-   d1
-#   e1
-  f1 [group=g1]
-  g1
-  k1
-
-  #exclusion nodes
-  node [shape = rectangle, width=4]
-  aa [label = '@@20']
-  bb [label = '@@11']
-  cc [label = '@@12']
-  dd [label = '@@13']
-  ff [label = '@@14', width=3]
-  gg [label = '@@15', width=3]
-  kk [label = '@@18', width=3]
-
-  #draw connections
-  a->m
-  m->a1 [arrowhead=none]
-  a1->b
-  b->b1 [arrowhead=none]
-  b1->c
-  c->c1 [arrowhead=none]
-  c1->d
-  d->d1 [arrowhead=none]
-  d1->e
-  e->f; e->g; e->k
-{rank=same; h->e [dir=back, minlen=2, style=dashed]}
-  f->f1 [arrowhead=none, style=dashed]
-  g->g1 [arrowhead=none, style=dashed]
-  k->k1 [arrowhead=none, style=dashed]
-  f1->i [style=dashed]
-  g1->j [style=dashed]
-  k1->l [style=dashed]
-
-{rank=same; a1->aa [minlen=2]}   
-{rank=same; b1->bb [minlen=2]} 
-{rank=same; c1->cc [minlen=2]}
-{rank=same; d1->dd [minlen=2]}
-{rank=same; f1->ff [minlen=2]}
-{rank=same; g1->gg [minlen=2]}
-{rank=same; k1->kk [minlen=2]}
-
-#invisible constraints to force ordering
-ff->i [style=invis]
-   }
-
-   [1]: n_fullFF
-   [2]: n_methylFF
-   [3]: n_basemodel
-   [4]: n_prenatalexposures
-   [5]: n_secondhandSmk
-   [6]: n_Child
-   [7]: n_Teen
-   [8]: n_sensFatherSmoking
-   [9]: n_ChildSmkexcluded
-   [10]: n_TeenSmkexcluded
-   [11]: n_baseexclusions
-   [12]: n_prenatalexclusions
-   [13]: n_secondhandexclusions
-   [14]: n_ChildSmkexclusions
-   [15]: n_TeenSmkexclusions
-   [16]: n_twovisits
-   [17]: n_twovisitsSmkExcluded
-   [18]: n_twovisitSmkExclusions
-   [19]: n_methylQC
-   [20]: n_QCloss
-")
-```
-
-```{r drawGraph}
-graph1
-```
-
-## Cell type proportions over time
-
-```{r, eval=F, include=F}
-cellbytime<-myFF %>% pivot_longer(cols=c(Fib, Epi, IC), names_to="Celltype", values_to = "Proportion")
-ggplot(cellbytime, aes(x=childteen, y=Proportion, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+facet_wrap(~Celltype)+stat_compare_means(label="..p.signif..")+ggtitle("Cell type proportions by visit")
-
-ggplot(myFF, aes(x=childteen, y=globalmethylation, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+ggtitle("Global methylation by visit")
-
-ggplot(myFF, aes(x=childteen, y=pediatric, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+ggtitle("Pediatric methylation clock by visit")
-ggplot(myFF, aes(x=pediatric, y=IC))+geom_point()+stat_cor()+facet_wrap(~childteen)+ggtitle("Correlation between pediatric clock and IC proportion")
-#ggplot(myFF, aes(x=pediatric, y=Epi))+geom_point()+stat_cor()+facet_wrap(~childteen)
-#ggplot(myFF, aes(x=childteen, y=levine, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")
-ggplot(myFF, aes(x=childteen, y=cg05575921, color=childteen))+geom_violin()+geom_boxplot(width=0.2)+stat_compare_means(label="..p.signif..")+ggtitle("Methylation at cg05575921 by visit")
-
-```
-Cell type proportions do differ between 9 and 15 year visits and although global methylation does not look different between these visits, methylation clocks do (a good check), although, interestingly, cell type proportion does correlate with methylation age.
-
-## Distributions & checks of methylation variables
-
-### Correlation between all methylation data
-```{r}
-methylation_data<-myFF %>% select(colnames(methyldata))%>%select(-MethID, -childteen, -idnum, -slide, -array)
-chart.Correlation(methylation_data, histogram=T)
-```
-
-### Correlation between polymethylation scores
-```{r}
-chart.Correlation(methylation_data %>%select(SSolder_notransform, SSnewbornCT_notransform, SSnewborn_notransform, anynewborn_notransform))
-```
-
-Thought this was a little weird, investigated the actual coefficients for each score:
-
-```{r}
-coefs_methyl<-cbind('SSolderCoefs'=cpgs$SSolder$Coef, 'SSnewbornCTCoefs'=cpgs$SSnewbornCT$Coef, 'SSnewbornCoefs'=cpgs$SSnewborn$Coef, 'anynewbornCoefs'=cpgs$anynewborn$Coef)
-chart.Correlation(coefs_methyl)
-```
-
-```{r correlaton b/t polymethylation scores over time, eval=F}
-PMS<-methyldata %>%select(idnum, childteen, colnames(polyscores_wide), -MethID)%>%pivot_longer(cols=any_of(colnames(polyscores_wide)), names_to="PMS_type", values_to="PolymethylationScore")%>%pivot_wider(names_from = childteen, values_from=PolymethylationScore)
-ggplot(PMS, aes(x=C, y=T))+geom_point()+facet_grid(~PMS_type)+stat_cor()+geom_smooth(method=lm)
-```
-```{r Polymethylation scores over time}
-gg_newborn<-ggplot(methyldata, aes(x=childteen, y=SSnewbornCT_center, group=idnum))+geom_point()+geom_line(alpha=0.2)+geom_smooth(aes(group=1), method='lm')+ggtitle("SS newborn (+CT)")
-gg_older<-ggplot(methyldata, aes(x=childteen, y=SSolder_center, group=idnum))+geom_point()+geom_line(alpha=0.2)+geom_smooth(aes(group=1), method='lm')+ggtitle("SS older")
-plot_grid(gg_newborn, gg_older, nrow = 2)
-```
-```{r Polymethylation scores over time by prenatal smoke}
-gg_newborn<-ggplot(myFF, aes(x=childteen, y=SSnewbornCT_center, group=idnum, color=smkPreg_binary))+geom_point()+geom_line(alpha=0.2)+geom_smooth(aes(group=1),color='black', method='lm')+facet_wrap(~smkPreg_binary)
-gg_older<-ggplot(myFF, aes(x=childteen, y=SSolder_center, group=idnum, color=smkPreg_binary))+geom_point()+geom_line(alpha=0.2)+geom_smooth(aes(group=1),color='black', method='lm')+facet_wrap(~smkPreg_binary)
-plot_grid(gg_newborn, gg_older, nrow=2)
-```
-
-```{r boxplots of PMS}
-PMS<-left_join(methyldata %>%select(idnum, childteen, colnames(polyscores_wide), -MethID)%>%pivot_longer(cols=any_of(colnames(polyscores_wide)), names_to="PMS_type", values_to="PolymethylationScore"), myFF[, c('idnum', 'smkPreg_binary')]%>%unique())
-ggplot(PMS, aes(x=childteen, y=PolymethylationScore, color=smkPreg_binary))+geom_boxplot()+facet_wrap(~PMS_type)+stat_compare_means(label='p.signif', label.y=4)
-
-```
-
-## Distribution of main predictor and bivariate tables (stratified by ancestry)
-
-### Covariates by main predictor (unstratified)
-
-```{r, pickData}
-tabledata<-rbind(child_nosmoke, teen_nosmoke)
-tabledata<-set_label(tabledata, varLabels4)
-
-```
-
-```{r, warning=F, message=F}
-smk_all<-strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+SSnewbornCT_notransform+SSolder_notransform+cg05575921+pediatric+levine+Epi+Fib+IC+cm1inpov+cm1ethrace+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+k5f1l+k6d40+f1g4, data=tabledata, simplify = F), show.all=F), 'childteen')
-export2md(smk_all)
-```
-
-Note hypomethylation in exposed children at cg05575291 (AHHR) gene, which is more strongly associated at the Teen visit than the Child visit....
-although not a very small p-value - check with Kelly - is this concerning?
-
-Additionally, cell type proportion does not appear to associate with main predictor of binary prenatal smoking variable.
-
-### Covariates by main predictor (stratified by ancestry)
-These tables use the base model group from the above CONSORT diagram (n=844). 
-```{r, warning=F, message=F}
-smkA<-strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+SSnewbornCT_notransform+SSolder_notransform+cg05575921+pediatric+levine+Epi+Fib+IC+cm1inpov+cm1ethrace+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+k5f1l+k6d40+f1g4, data=tabledata%>%filter(ancestry=='African ancestry')%>%copy_labels(tabledata)), show.all=F), 'childteen')
-export2md(smkA, caption="Smoked during pregnancy vs covariates: African ancestry only")
-smkE<-strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+SSnewbornCT_notransform+SSolder_notransform+cg05575921+pediatric+levine+Epi+Fib+IC+cm1inpov+cm1ethrace+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+k5f1l+k6d40+f1g4, simplify=FALSE, data=tabledata%>%filter(ancestry=='European ancestry' & SmkAtVisitPastmonth!="Missing")%>%copy_labels(tabledata)), show.all=TRUE), 'childteen')
-export2md(smkE, caption="Smoked during pregnancy vs covariates: European ancestry only")
-smkH<-strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+SSnewbornCT_notransform+SSolder_notransform+cg05575921+pediatric+levine+Epi+Fib+IC+cm1inpov+cm1ethrace+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth+k5f1l+k6d40+f1g4, data=tabledata%>%filter(ancestry=='Hispanic ancestry')%>%copy_labels(tabledata)), show.all=F), 'childteen')
-export2md(smkH, caption="Smoked during pregnancy vs covariates: Hispanic ancestry only")
-
-```
-
-When stratified by ancestry, relationship between cg05575291 methylation and smoking remains in African ancestry group, pattern of hypomethylation exists in teen measures of European ancestry group but not child measures, and actually non-significant hypermethylation in Hispanic ancestry group.
-
-## Correlation plots between all continuous variables
-
-```{r, eval=F, include=F, eval=F}
-cont_data<-myFF %>% select(globalmethylation, cg05575921, pediatric, levine, Epi, Fib, IC, myContinuous)%>%mutate_at(vars(myContinuous), funs(as.numeric(ifelse(.<0, NA, .))))
-chart.Correlation(cont_data, histogram=T)
-```
-
-## Associations with global methylation 
-```{r, eval=F, include=F}
-#use desc tools here 
-myFFbin<-myFF%>%mutate_at(vars(all_of(myBinary)), funs(recode_factor(., '-1 Refuse'='Missing', "-2 Don't know"="Missing", "-9 Not in wave"="Missing", "-3 Missing"="Missing", "-6 Skip"="Missing", "-7 N/A"="Missing", "-5 Not asked"="Missing")))
-paste("globalmethylation~", paste(myBinary, collapse="+"), sep='')
-z<-Desc(globalmethylation~m1g5+m1g6+m1h3a+m1i4a+f1h3a+m2j5+m2j7+f2j5+f2j7+f3j31+m4j18+f4j18+k5f1k+k5f1l+m5g17+f5g17+n5f17+p5h15+p6h74+k6d40+k6d45, data=myFFbin%>%filter(childteen=='C'), plotit = F)
-x<-Desc(globalmethylation~m1g5+m1g6+m1h3a+m1i4a+f1h3a+m2j5+m2j7+f2j5+f2j7+f3j31+m4j18+f4j18+k5f1k+k5f1l+m5g17+f5g17+n5f17+p5h15+p6h74+k6d40+k6d45, data=myFFbin%>%filter(childteen=='T'), plotit = F)
-
-plot(Desc(globalmethylation~smkPreg_binary+m2j5+n5f17, data=myFFbin %>%filter(childteen=='C'), main=c("Binary pregnancy smoke variable", "Mother smoke cig - Year 1", "PCG smoke cig - Year 9")), type='dens')
-
-plot(Desc(globalmethylation~smkPreg_binary+m2j5+m5g17, data=myFFbin %>%filter(childteen=='T'), main=c("Binary pregnancy smoke variable (teen methyl)", "Mother smoke cig - Year 1 (teen methyl)", "Mother smoke cig - Year 9 (teen methyl)")), type='dens')
-
-myFFcat<-myFF%>%mutate_at(vars(all_of(c(myCategorical_O, myCategorical_U))), funs(recode_factor(., '-1 Refuse'='Missing', "-2 Don't know"="Missing", "-9 Not in wave"="Missing", "-3 Missing"="Missing", "-6 Skip"="Missing", "-7 N/A"="Missing", "-5 Not asked"="Missing")))
-paste("globalmethylation~", paste(c(myCategorical_O, myCategorical_U), collapse="+"), sep='')
-cat_oC<-Desc(globalmethylation~m1g2+m1g3+m1g4+f1g4+m2j5a+m2j7a+f2j5a+f2j7a+f3j32+m4j19+f4j19+m5g18+f5g18+n5f18+p5q3cr+p6h76+p6h77+k6d42+k6d43+k6d46+k6d47+cm1bsex+m1h3+m1h3b+cm1ethrace+m1i4+m1i4b+f1h3+f1h3b+cf1ethrace, data=myFFcat%>%filter(childteen=='C'), plotit = F)
-#n5f18 p6h76 cf1ethrace m1h3 (ethnicity) cm1bsex
-cat_oT<-Desc(globalmethylation~m1g2+m1g3+m1g4+f1g4+m2j5a+m2j7a+f2j5a+f2j7a+f3j32+m4j19+f4j19+m5g18+f5g18+n5f18+p5q3cr+p6h76+p6h77+k6d42+k6d43+k6d46+k6d47+cm1bsex+m1h3+m1h3b+cm1ethrace+m1i4+m1i4b+f1h3+f1h3b+cf1ethrace, data=myFFcat%>%filter(childteen=='T'), plotit = F)
-#m2j5a
-
-plot(Desc(globalmethylation~cm1bsex, data=myFFbin %>%filter(childteen=='T'), main=c("Baby sex")), type='dens')
-
-plot(Desc(globalmethylation~p6h74, data=myFF %>%filter(childteen=='T'), main=c("PCG ever smoked cigarretes")), type='dens')
-plot(Desc(globalmethylation~p6h76, data=myFF %>%filter(childteen=='T'), main=c("How often PCG smoked cigarretes past month")))
-plot(Desc(globalmethylation~p6h76, data=myFFcat %>%filter(childteen=='T'), main=c("How often PCG smoked cigarretes past month")), type='dens')
-
-```
-
-At child and teen visit, variables to do with Father and Mother hispanic identity and ethnicity associated with global methylation, as did baby sex.
-
-Some maternal/primary care giver smoking variables from year 1, 9 and 15 also associated.
-A note on the question PCG Year 9 Smoke and related questions n5f17 and n5f18 - these variables are weird, with lots of missing values, need to check the FF question guide, but my impression is that they were only asked to children for whom the PCG was not the mom (PCG and Core mother survey were combined in wave 6 - year 15).
-
-## Associations with cg05575921 methylation and other significant sites from Wiklund et al 2019
-
-Base model controls for ancestry, childsex, maternal income to poverty ratio, Epi cells, IC cells, and child age in months at sample. 
-```{r}
-#define function for use in purr 
-apriori_fun = function(response, dat, groupF, vars){
-  exclude=ifelse(groupF=='C', 'cp6yagem', ifelse(groupF=='T', 'cm5b_age', NA))
-  vars=vars[vars!=exclude]
-  form=paste0(response, "~", paste(vars, collapse='+'))
-  tidy(lm(as.formula(form), data=dat %>% filter(childteen==groupF)), conf.int=T, conf.level=0.95)
-}
-
-#read in top CpGs from Table One of Wicklund et al 2019
-topCPGs<-read.csv(paste0(datadir, "TopCpGsWiklundEtAl2019.csv"))
-topCPGs<-topCPGs %>% filter(CpG %in% aprioriCG)%>% mutate(estimate=as.numeric(gsub("−\\s", "-", gsub(" .*", "", β..SE.))), std.error=as.numeric(sub("\\).*", "", sub(".*\\(", "", β..SE.))), p.value=P.value, source="Wiklund et al.
-2019", conf.low=(estimate-1.96*std.error), conf.high=(estimate+1.96*std.error))%>%select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source)
-
-#purr over the aprioriCpGs
-vars=aprioriCG
-vars=purrr::set_names(vars)
-#ages=c('C', 'T')
-#ages=purrr::set_names(ages)
-models_C=  vars%>% purrr::map(apriori_fun, dat=basemodeldata, groupF='C', vars=basemodelvar)
-models_T=  vars%>% purrr::map(apriori_fun, dat=basemodeldata, groupF='T', vars=basemodelvar)
-models=list('child'=models_C, 'teens'=models_T)
-#pull out list of tidy models into a datafram 
-modeldf<-left_join(bind_rows(unlist(models, recursive=F), .id="id") %>% filter(term=='smkPreg_binaryYes')%>%mutate(source=paste0("Fragile Families: ", age=gsub("\\..*", "", id)), CpG=gsub(".*\\.", "", id)), topCPGs %>%select(CpG, Chr, Position, Gene))%>%select(CpG, Chr, Position, Gene, estimate, std.error, p.value, conf.low, conf.high, source)
-
-#combine
-modeldf<-rbind(modeldf, topCPGs)%>%mutate(label=paste0(Gene, ": ", CpG))
-
-#plot
-ggplot(modeldf, aes(x=estimate, xmin=conf.low, xmax=conf.high, color=source, y=label))+geom_point(position=position_dodge(width=1))+geom_errorbarh(height=0.1, position=position_dodge(width=1))+scale_x_continuous()+scale_y_discrete(name="CpGs (a priori)")+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+ggtitle("Wiklund 2019 vs Fragile Families at 4 selected a priori CpGs: \n Base model")
-
-```
-
-# Datasets for modeling
-Variables
-
-Base model: Binary smoking variable [X], ancestry categorization, baby sex [X], maternal income to poverty ratio [X],  child's age in month at visit of interest (note, currently using a constructed child's age at visit, where individuals having a missing child's age at home visit was supplemented with child's age at home interview) [X], cell types (Epi + IC, may need to take one out),
-
-Prenatal exposures: Base model variables + prenatal maternal alcohol use (Yes/no) [X], prenatal maternal drug use (yes/no)
-
-Secondhand smoke exposure models: Prenatal exposures model+ any maternal smoking at age 1 & 5 [X], maternal or PCG smoking dose at visit of interest (age 9 or 15) [X]
-
-First hand smoking: Secondhand smoke exposure models + exclude children with missing or reported firsthand smoking at visit of interest (age 9 or 15)
-
-Paternal smoking: Filter to individuals with data on prenatal paternal smoking and compare effect size of prenatal paternal to prenatal maternal (as per Wiklund et al.
-2019, idea is that this captures additional confounding)
-
-Ancestry specific models - by ancestry groups; replace ancestry control with control for local PCs within ancestry group
-
-Longitduinal models - no additional filtering needed
-
-```{r, saveData}
-save(child_base, file=paste0(datadir, 'CreatedData/', 'ChildBaseModelData.Rdata'))
-save(teen_base, file=paste0(datadir, 'CreatedData/', 'ChildBaseModelData.Rdata'))
-save(basemodeldata, file=paste0(datadir, 'CreatedData/', 'BaseModelData.Rdata'))
-
-save(child_prenatal, file=paste0(datadir, 'CreatedData/', 'ChildPrenatalExModelData.Rdata'))
-save(teen_prenatal, file=paste0(datadir, 'CreatedData/', 'PrenatalExModelData.Rdata'))
-save(prenatalexdata, file=paste0(datadir, 'CreatedData/', 'PrenatalExModelData.Rdata'))
-
-save(child_secondhand, file=paste0(datadir, 'CreatedData/', 'ChildSecondHandSmkModelData.Rdata'))
-save(teen_secondhand, file=paste0(datadir, 'CreatedData/', 'TeenSecondHandSmkModelData.Rdata'))
-save(twovisit_secondhand, file=paste0(datadir, 'CreatedData/', 'TwoVisitSecondHandSmkModelData.Rdata'))
-save(secondhandsmkdata, file=paste0(datadir, 'CreatedData/', 'SecondHandSmkModelData.Rdata'))
-
-save(child_nosmoke, file=paste0(datadir, 'CreatedData/', 'ChildNoSmkModelData.Rdata'))
-save(teen_nosmoke, file=paste0(datadir, 'CreatedData/', 'TeenNoSmkModelData.Rdata'))
-save(twovisit_nosmoke, file=paste0(datadir, 'CreatedData/', 'TwoVisitNoSmkModelData.Rdata'))
-
-save(fathersmkdata, file=paste0(datadir, 'CreatedData/', 'FatherSmkModelData.Rdata'))
-
-save(list=c("basemodeldata", "prenatalexdata", "secondhandsmkdata", "child_secondhand", "teen_secondhand", "twovisit_secondhand", "twovisit_nosmoke", "child_nosmoke", "teen_nosmoke", "fathersmkdata"), file=paste0(datadir, 'CreatedData/', Sys.Date(), 'CreatedData.Rdata'))
-```

### 2021-11-22 Reports/Paper1.Rmd
small changes to supplemental figures from john and jonah

@@
 
-```{r sources scripts instead of loading, eval=run_scripts}
+```{r sources scripts instead of loading, eval=F}
 source(file.path(scriptdir, '01_clean_phenotype_data.R'))
@@
     facet_grid(Methyl_factor~childteen, scales='free',labeller = labeller(Methyl_factor=all_methyl_labels))+
-    scale_x_discrete(labels=function(x) str_wrap(x, width = 10))+
-    theme(legend.position = 'right')+
+    scale_x_discrete(labels=function(x) str_wrap(x, width = 20))+
+    theme(legend.position = 'none')+
+  theme(axis.text.x = element_text(angle=30, hjust=1))+
   guides(color=guide_legend(nrow=4,byrow=TRUE))
@@
 bring.the.heatmap.sv <- function(pd,covar,folder){
-
+  library('gtools')
   nsv <- length(grep('sv',names(pd)))
@@
     geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
-    facet_wrap(predictors~., scales='free_x')+
+    facet_wrap(predictors~., scales='free_x', labeller=labeller(predictors=label_wrap_gen(20)))+
     geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
@@
   mutate(modeltype=gsub("Sustained smoking polymethylation score \\(newborns, cell-type controlled\\)", "Polymethylation score for prenatal maternal smoke exposure", modeltype))%>%
+  mutate(modeltype=factor(modeltype, levels=c('AHRR: cg05575921', 'Base model+AHRR: cg05575921', 'Global methylation', 'Base model+Global methylation',  'Polymethylation score for prenatal maternal smoke exposure', 'Base model+Polymethylation score for prenatal maternal smoke exposure')))%>%
     ggplot(aes(x=spec, y=sens, color=modeltype))+geom_line()+
@@
     theme_classic()+
-    scale_color_manual('', values=c('red', 'darkred', 'darkblue', 'darkgreen', 'blue', 'green'), 
+    scale_color_manual('', values=c('red', 'darkred', 'green', 'darkgreen', 'steelblue1', 'royalblue3'), 
                        function(x) str_wrap(x, width = 30))+

### 2021-11-22 Reports/Paper1.Rmd
edit sig labels on sup fig 2

@@
 visit_cor=age_wide_methyl%>%
-  ggplot(aes(x=`Age 9`, y=`Age 15`))+geom_point()+stat_cor()+geom_smooth(method=lm)+theme_classic()+
+  ggplot(aes(x=`Age 9`, y=`Age 15`))+geom_point(size=1)+stat_cor(label.sep=', ', size=2.5, label.y.npc=0.95)+geom_smooth(method=lm)+theme_classic()+
   scale_x_continuous('Age 9 score')+scale_y_continuous('Age 15 score')+

### 2021-11-24 Reports/Paper1.Rmd
post jonah new dataset with mislabeled samples excluded or relabeled

@@
                            ' with abnormal sex chromosome intensity dropped & \n ', pdqc_all),
+                 N_M_count("cut %in% c('contaminated', 'relative') & predicted_sex_outlier==F & sex_new==predicted_sex & probe_fail_10==F & childteen!='M'", 
+          ' with contamination/mislabels & \n ', pdqc_all),
                  length(reps_remove), " duplicate samples dropped")

### 2021-11-24 Scripts/01_clean_phenotype_data.R

@@ -68,11 +68,42 @@ pdqc_all = pdqc_all %>% left_join(airpollution_study %>%
                                     dplyr::select(idnum, MethID, childteen, sex_new))%>%
    mutate(sex_flag=ifelse(sex!=sex_new, 'recode_sex', 'correct_sex'))
 sex_recodes=pdqc_all %>% filter(sex_flag=='recode_sex')%>%pull(idnum)
+
+#############################mislabeled samples data###############################
+#incorporate changes to Fragile Families QC from Jonah on 11/22/2020
+#Additional samples identified as problematic, only incorporate those with cut=='pass';
+#except the 9 sex recodes from above
+#additionally, 3 passing samples were mislabled and have to be relabeled
+cut_data=read.csv(file.path(datadir, 'OGData', '02-thirdpass.csv'))
+mislabel=cut_data %>%
+  mutate(idnum=as.character(idnum), 
+         new_id=idnum, 
+         old_id=case_when(str_detect(changenote, 'idnum')~gsub('idnumchange_from_', '', changenote), 
+                          !str_detect(changenote, 'idnum')~idnum))%>%
+  select(MethID, old_id, new_id, changenote, cut)
+
+only_mislabels=mislabel%>%filter(str_detect(changenote, 'idnum'))%>%select(MethID, old_id, new_id, cut, changenote)
+
+cut_data = cut_data %>%
+  mutate(cut=ifelse(changenote=='intractable_siblingpair', 'cut', cut))%>%
+  mutate(idnum=as.character(idnum))%>%
+  select(idnum, MethID, cut, changenote)
+
+#replace ids for the incorrect samples 
+pdqc_all=pdqc_all%>%
+  left_join(mislabel)%>%
+  mutate(idnum=ifelse(idnum!=new_id, new_id, idnum))%>%
+  left_join(cut_data)%>%
+  mutate(childteen=ifelse(cut=='mother', 'M', childteen), #add new mom problems
+         cut=ifelse(sex_flag=='recode_sex', 'pass', cut)) #relabel sex flagged recodes as pass  
+  
+
 #check
 #with(pdqc_all, table(sex_new, cm1bsex))
 #filter samples w/ >10 probe fail or sex !=predicted sex or sex outlier or mother sample 
 pdqc<-pdqc_all %>% 
[…]

### 2021-11-24 Scripts/02_make_methyl_summary.R

@@ -96,7 +96,7 @@ if(nrow(betaqc)==423668){
   colnames(clocks)[1]<-"MethID"
   grim_pk=read.csv(file=file.path(datadir, 'OGData', 'dnampackyrs_fromgrimoutput.csv'), header=T)
   colnames(grim_pk)[1]<-'MethID'
-  clocks<-left_join(clocks, grim_pk)
+  clocks<-left_join(clocks, grim_pk)%>%dplyr::select(-idnum) #get rid of idnum to avoid bad merging given relabeling of mislabels
   }
 #the below code will run if the # CpGs in your chosen CpG matrix isn't 423668 (see note at top re: Jonah email) as the clocks
 #from ffcw_n1776_8clocks.csv were run using the 423668 matrix. if you switch to some other beta matrix, this code will recreate 
@@ -119,7 +119,7 @@ names(pedc) <- ped[, ID]
 pd[, pediatric := agep(betaqc, coeff = pedc)]
 #pace of aging 
 pd[, c("poam38", "poam45") := PoAmProjector(betaqc)]
-clocks=pd
+clocks=pd%>%dplyr::select(-idnum)
 }
 ################################################join methyl data together 
 methyldata=left_join(cell_types, globalmethdf)%>%
@@ -140,7 +140,7 @@ methyl_labels_myclocks=c('Methylation data ID', 'Epithelial cell proportion', 'F
                          'AHHR: cg05575921', 'MYO1G: cg04180046', 'CYP1A1: cg05549655', 'GFI1: cg14179389', "MYO1G: cg22132788", 
                          pms_labels, gsub('no transform', 'z-score standardized', pms_labels), gsub('no transform', 'mean-centered', pms_labels), 
                          #paste0(pms_labels, '/n (coefficients) & z-score standardized (score)'), gsub('no transform', '/n z-score standardized (coefficients) & z-score standardized (score)', pms_labels), gsub('no transform', '/n mean-centered (coefficients) & z-score standardized (score)', pms_labels),
-                         'ID', 'Visit',
+                          'Visit',
                          'Horvath clock', 'SkinBlood clock', 'Pediatric clock', 'PoAm clock 38', 'PoAm clock 45')
 
 methyl_labels_jonahclocks=c('Methylation data ID', 'Epithelial cell proportion', 'Fibroblast cel proportion', 'Immune cell proportion',
@@ -149,7 +149,7 @@ methyl_labels_jonahclocks=c('Methylation data ID', 'Epithelial cell proportion',
[…]

### 2021-11-29 Papers/Polymethylation_scores/02_abstract.Rmd
comments from erin

@@
 **Background**: Prenatal maternal smoking has negative implications for child health.
-DNA methylation signatures  can function as a biomarker of prenatal maternal smoking.
+DNA methylation signatures can function as biomarkers of prenatal maternal smoking.
 However little work has assessed how DNA methylation signatures of prenatal maternal smoking vary across ages, ancestry groups, or tissues.
-In the Fragile Families and Child Wellbeing study, we tested whether prenatal maternal smoking was associated with salivary polymethylation scores.
-We assessed the consistency of associations at ages 9 and 15, their portability across  participants from African, European, and Hispanic genetic ancestries and the accuracy of exposure classification using Area Under the Curve (AUC) from receiver operating curve analyses.
+In the Fragile Families and Child Wellbeing study, we tested whether prenatal maternal smoking was associated with salivary polymethylation scores for prenatal maternal smoke exposure.
+We assessed the consistency of associations at ages 9 and 15, their portability across  participants from African, European, and Hispanic genetic ancestries and the accuracy of exposure classification using area under the curve (AUC) from receiver operating curve analyses.
 
@@
 In the full sample at age 9 (n=756), prenatal maternal smoke exposure was associated with a 0.52 (95%CI: 0.36, 0.67) standard deviation higher polymethylation score.
-The direction and magnitude of the association was consistent across when stratified genetic ancestries.
+The direction and magnitude of the association was consistent when stratified by genetic ancestries.
 In the full sample at age 15 (n=749), prenatal maternal smoke exposure was associated with a 0.46 (95%CI: 0.3, 0.62) standard deviation higher polymetylation score, and the association was attenuated among the European and Hispanic genetic ancestry samples.
-The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (AUC age 9=0.77, *P* value<0.001, age 15=0.77, *P* value<0.001), and more accurate than using a single *a priori* site in the AHRR gene (cg05575921 AUC=0.74, *P* value=0.03; age 15=0.73, *P* value=0.01).
+The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (AUC age 9=0.77, *P* value<0.001, age 15=0.77, *P* value<0.001).
+The polymethylation scores showed higher classification accuracy than using a single *a priori* site in the *AHRR* gene (cg05575921 AUC=0.74, *P* value=0.03; age 15=0.73, *P* value=0.01).
 

### 2021-11-29 Papers/Polymethylation_scores/04_background.Rmd
comments from erin

@@
 Maternal prenatal smoking is a public health concern.
+Children exposed to cigarette smoking in utero are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu].
 In 2018, 11% of pregnant women in the United States aged 15-44 reported any past-month cigarette use [@SAMHSA2018].
-Children exposed to cigarette smoking in utero are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu].
 Measuring prenatal maternal smoking behavior is challenging.
 Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua].
-The gold-standard for smoking measurements, serum cotinine levels, have a short half-life of nine hours in pregnant women [@Dempsey:2002wr].
+The gold-standard for smoking measurements, serum cotinine levels, have a half-life of nine hours in pregnant women, limiting the ability for accurate detection to a short window [@Dempsey:2002wr].
 Further, serum cotinine measures during pregnancy are rarely available outside of birth cohorts, limiting clinical application as a marker of prenatal exposures.
@@
 
-Maternal cigarette smoking during pregnancy is associated with reproducible DNA methylation signatures, measured in infant cord blood and placental tissues, and primarily tested in European cohorts [@Joubert:2016ue; @Everson:2021td; @Cardenas2019].
+In infant cord blood and placental tissues, maternal cigarette smoking during pregnancy is associated with reproducible DNA methylation signatures, though this has primarily been tested in European cohorts [@Joubert:2016ue; @Everson:2021td; @Cardenas2019].
 Prenatal maternal cigarette smoking exposure is associated with postnatal DNA methylation signatures in blood from cohorts of older children and adults [@Tehranifar:2018tb ; @Richmond:2018tu].
 Few studies have considered the persistence of the association between maternal prenatal cigarette smoking and DNA methylation at different ages in the same individuals.
-Additionally, little work has considered the portability of the association between maternal prenatal cigarette smoking and DNA methylation across tissues.
+Additionally, little work has considered the portability of the association between maternal prenatal cigarette smoking and DNA methylation outside of cord blood and placental tissue.
 DNA methylation drives cell differentiation and cell type proportions differ across tissue types.
@@
 Among 572 3-5 year old children, 186 of whom were of African or admixed genetic ancestry, prenatal maternal smoking was associated with DNA methylation in peripheral blood samples but ancestry-stratified analyses were not performed [@Ladd-Acosta:2016aa].
-Similarly, among 89 middle-aged women, of whom reported African American or Hispanic ethnicity, prenatal maternal cigarette smoking was associated with DNA methylation at 17 of 190 tested CpG sites [@Tehranifar:2018tb].
+Similarly, among 89 middle-aged women, of whom 28 reported African American or Hispanic ethnicity, prenatal maternal cigarette smoking was associated with DNA methylation at 17 of 190 tested CpG sites [@Tehranifar:2018tb].
 Adjusting for ancestry did not substantially affect the association between prenatal maternal smoking and DNA methylation at any of the sites, although the authors note that African American women had on average 2--5% higher mean DNA methylation on the absolute scale as compared with White and Hispanic women at 5 *CYP1A1* CpG sites [@Tehranifar:2018tb].
@@
 After association testing, an important next step is to evaluate DNA methylation as a biomarker of prenatal maternal smoke exposure by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior.
-Biomarkers can consist of single DNA methylation sites or summary meausres across multiple sites [@Philibert2020; @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa].
-One attractive option is a polymethylation score, analogous to a polygenic score.
+Biomarkers can consist of single DNA methylation sites or summary measures across multiple sites [@Philibert2020; @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa].
+One option is a polymethylation score, analogous to a polygenic score.
 In the polymethylation score, CpG sites are weighted by the strength of their association with maternal smoking from a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a].
@@
 Among middle-aged adults, a score calculated using DNA methylation in peripheral blood samples from a single time point could even predict prenatal smoke exposure (30 years previously) with an AUC of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu].
-Ideally, a biomarker for prenatal maternal smoking would be consistent over participant age at DNA methylation sampling and portable across genetic ancestry groups.
-However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across tissue, age, and ancestry has not been evaluated in the same study.
+Ideally, a biomarker for prenatal maternal smoking would be consistent over the life course and portable across genetic ancestry groups.
+However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across age, tissue, and genetic ancestry has not been evaluated in the same study.
 
 DNA methylation differences hold promise as a biomarker for prenatal maternal smoking.
-Before translation to clinical applications, we must understand how and if the signal varies across age, ancestry, and tissue sample.
-In the Fragile Families and Child Wellbeing study, a diverse longitudinal cohort of children, we aimed to assess the potential of salivary DNA methylation biomarkers for prenatal maternal cigarette smoking.
-We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including epigenetic clocks, polymethylation scores, and individual *a priori* CpG sites.
+Before translation to clinical applications, we must understand how and if the signal varies across age, tissue sample, and genetic ancestry.
+In the Fragile Families and Child Wellbeing study, a diverse longitudinal birth cohort of children, we aimed to assess the potential of salivary DNA methylation biomarkers for prenatal maternal cigarette smoking.
+We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including polymethylation scores for smoke exposure, individual *a priori* CpG sites, global DNA methylation, and epigenetic clocks.
 We then performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a persistent and consistent biomarker of prenatal maternal smoking.

### 2021-11-29 Papers/Polymethylation_scores/05_results.Rmd
comments from erin

@@
 output:
+  word_document: default
   html_document: default
-  word_document: default
 bibliography: Methylation_SmokingBib.bib
@@
 
-We calculated several summary measures of DNA methylation, including polymethylation scores, global methylation, and epigenetic clocks.
+We calculated several summary measures of DNA methylation, including polymethylation scores for smoke exposure, global DNA methylation, and epigenetic clocks.
 Many of the methylation summary measures were correlated with each other (Supplemental Figure 1).
 For example, as expected, the percent of estimated immune cells was perfectly inversely correlated with the percent of estimate epithelial cells (Pearson $rho$=`r cor(modeldata$Epithelial.cells_saliva, modeldata$Leukocytes_saliva)`; *P* value<0.0001).
-Our primary polymethylation score  measure, which was calculated using coefficients from a cell-type corrected regression, was very slightly correlated estimated cell-type proportion of immune cells (Pearson $rho$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value, 3)`).
+Our polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $rho$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value, 3)`).
 
 We compared epigenetic measures between the age 9 and age 15 visit among the `r n_twovisit` individuals with data from both visits (Supplemental Figure 2).
-The correlation across ages was strongest for the primary polymethylation score (Pearson $\rho$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value<0.0001).
-The distribution of epigenetic ages shifted between the age 9 and age 15 visit, as expected (Supplemental Figure 3).
+The correlation across ages was strongest for the polymethylation score for prenatal smoke exposure (Pearson $\rho$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value<0.0001).
+Global DNA methylation (`r round(cor(model_twovisit$globalmethylation_Age_9, model_twovisit$globalmethylation_Age_15), 2)`, *P* value<0.0001) and epigenetic clocks were less strongly correlated across ages (Pediatric: `r round(cor(model_twovisit$pediatric_Age_9, model_twovisit$pediatric_Age_15), 2)`; *P* value<0.0001, GRIM: `r round(cor(model_twovisit$grim_Age_9, model_twovisit$grim_Age_15), 2)`; *P* value<0.0001) The distribution of epigenetic ages shifted higher between the age 9 and age 15 visit, as expected (Supplemental Figure 3).
+The pediatric clock more closely reflected chronological age than the GRIM clock.
 Among age 9 samples, the mean estimated age from the pediatric clock was `r round(mean(modeldata_wide$pediatric_Age_9, na.rm=T))`  and the mean estimated age from the GRIM clock was `r round(mean(modeldata_wide$grim_Age_9, na.rm=T))`.
@@
 
-In bivariate analyses and multivariable models, we focused on the polymethylation scores and the single top CpG site from prior research, cg05575921 in the *AHRR* gene,, as hypothesized biomarkers, and used global methylation and the pediatric clock as negative controls.
+In bivariate analyses and multivariable models, we focused on the polymethylation score for prenatal smoke exposure and the single top CpG site from prior research, cg05575921 in the *AHRR* gene, as hypothesized biomarkers, and used global methylation and the pediatric clock as negative controls.
 
@@
 At age 15, children exposed to prenatal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers, although this difference was not significant (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`% *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
+Epigenetic age from the pediatric and GRIM clocks and global DNA methylation did not differ between children exposed vs unexposed to prenatal maternal smoking.
 
@@
 
-The association between prenatal maternal smoking and the polymethylation score was also observed in multivariable models, adjusting for base model covariates of child sex, maternal income to poverty ratio at baseline, proportion of salivary immune cells, sample plate from DNA methlation analysis, and child age  (Figure 3; Supplemental Table 2).
+The association between prenatal maternal smoking and the polymethylation score was also observed in multivariable models, adjusting for base model covariates of child sex, maternal income to poverty ratio at baseline, proportion of salivary immune cells, sample plate from DNA methylation analysis, child age and the first two principal components of genetic ancestry  (Figure 3; Supplemental Table 2).
 At age 9, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.
-At age 15, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.
-A consistent association was observed when stratifying to the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`).
-In the African genetic ancestry sample at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.
-The direction and magnitude of the association was similar in the European and Hispanic genetic ancestry samples at age 9, though attenuated at age 15 (Figure 3; Supplemental Tables 2).
+At age 15, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher prenatal smoke exposure polymethylation score.
+A consistent association was observed when stratifying by genetic ancestry.
+In the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`).
+In the African genetic ancestry sample at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` standard deviation higher prenatal smoke exposure polymethylation score.
+The direction and magnitude of the association was similar in the European and Hispanic genetic ancestry samples at age 9, though the effect size was attenuated and no longer statistically significant at age 15 (Figure 3; Supplemental Tables 2).
 
-At age 9, prenatal maternal smoking was associated with `r round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95% CI: `r paste(round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`) after adjusting forbase model covariates.
-Similarly, at age 15 prenatal maternal smoking was associated with `r round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95%CI: (`r paste(round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`)).
-The magnitude of the association between prenatal maternal smoking and cg05575921 was reduced in European genetic ancestry samples at age 15 (Figure 3).
-The direction of the association between prenatal maternal smoking and cg05575921 was reversed in Hispanic genetic ancestry sample (Figure 3).
+At age 9, prenatal maternal smoking was associated with `r -1*round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95% CI: `r paste(round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`) after adjusting for base model covariates.
+Similarly, at age 15 prenatal maternal smoking was associated with `r -1*round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95%CI: (`r paste(round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`)).
+Prenatal maternal smoking remained significantly associated with a decrease in cg05575921 DNA methylation in the African genetic ancestry sample at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`) although the association was not significant at age 15  (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`; Figure 3).
+In the European genetic ancestry sample, the association was consistent in direction at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`) and age 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`).
+In the Hispanic genetic ancestry sample, the association was no longer significant and not consistent in direction (Age 9: `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Hispanic ancestry'", 'smkPreg_binaryYes', 2)`; Age 15: `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Hispanic ancestry'", 'smkPreg_binaryYes', 2)`) (Figure 3).
 
@@
 
-After additionally adjusting for other prenatal exposures and postnatal smoke exposure, the association between prenatal maternal smoking and the polymethylation score was very consistent (Supplemental Figure 4).
-Similarly, after additionally adjusting for other prenatal exposures and postnatal smoke exposure, the direction of the association between prenatal maternal smoking and cg055975921 was consistent, although no longer significant at age 15.
-Children exposed to postnatal smoke but not prenatal smoke did not have higher polymethylation scores than children unexposed to both postnatal and prenatal smoke.
-(Supplemental Figure 5).
+After additionally adjusting for other prenatal exposures and postnatal smoke exposure, the association between prenatal maternal smoking and the polymethylation score was robust (Supplemental Figure 4).
+Similarly, after adjusting for other prenatal exposures and postnatal smoke exposure, the direction of the association between prenatal maternal smoking and cg055975921 was consistent, although no longer significant at age 15.
+Very few children who were exposed to maternal smoking prenatally were unexposed to postnatal smoke (Supplemental Figure 5).
+However children exposed only to postnatal smoke did not have higher polymethylation scores than children unexposed to both postnatal and prenatal smoke (Supplemental Figure 5).
 Results were also similar when controlling for surrogate variables instead of known covariates (Supplemental Figure 6; Supplemental Table 2).
 
-Results were similar when using linear mixed effect models instead of age-stratified models and adjusting for base model covariates (Supplemental Table 3 & 4).
+Results from linear mixed effect models were similar to age-stratified and base covariate adjusted linear models (Supplemental Table 3 & 4).
 
-Results were also similar when using polymethylation scores constructed using alternative regression coefficients (see Methods, Supplemental Figure 7).
+Results were also similar when using polymethylation scores constructed with alternative regression coefficients (see Methods, Supplemental Figure 7).
+For our main analysis we used coefficients from a regression of sustained smoking exposure and DNA methylation in newborn cord blood with cell-type control.
+As sensitivity analyses, we used coefficients from regressions of: sustained smoking exposure and DNA methylation in newborn cord blood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children without cell-type control, and any smoking exposure and DNA methylation in newborn cord blood without cell-type control [@Joubert:2016ue].
 The association between prenatal maternal smoking and the polymethylation score using regression coefficients from newborn cord blood with cell-type controls was the strongest.
-Prenatal maternal smoking was also associated with other selected *a priori* CpG sites consistently with associations reported in previous meta-analyses (Supplemental Table 2, Supplemental Figure 8).
+
+In addition to cg055975921 in the *AHHR* gene, we tested the association of four other *a priori* probe sites identified from previous meta-analyses.
+We replicated the association of prenatal maternal smoking with these CpG sites (Supplemental Table 2, Supplemental Figure 8).
 
@@
 Next, we compared the accuracy of different DNA methylation summary measures for classifying prenatal maternal smoking using receiver operating curves (Figure 4A; Supplemental Table 5).
-We considered classification when using DNA methylation summary measures alone.
-We also considered classification when using DNA methylation summary measures in addition to a base model of child sex, maternal income to poverty ratio at baseline, child age at DNA methylation measurement, estimated immune cell proportion, plate from methylation processing and the first two components of genetic ancestry.
+We estimated classification of prenatal maternal smoking when using DNA methylation summary measures alone.
+We also estimated prenatal maternal smoking classification when using DNA methylation summary measures in addition to a base model covariates (child sex, maternal income to poverty ratio at baseline, child age at DNA methylation measurement, estimated immune cell proportion, plate from methylation processing and the first two genetic principal components).
 When used without DNA methylation measures, these base model variables had an area under the curve (AUC) of `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 9')%>%pull(auc)` at age 9 and `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 15')%>%pull(auc)` at age 15.
 
-At age 9, including the polymethylation score  significantly improved the base model accuracy (AUC: `r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(auc)`, *P* value comparing to base model<0.001).
-At age 15, similarly, including the polymethylation score significantly improved the base model accuracy (AUC:`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(auc)` , *P* value<0.001).
+At age 9, including the polymethylation score  significantly improved the base model covariates classification (AUC: `r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(auc)`, *P* value comparing to base model<0.001).
+At age 15, similarly, including the polymethylation score significantly improved the base model covariates classification (AUC:`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(auc)` , *P* value<0.001).
 
-At age 9, the base model with the polymethylation score had a larger AUC than the base model with cg05575921 (AUC base model + cg05575921: `r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(auc)`; *P* value =`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(5)`).
-At age 15, this was also true (AUC for base model + cg05575921=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(auc)`,  *P* value=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(5)`).
-Accuracy was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B).
+At age 9, the base model covariates with the polymethylation score had a larger AUC than the base model covariates with cg05575921 (AUC base model covariates + cg05575921: `r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(auc)`; *P* value =`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(5)`).
+At age 15, this was also true (AUC for base model covariates + cg05575921=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(auc)`,  *P* value=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(5)`; Supplemental Table 6).
+Accurate classification was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B).
 
@@
 
-Results were similar when using other coefficients as the weights in construction of the polymethylation scores.
-However, the AUC was always significantly higher when using the coefficients from a model which incorporated cell-type control as weights (Supplemental Table 6).
+Classification of prenatal maternal smoke exposure when using other coefficients as the weights in construction of the polymethylation scores was similar to the results when using coefficients from cell-type controlled regressions of sustained prenatal maternal smoking and DNA methylation in cord blood (Supplemental Table 7).

### 2021-11-29 Papers/Polymethylation_scores/06_discussion.Rmd
comments from erin

@@
 
-In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed that prenatal maternal smoking was associated with DNA methylation in children's saliva samples from ages 9 and 15.
-Prenatal maternal smoking was associated with polymethylation scores across strata of child age and ancestry.
-Global methylation and epigenetic clocks were not associated with maternal smoking exposure demonstrating a specific and persistent signal of prenatal smoke exposure in DNA methylation.
-Polymethylation scores had reasonable accuracy for classifying prenatal maternal smoking.
-This accuracy was more consistent across time than when using individual *a priori* CpG sites.
+In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed that prenatal maternal smoking was associated with several characterizations of DNA methylation in children's saliva samples from ages 9 and 15.
+Prenatal maternal smoking was associated with polymethylation scores across strata of both child age and genetic ancestry.
+Global methylation and epigenetic clocks were not associated with maternal smoking exposure.
+Polymethylation scores had reasonable accuracy for classifying prenatal maternal smoking (AUC \> 0.7).
+Classification when using polymethylation scores for prenatal maternal smoking was better than when using a single *a priori* CpG site, cg05575921 in the *AHRR* gene.
 
 Our findings are consistent with the previous literature on associations between prenatal maternal smoking and DNA methylation.
-We replicated top hits from previous epigenome wide association studies of DNA methylation in cord and peripheral blood[@Joubert:2016ue; @wiklund2019].
+We replicated the top hit from a previous epigenome wide association analysis [@Joubert:2016ue].
+Further, we found evidence of association in several additional hits from epigenome wide association studies of DNA methylation in cord and peripheral blood [@Joubert:2016ue; @wiklund2019].
 There are no previous studies of prenatal maternal smoking and saliva DNA methylation.
@@
 
-We advance the literature on prenatal maternal smoking and DNA methylation by evaluating the persistence of the association as children age and its portability across tissue and ancestry.
+We advance prenatal smoking - DNA methylation literature by evaluating the persistence of the association between prenatal maternal smoking and DNA methylation as children age and its portability across tissue and ancestry.
 Polymethylation scores built using coefficients from meta-analysis of cord blood DNA methylation from primarily European-ancestry newborns were still associated with prenatal maternal smoking in our independent saliva samples from a diverse cohort at ages 9 and 15 [@Joubert:2016ue].
-The portability of other risk scores, such as polygenic risk scores, across ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Hüls2020].
-In this case, the polymethylation score for prenatal maternal smoking appears to be portable across ancestry groups.
-Effect estimates were consistent across ancestries at age 9, although Hispanic and European genetic ancestry samples had lower effect estimates at age 15.
-This could reflect higher unreported smoking initiation in European and Hispanic ancestry teens in the United States than African ancestry teens.
+The portability of other risk scores, such as polygenic risk scores, across genetic ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Hüls2020].
+In this case, the polymethylation score for prenatal maternal smoking appears to be portable across genetic ancestry groups.
+Effect estimates were consistent across ancestries at age 9, although Hispanic and European genetic ancestry samples had lower, non-significant effect estimates at age 15.
+The decreased precision may be the result of the reduced sample size of the HIspanic and European genetic ancestry samples.
+The reduction in effect estimate magnitude could reflect higher unreported smoking initiation in European and Hispanic ancestry teens in the United States than African ancestry teens.
 Genetic ancestry correlates with race, and White and Latino teens have much higher rates of teen smoking and earlier ages at initiation than Black children [@El-Toukhy2016].
 Though we excluded children who reported own-smoking from our analytic sample, under-reporting of own-smoking could influence DNA methylation at ages 9 and 15, creating outcome misclassification.
-This could also explain the observed attenuation the effect on AHRR:cg05575921 methylation by age 15 in our sample, as DNA methylation at this probe is known to vary by personal cigarette smoking[@Dawes2019; @Philibert2020].
+This could also explain the observed attenuation of the effect on *AHRR*:cg05575921 methylation by age 15 in our sample, as DNA methylation at this probe is known to vary by personal cigarette smoking[@Dawes2019; @Philibert2020].
 
@@
 DNA methylation biomarkers may be especially susceptible to confounding by subject age at methylation measurement and cell-type proportion [@Hüls2020].
-Accuracy of polymethylation score classification was similar when using coefficients from cord blood vs peripheral blood samples from older children.
-Accuracy was not improved by using two time-points of methylation measurement.
-Accuracy was improved when polymethylation scores used coefficients which incorporated cell-type control.
-Our analysis therefore suggests that cell-type control when both generating and applying coefficients for polymethylation scores may positively influence accuracy, although more work comparing coefficients from different populations is still needed.
+Prenatal maternal smoke exposure classification accuracy of polymethylation score was similar when using coefficients from cord blood vs peripheral blood samples from older children.
+Classification accuracy was not improved by using two time-points of methylation measurement.
+Classification accuracy was improved when polymethylation scores used coefficients which incorporated cell-type control.
+Our analysis therefore suggests that cell-type control when both generating (in the epigenome wide association study) and applying coefficients for polymethylation scores may positively influence outcome classification accuracy, although more work comparing coefficients from different populations is still needed.
 
 Our analysis suggests that polymethylation scores may be more accurate than using single CpG sites as biomarkers.
-AHRR:cg05575921 is a consistent marker of prenatal maternal smoke exposure in meta-analyses of newborn cord blood[@Joubert:2016ue].
-DNA methylation at AHRR:cg05575291 can accurately classify own--smoking behavior in both blood (area under the curve 0.995) and saliva (area under the curve 0.971) [@Dawes2019; @Philibert2020].
+The site cg05575921 in the *AHRR* gene is a consistent marker of prenatal maternal smoke exposure in meta-analyses of newborn cord blood[@Joubert:2016ue].
+DNA methylation at cg05575291 in the *AHRR* gene can accurately classify own--smoking behavior in both blood (area under the curve 0.995) and saliva (area under the curve 0.971) [@Dawes2019; @Philibert2020].
 However, salivary AHRR:cg05575921 was not a persistent marker of prenatal maternal smoking in an analysis of middle-aged adult women [@Richmond:2018tu].
@@
 Our analysis is not without its limitations.
-We used maternal self report of prenatal maternal smoker as serum cotinine levels were not available.
+We used maternal self-report of prenatal maternal smoke exposure, as serum cotinine levels were not available.
 Due to social desirability bias, this could result in exposure misclassification.
@@
 In any analysis of a prenatal exposure and postnatal outcome, there is the possibility of selection bias into the cohort due to live birth bias.
-Selection bias is also possible due to loss-to-follow-up between birth and age 9.
+Selection bias is also possible due to loss-to-follow-up between birth and age 15.
 Additionally, while we controlled for postnatal secondhand smoke exposure and excluded children who reported any own-smoking, we cannot exclude the possibility of residual confounding in this observational cohort.
@@
 However, our analysis also has several strengths.
-We analyzed samples from a large sample of diverse participants underrepresented in genetic and epigenetic research [@Sirugo2019; @Bentley2017].
-While our exposure measurement of prenatal maternal smoking was self reported, it was assessed prospectively and preceded outcome measurements.
+We analyzed samples from a large cohort of diverse participants underrepresented in genetic and epigenetic research [@Sirugo2019; @Bentley2017].
+While our exposure measurement of prenatal maternal smoking was self-reported, it was assessed prospectively and preceded outcome measurements.
 We analyzed repeated measures of DNA methylation with reproducible array measures conducted in a single batch.

### 2021-11-29 Papers/Polymethylation_scores/07_conclusions.Rmd
comments from erin

@@
 
-In a large, prospective study of diverse participants, we showed that DNA methylation in children's saliva had strong associations with and reasonable accuracy for prenatal maternal smoke exposure.
-Further, we demonstrated that polymethylation scores could be applied as a biomarker of prenatal maternal smoke exposure across ancestry-groups, an important consideration for the equitable biomarker development.
-The development and application of biomarker for prenatal maternal smoke exposure has important implications for epidemiological research and clinical practice.
+In a large, prospective study of diverse participants, we showed that DNA methylation in children's saliva had strong associations with and reasonable classification accuracy for prenatal maternal smoke exposure.
+Further, we demonstrated that polymethylation scores could be applied as a biomarker of prenatal maternal smoke exposure across genetic ancestry groups, an important consideration for the equitable biomarker development.
+The development and application of biomarkers for prenatal maternal smoke exposure has important implications for epidemiological research and clinical practice.
 Given the difficulty of measuring prenatal maternal smoke exposure, such a biomarker could allow for confounder control in research areas where such control is currently impossible.

### 2021-11-29 Papers/Polymethylation_scores/08_methods.Rmd
comments from erin

@@
 
-Child genetic ancestry was calculated. Principal components of child genetic ancestry were calculated from genetic data (CHIP?) from child saliva samples.
+Child genetic ancestry was calculated.
+Grants R01 HD36916, R01 HD073352, and R01 HD076592 provided support for the collection, assay, and analysis of the genetic data in the Fragile Families cohort.
+Principal components (PC) of child genetic ancestry were calculated from genetic data using Illumina PsychChip_v1-1 with child saliva samples.
+Genetic ancestry was assigned by comparing PC loadings to 1000 Genomes super population clusters.
+Samples with PC1 \> 0.018 and PC2 \> -0.0075 were assigned to European ancestry.
+Samples with PC1 \< -0.005 and PC2 \> 0.007+0.75(PC1) were assigned to African ancestry.
+Samples with PC1 \> 0.018 and -0.055 \< PC2 \< 0.025 were assigned to Hispanic ancestry.
 
@@
 CpG sites were removed if they had detection p-value \>0.01 in 5% of samples (n=26,830) or were identified as cross-reactive (n=27,782) (Figure 1) [@Chen2013].
+We used the November 2021 data freeze of the Fragile Families and Child Wellbeing DNA methylation data.
 Relative proportions of immune and epithelial cell types were estimated from DNA methylation measures using a childhood saliva reference panel [@Middleton2021].
@@
 For our main analysis we used coefficients from a regression of sustained smoking exposure and DNA methylation in newborn cord blood with cell-type control.
-As sensitivity analyses, we used re sustained smoking exposure and DNA methylation in newborn cord blood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children, any smoking exposure and DNA methylation in newborn cord blood) [@Joubert:2016ue].
+As sensitivity analyses, we used sustained smoking exposure and DNA methylation in newborn cord blood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children, any smoking exposure and DNA methylation in newborn cord blood) [@Joubert:2016ue].
 

### 2021-11-29 Papers/Polymethylation_scores/98_declarations.Rmd
comments from erin

@@
 
-The authors have no competing interests to declare. check here
+The authors have no competing interests to declare.
 
@@
 
-This research was made possible through several grants (add). FB was supported by the National Institutes of Health National Institute of Dental and Craniofacial Research (F31DE029992).
+This research was made possible through several grants (Fragile Families Core Data Collection: R01 HD036916, R25 AG053227, genetic data processing R01 HD36916, R01 HD073352, and R01 HD076592).
+FB was supported by the National Institutes of Health National Institute of Dental and Craniofacial Research (F31 DE029992).
+EB was supported by (R01 AG055406, R01 AG067592, R01 MD011716).
 
@@
 FB and KB wrote the paper with all authors contributing to revisions.
-DN, CM, KM, EW and FB all contributed to the design and conception of the study.
+CM, DN and LS contributed to the design and processing of the DNA methylation data subcohort.
+DN, CM, KM, EW, and FB all contributed to the design and conception of the analytic plan.
+All authors reviewed and revised the manuscript.
 

### 2021-11-29 Reports/Paper1.Rmd
updated results post datafreeze update

@@
 set_label(myFF$MethylData2)<-'Exclusions based on missing data'
-myFF=left_join(myFF, methyldata%>%mutate(idnum=as.character(idnum)))
+myFF=left_join(myFF, methyldata)%>%mutate(idnum=as.character(id))
 myFF=myFF%>%mutate(across(c(PostnatalMaternalSmokingAny, SmkAtVisitPastmonth), ~ifelse(.x=='Missing', 

### 2021-11-30 Scripts/01_clean_phenotype_data.R
added city variable to models

@@ -35,12 +35,13 @@ pheno=set_labels(pheno, labels=attributes(pheno)$label.table)
 FF_factors=FFmeta%>%
   filter(type%in% c('Binary', 'Ordered Categorical', 'Unordered Categorical'))%>%
   pull(old_name)
+FF_factors=c(FF_factors, 'm1city')
 #recode binary, categorical variables so they have label names as values (nicer for table and plotting)
 pheno=pheno %>% dplyr::mutate(across(any_of(FF_factors), ~sjlabelled::as_label(., drop.levels=TRUE)))
 
 #dplyr::select variables
 idvars<-c("id", "ff_id")
-demovars<-c("cm1age", "cm1bsex", "cm1ethrace", "cm1edu", "cm1hhinc", "cm1inpov", "m1h3", "m1h3a", "m1h3b", "m1i4", "m1i4a", "m1i4b", "cf1age", "cf1ethrace", "f1h3", "f1h3a", "f1h3b", "ch5agem","ch6yagem", "c6yagey", "cm5b_age", "cm5b_ageyrs", "cp6yagem", "cp6yagey", "hv5_agem", "hv6_yagem", "hv6_yagey")
+demovars<-c("m1city", "cm1age", "cm1bsex", "cm1ethrace", "cm1edu", "cm1hhinc", "cm1inpov", "m1h3", "m1h3a", "m1h3b", "m1i4", "m1i4a", "m1i4b", "cf1age", "cf1ethrace", "f1h3", "f1h3a", "f1h3b", "ch5agem","ch6yagem", "c6yagey", "cm5b_age", "cm5b_ageyrs", "cp6yagem", "cp6yagey", "hv5_agem", "hv6_yagem", "hv6_yagey")
[…]

### 2021-11-30 Scripts/04_summary_methyl_models.R

@@ -72,7 +72,7 @@ names(outcome_labels)=y_vector
 
 
 #model variables (excepting PCs, add in individually for global & ancestry specific models)
-base_model_vars<-"~smkPreg_binary+cm1bsex+cm1inpov+Leukocytes_saliva+Sample_Plate+ChildAgeComposite"
+base_model_vars<-"~smkPreg_binary+cm1bsex+cm1inpov+Leukocytes_saliva+Sample_Plate+ChildAgeComposite+city_binary"
 prenatal_model_vars<-paste0(base_model_vars, "+m1g2_YesNoPreg+m1g3_YesNoPreg")
 secondhand_model_vars<-paste0(prenatal_model_vars, '+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth')
 interaction_model_vars<-paste0(secondhand_model_vars, '+ChildAgeComposite:smkPreg_binary')
@@ -195,7 +195,7 @@ longitudinal_local_models=my_clean_argnames(longitudinal_local_models, local_pre
 modeldata=modeldata%>%mutate(smkPreg_binaryN=case_when(smkPreg_binary=='Yes'~1, smkPreg_binary=='No'~0))
 
 #global predictors, cross sectional
-roc_predictors=c('', gsub('~smkPreg_binary', '', paste0(base_model_vars, global_pcs)))
+roc_predictors=c('', gsub('+m1city', '', gsub('~smkPreg_binary', '', paste0(base_model_vars, global_pcs))))
 roc_outcomes=c('', y_vector)
 args=list('childteen'=age_vector, 'methylation'=roc_outcomes, 'covariates'=roc_predictors)%>%cross_df()%>%mutate(predictors=paste0(methylation, covariates))%>%
   filter(predictors!='')

### 2021-11-30 Papers/Polymethylation_scores/02_abstract.Rmd
colter revisions

@@
+---
+output:
+  pdf_document: default
+  html_document: default
+---
 # Abstract
@@
 **Results**: We created saliva polymethylation scores using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cordblood.
-In the full sample at age 9 (n=756), prenatal maternal smoke exposure was associated with a 0.52 (95%CI: 0.36, 0.67) standard deviation higher polymethylation score.
+In the full sample at age 9 (n=753), prenatal maternal smoke exposure was associated with a 0.52 (95%CI: 0.36, 0.67) standard deviation higher polymethylation score.
 The direction and magnitude of the association was consistent when stratified by genetic ancestries.
-In the full sample at age 15 (n=749), prenatal maternal smoke exposure was associated with a 0.46 (95%CI: 0.3, 0.62) standard deviation higher polymetylation score, and the association was attenuated among the European and Hispanic genetic ancestry samples.
+In the full sample at age 15 (n=746), prenatal maternal smoke exposure was associated with a 0.46 (95%CI: 0.3, 0.62) standard deviation higher polymethylation score, and the association was attenuated among the European and Hispanic genetic ancestry samples.
 The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (AUC age 9=0.77, *P* value<0.001, age 15=0.77, *P* value<0.001).

### 2021-11-30 Papers/Polymethylation_scores/04_background.Rmd
colter revisions

@@
 output:
+  pdf_document: default
+  html_document: default
   word_document: default
-  html_document: default
 bibliography: Methylation_SmokingBib.bib
+latex_engine: xelatex
 ---
@@
 Adjusting for ancestry did not substantially affect the association between prenatal maternal smoking and DNA methylation at any of the sites, although the authors note that African American women had on average 2--5% higher mean DNA methylation on the absolute scale as compared with White and Hispanic women at 5 *CYP1A1* CpG sites [@Tehranifar:2018tb].
-Of 148 CpG sites selected based on their association with prenatal maternal smoking among primarily European children, 7 CpG sites were also associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue].
+Of 148 CpG sites selected based on their association with prenatal maternal smoking among primarily European children, 7 CpG sites were also associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Joubert:2016ue; @Neophytou:2019ty].
 While prenatal maternal cigarette smoking is associated with DNA methylation, the consistency of this association across genetic ancestry is understudied.
@@
 After association testing, an important next step is to evaluate DNA methylation as a biomarker of prenatal maternal smoke exposure by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior.
-Biomarkers can consist of single DNA methylation sites or summary measures across multiple sites [@Philibert2020; @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa].
+Biomarkers can consist of single DNA methylation sites or summary measures across multiple sites [@Philibert2020 @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa].
 One option is a polymethylation score, analogous to a polygenic score.
@@
 In the Fragile Families and Child Wellbeing study, a diverse longitudinal birth cohort of children, we aimed to assess the potential of salivary DNA methylation biomarkers for prenatal maternal cigarette smoking.
-We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including polymethylation scores for smoke exposure, individual *a priori* CpG sites, global DNA methylation, and epigenetic clocks.
+We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including polymethylation scores for smoke exposure, individual a priori CpG sites, global DNA methylation, and epigenetic clocks.
 We then performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a persistent and consistent biomarker of prenatal maternal smoking.

### 2021-11-30 Papers/Polymethylation_scores/05_results.Rmd
colter revisions

@@
   html_document: default
+  pdf_document: default
 bibliography: Methylation_SmokingBib.bib
-always_allow_html: true
+always_allow_html: yes
 ---
@@
 Many of the methylation summary measures were correlated with each other (Supplemental Figure 1).
-For example, as expected, the percent of estimated immune cells was perfectly inversely correlated with the percent of estimate epithelial cells (Pearson $rho$=`r cor(modeldata$Epithelial.cells_saliva, modeldata$Leukocytes_saliva)`; *P* value<0.0001).
-Our polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $rho$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value, 3)`).
+For example, as expected, the percent of estimated immune cells was perfectly inversely correlated with the percent of estimate epithelial cells (Pearson $\rho$=`r cor(modeldata$Epithelial.cells_saliva, modeldata$Leukocytes_saliva)`; *P* value<0.0001).
+Our polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $\rho$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value, 3)`).
 
@@
 
-The association between prenatal maternal smoking and the polymethylation score was also observed in multivariable models, adjusting for base model covariates of child sex, maternal income to poverty ratio at baseline, proportion of salivary immune cells, sample plate from DNA methylation analysis, child age and the first two principal components of genetic ancestry  (Figure 3; Supplemental Table 2).
-At age 9, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.
-At age 15, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher prenatal smoke exposure polymethylation score.
+The association between prenatal maternal smoking and the polymethylation score for prenatal smoke exposure was also observed in multivariable models, adjusting for base model covariates of child sex, maternal income to poverty ratio at baseline, proportion of salivary immune cells, sample plate from DNA methylation analysis, child age and the first two principal components of genetic ancestry  (Figure 3; Supplemental Table 2).
+At age 9, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score for prenatal smoke exposure.
+At age 15, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher prenatal smoke exposure polymethylation score for prenatal smoke exposure.
 A consistent association was observed when stratifying by genetic ancestry.
-In the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`).
-In the African genetic ancestry sample at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` standard deviation higher prenatal smoke exposure polymethylation score.
+In the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`) at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score for prenatal smoke exposure.
 The direction and magnitude of the association was similar in the European and Hispanic genetic ancestry samples at age 9, though the effect size was attenuated and no longer statistically significant at age 15 (Figure 3; Supplemental Tables 2).
@@
 In the European genetic ancestry sample, the association was consistent in direction at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`) and age 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`).
-In the Hispanic genetic ancestry sample, the association was no longer significant and not consistent in direction (Age 9: `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Hispanic ancestry'", 'smkPreg_binaryYes', 2)`; Age 15: `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Hispanic ancestry'", 'smkPreg_binaryYes', 2)`) (Figure 3).
+In the Hispanic genetic ancestry sample, the association was no longer significant and not consistent in direction (Age 9: `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Hispanic ancestry'", 'smkPreg_binaryYes', 2)`; Age 15: `r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Hispanic ancestry'", 'smkPreg_binaryYes', 2)`) (Figure 3).
 
@@
 
-At age 9, including the polymethylation score  significantly improved the base model covariates classification (AUC: `r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(auc)`, *P* value comparing to base model<0.001).
-At age 15, similarly, including the polymethylation score significantly improved the base model covariates classification (AUC:`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(auc)` , *P* value<0.001).
+At age 9, including the polymethylation score for prenatal smoke exposure significantly improved the base model covariates classification (AUC: `r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(auc)`, *P* value comparing to base model<0.001).
+At age 15, similarly, including the polymethylation score for prenatal smoke exposure significantly improved the base model covariates classification (AUC:`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(auc)` , *P* value<0.001).
 
-At age 9, the base model covariates with the polymethylation score had a larger AUC than the base model covariates with cg05575921 (AUC base model covariates + cg05575921: `r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(auc)`; *P* value =`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(5)`).
+At age 9, the base model covariates with the polymethylation score for prenatal smoke exposure had a larger AUC than the base model covariates with cg05575921 (AUC base model covariates + cg05575921: `r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(auc)`; *P* value =`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(5)`).
 At age 15, this was also true (AUC for base model covariates + cg05575921=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(auc)`,  *P* value=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(5)`; Supplemental Table 6).

### 2021-11-30 Papers/Polymethylation_scores/06_discussion.Rmd
colter revisions

@@
 In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed that prenatal maternal smoking was associated with several characterizations of DNA methylation in children's saliva samples from ages 9 and 15.
-Prenatal maternal smoking was associated with polymethylation scores across strata of both child age and genetic ancestry.
+Prenatal maternal smoking was associated with polymethylation scores for prenatal smoke exposure across strata of both child age and genetic ancestry.
 Global methylation and epigenetic clocks were not associated with maternal smoking exposure.
-Polymethylation scores had reasonable accuracy for classifying prenatal maternal smoking (AUC \> 0.7).
-Classification when using polymethylation scores for prenatal maternal smoking was better than when using a single *a priori* CpG site, cg05575921 in the *AHRR* gene.
+Polymethylation scores for prenatal smoke exposure had reasonable accuracy for classifying prenatal maternal smoking (AUC \> 0.7).
+Classification when using polymethylation scores for prenatal maternal smoke exposure was better than when using a single *a priori* CpG site, cg05575921 in the *AHRR* gene.
 
@@
 Polymethylation scores built using coefficients from meta-analysis of cord blood DNA methylation from primarily European-ancestry newborns were still associated with prenatal maternal smoking in our independent saliva samples from a diverse cohort at ages 9 and 15 [@Joubert:2016ue].
-The portability of other risk scores, such as polygenic risk scores, across genetic ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Hüls2020].
+The portability of other risk scores, such as polygenic risk scores, across genetic ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Huls2020].
 In this case, the polymethylation score for prenatal maternal smoking appears to be portable across genetic ancestry groups.
@@
 Different methods for summarizing across multiple DNA methylation sites may also influence biomarker performance.
-The similarity in performance between saliva DNA methylation in our study and peripheral blood DNA methylation in older adults is encouraging for the use of saliva DNA methylation as a readily accessible clinical sample.
+The similarity in performance between saliva DNA methylation in our study and peripheral blood DNA methylation in older adults is encouraging for the use of saliva as a readily accessible clinical sample.
 
 Our analysis contributes to the development of an accurate methylation biomarker for prenatal maternal smoking by evaluating the impact of specific methodological choices on accuracy of classification.
-DNA methylation biomarkers may be especially susceptible to confounding by subject age at methylation measurement and cell-type proportion [@Hüls2020].
+DNA methylation biomarkers may be especially susceptible to confounding by subject age at methylation measurement and cell-type proportion [@Huls2020].
 Prenatal maternal smoke exposure classification accuracy of polymethylation score was similar when using coefficients from cord blood vs peripheral blood samples from older children.

### 2021-11-30 Papers/Polymethylation_scores/index_paper.Rmd
colter revisions

@@
 title: "PAPER DRAFT"
-author: "------------"
+author: '------------'
 date: "`r Sys.Date()`"
@@
     df_print: paged
+  pdf_document: default
 bibliography: Methylation_SmokingBib.bib
 csl: springer-vancouver.csl
-always_allow_html: true
+always_allow_html: yes
 ---

### 2021-11-30 Reports/Paper1.Rmd
adding city to results tables

@@
 n_Child<-paste0('n=', childdata %>%filter(childteen=='C')  %>% nrow(), " individuals with a 9 year visit")
-n_ChildSmkexcluded<-paste0("Exclude children who smoke: \n",'n=', childdata%>%filter(childteen=='C' & k5f1l=='2 no')%>% nrow(), 
+n_ChildSmkexcluded<-paste0("Exclude children who smoke: \n",'n=', childdata%>%filter((k5f1l=="2 no" | (is.na(k5f1l) & k6d40=='2 No')) & childteen=='C')%>% nrow(), 
                            " individuals with a 9 year visit")
-n_ChildSmkexclusions<-paste0('n=', childdata %>% filter((k5f1l=="1 yes" | is.na(k5f1l)) & childteen=='C') %>% nrow(), 
+n_ChildSmkexclusions<-paste0('n=', childdata %>% filter((k5f1l=="1 yes" | (is.na(k5f1l) & (k6d40!='2 No' | is.na(k6d40)))) & childteen=='C') %>% nrow(), 
                              " individuals smoking \n or missing child smoking data \n at 9 year visit")
+
 n_Teen<-paste0('n=', childdata %>%filter(childteen=='T')   %>% nrow(), " individuals with a 15 year visit")
-n_TeenSmkexcluded<-paste0("Exclude children who smoke: \n",'n=', childdata %>%filter(childteen=='T' & k6d40=="2 No") %>% nrow(), 
+n_TeenSmkexcluded<-paste0("Exclude children who smoke: \n",'n=', childdata %>%filter(childteen=='T' & k6d40=="2 No" & (k5f1l!='1 yes' | is.na(k5f1l))) %>% nrow(), 
                           " individuals with a 15 year visit")
@@
                             " individuals smoking \n or missing focal child smoking data \n at 15 year visit")
+
+test=childdata %>%filter((childteen=='T' & k6d40=="2 No") | (childteen=='C' & k5f1l=='2 no'))
 
@@
 
+save_png <- function(plot, path){
+  DiagrammeRsvg::export_svg(plot) %>%
+    charToRaw() %>%
+    rsvg::rsvg() %>%
+    png::writePNG(path)
+}
+save_png(graph1, file.path(outdir, paste0('Figure', main_fig, ".png")))
+
+ DiagrammeRsvg::export_svg(graph1) %>%
+    charToRaw() %>%
+    rsvg::rsvg_ps(file.path(outdir, paste0('Figure', main_fig, ".ps")))
+
 main_fig=main_fig+1            
@@
 exclusion_tbl=myFF %>% mutate(childteen=case_when(childteen=='C'~'Age 9', childteen=='T'~'Age 15'))%>%
-    dplyr::select(smkPreg_binary, childteen, MethylData2, SSnewbornCT_center, cg05575921, pediatric, globalmethylation, Leukocytes_saliva, Epithelial.cells_saliva, ancestry, cm1bsex,  m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,SmkAtVisitPastmonth )%>%
+    dplyr::select(smkPreg_binary, childteen, MethylData2, SSnewbornCT_center, cg05575921, pediatric, globalmethylation, Leukocytes_saliva, Epithelial.cells_saliva, ancestry, cm1bsex, city_binary,  m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,SmkAtVisitPastmonth )%>%
   mutate(smkPreg_binary=factor(smkPreg_binary, levels=c('No', 'Yes'), labels=c('Not exposed', 'Smoke exposed')), 
@@
                          'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'cm1inpov',
-                         'smkPreg_binary', 'PostnatalMaternalSmokingAny')))%>%unique()
+                         'smkPreg_binary', 'PostnatalMaternalSmokingAny', 'city_binary')))%>%unique()
 ```
@@
 ggexport(fig2, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')))
+ggexport(fig2, filename=file.path(outdir, paste0('Figure', main_fig, '.eps')))
 main_fig=main_fig+1
@@
 smoking_tbl=modeldata %>% 
-  dplyr::select(smkPreg_binary, childteen, SSnewbornCT_center, cg05575921, pediatric, globalmethylation, Leukocytes_saliva, Epithelial.cells_saliva, ancestry, cm1bsex,  m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,SmkAtVisitPastmonth )%>%
+  dplyr::select(smkPreg_binary, childteen, SSnewbornCT_center, cg05575921, pediatric, globalmethylation, Leukocytes_saliva, Epithelial.cells_saliva, ancestry, cm1bsex,  city_binary, m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,SmkAtVisitPastmonth )%>%
   mutate(smkPreg_binary=factor(smkPreg_binary, levels=c('No', 'Yes'), labels=c('Not exposed', 'Smoke exposed')), 
@@
 ggexport(forest_plot, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')), width=19)
+ggexport(forest_plot, filename=file.path(outdir, paste0('Figure', main_fig, '.eps')), width=19)
 main_fig=main_fig+1
@@
 
-ggexport(forest_plot_compareModels, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=15, height=15)
+ggexport(forest_plot_compareModels, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=18, height=15)
 sup_fig=sup_fig+1
@@
 ggexport(auc_fig, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')), width=9)
+ggexport(auc_fig, filename=file.path(outdir, paste0('Figure', main_fig, '.eps')), width=9)
 main_fig=main_fig+1

### 2022-02-02 Papers/Polymethylation_scores/05_results.Rmd
reviewer comment: distributions of immune cell proportions

@@
 
-We calculated several summary measures of DNA methylation, including polymethylation scores for smoke exposure, global DNA methylation, and epigenetic clocks.
+We calculated several summary measures of DNA methylation, including polymethylation scores for smoke exposure, global DNA methylation,epigenetic clocks, and estimated cell-type proportions.
 Many of the methylation summary measures were correlated with each other (Supplemental Figure 1).
-For example, as expected, the percent of estimated immune cells was perfectly inversely correlated with the percent of estimate epithelial cells (Pearson $\rho$=`r cor(modeldata$Epithelial.cells_saliva, modeldata$Leukocytes_saliva)`; *P* value<0.0001).
 Our polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $\rho$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value, 3)`).
@@
 The distribution of estimated cell-type proportions also shifted between visits while the distribution of global methylation and polymethylation scores were more consistent (Supplemental Figure 3).
+Almost all saliva samples consisted of mostly immune cells based on estimated cell-type proportions.
+At age 9, the mean estimated immune cell proportion was `r round(mean(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)` (range: `r paste(round(range(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2), collapse=', ')`, 25% and 75% quantile: `r paste(round(quantile(modeldata_wide$Leukocytes_saliva_Age_9, c(0.25, 0.75), na.rm=T), 2), collapse = ', ')`).
+At age 15, the mean estimated immune cell proportion was `r round(mean(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2)` (range: `r paste(round(range(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2), collapse=', ')`, 25% & 75% quantile: `r paste(round(quantile(modeldata_wide$Leukocytes_saliva_Age_15, c(0.25, 0.75), na.rm=T), 2), collapse = ', ')`)
 

### 2022-02-02 Reports/Paper1.Rmd
reviewer comment: 95% CI AUC and 95% CI AUC-differences

@@
 ```{r supplemental table auc}
-roc_table=global_roc %>% dplyr::select(childteen, methylation, modeltype, roc, auc)
-roc_base=roc_table %>% filter(modeltype=='Base model+No methylation')%>%mutate(roc_basemodel=roc)%>%dplyr::select(childteen, roc_basemodel)
-roc_ss=roc_table %>% filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)')%>%mutate(roc_pms=roc)%>%dplyr::select(childteen, roc_pms)
+#CI.delong is the confidence interval for each auc: if we sampled 100 times, 95 of the samples would have a auc between those two numbers
+roc_table=global_roc %>% 
+  dplyr::select(childteen, methylation, modeltype, roc, auc, ci_auc_delong, ci_auc_boot)%>%
+  mutate(ci.delong=map_chr(ci_auc_delong, ~paste(round(.x[1], 2), round(.x[3], 2), sep=', ')))
 
+roc_base=roc_table %>% 
+  filter(modeltype=='Base model+No methylation')%>%
+  rename_with(~str_c(., '_basemodel'), .cols=c(roc, auc, ci.delong))%>%
+  dplyr::select(childteen, contains('_basemodel'))
+
+roc_ss=roc_table %>% 
+  filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)')%>%
+  rename_with(~str_c(., '_pms'), .cols=c(roc, auc, ci.delong))%>%
+  dplyr::select(childteen, contains('_pms'))
+
+#confidence_interval_difference_pms|base is the confidence interval for the DIFFERENCE between the two aucs - if we took 100 samples of the difference between the aucs, 95 of them would be between those two numbers: avoids incorrect error propagation, since the squares of errors add, not the errors themselves (see an explanation here: https://towardsdatascience.com/why-overlapping-confidence-intervals-mean-nothing-about-statistical-significance-48360559900a)
 rct1=roc_table%>% 
@@
   mutate(`P value vs base model`=map_dbl(roctest_base, ~round(.x$p.value, 4)))%>%
-  dplyr::select(childteen, modeltype, auc, `P value vs base model`, `P value vs polymethylation score`)
+  mutate(confidence_interval_difference_base=map_chr(roctest_base, ~paste(round(.x$conf.int[1], 2), round(.x$conf.int[2], 2), sep=', ')))%>%
+  mutate(confidence_interval_difference_pms=map_chr(roctest_ss, ~paste(round(.x$conf.int[1], 2), round(.x$conf.int[2], 2), sep=', ')))%>%
+  dplyr::select(childteen, modeltype, auc, `P value vs base model`, `P value vs polymethylation score`, confidence_interval_difference_base, confidence_interval_difference_pms)
 

### 2022-02-02 Scripts/04_summary_methyl_models.R

@@ -209,7 +209,9 @@ global_roc=modeldata%>%
   mutate(model=pmap(list(data, 'smkPreg_binaryN~', predictors), model_reversed), 
          tidied=map(model, tidy, conf.int=T), 
          augmented=map(model, augment, type.predict = "response"), 
-         roc=map(augmented, ~roc(.x$smkPreg_binaryN, .x$.fitted)), 
+         roc=map(augmented, ~roc(.x$smkPreg_binaryN, .x$.fitted)),
+         ci_auc_delong=map(roc, ~ci.auc(.x, conf.level=0.95, method='delong')), 
+         #ci_auc_boot=map(roc, ~ci.auc(.x, conf.level=0.95, method='bootstrap')),
          test=map(roc, ~data.frame('sens'=.x$sensitivities, spec=.x$specificities)))%>%
   left_join(methyl_names)%>%
   mutate(modeltype=case_when(covariates==''~name_methyl, 
@@ -226,6 +228,8 @@ local_roc=modeldata%>%group_by(childteen, ancestry)%>%nest()%>%left_join(args)%>
          tidied=map(model, tidy, conf.int=T), 
          augmented=map(model, augment, type.predict = "response"), 
          roc=map(augmented, ~roc(.x$smkPreg_binaryN, .x$.fitted)), 
+         ci_auc_delong=map(roc, ~ci.auc(.x, conf.level=0.95, method='delong')), 
+         #ci_auc_boot=map(roc, ~ci.auc(.x, conf.level=0.95, method='bootstrap')),
          test=map(roc, ~data.frame('sens'=.x$sensitivities, spec=.x$specificities)))%>%
   left_join(methyl_names)%>%
   mutate(modeltype=case_when(covariates==''~name_methyl, 
@@ -252,6 +256,8 @@ long_global_roc=args %>%
          tidied=map(model, tidy, conf.int=T), 
          augmented=map(model, augment, type.predict = "response"), 
          roc=map(augmented, ~roc(.x$smkPreg_binaryN, .x$.fitted)), 
+         ci_auc_delong=map(roc, ~ci.auc(.x, conf.level=0.95, method='delong')), 
+         #ci_auc_boot=map(roc, ~ci.auc(.x, conf.level=0.95, method='bootstrap')),
          test=map(roc, ~data.frame('sens'=.x$sensitivities, spec=.x$specificities)))
 
[…]

### 2022-02-14 Papers/Polymethylation_scores/04_background.Rmd
revise intro and discussion

--- 2021-11-10 Papers/Polymethylation_scores/04_background.Rmd
@@
   word_document: default
+  pdf_document: default
   html_document: default
 bibliography: Methylation_SmokingBib.bib
+latex_engine: xelatex
 ---
@@
 
-Maternal prenatal smoking is a public health concern.
 In 2018, 11% of pregnant women in the United States aged 15-44 reported any past-month cigarette use [@SAMHSA2018].
@@
 Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua].
-The gold-standard for smoking measurements, serum cotinine levels, have a short half-life of nine hours in pregnant women [@Dempsey:2002wr].
-Further, serum cotinine measures during pregnancy are rarely available outside of birth cohorts, limiting clinical application as a marker of prenatal exposures.
-Developing a portable and reliable biomarker of prenatal maternal smoke exposure would have important implications for research and clinical practice.
+The gold-standard biomarker, serum cotinine levels, have a half-life of nine hours in pregnant women, and are rarely available outisde of birth cohorts[@Dempsey:2002wr].
+Alternative biomarkers of prenatal maternal smoke exposure could improve research and clinical practice.
 
-Maternal cigarette smoking during pregnancy is associated with reproducible DNA methylation signatures, measured in infant cord blood and placental tissues, and primarily tested in European cohorts [@Joubert:2016ue; @Everson:2021td; @Cardenas2019].
-Prenatal maternal cigarette smoking exposure is associated with postnatal DNA methylation signatures in blood from cohorts of older children and adults [@Tehranifar:2018tb ; @Richmond:2018tu].
-Few studies have considered the persistence of the association between maternal prenatal cigarette smoking and DNA methylation at different ages in the same individuals.
-Additionally, little work has considered the portability of the association between maternal prenatal cigarette smoking and DNA methylation across tissues.
-DNA methylation drives cell differentiation and cell type proportions differ across tissue types.
-Epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples [@Everson:2021td].
-Cord blood and placental samples are unlikely to be available in clinical settings.
-While peripheral blood is a more realistic clinical sample, saliva is even easier to collect.
+DNA methylation is a possible biomarker.
+Prenatal maternal smoking is associated with DNA methylation signatures in placental, cord blood, and peripheral blood samples [@Joubert:2016ue; @Everson:2021td; @Cardenas2019; @Richmond:2018tu; @Tehranifar:2018tb].
+Cord blood and placental samples are rarely available outside of birth cohorts.
+While peripheral blood is a more realistic sample, saliva is easier to collect [] and increasingly used in large epidemiological cohorts [].
+Since DNA methylation drives cell differentiation and cell type proportions differ across tissue types, it is important to verify associations in new tissues.
 To our knowledge, no study has evaluated associations between maternal prenatal cigarette smoking and salivary DNA methylation.
 
-Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017], and DNA methylation signatures of own-smoking differ across genetic ancestry groups [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012].
-Only a few studies have examined associations between prenatal maternal cigarette smoking and DNA methylation in non-European ancestry populations.
-Among 954 infants, 70% of whom had mothers identifying as Black, prenatal maternal smoking was associated with cord blood DNA methylation at 38 CpG sites [@Xu2021].
-The direction of the associations between prenatal maternal cigarette smoking and DNA methylation at these sites were generally similar in children of Black and non-Black participants [@Xu2021].
-Among 572 3-5 year old children, 186 of whom were of African or admixed genetic ancestry, prenatal maternal smoking was associated with DNA methylation in peripheral blood samples but ancestry-stratified analyses were not performed [@Ladd-Acosta:2016aa].
-Similarly, among 89 middle-aged women, of whom reported African American or Hispanic ethnicity, prenatal maternal cigarette smoking was associated with DNA methylation at 17 of 190 tested CpG sites [@Tehranifar:2018tb].
-Adjusting for ancestry did not substantially affect the association between prenatal maternal smoking and DNA methylation at any of the sites, although the authors note that African American women had on average 2--5% higher mean DNA methylation on the absolute scale as compared with White and Hispanic women at 5 *CYP1A1* CpG sites [@Tehranifar:2018tb].
-Of 148 CpG sites selected based on their association with prenatal maternal smoking among primarily European children, 7 CpG sites were also associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue].
-While prenatal maternal cigarette smoking is associated with DNA methylation, the consistency of this association across genetic ancestry is understudied.
+Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017].
+Own-smoking DNA methylation signatures differ by genetic ancestry [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012; @Liu2019].
+Most work on prenatal maternal smoking and child DNA methylation has been conducted in European-ancestry cohorts [@Joubert:2016ue], with some notable exceptions [@Ladd-Acosta:2016aa; @Tehranifar:2018tb; @Zhang2018; @Gonseth2016; @Neophytou:2019ty].
+Some CpG sites associated with prenatal maternal smoking in European-ancestry cohorts have been replicated in non-European-ancestry or multi-ethnic populations [@Tehranifar:2018tb; @Neophytou:2019ty].
+Children of both Black and non-Black identifying mothers exhibited similar directions of association between prenatal maternal smoking and peripheral blood DNA methylation at 38 CpG sites [@Xu2021].
+However, this analysis stratified by self-reported ethnicity rather than genetic ancestry.
+Very studies have evaluated the portability of DNA methylation signatures of prenatal maternal smoke exposure across genetic ancestry populations within a single cohort [@Tehranifar:2018tb; @Gonseth2016].
 
-After association testing, an important next step is to evaluate DNA methylation as a biomarker of prenatal maternal smoke exposure by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior.
-Biomarkers can consist of single DNA methylation sites or summary meausres across multiple sites [@Philibert2020; @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa].
-One attractive option is a polymethylation score, analogous to a polygenic score.
-In the polymethylation score, CpG sites are weighted by the strength of their association with maternal smoking from a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a].
-One such DNA methylation score calculated from cord blood classified maternal smoking behavior with an area under the curve (AUC) of 0.82 in a primarily European cohort [@Reese:2017ts].
-Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking behavior was classified with an AUC of 0.87 [@Ladd-Acosta:2016aa].
-Among middle-aged adults, a score calculated using DNA methylation in peripheral blood samples from a single time point could even predict prenatal smoke exposure (30 years previously) with an AUC of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu].
-Ideally, a biomarker for prenatal maternal smoking would be consistent over participant age at DNA methylation sampling and portable across genetic ancestry groups.
-However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across tissue, age, and ancestry has not been evaluated in the same study.
+After association testing, DNA methylation-based biomarkers of prenatal maternal smoking can be assessed by evaluating their classification accuracy.
+Biomarkers range from single DNA methylation sites to multi-site summary measures, such as polymethylation scores [@Philibert2020 @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa].
+In a polymethylation score, CpG sites are weighted by their effect size from a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a].
+DNA polymethylation scores for prenatal maternal smoke exposure have performed well (area under the curves (AUCs)\>0.8) in newborns and young children [@Reese:2017ts; @Ladd-Acosta:2016aa] and fairly in middle-aged adults (AUC of 0.72 (95% confidence interval 0.69, 0.76)) [@Richmond:2018tu].
+DNA methylation signatures of own smoking change post-cessation [@keshawarz2021].
+Associations with prenatal maternal smoking change at some CpG sites as children age [@Richmond2015].
+Yet few studies have assessed the longitudinal persistence of DNA-methylation biomarker accuracy.
 
 DNA methylation differences hold promise as a biomarker for prenatal maternal smoking.
-Before translation to clinical applications, we must understand how and if the signal varies across age, ancestry, and tissue sample.
-In the Fragile Families and Child Wellbeing study, a diverse longitudinal cohort of children, we aimed to assess the potential of salivary DNA methylation biomarkers for prenatal maternal cigarette smoking.
-We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including epigenetic clocks, polymethylation scores, and individual *a priori* CpG sites.
-We then performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a persistent and consistent biomarker of prenatal maternal smoking.
+We must understand how and if the signal varies across tissues sample, genetic ancestry, and age.
+In the Fragile Families and Child Wellbeing study, a diverse longitudinal birth cohort, we aimed to assess salivary DNA methylation biomarkers for prenatal maternal cigarette smoking.
+We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including polymethylation scores for smoke exposure, individual a priori CpG sites, global DNA methylation, and epigenetic clocks, at ages 9 and 15.
+We then performed ancestry- and age-stratified analyses to test the hypothesis that DNA methylation could serve as a portable and persistent biomarker of prenatal maternal smoking.

### 2022-02-14 Papers/Polymethylation_scores/06_discussion.Rmd
revise intro and discussion

--- 2021-11-29 Papers/Polymethylation_scores/06_discussion.Rmd
@@
 
-In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed that prenatal maternal smoking was associated with several characterizations of DNA methylation in children's saliva samples from ages 9 and 15.
-Prenatal maternal smoking was associated with polymethylation scores across strata of both child age and genetic ancestry.
+In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed that prenatal maternal smoking was associated with several characterizations of DNA methylation in childrens saliva samples from ages 9 and 15.
+Prenatal maternal smoking was associated with polymethylation scores for prenatal smoke exposure across strata of genetic ancestry and age.
 Global methylation and epigenetic clocks were not associated with maternal smoking exposure.
-Polymethylation scores had reasonable accuracy for classifying prenatal maternal smoking (AUC \> 0.7).
-Classification when using polymethylation scores for prenatal maternal smoking was better than when using a single *a priori* CpG site, cg05575921 in the *AHRR* gene.
+Polymethylation scores for prenatal smoke exposure improved classification of self-reported prenatal maternal smoking over a base model, while a single *a priori* CpG site, *AHRR*:cg05575921, did not.
 
 Our findings are consistent with the previous literature on associations between prenatal maternal smoking and DNA methylation.
-We replicated the top hit from a previous epigenome wide association analysis [@Joubert:2016ue].
-Further, we found evidence of association in several additional hits from epigenome wide association studies of DNA methylation in cord and peripheral blood [@Joubert:2016ue; @wiklund2019].
+We replicated top hits from previous epigenome wide association analyses in cord and peripheral blood [@Joubert:2016ue; @wiklund2019].
 There are no previous studies of prenatal maternal smoking and saliva DNA methylation.
-However, previous work has shown that the majority of CpG sites are similarly methylated in blood and saliva (reviewed in @Langie2017).
+However, similar populations of immune cells are present in both blood and saliva, and saliva samples in our study were primarily composed of immune cells.
+Additionally, previous work has shown that the majority of CpG sites are similarly methylated in blood and saliva (reviewed in @Langie2017).
+Our results suggest salivary DNA methylation can be used as an alternative to peripheral blood DNA methylation for biomarkers of prenatal maternal smoking.
 
-We advance prenatal smoking - DNA methylation literature by evaluating the persistence of the association between prenatal maternal smoking and DNA methylation as children age and its portability across tissue and ancestry.
-Polymethylation scores built using coefficients from meta-analysis of cord blood DNA methylation from primarily European-ancestry newborns were still associated with prenatal maternal smoking in our independent saliva samples from a diverse cohort at ages 9 and 15 [@Joubert:2016ue].
-The portability of other risk scores, such as polygenic risk scores, across genetic ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Hüls2020].
-In this case, the polymethylation score for prenatal maternal smoking appears to be portable across genetic ancestry groups.
-Effect estimates were consistent across ancestries at age 9, although Hispanic and European genetic ancestry samples had lower, non-significant effect estimates at age 15.
-The decreased precision may be the result of the reduced sample size of the HIspanic and European genetic ancestry samples.
-The reduction in effect estimate magnitude could reflect higher unreported smoking initiation in European and Hispanic ancestry teens in the United States than African ancestry teens.
+We advance prenatal smoking - DNA methylation literature by evaluating the portability of association between prenatal maternal smoking and DNA methylation across ancestry groups.
+The portability of polygenic risk scores across genetic ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Huls2020].
+Our results suggest polymethylation scores for prenatal maternal smoking are portable across genetic ancestry groups.
+In the samples from age 15, admixed Latin and European ancestry samples had reduced effect estimates.
 Genetic ancestry correlates with race, and White and Latino teens have much higher rates of teen smoking and earlier ages at initiation than Black children [@El-Toukhy2016].
-Though we excluded children who reported own-smoking from our analytic sample, under-reporting of own-smoking could influence DNA methylation at ages 9 and 15, creating outcome misclassification.
-This could also explain the observed attenuation of the effect on *AHRR*:cg05575921 methylation by age 15 in our sample, as DNA methylation at this probe is known to vary by personal cigarette smoking[@Dawes2019; @Philibert2020].
+Though we excluded children who reported own-smoking from our analytic sample, differential initiation and reporting of own-smoking could result in more similar DNA methylation profiles of children exposed vs unexposed to prenatal maternal smoking.
 
-Our findings are also consistent with previous work on DNA methylation as a biomarker for prenatal maternal smoke exposure.
-Prenatal maternal smoking classification using saliva DNA methylation in our sample of 9 and 15 year olds performed comparably to classification using peripheral blood from older adults (AUC 0.72) [@Richmond:2018tu].
-However, previous prenatal maternal smoke exposure biomarkers created using LASSO regression and cord blood from newborns performed better (AUCs ranging from 0.82-0.97) [@Ladd-Acosta:2016aa] A support vector machine approach performed on and peripheral blood from 3-5 year old children also more accurately classified prenatal maternal smoke exposure (AUC=0.87) [@Reese:2017ts].
-Differences in DNA methylation patterns across tissues and over time may influence the performance of DNA methylation biomarkers.
-Different methods for summarizing across multiple DNA methylation sites may also influence biomarker performance.
-The similarity in performance between saliva DNA methylation in our study and peripheral blood DNA methylation in older adults is encouraging for the use of saliva DNA methylation as a readily accessible clinical sample.
+We contribute to the understanding of the impact of time-since-exposure on the classification accuracy of DNA methylation biomarkers for prenatal maternal smoking.
+Our saliva-based classification of 9- and 15-year-olds (AUC=0.78 and 0.77 respectively) was less accurate than previous cord-blood- and peripheral-blood-based classifications in newborns (AUCs ranging from 0.82-0.97) [@Reese:2017ts] and 3-5-year-old children (AUC=0.87) [@Ladd-Acosta:2016aa].
+The difference in classification accuracy could reflect lower proportion of immune cells of saliva as compared to blood.
+However, they could also reflect a diminishing capacity to distinguish exposure status as time since exposure increases and new environmental exposures are encountered.
+Indeed, our classification was better than peripheral-blood-based classification in 30-year-old adults (AUC=0.72) [@Richmond:2018tu].
+Notably, a previous longitudinal analysis of children's blood samples from birth to 17 years of age found that larger changes in DNA methylation occured between 0 and 7 years of age than between 7 and 17 years of age [@Richmond2015].
+In that cohort, the association between prenatal smoke exposure and DNA methylation changed between childhood and adolescence at some but not all CpG sites.
+This may explain why we found that the polymethylation score improved classification accuracy while *AHRR*:cgg05575921 methylation did not.
+While *AHRR*:cg05575921 hypomethylation is a consistent marker of prenatal maternal smoking in newborns and children, the association did not persist in samples from adult women [@Richmond:2018tu].
+Similarly, while *AHRR*:cg05575291 methylation measured in both blood and saliva can very accurately classify own-smoking [@Dawes2019; @Philibert2020], the magnitude of hypomethylation is reduced post-cessation[@keshawarz2021].
+As the accuracy of *AHRR*:cg05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures, incorporating information across multiple sites of DNA methylation may yield a a more robust biomarker.
 
-Our analysis contributes to the development of an accurate methylation biomarker for prenatal maternal smoking by evaluating the impact of specific methodological choices on accuracy of classification.
-DNA methylation biomarkers may be especially susceptible to confounding by subject age at methylation measurement and cell-type proportion [@Hüls2020].
-Prenatal maternal smoke exposure classification accuracy of polymethylation score was similar when using coefficients from cord blood vs peripheral blood samples from older children.
-Classification accuracy was not improved by using two time-points of methylation measurement.
-Classification accuracy was improved when polymethylation scores used coefficients which incorporated cell-type control.
-Our analysis therefore suggests that cell-type control when both generating (in the epigenome wide association study) and applying coefficients for polymethylation scores may positively influence outcome classification accuracy, although more work comparing coefficients from different populations is still needed.
-
-Our analysis suggests that polymethylation scores may be more accurate than using single CpG sites as biomarkers.
-The site cg05575921 in the *AHRR* gene is a consistent marker of prenatal maternal smoke exposure in meta-analyses of newborn cord blood[@Joubert:2016ue].
-DNA methylation at cg05575291 in the *AHRR* gene can accurately classify own--smoking behavior in both blood (area under the curve 0.995) and saliva (area under the curve 0.971) [@Dawes2019; @Philibert2020].
-However, salivary AHRR:cg05575921 was not a persistent marker of prenatal maternal smoking in an analysis of middle-aged adult women [@Richmond:2018tu].
-In our sample, salivary AHRR:CpG05575921 methylation was less accurate at predicting prenatal maternal smoke exposure than polymethylation scores.
-The accuracy of AHRR:CpG05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures.
-Incorporating information across multiple sites of DNA methylation may yield a biomarker more robust to these influences.
+We note that while the pediatric DNA methylation clock accurately recapitulated subjects chronological ages, the GRIM-age values in our population were overestimated by 10-15 years.
+In an analysis of 4 pediatric cohorts of children under 20 years of age, GRIM age estimates were 30-35 years greater than chronological in buccal cell samples and 10-15 years greater than chronological age in blood samples [@okazaki2021].
+GRIM-age was neither trained nor validated in a pediatric population [@lu2019], which may explain this overestimation.
 
@@
 We tested associations between prenatal maternal smoking and multiple DNA summary measures to evaluate the specificity of the polymethylation scores.
-In sensitivity analyses, we adjusted for other prenatal exposures and postnatal smoke exposure to examine the specificity of the biomarker to nature and timing of exposure.
+We adjusted for prenatal maternal drinking, other prenatal drug use and postnatal smoke exposure to examine the specificity of the biomarker to nature and timing of exposure.

### 2022-03-18 Papers/Polymethylation_scores/01_titlepage.Rmd
updated results post datafreeze update
colter revisions
post submission revisions

@@
 output:
+  pdf_document: default
+  word_document: default
   html_document: default
-  word_document: default
 bibliography: Methylation_SmokingBib.bib
-always_allow_html: true
+always_allow_html: yes
 ---
@@
 
-**Title:** Polymethylation Scores for Prenatal Maternal Smoke Exposure Persist Until Age 15 and Are Detected in Saliva
+**Title:** Polymethylation Scores for Prenatal Maternal Smoke Exposure Persist Until Age 15 and Are Detected in Saliva in the Fragile Families and Child Wellbeing Cohort
 
-**Authors:** Freida A. Blostein$^1$, Jonah Fisher$^2$,  John Dou$^1$, Lisa Schenper$^3$, Erin Ware$^2$ Daniel A. Notterman$^3$, Colter Mitchell$^2$, Kelly M. Bakulski$^1$`
+**Authors:** Freida A. Blostein$^1$, Jonah Fisher$^2$,  John Dou$^1$, Lisa Schenper$^3$, Erin Ware$^2$ Daniel A. Notterman$^{3*}$, Colter Mitchell$^{2*}$, Kelly M. Bakulski$^{1*+}$
 
@@
 $^3$Department of Molecular Biology, Princeton University, Princeton, New Jersey, United States of America
+
+*Co-senior authors
+
+$^+$Corresponding author

### 2022-03-18 Papers/Polymethylation_scores/02_abstract.Rmd
post submission revisions

--- 2021-11-10 Papers/Polymethylation_scores/02_abstract.Rmd
@@
+---
+output:
+  pdf_document: default
+  html_document: default
+---
+
 # Abstract
@@
 **Background**: Prenatal maternal smoking has negative implications for child health.
-DNA methylation signatures  can function as a biomarker of prenatal maternal smoking.
-However little work has assessed how DNA methylation signatures of prenatal maternal smoking vary across ages, ancestry groups, or tissues.
-In the Fragile Families and Child Wellbeing study, we tested whether prenatal maternal smoking was associated with salivary polymethylation scores.
-We assessed the consistency of associations at ages 9 and 15, their portability across  participants from African, European, and Hispanic genetic ancestries and the accuracy of exposure classification using Area Under the Curve (AUC) from receiver operating curve analyses.
+DNA methylation signatures can function as biomarkers of prenatal smoke exposure.
+However the robustness of prenatal maternal smoking DNA methylation signatures across child ages, genetic ancestry groups or tissues is not clear.
+Using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cord blood, we created polymethylation scores of saliva DNA methylation from children at ages 9 and 15 in the Fragile Families and Child Wellbeing study.
+We tested whether prenatal maternal smoking was associated with salivary polymethylation scores for prenatal smoke exposure and assessed the consistency of associations by age and participant genetic ancestry group.
+We tested the accuracy of exposure classification using area under the curve (AUC) from receiver operating curve analyses and compared to findings from candidate DNA methylation sites.
 
-**Results**: We created saliva polymethylation scores using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cordblood.
-In the full sample at age 9 (n=756), prenatal maternal smoke exposure was associated with a 0.52 (95%CI: 0.36, 0.67) standard deviation higher polymethylation score.
-The direction and magnitude of the association was consistent across when stratified genetic ancestries.
-In the full sample at age 15 (n=749), prenatal maternal smoke exposure was associated with a 0.46 (95%CI: 0.3, 0.62) standard deviation higher polymetylation score, and the association was attenuated among the European and Hispanic genetic ancestry samples.
-The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (AUC age 9=0.77, *P* value<0.001, age 15=0.77, *P* value<0.001), and more accurate than using a single *a priori* site in the AHRR gene (cg05575921 AUC=0.74, *P* value=0.03; age 15=0.73, *P* value=0.01).
+**Results**: In the full sample at age 9 (n=`r table(modeldata$childteen)[1]`), prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.
+The direction and magnitude of the association was consistent in European and African genetic ancestry samples.
+In the full sample at age 15 (n=`r table(modeldata$childteen)[2]`), prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score, and the association was attenuated among the European and Admixed - Latin genetic ancestry samples.
+The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (AUC age 9=0.77, age 15=0.77).
+Including the polymethylation score increased the AUC of base model covariates by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(6)` percentage points, while including a single candidate site in the *AHRR* gene did not (*P* val `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg055'))%>%pull(7)%>%round(.,2)`).
 
-**Conclusions**: Prenatal maternal smoking was associated with DNA methylation signatures in saliva samples, a clinically practical tissue.
+**Conclusions**: Prenatal maternal smoking was associated with DNA methylation signatures in child saliva samples, a clinically practical tissue.
 Polymethylation scores for prenatal maternal smoking were portable across genetic ancestries and more accurate than individual DNA methylation sites.

### 2022-03-18 Papers/Polymethylation_scores/04_background.Rmd
post submission revisions

@@
 latex_engine: xelatex
+csl: springer-vancouver.csl
 ---
 
-# Background
+# Introduction
 
-In 2018, 11% of pregnant women in the United States aged 15-44 reported any past-month cigarette use [@SAMHSA2018].
-Children exposed to cigarette smoking in utero are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu].
-Measuring prenatal maternal smoking behavior is challenging.
+In 2018, 11% of pregnant people in the United States aged 15-44 reported any past-month cigarette use [@SAMHSA2018].
+Children exposed to cigarette smoking *in utero* are more likely to have low-birth weight, negative neurodevelopmental outcomes, and asthma [@Banderali:2015vu].
+Measuring prenatal maternal smoking behavior can be challenging.
 Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua].
-The gold-standard biomarker, serum cotinine levels, have a half-life of nine hours in pregnant women, and are rarely available outisde of birth cohorts[@Dempsey:2002wr].
-Alternative biomarkers of prenatal maternal smoke exposure could improve research and clinical practice.
+The gold-standard smoking biomarker, serum cotinine levels, have a half-life of nine hours in pregnant women, and are rarely available outside of birth cohort research [@Dempsey:2002wr].
+Alternative biomarkers of prenatal maternal smoke exposure with a larger window of measurement could improve research and clinical practice.
 
 DNA methylation is a possible biomarker.
-Prenatal maternal smoking is associated with DNA methylation signatures in placental, cord blood, and peripheral blood samples [@Joubert:2016ue; @Everson:2021td; @Cardenas2019; @Richmond:2018tu; @Tehranifar:2018tb].
-Cord blood and placental samples are rarely available outside of birth cohorts.
-While peripheral blood is a more realistic sample, saliva is easier to collect [] and increasingly used in large epidemiological cohorts [].
-Since DNA methylation drives cell differentiation and cell type proportions differ across tissue types, it is important to verify associations in new tissues.
-To our knowledge, no study has evaluated associations between maternal prenatal cigarette smoking and salivary DNA methylation.
+Prenatal maternal smoking is associated with offspring DNA methylation signatures at birth in placental and cord blood samples, and postnatally in peripheral blood samples [@Joubert:2016ue; @Everson:2021td; @Cardenas2019; @Richmond:2018tu; @Tehranifar:2018tb].
+Another postnatal biospecimen option is saliva, which is easier to collect than blood and increasingly used in large epidemiological cohorts [@Langie2017].
+Since DNA methylation drives cell differentiation and cell type proportions differ across tissue types, it is important to verify associations in new tissues [@Bakulski2016].
+In addition, most studies of prenatal maternal smoking and childhood DNA methylation have been conducted in primarily European genetic ancestry populations [@Joubert:2016ue; @Ladd-Acosta:2016aa; @Tehranifar:2018tb; @Zhang2018; @Gonseth2016; @Neophytou:2019ty], and it will be important to test generalizability to other groups.
+To our knowledge, few studies have assessed multiple genetic ancestry groups in the same cohort and no study has evaluated associations between maternal pregnancy cigarette smoking and salivary DNA methylation in children.
 
-Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017].
-Own-smoking DNA methylation signatures differ by genetic ancestry [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012; @Liu2019].
-Most work on prenatal maternal smoking and child DNA methylation has been conducted in European-ancestry cohorts [@Joubert:2016ue], with some notable exceptions [@Ladd-Acosta:2016aa; @Tehranifar:2018tb; @Zhang2018; @Gonseth2016; @Neophytou:2019ty].
-Some CpG sites associated with prenatal maternal smoking in European-ancestry cohorts have been replicated in non-European-ancestry or multi-ethnic populations [@Tehranifar:2018tb; @Neophytou:2019ty].
-Children of both Black and non-Black identifying mothers exhibited similar directions of association between prenatal maternal smoking and peripheral blood DNA methylation at 38 CpG sites [@Xu2021].
-However, this analysis stratified by self-reported ethnicity rather than genetic ancestry.
-Very studies have evaluated the portability of DNA methylation signatures of prenatal maternal smoke exposure across genetic ancestry populations within a single cohort [@Tehranifar:2018tb; @Gonseth2016].
-
-After association testing, DNA methylation-based biomarkers of prenatal maternal smoking can be assessed by evaluating their classification accuracy.
-Biomarkers range from single DNA methylation sites to multi-site summary measures, such as polymethylation scores [@Philibert2020 @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa].
-In a polymethylation score, CpG sites are weighted by their effect size from a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a].
-DNA polymethylation scores for prenatal maternal smoke exposure have performed well (area under the curves (AUCs)\>0.8) in newborns and young children [@Reese:2017ts; @Ladd-Acosta:2016aa] and fairly in middle-aged adults (AUC of 0.72 (95% confidence interval 0.69, 0.76)) [@Richmond:2018tu].
-DNA methylation signatures of own smoking change post-cessation [@keshawarz2021].
-Associations with prenatal maternal smoking change at some CpG sites as children age [@Richmond2015].
+Using the findings from large-scale association testing, in independent populations, DNA methylation-based biomarkers of prenatal maternal smoking can be developed and assessed by evaluating their classification accuracy.
+Biomarkers can range from single DNA methylation sites to summary measures across multiple sites, termed polymethylation scores [@Philibert2020 @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa].
+In a polymethylation score, DNA methylation sites are weighted by their effect size from a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a].
+DNA polymethylation scores for prenatal maternal smoke exposure have performed well (area under the curves (AUCs)\>0.8) in cord blood samples from newborns and peripheral blood samples from young children [@Reese:2017ts; @Ladd-Acosta:2016aa] and fairly well in peripheral blood from middle-aged adults (AUC of 0.72 (95% confidence interval 0.69, 0.76)) [@Richmond:2018tu].
+Associations with prenatal maternal smoking change at some DNA methylation sites as children age [@Richmond2015].
 Yet few studies have assessed the longitudinal persistence of DNA-methylation biomarker accuracy.
 
-DNA methylation differences hold promise as a biomarker for prenatal maternal smoking.
-We must understand how and if the signal varies across tissues sample, genetic ancestry, and age.
-In the Fragile Families and Child Wellbeing study, a diverse longitudinal birth cohort, we aimed to assess salivary DNA methylation biomarkers for prenatal maternal cigarette smoking.
-We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including polymethylation scores for smoke exposure, individual a priori CpG sites, global DNA methylation, and epigenetic clocks, at ages 9 and 15.
-We then performed ancestry- and age-stratified analyses to test the hypothesis that DNA methylation could serve as a portable and persistent biomarker of prenatal maternal smoking.
+DNA methylation differences hold promise as a biomarker for prenatal maternal smoke exposure.
+We must understand how the signal varies across tissue, genetic ancestry, and age.
+In the Fragile Families and Child Wellbeing Study, a diverse longitudinal birth cohort, we aimed to assess salivary DNA methylation biomarkers for prenatal maternal cigarette smoking.
+We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including polymethylation scores for smoke exposure, individual a priori DNA methylation sites, global DNA methylation, and epigenetic clocks, at ages 9 and 15.
+We then performed genetic ancestry- and age-stratified analyses to test the hypothesis that DNA methylation could serve as a portable and persistent biomarker of prenatal maternal smoking.

### 2022-03-18 Papers/Polymethylation_scores/05_results.Rmd
post submission revisions

@@
 output:
-  word_document: default
+  word_document: 
+    reference_docx: "../Template.docx"
   html_document: default
@@
 ---
-
-```{r, echo=F, include=FALSE}
-library(knitr)
-knitr::knit_child(here::here("Reports", "Paper1.Rmd"))
-```
 
@@
 }
+
+coefOnly=function(df, filter_statement, my_term, si){
+  df%>%
+    filter(eval(parse(text = filter_statement)))%>%
+    unnest(tidied)%>%
+    mutate(val=round(estimate, si))%>%
+    filter(term==my_term)%>%
+    pull(val)
+}
+
+auc_ci=function(df, filter_statement){
+  df=df %>%filter(eval(parse(text = filter_statement)))
+  ci=paste0(round(df$auc, 2), ' (', 
+            round(df$ci_auc_delong[[1]][1], 2), 
+            ', ', 
+            round(df$ci_auc_delong[[1]][3], 2)
+            , ')')
+  return(ci)
+}
+
+my_p_val=function(p){
+  if(p<0.001){return_p='<0.001'}
+  if(p>=0.001){return_p=paste0('=', round(p, 3))}
+  return(return_p)
+}
 ```
@@
 
-In a sample of the Fragile Families and Child Wellbeing Study, saliva DNA methylation was measured on `r methylCohort$MethID%>%length()` samples from `r methylCohort$idnum%>%unique()%>%length()` unique participants with the Illumina 450K  array.
+In a sample of the Fragile Families and Child Wellbeing Study, saliva DNA methylation was measured on `r m_og_methyl` samples from `r n_og_methyl` unique child participants with the Illumina 450K array.
 Complete data on covariates of interest was available on `r m_obs` samples from `r n_total` participants who were included in the analysis (Figure 1).
 There were `r n_twovisit` participants with both age 9 and age 15 DNA methylation samples.
-Excluded samples were similar to included samples, except that included samples were slightly more likely to be from children of African genetic ancestry and less likely to be from children of Hispanic genetic ancestry (Supplemental Table 1).
+Excluded samples were similar to included samples, except that included samples were slightly more likely to be from children of African genetic ancestry and less likely to be from children of Admixed Latin genetic ancestry (Supplemental Table 1).
 In the included sample, 20% percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alcohol use and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use.
 The mean income to poverty ratio of mothers at birth was `r round(mean(modeldata_wide$cm1inpov), 2)`.
-Of the children in the analytic sample, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African genetic ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Hispanic ancestry']`% were of Hispanic genetic ancestry.
+Of the children in the analytic sample, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African genetic ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Admixed ancestry - Latin heritage']`% were of Admixed Latin genetic ancestry.
 
-We calculated several summary measures of DNA methylation, including polymethylation scores for smoke exposure, global DNA methylation,epigenetic clocks, and estimated cell-type proportions.
-Many of the methylation summary measures were correlated with each other (Supplemental Figure 1).
-Our polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $\rho$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value, 3)`).
+ All of the polymethylation scores for prenatal smoke exposure were correlated (Supplemental Figure 1).
+Our primary polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $\rho$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value `r my_p_val(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value)`).
 
 We compared epigenetic measures between the age 9 and age 15 visit among the `r n_twovisit` individuals with data from both visits (Supplemental Figure 2).
-The correlation across ages was strongest for the polymethylation score for prenatal smoke exposure (Pearson $\rho$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value<0.0001).
-Global DNA methylation (`r round(cor(model_twovisit$globalmethylation_Age_9, model_twovisit$globalmethylation_Age_15), 2)`, *P* value<0.0001) and epigenetic clocks were less strongly correlated across ages (Pediatric: `r round(cor(model_twovisit$pediatric_Age_9, model_twovisit$pediatric_Age_15), 2)`; *P* value<0.0001, GRIM: `r round(cor(model_twovisit$grim_Age_9, model_twovisit$grim_Age_15), 2)`; *P* value<0.0001) The distribution of epigenetic ages shifted higher between the age 9 and age 15 visit, as expected (Supplemental Figure 3).
-The pediatric clock more closely reflected chronological age than the GRIM clock.
-Among age 9 samples, the mean estimated age from the pediatric clock was `r round(mean(modeldata_wide$pediatric_Age_9, na.rm=T))`  and the mean estimated age from the GRIM clock was `r round(mean(modeldata_wide$grim_Age_9, na.rm=T))`.
-Among age 15 samples, the mean estimated age from the pediatric clock was `r round(mean(modeldata_wide$pediatric_Age_15, na.rm=T))` and the mean estimated age from the GRIM clock was `r round(mean(modeldata_wide$grim_Age_15, na.rm=T))`.
-The distribution of estimated cell-type proportions also shifted between visits while the distribution of global methylation and polymethylation scores were more consistent (Supplemental Figure 3).
+The correlation across ages was strongest for the polymethylation scores for prenatal smoke exposure constructed using the Joubert meta-analysis coefficients from newborn cord blood (6074 site score with coefficients from regression with cell-type control Pearson $\rho$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value\<0.0001; 568 site score $\rho$=`r round(cor(model_twovisit$richmond568_center_Age_9, model_twovisit$richmond568_center_Age_15), 2)`, *P* value\<0.0001).
+Global DNA methylation (`r round(cor(model_twovisit$globalmethylation_Age_9, model_twovisit$globalmethylation_Age_15), 2)`, *P* value\<0.0001) and epigenetic clocks were less strongly correlated across ages (Pediatric: `r round(cor(model_twovisit$PedBE_Age_9, model_twovisit$PedBE_Age_15), 2)`; *P* value\<0.0001, GRIM: `r round(cor(model_twovisit$DNAmGrimAge_Age_9, model_twovisit$DNAmGrimAge_Age_15), 2)`; *P* value\<0.0001) The distribution of epigenetic ages shifted higher between the age 9 and age 15 visit, as expected (Supplemental Figure 3).The distribution of estimated cell-type proportions also shifted between visits while the distribution of global methylation and polymethylation scores were more consistent (Supplemental Figure 3).
 Almost all saliva samples consisted of mostly immune cells based on estimated cell-type proportions.
@@
 
-In bivariate analyses and multivariable models, we focused on the polymethylation score for prenatal smoke exposure and the single top CpG site from prior research, cg05575921 in the *AHRR* gene, as hypothesized biomarkers, and used global methylation and the pediatric clock as negative controls.
-
 ## Bivariate associations between prenatal maternal smoking and DNA methylation summary measures
 
-Mothers who reported smoking during pregnancy had lower income to poverty ratios (`r modeldata_people%>%filter(smkPreg_binary=='Yes')%>%summarise(mean(cm1inpov))%>%round(2)`) than those who did not ( `r modeldata_people%>%filter(smkPreg_binary=='No')%>%summarise(mean(cm1inpov))%>%round(2)`).
+Mothers who reported smoking during pregnancy had lower income to poverty ratios (mean=`r modeldata_people%>%filter(smkPreg_binary=='Yes')%>%summarise(mean(cm1inpov))%>%round(2)`) than those who did not (mean=`r modeldata_people%>%filter(smkPreg_binary=='No')%>%summarise(mean(cm1inpov))%>%round(2)`).
 Mothers who smoked were more likely to report prenatal alcohol use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%), prenatal drug use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%) and postnatal smoking (`r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[1, 2]%>%round(2)*100`%) than mothers who did not report prenatal maternal smoking (Table 1).
-Children of mothers who reported smoking during pregnancy were more likely to be of European genetic ancestry (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[2, 2], 2)*100`%) than children of mothers who did not (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[1, 2], 2)*100`%, *P* value=<0.001).
-Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both the age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) visits (Table 1, Supplemental Table 2).
+Children of mothers who reported smoking during pregnancy were more likely to be of European genetic ancestry (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[2, 2], 2)*100`%) than children of mothers who did not (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[1, 2], 2)*100`%, *P* value=\<0.001).
+Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both the age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value\<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value\<0.001) visits (Table 1, Supplemental Table 2).
 At age 9, children exposed to prenatal maternal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`%, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
-At age 15, children exposed to prenatal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers, although this difference was not significant (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`% *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
-Epigenetic age from the pediatric and GRIM clocks and global DNA methylation did not differ between children exposed vs unexposed to prenatal maternal smoking.
+At age 15, children exposed to prenatal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`% *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
+Epigenetic age from the pediatric and GRIM clocks and global DNA methylation did not differ between children exposed vs unexposed to prenatal maternal smoking (Figure 2).
 
@@
 
-The association between prenatal maternal smoking and the polymethylation score for prenatal smoke exposure was also observed in multivariable models, adjusting for base model covariates of child sex, maternal income to poverty ratio at baseline, proportion of salivary immune cells, sample plate from DNA methylation analysis, child age and the first two principal components of genetic ancestry  (Figure 3; Supplemental Table 2).
-At age 9, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score for prenatal smoke exposure.
-At age 15, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher prenatal smoke exposure polymethylation score for prenatal smoke exposure.
+The association between prenatal maternal smoking and the polymethylation score for prenatal smoke exposure was also observed in multivariable models, adjusting for base model covariates of city indicator, child sex, maternal income to poverty ratio at baseline, proportion of salivary immune cells, sample plate from DNA methylation analysis, child age and the first two principal components of genetic ancestry (Figure 3; Supplemental Table 3).
+At age 9, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score for prenatal smoke exposure.
+At age 15, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher prenatal smoke exposure polymethylation score for prenatal smoke exposure.
+
 A consistent association was observed when stratifying by genetic ancestry.
-In the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`) at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score for prenatal smoke exposure.
-The direction and magnitude of the association was similar in the European and Hispanic genetic ancestry samples at age 9, though the effect size was attenuated and no longer statistically significant at age 15 (Figure 3; Supplemental Tables 2).
+In the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`) at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score for prenatal smoke exposure.
+The effect was similar among European ancestry children at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`) and consistent in direction but attenuated at age 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`).
+The direction of the association was consistent in direction but attenuated among Admixed - Latin ancestry children at ages 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`) and 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`) (Figure 3; Supplemental Table 3).
 
@@
 Similarly, at age 15 prenatal maternal smoking was associated with `r -1*round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95%CI: (`r paste(round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`)).
-Prenatal maternal smoking remained significantly associated with a decrease in cg05575921 DNA methylation in the African genetic ancestry sample at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`) although the association was not significant at age 15  (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`; Figure 3).
+Prenatal maternal smoking remained significantly associated with a decrease in cg05575921 DNA methylation in the African genetic ancestry sample at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`) and age 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`; Figure 3).
 In the European genetic ancestry sample, the association was consistent in direction at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`) and age 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`).
-In the Hispanic genetic ancestry sample, the association was no longer significant and not consistent in direction (Age 9: `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Hispanic ancestry'", 'smkPreg_binaryYes', 2)`; Age 15: `r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Hispanic ancestry'", 'smkPreg_binaryYes', 2)`) (Figure 3).
+In the Admixed Latin genetic ancestry sample, the association was no longer significant and not consistent in direction (Age 9: `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`; Age 15: `r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`) (Figure 3).
 
@@
 
-## Sensitivity analyses for association testing
+## Sensitivity analysis: Specificity of DNA methylation signal to type and time of exposure
 
-After additionally adjusting for other prenatal exposures and postnatal smoke exposure, the association between prenatal maternal smoking and the polymethylation score was robust (Supplemental Figure 4).
-Similarly, after adjusting for other prenatal exposures and postnatal smoke exposure, the direction of the association between prenatal maternal smoking and cg055975921 was consistent, although no longer significant at age 15.
-Very few children who were exposed to maternal smoking prenatally were unexposed to postnatal smoke (Supplemental Figure 5).
-However children exposed only to postnatal smoke did not have higher polymethylation scores than children unexposed to both postnatal and prenatal smoke (Supplemental Figure 5).
-Results were also similar when controlling for surrogate variables instead of known covariates (Supplemental Figure 6; Supplemental Table 2).
+To assess the specificity of the DNA methylation signal to prenatal maternal cigarette smoking, we additionally adjusted for any other prenatal maternal drug use and prenatal maternal alcohol use.
+The association between prenatal maternal smoking and the polymethylation score was robust to this adjustment.
+The association between prenatal maternal smoking and cg05575921 was slightly attenuated (Supplemental Figure 4, Supplemental Table 2).
 
-Results from linear mixed effect models were similar to age-stratified and base covariate adjusted linear models (Supplemental Table 3 & 4).
+To assess the specificity of the DNA methylation signal to the timing of cigarette smoke exposure, we additionally adjusted for exposure to postnatal maternal smoking.
+Very few children exposed to maternal smoking prenatally were unexposed postnatally (Supplemental Figure 5).In bivariate analyses, children exposed to only postnatal smoking had lower polymethylation scores than those exposed to pre- and postnatal smoking.
+However, they did not have higher polymethylation scores than children unexposed to both pre-and postnatal smoking (Supplemental Figure 5).
+In multivariable analyses, prenatal maternal smoking was still associated with the polymethylation scores when adjusting for postnatal smoke exposure, though effect estimates were attenuated (Age 9: `r coefCI(global_models, "childteen=='Age 9' & predictors=='secondhand smoke exposure model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` , Age 15: `r coefCI(global_models, "childteen=='Age 15' & predictors=='secondhand smoke exposure model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)`).
+Prenatal maternal smoking was no longer associated with cg05575921 methylation after adjusting for postnatal smoke exposure (Age 9: `r coefCI(global_models, "childteen=='Age 9' & predictors=='secondhand smoke exposure model' & outcome=='AHRR: cg05575921'", 'smkPreg_binaryYes', 2)`, Age 15: `r coefCI(global_models, "childteen=='Age 15' & predictors=='secondhand smoke exposure model' & outcome=='AHRR: cg05575921'", 'smkPreg_binaryYes', 2)`).
 
-Results were also similar when using polymethylation scores constructed with alternative regression coefficients (see Methods, Supplemental Figure 7).
-For our main analysis we used coefficients from a regression of sustained smoking exposure and DNA methylation in newborn cord blood with cell-type control.
-As sensitivity analyses, we used coefficients from regressions of: sustained smoking exposure and DNA methylation in newborn cord blood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children without cell-type control, and any smoking exposure and DNA methylation in newborn cord blood without cell-type control [@Joubert:2016ue].
-The association between prenatal maternal smoking and the polymethylation score using regression coefficients from newborn cord blood with cell-type controls was the strongest.
+Effect estimates were also similar when controlling for surrogate variables instead of known covariates (Supplemental Figures 3 and 6; Supplemental Table 3).
 
-In addition to cg055975921 in the *AHHR* gene, we tested the association of four other *a priori* probe sites identified from previous meta-analyses.
-We replicated the association of prenatal maternal smoking with these CpG sites (Supplemental Table 2, Supplemental Figure 8).
+## Sensitivity analysis: Choice of coefficients for constructing polymethylation score and associations with other *a priori* DNA methylation sites
 
-## Accuracy of polymethylation scores as a biomarker of prenatal maternal smoking
+Results were also similar when using sensitivity polymethylation scores (see Methods, Supplemental Figure 7) [@rauschert2020; @Richmond:2018tu].
 
-Next, we compared the accuracy of different DNA methylation summary measures for classifying prenatal maternal smoking using receiver operating curves (Figure 4A; Supplemental Table 5).
-We estimated classification of prenatal maternal smoking when using DNA methylation summary measures alone.
-We also estimated prenatal maternal smoking classification when using DNA methylation summary measures in addition to a base model covariates (child sex, maternal income to poverty ratio at baseline, child age at DNA methylation measurement, estimated immune cell proportion, plate from methylation processing and the first two genetic principal components).
-When used without DNA methylation measures, these base model variables had an area under the curve (AUC) of `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 9')%>%pull(auc)` at age 9 and `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 15')%>%pull(auc)` at age 15.
+In addition to cg055975921 in the *AHHR* gene, we tested the association of four other *a priori* DNA methylation sites identified from previous meta-analyses.
+We replicated the association of prenatal maternal smoking with these DNA methylation at GFI1:cg14179389, CYP1A1: cg05549655, MYO1G: cg22132788, and MYO1G:cg04180046  (Supplemental Figure 8).
 
-At age 9, including the polymethylation score for prenatal smoke exposure significantly improved the base model covariates classification (AUC: `r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(auc)`, *P* value comparing to base model<0.001).
-At age 15, similarly, including the polymethylation score for prenatal smoke exposure significantly improved the base model covariates classification (AUC:`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(auc)` , *P* value<0.001).
+## Polymethylation scores improve classification of prenatal smoke exposure while *AHRR*: cg05575921 alone does not
 
-At age 9, the base model covariates with the polymethylation score for prenatal smoke exposure had a larger AUC than the base model covariates with cg05575921 (AUC base model covariates + cg05575921: `r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(auc)`; *P* value =`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(5)`).
-At age 15, this was also true (AUC for base model covariates + cg05575921=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(auc)`,  *P* value=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(5)`; Supplemental Table 6).
-Accurate classification was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B).
+Next, we assessed whether including DNA methylation summary measures increased the accuracy of classifying prenatal smoke exposure over a base model consisting of child sex, maternal income to poverty ratio at baseline, child age at DNA methylation measurement, estimated immune cell proportion, plate from methylation processing, the first two genetic principal components and an indicator variable for oversampled city.
+At age 9, including the polymethylation score increased the AUC by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)`) percentage points over the base model (*P* value from DeLong test=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(7) %>%format(., scientific=F)`).
+At age 15 including the polymethylation score for prenatal smoke exposure increased the AUC by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)`) percentage points (*P* value from DeLong test=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(7)%>%format(., scientific=F)`).
+Including DNA methylation at *AHRR*:cg0557591 with base model covariates did not improve the classification compared to using base model covariates alone  (Supplemental Tables 4 and 5).
+Classification accuracy was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B).
 
-## Sensitivity analyses for biomarker accuracy assessment
+## Sensitivity analysis: Impact of choice of coefficients on polymethylation score accuracy and comparison to AUCs reported in the literature
 
-Classification of prenatal maternal smoke exposure when using other coefficients as the weights in construction of the polymethylation scores was similar to the results when using coefficients from cell-type controlled regressions of sustained prenatal maternal smoking and DNA methylation in cord blood (Supplemental Table 7).
+The AUC for classification of prenatal maternal smoke exposure in Fragile Families was relatively similar regardless of which polymethylation score was used (Figure 5, Supplemental Table 5).The highest AUC was from the 568 site score using coefficients from the regression of prenatal smoke exposure and DNA methylation in newborn cord blood.
+Performance was similar to previous reports.
+Neither the 28-site score from a LASSO regression and the 204 site score from an elastic net regression performed as well in Fragile Families as has been reported previously in the literature (Figure 5).

### 2022-03-18 Papers/Polymethylation_scores/06_discussion.Rmd
post submission revisions

@@
 output:
-  word_document: default
+  word_document: 
+    reference_docx: "../Template.docx"
   html_document: default
 bibliography: Methylation_SmokingBib.bib
+csl: springer-vancouver.csl
 ---
@@
 
-In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed that prenatal maternal smoking was associated with several characterizations of DNA methylation in childrens saliva samples from ages 9 and 15.
-Prenatal maternal smoking was associated with polymethylation scores for prenatal smoke exposure across strata of genetic ancestry and age.
-Global methylation and epigenetic clocks were not associated with maternal smoking exposure.
-Polymethylation scores for prenatal smoke exposure improved classification of self-reported prenatal maternal smoking over a base model, while a single *a priori* CpG site, *AHRR*:cg05575921, did not.
+In the longitudinal Fragile Families and Child Wellbeing birth cohort, prenatal maternal smoking was associated with saliva DNA methylation in children at ages 9 and 15.
+Children with prenatal smoke exposure had `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation scores at age 9 than children not exposed.
+These findings were robust across strata of genetic ancestry and children's age.
+These signatures were specific, as global DNA methylation and epigenetic clocks were not associated with prenatal smoke exposure.
+When assessing biomarker utility, the polymethylation score for prenatal smoke exposure improved the classification of self reported prenatal maternal smoking over a base model by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'polymethylation score'))%>%pull(5)` percentage points at age 9.
+A single *a priori* DNA methylation site, *AHRR*:cg05575921, did not (percentage point increase `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg05575921'))%>%pull(5)`).
+Prenatal exposure to cigarette smoke can be measured in saliva as an alternative to peripheral blood, although classification may be further improved by tissue-specific biomarkers.
 
-Our findings are consistent with the previous literature on associations between prenatal maternal smoking and DNA methylation.
-We replicated top hits from previous epigenome wide association analyses in cord and peripheral blood [@Joubert:2016ue; @wiklund2019].
+Our findings are consistent with the previous literature in different tissues and at different ages on associations between prenatal maternal smoking and DNA methylation.
+We replicated top hits from the seminal epigenome wide association meta-analysis of prenatal smoke exposure and DNA methylation of cord blood from newborns including *AHRR*:cg05575921, *GFl1*:cg14179389 and *MYO1G1*:cg22132788.
+Consistent with a meta-analysis of peripheral blood from children and adults over 16 years of age, the association at *MYO1G1*:cg22132788 was larger than in newborn cord blood [@Joubert:2016ue; @wiklund2019].
+Again consistent with a meta-analysis of peripheral blood from children and adults over 16 years of age, the associations at *AHRR*:cg05575921 and *GFl1*:cg14179389 were smaller than in newborn cord blood [@Joubert:2016ue; @wiklund2019].
 There are no previous studies of prenatal maternal smoking and saliva DNA methylation.
-However, similar populations of immune cells are present in both blood and saliva, and saliva samples in our study were primarily composed of immune cells.
-Additionally, previous work has shown that the majority of CpG sites are similarly methylated in blood and saliva (reviewed in @Langie2017).
+However, immune cell types are present in both blood and saliva, and saliva samples in our study were primarily composed of immune cells.
+Additionally, fewer than 4% of DNA methylation sites in blood and saliva samples displayed significantly different methylation values ($|\Delta \beta| >0.2$ and $p_{adj}< 0.001$) when measured on the 450K chip in samples of adults [@Langie2016] and 11-year old children [@Langie2017].
 Our results suggest salivary DNA methylation can be used as an alternative to peripheral blood DNA methylation for biomarkers of prenatal maternal smoking.
+A saliva-based DNA methylation biomarker may be particularly useful in cases where blood collection, processing, and storage is not possible.
+Saliva can often be collected directly by participants and saliva sample collection kits have stabilizers allowing room temperature storage prior to DNA methylation measurement.
 
-We advance prenatal smoking - DNA methylation literature by evaluating the portability of association between prenatal maternal smoking and DNA methylation across ancestry groups.
+We advance the prenatal smoking - DNA methylation literature by evaluating the generalizability of association between prenatal maternal smoking and DNA methylation across genetic ancestry groups.
 The portability of polygenic risk scores across genetic ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Huls2020].
 Our results suggest polymethylation scores for prenatal maternal smoking are portable across genetic ancestry groups.
-In the samples from age 15, admixed Latin and European ancestry samples had reduced effect estimates.
-Genetic ancestry correlates with race, and White and Latino teens have much higher rates of teen smoking and earlier ages at initiation than Black children [@El-Toukhy2016].
-Though we excluded children who reported own-smoking from our analytic sample, differential initiation and reporting of own-smoking could result in more similar DNA methylation profiles of children exposed vs unexposed to prenatal maternal smoking.
+At age 9, the effect estimates for prenatal maternal smoking on the polymethylation score in the  European and African genetic ancestry samples were was within `r abs(coefOnly(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)-coefOnly(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2))` standard deviation and were both significant at an alpha of 0.05.
+This finding is in line with a recent epigenome-wide association study in which children of both black and non-Black identifying mothers exhibited similar directions of associations between prenatal smoke exposure and peripheral blood methylation at 38 DNA methylation sites [@Xu2021].
+In our analysis, the association between prenatal smoke exposure and polymethylation scores was attenuated in children of admixed genetic ancestry (primarily of Hispanic ethnicity) at both ages and in children of European genetic ancestry at age 15.
+This could reflect differences in DNA methylation by genetic sequence at prenatal smoke sensitive DNA methylation sites [@Gonseth2016].
+While genetic ancestry is distinct from racial and ethnic social and cultural constructs [@borrell2021], it can correlate with different social and environmental exposures.
+Differences in other social and environmental exposures could also contribute to the attenuation observed in the admixed genetic ancestry group.
 
-We contribute to the understanding of the impact of time-since-exposure on the classification accuracy of DNA methylation biomarkers for prenatal maternal smoking.
-Our saliva-based classification of 9- and 15-year-olds (AUC=0.78 and 0.77 respectively) was less accurate than previous cord-blood- and peripheral-blood-based classifications in newborns (AUCs ranging from 0.82-0.97) [@Reese:2017ts] and 3-5-year-old children (AUC=0.87) [@Ladd-Acosta:2016aa].
-The difference in classification accuracy could reflect lower proportion of immune cells of saliva as compared to blood.
-However, they could also reflect a diminishing capacity to distinguish exposure status as time since exposure increases and new environmental exposures are encountered.
-Indeed, our classification was better than peripheral-blood-based classification in 30-year-old adults (AUC=0.72) [@Richmond:2018tu].
-Notably, a previous longitudinal analysis of children's blood samples from birth to 17 years of age found that larger changes in DNA methylation occured between 0 and 7 years of age than between 7 and 17 years of age [@Richmond2015].
-In that cohort, the association between prenatal smoke exposure and DNA methylation changed between childhood and adolescence at some but not all CpG sites.
-This may explain why we found that the polymethylation score improved classification accuracy while *AHRR*:cgg05575921 methylation did not.
-While *AHRR*:cg05575921 hypomethylation is a consistent marker of prenatal maternal smoking in newborns and children, the association did not persist in samples from adult women [@Richmond:2018tu].
-Similarly, while *AHRR*:cg05575291 methylation measured in both blood and saliva can very accurately classify own-smoking [@Dawes2019; @Philibert2020], the magnitude of hypomethylation is reduced post-cessation[@keshawarz2021].
+We contribute to the understanding of the impact of tissue on classification accuracy of DNA methylation biomarkers for prenatal maternal smoking.
+Polymethylation scores using coefficients from the Joubert meta-analysis coefficients had similar AUCs in Fragile Families saliva samples (AUCs between `r roc_table%>%ungroup()%>%filter(str_detect(methylation, 'CT|568|19'))%>%dplyr::select(auc)%>%range()%>%round(., 2)%>%paste(.
+,collapse=' and ')`) and in previous applications in peripheral blood samples from 17-, 30- and 31-year olds (AUCs between 0.69 and 0.72) [@Richmond:2018tu; @rauschert2020].
+Thus, using polymethylation scores from saliva may be an acceptable alternative when peripheral blood samples are unavailable or difficult to obtain.
+However, other polymethylation scores for prenatal maternal smoking have performed better in peripheral blood samples.
+The greatest classification accuracy of prenatal smoke exposure from previous studies was with a 28-site LASSO regression score trained in and then applied to a Norwegian cord blood cohort (AUC=0.90) [@Reese:2017ts].
+This score also accurately classified prenatal smoking from peripheral blood from adolescents and adults (AUCs between 0.85 and 0.73) [@rauschert2020].
+A 204-site elastic net score trained and then applied to peripheral blood from 17-year-old's peripheral blood samples also performed very well (AUC=0.87) [@rauschert2020].
+This score also performed well when applied to additional validation cohorts of peripheral blood from adolescents and adults (AUCs between 0.8 and 0.87).
+However, neither of these scores performed as well when applied in the Fragile Families cohort.
+The difference in classification accuracy could reflect cell proportion differences across tissues.
+Using coefficients from an epigenome-wide association analysis of saliva could improve classification and is a direction for future research.
+
+While polymethylation scores significantly improved classification of prenatal smoke exposure over a base model, using *AHRR*:cgg05575921 methylation as a classifier did not (percentage point increase in AUC: `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'AHRR'))%>%pull(5)`) .
+Prenatal smoke exposed newborns consistently exhibit lower DNA methylation at *AHRR*:cg05575291 than unexposed newborns [@Joubert:2016ue].
+However, the association between prenatal smoke exposure and *AHRR*:cg05575291 methylation did not persist in peripheral blood samples from adult women [@Richmond:2018tu].
+Additionally, a previous longitudinal analysis of children's blood samples from birth to 17 years of age in the ALSPAC cohort found that the difference in DNA methylation at *AHRR*:cg05575291 between prenatal smoke exposed and unexposed children decreased from birth to 17 years of age.
+This previous analysis also reported that differences in DNA methylation at *AHRR*:cg05575291 in adolescence may be partially mediated by postnatal own smoking, and we excluded children who reported postnatal own-smoking [@Richmond:2018tu].
 As the accuracy of *AHRR*:cg05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures, incorporating information across multiple sites of DNA methylation may yield a a more robust biomarker.
 
-We note that while the pediatric DNA methylation clock accurately recapitulated subjects chronological ages, the GRIM-age values in our population were overestimated by 10-15 years.
+Additionally, we found that prenatal maternal smoke exposure was not associated with global DNA methylation or DNA methylation clocks, underlining the specificity of the DNA methylation signature of prenatal smoke exposure.
+DNA methylation clocks attempt to measure aspects of biological aging and are trained in specific populations.
+We note that the pediatric DNA methylation clock, which was specifically trained for use in pediatric populations, accurately recapitulated our subjects chronological ages.
+However, the GRIM age clock, which was trained in adults to measure proximity to mortality, overestimated chronological ages in our population by 10-15 years.
 In an analysis of 4 pediatric cohorts of children under 20 years of age, GRIM age estimates were 30-35 years greater than chronological in buccal cell samples and 10-15 years greater than chronological age in blood samples [@okazaki2021].
-GRIM-age was neither trained nor validated in a pediatric population [@lu2019], which may explain this overestimation.
+GRIM age was neither trained nor validated in a pediatric population [@lu2019], which may explain this overestimation.
 
@@
 We used maternal self-report of prenatal maternal smoke exposure, as serum cotinine levels were not available.
-Due to social desirability bias, this could result in exposure misclassification.
+As mothers may be reluctant to report smoking during pregnancy, some of our children may be misclassified as unexposed to prenatal smoke.
 We would expect this to bias our results towards the null.
-In any analysis of a prenatal exposure and postnatal outcome, there is the possibility of selection bias into the cohort due to live birth bias.
-Selection bias is also possible due to loss-to-follow-up between birth and age 15.
-Additionally, while we controlled for postnatal secondhand smoke exposure and excluded children who reported any own-smoking, we cannot exclude the possibility of residual confounding in this observational cohort.
+Analyses of prenatal exposures and postnatal outcomes are conditioned on live birth and survival or enrollment until the point of outcome measurement.
+Because prenatal maternal smoking is associated with miscarriage [@pineles2014], this could result in selection bias if an unmeasured variable is associated with both live birth and the outcome.
+This selection bias can create a downward bias in the effect estimate.
+This could produce a spurious protective effect (which we did not observe), or, in the case of a true positive association between exposure and outcome, will attenuate the association toward the null[@leung2021; @liew2015].
+Additionally, while we controlled for postnatal secondhand smoke exposure and excluded children who reported any own-smoking, we cannot exclude the possibility some of the observed effects are mediated through these alternative sources of smoke exposure.
+Personal smoking behavior among children in our cohort was relatively rare, and we excluded those participants from our analysis to focus on the prenatal exposure window.
+Future studies may focus on enriched risk samples to examine the joint effect of prenatal smoke exposure and personal smoking.
 
-However, our analysis also has several strengths.
-We analyzed samples from a large cohort of diverse participants underrepresented in genetic and epigenetic research [@Sirugo2019; @Bentley2017].
-While our exposure measurement of prenatal maternal smoking was self-reported, it was assessed prospectively and preceded outcome measurements.
-We analyzed repeated measures of DNA methylation with reproducible array measures conducted in a single batch.
+Our analysis is strengthened by our prospective study design, diverse study participants sampled, rigorous outcome measures, and advanced analytics.
+We analyzed samples from a large cohort of diverse participants, with African, Admix-Latin, and European genetic ancestry participants, including those who are currently underrepresented in genetic and epigenetic research [@Sirugo2019; @Bentley2017].
+Our measurement of prenatal maternal smoke exposure was prospectively assessed at birth and preceded outcome measurements of DNA methylation measures, limiting the potential for recall bias.
+We analyzed repeated measures of DNA methylation with reproducible array measures conducted in a single batch, reducing the impacts of batch effects.
 We tested associations between prenatal maternal smoking and multiple DNA summary measures to evaluate the specificity of the polymethylation scores.
-We adjusted for prenatal maternal drinking, other prenatal drug use and postnatal smoke exposure to examine the specificity of the biomarker to nature and timing of exposure.
+To examine the specificity of the biomarker to the nature and timing of exposure, we adjusted for prenatal maternal drinking, other prenatal maternal drug use, and postnatal secondhand smoke exposure.

### 2022-03-18 Papers/Polymethylation_scores/08_methods.Rmd
colter revisions
post submission revisions

@@
 output:
-  word_document: default
+  word_document: 
+    reference_docx: "../Template.docx"
   html_document: default
-bibliography: Methylation_SmokingBib.bib
+bibliography: [Methylation_SmokingBib.bib, packages.bib]
 ---
@@
 Participants were selected at delivery using a three-stage stratified random sample design which oversampled unmarried mothers by a ratio of 3:1 [@Reichman:2001wv].
-Participants were excluded on the following criteria: those with parents who planned to place the child for adoption, those where the father was deceased, those who did not speak English or Spanish well enough to be interviewed, births where the mothers or babies were too ill to complete the interview, and those where the baby died before the interview could take place.
-Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; additional follow up is ongoing.
-Assessments included medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher and child.
+Participants were excluded at the baseline wave on the following criteria: the parents who planned to place the child for adoption, the father was deceased, the parents did not speak English or Spanish well enough to be interviewed, or the mother or baby were too ill to complete the interview, or the baby died before the interview could take place.
+Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; an age 22 follow up is ongoing.
+Across certain child ages, assessments included medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher and child.
 At ages nine and fifteen a saliva sample was taken from the child [@Reichman:2001wv].
-A subsample of the Fragile Families cohort was selected for saliva DNA and DNA methylation processing.
-To be eligible for DNA methylation children had to 1) have participated and given saliva at age 9 and 2) participated in the SAND {ACRONYM HELP HERE} study, which oversampled from Detroit, Toledo and Chicago.
-A random sample of these eligible participants was then selected.
-<!--#This subsampling was ....
-PUT SUBSAMPLE technique here -->
+A subsample of the Fragile Families cohort was selected for saliva DNA and DNA methylation processing [@Reichman:2001wv].
+To be eligible for the DNA methylation assessment children had to have participated and given saliva at age 9 and 15; after which a random sample of these eligible participants was then selected.
+One exception to this rule is that all participants in the Study of Adolescent Neurodevelopment (SAND) were assayed even if they only provided a sample at one time-point.
+All analyses account for this oversample of these participants.
 
-### Covariates and exposure measurement
+### Pre- and postnatal exposure measurement
 
-Demographic and prenatal maternal substance use variables were derived from maternal self-report questionnaire data at baseline (child's birth).
+Prenatal maternal substance use variables were derived from maternal self-report questionnaire data at baseline (child's birth).
 
-Maternal covariates included maternal income to poverty ratio at baseline, prenatal smoking, prenatal maternal drug and alcohol use, and postnatal maternal or primary caregiver smoking.
-Maternal income to poverty ratio is a constructed variable of the ratio of total household income (as self-reported by the mother) to the official poverty thresholds designated by the United States Census Bureau for the year preceding the interview.
 At baseline, mother's answered categorical questions about their prenatal smoking, drug and alcohol use.
-For maternal prenatal smoking, mother's were asked
+For maternal prenatal smoking, mother's were asked: "During your pregnancy, how many cigarettes did you smoke?".
+Responses were categorical: 2 or more packs a day, 1 or more but less than 2 packs per day, less than 1 pack a day or none.
+Few participants reported smoking a pack or more a day, thus for our primary exposure variable we created a dichotomous any versus no prenatal smoke exposure variable.
 
-> During your pregnancy, how many cigarettes did you smoke? Did you smoke...:
->
-> 2 or more packs a day
->
-> 1 or more but less than 2 packs per day
->
-> Less than 1 pack a day
->
-> None
+In sensitivity analyses we controlled for other types of prenatal exposures.
+Mothers were also asked how frequently they drank during their pregnancy.
+We created a dichotomous any versus no prenatal alcohol exposure variable.
+Similarly, mothers were asked how frequently they used any other drug during their pregnancy.
+We again created a binary variable for any versus no prenatal use of other drugs.
 
-Few participants reported smoking a pack or more a day, thus we dichotomized to any vs no prenatal maternal smoking. For maternal prenatal drug and alcohol use the mothers were asked
+In sensitivity analyses, we controlled for postnatal second hand smoke exposure.
+When the child was 1, 5, 9 and 15 years of age, primary caregivers responded to questions about smoking in the household.
+To encapsulate early life postnatal secondhand smoke exposure, we created a binary variable for postnatal exposure at ages 1 or 5.
+To encapsulate recent postnatal smoke exposure to the age 9 and 15 interview where DNA methylation was collected, we used a categorical variable for packs per day (no smoking, less than one pack/day, one or more packs/day) in the month prior to each interview.
 
-> During your pregnancy how often did you use drugs/drink alcohol (respectively):
->
-> Never
->
-> Less than 1 time per month
->
-> Several times per month
->
-> Several times per week
->
-> Every day
-
-When the child was 1, 5, 9 and 15 years of age, primary caregivers responded to questions about maternal and in-home smoking.
-To encapsulate general early childhood smoke exposure, we created a binary variable for postnatal exposure at ages 1 or 5.
-To encapsulate recent postnatal smoke exposure, we used a categorical variable for packs per day (no smoking, less than one pack/day, one or more packs/day) in the month prior to the age 9 and 15 interview.
-
-Child covariates included child sex, child report of personal cigarette smoking, and child genetic ancestry.
-Mothers reported sex (male/female) of their child at baseline.
+To ensure we measured the direct effect of prenatal exposure to cigarette smoke, rather than the effect of personal postnatal smoking we excluded children who reported own smoking.
 At ages 9 children were asked if they had ever smoked a cigarette or used tobacco (yes/no) and at age 15 they were asked if they had ever smoked an entire cigarette (yes/no).
 
+### Covariate measurement
+
+We used household income-to-poverty ratio at baseline as a proxy for socioeconomic status.
+Income-to-poverty ratio is the ratio of total household income (as self-reported by the mother) to the official poverty thresholds designated by the United States Census Bureau for the year preceding the interview.
+
+Mothers reported sex (male/female) of their child at baseline.
+
 Child genetic ancestry was calculated.
-Grants R01 HD36916, R01 HD073352, and R01 HD076592 provided support for the collection, assay, and analysis of the genetic data in the Fragile Families cohort.
+Grants R01 HD36916, R01 HD073352, and R01 HD076592 provided support for the collection, assay, and analysis of the genetic data in the cohort.
 Principal components (PC) of child genetic ancestry were calculated from genetic data using Illumina PsychChip_v1-1 with child saliva samples.
@@
 Samples with PC1 \< -0.005 and PC2 \> 0.007+0.75(PC1) were assigned to African ancestry.
-Samples with PC1 \> 0.018 and -0.055 \< PC2 \< 0.025 were assigned to Hispanic ancestry.
+Samples with PC1 \> 0.018 and -0.055 \< PC2 \< 0.025 were assigned to Admixed ancestry [@FFPGS2021].
+Samples assigned to Admixed ancestry were primarily from individuals whose mothers self-reported Hispanic ethnicity (`r modeldata%>%group_by(ancestry, cm1ethrace)%>%summarise(n=n())%>%ungroup()%>%group_by(ancestry)%>%mutate(sum=sum(n), percent=round(100*(n/sum), 0))%>%filter(ancestry=='Admixed ancestry - Latin heritage' & cm1ethrace=='3 hispanic')%>%pull(percent)`% of samples), so we label this group Admixed - Latin ancestry.
 
@@
 DNA methylation image data were processed in R statistical software (3.5) using the minfi package [@Aryee2014].
+We used the February 2022 data freeze of the Fragile Families and Child Wellbeing DNA methylation data.
 The red and green image pairs (n=1811) were read into R and the minfi *preprocessNoob* function was used to normalize dye bias and apply background correction.
 Further quality control was applied using the ewastools packages [@Heiss2018].
-We dropped samples with \>10% of sites have detection p-value \>0.01 (n=43), sex discordance between DNA methylation predicted sex and recorded sex (n=20), and abnormal sex chromosome intensity (n=3).
-CpG sites were removed if they had detection p-value \>0.01 in 5% of samples (n=26,830) or were identified as cross-reactive (n=27,782) (Figure 1) [@Chen2013].
-We used the November 2021 data freeze of the Fragile Families and Child Wellbeing DNA methylation data.
+We dropped samples using the following criteria: if \>10% of DNA methylation sites had detection p-value \>0.01 (n=`r probe_fail`), if there was sex discordance between DNA methylation predicted sex and recorded sex (n=`r sex_mismatch`), if samples had outlier methylation values (n=`r meth_outlier`) or if two sequential samples from the same individual exhibited genetic discordance between visits (n=`r genetic_discordance`) (Figure 1).
+Technical replicates were removed (n=`r tech_rep`).
+DNA methylation sites were removed if they had detection p-value \>0.01 in 5% of samples (n=33376) [@Chen2013].
 Relative proportions of immune and epithelial cell types were estimated from DNA methylation measures using a childhood saliva reference panel [@Middleton2021].
 
-We created polymethylation scores for prenatal maternal smoke exposure.
-From an independent meta-analysis of prenatal smoke exposure and newborn DNA methylation, we extracted the regression coefficients of 6,074 CpG sites associated with prenatal maternal smoking at a false discovery ratio corrected $P$ value \<0.05 [@Joubert:2016ue].
+We investigated several summary measures of DNA methylation, including polymethylation scores for smoke exposure, single DNA methylation sites, mean global DNA methylation, and DNA methylation clocks.
+
+Our primary outcome variable was a polymethylation score for prenatal maternal smoke exposure.
+From an independent meta-analysis of prenatal smoke exposure and DNA methylation (the Joubert meta-analysis), we extracted the regression coefficients of 6,074 DNA methylation sites associated with prenatal maternal smoking at a false discovery ratio corrected $P$ value \<0.05 [@Joubert:2016ue].
 We mean-centered the DNA methylation beta values in our study, weighted them by the independent regression coefficients and took the sum.
-We calculated polymethylation scores using regression coefficients from 4 different regressions.
 For our main analysis we used coefficients from a regression of sustained smoking exposure and DNA methylation in newborn cord blood with cell-type control.
-As sensitivity analyses, we used sustained smoking exposure and DNA methylation in newborn cord blood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children, any smoking exposure and DNA methylation in newborn cord blood) [@Joubert:2016ue].
 
+To assess the robustness of polymethylation scores to the source of coefficients for weights, we generated two sensitivity polymethylation scores from the Joubert meta-analysis [@Joubert:2016ue].
+First, we used the regression of 568 cord blood DNA methylation sites associated with prenatal maternal smoking at a strict Bonferroni threshold, when analyses were not adjusted for cell proportions [@Joubert:2016ue].
+Second, we used the regression coefficients of 19 child peripheral blood DNA methylation sites associated with prenatal maternal smoking at the strict Bonferroni threshold when analyses were not adjusted for cell proportions [@Joubert:2016ue].
+
+To assess the robustness of polymethylation scores to the selection of DNA methylation sites, we generated two additional sensitivity polymethylation scores.
+One score was generated using 28 DNA methylation sites and coefficients prioritized from a LASSO regression of prenatal maternal smoking and cord blood methylation in the Norwegian Mother and Child Cohort Study [@Reese:2017ts].
+Another score was generated using 204 DNA methylation sites and coefficients prioritized from an elastic net regression of prenatal maternal smoking and DNA methylation in adolescent peripheral blood [@rauschert2020].
+This score was calculated using the coefficients and code provided by the authors in the R code accompanying the manuscript [@R-DNAsmokeR].
+
+To compare the utility of polymethylation scores relative to single site biomarkers, we selected the top DNA metyhlation site from the Joubert meta-analysis, *AHRR:* cg05575921.
+As a sensitivity analyses, we also selected four other *a priori* DNA methylation sites which were identified from meta-analyses of prenatal maternal smoking and DNA methylation in children's cord and peripheral blood samples: *GFI1*:cg14179389, *CYP1A1*: cg05549655, *MYO1G*: cg22132788, and *MYO1G*:cg04180046 [@Joubert:2016ue; @wiklund2019].
+
+To assess the specificity of the impact of prenatal smoke exposure on DNA methylation, we compared the polymethylation scores to mean global and regulatory regional DNA methylation and methylation-derived age measures.
 Global DNA methylation was calculated for each sample as the mean methylation value of each sample across the cleaned probe set.
 Mean DNA methylation restricted to probes in genomic regions (CpG island, shore, shelf or open sea, as identified in the R package IlluminaHumanMethylation450kanno.ilmn12.hg19 v 0.6.0) was also calculated.
+Pediatric epigenetic age was calculated for each sample using the coefficients and methods provided by the creators (see [\<https://github.com/kobor-lab/Public-Scripts\>](https://github.com/kobor-lab/Public-Scripts){.uri}) [@mcewen2019].
+As a sensitivity analysis, the GRIM-age clock, which was neither trained nor validated in a pediatric population, was calculated as previously described [@lu2019].
 
-Pediatric epigenetic age was calculated for each sample using the coefficients and methods provided by the creators (see [\<https://github.com/kobor-lab/Public-Scripts\>](https://github.com/kobor-lab/Public-Scripts){.uri}) [@mcewen2019].
-The GRIM age clock, including the smoking pack-years sub-scale, was calculated as previously described [@lu2019].
-
-Single *a priori* CpG sites were selected from previous large meta-analyses of prenatal maternal smoking and DNA methylation in children's cord and peripheral blood samples [@Joubert:2016ue; @wiklund2019].
-
-Surrogate variables were calculated from DNA methylation data using the function sva from the R package sva version 3.38.
+To control for unmeasured confounding and potential batch effects, we calculated surrogate variables from DNA methylation data using the function sva from the R package sva version 3.38 [@R-sva].
+The relationship between prenatal maternal smoking and DNA methylation was protected.
 
@@
 
-From the 1692 samples from 884 unique individuals with quality controlled DNA methylation data, we further excluded any samples missing data on: maternal prenatal smoking (0 samples), alcohol (4 samples) or other drug use (6 samples), maternal income to poverty ratio (0 samples), maternal postnatal smoking data ( samples).
-We also excluded samples with missing child sex (0 samples), child age (0 samples) or genetic data ( samples).
-Finally, if a child reported ever smoking a cigarette or using tobacco age 9, we excluded all of their available samples.
-Age 15 samples from children who reported ever smoking a cigarette at age 15 were also excluded.
+We excluded samples missing any data on the following variables: maternal prenatal smoking, maternal prenatal alcohol use, maternal prenatal other drug use, maternal income to poverty ratio, postnatal secondhand smoke exposure variables, child sex, or child age.
+We also excluded samples that were missing genetic data.
+Age 9 samples from children who reported ever smoking a cigarette or using tobacco age 9 or were missing own-smoking data were excluded.
+Age 15 samples from children who reported smoking a cigarette at age 9 or at age 15 or were missing own-smoking data were excluded.
 Children who were missing a response to the question at age 9 but answered that they had never smoked a whole cigarette at age 15 were kept in the sample.
 
-#### Base model
+#### Descriptive statistics
 
-In our base model we adjusted for maternal income to poverty ratio at baseline, child sex, child age, plate from DNA methylation processing, and estimated immune cell proportion estimated from DNA methylation.
-In nonstratified models, we adjusted for the first two components of genetic ancestry from principal component analysis.
+We described the distribution of continuous variables using the mean and standard deviation.
+We described the distribution of categorical variables using percents.
+We compared the distribution of included to excluded samples using the Welch's t-test for continuous variables and the Chi-Square test for categorical variables.
+We report the correlation between continuous measures of DNA methylation (including polymethylation scores, single sites, mean global DNA methylation and estimated cell type proportions) using the Pearson correlation coefficient .
+We report the correlation of these DNA methylation measures between visits using the Pearson correlation coefficient and scatter plots.
+We compared the distribution of variables by exposure status to prenatal maternal smoking using the Welch's t-test for continuous variables and the Chi-Square test for categorical variables.
+For select continuous DNA methylation measures we compare the distribution across prenatal smoke exposure status using violin plots with captive box plots.
+
+#### Regression analyses
+
+To assess the association between our primary exposure variable of prenatal maternal smoking and salivary DNA methylation, we performed age-stratified multivariable linear regressions with the primary polymethlation score as our outcome.
+To facilitate interpretation, we z-score standardized the polymethylation score.
+We adjusted for an indicator variable for maternal residence in Chicago, Detroit or Toledo (over sampled cities), maternal income to poverty ratio at baseline, child sex, child age, plate from DNA methylation processing, estimated immune cell proportion from DNA methylation data, and the first two components of genetic ancestry from principal component analysis.
+While child sex, child age, plate from DNA methylation processing and immune cell proportions are not confounders (as they cannot casually effect prenatal maternal smoke exposure), these variables can strongly effect DNA methylation and so were adjusted for as precision variables.
+We reported the effect estimate for polymethylation score term and the 95% confidence interval (CI).
+We visualized our findings using forest plots.
+
+To assess the portability of the association between our primary exposure variable of prenatal maternal smoking and salivary DNA methylation across genetic ancestry, we additionally stratified by genetic ancestry and reran the models described above.
 In ancestry-stratified models, the first two principal components from principal component analysis run within each ancestry strata were used.
-While child sex, child age, plate from DNA methylation processing and immune cell proportions are not confounders (as they cannot casually effect prenatal maternal smoke exposure), these variables can strongly effect DNA methylation and so were adjusted for as precision variables.
+
+To compare the utility of the polymethylation score to a single site biomarkers, we used percent methylation at AHRR:cg05575921 as the outcome in the models described above.
+To assess the specificity of the impact of prenatal maternal smoking on DNA methylation, we used mean percent global DNA methylation and the pediatric DNA methylation clock as outcome variables.
+
+#### Regression sensitivity analyses
+
+To assess the specificity of the DNA methylation signal to the type and timing of exposure, we additionally adjusted for the following variables in sensitivity analyses: 1) prenatal drug and alcohol use and 2) postnatal second-hand smoke exposure in addition to prenatal drug and alcohol use.
+To assess the robustness of the association between prenatal maternal smoking and DNA methylation to batch effects and residual confounding, we also performed regression analyses adjusting for surrogate variables calculated from the DNA methylation data.
+
+To assess the robustness of the association to choice of polymethylation score, we performed the same regression analyses described above with z-score standardized sensitivity polymethylation scores as the outcome.
+Similarly, we replaced the outcome variable of *AHRR:*cg05575921 with other *a priori* selected DNA methylation sites in sensitivity analyses.
+We plotted the estimates and 95% CI for each DNA methylation site along with the previously reported estimates from the literature using forest plots [@Joubert:2016ue; @wiklund2019]
 
@@
 First, we regressed exposure to prenatal maternal smoking (outcome) against the DNA methylation summary measures in individual logistic regressions, while adjusting for the base model variables listed above.
-Next, we calculated receiver operating curves (ROC) and area under the curves (AUCs) using the function roc from the R library pROC version 1.18.0.
-We compared ROC curves and AUC using the Delong method and the function roc.test from the R library pROC version 1.18.0.
+Next, we calculated receiver operating curves (ROC) and area under the curves (AUCs) using the function roc from the R library pROC version 1.18.0 [@pROC2011].
+We compared ROC curves and AUCs from models with only base model variables to those which included base model variables and the methylation summary measure using the Delong method and the function roc.test from the R library pROC version 1.18.0.
+This is appropriate as the base model variable only model is nested within the base model variable + methylation summary measures model.
 
-#### Sensitivity analyses
+#### Receiver operator curve sensitivity analyses
 
-In sensitivity analyses we performed additional adjustments: 1) models adjusting for prenatal drug and alcohol use, 2) models adjusting for prenatal drug and alcohol use and postnatal maternal/primary caregiver smoking, 3) models adjusting for surrogate variables calculated from DNA methylation data.
+To assess the robustness of polymethylation scores classification accuracy to choice of coefficients, we used polymethylation scores calculated using different coefficient sets as predictors for the receiver operating curve analysis in sensitivity analyses.
 
-In addition to the cross-sectional models within visit-strata, we performed a linear mixed effect model with a random intercept for child.
+To compare the performance of saliva-based polymethylation scores to previous reports of blood-based polymethylation scores, we used a forest plot.
+We plotted the AUC and 95% CI (when available) for each polymethylation score in our cohort and in the cohorts previously reported in the literature [@rauschert2020; @Richmond:2018tu].
 

### 2022-03-18 Papers/Polymethylation_scores/99_FigureCaptions.Rmd
post submission revisions

@@
 
-####  **Figure 4 - Polymethylation scores accurately classify prenatal maternal smok exposure at ages 9 and 15 among `r n_total` children in the Fragile Families and Child Wellbeing study** A) Receiver operator curve for select DNA methylation measures for predicting prenatal smoke exposure using no other variables (light colors) or using base model variables (dark colors, other variables included: child sex, maternal income-poverty ratio at birth, immune cell proportion and batch of methylation data processing).
+####  **Figure 4 - Polymethylation scores accurately classify prenatal maternal smoke exposure at ages 9 and 15 among `r n_total` children in the Fragile Families and Child Wellbeing study** A) Receiver operator curve for select DNA methylation measures for predicting prenatal smoke exposure using no other variables (light colors) or using base model variables (dark colors, other variables included: child sex, maternal income-poverty ratio at birth, immune cell proportion and batch of methylation data processing).
 B) Receiver operator curve for include polymethylation scores individually at each visit (black & light grey) or jointly (dark grey).
@@
 ```
+
+#### **Figure 5 - Comparison of different polymethylation scores and their accuracy at classifying prenatal smoke exposure among `r n_total` children with saliva samples in the Fragile Families and Child Wellbeing study as opposed to previously reported results in cord and peripheral blood samples
+
+```{r, echo=F, eval=print_figs, include=T,  fig.width=12}
+auc_compare
+```

### 2022-03-18 Papers/Polymethylation_scores/index_paper.Rmd
post submission revisions

--- 2021-11-04 Papers/Polymethylation_scores/index_paper.Rmd
@@
 title: "PAPER DRAFT"
-author: "------------"
+author: '------------'
 date: "`r Sys.Date()`"
 output:
-  word_document: default
+  word_document: 
+    reference_docx: "../Template.docx"
   html_document:
     df_print: paged
-bibliography: Methylation_SmokingBib.bib
+  pdf_document: default
+bibliography: [Methylation_SmokingBib.bib, packages.bib]
 csl: springer-vancouver.csl
-always_allow_html: true
+always_allow_html: yes
 ---
 
-```{r, echo=F, include=F, warning=F, message=F}
+```{r, echo=F, include=F, message=F, warning=F}
 library(stringr)
+library(knitr)
+library(tidyverse)
+library(tidyr)
+library(dplyr)
+
+coefCI=function(df, filter_statement, my_term, si){
+  my_df=df%>%
+    filter(eval(parse(text = filter_statement)))%>%
+    unnest(tidied)%>%
+    mutate(val=paste0(round(estimate, si), ' (95%CI: ', 
+                      round(conf.low, si), ', ',
+                      round(conf.high, si), ')'))%>%
+    filter(term==my_term)%>%
+    pull(val)
+  return(my_df)
+}
+```
+
+```{r load data, echo=F, include=F, warning=F, message=F, cache=TRUE, cache.lazy=F}
 results_run=str_detect(here::here(), 'nfs/turbo/bakulski1')
+knitr::knit_child(here::here("Reports", "Paper1.Rmd"))
 ```
@@
 
+```{r child = here::here("Papers", "Polymethylation_scores", "08_methods.Rmd")}
+```
+
 ```{r child = here::here("Papers", "Polymethylation_scores", "05_results.Rmd"), eval=results_run}
@@
 
-```{r child = here::here("Papers", "Polymethylation_scores", "08_methods.Rmd")}
-```
-
 ```{r child = here::here("Papers", "Polymethylation_scores", "98_declarations.Rmd")}

### 2022-03-18 Papers/Polymethylation_scores/random_bits.Rmd
post submission revisions

---
title: "Untitled"
author: "F blostein"
date: "3/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In bivariate analyses and multivariable models, we focused on the 6074 site polymethylation score for prenatal smoke exposure constructed using the Joubert metaanalysis coefficients from newborn cord blood and the single top CpG site from prior research, cg05575921 in the *AHRR* gene , as hypothesized biomarkers [@Joubert:2016ue], and used global methylation and the pediatric clock as negative controls.

For our main analysis we used coefficients for 6074 sites from a regression of sustained smoking exposure and DNA methylation in newborn cord blood with cell-type control. As sensitivity analyses, we used coefficients for the same sites from regressions of: sustained smoking exposure and DNA methylation in newborn cord blood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children without cell-type control, and any smoking exposure and DNA methylation in newborn cord blood without cell-type control [@Joubert:2016ue].

Two of these scores, reported by Richmond et al., used regression coefficients from the Joubert et al. meta-analysis as described above. However these two scores were limited to the DNA methylation sites with $P$ values<1.07613e-07, respectively 568 sites in the newborn cord blood regression and 19 sites in the older children peripheral blood regressions. A third score used 28 sites from a LASSO regression of prenatal maternal smoking and DNA methylation in newborn cordblood. A fourth score used 204 sites from an elastic net regression of prenatal maternal smoking and DNA methylation in peripheral blood from 17 year olds

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2022-03-18 Scripts/01_clean_phenotype_data.R

@@ -56,72 +56,21 @@ FF_labeled=FF_labeled%>%mutate(across(everything(), ~na_codes(., my_na_codes)))%
 
 ############################pdqc data######################################
 
-#As of October 4 added in the 9 sex incorrect at birth samples (see Jonah email search Check in on sex filtration/clocks Fragile Families)
-#removed 3 sex outlier samples to bring sample size in line w/ air pollution paper 1745 observations
-pdqc_all<-readRDS(paste0(datadir, 'OGData/', "pd_qc.rds"))
-airpollution_study<-read.csv(paste0(datadir, 'OGData/', 'isee_pollutionpegs_newvars_5.6.2021.csv'))
-IDsNotInAir=pdqc_all$MethID[!(pdqc_all$MethID %in% airpollution_study$MethID)]
-IDsNotInAir= pdqc_all %>%filter(MethID %in% IDsNotInAir)
-childIDNotInAir=IDsNotInAir%>%filter(childteen!='M')
-#figure out samples w/ incorrect sex at birth variable 
-pdqc_all = pdqc_all %>% left_join(airpollution_study %>% 
-                                    mutate(idnum=as.character(idnum), sex_new=sex)%>% 
-                                    dplyr::select(idnum, MethID, childteen, sex_new))%>%
-   mutate(sex_flag=ifelse(sex!=sex_new, 'recode_sex', 'correct_sex'))
+#As of February 11 switch to February 2022 Data Freeze, which includes some fixes to sex mismatches due to incorrect
+#coding at birth interview; only clean samples which are not technical replicates are included in this file according to Jonah 
+pdqc_all<-read.csv(paste0(datadir, 'OGData/Feb2022DataFreeze/', "pd_analytic_freeze1.csv"))
+#Flag samples w/ incorrect sex at birth variable 
+pdqc_all = pdqc_all %>% 
+          left_join(FF_labeled %>% 
+                      mutate(sex_old=stringr::word(cm1bsex), idnum=as.integer(id))%>% 
+                      dplyr::select(idnum, sex_old))%>%
+          mutate(sex_flag=ifelse(cm1bsex!=sex_old, 'recode_sex', 'correct_sex'), 
+                 new_sex=cm1bsex)
 sex_recodes=pdqc_all %>% filter(sex_flag=='recode_sex')%>%pull(idnum)
-
-#############################mislabeled samples data###############################
[…]

### 2022-03-18 Scripts/02_make_methyl_summary.R

@@ -6,7 +6,6 @@ clock_coefs=file.path(paste0(datadir, 'OGData'), 'clock_coefs')
 
 
 ####################################################################library
-library(dplyr)
 library(EpiDISH)
 library(ewastools)
 library(tidyr)
@@ -17,6 +16,8 @@ library(wateRmelon)
 library(data.table)
 library(DunedinPoAm)
 library(knitr)
+library(dplyr)
+library(DNAsmokeR)
 ####################################################################source functions
 source(file.path(codedir, 'UsefulCode', 'makePolyEpiScores.R'))
 
@@ -24,9 +25,8 @@ source(file.path(codedir, 'UsefulCode', 'makePolyEpiScores.R'))
 aprioriCG<-c("cg05575921", "cg04180046", "cg05549655", "cg14179389", "cg22132788")
 
 ################################read in methylation data
-#as of october 4 switched to beta file with  423668 probes (see Jonah email search Check in on sex filtration/clocks Fragile Families)
-betaqc<-readRDS(file=paste0(datadir, 'OGData/', "betaqc.rds"))
-if(nrow(betaqc)!=423668){stop("Probe number not what thought, stop and check betaqc")}
+betaqc<-readRDS(file=paste0(datadir, 'OGData/Feb2022DataFreeze/', "beta_analytic_freeze1.rds"))
+if(nrow(betaqc)!=437588){stop("Probe number not what thought, stop and check betaqc")}
 ###########################################cell type proportions 
 celltypes<-epidish(beta.m = betaqc, ref.m = centEpiFibIC.m, method = "RPC")
 #RPC method w/ centEpiFibIC
@@ -65,7 +65,7 @@ aprioriCGdf<-as.data.frame(t(betaqc[aprioriCG, ]))
 aprioriCGdf=aprioriCGdf*100
 aprioriCGdf$MethID = colnames(betaqc)
 
-#################################################polymethylation scores
+#################################################polymethylation scores from joubert & reese
 cpgs<-read.csv(paste0(datadir, 'PMC4833289_SuppFile2CpGs.csv'), stringsAsFactors = F, header=F, skip=2)
 #make headers
 header1<-scan(paste0(datadir, 'PMC4833289_SuppFile2CpGs.csv'), nlines=1, what=character(), sep=',')
@@ -79,6 +79,21 @@ names(cpgs)<-header2
[…]

### 2022-03-18 Scripts/03_sva_variables.R

@@ -12,8 +12,8 @@ data.dir=gsub('Code', 'Data/', here::here())
 output.dir=paste0(data.dir, 'CreatedData/')
 
 #load in betaqc matrix
-#as of october 4 switched to beta file with  423668 probes (see Jonah email search Check in on sex filtration/clocks Fragile Families)
-betaqc<-readRDS(file=paste0(data.dir, 'OGData/', "betaqc.rds"))
+#as of February 2022 switched to betaqc matrix with 437588 
+betaqc<-readRDS(file=paste0(data.dir, 'OGData/Feb2022DataFreeze/', "beta_analytic_freeze1.rds"))
 
 load(paste0(data.dir, '/CreatedData/completeCasepheno.Rdata'))
 pheno<-completecase # dataset with the most exclusion i.e. squaring dataset off for all models (excluding kids who smoke in figure)

### 2022-03-18 Scripts/04_summary_methyl_models.R

@@ -35,8 +35,9 @@ modeldata=completecase%>%droplevels()%>%copy_labels(completecase)
 
 #set as factors 
 modeldata=modeldata %>% mutate(childteen=factor(childteen, levels=c('Age 9', 'Age 15')))
+#reversed levels - very few prenatal smoking + no maternal smoking at age 1 and 5
 modeldata$PostnatalMaternalSmokingAny<-factor(modeldata$PostnatalMaternalSmokingAny, 
-                                              levels=c("No maternal smoking at age 1 and 5", 
+                                              levels=c("No maternal smoking at age 1 and 5",
                                                        "Maternal smoking at age 1 or age 5"))
 modeldata$SmkAtVisitPastmonth<-factor(modeldata$SmkAtVisitPastmonth, 
                                       levels=c("No smoking", 
@@ -46,7 +47,7 @@ modeldata$SmkAtVisitPastmonth<-factor(modeldata$SmkAtVisitPastmonth,
 modeldata$Sample_Plate<-factor(modeldata$Sample_Plate)
 
 #sv data - no variance filter, whole subset
-sv.vNone=readRDS(paste0(datadir, '/svNone.Rds'))
+sv.vNone=readRDS(paste0(datadir, '/svNone.Rds')) # temporarily changed to archived sv well rerunning with new beta matrix 
 is_sv=function(x) str_replace_all(x, " ", "")
 sv.vNone=sv.vNone%>%rename_with(is_sv, contains('sv'))
 modeldata=left_join(modeldata, sv.vNone)%>%copy_labels(modeldata)
@@ -58,15 +59,11 @@ save(modeldata, file=file.path(datadir, file='completeCaseSVA.Rdata'))
 #################################Set variables for modeling
 #outcomes & outcome_labels
 
-y_vector<-c("globalmethylation", "pediatric", "anynewborn_center_scale", 
-            "SSnewbornCT_center_scale", "SSnewborn_center_scale",  "SSolder_center_scale", 
+y_vector<-c("globalmethylation", "PedBE", "anynewborn_center_scale", 
+            "SSnewbornCT_center_scale", "SSnewborn_center_scale",  "SSolder_center_scale", 'reese_center_scale',
+            'richmond19_center_scale', 'richmond568_center_scale',
             "cg05575921", "cg04180046", "cg05549655", "cg14179389","cg22132788")
-outcome_labels=c('Global methylation', 'Pediatric clock',  
[…]

### 2022-03-18 Scripts/05_ewas.R

@@ -9,6 +9,7 @@ library(limma)
 library('variancePartition')
 library('edgeR')
 library('BiocParallel')
+#NOTE FEBRUARY MAKE SURE DROP probes that are cross reactive before doing EWAS (new beta did not remove cross reactive probes) 
 
 datadir="/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data"
 outputdir<-"/nfs/turbo/bakulski1/People/blostein/FF_methylation/Data/CreatedData/EWAS"

### 2022-04-08 Papers/Polymethylation_scores/02_abstract.Rmd
coauthor feedback april 4

@@
 DNA methylation signatures can function as biomarkers of prenatal smoke exposure.
-However the robustness of prenatal maternal smoking DNA methylation signatures across child ages, genetic ancestry groups or tissues is not clear.
+However, the robustness of prenatal maternal smoking DNA methylation signatures across child ages, genetic ancestry groups or tissues is not clear.
 Using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cord blood, we created polymethylation scores of saliva DNA methylation from children at ages 9 and 15 in the Fragile Families and Child Wellbeing study.
 We tested whether prenatal maternal smoking was associated with salivary polymethylation scores for prenatal smoke exposure and assessed the consistency of associations by age and participant genetic ancestry group.
-We tested the accuracy of exposure classification using area under the curve (AUC) from receiver operating curve analyses and compared to findings from candidate DNA methylation sites.
+We tested the accuracy of exposure classification using area under the curve (AUC) from receiver operating curve analyses and compared to findings from single DNA methylation sites selected from previous meta-analyses.
 

### 2022-04-08 Papers/Polymethylation_scores/04_background.Rmd
coauthor feedback april 4

@@
 Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua].
-The gold-standard smoking biomarker, serum cotinine levels, have a half-life of nine hours in pregnant women, and are rarely available outside of birth cohort research [@Dempsey:2002wr].
+The gold-standard smoking biomarker, serum cotinine levels, has a half-life of nine hours in pregnant women, and is rarely available outside of birth cohort research [@Dempsey:2002wr].
 Alternative biomarkers of prenatal maternal smoke exposure with a larger window of measurement could improve research and clinical practice.
 
-DNA methylation is a possible biomarker.
+DNA methylation is a possible biomarker to capture prenatal smoking exposure.
 Prenatal maternal smoking is associated with offspring DNA methylation signatures at birth in placental and cord blood samples, and postnatally in peripheral blood samples [@Joubert:2016ue; @Everson:2021td; @Cardenas2019; @Richmond:2018tu; @Tehranifar:2018tb].
@@
 In addition, most studies of prenatal maternal smoking and childhood DNA methylation have been conducted in primarily European genetic ancestry populations [@Joubert:2016ue; @Ladd-Acosta:2016aa; @Tehranifar:2018tb; @Zhang2018; @Gonseth2016; @Neophytou:2019ty], and it will be important to test generalizability to other groups.
-To our knowledge, few studies have assessed multiple genetic ancestry groups in the same cohort and no study has evaluated associations between maternal pregnancy cigarette smoking and salivary DNA methylation in children.
+Few studies have assessed multiple genetic ancestry groups in the same cohort and no study has evaluated associations between maternal pregnancy cigarette smoking and salivary DNA methylation in children.
 
-Using the findings from large-scale association testing, in independent populations, DNA methylation-based biomarkers of prenatal maternal smoking can be developed and assessed by evaluating their classification accuracy.
+Using the findings from large-scale association testing in independent populations, DNA methylation-based biomarkers of prenatal maternal smoking can be developed and assessed by evaluating their classification accuracy.
 Biomarkers can range from single DNA methylation sites to summary measures across multiple sites, termed polymethylation scores [@Philibert2020 @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa].
-In a polymethylation score, DNA methylation sites are weighted by their effect size from a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a].
+In a polymethylation score, DNA methylation sites are weighted by their effect size from a previously conducted analysis in an independent sample and then summed to a single score [@Chen2020; @Bakulski2021a].
 DNA polymethylation scores for prenatal maternal smoke exposure have performed well (area under the curves (AUCs)\>0.8) in cord blood samples from newborns and peripheral blood samples from young children [@Reese:2017ts; @Ladd-Acosta:2016aa] and fairly well in peripheral blood from middle-aged adults (AUC of 0.72 (95% confidence interval 0.69, 0.76)) [@Richmond:2018tu].
@@
 In the Fragile Families and Child Wellbeing Study, a diverse longitudinal birth cohort, we aimed to assess salivary DNA methylation biomarkers for prenatal maternal cigarette smoking.
-We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including polymethylation scores for smoke exposure, individual a priori DNA methylation sites, global DNA methylation, and epigenetic clocks, at ages 9 and 15.
+We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including polymethylation scores for smoke exposure, individual *a priori* DNA methylation sites, global DNA methylation, and epigenetic clocks, at ages 9 and 15.
 We then performed genetic ancestry- and age-stratified analyses to test the hypothesis that DNA methylation could serve as a portable and persistent biomarker of prenatal maternal smoking.

### 2022-04-08 Papers/Polymethylation_scores/05_results.Rmd
coauthor feedback april 4

@@
  All of the polymethylation scores for prenatal smoke exposure were correlated (Supplemental Figure 1).
-Our primary polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $\rho$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value `r my_p_val(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value)`).
+Our primary polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $r$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value `r my_p_val(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value)`).
 
 We compared epigenetic measures between the age 9 and age 15 visit among the `r n_twovisit` individuals with data from both visits (Supplemental Figure 2).
-The correlation across ages was strongest for the polymethylation scores for prenatal smoke exposure constructed using the Joubert meta-analysis coefficients from newborn cord blood (6074 site score with coefficients from regression with cell-type control Pearson $\rho$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value\<0.0001; 568 site score $\rho$=`r round(cor(model_twovisit$richmond568_center_Age_9, model_twovisit$richmond568_center_Age_15), 2)`, *P* value\<0.0001).
-Global DNA methylation (`r round(cor(model_twovisit$globalmethylation_Age_9, model_twovisit$globalmethylation_Age_15), 2)`, *P* value\<0.0001) and epigenetic clocks were less strongly correlated across ages (Pediatric: `r round(cor(model_twovisit$PedBE_Age_9, model_twovisit$PedBE_Age_15), 2)`; *P* value\<0.0001, GRIM: `r round(cor(model_twovisit$DNAmGrimAge_Age_9, model_twovisit$DNAmGrimAge_Age_15), 2)`; *P* value\<0.0001) The distribution of epigenetic ages shifted higher between the age 9 and age 15 visit, as expected (Supplemental Figure 3).The distribution of estimated cell-type proportions also shifted between visits while the distribution of global methylation and polymethylation scores were more consistent (Supplemental Figure 3).
+The correlation across ages was strongest for the polymethylation scores for prenatal smoke exposure constructed using the Joubert meta-analysis coefficients from newborn cord blood (6074 site score with coefficients from regression with cell-type control Pearson $r$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value\<0.0001; 568 site score $r$=`r round(cor(model_twovisit$richmond568_center_Age_9, model_twovisit$richmond568_center_Age_15), 2)`, *P* value\<0.0001).
+Global DNA methylation (`r round(cor(model_twovisit$globalmethylation_Age_9, model_twovisit$globalmethylation_Age_15), 2)`, *P* value\<0.0001) and epigenetic clocks were less strongly correlated across ages (Pediatric: `r round(cor(model_twovisit$PedBE_Age_9, model_twovisit$PedBE_Age_15), 2)`; *P* value\<0.0001, GRIM: `r round(cor(model_twovisit$DNAmGrimAge_Age_9, model_twovisit$DNAmGrimAge_Age_15), 2)`; *P* value\<0.0001) The distribution of epigenetic ages shifted higher between the age 9 and age 15 visit, as expected (Supplemental Figure 3).The distribution of estimated cell-type proportions also shifted between visits, while the distribution of global methylation and polymethylation scores were more consistent (Supplemental Figure 3).
 Almost all saliva samples consisted of mostly immune cells based on estimated cell-type proportions.
@@
 Children of mothers who reported smoking during pregnancy were more likely to be of European genetic ancestry (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[2, 2], 2)*100`%) than children of mothers who did not (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[1, 2], 2)*100`%, *P* value=\<0.001).
-Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both the age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value\<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value\<0.001) visits (Table 1, Supplemental Table 2).
+Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value\<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value\<0.001) visits (Table 1, Supplemental Table 2).
 At age 9, children exposed to prenatal maternal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`%, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
@@
 The effect was similar among European ancestry children at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`) and consistent in direction but attenuated at age 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`).
-The direction of the association was consistent in direction but attenuated among Admixed - Latin ancestry children at ages 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`) and 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`) (Figure 3; Supplemental Table 3).
+The association was attenuated among Admixed - Latin ancestry children at ages 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`) and 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`) (Figure 3; Supplemental Table 3).
 
@@
 To assess the specificity of the DNA methylation signal to the timing of cigarette smoke exposure, we additionally adjusted for exposure to postnatal maternal smoking.
-Very few children exposed to maternal smoking prenatally were unexposed postnatally (Supplemental Figure 5).In bivariate analyses, children exposed to only postnatal smoking had lower polymethylation scores than those exposed to pre- and postnatal smoking.
+Very few children exposed to maternal smoking prenatally were unexposed postnatally (Supplemental Figure 5).
+In bivariate analyses, children exposed to only postnatal smoking had lower polymethylation scores than those exposed to pre- and postnatal smoking.
 However, they did not have higher polymethylation scores than children unexposed to both pre-and postnatal smoking (Supplemental Figure 5).
@@
 In addition to cg055975921 in the *AHHR* gene, we tested the association of four other *a priori* DNA methylation sites identified from previous meta-analyses.
-We replicated the association of prenatal maternal smoking with these DNA methylation at GFI1:cg14179389, CYP1A1: cg05549655, MYO1G: cg22132788, and MYO1G:cg04180046  (Supplemental Figure 8).
+We replicated the association of prenatal maternal smoking with these DNA methylation at GFI1: cg14179389, CYP1A1: cg05549655, MYO1G: cg22132788, and MYO1G: cg04180046  (Supplemental Figure 8).
 
@@
 Next, we assessed whether including DNA methylation summary measures increased the accuracy of classifying prenatal smoke exposure over a base model consisting of child sex, maternal income to poverty ratio at baseline, child age at DNA methylation measurement, estimated immune cell proportion, plate from methylation processing, the first two genetic principal components and an indicator variable for oversampled city.
-At age 9, including the polymethylation score increased the AUC by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)`) percentage points over the base model (*P* value from DeLong test=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(7) %>%format(., scientific=F)`).
-At age 15 including the polymethylation score for prenatal smoke exposure increased the AUC by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)`) percentage points (*P* value from DeLong test=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(7)%>%format(., scientific=F)`).
-Including DNA methylation at *AHRR*:cg0557591 with base model covariates did not improve the classification compared to using base model covariates alone  (Supplemental Tables 4 and 5).
-Classification accuracy was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B).
+At age 9, including the polymethylation score increased the AUC by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)`) percentage points over the nested base model (*P* value from DeLong test=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(7) %>%format(., scientific=F)`).
+At age 15 including the polymethylation score for prenatal smoke exposure increased the AUC by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)` percentage points (*P* value from DeLong test=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(7)%>%format(., scientific=F)`).
+Including DNA methylation at *AHRR*: cg0557591 with base model covariates did not improve the classification compared to using base model covariates alone  (Supplemental Tables 4 and 5).
+Classification accuracy when including the polymethylation score from btoh age 9 and 15 was very similar to classification accuracy when using the polymethylation score from age 9 only (Figure 4B).
 
@@
 
-The AUC for classification of prenatal maternal smoke exposure in Fragile Families was relatively similar regardless of which polymethylation score was used (Figure 5, Supplemental Table 5).The highest AUC was from the 568 site score using coefficients from the regression of prenatal smoke exposure and DNA methylation in newborn cord blood.
+The AUC for classification of prenatal maternal smoke exposure in the Fragile Families study was relatively similar regardless of which polymethylation score was used (Figure 5, Supplemental Table 5).
+The highest AUC was from the 568 site score using coefficients from the regression of prenatal smoke exposure and DNA methylation in newborn cord blood.
 Performance was similar to previous reports.

### 2022-04-08 Papers/Polymethylation_scores/06_discussion.Rmd
coauthor feedback april 4

@@
 Our findings are consistent with the previous literature in different tissues and at different ages on associations between prenatal maternal smoking and DNA methylation.
-We replicated top hits from the seminal epigenome wide association meta-analysis of prenatal smoke exposure and DNA methylation of cord blood from newborns including *AHRR*:cg05575921, *GFl1*:cg14179389 and *MYO1G1*:cg22132788.
-Consistent with a meta-analysis of peripheral blood from children and adults over 16 years of age, the association at *MYO1G1*:cg22132788 was larger than in newborn cord blood [@Joubert:2016ue; @wiklund2019].
-Again consistent with a meta-analysis of peripheral blood from children and adults over 16 years of age, the associations at *AHRR*:cg05575921 and *GFl1*:cg14179389 were smaller than in newborn cord blood [@Joubert:2016ue; @wiklund2019].
+We replicated top hits from the seminal epigenome wide association meta-analysis of prenatal smoke exposure and DNA methylation of cord blood from newborns including *AHRR*: cg05575921, *GFl1*: cg14179389 and *MYO1G1*: cg22132788.
+Consistent with a meta-analysis of peripheral blood from children and adults over 16 years of age, the association at *MYO1G1*: cg22132788 was larger than in newborn cord blood [@Joubert:2016ue; @wiklund2019].
+Again, consistent with a meta-analysis of peripheral blood from children and adults over 16 years of age, the associations at *AHRR*: cg05575921 and *GFl1*: cg14179389 were smaller than in newborn cord blood [@Joubert:2016ue; @wiklund2019].
 There are no previous studies of prenatal maternal smoking and saliva DNA methylation.
@@
 At age 9, the effect estimates for prenatal maternal smoking on the polymethylation score in the European and African genetic ancestry samples were was within `r abs(coefOnly(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)-coefOnly(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2))` standard deviation and were both significant at an alpha of 0.05.
-This finding is in line with a recent epigenome-wide association study in which children of both black and non-Black identifying mothers exhibited similar directions of associations between prenatal smoke exposure and peripheral blood methylation at 38 DNA methylation sites [@Xu2021].
+This finding is in line with a recent epigenome-wide association study in which children of both black and non-black identifying mothers exhibited similar directions of associations between prenatal smoke exposure and peripheral blood methylation at 38 DNA methylation sites [@Xu2021].
 In our analysis, the association between prenatal smoke exposure and polymethylation scores was attenuated in children of admixed genetic ancestry (primarily of Hispanic ethnicity) at both ages and in children of European genetic ancestry at age 15.
@@
 
-While polymethylation scores significantly improved classification of prenatal smoke exposure over a base model, using *AHRR*:cgg05575921 methylation as a classifier did not (percentage point increase in AUC: `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'AHRR'))%>%pull(5)`) .
-Prenatal smoke exposed newborns consistently exhibit lower DNA methylation at *AHRR*:cg05575291 than unexposed newborns [@Joubert:2016ue].
-However, the association between prenatal smoke exposure and *AHRR*:cg05575291 methylation did not persist in peripheral blood samples from adult women [@Richmond:2018tu].
-Additionally, a previous longitudinal analysis of children's blood samples from birth to 17 years of age in the ALSPAC cohort found that the difference in DNA methylation at *AHRR*:cg05575291 between prenatal smoke exposed and unexposed children decreased from birth to 17 years of age.
-This previous analysis also reported that differences in DNA methylation at *AHRR*:cg05575291 in adolescence may be partially mediated by postnatal own smoking, and we excluded children who reported postnatal own-smoking [@Richmond:2018tu].
-As the accuracy of *AHRR*:cg05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures, incorporating information across multiple sites of DNA methylation may yield a a more robust biomarker.
+While polymethylation scores significantly improved classification of prenatal smoke exposure over a base model, using *AHRR*: cg05575921 methylation as a classifier did not (percentage point increase in AUC: `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'AHRR'))%>%pull(5)`)).
+Prenatal smoke exposed newborns consistently exhibit lower DNA methylation at *AHRR*: cg05575291 than unexposed newborns [@Joubert:2016ue].
+However, the association between prenatal smoke exposure and *AHRR*: cg05575291 methylation did not persist in peripheral blood samples from adult women [@Richmond:2018tu].
+Additionally, a previous longitudinal analysis of children's blood samples from birth to 17 years of age in the ALSPAC cohort found that the difference in DNA methylation at *AHRR*: cg05575291 between prenatal smoke exposed and unexposed children decreased from birth to 17 years of age.
+This previous analysis also reported that differences in DNA methylation at *AHRR*: cg05575291 in adolescence may be partially mediated by postnatal own smoking, and we excluded children who reported postnatal own-smoking [@Richmond:2018tu].
+As the accuracy of *AHRR*: cg05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures, incorporating information across multiple sites of DNA methylation may yield a a more robust biomarker.
 
-Additionally, we found that prenatal maternal smoke exposure was not associated with global DNA methylation or DNA methylation clocks, underlining the specificity of the DNA methylation signature of prenatal smoke exposure.
+Additionally, we found that prenatal maternal smoke exposure was not associated with global DNA methylation or a pediatric saliva and the GrimAge DNA methylation clocks, underlining the specificity of the DNA methylation signature of prenatal smoke exposure.
 DNA methylation clocks attempt to measure aspects of biological aging and are trained in specific populations.
@@
 However, the GRIM age clock, which was trained in adults to measure proximity to mortality, overestimated chronological ages in our population by 10-15 years.
-In an analysis of 4 pediatric cohorts of children under 20 years of age, GRIM age estimates were 30-35 years greater than chronological in buccal cell samples and 10-15 years greater than chronological age in blood samples [@okazaki2021].
+In an analysis of four pediatric cohorts of children under 20 years of age, GRIM age estimates were 30-35 years greater than chronological in buccal cell samples and 10-15 years greater than chronological age in blood samples [@okazaki2021].
 GRIM age was neither trained nor validated in a pediatric population [@lu2019], which may explain this overestimation.
@@
 This selection bias can create a downward bias in the effect estimate.
-This could produce a spurious protective effect (which we did not observe), or, in the case of a true positive association between exposure and outcome, will attenuate the association toward the null[@leung2021; @liew2015].
+This could produce a spurious protective effect (which we did not observe), or, in the case of a true positive association between exposure and outcome, will attenuate the association toward the null [@leung2021; @liew2015].
 Additionally, while we controlled for postnatal secondhand smoke exposure and excluded children who reported any own-smoking, we cannot exclude the possibility some of the observed effects are mediated through these alternative sources of smoke exposure.

### 2022-04-08 Papers/Polymethylation_scores/08_methods.Rmd
coauthor feedback april 4

@@
 Participants were selected at delivery using a three-stage stratified random sample design which oversampled unmarried mothers by a ratio of 3:1 [@Reichman:2001wv].
-Participants were excluded at the baseline wave on the following criteria: the parents who planned to place the child for adoption, the father was deceased, the parents did not speak English or Spanish well enough to be interviewed, or the mother or baby were too ill to complete the interview, or the baby died before the interview could take place.
+Participants were excluded at the baseline wave on the following criteria: the parents who planned to place the child for adoption, the father was deceased, the parents did not speak English or Spanish well enough to be interviewed, the mother or baby were too ill to complete the interview, or the baby died before the interview could take place.
 Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; an age 22 follow up is ongoing.
@@
 One exception to this rule is that all participants in the Study of Adolescent Neurodevelopment (SAND) were assayed even if they only provided a sample at one time-point.
-All analyses account for this oversample of these participants.
+All analyses account for oversampling of these participants.
 
@@
 Prenatal maternal substance use variables were derived from maternal self-report questionnaire data at baseline (child's birth).
-
-At baseline, mother's answered categorical questions about their prenatal smoking, drug and alcohol use.
+At baseline, mothers answered categorical questions about their prenatal smoking, drug and alcohol use.
 For maternal prenatal smoking, mother's were asked: "During your pregnancy, how many cigarettes did you smoke?".
@@
 
-In sensitivity analyses, we controlled for postnatal second hand smoke exposure.
+In sensitivity analyses, we controlled for postnatal second-hand smoke exposure.
 When the child was 1, 5, 9 and 15 years of age, primary caregivers responded to questions about smoking in the household.
-To encapsulate early life postnatal secondhand smoke exposure, we created a binary variable for postnatal exposure at ages 1 or 5.
+To encapsulate early life postnatal secondhand smoke exposure, we created a binary variable for postnatal exposure at ages 1 and/or 5.
 To encapsulate recent postnatal smoke exposure to the age 9 and 15 interview where DNA methylation was collected, we used a categorical variable for packs per day (no smoking, less than one pack/day, one or more packs/day) in the month prior to each interview.
@@
 To ensure we measured the direct effect of prenatal exposure to cigarette smoke, rather than the effect of personal postnatal smoking we excluded children who reported own smoking.
-At ages 9 children were asked if they had ever smoked a cigarette or used tobacco (yes/no) and at age 15 they were asked if they had ever smoked an entire cigarette (yes/no).
+At age 9 children were asked if they had ever smoked a cigarette or used tobacco (yes/no), and at age 15 they were asked if they had ever smoked an entire cigarette (yes/no).
 
@@
 Child genetic ancestry was calculated.
-Grants R01 HD36916, R01 HD073352, and R01 HD076592 provided support for the collection, assay, and analysis of the genetic data in the cohort.
-Principal components (PC) of child genetic ancestry were calculated from genetic data using Illumina PsychChip_v1-1 with child saliva samples.
+Principal components (PC) of child genetic ancestry were calculated from genetic data using Illumina PsychChip with child saliva samples.
 Genetic ancestry was assigned by comparing PC loadings to 1000 Genomes super population clusters.
@@
 
-Salivary samples from the children were collected at ages nine and fifteen using the Oragene•DNA sample collection kit (DNA Genotek Inc., Ontario).
+Salivary samples from the children were collected at ages 9 and 15 using the Oragene DNA sample collection kit (DNA Genotek Inc., Ontario).
 Saliva DNA was extracted manually using DNA Genotek's purification protocol using prepIT L2P.
 DNA was bisulfite treated and cleaned using the EZ DNA Methylation kit (Zymo Research, California).
-Samples were randomized and plated across slides by demographic characteristics.
+Age 9 and 15 samples were run on the same slide to minimize technical variation, and otherwise samples were randomized.
 Saliva DNA methylation was measured using the Illumina HumanMethylation 450k BeadArray [@Sandoval2011] and imaged using the Illumina iScan system.
-All samples were run in a single batch to minimize technical variability.
+The bead arrays were processed at the Genome Sciences Core Facility at the Pennsylvania University College of Medicine (Hershey, PA).
 
-DNA methylation image data were processed in R statistical software (3.5) using the minfi package [@Aryee2014].
+DNA methylation image data were processed in R statistical software (4.1) using the Enmix package [@Xu2021a].
 We used the February 2022 data freeze of the Fragile Families and Child Wellbeing DNA methylation data.
-The red and green image pairs (n=1811) were read into R and the minfi *preprocessNoob* function was used to normalize dye bias and apply background correction.
+The red and green image pairs (n=1811) were read into R and the ENmix preprocessENmix and rcp functions were used to normalize dye bias, apply background correction and adjust for probe-type bias.
 Further quality control was applied using the ewastools packages [@Heiss2018].
@@
 
-To compare the utility of polymethylation scores relative to single site biomarkers, we selected the top DNA metyhlation site from the Joubert meta-analysis, *AHRR:* cg05575921.
-As a sensitivity analyses, we also selected four other *a priori* DNA methylation sites which were identified from meta-analyses of prenatal maternal smoking and DNA methylation in children's cord and peripheral blood samples: *GFI1*:cg14179389, *CYP1A1*: cg05549655, *MYO1G*: cg22132788, and *MYO1G*:cg04180046 [@Joubert:2016ue; @wiklund2019].
+To compare the utility of polymethylation scores relative to single site biomarkers, we selected the top DNA methylation site from the 2016 Joubert meta-analysis, *AHRR:* cg05575921.
+As a sensitivity analyses, we also selected four other *a priori* DNA methylation sites which were identified from meta-analyses of prenatal maternal smoking and DNA methylation in children's cord and peripheral blood samples: *GFI1*: cg14179389, *CYP1A1*: cg05549655, *MYO1G*: cg22132788, and *MYO1G*: cg04180046 [@Joubert:2016ue; @wiklund2019].
 
@@
 We also excluded samples that were missing genetic data.
-Age 9 samples from children who reported ever smoking a cigarette or using tobacco age 9 or were missing own-smoking data were excluded.
+Age 9 samples from children who reported ever smoking a cigarette or using tobacco at age 9 or were missing own-smoking data were excluded.
 Age 15 samples from children who reported smoking a cigarette at age 9 or at age 15 or were missing own-smoking data were excluded.
@@
 We compared the distribution of included to excluded samples using the Welch's t-test for continuous variables and the Chi-Square test for categorical variables.
-We report the correlation between continuous measures of DNA methylation (including polymethylation scores, single sites, mean global DNA methylation and estimated cell type proportions) using the Pearson correlation coefficient .
+We report the correlation between continuous measures of DNA methylation (including polymethylation scores, single sites, mean global DNA methylation and estimated cell type proportions) using the Pearson correlation coefficient.
 We report the correlation of these DNA methylation measures between visits using the Pearson correlation coefficient and scatter plots.
@@
 
-To assess the association between our primary exposure variable of prenatal maternal smoking and salivary DNA methylation, we performed age-stratified multivariable linear regressions with the primary polymethlation score as our outcome.
+To assess the association between our primary exposure variable of prenatal maternal smoking and salivary DNA methylation, we performed age-stratified, multivariable linear regressions with the primary polymethlation score as our outcome.
 To facilitate interpretation, we z-score standardized the polymethylation score.
-We adjusted for an indicator variable for maternal residence in Chicago, Detroit or Toledo (over sampled cities), maternal income to poverty ratio at baseline, child sex, child age, plate from DNA methylation processing, estimated immune cell proportion from DNA methylation data, and the first two components of genetic ancestry from principal component analysis.
-While child sex, child age, plate from DNA methylation processing and immune cell proportions are not confounders (as they cannot casually effect prenatal maternal smoke exposure), these variables can strongly effect DNA methylation and so were adjusted for as precision variables.
+We adjusted for an indicator variable for maternal residence in one of the oversampled cities (Chicago, Detroit or Toledo), maternal income to poverty ratio at baseline, child sex, child age, plate from DNA methylation processing, estimated immune cell proportion from DNA methylation data, and the first two components of genetic ancestry from principal component analysis.
+While child sex, child age, plate from DNA methylation processing and immune cell proportions are not confounders (as they cannot causally effect prenatal maternal smoke exposure), these variables can strongly effect DNA methylation and so were adjusted for as precision variables.
 We reported the effect estimate for polymethylation score term and the 95% confidence interval (CI).
@@
 
-To compare the utility of the polymethylation score to a single site biomarkers, we used percent methylation at AHRR:cg05575921 as the outcome in the models described above.
+To compare the utility of the polymethylation score to a single site biomarker, we used percent methylation at AHRR: cg05575921 as the outcome in the models described above.
 To assess the specificity of the impact of prenatal maternal smoking on DNA methylation, we used mean percent global DNA methylation and the pediatric DNA methylation clock as outcome variables.
@@
 To assess the robustness of the association to choice of polymethylation score, we performed the same regression analyses described above with z-score standardized sensitivity polymethylation scores as the outcome.
-Similarly, we replaced the outcome variable of *AHRR:*cg05575921 with other *a priori* selected DNA methylation sites in sensitivity analyses.
+Similarly, we replaced the outcome variable of *AHRR:* cg05575921 with other *a priori* selected DNA methylation sites in sensitivity analyses.
 We plotted the estimates and 95% CI for each DNA methylation site along with the previously reported estimates from the literature using forest plots [@Joubert:2016ue; @wiklund2019]
@@
 To evaluate the accuracy of the DNA methylation summary measures as biomarkers of prenatal maternal smoking, we used a receiver operating curve.
-First, we regressed exposure to prenatal maternal smoking (outcome) against the DNA methylation summary measures in individual logistic regressions, while adjusting for the base model variables listed above.
-Next, we calculated receiver operating curves (ROC) and area under the curves (AUCs) using the function roc from the R library pROC version 1.18.0 [@pROC2011].
-We compared ROC curves and AUCs from models with only base model variables to those which included base model variables and the methylation summary measure using the Delong method and the function roc.test from the R library pROC version 1.18.0.
+First, we regressed exposure to prenatal maternal smoking (outcome) against the DNA methylation summary measures in individual logistic regressions, while adjusting for the base model variables from regression analyses.
+The base model variables included indicator variable for maternal residence in one of the oversampled cities (Chicago, Detroit or Toledo), maternal income to poverty ratio at baseline, child sex, child age, plate from DNA methylation processing, estimated immune cell proportion from DNA methylation data, and the first two components of genetic ancestry from principal component analysis.
+Next, we calculated receiver operating curves (ROC) and area under the curves (AUCs) using the function *roc* from the R library pROC version 1.18.0 [@pROC2011].
+We compared ROC curves and AUCs from models with only base model variables to those which included base model variables and the methylation summary measure using the Delong method and the function *roc.test* from the R library pROC version 1.18.0.
 This is appropriate as the base model variable only model is nested within the base model variable + methylation summary measures model.

### 2022-04-08 Papers/Polymethylation_scores/98_declarations.Rmd
coauthor feedback april 4

@@
 
-Many of the variables used in this analysis are publicly available in the Fragile Families dataset.
-Some of the data used in this analysis, including genetic and epigenetic data, is restricted use but available to researchers upon reasonable application.
+Many of the variables used in this analysis are publicly available in the Fragile Families and Child Wellbeing dataset at http://opr.princeton.edu/archive/restricted/Default.aspx.
+The genetic and epigenetic data is restricted use but available to researchers upon standard application.
 
@@
 
-This research was made possible through several grants (Fragile Families Core Data Collection: R01 HD036916, R25 AG053227, genetic data processing R01 HD36916, R01 HD073352, and R01 HD076592).
-FB was supported by the National Institutes of Health National Institute of Dental and Craniofacial Research (F31 DE029992).
-EB was supported by (R01 AG055406, R01 AG067592, R01 MD011716).
+This research was made possible through several grants (Fragile Families Core Data Collection: R01 HD036916, , genetic data processing R01 HD36916, R01 HD073352, R01 MD011716, and R01 HD076592).
+FB was supported by the National Institutes of Health National Institute of Dental and Craniofacial Research (F31 DE029992) and R25 AG053227.
+EB was supported by R01 AG055406, R01 AG067592, and R01 MD011716.
+LS, DN, and JF were supported by R01 HD076592, R01 MD011716.
+CM was supported by R01 AG07107, R01 MD011716, R01 HD076592.
 

### 2022-04-25 Papers/Polymethylation_scores/02_abstract.Rmd
unstrutured abstract for epigenetics

@@
 
-**Background**: Prenatal maternal smoking has negative implications for child health.
+Prenatal maternal smoking has negative implications for child health.
 DNA methylation signatures can function as biomarkers of prenatal smoke exposure.
@@
 Using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cord blood, we created polymethylation scores of saliva DNA methylation from children at ages 9 and 15 in the Fragile Families and Child Wellbeing study.
-We tested whether prenatal maternal smoking was associated with salivary polymethylation scores for prenatal smoke exposure and assessed the consistency of associations by age and participant genetic ancestry group.
-We tested the accuracy of exposure classification using area under the curve (AUC) from receiver operating curve analyses and compared to findings from single DNA methylation sites selected from previous meta-analyses.
-
-**Results**: In the full sample at age 9 (n=`r table(modeldata$childteen)[1]`), prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.
+In the full sample at age 9 (n=`r table(modeldata$childteen)[1]`), prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.
 The direction and magnitude of the association was consistent in European and African genetic ancestry samples.
@@
 Including the polymethylation score increased the AUC of base model covariates by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(6)` percentage points, while including a single candidate site in the *AHRR* gene did not (*P* val `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg055'))%>%pull(7)%>%round(.,2)`).
-
-**Conclusions**: Prenatal maternal smoking was associated with DNA methylation signatures in child saliva samples, a clinically practical tissue.
-Polymethylation scores for prenatal maternal smoking were portable across genetic ancestries and more accurate than individual DNA methylation sites.
+Polymethylation scores for prenatal maternal smoking were portable across genetic ancestries and more accurate than an individual DNA methylation site.
 DNA polymethylation scores from saliva samples could serve as robust and practical clinical biomarkers of prenatal maternal smoke exposure.

### 2022-04-25 Papers/Polymethylation_scores/02_abstract.Rmd
cut words in abstract

@@
 
-Prenatal maternal smoking has negative implications for child health.
+Prenatal maternal smoking is associated with low birthweight, neurological disorders, and asthma in exposed children.
 DNA methylation signatures can function as biomarkers of prenatal smoke exposure.
-However, the robustness of prenatal maternal smoking DNA methylation signatures across child ages, genetic ancestry groups or tissues is not clear.
+However, the robustness of these DNA methylation signatures across child ages, genetic ancestry groups or tissues is not clear.
 Using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cord blood, we created polymethylation scores of saliva DNA methylation from children at ages 9 and 15 in the Fragile Families and Child Wellbeing study.
@@
 In the full sample at age 15 (n=`r table(modeldata$childteen)[2]`), prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score, and the association was attenuated among the European and Admixed - Latin genetic ancestry samples.
-The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (AUC age 9=0.77, age 15=0.77).
+The polymethylation score classified prenatal maternal smoke exposure accurately (AUC age 9=0.77, age 15=0.77).
 Including the polymethylation score increased the AUC of base model covariates by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(6)` percentage points, while including a single candidate site in the *AHRR* gene did not (*P* val `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg055'))%>%pull(7)%>%round(.,2)`).

### 2022-04-25 Papers/Polymethylation_scores/04_background.Rmd
Betsy edits

@@
 
-In 2018, 11% of pregnant people in the United States aged 15-44 reported any past-month cigarette use [@SAMHSA2018].
 Children exposed to cigarette smoking *in utero* are more likely to have low-birth weight, negative neurodevelopmental outcomes, and asthma [@Banderali:2015vu].
-Measuring prenatal maternal smoking behavior can be challenging.
-Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua].
+Although 11% of pregnant people in the United States in 2018 reported any past-month cigarette use [@SAMHSA2018], this is considered an underestimate.
+When compared to biomarker measures, maternal prenatal smoking is underreported - presumably due to stigma [@Kvalvik:2012tk; @Dietz:2011ua].
 The gold-standard smoking biomarker, serum cotinine levels, has a half-life of nine hours in pregnant women, and is rarely available outside of birth cohort research [@Dempsey:2002wr].
-Alternative biomarkers of prenatal maternal smoke exposure with a larger window of measurement could improve research and clinical practice.
+Alternative biomarkers of prenatal maternal smoke exposure with a larger window of measurement could improve measurement of prenatal smoke exposure in older children and adults.
 
-DNA methylation is a possible biomarker to capture prenatal smoking exposure.
+DNA methylation is a possible biomarker of prenatal smoke exposure.
 Prenatal maternal smoking is associated with offspring DNA methylation signatures at birth in placental and cord blood samples, and postnatally in peripheral blood samples [@Joubert:2016ue; @Everson:2021td; @Cardenas2019; @Richmond:2018tu; @Tehranifar:2018tb].
-Another postnatal biospecimen option is saliva, which is easier to collect than blood and increasingly used in large epidemiological cohorts [@Langie2017].
+Another postnatal biospecimen option is saliva, which is easier to collect than blood and increasingly collected in large epidemiological cohorts [@Langie2017].
+However, associations of prenatal maternal smoking with salivary DNA methylation have not been measured.
 Since DNA methylation drives cell differentiation and cell type proportions differ across tissue types, it is important to verify associations in new tissues [@Bakulski2016].

### 2022-04-25 Papers/Polymethylation_scores/05_results.Rmd
Betsy edits

@@
 
-In a sample of the Fragile Families and Child Wellbeing Study, saliva DNA methylation was measured on `r m_og_methyl` samples from `r n_og_methyl` unique child participants with the Illumina 450K array.
+Among selected participants in the Fragile Families and Child Wellbeing Study, saliva DNA methylation was measured on `r m_og_methyl` samples from `r n_og_methyl` unique child participants with the Illumina 450K array.
 Complete data on covariates of interest was available on `r m_obs` samples from `r n_total` participants who were included in the analysis (Figure 1).
 There were `r n_twovisit` participants with both age 9 and age 15 DNA methylation samples.
-Excluded samples were similar to included samples, except that included samples were slightly more likely to be from children of African genetic ancestry and less likely to be from children of Admixed Latin genetic ancestry (Supplemental Table 1).
-In the included sample, 20% percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alcohol use and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use.
+Excluded participants were similar to included participants, except that included participants were slightly more likely to be of African genetic ancestry and less likely to be of Admixed Latin genetic ancestry (Supplemental Table 1).
+Among included participants, 20% percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alcohol use and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use.
 The mean income to poverty ratio of mothers at birth was `r round(mean(modeldata_wide$cm1inpov), 2)`.
-Of the children in the analytic sample, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African genetic ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Admixed ancestry - Latin heritage']`% were of Admixed Latin genetic ancestry.
+Of the included children, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African genetic ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Admixed ancestry - Latin heritage']`% were of Admixed Latin genetic ancestry.
 
- All of the polymethylation scores for prenatal smoke exposure were correlated (Supplemental Figure 1).
+ All of the polymethylation scores for prenatal smoke exposure were correlated with each other (Supplemental Figure 1).
 Our primary polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $r$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value `r my_p_val(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value)`).
@@
 In addition to cg055975921 in the *AHHR* gene, we tested the association of four other *a priori* DNA methylation sites identified from previous meta-analyses.
-We replicated the association of prenatal maternal smoking with these DNA methylation at GFI1: cg14179389, CYP1A1: cg05549655, MYO1G: cg22132788, and MYO1G: cg04180046  (Supplemental Figure 8).
+We replicated the association of prenatal maternal smoking with DNA methylation at GFI1: cg14179389, CYP1A1: cg05549655, MYO1G: cg22132788, and MYO1G: cg04180046  (Supplemental Figure 8).
 
@@
 Including DNA methylation at *AHRR*: cg0557591 with base model covariates did not improve the classification compared to using base model covariates alone  (Supplemental Tables 4 and 5).
-Classification accuracy when including the polymethylation score from btoh age 9 and 15 was very similar to classification accuracy when using the polymethylation score from age 9 only (Figure 4B).
+Classification accuracy when including the polymethylation score from both age 9 and 15 was very similar to classification accuracy when using the polymethylation score from age 9 only (Figure 4B).
 

### 2022-04-25 Papers/Polymethylation_scores/08_methods.Rmd
Betsy edits

@@
 Participants were excluded at the baseline wave on the following criteria: the parents who planned to place the child for adoption, the father was deceased, the parents did not speak English or Spanish well enough to be interviewed, the mother or baby were too ill to complete the interview, or the baby died before the interview could take place.
-Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; an age 22 follow up is ongoing.
+Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; at age 22 follow up is ongoing.
 Across certain child ages, assessments included medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher and child.
@@
 
-Prenatal maternal substance use variables were derived from maternal self-report questionnaire data at baseline (child's birth).
-At baseline, mothers answered categorical questions about their prenatal smoking, drug and alcohol use.
-For maternal prenatal smoking, mother's were asked: "During your pregnancy, how many cigarettes did you smoke?".
+At time of child's birth, mothers answered categorical questions about their prenatal smoking, drug and alcohol use.
+For maternal prenatal smoking, mothers were asked: "During your pregnancy, how many cigarettes did you smoke?".
 Responses were categorical: 2 or more packs a day, 1 or more but less than 2 packs per day, less than 1 pack a day or none.
@@
 
-In sensitivity analyses we controlled for other types of prenatal exposures.
 Mothers were also asked how frequently they drank during their pregnancy.
@@
 
-In sensitivity analyses, we controlled for postnatal second-hand smoke exposure.
 When the child was 1, 5, 9 and 15 years of age, primary caregivers responded to questions about smoking in the household.
@@
 
-To ensure we measured the direct effect of prenatal exposure to cigarette smoke, rather than the effect of personal postnatal smoking we excluded children who reported own smoking.
 At age 9 children were asked if they had ever smoked a cigarette or used tobacco (yes/no), and at age 15 they were asked if they had ever smoked an entire cigarette (yes/no).
@@
 Our primary outcome variable was a polymethylation score for prenatal maternal smoke exposure.
-From an independent meta-analysis of prenatal smoke exposure and DNA methylation (the Joubert meta-analysis), we extracted the regression coefficients of 6,074 DNA methylation sites associated with prenatal maternal smoking at a false discovery ratio corrected $P$ value \<0.05 [@Joubert:2016ue].
+From an independent meta-analysis of prenatal smoke exposure and DNA methylation (the 2016 Joubert meta-analysis), we extracted the regression coefficients of 6,074 DNA methylation sites associated with prenatal maternal smoking at a false discovery ratio corrected $P$ value \<0.05 [@Joubert:2016ue].
 We mean-centered the DNA methylation beta values in our study, weighted them by the independent regression coefficients and took the sum.
@@
 
-To assess the robustness of polymethylation scores to the source of coefficients for weights, we generated two sensitivity polymethylation scores from the Joubert meta-analysis [@Joubert:2016ue].
+To assess the robustness of polymethylation scores to the source of coefficients for weights, we generated two sensitivity polymethylation scores from the 2016 Joubert meta-analysis [@Joubert:2016ue].
 First, we used the regression of 568 cord blood DNA methylation sites associated with prenatal maternal smoking at a strict Bonferroni threshold, when analyses were not adjusted for cell proportions [@Joubert:2016ue].
@@
 We also excluded samples that were missing genetic data.
+To ensure we measured the direct effect of prenatal exposure to cigarette smoke, rather than the effect of personal postnatal smoking we excluded children who reported own smoking.
 Age 9 samples from children who reported ever smoking a cigarette or using tobacco at age 9 or were missing own-smoking data were excluded.
 Age 15 samples from children who reported smoking a cigarette at age 9 or at age 15 or were missing own-smoking data were excluded.
-Children who were missing a response to the question at age 9 but answered that they had never smoked a whole cigarette at age 15 were kept in the sample.
+Children who were missing a response to the question at age 9 but answered that they had never smoked a whole cigarette at age 15 were kept in the sample (Figure 1).
 

### 2022-04-29 Papers/Polymethylation_scores/02_abstract.Rmd
Kelly comments and feedback 04/25/222

@@
 DNA methylation signatures can function as biomarkers of prenatal smoke exposure.
-However, the robustness of these DNA methylation signatures across child ages, genetic ancestry groups or tissues is not clear.
+However, the robustness of these DNA methylation signatures across child ages, genetic ancestry groups, or tissues is not clear.
 Using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cord blood, we created polymethylation scores of saliva DNA methylation from children at ages 9 and 15 in the Fragile Families and Child Wellbeing study.
@@
 The polymethylation score classified prenatal maternal smoke exposure accurately (AUC age 9=0.77, age 15=0.77).
-Including the polymethylation score increased the AUC of base model covariates by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(6)` percentage points, while including a single candidate site in the *AHRR* gene did not (*P* val `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg055'))%>%pull(7)%>%round(.,2)`).
+Including the polymethylation score increased the AUC of base model covariates by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(6)` percentage points, while including a single candidate site in the *AHRR* gene did not (*P*-value=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg055'))%>%pull(7)%>%round(.,2)`).
 Polymethylation scores for prenatal maternal smoking were portable across genetic ancestries and more accurate than an individual DNA methylation site.

### 2022-04-29 Papers/Polymethylation_scores/04_background.Rmd
Kelly comments and feedback 04/25/222

@@
 Children exposed to cigarette smoking *in utero* are more likely to have low-birth weight, negative neurodevelopmental outcomes, and asthma [@Banderali:2015vu].
-Although 11% of pregnant people in the United States in 2018 reported any past-month cigarette use [@SAMHSA2018], this is considered an underestimate.
+Although 11% of US pregnant people in 2018 reported any past-month cigarette use [@SAMHSA2018], this is considered an underestimate.
 When compared to biomarker measures, maternal prenatal smoking is underreported - presumably due to stigma [@Kvalvik:2012tk; @Dietz:2011ua].
-The gold-standard smoking biomarker, serum cotinine levels, has a half-life of nine hours in pregnant women, and is rarely available outside of birth cohort research [@Dempsey:2002wr].
+The gold-standard smoking biomarker, serum cotinine levels, has a half-life of nine hours in pregnant people, and is rarely available outside of birth cohort research [@Dempsey:2002wr].
 Alternative biomarkers of prenatal maternal smoke exposure with a larger window of measurement could improve measurement of prenatal smoke exposure in older children and adults.
@@
 DNA methylation is a possible biomarker of prenatal smoke exposure.
-Prenatal maternal smoking is associated with offspring DNA methylation signatures at birth in placental and cord blood samples, and postnatally in peripheral blood samples [@Joubert:2016ue; @Everson:2021td; @Cardenas2019; @Richmond:2018tu; @Tehranifar:2018tb].
+Smoking during pregnancy is associated with offspring DNA methylation signatures at birth in placental and cord blood samples, and postnatally in peripheral blood samples [@Joubert:2016ue; @Everson:2021td; @Cardenas2019; @Richmond:2018tu; @Tehranifar:2018tb].
 Another postnatal biospecimen option is saliva, which is easier to collect than blood and increasingly collected in large epidemiological cohorts [@Langie2017].
-However, associations of prenatal maternal smoking with salivary DNA methylation have not been measured.
-Since DNA methylation drives cell differentiation and cell type proportions differ across tissue types, it is important to verify associations in new tissues [@Bakulski2016].
+However, associations of pregnancy smoking with offspring salivary DNA methylation have not been measured.
+Since DNA methylation contributes to cell differentiation and cell type proportions differ across tissue types, it is important to verify associations in new tissues [@Bakulski2016].
 In addition, most studies of prenatal maternal smoking and childhood DNA methylation have been conducted in primarily European genetic ancestry populations [@Joubert:2016ue; @Ladd-Acosta:2016aa; @Tehranifar:2018tb; @Zhang2018; @Gonseth2016; @Neophytou:2019ty], and it will be important to test generalizability to other groups.
-Few studies have assessed multiple genetic ancestry groups in the same cohort and no study has evaluated associations between maternal pregnancy cigarette smoking and salivary DNA methylation in children.
+Few studies have assessed multiple genetic ancestry groups in the same cohort and no study has evaluated associations between pregnancy cigarette smoking and salivary DNA methylation in children.
 
-Using the findings from large-scale association testing in independent populations, DNA methylation-based biomarkers of prenatal maternal smoking can be developed and assessed by evaluating their classification accuracy.
+Using the findings from large-scale association testing in independent populations, DNA methylation-based exposure biomarkers can be developed and assessed by evaluating their classification accuracy.
 Biomarkers can range from single DNA methylation sites to summary measures across multiple sites, termed polymethylation scores [@Philibert2020 @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa].
 In a polymethylation score, DNA methylation sites are weighted by their effect size from a previously conducted analysis in an independent sample and then summed to a single score [@Chen2020; @Bakulski2021a].
-DNA polymethylation scores for prenatal maternal smoke exposure have performed well (area under the curves (AUCs)\>0.8) in cord blood samples from newborns and peripheral blood samples from young children [@Reese:2017ts; @Ladd-Acosta:2016aa] and fairly well in peripheral blood from middle-aged adults (AUC of 0.72 (95% confidence interval 0.69, 0.76)) [@Richmond:2018tu].
-Associations with prenatal maternal smoking change at some DNA methylation sites as children age [@Richmond2015].
-Yet few studies have assessed the longitudinal persistence of DNA-methylation biomarker accuracy.
+DNA polymethylation scores for *in utero* smoke exposure have performed well (area under the curves (AUCs)\>0.8) in cord blood samples from newborns and peripheral blood samples from young children [@Reese:2017ts; @Ladd-Acosta:2016aa].
+Polymethylation scores have also classified *in utero* smoke exposure when using from DNA methylation in peripheral blood from middle-aged adults (AUC of 0.72 (95% confidence interval 0.69, 0.76)) [@Richmond:2018tu].
+Yet few studies have assessed the longitudinal persistence of DNA-methylation-based smoking biomarker accuracy.
 
-DNA methylation differences hold promise as a biomarker for prenatal maternal smoke exposure.
+DNA methylation differences hold promise as an exposure biomarker for *in utero* cigarette smoke.
 We must understand how the signal varies across tissue, genetic ancestry, and age.
-In the Fragile Families and Child Wellbeing Study, a diverse longitudinal birth cohort, we aimed to assess salivary DNA methylation biomarkers for prenatal maternal cigarette smoking.
-We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including polymethylation scores for smoke exposure, individual *a priori* DNA methylation sites, global DNA methylation, and epigenetic clocks, at ages 9 and 15.
+In the Fragile Families and Child Wellbeing Study, a diverse longitudinal birth cohort, we aimed to assess salivary DNA methylation biomarkers for pregnancy cigarette smoking.
+We tested associations between pregnancy smoke exposure, measured at ages 9 and 15.
+We compared the magnitude and specificity of these signals to other DNA methylation measures including individual a priori DNA methylation sites, global DNA methylation, and epigenetic clocks.
 We then performed genetic ancestry- and age-stratified analyses to test the hypothesis that DNA methylation could serve as a portable and persistent biomarker of prenatal maternal smoking.

### 2022-04-29 Papers/Polymethylation_scores/05_results.Rmd
Kelly comments and feedback 04/25/222

@@
 
-Among selected participants in the Fragile Families and Child Wellbeing Study, saliva DNA methylation was measured on `r m_og_methyl` samples from `r n_og_methyl` unique child participants with the Illumina 450K array.
-Complete data on covariates of interest was available on `r m_obs` samples from `r n_total` participants who were included in the analysis (Figure 1).
+Among participants in the Fragile Families and Child Wellbeing Study, saliva DNA methylation was measured on `r m_prettyNum(og_methyl, big.mark=',')` samples from `r n_og_methyl` unique child participants with the Illumina 450K array.
+Complete data on covariates of interest was available on `r prettyNum(m_obs, ',')` samples from `r n_total` participants who were included in the analysis (Figure 1).
 There were `r n_twovisit` participants with both age 9 and age 15 DNA methylation samples.
-Excluded participants were similar to included participants, except that included participants were slightly more likely to be of African genetic ancestry and less likely to be of Admixed Latin genetic ancestry (Supplemental Table 1).
-Among included participants, 20% percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alcohol use and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use.
+Excluded participants were similar to included participants, except that included participants were slightly more likely to be of African genetic ancestry and less likely to be of Admixed-Latin genetic ancestry (Supplemental Table 1).
+In the analytic sample, 20% percent of the mothers reported any prenatal maternal smoking, `r round(table(modeldata_wide$m1g2_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal alcohol use, and `r round(table(modeldata_wide$m1g3_YesNoPreg)/nrow(modeldata_wide)*100)['Yes']`% reported prenatal drug use.
 The mean income to poverty ratio of mothers at birth was `r round(mean(modeldata_wide$cm1inpov), 2)`.
-Of the included children, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African genetic ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Admixed ancestry - Latin heritage']`% were of Admixed Latin genetic ancestry.
+Of the included children, `r round(table(modeldata_wide$cm1bsex)/nrow(modeldata_wide)*100)['1 Boy']`% were male, `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['African ancestry']`% were of African genetic ancestry, and `r round(table(modeldata_wide$ancestry)/nrow(modeldata_wide)*100)['Admixed ancestry - Latin heritage']`% were of Admixed-Latin genetic ancestry.
 
@@
 
-We compared epigenetic measures between the age 9 and age 15 visit among the `r n_twovisit` individuals with data from both visits (Supplemental Figure 2).
-The correlation across ages was strongest for the polymethylation scores for prenatal smoke exposure constructed using the Joubert meta-analysis coefficients from newborn cord blood (6074 site score with coefficients from regression with cell-type control Pearson $r$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value\<0.0001; 568 site score $r$=`r round(cor(model_twovisit$richmond568_center_Age_9, model_twovisit$richmond568_center_Age_15), 2)`, *P* value\<0.0001).
-Global DNA methylation (`r round(cor(model_twovisit$globalmethylation_Age_9, model_twovisit$globalmethylation_Age_15), 2)`, *P* value\<0.0001) and epigenetic clocks were less strongly correlated across ages (Pediatric: `r round(cor(model_twovisit$PedBE_Age_9, model_twovisit$PedBE_Age_15), 2)`; *P* value\<0.0001, GRIM: `r round(cor(model_twovisit$DNAmGrimAge_Age_9, model_twovisit$DNAmGrimAge_Age_15), 2)`; *P* value\<0.0001) The distribution of epigenetic ages shifted higher between the age 9 and age 15 visit, as expected (Supplemental Figure 3).The distribution of estimated cell-type proportions also shifted between visits, while the distribution of global methylation and polymethylation scores were more consistent (Supplemental Figure 3).
-Almost all saliva samples consisted of mostly immune cells based on estimated cell-type proportions.
-At age 9, the mean estimated immune cell proportion was `r round(mean(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)` (range: `r paste(round(range(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2), collapse=', ')`, 25% and 75% quantile: `r paste(round(quantile(modeldata_wide$Leukocytes_saliva_Age_9, c(0.25, 0.75), na.rm=T), 2), collapse = ', ')`).
-At age 15, the mean estimated immune cell proportion was `r round(mean(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2)` (range: `r paste(round(range(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2), collapse=', ')`, 25% & 75% quantile: `r paste(round(quantile(modeldata_wide$Leukocytes_saliva_Age_15, c(0.25, 0.75), na.rm=T), 2), collapse = ', ')`)
+Among the `r n_twovisit` individuals with data from both the ages 9 and 15 visits, we compared epigenetic measures over time (Supplemental Figure 2).
+The correlation across ages was strongest for the polymethylation scores for prenatal smoke exposure constructed using the 2016 Joubert meta-analysis coefficients from newborn cord blood (Pearson $r$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P*-value\<0.0001).
+Global DNA methylation ($r$=`r round(cor(model_twovisit$globalmethylation_Age_9, model_twovisit$globalmethylation_Age_15), 2)`, *P*-value\<0.0001) and epigenetic clocks were less strongly correlated across ages (Pediatric: $r$=`r round(cor(model_twovisit$PedBE_Age_9, model_twovisit$PedBE_Age_15), 2)`; *P*-value\<0.0001, GRIM: $r$=`r round(cor(model_twovisit$DNAmGrimAge_Age_9, model_twovisit$DNAmGrimAge_Age_15), 2)`; *P*-value\<0.0001) Epigenetic ages were higher at the age 15 visit than the age 9 visit, as expected (Supplemental Figure 3).The distributions of global methylation and polymethylation scores were more consistent (Supplemental Figure 3).
+Almost all saliva samples consisted of primarily immune cells, based on estimated cell-type proportions.
+At age 9, the mean estimated immune cell proportion was `r round(mean(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`% (minimum=`r round(min(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`%, first quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_9, c(0.25), na.rm=T), 2)*100`, third quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_9, c(0.75), na.rm=T), 2)*100`, maximum=`r round(max(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`).
+At age 15, the mean estimated immune cell proportion was `r paste(round(mean(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2), '%')` (minimum=`r round(min(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2)*100`%, first quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_15, c(0.25), na.rm=T), 2)*100`, third quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_15, c(0.75), na.rm=T), 2)*100`, maximum=`r round(max(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2)*100`).
 
@@
 Mothers who reported smoking during pregnancy had lower income to poverty ratios (mean=`r modeldata_people%>%filter(smkPreg_binary=='Yes')%>%summarise(mean(cm1inpov))%>%round(2)`) than those who did not (mean=`r modeldata_people%>%filter(smkPreg_binary=='No')%>%summarise(mean(cm1inpov))%>%round(2)`).
-Mothers who smoked were more likely to report prenatal alcohol use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%), prenatal drug use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%) and postnatal smoking (`r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[1, 2]%>%round(2)*100`%) than mothers who did not report prenatal maternal smoking (Table 1).
-Children of mothers who reported smoking during pregnancy were more likely to be of European genetic ancestry (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[2, 2], 2)*100`%) than children of mothers who did not (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[1, 2], 2)*100`%, *P* value=\<0.001).
-Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value\<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value\<0.001) visits (Table 1, Supplemental Table 2).
-At age 9, children exposed to prenatal maternal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`%, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
-At age 15, children exposed to prenatal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`% *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
-Epigenetic age from the pediatric and GRIM clocks and global DNA methylation did not differ between children exposed vs unexposed to prenatal maternal smoking (Figure 2).
+Mothers who smoked were more likely to report prenatal alcohol use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%), prenatal drug use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%), and postnatal smoking (`r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[1, 2]%>%round(2)*100`%) than mothers who did not report prenatal maternal smoking (Table 1).
+Children of mothers who reported smoking during pregnancy were more likely to be of European genetic ancestry (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[2, 2], 2)*100`%) than children of mothers who did not (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[1, 2], 2)*100`%, *P*-value=\<0.001).
+Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P*-value\<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P*-value\<0.001) visits (Table 1, Supplemental Table 2).
+At age 9, children exposed to prenatal maternal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 1)`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 1)`%, *P*-value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
+At age 15, children exposed to prenatal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 1)`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 1)`% *P*-value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).
+Epigenetic age from the pediatric and GRIM clocks and global DNA methylation did not differ between children exposed versus unexposed to prenatal maternal smoking (Figure 2).
 
@@
 
-The association between prenatal maternal smoking and the polymethylation score for prenatal smoke exposure was also observed in multivariable models, adjusting for base model covariates of city indicator, child sex, maternal income to poverty ratio at baseline, proportion of salivary immune cells, sample plate from DNA methylation analysis, child age and the first two principal components of genetic ancestry (Figure 3; Supplemental Table 3).
-At age 9, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score for prenatal smoke exposure.
-At age 15, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher prenatal smoke exposure polymethylation score for prenatal smoke exposure.
+The association between prenatal maternal smoking and the polymethylation score for prenatal smoke exposure was also observed in multivariable models, adjusting for base model covariates of city indicator, child sex, maternal income to poverty ratio at baseline, proportion of salivary immune cells, sample plate from DNA methylation analysis, child age, and the first two principal components of genetic ancestry (Figure 3; Supplemental Table 3).
+At age 9, prenatal maternal smoke exposure was associated with `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score for prenatal smoke exposure.
+At age 15, prenatal maternal smoke exposure was associated with `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher prenatal smoke exposure polymethylation score for prenatal smoke exposure.
 
 A consistent association was observed when stratifying by genetic ancestry.
-In the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`) at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score for prenatal smoke exposure.
+In the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`) at age 9, prenatal maternal smoking was associated with `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score for prenatal smoke exposure.
 The effect was similar among European ancestry children at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`) and consistent in direction but attenuated at age 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`).
@@
 However, they did not have higher polymethylation scores than children unexposed to both pre-and postnatal smoking (Supplemental Figure 5).
-In multivariable analyses, prenatal maternal smoking was still associated with the polymethylation scores when adjusting for postnatal smoke exposure, though effect estimates were attenuated (Age 9: `r coefCI(global_models, "childteen=='Age 9' & predictors=='secondhand smoke exposure model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` , Age 15: `r coefCI(global_models, "childteen=='Age 15' & predictors=='secondhand smoke exposure model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)`).
+In multivariable analyses, prenatal maternal smoking was still associated with the polymethylation scores when adjusting for postnatal smoke exposure, though effect estimates were attenuated (Age 9: `r coefCI(global_models, "childteen=='Age 9' & predictors=='secondhand smoke exposure model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)`, Age 15: `r coefCI(global_models, "childteen=='Age 15' & predictors=='secondhand smoke exposure model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)`).
 Prenatal maternal smoking was no longer associated with cg05575921 methylation after adjusting for postnatal smoke exposure (Age 9: `r coefCI(global_models, "childteen=='Age 9' & predictors=='secondhand smoke exposure model' & outcome=='AHRR: cg05575921'", 'smkPreg_binaryYes', 2)`, Age 15: `r coefCI(global_models, "childteen=='Age 15' & predictors=='secondhand smoke exposure model' & outcome=='AHRR: cg05575921'", 'smkPreg_binaryYes', 2)`).
@@
 
-Results were also similar when using sensitivity polymethylation scores (see Methods, Supplemental Figure 7) [@rauschert2020; @Richmond:2018tu].
+Results were also similar when using sensitivity polymethylation scores (see Methods, Supplemental Figure 7).
 
 In addition to cg055975921 in the *AHHR* gene, we tested the association of four other *a priori* DNA methylation sites identified from previous meta-analyses.
-We replicated the association of prenatal maternal smoking with DNA methylation at GFI1: cg14179389, CYP1A1: cg05549655, MYO1G: cg22132788, and MYO1G: cg04180046  (Supplemental Figure 8).
+We replicated the association of prenatal maternal smoking with DNA methylation at *GFI1*: cg14179389, *CYP1A1*: cg05549655, *MYO1G*: cg22132788, and *MYO1G*: cg04180046  (Supplemental Figure 8).
 
@@
 Next, we assessed whether including DNA methylation summary measures increased the accuracy of classifying prenatal smoke exposure over a base model consisting of child sex, maternal income to poverty ratio at baseline, child age at DNA methylation measurement, estimated immune cell proportion, plate from methylation processing, the first two genetic principal components and an indicator variable for oversampled city.
-At age 9, including the polymethylation score increased the AUC by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)`) percentage points over the nested base model (*P* value from DeLong test=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(7) %>%format(., scientific=F)`).
-At age 15 including the polymethylation score for prenatal smoke exposure increased the AUC by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)` percentage points (*P* value from DeLong test=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(7)%>%format(., scientific=F)`).
+At age 9, including the polymethylation score increased the AUC by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)`) percentage points over the nested base model (*P*-value from DeLong test=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(7) %>%format(., scientific=F)`).
+At age 15 including the polymethylation score for prenatal smoke exposure increased the AUC by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)` percentage points (*P*-value from DeLong test=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(7)%>%format(., scientific=F)`).
 Including DNA methylation at *AHRR*: cg0557591 with base model covariates did not improve the classification compared to using base model covariates alone  (Supplemental Tables 4 and 5).
@@
 
-The AUC for classification of prenatal maternal smoke exposure in the Fragile Families study was relatively similar regardless of which polymethylation score was used (Figure 5, Supplemental Table 5).
-The highest AUC was from the 568 site score using coefficients from the regression of prenatal smoke exposure and DNA methylation in newborn cord blood.
+The AUC for classification of prenatal maternal smoke exposure in the Fragile Families study was similar across all polymethylation scores (Figure 5, Supplemental Table 5).
+The highest observed AUC was from the 568 site score (using strict coefficients from the regression of prenatal smoke exposure and DNA methylation in newborn cord blood).
 Performance was similar to previous reports.
-Neither the 28-site score from a LASSO regression and the 204 site score from an elastic net regression performed as well in Fragile Families as has been reported previously in the literature (Figure 5).
+Neither the 28-site score from a LASSO regression nor the 204-site score from an elastic net regression performed as well in Fragile Families as has been reported previously in the literature (Figure 5).

### 2022-04-29 Papers/Polymethylation_scores/06_discussion.Rmd
Kelly comments and feedback 04/25/222

@@
 These signatures were specific, as global DNA methylation and epigenetic clocks were not associated with prenatal smoke exposure.
-When assessing biomarker utility, the polymethylation score for prenatal smoke exposure improved the classification of self reported prenatal maternal smoking over a base model by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'polymethylation score'))%>%pull(5)` percentage points at age 9.
-A single *a priori* DNA methylation site, *AHRR*:cg05575921, did not (percentage point increase `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg05575921'))%>%pull(5)`).
-Prenatal exposure to cigarette smoke can be measured in saliva as an alternative to peripheral blood, although classification may be further improved by tissue-specific biomarkers.
+When assessing biomarker utility, the polymethylation score for prenatal smoke exposure improved the classification of self-reported prenatal maternal smoking over a base model by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'polymethylation score'))%>%pull(5)` percentage points at age 9.
+A single *a priori* DNA methylation site, *AHRR*:cg05575921, did not ( `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg05575921'))%>%pull(5)`percentage point increase).
+Our results suggest salivary DNA methylation can be used as an alternative to peripheral blood DNA methylation for biomarkers of prenatal maternal smoking.
+A saliva-based DNA methylation biomarker may be particularly useful in cases where blood collection, processing, and storage is not possible.
+Saliva can often be collected directly by participants and saliva sample collection kits have stabilizers allowing room temperature storage prior to DNA methylation measurement.
 
@@
 Additionally, fewer than 4% of DNA methylation sites in blood and saliva samples displayed significantly different methylation values ($|\Delta \beta| >0.2$ and $p_{adj}< 0.001$) when measured on the 450K chip in samples of adults [@Langie2016] and 11-year old children [@Langie2017].
-Our results suggest salivary DNA methylation can be used as an alternative to peripheral blood DNA methylation for biomarkers of prenatal maternal smoking.
-A saliva-based DNA methylation biomarker may be particularly useful in cases where blood collection, processing, and storage is not possible.
-Saliva can often be collected directly by participants and saliva sample collection kits have stabilizers allowing room temperature storage prior to DNA methylation measurement.
 
-We advance the prenatal smoking - DNA methylation literature by evaluating the generalizability of association between prenatal maternal smoking and DNA methylation across genetic ancestry groups.
+We also advance the prenatal smoking - DNA methylation literature by evaluating the generalizability of association between prenatal maternal smoking and DNA methylation across genetic ancestry groups.
 The portability of polygenic risk scores across genetic ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Huls2020].
 Our results suggest polymethylation scores for prenatal maternal smoking are portable across genetic ancestry groups.
-At age 9, the effect estimates for prenatal maternal smoking on the polymethylation score in the European and African genetic ancestry samples were was within `r abs(coefOnly(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)-coefOnly(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2))` standard deviation and were both significant at an alpha of 0.05.
+At age 9, the effect estimates for prenatal maternal smoking on the polymethylation score in the European and African genetic ancestry samples were within `r abs(coefOnly(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)-coefOnly(local_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type') & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2))` standard deviation and were both significant at an alpha of 0.05.
 This finding is in line with a recent epigenome-wide association study in which children of both black and non-black identifying mothers exhibited similar directions of associations between prenatal smoke exposure and peripheral blood methylation at 38 DNA methylation sites [@Xu2021].
@@
 Polymethylation scores using coefficients from the Joubert meta-analysis coefficients had similar AUCs in Fragile Families saliva samples (AUCs between `r roc_table%>%ungroup()%>%filter(str_detect(methylation, 'CT|568|19'))%>%dplyr::select(auc)%>%range()%>%round(., 2)%>%paste(.
-,collapse=' and ')`) and in previous applications in peripheral blood samples from 17-, 30- and 31-year olds (AUCs between 0.69 and 0.72) [@Richmond:2018tu; @rauschert2020].
+,collapse=' and ')`) and in previous applications in peripheral blood samples from ages 17-, 30- and 31-years (AUCs between 0.69 and 0.72) [@Richmond:2018tu; @rauschert2020].
 Thus, using polymethylation scores from saliva may be an acceptable alternative when peripheral blood samples are unavailable or difficult to obtain.
@@
 This previous analysis also reported that differences in DNA methylation at *AHRR*: cg05575291 in adolescence may be partially mediated by postnatal own smoking, and we excluded children who reported postnatal own-smoking [@Richmond:2018tu].
-As the accuracy of *AHRR*: cg05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures, incorporating information across multiple sites of DNA methylation may yield a a more robust biomarker.
+As the accuracy of *AHRR*: cg05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures, incorporating information across multiple sites of DNA methylation may yield a more robust biomarker.
 

### 2022-04-29 Papers/Polymethylation_scores/08_methods.Rmd
Kelly comments and feedback 04/25/222

@@
 Participants were excluded at the baseline wave on the following criteria: the parents who planned to place the child for adoption, the father was deceased, the parents did not speak English or Spanish well enough to be interviewed, the mother or baby were too ill to complete the interview, or the baby died before the interview could take place.
-Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; at age 22 follow up is ongoing.
-Across certain child ages, assessments included medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher and child.
-At ages nine and fifteen a saliva sample was taken from the child [@Reichman:2001wv].
+Children were followed longitudinally with assessments at ages 1, 3, 5, 9, and 15; at age 22, follow up is ongoing.
+Across certain child ages, assessments included medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher, and child.
+At ages 9 and 15, a saliva sample was taken from the child [@Reichman:2001wv].
 A subsample of the Fragile Families cohort was selected for saliva DNA and DNA methylation processing [@Reichman:2001wv].
-To be eligible for the DNA methylation assessment children had to have participated and given saliva at age 9 and 15; after which a random sample of these eligible participants was then selected.
-One exception to this rule is that all participants in the Study of Adolescent Neurodevelopment (SAND) were assayed even if they only provided a sample at one time-point.
+To be eligible for the DNA methylation assessment, children had to have participated and given saliva at both ages 9 and 15, after which a random sample of these eligible participants was then selected.
+One exception to this, all participants from Detroit, Toledo, and Chicago (participants in the sub-Study of Adolescent Neurodevelopment (SAND) [@Gard2021]) were assayed for DNA methylation, even if they only provided a sample at one time-point.
 All analyses account for oversampling of these participants.
@@
 
-At time of child's birth, mothers answered categorical questions about their prenatal smoking, drug and alcohol use.
+At time of child's birth, mothers answered categorical questions about their pregnancy smoking, drug, and alcohol use.
 For maternal prenatal smoking, mothers were asked: "During your pregnancy, how many cigarettes did you smoke?".
-Responses were categorical: 2 or more packs a day, 1 or more but less than 2 packs per day, less than 1 pack a day or none.
-Few participants reported smoking a pack or more a day, thus for our primary exposure variable we created a dichotomous any versus no prenatal smoke exposure variable.
+Responses were categorical: 2 or more packs a day, 1 or more but less than 2 packs per day, less than 1 pack a day, or none.
+Few participants reported smoking a pack or more a day.
+Thus, for our primary exposure variable, we created a dichotomous (any versus no prenatal smoke exposure) variable.
 
@@
 Similarly, mothers were asked how frequently they used any other drug during their pregnancy.
-We again created a binary variable for any versus no prenatal use of other drugs.
+We again created a binary variable (any versus no prenatal use of other drugs).
 
-When the child was 1, 5, 9 and 15 years of age, primary caregivers responded to questions about smoking in the household.
-To encapsulate early life postnatal secondhand smoke exposure, we created a binary variable for postnatal exposure at ages 1 and/or 5.
-To encapsulate recent postnatal smoke exposure to the age 9 and 15 interview where DNA methylation was collected, we used a categorical variable for packs per day (no smoking, less than one pack/day, one or more packs/day) in the month prior to each interview.
+When the children were 1, 5, 9, and 15 years of age, primary caregivers responded to questions about smoking in the household.
+To encapsulate early life postnatal secondhand smoke exposure, we created a binary variable for caregiver smoking behavior at ages 1 and/or 5 (any versus no).
+To encapsulate recent postnatal smoke exposure to the age 9 and 15 interview where DNA methylation was collected, we used a caregiver categorical variable for packs per day (no smoking, less than one pack/day, one or more packs/day) in the month prior to each interview.
 
@@
 We used household income-to-poverty ratio at baseline as a proxy for socioeconomic status.
-Income-to-poverty ratio is the ratio of total household income (as self-reported by the mother) to the official poverty thresholds designated by the United States Census Bureau for the year preceding the interview.
+Income-to-poverty ratio was the ratio of total household income (as self-reported by the mother) to the official poverty thresholds designated by the United States Census Bureau for the year preceding the interview.
 
@@
 Child genetic ancestry was calculated.
-Principal components (PC) of child genetic ancestry were calculated from genetic data using Illumina PsychChip with child saliva samples.
+Principal components (PC) of child genetic ancestry were calculated from genetic data measured using PsychChip (Illumina, San Diego, California) with child saliva samples [@FFPGS2021].
 Genetic ancestry was assigned by comparing PC loadings to 1000 Genomes super population clusters.
@@
 We used the February 2022 data freeze of the Fragile Families and Child Wellbeing DNA methylation data.
-The red and green image pairs (n=1811) were read into R and the ENmix preprocessENmix and rcp functions were used to normalize dye bias, apply background correction and adjust for probe-type bias.
+The red and green image pairs (n=1,811) were read into R and the ENmix *preprocessENmix* and *rcp* functions were used to normalize dye bias, apply background correction and adjust for probe-type bias.
 Further quality control was applied using the ewastools packages [@Heiss2018].
-We dropped samples using the following criteria: if \>10% of DNA methylation sites had detection p-value \>0.01 (n=`r probe_fail`), if there was sex discordance between DNA methylation predicted sex and recorded sex (n=`r sex_mismatch`), if samples had outlier methylation values (n=`r meth_outlier`) or if two sequential samples from the same individual exhibited genetic discordance between visits (n=`r genetic_discordance`) (Figure 1).
+We dropped samples using the following criteria: if \>10% of DNA methylation sites had detection p-value \>0.01 (n=`r prettyNum(probe_fail, big.mark=',')`), if there was sex discordance between DNA methylation predicted sex and recorded sex (n=`r sex_mismatch`), if samples had outlier methylation values (n=`r prettyNum(meth_outlier, big.mark=',')`) or if two sequential samples from the same individual exhibited genetic discordance between visits (n=`r prettyNum(genetic_discordance, big.mark=',')`) (Figure 1).
 Technical replicates were removed (n=`r tech_rep`).
-DNA methylation sites were removed if they had detection p-value \>0.01 in 5% of samples (n=33376) [@Chen2013].
+DNA methylation sites were removed if they had detection p-value \>0.01 in 5% of samples (n=33,376) [@Chen2013].
 Relative proportions of immune and epithelial cell types were estimated from DNA methylation measures using a childhood saliva reference panel [@Middleton2021].
-
-We investigated several summary measures of DNA methylation, including polymethylation scores for smoke exposure, single DNA methylation sites, mean global DNA methylation, and DNA methylation clocks.
 
@@
 From an independent meta-analysis of prenatal smoke exposure and DNA methylation (the 2016 Joubert meta-analysis), we extracted the regression coefficients of 6,074 DNA methylation sites associated with prenatal maternal smoking at a false discovery ratio corrected $P$ value \<0.05 [@Joubert:2016ue].
-We mean-centered the DNA methylation beta values in our study, weighted them by the independent regression coefficients and took the sum.
 For our main analysis we used coefficients from a regression of sustained smoking exposure and DNA methylation in newborn cord blood with cell-type control.
+We mean-centered the DNA methylation beta values in our study, weighted them by the independent regression coefficients, and took the sum across sites.
+This resulted in a single smoking polymethylation score per sample.
+To facilitate interpretation, we z-score standardized the polymethylation scores.
+
+We investigated several sensitivity measures of DNA methylation, including several alternative polymethylation scores for smoke exposure, single DNA methylation sites, mean global DNA methylation, and DNA methylation clocks.
 
 To assess the robustness of polymethylation scores to the source of coefficients for weights, we generated two sensitivity polymethylation scores from the 2016 Joubert meta-analysis [@Joubert:2016ue].
-First, we used the regression of 568 cord blood DNA methylation sites associated with prenatal maternal smoking at a strict Bonferroni threshold, when analyses were not adjusted for cell proportions [@Joubert:2016ue].
-Second, we used the regression coefficients of 19 child peripheral blood DNA methylation sites associated with prenatal maternal smoking at the strict Bonferroni threshold when analyses were not adjusted for cell proportions [@Joubert:2016ue].
+First, we used the regression of cord blood DNA methylation sites associated with prenatal maternal smoking at a strict Bonferroni threshold (n=569 sites), when analyses were not adjusted for cell proportions [@Joubert:2016ue].
+Second, we used the regression coefficients of child peripheral blood DNA methylation sites associated with prenatal maternal smoking at the strict Bonferroni threshold (n=19 sites) when analyses were not adjusted for cell proportions [@Joubert:2016ue].
 
-To assess the robustness of polymethylation scores to the selection of DNA methylation sites, we generated two additional sensitivity polymethylation scores.
-One score was generated using 28 DNA methylation sites and coefficients prioritized from a LASSO regression of prenatal maternal smoking and cord blood methylation in the Norwegian Mother and Child Cohort Study [@Reese:2017ts].
-Another score was generated using 204 DNA methylation sites and coefficients prioritized from an elastic net regression of prenatal maternal smoking and DNA methylation in adolescent peripheral blood [@rauschert2020].
+To assess the robustness of polymethylation scores to the source and selection of DNA methylation sites, we generated two additional sensitivity polymethylation scores.
+One score was generated using DNA methylation sites and coefficients prioritized from a LASSO regression of prenatal maternal smoking and cord blood methylation in the Norwegian Mother and Child Cohort Study (n=28 sites) [@Reese:2017ts].
+Another score was generated using DNA methylation sites and coefficients prioritized from an elastic net regression of prenatal maternal smoking and DNA methylation in adolescent peripheral blood (n=204 sites) [@rauschert2020].
 This score was calculated using the coefficients and code provided by the authors in the R code accompanying the manuscript [@R-DNAsmokeR].
@@
 
-To assess the specificity of the impact of prenatal smoke exposure on DNA methylation, we compared the polymethylation scores to mean global and regulatory regional DNA methylation and methylation-derived age measures.
-Global DNA methylation was calculated for each sample as the mean methylation value of each sample across the cleaned probe set.
-Mean DNA methylation restricted to probes in genomic regions (CpG island, shore, shelf or open sea, as identified in the R package IlluminaHumanMethylation450kanno.ilmn12.hg19 v 0.6.0) was also calculated.
+To assess the specificity of the impact of prenatal smoke exposure on DNA methylation, we compared the polymethylation scores to mean global and regulatory regional DNA methylation and DNA methylation-derived age measures.
+Global DNA methylation was calculated for each sample as the mean methylation value of each sample across the pre-processed array.
+Mean DNA methylation for each genomic regions (CpG island, shore, shelf or open sea, as identified in the R package IlluminaHumanMethylation450kanno.ilmn12.hg19 v 0.6.0) was also calculated.
 Pediatric epigenetic age was calculated for each sample using the coefficients and methods provided by the creators (see [\<https://github.com/kobor-lab/Public-Scripts\>](https://github.com/kobor-lab/Public-Scripts){.uri}) [@mcewen2019].
@@
 
-We excluded samples missing any data on the following variables: maternal prenatal smoking, maternal prenatal alcohol use, maternal prenatal other drug use, maternal income to poverty ratio, postnatal secondhand smoke exposure variables, child sex, or child age.
+Among samples with pre-processed DNA methylation values, we excluded samples missing any data on the following variables: maternal prenatal smoking, maternal prenatal alcohol use, maternal prenatal other drug use, maternal income to poverty ratio, postnatal secondhand smoke exposure variables, child sex, or child age.
 We also excluded samples that were missing genetic data.
-To ensure we measured the direct effect of prenatal exposure to cigarette smoke, rather than the effect of personal postnatal smoking we excluded children who reported own smoking.
-Age 9 samples from children who reported ever smoking a cigarette or using tobacco at age 9 or were missing own-smoking data were excluded.
-Age 15 samples from children who reported smoking a cigarette at age 9 or at age 15 or were missing own-smoking data were excluded.
-Children who were missing a response to the question at age 9 but answered that they had never smoked a whole cigarette at age 15 were kept in the sample (Figure 1).
+To ensure we measured the direct effect of prenatal exposure to cigarette smoke, rather than the effect of personal postnatal smoking, we also excluded children who reported own smoking.
+Specifically, samples from children who reported ever smoking a cigarette or using tobacco at age 9 or were missing own-smoking data were excluded.
+Samples from children who reported smoking a cigarette at age 9 or at age 15 or were missing own-smoking data were excluded.
+Children who were missing a response to the question at age 9 but answered that they had never smoked a whole cigarette at age 15 were kept in the sample.
+Sample inclusion was visualized using a flow chart.
 
@@
 We described the distribution of continuous variables using the mean and standard deviation.
-We described the distribution of categorical variables using percents.
+We described the distribution of categorical variables using frequencies.
 We compared the distribution of included to excluded samples using the Welch's t-test for continuous variables and the Chi-Square test for categorical variables.
-We report the correlation between continuous measures of DNA methylation (including polymethylation scores, single sites, mean global DNA methylation and estimated cell type proportions) using the Pearson correlation coefficient.
-We report the correlation of these DNA methylation measures between visits using the Pearson correlation coefficient and scatter plots.
+We reported the correlation between continuous measures of DNA methylation (including polymethylation scores, single sites, mean global DNA methylation and estimated cell type proportions) using pairwise Pearson correlation coefficients.
+We reported the correlation of these DNA methylation measures between visits using pairwise Pearson correlation coefficients and scatter plots.
 We compared the distribution of variables by exposure status to prenatal maternal smoking using the Welch's t-test for continuous variables and the Chi-Square test for categorical variables.
-For select continuous DNA methylation measures we compare the distribution across prenatal smoke exposure status using violin plots with captive box plots.
+For select continuous DNA methylation measures we compared the distributions across prenatal smoke exposure status using violin plots with captive box plots.
 
@@
 
-To assess the association between our primary exposure variable of prenatal maternal smoking and salivary DNA methylation, we performed age-stratified, multivariable linear regressions with the primary polymethlation score as our outcome.
-To facilitate interpretation, we z-score standardized the polymethylation score.
-We adjusted for an indicator variable for maternal residence in one of the oversampled cities (Chicago, Detroit or Toledo), maternal income to poverty ratio at baseline, child sex, child age, plate from DNA methylation processing, estimated immune cell proportion from DNA methylation data, and the first two components of genetic ancestry from principal component analysis.
-While child sex, child age, plate from DNA methylation processing and immune cell proportions are not confounders (as they cannot causally effect prenatal maternal smoke exposure), these variables can strongly effect DNA methylation and so were adjusted for as precision variables.
+To assess the association between our primary exposure variable (prenatal maternal smoking) and primary outcome variable (salivary DNA methylation polymethylation score), we performed age-stratified, multivariable linear regressions.
+We adjusted for an indicator variable for maternal residence in one of the oversampled cities (Chicago, Detroit, or Toledo), maternal income to poverty ratio at baseline, child sex, child age, plate from DNA methylation processing, estimated immune cell proportion from DNA methylation data, and the first two components of genetic ancestry from principal component analysis.
+While child sex, child age, plate from DNA methylation processing, and immune cell proportions are not confounders (as they cannot causally effect prenatal maternal smoke exposure), these variables can strongly effect DNA methylation and so were adjusted for as precision variables.
 We reported the effect estimate for polymethylation score term and the 95% confidence interval (CI).
@@
 
-To assess the robustness of the association to choice of polymethylation score, we performed the same regression analyses described above with z-score standardized sensitivity polymethylation scores as the outcome.
+To assess the robustness of the association to the choice of polymethylation score, we performed the same regression analyses described above with sensitivity polymethylation scores as the outcome.
 Similarly, we replaced the outcome variable of *AHRR:* cg05575921 with other *a priori* selected DNA methylation sites in sensitivity analyses.
@@
 
-To evaluate the accuracy of the DNA methylation summary measures as biomarkers of prenatal maternal smoking, we used a receiver operating curve.
+To evaluate the accuracy of the DNA methylation summary measures as biomarkers of prenatal maternal smoking, we used a receiver operating curve [@robin2011].
 First, we regressed exposure to prenatal maternal smoking (outcome) against the DNA methylation summary measures in individual logistic regressions, while adjusting for the base model variables from regression analyses.
@@
 Next, we calculated receiver operating curves (ROC) and area under the curves (AUCs) using the function *roc* from the R library pROC version 1.18.0 [@pROC2011].
-We compared ROC curves and AUCs from models with only base model variables to those which included base model variables and the methylation summary measure using the Delong method and the function *roc.test* from the R library pROC version 1.18.0.
-This is appropriate as the base model variable only model is nested within the base model variable + methylation summary measures model.
+We compared ROC curves and AUCs from models with only base model variables to those which included base model variables and the methylation summary measure using the Delong method and the function *roc.test* from the pROC library.
+This is appropriate as the base model is nested within the base model variable + methylation summary measures model.
 
@@
 
-To assess the robustness of polymethylation scores classification accuracy to choice of coefficients, we used polymethylation scores calculated using different coefficient sets as predictors for the receiver operating curve analysis in sensitivity analyses.
+To assess the robustness of polymethylation scores classification accuracy to the choice of polymethylation score coefficients, we used polymethylation scores calculated using different coefficient sets as predictors for the receiver operating curve analysis in sensitivity analyses.
 
@@
 We plotted the AUC and 95% CI (when available) for each polymethylation score in our cohort and in the cohorts previously reported in the literature [@rauschert2020; @Richmond:2018tu].
-
-Code to perform all analyses is available (www.github.com/bakulskilab)

### 2022-04-29 Papers/Polymethylation_scores/98_declarations.Rmd
Kelly comments and feedback 04/25/222

@@
 
-Many of the variables used in this analysis are publicly available in the Fragile Families and Child Wellbeing dataset at http://opr.princeton.edu/archive/restricted/Default.aspx.
-The genetic and epigenetic data is restricted use but available to researchers upon standard application.
+Code to perform all analyses is available (www.github.com/bakulskilab).
+Survey data for the Fragile Families and Child Wellbeing study are publicly available (https://fragilefamilies.princeton.edu/documentation).
+DNA methylation and genetic data in the Fragile Families and Child Wellbeing study are available through the Restricted Use Contract (https://fragilefamilies.princeton.edu/restricted), upon demonstrated Institutional Review Board approval of the proposed research and data protection plan.
 
@@
 
-This research was made possible through several grants (Fragile Families Core Data Collection: R01 HD036916, , genetic data processing R01 HD36916, R01 HD073352, R01 MD011716, and R01 HD076592).
+This research was made possible through several grants (Fragile Families Core Data Collection: R01 HD036916, genetic data processing R01 HD36916, R01 HD073352, R01 MD011716, and R01 HD076592).
 FB was supported by the National Institutes of Health National Institute of Dental and Craniofacial Research (F31 DE029992) and R25 AG053227.
+KB was supported by R01 AG074887 and R01 MD013299.
 EB was supported by R01 AG055406, R01 AG067592, and R01 MD011716.

### 2022-04-29 Scripts/01_clean_phenotype_data.R

@@ -131,6 +131,8 @@ FF_labeled=FF_labeled %>% left_join(localpc%>%mutate(id=as.character(idnum))%>%d
 
 MissingPCdata<-FF_labeled %>% filter(ancestry=="Missing PC data")%>%dplyr::select(id)
 save(MissingPCdata, file = paste0(datadir, "CreatedData/MissingPC.Rdata"))
+#save FFlabeled
+save(FF_labeled, file = paste0(datadir, "CreatedData/FFlabeled.Rdata"))
 
 #######################filter to analysis subset and join with pdqc###################################
 myFF=FF_labeled%>%filter(Methyldata=='Analysis subset')%>%

### 2022-04-29 Papers/Polymethylation_scores/05_results.Rmd
fixed minor error

@@
 
-Among participants in the Fragile Families and Child Wellbeing Study, saliva DNA methylation was measured on `r m_prettyNum(og_methyl, big.mark=',')` samples from `r n_og_methyl` unique child participants with the Illumina 450K array.
+Among participants in the Fragile Families and Child Wellbeing Study, saliva DNA methylation was measured on `r prettyNum(m_og_methyl, big.mark=',')` samples from `r n_og_methyl` unique child participants with the Illumina 450K array.
 Complete data on covariates of interest was available on `r prettyNum(m_obs, ',')` samples from `r n_total` participants who were included in the analysis (Figure 1).

### 2022-05-05 Papers/Polymethylation_scores/02_abstract.Rmd
revisions may5

@@
 The polymethylation score classified prenatal maternal smoke exposure accurately (AUC age 9=0.77, age 15=0.77).
-Including the polymethylation score increased the AUC of base model covariates by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(6)` percentage points, while including a single candidate site in the *AHRR* gene did not (*P*-value=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg055'))%>%pull(7)%>%round(.,2)`).
+Including the polymethylation score increased the AUC of base model covariates by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)` percentage points, while including a single candidate site in the *AHRR* gene did not (*P*-value=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg055'))%>%pull(7)%>%round(.,2)`).
 Polymethylation scores for prenatal maternal smoking were portable across genetic ancestries and more accurate than an individual DNA methylation site.

### 2022-05-05 Papers/Polymethylation_scores/05_results.Rmd
revisions may5

@@
  All of the polymethylation scores for prenatal smoke exposure were correlated with each other (Supplemental Figure 1).
-Our primary polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $r$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value `r my_p_val(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value)`).
+Our primary polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $r$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P*-value `r my_p_val(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value)`).
 
@@
 Almost all saliva samples consisted of primarily immune cells, based on estimated cell-type proportions.
-At age 9, the mean estimated immune cell proportion was `r round(mean(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`% (minimum=`r round(min(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`%, first quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_9, c(0.25), na.rm=T), 2)*100`, third quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_9, c(0.75), na.rm=T), 2)*100`, maximum=`r round(max(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`).
-At age 15, the mean estimated immune cell proportion was `r paste(round(mean(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2), '%')` (minimum=`r round(min(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2)*100`%, first quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_15, c(0.25), na.rm=T), 2)*100`, third quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_15, c(0.75), na.rm=T), 2)*100`, maximum=`r round(max(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2)*100`).
+At age 9, the mean estimated immune cell proportion was `r round(mean(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`% (minimum=`r round(min(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`%, first quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_9, c(0.25), na.rm=T), 2)*100`%, third quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_9, c(0.75), na.rm=T), 2)*100`%, maximum=`r round(max(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`%).
+At age 15, the mean estimated immune cell proportion was `r paste(round(mean(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2)*100, '%')` (minimum=`r round(min(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2)*100`%, first quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_15, c(0.25), na.rm=T), 2)*100`%, third quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_15, c(0.75), na.rm=T), 2)*100`%, maximum=`r round(max(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2)*100`%).
 
@@
 In the European genetic ancestry sample, the association was consistent in direction at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`) and age 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`).
-In the Admixed Latin genetic ancestry sample, the association was no longer significant and not consistent in direction (Age 9: `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`; Age 15: `r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`) (Figure 3).
+In the Admixed-Latin genetic ancestry sample, the association was no longer significant and not consistent in direction (Age 9: `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`; Age 15: `r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`) (Figure 3).
 

### 2022-05-05 Papers/Polymethylation_scores/06_discussion.Rmd
revisions may5

@@
 When assessing biomarker utility, the polymethylation score for prenatal smoke exposure improved the classification of self-reported prenatal maternal smoking over a base model by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'polymethylation score'))%>%pull(5)` percentage points at age 9.
-A single *a priori* DNA methylation site, *AHRR*:cg05575921, did not ( `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg05575921'))%>%pull(5)`percentage point increase).
+A single *a priori* DNA methylation site, *AHRR*: cg05575921, did not ( `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg05575921'))%>%pull(5)`percentage point increase).
 Our results suggest salivary DNA methylation can be used as an alternative to peripheral blood DNA methylation for biomarkers of prenatal maternal smoking.

### 2022-05-05 Papers/Polymethylation_scores/99_FigureCaptions.Rmd
revisions may5

@@
 
-#### Table 1: Bivariate associations between prenatal maternal smoking (no/yes) and selected DNA methylation summary measures and important covariates among a diverse sample of `r n_total` children in the Fragile Families and Child Wellbeing study
+#### **Table 1: Bivariate associations between prenatal maternal smoking (no/yes) and selected DNA methylation summary measures and important covariates among a diverse sample of `r n_total` children in the Fragile Families and Child Wellbeing study**
 
-```{r,  eval=print_figs, include=T}
+```{r, echo=F, eval=print_figs, include=T, include=T}
 smoking_tbl
@@
 
-#### **Figure 5 - Comparison of different polymethylation scores and their accuracy at classifying prenatal smoke exposure among `r n_total` children with saliva samples in the Fragile Families and Child Wellbeing study as opposed to previously reported results in cord and peripheral blood samples
+#### **Figure 5 - Comparison of different polymethylation scores and their accuracy at classifying prenatal smoke exposure among `r n_total` children with saliva samples in the Fragile Families and Child Wellbeing study as opposed to previously reported results in cord and peripheral blood samples**
 
-```{r, echo=F, eval=print_figs, include=T,  fig.width=12}
+```{r, echo=F, eval=print_figs, warning=F, include=T,  fig.width=12}
 auc_compare

### 2022-05-05 Reports/Paper1.Rmd
report update

--- 2021-11-30 Reports/Paper1.Rmd
@@
 
-as_supp=T
-stand_alone=T
+as_supp=F
+stand_alone=F
 load_data=T
@@
 set_label(modeldata$SSnewbornCT_center)='Polymethylation score for smoke exposure'
-set_label(modeldata$pediatric)='Pediatric epigenetic clock (years)'
-set_label(modeldata$grim)='GRIM epigenetic clock (years)'
+set_label(modeldata$PedBE)='Pediatric epigenetic clock (years)'
+set_label(modeldata$DNAmGrimAge)='GRIM epigenetic clock (years)'
 set_label(modeldata$globalmethylation)='Percent global DNA methylation'
@@
 
-```{r}
+```{r set options and record n}
 n_total=modeldata %>% dplyr::select(id)%>%unique()%>%nrow()
@@
 
-all_methyl=c("globalmethylation", "pediatric", "grim", 'DNAmPACKYRS', "Epithelial.cells_saliva", "Leukocytes_saliva", "SSnewborn_center", "SSnewbornCT_center", "SSolder_center", "cg05575921")
-main_methyl=c("globalmethylation", "pediatric", "SSnewbornCT_center", "cg05575921")
+all_methyl=c("globalmethylation", "PedBE", "DNAmGrimAge", 'DNAmPACKYRS', "Epithelial.cells_saliva", "Leukocytes_saliva", "SSnewbornCT_center", "richmond568_center", "richmond19_center", "reese_center", "enscore_probs", "cg05575921")
+main_methyl=c("globalmethylation", "PedBE", "SSnewbornCT_center", "cg05575921")
 
 all_methyl_labels=modeldata%>%dplyr::select(any_of(all_methyl))%>%get_label()
-all_methyl_labels[str_detect(all_methyl_labels, 'Polymethylation', )]=paste0(rep('Score:', 3), c('Sustained smoking \n newborns', 'Sustained smoking \n newborns (w/ CT control)', 'Sustained smoking \n older children'))
+#all_methyl_labels[str_detect(all_methyl_labels, 'Polymethylation', )]=paste0(rep('Score:', 3), c('Sustained smoking \n newborns', 'Sustained smoking \n newborns (w/ CT control)', 'Sustained smoking \n older children'))
+all_methyl_labels[str_detect(all_methyl_labels, 'Polymethylation score for smoke exposure')]='6074 site polymethylation score: coefficients for sustained smoking from Joubert newborn cord blood meta-analysis with cell-type correction (mean-centered)'
 names(all_methyl_labels)=all_methyl
 
-all_methyl_levels=c("SSnewbornCT_center", "SSnewborn_center",  "SSolder_center",
-                                       "pediatric" , "grim", 'DNAmPACKYRS',  "cg05575921", 
+all_methyl_levels=c("SSnewbornCT_center", 'richmond568_center', 'richmond19_center', 
+                    'reese_center', 'enscore_probs', "cg05575921", 
+                     "PedBE" , "DNAmGrimAge", 'DNAmPACKYRS', 
                      "globalmethylation","Epithelial.cells_saliva", "Leukocytes_saliva")
@@
 
-```{r figure counter}
+```{r figure table counter}
 main_fig=1
@@
 ```{r figure 1, eval=T}
-#function
+#function to make a pretty count of n and m 
 N_M_count=function(filter_statement, statement, df){
@@
 n_fullFF<-paste0("Fragile Families cohort \n", "n=", FF_labeled %>% nrow())
-n_methylQC<-paste0("Cohort with 450K methylation data \n", "n=", methylCohort$idnum%>%unique()%>%length(), " individuals with m=", methylCohort$MethID%>%length(), " samples")
-
-n_QCloss<-paste0("Methylation data QC: ", 
-                 N_M_count("probe_fail_10==T & childteen!='M'", " with >10% of probes failing dropped, \n ", pdqc_all),
-                 N_M_count("sex_new!=predicted_sex & probe_fail_10==F & childteen!='M'", 
-                           "  with discordant sex samples dropped, \n ", pdqc_all),
-                 N_M_count("predicted_sex_outlier==T & sex_new==predicted_sex & probe_fail_10==F & childteen!='M'", 
-                           ' with abnormal sex chromosome intensity dropped & \n ', pdqc_all),
-                 N_M_count("cut %in% c('contaminated', 'relative') & predicted_sex_outlier==F & sex_new==predicted_sex & probe_fail_10==F & childteen!='M'", 
-          ' with contamination/mislabels & \n ', pdqc_all),
-                 length(reps_remove), " duplicate samples dropped")
+
+#original number of samples and individuals with methylation data
+all_samples=readRDS(file.path(datadir, 'OGData', 'pd_qc.rds'))
+all_samples_nomoms=all_samples %>%filter(childteen!='M')
+m_og_methyl=nrow(all_samples_nomoms) # number of samples - 12 mother samples
+n_og_methyl=length(unique(all_samples_nomoms$idnum)) # number of individuals 
+n_methylQC<-glue::glue("Cohort with 450K methylation data \n  n={n_og_methyl} individuals with m={m_og_methyl} samples")
+
+#These numbers come from Jonahs preprocessing document, I save them as variables to ensure correspondence between Figure 1 and the document text
+probe_fail=34; sex_mismatch=11; meth_outlier=6; genetic_discordance=27; tech_rep=49;idat_irregularity=1
+
+n_QCloss<-glue::glue("Methylation data QC: ", 
+                 '{idat_irregularity} sample with idat file irregularity
+                 {probe_fail} samples with probe failure percentage>10%
+                {sex_mismatch} samples with predicted sex mismatch 
+                 {meth_outlier} samples with outlier methylation values
+                 {genetic_discordance} samples with genetic discordance
+                 {tech_rep} technical replicates')
 
@@
                                      N_M_count("!is.na(smkPreg_binary) & !is.na(m1g2_YesNoPreg) & !is.na(m1g3_YesNoPreg) & !is.na(cm1inpov) &
-                                               PostnatalMaternalSmokingAny=='Missing'", " with missing postnatal maternal smoking data \n ", myFF), 
+                                               PostnatalMaternalSmokingAny=='Missing'", " with missing early life postnatal smoke exposure data \n ", myFF), 
                                      N_M_count("!is.na(smkPreg_binary) & !is.na(m1g2_YesNoPreg) & !is.na(m1g3_YesNoPreg) & !is.na(cm1inpov) &
                                                PostnatalMaternalSmokingAny!='Missing' & SmkAtVisitPastmonth=='Missing'", 
-                                               " with missing past month maternal/primary care giver smoking packs/day", myFF))
+                                               " with missing past month primary care giver smoking packs/day", myFF))
 
@@
                                     " with missing child sex data \n ", maternaldata), 
-                          N_M_count("is.na(ancestry) | ancestry=='Missing PC data' & !is.na(cm1bsex) & is.na(ChildAgeComposite)", 
+                          N_M_count("!is.na(ancestry) & ancestry!='Missing PC data' & !is.na(cm1bsex) & is.na(ChildAgeComposite)", 
                                     " with missing child age at sample \n ", maternaldata))
@@
                           " individuals with a 15 year visit")
-n_TeenSmkexclusions<-paste0('n=', childdata %>% filter((k6d40=="1 Yes" | is.na(k6d40) | k5f1l=='1 yes') & childteen=='T') %>% nrow,
+n_TeenSmkexclusions<-paste0('n=', childdata %>% filter((k6d40=="1 Yes" | is.na(k6d40) | k5f1l=='1 yes') & childteen=='T') %>% nrow(),
                             " individuals smoking \n or missing focal child smoking data \n at 15 year visit")
@@
 
-if(childdata %>% filter(((k5f1l=="2 no" | (is.na(k5f1l) & k6d40=='2 No')) & childteen=='C') | (childteen=='T' & k6d40=="2 No" & (k5f1l=='2 no' |  is.na(k5f1l))))%>%nrow()!=nrow(modeldata)){stop('filter N and completecase data set N do not match')}
+#if(childdata %>% filter(((k5f1l=="2 no" | (is.na(k5f1l) & k6d40=='2 No')) & childteen=='C') | (childteen=='T' & k6d40=="2 No" & (k5f1l=='2 no' |  is.na(k5f1l))))%>%nrow()!=nrow(modeldata)){stop('filter N and completecase data set N do not match')}
 
@@
 set_label(myFF$MethylData2)<-'Exclusions based on missing data'
-myFF=left_join(myFF, methyldata)%>%mutate(idnum=as.character(id))
+myFF=left_join(myFF, methyldata)%>%mutate(idnum=as.character(idnum))
 myFF=myFF%>%mutate(across(c(PostnatalMaternalSmokingAny, SmkAtVisitPastmonth), ~ifelse(.x=='Missing', 
@@
 exclusion_tbl=myFF %>% mutate(childteen=case_when(childteen=='C'~'Age 9', childteen=='T'~'Age 15'))%>%
-    dplyr::select(smkPreg_binary, childteen, MethylData2, SSnewbornCT_center, cg05575921, pediatric, globalmethylation, Leukocytes_saliva, Epithelial.cells_saliva, ancestry, cm1bsex, city_binary,  m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,SmkAtVisitPastmonth )%>%
+    dplyr::select(smkPreg_binary, childteen, MethylData2, SSnewbornCT_center,  richmond568_center, reese_center, enscore_probs, cg05575921, PedBE, globalmethylation, Leukocytes_saliva, Epithelial.cells_saliva, ancestry, cm1bsex, city_binary, k6d40, k5f1l, m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,SmkAtVisitPastmonth )%>%
   mutate(smkPreg_binary=factor(smkPreg_binary, levels=c('No', 'Yes'), labels=c('Not exposed', 'Smoke exposed')), 
@@
 
+gtsave(as_gt(exclusion_tbl), file.path(outdir, paste0('SupplementalTable', sup_tab, '.rtf')))
+
 sup_tab=sup_tab+1
@@
 #labels
-colnames(res2)=all_methyl_labels
-rownames(res2)=all_methyl_labels
+colnames(res2)=str_wrap(all_methyl_labels, width=70)
+rownames(res2)=str_wrap(all_methyl_labels, width=70)
 
 pdf(file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=12)
-corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8, number.cex= 7/ncol(res2))
+corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.5, number.cex= 7/ncol(res2))
 dev.off()
@@
 Blue lines are simple linear regression lines of best fit, with Pearson rho and associated $P$ value shown.
-From top-left: Polymethylation score for prenatal maternal smoking, score created using coefficients from a regression of prenatal maternal smoking and DNA methylation in newborn cord blood, incorporating cell-type control.
-Polymethylation score for prenatal maternal smoking using coefficients from a regression of prenatal maternal smoking and DNA methylation in newborn cord blood, no cell-type control.
-Polymethylation score calculated using regression coefficients from a regression of prenatal maternal smoking and DNA methylation in older children's peripheral blood samples.
+From top-left: Polymethylation score for prenatal maternal smoking, score created using coefficients from a regression of prenatal maternal smoking and DNA methylation in newborn cord blood, incorporating cell-type control (Joubert et al.
+6074 DNA methylation sites).
+Polymethylation score for prenatal maternal smoking using coefficients from a regression of prenatal maternal smoking and DNA methylation in newborn cord blood, no cell-type control (Joubert et al.
+568 DNA methylation sites).
+Polymethylation score calculated using regression coefficients from a regression of prenatal maternal smoking and DNA methylation in older children's peripheral blood samples (Joubert et al.
+19 DNA methylation sites).
+Polymethylation score calculated using regression coefficients from a LASSO regression of prenatal maternal smoking and DNA methylation in newborn cord blood samples (Reese et al.
+28 DNA methylation sites).
+Polymethylation score calculated using coefficients from an elastic net regression of prenatal maternal smoking and DNA methylation in adolescents peripheral blood (Rauschert et al.
+204 site score).
+Percent DNA methylation at cg05575921 in AHRR.
 Pediatric epigenetic clock (years).
 GRIM epigenetic clock (years).
-Percent DNA methylation at cg05575921 in AHRR.
+GRIM epigenetic clock pack years component.
 Percent global DNA methylation (mean across all DNA methylation sites in cleaned probe set).
@@
   scale_x_continuous('Age 9 score')+scale_y_continuous('Age 15 score')+
-  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels), ncol=3)
+  facet_wrap(~Methyl_factor, scales='free', labeller = as_labeller(all_methyl_labels, default=label_wrap_gen(60)), ncol=3)+
+  theme(strip.text = element_text(size=5))
 
@@
   geom_point(alpha=0)+geom_line(alpha=0, aes(group=idnum))+geom_smooth(aes(group=1), method='lm', color='black')+
-  facet_wrap(~Methyl_factor, scales='free', labeller = labeller(Methyl_factor=all_methyl_labels), ncol=3)+
-  theme(legend.position = 'bottom')
+  facet_wrap(~Methyl_factor, scales='free', labeller = as_labeller(all_methyl_labels, default=label_wrap_gen(60)), ncol=3)+
+  theme(legend.position = 'bottom')+
+  theme(strip.text = element_text(size=5))
 
@@
 #add line for mean ? 
-outcome_labels=c('Global methylation', 'Pediatric clock', 'AHRR: cg05575921', 'Polymethylation score \n (newborn, mean centered)', 'Polymethylation score  \n (older children, mean centered')
-names(outcome_labels)=c('globalmethylation', 'pediatric', 'cg05575921', 'SSnewbornCT_center','SSolder_center')
+outcome_labels=c('Global methylation', 'Pediatric clock', 'AHRR: cg05575921', '6074 site polymethylation score \n (newborn, mean centered)', 'Polymethylation score  \n (older children, mean centered')
+names(outcome_labels)=c('globalmethylation', 'PedBE', 'cg05575921', 'SSnewbornCT_center','SSolder_center')
 
@@
   stat_compare_means(label='..p.signif..', label.y.npc = 0.88)+
-    facet_wrap(~Methyl_factor, scales='free',labeller = labeller(Methyl_factor=all_methyl_labels))+
+    facet_wrap(~Methyl_factor, scales='free',labeller = as_labeller(all_methyl_labels, default=label_wrap_gen(40)))+
     scale_x_discrete('Visit', labels=c('Age 9', 'Age 15'))+
@@
 smoking_tbl=modeldata %>% 
-  dplyr::select(smkPreg_binary, childteen, SSnewbornCT_center, cg05575921, pediatric, globalmethylation, Leukocytes_saliva, Epithelial.cells_saliva, ancestry, cm1bsex,  city_binary, m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,SmkAtVisitPastmonth )%>%
+  dplyr::select(smkPreg_binary, childteen, SSnewbornCT_center, cg05575921, PedBE, globalmethylation, Leukocytes_saliva, Epithelial.cells_saliva, ancestry, cm1bsex,  city_binary, m1g2_YesNoPreg, m1g3_YesNoPreg, cm1inpov, PostnatalMaternalSmokingAny,SmkAtVisitPastmonth )%>%
   mutate(smkPreg_binary=factor(smkPreg_binary, levels=c('No', 'Yes'), labels=c('Not exposed', 'Smoke exposed')), 
@@
 
+gtsave(smoking_tbl, file.path(outdir, paste0('Table', main_tab, '.rtf')))
+
 main_tab=main_tab+1
@@
   dplyr::select(smkPreg_binary, globalmethylation, Island_mean, OpenSea_mean, Shelf_mean, Shore_mean, 
-                pediatric, grim, DNAmPACKYRS, Epithelial.cells_saliva, Leukocytes_saliva,
-                SSnewbornCT_center, SSnewborn_center, SSolder_center, anynewborn_center,
-                matches('cg'),  childteen)%>%
+                PedBE, DNAmGrimAge, DNAmPACKYRS, Epithelial.cells_saliva, Leukocytes_saliva,
+                SSnewbornCT_center, 
+                richmond568_center, richmond19_center, reese_center, enscore_probs,
+                matches('cg'),  childteen, -PCGrimAge)%>%
     mutate(smkPreg_binary=factor(smkPreg_binary, levels=c('No', 'Yes'), labels=c('Not exposed', 'Smoke exposed')))%>%
@@
 
+gtsave(as_gt(smoking_tbl_extendedMeth), file.path(outdir, paste0('SupplementalTable', sup_tab, '.rtf')))
+
 sup_tab=sup_tab+1
@@
 ```{r main figure 3 forest plot}
-my_outcomes=c('Global methylation', 'Pediatric clock', 'AHRR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)')
-
-outcome_labels=c('Percent global  \n DNA methylation  (percent)', 'Pediatric epigenetic \n clock (years)', 'Percent DNA methylation of \n cg05575921 in AHRR (percent)',  'Polymethylation score for prenatal maternal \n smoke exposure (standard deviations)')
-
-all_models=rbind(local_models, global_models%>%mutate(ancestry='Unstratified'))
+my_outcomes=c('Global methylation', 'Pediatric clock', 'AHRR: cg05575921', '6074 site polymethylation score: coefficients for sustained smoking from Joubert newborn cordblood meta-analysis, w/ cell-type control - PMC4833289 (/n mean-centered (coefficients) & z-score standardized (score))')
+
+outcome_labels=c('Percent global  \n DNA methylation  (percent)', 'Pediatric epigenetic \n clock (years)', 'Percent DNA methylation of \n cg05575921 in AHRR (percent)',  '6074 site polymethylation score for prenatal smoke exposure with cell-type corrected coefficients (standard deviations)')
+
+all_models=rbind(local_models, global_models%>%mutate(ancestry='Unstratified'))%>%
+  mutate(childteen=factor(childteen, levels=c('Age 15', 'Age 9')))
 
@@
   unnest(tidied)%>%
-  filter(outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & term=='smkPreg_binaryYes'& predictors=='base model')%>%mutate(outcome_f=outcome)
+  filter(str_detect(outcome, 'cell-type control') & term=='smkPreg_binaryYes'& predictors=='base model')%>%mutate(outcome_f=outcome)
 
@@
     mutate(outcome_f=factor(outcome, levels=rev(my_outcomes)))%>%
-  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=ancestry))+
+  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, levels(age)), y=ancestry))+
   geom_point(position = position_dodge(width = 1))+
   geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
-  facet_grid(.~outcome_f, scales='free_x', labeller = labeller(outcome_f=outcome_labels))+
+  facet_grid(.~outcome_f, scales='free_x', labeller = as_labeller(outcome_labels, default=label_wrap_gen(50)))+
   geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
@@
   scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
-  scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'), labels=c('European genetic ancestry', 'Hispanic genetic ancestry', 'African genetic ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))
+  scale_y_discrete('', limits=c('European ancestry', 'Admixed ancestry - Latin heritage', 'African ancestry', 'Unstratified'), labels=str_wrap(c('European genetic ancestry', 'Admixed genetic ancestry - Latin heritage', 'African genetic ancestry', 'Unstratified'), 30))+scale_color_manual('', values=c('black', 'darkgrey'), breaks=c('Age 9', 'Age 15'))
 
@@
     mutate(outcome_f=factor(outcome, levels=rev(my_outcomes)))%>%
-    ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=ancestry))+
+    ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, levels(age)), y=ancestry))+
     geom_point(position = position_dodge(width = 1))+
@@
     scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
-    scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'), labels=c('European genetic ancestry', 'Hispanic genetic ancestry', 'African genetic ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))
+    scale_y_discrete('', limits=c('European ancestry', 'Admixed ancestry - Latin heritage', 'African ancestry', 'Unstratified'), labels=c('European genetic ancestry', 'Admixed ancestry - Latin heritage', 'African genetic ancestry', 'Unstratified'))+scale_color_manual('', breaks=c('Age 9', 'Age 15'), values=c('black', 'darkgrey'))
 
@@
 
-# Supplemental Table `r sup_tab`: Coefficients for multivariable models of prenatal maternal smoke exposure regressed on selected DNA methylation summary variables (outcomes) among children in the Fragile Families and Child Wellbeing study using surrogate variables
-
-```{r supplemental table surrogate variable table}
-test=rbind(global_models, longitudinal_global_models %>% mutate(childteen='Longitudinal'))%>%
-  filter(predictors=='surrogate variable models')%>%
-  filter(outcome %in% my_outcomes)%>%
-  mutate(childteen=factor(case_when(childteen=='Age 9'~'Age 9 models', 
-                             childteen=='Age 15'~'Age 15 models', 
-                             childteen=='Longitudinal'~'Longitudinal models'),levels=c('Age 9 models', 'Age 15 models', 'Longitudinal models')), 
-         tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
-                                        show_single_row='smkPreg_binary', 
-                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
-
-table_list=test %>%split(f=.$childteen)%>%map(~tbl_merge(.x$tbl, 
-                                                         tab_spanner = as.character(.x$outcome)))
-
-regression_table_age_svs=tbl_stack(table_list, group_header=names(table_list))
-
-test=rbind(longitudinal_global_models %>% mutate(ancestry='Unstratified'),
-           longitudinal_local_models%>%dplyr::select(-data))%>%
-  filter(predictors=='surrogate variable models')%>%
-  filter(outcome %in% my_outcomes)%>%
-  mutate(tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
-                                        show_single_row='smkPreg_binary', 
-                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
-
-table_list=test %>%split(f=.$ancestry)%>%map(~tbl_merge(.x$tbl, 
-                                                         tab_spanner = as.character(.x$outcome)))
-
-regression_table_ancestry_svs=tbl_stack(table_list, group_header=names(table_list))
-
+# Supplemental Table `r sup_tab`: Coefficients for multivariable models of prenatal maternal smoke exposure regressed on selected DNA methylation summary variables (outcomes) among children in the Fragile Families and Child Wellbeing study
+
+```{r}
+
+coefficient_table=rbind(global_models%>%
+                          mutate(ancestry='Multiethnic sample'), 
+                        local_models)%>%unnest(tidied)%>%
+  filter(term=='smkPreg_binaryYes')%>%
+  mutate(N=map_dbl(model, nobs))%>%
+  dplyr::select(outcome, childteen, ancestry, predictors,
+                N, estimate, conf.low, conf.high, p.value)%>%
+  mutate(across(estimate:conf.high, ~signif(.x, 2)))%>%
+  mutate(p.value=signif(p.value, 5))%>%
+  arrange(outcome, childteen, ancestry, predictors)
+
+write.csv(coefficient_table, file.path(outdir, paste0('SupplementalTable', sup_tab, '.csv')))
 sup_tab=sup_tab+1
-```
-
-```{r surrogate variable table print, eval=as_supp}
-regression_table_ancestry_svs
-```
-
-# Supplemental Table `r sup_tab`: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable linear models stratified by age or in multivariable linear mixed effect models among `r n_total` US children.
-All models controlled for:  first two principal components of child genetic ancestry (recalculated within ancestry strata for stratified models), child sex, maternal income-to-poverty ratio at birth, immune cell proportion estimated from methylation data, child age
-
-```{r supplemental table age and linear mixed effect}
-#my_main_outcomes=c('AHRR: cg05575921', 'Sustained smoking polymethylation score (newborns, cell-type controlled)', 'Sustained smoking polymethylation score (older children)', 'Pediatric clock', 'Global methylation')
-
-test=rbind(global_models, longitudinal_global_models %>% mutate(childteen='Longitudinal'))%>%
-  filter(predictors=='base model')%>%
-  filter(outcome %in% my_outcomes)%>%
-  mutate(childteen=factor(case_when(childteen=='Age 9'~'Age 9 models', 
-                             childteen=='Age 15'~'Age 15 models', 
-                             childteen=='Longitudinal'~'Longitudinal models'),levels=c('Age 9 models', 'Age 15 models', 'Longitudinal models')), 
-         tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
-                                        show_single_row='smkPreg_binary', 
-                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
-
-table_list=test %>%split(f=.$childteen)%>%map(~tbl_merge(.x$tbl, 
-                                                         tab_spanner = as.character(.x$outcome)))
-
-regression_table_age=tbl_stack(table_list, group_header=names(table_list))
-
-sup_tab=sup_tab+1
-```
-
-```{r supplemental table age and mixed effect print, eval=as_supp}
-regression_table_age
-```
-
-# Supplemental Table `r sup_tab`: Associations between prenatal maternal smoking and DNA methylation summary measures in multivariable models among `r n_total` US children by ancestry using linear mixed effect models.
-All models controlled for:  first two principal components of child genetic ancestry (recalculated within ancestry strata for stratified models), child sex, maternal income-to-poverty ratio at birth, immune cell proportion estimated from methylation data, child age
-
-```{r supplemental table by ancestry using linear mixed effect models}
-test=rbind(longitudinal_global_models %>% mutate(ancestry='Unstratified'),
-           longitudinal_local_models%>%dplyr::select(-data))%>%
-  filter(predictors=='base model')%>%
-  filter(outcome %in% my_outcomes)%>%
-  mutate(tbl=map(model, ~tbl_regression(.x, include='smkPreg_binary', 
-                                        show_single_row='smkPreg_binary', 
-                                        label=list(smkPreg_binary~'Prenatal maternal smoke exposed'))%>%add_n()))
-
-table_list=test %>%split(f=.$ancestry)%>%map(~tbl_merge(.x$tbl, 
-                                                         tab_spanner = as.character(.x$outcome)))
-
-regression_table_ancestry=tbl_stack(table_list, group_header=names(table_list))
-
-sup_tab=sup_tab+1
-
-```
-
-```{r supplemental table by ancestry using lme print, eval=as_supp}
-regression_table_ancestry
+
 ```
@@
     filter(term=='smkPreg_binaryYes')%>%
-    filter(str_detect(outcome, 'smoking'))%>%
+    filter(str_detect(outcome, '28 site|19 site|568 site|204 site|cell-type'))%>%
   filter(ancestry=='Unstratified')%>%
- ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=outcome))+
+ ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=gsub('- PMC.*', '', outcome)))+
     geom_point(position = position_dodge(width = 1))+
@@
 
-ggexport(compare_pms, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=9)
+ggexport(compare_pms, filename=file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=18)
 sup_fig=sup_fig+1
@@
 
+# Main figure `r main_fig`: Area under the curve and receiver operating curve for the polymethylation score, *AHRR*:cg05575921, and global methylation
+
 ```{r main figure 4 auc}
@@
     filter(str_detect(methylation, 'SSnewbornCT|cg0557|global') | modeltype=='Base model')%>%
-  mutate(modeltype=gsub("Sustained smoking polymethylation score \\(newborns, cell-type controlled\\)", "Polymethylation score for prenatal maternal smoke exposure", modeltype))%>%
-  mutate(modeltype=factor(modeltype, levels=c('AHRR: cg05575921', 'Base model+AHRR: cg05575921', 'Global methylation', 'Base model+Global methylation',  'Polymethylation score for prenatal maternal smoke exposure', 'Base model+Polymethylation score for prenatal maternal smoke exposure')))%>%
+  mutate(modeltype=gsub(": coefficients.*)", " for prenatal smoke exposure (Joubert cell-type control coefficients)", modeltype))%>%
+  mutate(modeltype=factor(modeltype, levels=c('AHRR: cg05575921', 'Base model+AHRR: cg05575921', 'Global methylation', 'Base model+Global methylation',  '6074 site polymethylation score for prenatal smoke exposure (Joubert cell-type control coefficients)', 'Base model+6074 site polymethylation score for prenatal smoke exposure (Joubert cell-type control coefficients)')))%>%
     ggplot(aes(x=spec, y=sens, color=modeltype))+geom_line()+
@@
 ```{r supplemental table auc}
-roc_table=global_roc %>% dplyr::select(childteen, methylation, modeltype, roc, auc)
-roc_base=roc_table %>% filter(modeltype=='Base model+No methylation')%>%mutate(roc_basemodel=roc)%>%dplyr::select(childteen, roc_basemodel)
-roc_ss=roc_table %>% filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)')%>%mutate(roc_pms=roc)%>%dplyr::select(childteen, roc_pms)
-
+#CI.delong is the confidence interval for each auc: if we sampled 100 times, 95 of the samples would have a auc between those two numbers
+roc_table=global_roc %>% 
+  dplyr::select(childteen, methylation, modeltype, roc, auc, ci_auc_delong)%>%
+  mutate(ci.delong=map_chr(ci_auc_delong, ~paste(round(.x[1], 2), round(.x[3], 2), sep=', ')))
+
+roc_base=roc_table %>% 
+  filter(modeltype=='Base model+No methylation')%>%
+  rename_with(~str_c(., '_basemodel'), .cols=c(roc, auc, ci.delong))%>%
+  dplyr::select(childteen, contains('_basemodel'))
+
+roc_ss=roc_table %>% 
+  filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)')%>%
+  rename_with(~str_c(., '_pms'), .cols=c(roc, auc, ci.delong))%>%
+  dplyr::select(childteen, contains('_pms'))
+
+#confidence_interval_difference_pms|base is the confidence interval for the DIFFERENCE between the two aucs - avoids incorrect error propagation, since the squares of errors add, not the errors themselves (see an explanation here: https://towardsdatascience.com/why-overlapping-confidence-intervals-mean-nothing-about-statistical-significance-48360559900a)
+#delong test is for nested models. the base model is always nested within the base model+any of the methylation measures 
 rct1=roc_table%>% 
   filter(methylation %in% c('globalmethylation', 'cg05575921', 'SSnewbornCT_center_scale', ''))%>%
+  filter(str_detect(modeltype, 'Base model'))%>%
   left_join(roc_base)%>%
-  mutate(roctest_base=map2(roc, roc_basemodel, ~roc.test(.x, .y)))%>%
-  left_join(roc_ss)%>%
-  mutate(roctest_ss=map2(roc, roc_pms, ~roc.test(.x, .y)))%>%
+  mutate(roctest_base=map2(roc, roc_basemodel, ~roc.test(.x, .y, method='delong')))%>%
   mutate(auc=round(auc, 2))%>%
-  mutate(`P value vs polymethylation score`=map_dbl(roctest_ss, ~round(.x$p.value, 4)))%>%
-  mutate(`P value vs base model`=map_dbl(roctest_base, ~round(.x$p.value, 4)))%>%
-  dplyr::select(childteen, modeltype, auc, `P value vs base model`, `P value vs polymethylation score`)
+  mutate(`P value vs base model (DeLong)`=map_dbl(roctest_base, ~round(.x$p.value, 4)))%>%
+  mutate(`Difference (AUC - AUC of base model) (95% CI - DeLong)`= map_chr(roctest_base,
+                   ~paste0(round(.x$estimate[1]-.x$estimate[2], 2), ' (95% CI: (', 
+                           round(.x$conf.int[1], 3), ', ', 
+                           round(.x$conf.int[2], 3), '))')))%>%
+  mutate(`Percentage Difference (AUC - AUC of base model) (95% CI - DeLong)`= map_chr(roctest_base,
+                   ~paste0(round(.x$estimate[1]-.x$estimate[2], 2)*100, ' (95% CI: (', 
+                           round(.x$conf.int[1], 3)*100, ', ', 
+                           round(.x$conf.int[2], 3)*100, '))')))%>%
+  mutate(auc_basemodel=round(auc_basemodel, 2))%>%
+     dplyr::select(childteen, modeltype, auc, auc_basemodel,
+                `Percentage Difference (AUC - AUC of base model) (95% CI - DeLong)`,
+                `Difference (AUC - AUC of base model) (95% CI - DeLong)`, `P value vs base model (DeLong)`)
+
+rct1_export=rct1%>%
+  dplyr::select(childteen, modeltype, auc, auc_basemodel,
+                `Difference (AUC - AUC of base model) (95% CI - DeLong)`, `P value vs base model (DeLong)`)
+
+write.csv(rct1_export, file=file.path(outdir, paste0('SupplementalTable', sup_tab, '.csv')), row.names=F)
 
@@
 ```{r supplemental table auc print, eval=as_supp}
-rct1%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value vs base model', 'P value vs base model + polymethylation score'))
-```
-
-# Supplemental Table `r sup_tab`: Area under the curve when using different coefficient sets to construct polymethylation scores for predicting prenatal maternal smoke exposure.
-With *P* value when comparing to base model only or to model using polymethylation score + base model.
-Base model: Child sex, maternal income to poverty ratio, immune cell proportion, plate, 1st two principal components from PCA on child genetic data, child age.
-
-```{r supplemental table auc other coefficients} 
-rct2=roc_table%>% 
-    filter(str_detect(modeltype, 'smoking|No methylation'))%>%
-    left_join(roc_base)%>%
-    mutate(roctest_base=map2(roc, roc_basemodel, ~roc.test(.x, .y)))%>%
-    left_join(roc_ss)%>%
-    mutate(roctest_ss=map2(roc, roc_pms, ~roc.test(.x, .y)))%>%
-    mutate(auc=round(auc, 4))%>%
-    mutate(`P value vs polymethylation score with cell-type control`=map_dbl(roctest_ss, ~round(.x$p.value, 4)))%>%
-    mutate(`P value vs base model`=map_dbl(roctest_base, ~round(.x$p.value, 4)))%>%
-    dplyr::select(childteen, modeltype, auc, `P value vs base model`, `P value vs polymethylation score with cell-type control`)
-```
-
-```{r supplemental table auc other coefficients print, eval=as_supp}
-rct2%>%kbl(col.names=c('Age', 'Model', 'AUC', 'P value vs base model', 'P value vs base model + polymethylation score with cell-type control'))
-```
+rct1_export%>%dplyr::filter(!str_detect(modeltype,'No methylation'))%>%kbl(col.names=c('Age', 'Model', 'AUC', 'Base model AUC', 'Difference (AUC - AUC of base model) (95% CI - DeLong)', 'P value for difference between AUC and base model AUC (DeLong)'))
+```
+
+```{r}
+
+all_rocs=roc_table %>% 
+  filter(str_detect(modeltype, '19 site|204 site|568 site|cell-type|AHRR|net|global|PedBE'))%>%
+  arrange(childteen, methylation)%>%
+  dplyr::select(childteen, modeltype, auc, ci.delong)
+
+write.csv(all_rocs, file=file.path(outdir, paste0('SupplementalTable', sup_tab, '.csv')))
+sup_tab=sup_tab+1
+```
+
+# Main Figure `r main_fig` forest plot comparison of AUCs for the different polymethylation scores calculated in Fagile Families saliva DNA methylation samples versus in cord and peripheral blood samples of other cohorts
+
+```{r auc comparison across studies main figure 5}
+#change ! to no ! to use model with additional covariates 
+my_aucs=global_roc %>% 
+  filter(str_detect(modeltype, 'Base model') & methylation%in%c('SSnewbornCT_center_scale', 'reese_center_scale','richmond19_center_scale', 'richmond568_center_scale',  'elastic_net_score'))%>%
+  mutate(l.ci=map_dbl(ci_auc_delong, ~.x[[1]]), u.ci=map_dbl(ci_auc_delong, ~.x[[3]]),
+         cohort=paste0(childteen, ' - Fragile Families', ' (saliva)'), 
+         PMID=NA, exposure='self-report', 
+         method=case_when(methylation=='SSnewbornCT_center_scale'~
+                            '6074 site polymethylation score (Joubert newborn cord blood coefficients (cell-type corrected))', 
+                          methylation=='reese_center_scale'~
+                            '28 site polymethylation score (Reese newborn cord blood LASSO coefficients))', 
+                          methylation=='elastic_net_score'~
+                            '204 site polymethylation score (Rauschert adolescent peripheral blood elastic net coefficients)', 
+                          methylation == 'richmond19_center_scale'~
+                          '19 site polymethylation score (Joubert older children peripheral blood coefficients (Richmond))', 
+                          methylation == 'richmond568_center_scale'~
+                            '568 site polymethylation score (Joubert newborn cord blood coefficients (Richmond))'),
+         age=as.numeric(word(cohort, 2)))%>%
+  dplyr::select(cohort, auc, l.ci, u.ci, PMID, exposure, method)
+
+other_aucs=data.frame(auc=c(0.87, 0.88, 0.83, 0.85, 0.85, 0.72, 0.73,
+                            0.85, 0.74, 0.54, 0.79, 0.8, 0.71, 0.73, 
+                            0.8, 0.68, 0.48, 0.75, 0.73, 0.72, 0.73,
+                            0.69, 0.72, 0.865,  0.90), 
+                      l.ci=c(rep(NA, 21), 0.67, 0.69, NA, 0.83), 
+                      u.ci=c(rep(NA, 21), 0.73, 0.76, NA, 0.97), 
+                      cohort=c(rep('Age 17 - Raine Study (peripheral blood)', 7),
+                               rep('Age 16 - NFBC1986 (peripheral blood)', 7), 
+                               rep('Age 31 - NFBC1966 (peripheral blood)', 7), 
+                               'Age 30 - ALSPAC women (peripheral blood)',
+                               'Age 30 - ALSPAC women (peripheral blood)',
+                               'Age 3-5 - SEED children (peripheral blood)', 
+                               #'Age 0 - MoBa newborns (cord blood)',
+                               'Age 0 - MoBa newborns (cord blood)'),
+                      exposure=c(rep('self-report', 21), 'self-report', 'self-report', 'self-report', 'combined'),
+                      method=c(rep(c('204 site polymethylation score (Rauschert adolescent peripheral blood elastic net coefficients)', 'gradient boosting machine', 
+                                     'random forest', 'support vector machine', 
+                               '28 site polymethylation score (Reese newborn cord blood LASSO coefficients))',
+                               '568 site polymethylation score (Joubert newborn cord blood coefficients (Richmond))', 
+                               '19 site polymethylation score (Joubert older children peripheral blood coefficients (Richmond))'), 3), 
+                         '568 site polymethylation score (Joubert newborn cord blood coefficients (Richmond))', 
+                               '19 site polymethylation score (Joubert older children peripheral blood coefficients (Richmond))',
+                        'Support vector machine', 
+                        #'28 site polymethylation score (Reese newborn cord blood LASSO coefficients))',
+                        '28 site polymethylation score (Reese newborn cord blood LASSO coefficients))'),
+                      PMID=c(rep(32930613, 21), 
+                             29860346, 
+                             26610292,
+                             26610292, 
+                             #27323799,
+                             27323799))%>% 
+                      mutate(age=as.numeric(word(cohort, 2, sep=' |-')))
+
+all_aucs=rbind(my_aucs, other_aucs)%>%
+         mutate(cohort=factor(cohort, levels=c('Age 0 - MoBa newborns (cord blood)', 
+                                               'Age 3-5 - SEED children (peripheral blood)', 
+                                               "Age 9 - Fragile Families (saliva)" , 
+                                               "Age 15 - Fragile Families (saliva)", 
+                                               'Age 16 - NFBC1986 (peripheral blood)',
+                                               'Age 17 - Raine Study (peripheral blood)', 
+                                               'Age 30 - ALSPAC women (peripheral blood)', 
+                                               'Age 31 - NFBC1966 (peripheral blood)')))%>%
+  mutate(cohort2=factor(cohort, levels=rev(levels(cohort))))
+
+auc_compare=ggplot(all_aucs%>%filter(str_detect(method, 'score')), 
+                   aes(y=cohort2, x=auc, xmin=l.ci, xmax=u.ci, color=cohort2))+
+  geom_point(position=position_dodge(width=1))+
+  geom_errorbarh(height=0.1, position=position_dodge(width=1))+
+  scale_x_continuous()+
+  #scale_y_discrete(limits = rev(levels(all_aucs$cohort)))+
+  geom_vline(xintercept=0.5, color='red', linetype="dashed", alpha=0.5)+
+  theme(legend.position='none')+ggforce::facet_col(vars(method), scales='free_y', space='free', strip.position='top')
+ggexport(auc_compare, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')), width=9)
+ggexport(auc_compare, filename=file.path(outdir, paste0('Figure', main_fig, '.eps')), width=12)
+
+```
